### 2021-06-20 Exemples and Implementations.Rmd
Ho apportato nuove modifiche

---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

\textbf{Implementation}

In this section we present some practical illustrations of the algorithm discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As we already know, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be the basis for the implementations of other filtering strategies since we think that it is useful to understand the logic of the algorithms and to compare their performances.
The typical observed process for this kind of model is the one presented in Figure XX. We simulated 500 observation imposing $\sigma^{2}=\tau^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Toy example: the true states and the observation sequence", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
library(ggplot2)
q025=function(x){quantile(x,0.025)}
q975=function(x){quantile(x,0.975)}

set.seed(1991)
dlm.sim = function(n,sigma,tau,x0){
  x    = rep(0,n)  #state process
  y    = rep(0,n)  #obs process
  x[1] = rnorm(1,x0,tau)
  y[1] = rnorm(1,x[1],sigma)
  for (t in 2:n){
    x[t] = rnorm(1,x[t-1],tau)
    y[t] = rnorm(1,x[t],sigma)
  }
  return(list(x=x,y=y))
}

n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

sim<-dlm.sim(n,sigma,tau,x0)
y<-sim$y
x<-sim$x

timeframe<-c(1:length(y))
df<-data.frame(timeframe,y,x)

ggplot(df,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```
The Kalman Filter for this model can be easily implemented. Starting from the filtering distribution at period $t-1$, $x_{t-1}|y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, we compute:
\begin{itemize}
\item the one-step-ahead predictive distribution at time $t-1$
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{itemize}
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$. The filtered states follow the observations closely and they provide a good approximation of the true states.
```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Put everything in a data frate (for ggplot)
timeframe<-c(1:length(y))
DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)

#Plots (observetions in black and filtered states in red)
ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```

\textbf{Implementation}
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
```{r}
SISfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  for(t in 1:length(data)){
    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
    w    = w*dnorm(data[t],x,sigma)         #update weight
    xs = rbind(xs,x)
    ws = rbind(ws,w)

    wnorm= w/sum(w)                         #normalized weight
    ESS  = 1/sum(wnorm^2)                   #effective sample size

    ess =rbind(ess,ESS)
  }

  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
N=1000
sis<-SISfun(y,N,m0,C0,tau,sigma)
sisxs<-sis$xs
sisws<-sis$ws
sisess<-sis$ess
sis.data<-data.frame(df,sisxs,sisws,sisess)

ggplot(sis.data,aes(x=timeframe))+
  geom_line(aes(y=sisess))+
    labs(x="Time",
       y="")+
  ggtitle("Effective Sample size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, fig.align='center', fig.cap="SIS Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtervalues<-function(fun){
  xhat<-sapply(1:n,function(i)
    weighted.mean(fun$xs[i,],fun$ws[i,]))
  sdhat<-sapply(1:n,function(i)
    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))

  return(list(mean=xhat,sd=sdhat))

}

Filtplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

Filtplot(df,sis,title="Sequential Importance Sampling Filter")

```
\
\textbf{Particle Filter}
\
With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
\begin{itemize}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{itemize}
In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
N=1000

pf<-PFfun(y,N,m0,C0,tau,sigma)

ESSplot<-function(dataframe,fun,threshold){
if(missing(threshold)){threshold=T}else{}
xs<-fun$xs
ws<-fun$ws
ess<-fun$ess
dt<-data.frame(dataframe,xs,ws,ess)

ggplot(dt,aes(x=timeframe))+
  geom_line(aes(y=ess))+
  geom_line(aes(y=500),col="orange")+
  labs(x="Time",
       y="")+
  ggtitle("Effective Sample Size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
}
ESSplot(df,pf)

```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,pf,"Particle Filter")
```

\textbf{Guided Particle Filter}

Let's consider another toy example. Suppose that we have the following model

```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    xprev<-x
    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)*dnorm(x,xprev,tau)*I(x>0)/dnorm(x,xprev,tau)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

\textbf{}

With reference to the linear gaussian model of section XX, a very basic auxiliary particle sampling technique is 
\begin{itemize}
\item Let $\{(x_{t-1},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t-1}|y_{t-1})$
\end{itemize}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $ where $g(x_{t-1}^{(i)})=E(X_{t}|X_{t-1})$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
\item Set  $w_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
\end{itemize}
The remaining steps are the same of the particle filter already seen in section XX.

```{r}
APFfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    weight = w*dnorm(data[t],x,sigma)
    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
    x1   = rnorm(N,x[k],tau)
    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
    w   = exp(lw)
    w   = w/sum(w)
    ESS  = 1/sum(w^2)

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      x1<-x1[index]
      w<-rep(1/N,N)
    }else{}

    x <- x1
    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)

  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
apf<-APFfun(y,N,m0,C0,tau,sigma)
ESSplot(df,apf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,apf,"Particle Filter")
```
\
\textbf{Liu and West Filter}
\
Consider the toy example of section XX. Let $\psi=(\sigma^{2},\tau^{2})$ be unknown and assign a gamma prior on these parameters
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
or let assign them a uniform prior if we have no knowledge on hyperparameters.
After having drown the hyperparameters indipendently form their priors and having set $w_{0}^{(i)}=N^{-1}$, $i=1,...,N$, and $\hat{\pi}_{0}=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$, for t=1,..T 
\begin{itemize}
\item Compute $\hat{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=Var_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$ set
\begin{align*}
m(\psi^{(i)}) & =a\psi^{(i)}+(1-a)\overline{\psi}\\
v(\psi^{(i)}) & =(1-a^2)\Sigma
\end{align*}
and
\begin{align*}
\alpha(\psi^{(i)}) & =\frac{m(\psi^{(i)})^2}{v(\psi^{(i)})} \\
\beta(\psi^{(i)}) & =\frac{m(\psi^{(i)})}{v(\psi^{(i)})}
\end{align*}
For $k=1,...,N$
\begin{itemize}
\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1^{(i)}}f_{N}(y_{t}|g(x_{t}^{(i)}),m(\psi^{(i)}))$ where for simplicity $g(x_{t}^{(i)})=E(x_{t}|x_{t-1},m(\psi^{(i)}))$
\item Draw $\psi^{(k)} \sim G(\alpha(\psi^{(I_{k})}),\beta(\psi^{(I_{k})}))$
\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^{{2}^{(k)}})$
\item Set $\tilde{w}_{t}^{k}=\frac{f_N(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_N(y_{t}|g(x_{t}^{(I_{k})}),\psi=m(\psi)^{(I_k)})}$
\end{itemize}
\item Normalize the weights
\item Compute the effective sample size ($ESS$)
\item If $ESS<N/2$, resample:
\begin{itemize}
\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P((x_{t},\psi)=(x_{t}^{(i)},\psi^{(i)}))=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{itemize}

\end{itemize}
```{r}
LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif){

  xs     = rnorm(N,m0,sqrt(C0))
  if(unif==T){
  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
  a      = (3*delta-1)/(2*delta)
  h2     = 1-a^2
  parss  = array(0,c(N,2,n))
  xss    = NULL
  ws     = NULL
  ess    = NULL
  w      = rep(1/N,N)
  for (t in 1:length(data)){
    meanV = weighted.mean(pars[,1],w)
    varV  = weighted.mean((pars[,1]-meanV)^2,w)
    meanW = weighted.mean(pars[,2],w)
    varW  = weighted.mean((pars[,2]-meanW)^2,w)

    muV = a*pars[,1]+(1-a)*meanV
    sigma2V = (1-a^2)*varV
    alphaV = muV^2/sigma2V
    betaV = muV/sigma2V

    muW = a*pars[,1]+(1-a)*meanW
    sigma2W = (1-a^2)*varW
    alphaW = muW^2/sigma2W
    betaW = muW/sigma2W

    weight      = w*dnorm(data[t],xs,sqrt(muV))
    k           = sample(1:N,size=N,replace=T,prob=weight)

    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])

    xsprevious<-xs[k]
    xs = rnorm(N,xs[k],sqrt(pars[,2]))

    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
    w           = w/sum(w)
    ESS         = 1/sum(w^2)

    if(ESS<(N/2)){
      index<-sample(N,size=N,replace=T,prob=w)
      xs<-xs[index]
      pars<-pars[index,]
      w<-rep(1/N,N)
    }else{
      xs<-xs
      pars<-pars
    }

    xss         = rbind(xss,xs)
    parss[,,t]  = pars 
    ws          = rbind(ws,w)
    ess         = rbind(ess,ESS)
  }
  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
}
```

```{r echo=FALSE}
delta=0.7
lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
ESSplot(df,lwf)
```

```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtplot(df,lwf,"Liu and West Filter")
```

### 2021-06-20 Filtering strategies new.R

@@ -29,7 +29,7 @@ dlm.sim = function(n,sigma,tau,x0){
 #---------------------
 #y = observed data
 
-DLM<-function(data,sig2,tau2,m0,C0){
+DLM<-function(data,sig2,tau2,m0,C0,smooth){
   n  = length(data)
   m  = rep(0,n)
   C  = rep(0,n)
@@ -214,6 +214,79 @@ PFfun<-function(data,N,m0,C0,tau,sigma){
   }
   return(list(xs=xs,ws=ws,ess=ess))
 }
+
+
+
+#Guided Particle Filter
+#----------------------
+GPFfun<-function(data,N,m0,C0,tau,sigma){
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    xprev<-x
+    x<-rnorm(N,x,tau)
+    w1<-w*dnorm(data[t],x,sigma)*dnorm(x,xprev,tau)*I(x>0)/dnorm(x,xprev,tau)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/2)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+a<-PFfun(y,1000,0,100,1,1)
+s<-sample(a$xs[200,],replace=T,prob=a$ws[200,])
+#
+#
+#Guided Particle Filter
+#----------------------
+GPFfun<-function(data,N,m0,C0,tau,sigma){
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    xprev<-x
+    x<-rnorm(N,x,tau)
+    w1<-w*dnorm(data[t],x,sigma)*dunif(x,xprev,xprev+10)/dnorm(x,xprev,tau)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/2)){
[…]

### 2021-06-20 Filtering strategies new.R
altre modifiche

@@ -285,9 +285,6 @@ GPFfun<-function(data,N,m0,C0,tau,sigma){
 
 
 
-
-
-
 #Auxiliary Particle Filter (ESS-based resampling)
 #------------------------------------------------
 APFfun<-function(data,N,m0,C0,tau,sigma){
@@ -366,11 +363,14 @@ NAMEfun<-function(data,N,m0,C0,tau,sigma,smooth){
 set.seed(8642)
 N      = 10
 
-LWfun<-function(data,N,m0,C0,v0,V0,w0,W0,delta){
+LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta){
   
   xs     = rnorm(N,m0,sqrt(C0))
-  pars   = cbind(rgamma(N,v0,V0),rgamma(N,w0,W0))
-  #pars   = cbind(runif(N,0,10),runif(N,0,10))
+  if(missing(c(alphav,betav,alphaw,betaw))){
+    pars   = cbind(runif(N,0,10),runif(N,0,10))
+  }else{
+  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
+  }
   a      = (3*delta-1)/(2*delta)
   h2     = 1-a^2
   parss  = array(0,c(N,2,n))

### 2021-06-20 script serio.R
Delete script serio.R

@@ -1,199 +1,0 @@
-#@@@@@@@@@@@@@@@@@@@@##
-#SEQUENTIAL MONTECARLO#
-#@@@@@@@@@@@@@@@@@@@@##
-library(ggplot2)
-q025=function(x){quantile(x,0.025)}
-q975=function(x){quantile(x,0.975)}
-
-
-#Generate a random walk plus noise
-#---------------------------------
-#n = sample dim
-#v = obs noise
-#w = state noise
-#x0 = init state
-
-dlm.sim = function(n,sigma,tau,x0){
-  x    = rep(0,n)
-  y        = rep(0,n)
-  x[1] = rnorm(1,x0,tau)
-  y[1] = rnorm(1,x[1],sigma)
-  for (t in 2:n){
-    x[t] = rnorm(1,x[t-1],tau)
-    y[t] = rnorm(1,x[t],sigma)
-  }
-  return(list(x=x,y=y))
-}
-
-#Filtering, using data
-#---------------------
-#y = observed data
-
-DLM = function(y,sigma2,tau2,m0,C0){
-  n = length(y)
-  m = rep(0,n)
-  C = rep(0,n)
-  for (t in 1:n){
-    if (t==1){
-      a = m0
-      R = C0 + tau2
-    }else{
-      a = m[t-1]
-      R = C[t-1] + tau2
-    }
-    A = R/(R+sigma2)
-    m[t] = (1-A)*a + A*y[t]
-    C[t] = A*sigma2
-  }
-  return(list(m=m,C=C))
-}
-
-
-########################################################
-
-#Generate a random walk plus noise
-#---------------------------------
-n=500
-tau2=1
-sigma2=1
-sigma=sqrt(sigma2)
-tau=sqrt(tau2)
-x0=0
-
-sim<-dlm.sim(n,sigma,tau,x0)
-y<-sim$y
-x<-sim$x
-
-#Kalman Filter
-#-------------
-m0=0
-C0=100
-filtval<-DLM(y,sigma2,tau2,m0,C0)
-m<-filtval$m
-C<-filtval$C
-Low = m + qnorm(0.025)*sqrt(C)
-Up = m + qnorm(0.975)*sqrt(C)
-
-#Put everything in a data frate (for ggplot)
-timeframe<-c(1:length(y))
-DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)
-
-#Plots (observetions in black and filtered states in red)
[…]

### 2021-06-20 Exemples and Implementations.Rmd
nuovi files

@@
 ```{r}
-PFfun<-function(data,N,m0,C0,tau,sigma){
+PFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
   xs<-NULL
@@
 
-    if(ESS<(N/2)){
+    if(ESS<(N/r)){
       index<-sample(N,size=N,replace=T,prob=w)
@@
 ```{r}
-GPFfun<-function(data,N,m0,C0,tau,sigma){
+GPFfun<-function(data,N,m0,C0,tau,sigma,r){
+ if(missing(r)){r=2}else{}
    xs<-NULL
@@
 ```{r}
-APFfun<-function(data,N,m0,C0,tau,sigma){
+APFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
   xs<-NULL
@@
 ```{r}
-LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif){
-
+LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
+  if(missing(r)){r=2}else{}
   xs     = rnorm(N,m0,sqrt(C0))
@@
 ```
+
+# COMPARISON (ANOTHER SECTION)
+```{r}
+library(ggpubr)
+library(kableExtra)
+
+Kalmanvarplot<-ggplot(DLM.df,aes(x=timeframe))+
+  geom_line(aes(y=C))+
+  labs(x="Time",
+       y="")+
+  ggtitle("Kalman Filter")+
+  theme_bw()
+
+Varplot<-function(dataframe,fun,title){
+
+Filt<-Filtervalues(fun)
+mean<-Filt$mean
+sd<-Filt$sd
+dataframe<-data.frame(dataframe,mean,sd)
+
+ggplot(dataframe,aes(x=timeframe))+
+geom_line(aes(y=sd))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()
+
+}
+
+Varsisplot<-Varplot(df,sis,"SIS")
+Varpfplot<-Varplot(df,pf,"PF")
+Varapfplot<-Varplot(df,pf,"APF")
+Varlwplot<-Varplot(df,pf,"LW")
+
+ggarrange(Kalmanvarplot,Varsisplot,Varpfplot,Varapfplot,Varlwplot)
+
+```
+
+```{r}
+truex<-x
+estimatedx<-matrix(NA,ncol=6,nrow=6)
+colnames(estimatedx)<-c("N","Threshold","KF","PF","APF","LWF")
+estimatedx[,1]<-c(20,100,500,1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
+RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
+
+i=1
+for(N in c(20,100,500)){
+
+DLM1<-DLM(y,sigma2,tau2,m0,C0)
+pf1<-PFfun(y,N,m0,C0,tau,sigma)
+apf1<-APFfun(y,N,m0,C0,tau,sigma)
+lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
+
+estimatedx[i,3]<-RMSE(truex,DLM1$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
+estimatedx[i,5]<-RMSE(truex,Filtervalues(apf1)$mean)
+estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf1)$mean)
+i=i+1
+
+}
+
+i=4
+for(k in c(2,5,10)){
+  N=1000
+DLM2<-DLM(y,sigma2,tau2,m0,C0)
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
+lwf2<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T,r=k)
+
+estimatedx[i,3]<-RMSE(truex,DLM2$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
+estimatedx[i,5]<-RMSE(truex,Filtervalues(apf2)$mean)
+estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf2)$mean)
+i=i+1
+
+}
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[4:6,], align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
+```
+
+# STOCHASTIC VOLATILITY
+
+Basic specification of Stochastic Volatilty Model
+\begin{align}
+y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
+x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align}
+\
+```{r include=FALSE}
+#############################
+#
+#   y(t) = exp(h(t)/2)*v(t)
+#   x(t) = alpha + beta*x(t-1) + tau*w(t)
+#
+#   v(t) ~ N(0,1)
+#   w(t) ~ N(0,1)
+#   x(0) ~ N(0,100)
+#
+#   PRIOR ON alpha,beta,tau2
+
+# Simulating the data
+# -------------------
+set.seed(12345)
+n     =  50
+alpha = -0.0031
+beta  =  0.9951
+tau2  =  0.0074
+tau   = sqrt(tau2)
+y     = rep(0,n)
+x     = rep(0,n)
+x[1]  = alpha/(1-beta)
+y[1]  = rnorm(1,0,exp(x[1]/2))
+for (t in 2:n){
+  x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
+  y[t] = rnorm(1,0,exp(x[t]/2))
+}
+alpha.true = alpha
+beta.true  = beta
+tau2.true  = tau2
+
+timeframe<-c(1:length(y))
+dfsv<-data.frame(timeframe,y,x)
+```
+
+```{r echo=FALSE}
+n=50
+tau2=1
+sigma2=1
+sigma=sqrt(sigma2)
+tau=sqrt(tau2)
+x0=0
+
+ggplot(dfsv,aes(x=timeframe))+
+  geom_line(aes(y=y, linetype="Observations"))+
+  geom_line(aes(y=x, linetype="True States"))+
+  scale_linetype_manual(name="",
+                      values=c("Observations" = 1,
+                               "True States" = 3))+
+  labs(x="Time",
+       y="")+
+  ggtitle("States and Observations")+
+  theme_bw()+
+  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
+
+\textbf{Particle Filter}
+\
+```{r}
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\textbf{Auxiliary Particle Filter}
+
+```{r}
+SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],0,exp(x/2))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k],tau)
+    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x1<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\textbf{Liu and West}
+
+```{r include=FALSE}
+# Data and prior hyperparameters 
+# ------------------------------
+m0      = 0.0
+C0      = 0.1
+sC0     = sqrt(C0)
+ealpha  = alpha
+valpha  = 0.01
+ebeta    = beta
+vbeta    = 0.01
+nu      = 3
+lambda  = tau2
+
+# Liu and West filter
+# -------------------
+set.seed(8642)
+N      = 1000
+```
+
+```{r}
+SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+xs     = rnorm(N,m0,sqrt(C0))
+pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
+               log(1/rgamma(N,nu/2,nu*lambda/2)))
+delta  = 0.75
+a      = (3*delta-1)/(2*delta)
+h2     = 1-a^2
+parss  = array(0,c(N,3,n))
+xss    = NULL
+ws     = NULL
+ESS    = NULL
+w      = rep(1/N,N)
+for (t in 1:length(data)){
+  mpar        = apply(pars,2,mean)
+  vpar        = var(pars)
+  ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
+  mus         = pars[,1]+pars[,2]*xs 
+  weight      = w*dnorm(data[t],0,exp(mus/2))
+  k           = sample(1:N,size=N,replace=T,prob=weight)
+  ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
+  xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
+  w           = dnorm(y[t],0,exp(xt/2))/dnorm(y[t],0,exp(mus/2))
+  w           = w/sum(w)
+  ESS         = 1/sum(w^2)
+
+  if(ESS<(N/2)){
+    index<-sample(N,size=N,replace=T,prob=w)
+    xs<-xt[index]
+    pars<-ms1[index,]
+    w<-rep(1/N,N)
+  }else{
+    xs<-xt
+    pars<-ms1
+  }
+
+  xss         = rbind(xss,xs)
+  parss[,,t]  = pars 
+  ws          = rbind(ws,w)
+}
+return(list(xs=xss,pars=parss,ws=ws))
+}
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+#Kalman Filter
+#-------------
+m0=0
+C0=100
+filtval<-DLM(y,sigma2,tau2,m0,C0)
+m<-filtval$m
+C<-filtval$C
+Low = m + qnorm(0.025)*sqrt(C)
+Up = m + qnorm(0.975)*sqrt(C)
+
+#Put everything in a data frate (for ggplot)
+timeframe<-c(1:length(y))
+DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)
+
+#Plots (observetions in black and filtered states in red)
+ggplot(DLM.df,aes(x=timeframe))+
+  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
+  geom_ribbon(aes(ymin = Low, ymax = Up),
+              fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("Observations" = "black",
+                               "True States" = "black",
+                               "Filtered States"= "red"))+
+  scale_linetype_manual("",
+                      values=c("Observations" = 1,
+                               "True States" = 3,
+                               "Filtered States"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle("Kalman Filter")+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
+
+```{r include=FALSE}
+N=1000
+svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
+svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
+#svlw<-SVLWfun(y,N=5000000,m0,C0,ealpha=0,valpha=1,ebeta=1,vbeta=1,nu=1,lambda=1)
+```
+
+```{r}
+#Filtplot(dfsv,,"")
+```

### 2021-06-20 Implementatione.Rmd
nuovi files

--- 2021-06-20 Exemples and Implementations.Rmd
@@
 ---
+```{r include=FALSE}
+library(ggplot2)
+q025=function(x){quantile(x,0.025)}
+q975=function(x){quantile(x,0.975)}
+```
 
-\textbf{Implementation}
+\section{Implementation}
 
-In this section we present some practical illustrations of the algorithm discussed.
-For the sake of simplicity we use a random walk plus noise model, i.e.
+In this section we present some practical illustrations of the algorithms previously discussed.
+We start by considering a random walk plus noise model, i.e.
 the most basic form of a linear Gaussian state-space model.
@@
 As we already know, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
-However, this toy example will be the basis for the implementations of other filtering strategies since we think that it is useful to understand the logic of the algorithms and to compare their performances.
-The typical observed process for this kind of model is the one presented in Figure XX. We simulated 500 observation imposing $\sigma^{2}=\tau^{2}=1$.
-```{r echo=FALSE, fig.align='center', fig.cap="Toy example: the true states and the observation sequence", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-library(ggplot2)
-q025=function(x){quantile(x,0.025)}
-q975=function(x){quantile(x,0.975)}
-
-set.seed(1991)
-dlm.sim = function(n,sigma,tau,x0){
-  x    = rep(0,n)  #state process
-  y    = rep(0,n)  #obs process
-  x[1] = rnorm(1,x0,tau)
-  y[1] = rnorm(1,x[1],sigma)
-  for (t in 2:n){
-    x[t] = rnorm(1,x[t-1],tau)
-    y[t] = rnorm(1,x[t],sigma)
-  }
-  return(list(x=x,y=y))
-}
-
-n=50
-tau2=1
-sigma2=1
-sigma=sqrt(sigma2)
-tau=sqrt(tau2)
-x0=0
-
-sim<-dlm.sim(n,sigma,tau,x0)
-y<-sim$y
-x<-sim$x
-
-timeframe<-c(1:length(y))
-df<-data.frame(timeframe,y,x)
-
-ggplot(df,aes(x=timeframe))+
-  geom_line(aes(y=y, linetype="Observations"))+
-  geom_line(aes(y=x, linetype="True States"))+
-  scale_linetype_manual(name="",
-                      values=c("Observations" = 1,
-                               "True States" = 3))+
-  labs(x="Time",
-       y="")+
-  ggtitle("States and Observations")+
-  theme_bw()+
-  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-        legend.background = element_rect(fill = "white", colour = "gray30")) +
-  theme(plot.title = element_text(hjust = 0.5))
-```
+However, this toy example is useful to understand the logic of the algorithms and compare the performances of different filtering strategies.
+\
+\
+\textbf{Kalman Filter}
+\
 The Kalman Filter for this model can be easily implemented. Starting from the filtering distribution at period $t-1$, $x_{t-1}|y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, we compute:
@@
 ```
-In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$.
-The filtered states follow the observations closely and they provide a good approximation of the true states.
-```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-#Kalman Filter
-#-------------
-m0=0
-C0=100
-filtval<-DLM(y,sigma2,tau2,m0,C0)
-m<-filtval$m
-C<-filtval$C
-Low = m + qnorm(0.025)*sqrt(C)
-Up = m + qnorm(0.975)*sqrt(C)
-
-#Put everything in a data frate (for ggplot)
-timeframe<-c(1:length(y))
-DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)
-
-#Plots (observetions in black and filtered states in red)
-ggplot(DLM.df,aes(x=timeframe))+
-  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
-  geom_line(aes(y=x, col="True States", linetype="True States"))+
-  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
-  geom_ribbon(aes(ymin = Low, ymax = Up),
-              fill="red",alpha=0.16) +
-    scale_color_manual("",
-                      values=c("Observations" = "black",
-                               "True States" = "black",
-                               "Filtered States"= "red"))+
-  scale_linetype_manual("",
-                      values=c("Observations" = 1,
-                               "True States" = 3,
-                               "Filtered States"=1))+
-  labs(x="Time",
-       y="")+
-  ggtitle("Kalman Filter")+
-  theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-        legend.background = element_rect(fill = "white", colour = "gray30")) +
-  theme(plot.title = element_text(hjust = 0.5))
-```
-
-\textbf{Implementation}
-With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+\
+\textbf{Sequential Importance Sampling}
+\
+Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
 For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
@@
 
-```{r echo=FALSE}
-N=1000
-sis<-SISfun(y,N,m0,C0,tau,sigma)
-sisxs<-sis$xs
-sisws<-sis$ws
-sisess<-sis$ess
-sis.data<-data.frame(df,sisxs,sisws,sisess)
-
-ggplot(sis.data,aes(x=timeframe))+
-  geom_line(aes(y=sisess))+
-    labs(x="Time",
-       y="")+
-  ggtitle("Effective Sample size")+
-  theme_bw()+
-  theme(plot.title = element_text(hjust = 0.5))
-```
-
-```{r echo=FALSE, fig.align='center', fig.cap="SIS Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-Filtervalues<-function(fun){
-  xhat<-sapply(1:n,function(i)
-    weighted.mean(fun$xs[i,],fun$ws[i,]))
-  sdhat<-sapply(1:n,function(i)
-    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))
-
-  return(list(mean=xhat,sd=sdhat))
-
-}
-
-Filtplot<-function(dataframe,fun,title){
-
-Filt<-Filtervalues(fun)
-mean<-Filt$mean
-sd<-Filt$sd
-dataframe<-data.frame(dataframe,mean,sd)
-
-ggplot(dataframe,aes(x=timeframe))+
-  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
-  geom_line(aes(y=x, col="True States", linetype="True States"))+
-  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
-  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
-              fill="red",alpha=0.16) +
-    scale_color_manual("",
-                      values=c("Observations" = "black",
-                               "True States" = "black",
-                               "Filtered States"= "red"))+
-  scale_linetype_manual("",
-                      values=c("Observations" = 1,
-                               "True States" = 3,
-                               "Filtered States"=1))+
-  labs(x="Time",
-       y="")+
-  ggtitle(title)+
-  theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-        legend.background = element_rect(fill = "white", colour = "gray30")) +
-  theme(plot.title = element_text(hjust = 0.5))
-}
-
-Filtplot(df,sis,title="Sequential Importance Sampling Filter")
-
-```
-\
-\textbf{Particle Filter}
-\
-With reference to the random walk plus noise of section XX, let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \frac{1}{N}\sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
-For $t=1,...,n$ where $n$ is the length of the sample, at any iteration
-\begin{itemize}
-\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
-\end{itemize}
-In addition, when $ESS<ESS_{0}$ \footnote{In our example we fix $ESS_{0}=N/2$, this is an arbitrary common rule of thumb.}, resempling applies
-\begin{itemize}
-\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
-\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
-\end{itemize}
 ```{r}
-PFfun<-function(data,N,m0,C0,tau,sigma){
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    x<-rnorm(N,x,tau)
-    w1<-w*dnorm(data[t],x,sigma)
-
-    w = w1/sum(w1)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/2)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-```
-
-```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-N=1000
-
-pf<-PFfun(y,N,m0,C0,tau,sigma)
-
-ESSplot<-function(dataframe,fun,threshold){
-if(missing(threshold)){threshold=T}else{}
-xs<-fun$xs
-ws<-fun$ws
-ess<-fun$ess
-dt<-data.frame(dataframe,xs,ws,ess)
-
-ggplot(dt,aes(x=timeframe))+
-  geom_line(aes(y=ess))+
-  geom_line(aes(y=500),col="orange")+
-  labs(x="Time",
-       y="")+
-  ggtitle("Effective Sample Size")+
-  theme_bw()+
-  theme(plot.title = element_text(hjust = 0.5))
-}
-ESSplot(df,pf)
 
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-Filtplot(df,pf,"Particle Filter")
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+#Generate a random walk plus noise
+#---------------------------------
+#n = sample dim
+#sigma = obs noise variance
+#tau = state noise variance
+#x0 = init state
+
+dlm.sim = function(n,sigma,tau,x0){
+  x    = rep(0,n)  #state process
+  y    = rep(0,n)  #obs process
+  x[1] = rnorm(1,x0,tau)
+  y[1] = rnorm(1,x[1],sigma)
+  for (t in 2:n){
+    x[t] = rnorm(1,x[t-1],tau)
+    y[t] = rnorm(1,x[t],sigma)
+  }
+  return(list(x=x,y=y))
+}
+n=500
+tau2=1
+sigma2=1
+sigma=sqrt(sigma2)
+tau=sqrt(tau2)
+x0=0
+set.seed(1990)
+sim<-dlm.sim(n,sigma,tau,x0)
+y<-sim$y
+x<-sim$x
+
+#Kalman Filter
+#-------------
+m0=0
+C0=100
+filtval<-DLM(y,sigma2,tau2,m0,C0)
+m<-filtval$m
+C<-filtval$C
+Low = m + qnorm(0.025)*sqrt(C)
+Up = m + qnorm(0.975)*sqrt(C)
+
+#Put everything in a data frate (for ggplot)
+timeframe<-c(1:length(y))
+DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)
+
+#Plots (observetions in black and filtered states in red)
+ggplot(DLM.df,aes(x=timeframe))+
+  geom_line(aes(y=y))+
+  geom_line(aes(y=m),col="red")+
+  geom_ribbon(aes(ymin = Low, ymax = Up),
+              fill="red",alpha=0.16) +
+  labs(x="Periods",
+       y="")+
+  ggtitle("Kalman Filter")+
+  theme_bw()+
+  theme(plot.title = element_text(hjust = 0.5))
+
 ```
-
-\textbf{Guided Particle Filter}
-
-Let's consider another toy example. Suppose that we have the following model
-
-```{r}
-GPFfun<-function(data,N,m0,C0,tau,sigma){
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    xprev<-x
-    x<-rnorm(N,x,tau)
-    w1<-w*dnorm(data[t],x,sigma)*dnorm(x,xprev,tau)*I(x>0)/dnorm(x,xprev,tau)
-
-    w = w1/sum(w1)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/2)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-```
-
-\textbf{}
-
-With reference to the linear gaussian model of section XX, a very basic auxiliary particle sampling technique is 
-\begin{itemize}
-\item Let $\{(x_{t-1},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t-1}|y_{t-1})$
-\end{itemize}
-For $k=1,...,N$
-\begin{itemize}
-\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $ where $g(x_{t-1}^{(i)})=E(X_{t}|X_{t-1})$
-\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
-\item Set  $w_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
-\end{itemize}
-The remaining steps are the same of the particle filter already seen in section XX.
-
-```{r}
-APFfun<-function(data,N,m0,C0,tau,sigma){
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    weight = w*dnorm(data[t],x,sigma)
-    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-    x1   = rnorm(N,x[k],tau)
-    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
-    w   = exp(lw)
-    w   = w/sum(w)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/2)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x1<-x1[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    x <- x1
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-```
-
-```{r echo=FALSE}
-apf<-APFfun(y,N,m0,C0,tau,sigma)
-ESSplot(df,apf)
-```
-
-```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-Filtplot(df,apf,"Particle Filter")
-```
-\
-\textbf{Liu and West Filter}
-\
-Consider the toy example of section XX. Let $\psi=(\sigma^{2},\tau^{2})$ be unknown and assign a gamma prior on these parameters
-\begin{align*}
-\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
-\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
-\end{align*}
-or let assign them a uniform prior if we have no knowledge on hyperparameters.
-After having drown the hyperparameters indipendently form their priors and having set $w_{0}^{(i)}=N^{-1}$, $i=1,...,N$, and $\hat{\pi}_{0}=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$, for t=1,..T
-\begin{itemize}
-\item Compute $\hat{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=Var_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$ set
-\begin{align*}
-m(\psi^{(i)}) & =a\psi^{(i)}+(1-a)\overline{\psi}\\
-v(\psi^{(i)}) & =(1-a^2)\Sigma
-\end{align*}
-and
-\begin{align*}
-\alpha(\psi^{(i)}) & =\frac{m(\psi^{(i)})^2}{v(\psi^{(i)})} \\
-\beta(\psi^{(i)}) & =\frac{m(\psi^{(i)})}{v(\psi^{(i)})}
-\end{align*}
-For $k=1,...,N$
-\begin{itemize}
-\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1^{(i)}}f_{N}(y_{t}|g(x_{t}^{(i)}),m(\psi^{(i)}))$ where for simplicity $g(x_{t}^{(i)})=E(x_{t}|x_{t-1},m(\psi^{(i)}))$
-\item Draw $\psi^{(k)} \sim G(\alpha(\psi^{(I_{k})}),\beta(\psi^{(I_{k})}))$
-\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^{{2}^{(k)}})$
-\item Set $\tilde{w}_{t}^{k}=\frac{f_N(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_N(y_{t}|g(x_{t}^{(I_{k})}),\psi=m(\psi)^{(I_k)})}$
-\end{itemize}
-\item Normalize the weights
-\item Compute the effective sample size ($ESS$)
-\item If $ESS<N/2$, resample:
-\begin{itemize}
-\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P((x_{t},\psi)=(x_{t}^{(i)},\psi^{(i)}))=w_{t}^{(i)},\ \ i=1,...,N$
-\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
-\end{itemize}
-
-\end{itemize}
-```{r}
-LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif){
-
-  xs     = rnorm(N,m0,sqrt(C0))
-  if(unif==T){
-  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
-  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
-  a      = (3*delta-1)/(2*delta)
-  h2     = 1-a^2
-  parss  = array(0,c(N,2,n))
-  xss    = NULL
-  ws     = NULL
-  ess    = NULL
-  w      = rep(1/N,N)
-  for (t in 1:length(data)){
-    meanV = weighted.mean(pars[,1],w)
-    varV  = weighted.mean((pars[,1]-meanV)^2,w)
-    meanW = weighted.mean(pars[,2],w)
-    varW  = weighted.mean((pars[,2]-meanW)^2,w)
-
-    muV = a*pars[,1]+(1-a)*meanV
-    sigma2V = (1-a^2)*varV
-    alphaV = muV^2/sigma2V
-    betaV = muV/sigma2V
-
-    muW = a*pars[,1]+(1-a)*meanW
-    sigma2W = (1-a^2)*varW
-    alphaW = muW^2/sigma2W
-    betaW = muW/sigma2W
-
-    weight      = w*dnorm(data[t],xs,sqrt(muV))
-    k           = sample(1:N,size=N,replace=T,prob=weight)
-
-    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
-    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])
-
-    xsprevious<-xs[k]
-    xs = rnorm(N,xs[k],sqrt(pars[,2]))
-
-    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
-                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
-    w           = w/sum(w)
-    ESS         = 1/sum(w^2)
-
-    if(ESS<(N/2)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      xs<-xs[index]
-      pars<-pars[index,]
-      w<-rep(1/N,N)
-    }else{
-      xs<-xs
-      pars<-pars
-    }
-
-    xss         = rbind(xss,xs)
-    parss[,,t]  = pars 
-    ws          = rbind(ws,w)
-    ess         = rbind(ess,ESS)
-  }
-  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
-}
-```
-
-```{r echo=FALSE}
-delta=0.7
-lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
-ESSplot(df,lwf)
-```
-
-```{r echo=FALSE, fig.align='center', fig.cap="Particle Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-Filtplot(df,lwf,"Liu and West Filter")
-```

### 2021-06-20 comparison.Rmd
nuovi files

---
title: "Comparison"
output: html_document
---

```{r}
library(ggplot2)
library(ggpubr)
library(kableExtra)
Kalmanvarplot<-ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=C))+
  labs(x="Time",
       y="")+
  ggtitle("Kalman Filter")+
  theme_bw()

Varplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sd))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()

}

Varsisplot<-Varplot(dt,sis,"SIS")
Varpfplot<-Varplot(dt,pf,"PF")
Varapfplot<-Varplot(dt,pf,"APF")
Varlwplot<-Varplot(dt,pf,"LW")

ggarrange(Kalmanvarplot,Varsisplot,Varpfplot,Varapfplot,Varlwplot)

```

```{r}
truex<-x
estimatedx<-matrix(NA,ncol=5,nrow=3)
colnames(estimatedx)<-c("N","KF","PF","APF","LWF")
estimatedx[,1]<-c(20,100,500)
RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}

i=1
for(N in c(20,100,500)){

DLM1<-DLM(y,sigma2,tau2,m0,C0)
pf1<-PFfun(y,N,m0,C0,tau,sigma)
apf1<-APFfun(y,N,m0,C0,tau,sigma)
lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)

estimatedx[i,2]<-RMSE(truex,DLM1$m)
estimatedx[i,3]<-RMSE(truex,Filtervalues(pf1)$mean)
estimatedx[i,4]<-RMSE(truex,Filtervalues(apf1)$mean)
estimatedx[i,5]<-RMSE(truex,Filtervalues(lwf1)$mean)
i=i+1

}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
```

### 2021-06-20 Filtering+Smoothing.R

@@ -156,12 +156,15 @@ SISRfun<-function(data,N,m0,C0,tau,sigma,smooth){
   n   = length(data)
   for (t in 1:n){
     x1  = rnorm(N,x,tau)
-    w   = dnorm(data[t],x1,sigma)
-    w   = w/sum(w)
-    ESS = 1/sum(w^2)
-    x   = sample(x1,size=N,replace=T,prob=w)
+    w1   = w*dnorm(data[t],x1,sigma)
+    x   = sample(x1,size=N,replace=T,prob=w1)
+    w   = rep(1/N,N)
+    
     xs  = rbind(xs,x)
-    ws  = rbind(ws,w/sum(w))
+    ws  = rbind(ws,w)
+    
+    wnorm   = w/sum(w)
+    ESS = 1/sum(wnorm^2)
     ess = c(ess,ESS)
   }
   if(smooth){
@@ -184,6 +187,7 @@ SISRfun<-function(data,N,m0,C0,tau,sigma,smooth){
 #-----------------------------
 g<-ESSARfun(y,1000,0,100,1,1)
 s<-SISRfun(y,1000,0,100,1,1)
+
 ESSARfun<-function(data,N,m0,C0,tau,sigma){
   xs<-NULL
   ws<-NULL
@@ -194,15 +198,24 @@ ESSARfun<-function(data,N,m0,C0,tau,sigma){
   ESS  = 1/sum(wnorm^2)  
   for(t in 1:length(data)){
     
+    
+    
+    
+    
+    
+    
+    
+    
+    
     if(ESS<(N/2)){
-      x  =  rnorm(N,draw,tau)
-      w   = dnorm(data[t],x,sigma)
-      wnorm= w/sum(w)
+      x   =  rnorm(N,draw,tau)
+      w   = wnorm*dnorm(data[t],x,sigma)
+      wnorm = w/sum(w)
       draw   = sample(x,size=N,replace=T,prob=wnorm)
       xs = rbind(xs,draw)
     }else{
       x    = rnorm(N,x,tau) 
-      w    = w*dnorm(data[t],x,sigma)
+      w    = wnorm*dnorm(data[t],x,sigma)
       wnorm= w/sum(w)
       draw = sample(x,size=N,replace=T,prob=wnorm)
       xs = rbind(xs,draw)}
@@ -258,7 +271,7 @@ APFfun<-function(data,N,m0,C0,tau,sigma){
[…]

### 2021-06-21 Exemples and Implementations.Rmd
ho inserito examples e forecast

@@
 set.seed(8642)
-N      = 1000
+N      = 10
 ```
@@
 for (t in 1:length(data)){
+
   mpar        = apply(pars,2,mean)
@@
   mus         = pars[,1]+pars[,2]*xs 
-  weight      = w*dnorm(data[t],0,exp(mus/2))
+  weight      = w*dnorm(y[1],0,exp(mus/2))
   k           = sample(1:N,size=N,replace=T,prob=weight)
@@
   xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
-  w           = dnorm(y[t],0,exp(xt/2))/dnorm(y[t],0,exp(mus/2))
+  w           = dnorm(y[1],0,exp(xt/2))/dnorm(y[1],0,exp(mus[k]/2))
   w           = w/sum(w)
@@
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
-#svlw<-SVLWfun(y,N=5000000,m0,C0,ealpha=0,valpha=1,ebeta=1,vbeta=1,nu=1,lambda=1)
+svlw<-SVLWfun(y,N=1000,m0,C0,ealpha=0.12,valpha=0.4,ebeta=1,vbeta=0.3,nu=0.11,lambda=0.1)
 ```
@@
 ```{r}
-#Filtplot(dfsv,,"")
+Filtplot(dfsv,svlw,"")
 ```

### 2021-06-21 Forecast.R

@@ -1,0 +1,402 @@
+################################
+#
+# FUNZIONI CON FORECAST
+#
+###############################
+
+
+
+DLM<-function(data,sig2,tau2,m0,C0){
+  n  = length(data)
+  m  = rep(0,n)
+  C  = rep(0,n)
+  f  = rep(0,n)
+  Q  = rep(0,n)
+  for (t in 1:n){
+    if (t==1){
+      a = m0
+      R = C0 + tau2
+      j = a
+      K = R + sig2
+    }else{
+      a = m[t-1]
+      R = C[t-1] + tau2
+      j = a
+      K = R + sig2
+    }
+    A = R/(R+sig2)
+    m[t] = (1-A)*a + A*y[t]
+    C[t] = A*sig2
+    f[t] = j
+    Q[t] = K
+  }
+  return(list(m=m,C=C,f=f,Q=Q))
+}
+
+PFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ys<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,x,tau)
+    y<-rnorm(N,x,sigma)
+    w1<-w*dnorm(data[t],x,sigma)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    ys = rbind(ys,y)
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+}
+
+APFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    weight = w*dnorm(data[t],x,sigma)
[…]

### 2021-06-21 Exemples and Implementations.Rmd
example and implementation aggiornamenti in arrivo

@@
 lambda  = tau2
-
-# Liu and West filter
-# -------------------
-set.seed(8642)
-N      = 10
 ```
@@
 
-  mpar        = apply(pars,2,mean)
+  #mpar        = apply(pars,2,mean) #alternatively use this 
+  mpar<-c()
+  for(i in 1:3){
+  mpar[i]    = weighted.mean(pars[,i],w)
+  }
+
   vpar        = var(pars)
@@
   xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
-  w           = dnorm(y[1],0,exp(xt/2))/dnorm(y[1],0,exp(mus[k]/2))
+  w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
   w           = w/sum(w)
@@
 ```{r include=FALSE}
-N=1000
+set.seed(12345)
+N=5000
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(1235)
 svlw<-SVLWfun(y,N=1000,m0,C0,ealpha=0.12,valpha=0.4,ebeta=1,vbeta=0.3,nu=0.11,lambda=0.1)
@@
 
-```{r}
-Filtplot(dfsv,svlw,"")
+```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=8, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+plot1<-Filtplot(dfsv,svpf,"Particle Filter")
+plot2<-Filtplot(dfsv,svapf,"Auxiliary Particle Filter")
+plot3<-Filtplot(dfsv,svlw,"Liu e West Filter")
+
+ggarrange(plot1,plot2,plot3)
 ```

### 2021-06-22 Exemples and Implementations.Rmd
Tutto funzionante

@@
   mus         = pars[,1]+pars[,2]*xs 
-  weight      = w*dnorm(y[1],0,exp(mus/2))
+  weight      = w*dnorm(data[t],0,exp(mus/2))
   k           = sample(1:N,size=N,replace=T,prob=weight)
@@
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
-set.seed(1235)
-svlw<-SVLWfun(y,N=1000,m0,C0,ealpha=0.12,valpha=0.4,ebeta=1,vbeta=0.3,nu=0.11,lambda=0.1)
+svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 ```

### 2021-06-22 Forecast.R

@@ -45,7 +45,8 @@ PFfun<-function(data,N,m0,C0,tau,sigma,r){
   for(t in 1:length(data)){
     
     x<-rnorm(N,x,tau)
-    y<-rnorm(N,x,sigma)
+    z<-sample(x,replace=T,prob=w)
+    y<-rnorm(N,z,sigma)
     w1<-w*dnorm(data[t],x,sigma)
     
     w = w1/sum(w1)
@@ -176,7 +177,8 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   for(t in 1:length(data)){
     
     x<-rnorm(N,alpha+beta*x,tau)
-    y<-rnorm(N,0,exp(x/2))
+    z<-sample(x,N,replace = T,prob=w)
+    y<-rnorm(N,0,exp(z/2))
     w1<-w*dnorm(data[t],0,exp(x/2))
     
     w = w1/sum(w1)
@@ -357,10 +359,10 @@ ggplot(DLM.df,aes(x=timeframe))+
 
 
 Forecastvalues<-function(fun){
-  yhat<-sapply(1:n,function(i)
-    weighted.mean(fun$ys[i,],fun$ws[i,]))
-  sdhat<-sapply(1:n,function(i)
-    sqrt(weighted.mean((fun$ys[i,]-yhat[i])^2,fun$ws[i,])))
+  yhat<-sapply(2:n,function(i)
+    weighted.mean(fun$ys[i,],fun$ws[i-1,]))
+  sdhat<-sapply(2:n,function(i)
+    sqrt(weighted.mean((fun$ys[i,]-yhat[i-1])^2,fun$ws[i-1,])))
   
   return(list(mean=yhat,sd=sdhat))
   

### 2021-06-22 Script su SV.R

@@ -1,0 +1,238 @@
+#############################
+#
+#   y(t) = exp(x(t)/2)*v(t)
+#   x(t) = alpha + beta*x(t-1) + tau*w(t)
+#
+#   v(t) ~ N(0,1)
+#   w(t) ~ N(0,1)
+#   x(0) ~ N(0,100)
+#
+#   PRIOR ON alpha,beta,tau2
+
+#Particle Filter
+#---------------
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ys<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,alpha+beta*x,tau)
+    z<-sample(x,replace=T,prob=w)
+    y<-rnorm(N,0,exp(z/2))
+    w1<-w*dnorm(data[t],0,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    ys = rbind(ys,y)
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+}
+
+#Auxiliary Particle Filter
+#-------------------------
+SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    weight = w*dnorm(data[t],0,exp(x/2))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k],tau)
+    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
[…]

### 2021-06-24 Exemples and Implementations.Rmd
Particle Filter aggiunto quello della Petrone

@@
 ---
+```{r include=FALSE}
+library(ggplot2)
+library(ggpubr)
+library(kableExtra)
+```
+
+```{r include=FALSE}
+#Calcolo intevalli di credibilità
+#--------------------------------
+CredInt<-function(fun){
+
+  Filt<-Filtervalues(fun)
+  mean<-Filt$mean
+  sd<-Filt$sd
+  vec<-NULL
+  for(t in 1:length(mean)){
+  if(mean[t]-1.96*sd[t]<truex[t] & truex[t]<mean[t]+1.96*sd[t]){vec[t]<-1}else{vec[t]<-0}
+  }
+  return(mean(vec))
+}
+
+#RMSE
+#----
+RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
+```
 
@@
 x<-sim$x
+truex<-x
 
@@
 
+\textbf{Particle Filter Petroni more efficient}
+```{r}
+PF2fun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    means<-x+(tau/(tau+sigma))*(data[t]-x)
+    x<-rnorm(N,means,importancesd)
+    w1<-w*dnorm(data[t],x,predsd)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\textbf{comparison between Particle Filters}
+
+```{r echo=FALSE}
+pf1<-PFfun(y,1000,0,100,1,1)
+pf2<-PF2fun(y,1000,0,100,1,1)
+plotpf1<-Filtplot(df,pf1,"transition")
+plotpf2<-Filtplot(df,pf2,"more efficient")
+
+ggarrange(plotpf1,plotpf2)
+```
+
+```{r}
+Comparethetwo<-matrix(NA,nrow=2,ncol=2)
+colnames(Comparethetwo)<-c("Credible Interval","RMSE")
+rownames(Comparethetwo)<-c("transition","more efficient")
+Comparethetwo[1,1]<-CredInt(pf1)
+Comparethetwo[2,1]<-CredInt(pf2)
+Comparethetwo[1,2]<-RMSE(truex,Filtervalues(pf1)$mean)
+Comparethetwo[2,2]<-RMSE(truex,Filtervalues(pf2)$mean)
+```
+
+```{r echo=FALSE}
+kbl(Comparethetwo, align="c", longtable = T, booktabs = T, caption="Comparison",digits=3)
+```
+
 \textbf{Guided Particle Filter}
@@
 # COMPARISON (ANOTHER SECTION)
-```{r}
-library(ggpubr)
-library(kableExtra)
+
+```{r echo=FALSE, fig.align='center', fig.cap="Variances", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
 
@@
 
-```{r}
+```{r include=FALSE}
 truex<-x
@@
 estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
-RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
 
@@
 
-```{r echo=FALSE}
+```{r echo=FALSE, fig.align='center', fig.cap="SV", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
 n=50
@@
 #Plots (observetions in black and filtered states in red)
-ggplot(DLM.df,aes(x=timeframe))+
+plot0<-ggplot(DLM.df,aes(x=timeframe))+
   geom_line(aes(y=y, col="Observations", linetype="Observations"))+
@@
                                "Filtered States"=1))+
+    coord_cartesian(ylim=c(-6,6))+
   labs(x="Time",
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=8, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison filter performances", fig.height=8, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
 plot1<-Filtplot(dfsv,svpf,"Particle Filter")
@@
 
-ggarrange(plot1,plot2,plot3)
+ggarrange(plot0,plot1,plot2,plot3)
 ```

### 2021-06-24 Filtering strategies (SIS,SISR,APF).R

@@ -65,7 +65,7 @@ x<-sim$x
 #-------------
 m0=0
 C0=100
-filtval<-DLMsmooth(y,sigma2,tau2,m0,C0)
+filtval<-DLM(y,sigma2,tau2,m0,C0)
 m<-filtval$m
 C<-filtval$C
 Low = m + qnorm(0.025)*sqrt(C)
@@ -219,6 +219,10 @@ ESSARfun<-function(data,N,m0,C0,tau,sigma){
   }
   return(list(xs=xs,ws=ws,ess=ess))
 }
+
+
+
+
 
 #ESS-based adaptive resampling specificato da libro Petris
 #---------------------------------------------------------

### 2021-06-24 Filtering strategies new.R

@@ -71,6 +71,7 @@ x0=0
 sim<-dlm.sim(n,sigma,tau,x0)
 y<-sim$y
 x<-sim$x
+truex<-x
 
 #Kalman Filter
 #-------------
@@ -215,6 +216,41 @@ PFfun<-function(data,N,m0,C0,tau,sigma){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 
+#Petrone more efficient PF
+#-------------------------
+PF2fun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    means<-x+(tau/(tau+sigma))*(data[t]-x)
+    x<-rnorm(N,means,importancesd)
+    w1<-w*dnorm(data[t],x,predsd)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+pf2<-PF2fun(y,1000,0,100,1,1)
 
 
 #Guided Particle Filter
@@ -283,6 +319,44 @@ GPFfun<-function(data,N,m0,C0,tau,sigma){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 
+#
+#Guided Particle Filter example C-P
+#----------------------------------
+GPF2fun<-function(data,N,m0,C0,tau,sigma){
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    xprev<-x
+    
+    
+    if(data[t]==1){x<-rnorm(N,x,tau)*I(x>0)
+    }else if(data[t]==-1){
+      rnorm(N,x,tau)*I(x<0)
[…]

### 2021-06-27 Exemples and Implementations.Rmd
Recenti modifiche con Optimal kernel SV

@@
 #--------------------------------
-CredInt<-function(fun){
+Coverage<-function(fun){
 
@@
 Comparethetwo<-matrix(NA,nrow=2,ncol=2)
-colnames(Comparethetwo)<-c("Credible Interval","RMSE")
+colnames(Comparethetwo)<-c("Coverage","RMSE")
 rownames(Comparethetwo)<-c("transition","more efficient")
-Comparethetwo[1,1]<-CredInt(pf1)
-Comparethetwo[2,1]<-CredInt(pf2)
+Comparethetwo[1,1]<-Coverage(pf1)
+Comparethetwo[2,1]<-Coverage(pf2)
 Comparethetwo[1,2]<-RMSE(truex,Filtervalues(pf1)$mean)

### 2021-06-27 Script su SV.R

@@ -44,6 +44,48 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   return(list(xs=xs,ws=ws,ess=ess,ys=ys))
 }
 
+#
+#-----------------------
+#Particle Filter Optimal Kernel
+#------------------------------
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ys<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*(y^2*exp(-alpha+beta*x)-2),tau^2)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    ys = rbind(ys,y)
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+}
+
+
+
+
+
+
+
+
 #Auxiliary Particle Filter
 #-------------------------
 SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){

### 2021-06-27 Script su SV.R
aggiunto SVPFoptfun

@@ -48,7 +48,7 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 #-----------------------
 #Particle Filter Optimal Kernel
 #------------------------------
-SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
   ys<-NULL
@@ -59,7 +59,7 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   
   for(t in 1:length(data)){
     
-    x<-rnorm(N,alpha+beta*x+(tau^2)/4*(y^2*exp(-alpha+beta*x)-2),tau^2)
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau^2)
     w1<-w*dnorm(data[t],0,exp(x/2))
     
     w = w1/sum(w1)
@@ -217,6 +217,7 @@ N=5000
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
 svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+svopt<-SVPFoptfun(y,N,m0,C0,alpha,beta,tau)
 
 #Filtering Values and Plot
 #-------------------------
@@ -267,10 +268,11 @@ plot2<-Filtplot(dfsv,svapf,"Auxiliary Particle Filter")
 #set.seed(123)
 #svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 plot3<-Filtplot(dfsv,svlw,"Liu e West Filter")
+plot4<-Filtplot(dfsv,svopt,"Opt Ker")
 
 ggarrange(plot1,plot2,plot3)
 
-
+ggarrange(plot1,plot4)
 
 
 

### 2021-06-27 Script su SV.R
Opt kernel 1 e Opt kernel 2

@@ -59,7 +59,7 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   
   for(t in 1:length(data)){
     
-    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau^2)
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
     w1<-w*dnorm(data[t],0,exp(x/2))
     
     w = w1/sum(w1)
@@ -79,6 +79,40 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   return(list(xs=xs,ws=ws,ess=ess,ys=ys))
 }
 
+
+#Particle Filter Optimal Kernel 2
+#------------------------------
+SVPFopt1fun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ys<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    ys = rbind(ys,y)
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+}
 
 
 
@@ -218,6 +252,7 @@ svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
[…]

### 2021-06-28 Exemples and Implementations.Rmd
some modifications applied to the text

@@
 
-In this section we present some practical illustrations of the algorithm discussed.
+In this section we present a practical illustration of the algorithm discussed.
 For the sake of simplicity we use a random walk plus noise model, i.e.
@@
 ```
-In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$.
+In the Figure below we show the filtered states states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and  $\sigma^{2}=\tau^{2}=1$.
 The filtered states follow the observations closely and they provide a good approximation of the true states.

### 2021-06-28 Script su SV.R
Per Bolla

@@ -59,40 +59,6 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   
   for(t in 1:length(data)){
     
-    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))
-    
-    w = w1/sum(w1)
-    ESS  = 1/sum(w^2)
-    
-    if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x[index]
-      w<-rep(1/N,N)
-    }else{}
-    
-    ys = rbind(ys,y)
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-  }
-  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
-}
-
-
-#Particle Filter Optimal Kernel 2
-#------------------------------
-SVPFopt1fun<-function(data,N,m0,C0,alpha,beta,tau,r){
-  if(missing(r)){r=2}else{}
-  xs<-NULL
-  ys<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-  
-  for(t in 1:length(data)){
-    
     xprev<-x
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
     w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
@@ -135,14 +101,14 @@ SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
     weight = w*dnorm(data[t],0,exp(x/2))
     k   = sample(1:N,size=N,replace=TRUE,prob=weight)
     x1   = rnorm(N,alpha+beta*x[k],tau)
-    lw  = dnorm(data[t],0,exp(x/2),log=TRUE)-dnorm(data[t],0,exp(x/2),log=TRUE)
[…]

### 2021-06-29 Script su SV.R
SVPFopt1fun non definito prima. Sia APF che GPF con optimal kernel hanno la funzione SVPFoptfun, che viene sovrascritta

@@ -257,7 +257,8 @@ svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
 svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 svopt<-SVPFoptfun(y,N,m0,C0,alpha,beta,tau)
-svopt1<-SVPFopt1fun(y,N,m0,C0,alpha,beta,tau)
+svopt1<-SVPFopt1fun(y,N,m0,C0,alpha,beta,tau) 
+#SVPFopt1fun non definito prima. Sia APF che GPF con optimal kernel hanno la funzione SVPFoptfun, che viene sovrascritta
 svapfopt1<-SVAPFoptfun(y,N,m0,C0,alpha,beta,tau)
 
 #Filtering Values and Plot

### 2021-06-29 Script su SV.R
correct

@@ -14,7 +14,6 @@
 SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
-  ys<-NULL
   ws<-NULL
   ess<-NULL
   x  = rnorm(N,m0,sqrt(C0))
@@ -23,8 +22,6 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   for(t in 1:length(data)){
     
     x<-rnorm(N,alpha+beta*x,tau)
-    z<-sample(x,replace=T,prob=w)
-    y<-rnorm(N,0,exp(z/2))
     w1<-w*dnorm(data[t],0,exp(x/2))
     
     w = w1/sum(w1)
@@ -36,12 +33,11 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
       w<-rep(1/N,N)
     }else{}
     
-    ys = rbind(ys,y)
     xs = rbind(xs,x)
     ws = rbind(ws,w)
     ess =rbind(ess,ESS)
   }
-  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+  return(list(xs=xs,ws=ws,ess=ess))
 }
 
 #
@@ -51,7 +47,6 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
-  ys<-NULL
   ws<-NULL
   ess<-NULL
   x  = rnorm(N,m0,sqrt(C0))
@@ -72,12 +67,11 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
       w<-rep(1/N,N)
     }else{}
     
-    ys = rbind(ys,y)
     xs = rbind(xs,x)
     ws = rbind(ws,w)
     ess =rbind(ess,ESS)
   }
-  return(list(xs=xs,ws=ws,ess=ess,ys=ys))
+  return(list(xs=xs,ws=ws,ess=ess))
 }
 
 
@@ -257,9 +251,7 @@ svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
[…]

### 2021-06-29 SV real dataset (in fieri).R
Ho iniziato a usare il dataset reale, per ora limitandomi allo S&P500 da giugno 2017 a giugno 2021, 996 osservazioni. C'è un problema di scala, ma i grafici mostrano una distribuzione di filtering incoraggiante

@@ -1,0 +1,310 @@
+#############################
+#
+#   y(t) = exp(x(t)/2)*v(t)
+#   x(t) = alpha + beta*x(t-1) + tau*w(t)
+#
+#   v(t) ~ N(0,1)
+#   w(t) ~ N(0,1)
+#   x(0) ~ N(0,100)
+#
+#   PRIOR ON alpha,beta,tau2
+
+#Particle Filter
+#---------------
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#
+#-----------------------
+#Particle Filter Optimal Kernel
+#------------------------------
+SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
[…]

### 2021-06-30 Script su SV.R
(alex) add SVAPFfun_2, attempt at implementing PETRONE P 218

@@ -74,12 +74,6 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 
-
-
-
-
-
-
 #Auxiliary Particle Filter
 #-------------------------
 SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
@@ -107,6 +101,41 @@ SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
     }else{}
     
     x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+    
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Auxiliary Particle Filter (PETRONE p 218)
+#-------------------------
+SVAPFfun_2<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    x_prev=x
+    x_hat = alpha+beta*x #means of the transition from the previous particle 
+    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
+    for(k in 1:N){
+      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=1)
+      w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
+    }
+    w = w/sum(w)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
     xs = rbind(xs,x)
     ws = rbind(ws,w)
     ess =rbind(ess,ESS)

### 2021-06-30 Script su SV.R
SVAPFfun_2 correction

@@ -124,7 +124,7 @@ SVAPFfun_2<-function(data,N,m0,C0,alpha,beta,tau,r){
     x_hat = alpha+beta*x #means of the transition from the previous particle 
     I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
     for(k in 1:N){
-      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=1)
+      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
       w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
     }
     w = w/sum(w)

### 2021-07-09 Real Dataset Script.R
Dataset unico

@@ -257,11 +257,11 @@ tau   = sqrt(tau2)
 #Real dataset
 
 library(openxlsx)
-realdatasetSPX<-read.xlsx("Dataset 2 x100 17-21.xlsx", sheet = 1)
-y<-realdatasetSPX[,1]
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y<-realdatasetSPX[,2]
 n = length(y)
-volatilitySPX<-read.xlsx("Dataset 3 scale 17-21.xlsx", sheet = 1)
-x<-volatilitySPX[,1]
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x<-volatilitySPX[,3]
 alpha.true = alpha
 beta.true  = beta
 tau2.true  = tau2

### 2021-07-09 SV/SV_relative_ll.R
add SV folder with relative l-likelihood analysis

@@ -1,0 +1,92 @@
+## model
+#   SV
+#   y(t) = a+exp(x(t)/2)*v(t)
+#   x(t) = a.v + b.v*x(t-1) + s.v*w(t)
+#
+#   v(t) ~ N(0,1)
+#   w(t) ~ N(0,1)
+#   x(0) ~ N(0,100)
+#
+#   PRIOR ON alpha,beta,tau2
+## packages
+library(tidyverse)
+## functions
+SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,a.v+b.v*x,s.v)
+    w1<-w*dnorm(data[t],a,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+## data
+y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
+
+n<-dim(y)[1]
+
+# find parameters for constant and stochastic vol models
+mean=sum(y)/n
+res<-(y-mean)^2
+lres<-c(NA,res[1:n-1,1])
+resdf=data.frame(res,lres)
+colnames(resdf)=c("res","lres")
+lm_res<-lm(res~1+lres,data=resdf)
+c_res<-lm_res[[1]]
+sigma_res<-sqrt(mean(summary(lm_res)$residuals^2))
+meansq_res<-sqrt(sum(res)/n)
+
+a<-mean
+s<-meansq_res
+
+a.v<-c_res[1]
+b.v<-c_res[2]
+s.v<-sigma_res
+# calculate cumulative marginal likelihood
+# CV model
+l_cv<-dnorm(y,a,s)
[…]

### 2021-07-09 SV/SV_relative_ll.R
- SV parameters calibrated with AR regression using *log* squared residuals instead of squared residuals (following JKP p 624 - added test with homoskedastic dgp. Relative log likelihood still prefers SV, even though not by as much. WHY??

@@ -40,17 +40,19 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 ## data
-y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
+#y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
+y<-as.matrix(rnorm(1000,mean=1,sd=1)) #test
 
 n<-dim(y)[1]
 
 # find parameters for constant and stochastic vol models
 mean=sum(y)/n
 res<-(y-mean)^2
-lres<-c(NA,res[1:n-1,1])
-resdf=data.frame(res,lres)
-colnames(resdf)=c("res","lres")
-lm_res<-lm(res~1+lres,data=resdf)
+lres<-log(res)
+l.lres<-c(NA,lres[1:n-1,1])
+resdf=data.frame(lres,l.lres)
+colnames(resdf)=c("lres","l.lres")
+lm_res<-lm(lres~1+l.lres,data=resdf)
 c_res<-lm_res[[1]]
 sigma_res<-sqrt(mean(summary(lm_res)$residuals^2))
 meansq_res<-sqrt(sum(res)/n)
@@ -74,7 +76,7 @@ N=1000
 pf<-SVPFfun(data=y,N=N,m0=0,C0=100,a=a,a.v=a.v,b.v=b.v,s.v=s.v,r=4)
 xs<-pf$xs
 ws<-pf$ws
-l_sv<-matrix(0,996,1)
+l_sv<-matrix(0,n,1)
 for (i in 1:n){
   for (j in 1:N){
   l_sv[i]=l_sv[i]+dnorm(y[i],a,exp(xs[i,j]/2))*ws[i,j]
@@ -87,6 +89,7 @@ for (i in 2:n){
 }
 # Difference
 cll_diff=cll_cv-cll_sv
-cll_diff.df=data.frame(1:996,cll_diff)
+ll_diff=ll_cv-ll_sv
+cll_diff.df=data.frame(1:n,cll_diff)
 colnames(cll_diff.df)<-c('t','cll_diff')
 ggplot(data = cll_diff.df, aes(x = t, y = cll_diff))+geom_line()

### 2021-07-10 SV/SV_relative_ll.R
SV - predictive state density p(x_t+1|y_1:t) now used, correcting mistake in previous version. Test on homoscedastic simulated series now gives expected result (CV does better).

@@ -22,6 +22,7 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
   for(t in 1:length(data)){
     
     x<-rnorm(N,a.v+b.v*x,s.v)
+    xp.1<-x
     w1<-w*dnorm(data[t],a,exp(x/2))
     
     w = w1/sum(w1)
@@ -37,10 +38,41 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
     ws = rbind(ws,w)
     ess =rbind(ess,ESS)
   }
-  return(list(xs=xs,ws=ws,ess=ess))
+  return(list(xs=xs,ws=ws,ess=ess,xp.1=xp.1))
 }
+# SVPFfun.f<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
+#   if(missing(r)){r=2}else{}
+#   xs<-NULL
+#   xf<-NULL
+#   ws<-NULL
+#   ess<-NULL
+#   x  = rnorm(N,m0,sqrt(C0))
+#   w  = rep(1/N,N)
+#   
+#   for(t in 1:length(data)){
+#     
+#     x<-rnorm(N,a.v+b.v*x,s.v)
+#     x.f1<-x
+#     w1<-w*dnorm(data[t],a,exp(x/2))
+#     x.f<-rnorm(N,a.v+b.v*x,s.v)
+#     w = w1/sum(w1)
+#     ESS  = 1/sum(w^2)
+#     
+#     if(ESS<(N/r)){
+#       index<-sample(N,size=N,replace=T,prob=w)
+#       x<-x[index]
+#       w<-rep(1/N,N)
+#     }else{}
+#     
+#     xs = rbind(xs,x)
+#     xs.f = rbind(xs.f,x.f) 
+#     ws = rbind(ws,w)
+#     ess =rbind(ess,ESS)
+#   }
+#   return(list(xs=xs,xs.f=xs.f,ws=ws,ess=ess,x.f1=x.f1))
+# }
 ## data
-#y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
+# y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
 y<-as.matrix(rnorm(1000,mean=1,sd=1)) #test
 
[…]

### 2021-07-11 script di prova.R
fatto modifica

@@ -4,6 +4,8 @@ use_git_config(user.name = "edoardo-marcelli",
                user.email = "edoardomarcelli2@gmail.com")
 use_git_config(user.name = "michelebolla", 
                user.email = "michele.bolla@studbocconi.it")
+use_git_config(user.name = "RobertoColarieti", 
+               user.email = "roberto.colarieti1998@gmail.com")
 
 
 
@@ -101,3 +103,4 @@ ggplot(DLM.df,aes(x=timeframe))+
 #ciao sono Michele
 # test (Alexei)
 # alexei teast from local
+#ciao 

### 2021-07-11 script di prova.R
Quanto è bello Robbi

@@ -6,7 +6,8 @@ use_git_config(user.name = "michelebolla",
                user.email = "michele.bolla@studbocconi.it")
 use_git_config(user.name = "RobertoColarieti", 
                user.email = "roberto.colarieti1998@gmail.com")
-
+use_git_config(user.name = "gabrieleromano111", 
+               user.email = "gabrieleromano111@gmail.com")
 
 
 #@@@@@@@@@@@@@@@@@@@@##
@@ -104,3 +105,4 @@ ggplot(DLM.df,aes(x=timeframe))+
 # test (Alexei)
 # alexei teast from local
 #ciao 
+#Robbi

### 2021-07-11 Exemples and Implementations.Rmd
Two drafts update

--- 2021-06-27 Exemples and Implementations.Rmd
@@
 ---
+header-includes: 
+- \newcommand{\R}{\textnormal{\sffamily\bfseries R}}}
 title: "R Notebook"
@@
 ```{r include=FALSE}
+#Quantiles functions
+#-------------------
+q025=function(x){quantile(x,0.025)}
+q975=function(x){quantile(x,0.975)}
+
 #Calcolo intevalli di credibilità
@@
 
-In this section we present some practical illustrations of the algorithm discussed.
+In this section we present a practical illustration of the algorithms discussed.
 For the sake of simplicity we use a random walk plus noise model, i.e.
@@
 \end{align}
-As we already know, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
-However, this toy example will be the basis for the implementations of other filtering strategies since we think that it is useful to understand the logic of the algorithms and to compare their performances.
-The typical observed process for this kind of model is the one presented in Figure XX. We simulated 500 observation imposing $\sigma^{2}=\tau^{2}=1$.
-```{r echo=FALSE, fig.align='center', fig.cap="Toy example: the true states and the observation sequence", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
-library(ggplot2)
-q025=function(x){quantile(x,0.025)}
-q975=function(x){quantile(x,0.975)}
+As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
+However, this toy example will be used also to illustrate more involving filtering strategies described in this work.
+We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
+\
+The filtering strategy is applied to 50 simulated data.
+Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
+```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
 
@@
   geom_line(aes(y=x, linetype="True States"))+
+  coord_cartesian(ylim = c(-4, +12)) +
   scale_linetype_manual(name="",
@@
        y="")+
-  ggtitle("States and Observations")+
+  #ggtitle("States and Observations")+
   theme_bw()+
@@
 ```
-The Kalman Filter for this model can be easily implemented. Starting from the filtering distribution at period $t-1$, $x_{t-1}|y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, we compute:
+The Kalman Filter for this model is implemented as described below.
 \begin{itemize}
-\item the one-step-ahead predictive distribution at time $t-1$
+\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
+\item For $t=1,..N$
+\begin{enumerate}
+\item Compute the one-step-ahead state predictive distribution at time $t-1$
 \begin{align*}
@@
 \end{align*}
-\item the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$
+\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e.
+the product of the one-step-ahead state predictive distribution and the likelihood
 \begin{align*}
@@
 \end{align*}
+\end{enumerate}
 \end{itemize}
+These steps are summarize in the 
 ```{r}
@@
 i=1
-for(N in c(20,100,500)){
+for(N in c(100,1000,10000)){
 

### 2021-07-11 draft implementation.Rmd
Two drafts update

---
header-includes: 
- \newcommand{\R}{\textnormal{\sffamily\bfseries R}}
- \def\code#1{\texttt{#1}}
- \newtheorem{algorithm}{Algorithm}[section]
title: "R Notebook"

output:
  pdf_document: 
   latex_engine: xelatex
   fig_width: 5
   fig_height: 3.5
   fig_caption: true
   extra_dependencies: ["float"]
   number_sections: true 
---
```{r include=FALSE}
library(ggplot2)
library(ggpubr)
library(kableExtra)
```

```{r include=FALSE}
#Quantiles functions
#-------------------
q025=function(x){quantile(x,0.025)}
q975=function(x){quantile(x,0.975)}

#Calcolo intevalli di credibilità
#--------------------------------
Coverage<-function(fun){

  Filt<-Filtervalues(fun)
  mean<-Filt$mean
  sd<-Filt$sd
  vec<-NULL
  for(t in 1:length(mean)){
  if(mean[t]-1.96*sd[t]<truex[t] & truex[t]<mean[t]+1.96*sd[t]){vec[t]<-1}else{vec[t]<-0}
  }
  return(mean(vec))
}

#RMSE
#----
RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
```

\textbf{Implementation}

In this section we present a practical illustration of the algorithms discussed. For the sake of simplicity we use a random walk plus noise model, i.e. the most basic form of a linear Gaussian state-space model. 
\begin{align}
y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
x_{0} & \sim N(m_{0},C_{0})
\end{align}
As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter. However, this toy example will be used also to illustrate more involving filtering strategies described in this work. We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models. \
The filtering strategy is applied to 50 simulated data. Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}

set.seed(1991)
dlm.sim = function(n,sigma,tau,x0){
  x    = rep(0,n)  #state process
  y    = rep(0,n)  #obs process
  x[1] = rnorm(1,x0,tau)
  y[1] = rnorm(1,x[1],sigma)
  for (t in 2:n){
    x[t] = rnorm(1,x[t-1],tau)
    y[t] = rnorm(1,x[t],sigma)
  }
  return(list(x=x,y=y))
}

n=50
tau2=1
sigma2=1
sigma=sqrt(sigma2)
tau=sqrt(tau2)
x0=0

sim<-dlm.sim(n,sigma,tau,x0)
y<-sim$y
x<-sim$x
truex<-x

timeframe<-c(1:length(y))
df<-data.frame(timeframe,y,x)

ggplot(df,aes(x=timeframe))+
  geom_line(aes(y=y, linetype="Observations"))+
  geom_line(aes(y=x, linetype="True States"))+
  coord_cartesian(ylim = c(-4, +12)) +
  scale_linetype_manual(name="",
                      values=c("Observations" = 1,
                               "True States" = 3))+
  labs(x="Time",
       y="")+
  #ggtitle("States and Observations")+
  theme_bw()+
  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```
The Kalman Filter for this model is implemented as described below.
\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
\begin{align*}
x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
a_{t} & = m_{t-1}\\
R_{t} & = C_{t-1}+\tau^2
\end{align*}
\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e. the product of the one-step-ahead state predictive distribution and the likelihood
\begin{align*}
x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
\end{align*}
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{DLM} function implement in \R replicate this steps.
\
\hrule
\hrule
\code{DLM(data,sig2,tau2,m0,C0)}
\hrule
\textbf{Arguments}
\
\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
\code{sig2} \ \ the variance $\sigma^{2}$ in Equation (1)
\
\code{tau2} \ \ the variance $\tau^{2}$ in Equation (2)
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\hrule
\hrule
```{r}
DLM<-function(data,sig2,tau2,m0,C0){
  n  = length(data)
  m  = rep(0,n)
  C  = rep(0,n)
  for (t in 1:n){
    if (t==1){
      a = m0
      R = C0 + tau2
    }else{
      a = m[t-1]
      R = C[t-1] + tau2
    }
    A = R/(R+sig2)
    m[t] = (1-A)*a + A*y[t]
    C[t] = A*sig2
  }
  return(list(m=m,C=C))
}
```
In Figure XX below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
#Kalman Filter
#-------------
m0=0
C0=100
filtval<-DLM(y,sigma2,tau2,m0,C0)
m<-filtval$m
C<-filtval$C
Low = m + qnorm(0.025)*sqrt(C)
Up = m + qnorm(0.975)*sqrt(C)

#Put everything in a data frate (for ggplot)
timeframe<-c(1:length(y))
DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)

#Plots (observetions in black and filtered states in red)
ggplot(DLM.df,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = Low, ymax = Up),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  #ggtitle("Kalman Filter")+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
```
The discussion of Kalman Filter will continue in Section XX where we compare it to the Particle Filter.

\
\textbf{Implementation}
Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm. The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm. We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo. 
\begin{algorithm} SIS filter for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{itemize}
\end{algorithm}
Our \code{SIS} function implemented in \R replicate this steps.
\
\hrule
\hrule
\code{SIS(data,N,m0,C0,tau,sigma)}
\hrule
\textbf{Arguments}
\
\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
\code{N} \ \ number of particles generated at each step
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\
\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
\
\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
\hrule
\hrule
```{r}
SISfun<-function(data,N,m0,C0,tau,sigma){
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)
  for(t in 1:length(data)){
    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
    w    = w*dnorm(data[t],x,sigma)         #update weight
    xs = rbind(xs,x)
    ws = rbind(ws,w)

    wnorm= w/sum(w)                         #normalized weight
    ESS  = 1/sum(wnorm^2)                   #effective sample size

    ess =rbind(ess,ESS)
  }

  return(list(xs=xs,ws=ws,ess=ess))
}
```
We have already discussed the reasons why the SIS algorithm does not provide a good strategy in the filtering problem. We provide a graphical intuition of what happens when we use such filtering strategy on a simulated dataset. We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$. The results shown in the following two plots shows a clear degeneration of the effective sample size and bad fit of filtered states with respect to the true values.
```{r echo=FALSE}
N=1000
sis<-SISfun(y,N,m0,C0,tau,sigma)
sisxs<-sis$xs
sisws<-sis$ws
sisess<-sis$ess
sis.data<-data.frame(df,sisxs,sisws,sisess)

ESS.SIS<-ggplot(sis.data,aes(x=timeframe))+
  geom_line(aes(y=sisess))+
    labs(x="Time",
       y="")+
  ggtitle("Effective Sample size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

```{r echo=FALSE, fig.align='center', fig.cap="a) SIS Filtered States with credible interval (in red). b) Effective sample size.", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
Filtervalues<-function(fun){
  xhat<-sapply(1:n,function(i)
    weighted.mean(fun$xs[i,],fun$ws[i,]))
  sdhat<-sapply(1:n,function(i)
    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))

  return(list(mean=xhat,sd=sdhat))

}

Filtplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
              fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("Observations" = "black",
                               "True States" = "black",
                               "Filtered States"= "red"))+
  scale_linetype_manual("",
                      values=c("Observations" = 1,
                               "True States" = 3,
                               "Filtered States"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

PlotSIS<-Filtplot(df,sis,title="")

ggarrange(PlotSIS,ESS.SIS,labels=c("(a)","(b)"))
```

\textbf{Implementation}
In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX. The overall strategy replicate the SIS filter with the addition of a ESS-based resampling step that applies when the effective sample size is smaller than a predetermined threshold opportunely chosen (in our example we decide to follow a common rule of thumb consisting in setting the threshold at $N/2$). The steps are presented in the following algorithm.
\begin{algorithm} BPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item For $t=1,...,n$:
\begin{enumerate}
\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{enumerate}
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{itemize}
\end{algorithm}
These steps are resumed in our \code{PF} function.
\
\hrule
\hrule
\code{PF(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}
\
\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
\code{N} \ \ number of particles generated at each step
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\
\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
\
\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
\
\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
\hrule
\hrule
```{r}
PFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    x<-rnorm(N,x,tau)
    w1<-w*dnorm(data[t],x,sigma)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```
The estimated states of the Boostrap Particle Filter togheter with the effective sample size are shown in Figure XX. We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$. Notice how the resampling step allows the effective sample size not to drop, improving results.
```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
N=1000

pf<-PFfun(y,N,m0,C0,tau,sigma)

ESSplot<-function(dataframe,fun,threshold){
if(missing(threshold)){threshold=T}else{}
xs<-fun$xs
ws<-fun$ws
ess<-fun$ess
dt<-data.frame(dataframe,xs,ws,ess)

ggplot(dt,aes(x=timeframe))+
  geom_line(aes(y=ess))+
  geom_line(aes(y=500),col="orange")+
  labs(x="Time",
       y="")+
  ggtitle("Effective Sample Size")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
}
ESS.PF<-ESSplot(df,pf)

```

```{r echo=FALSE, fig.align='center', fig.cap="a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
FILT.PF <- Filtplot(df,pf,"")
ggarrange(FILT.PF,ESS.PF,labels=c("(a)","(b)"))
```

Since the Random Walk plus Noise model allows for closed form solutions, we want to compare the results of Boostrap Particle Filter(BPF) with Kalman Filter(KF). As we can see from figure XX, when the number of particles generated at any interaction increases the results for both the estimated mean and the variance tend to converge [[SPIEGARE PERCHE']]. Moreover, comparing the Root Mean Square Errors of the filtered states with respect to the true state values, when the sample size increases the BPF decreases and, for large N, it reaches the accuracy of the Kalman Filter.

```{r include=FALSE}
truex<-x
estimatedx<-matrix(NA,ncol=4,nrow=6)
colnames(estimatedx)<-c("N","Threshold","KF","BPF")
estimatedx[,1]<-c(100,1000,10000,1000,1000,1000)
estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)

i=1
for(N in c(100,1000,10000)){

DLM1<-DLM(y,sigma2,tau2,m0,C0)
pf1<-PFfun(y,N,m0,C0,tau,sigma)

estimatedx[i,3]<-RMSE(truex,DLM1$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
i=i+1

}

i=4
for(k in c(2,5,10)){
  N=1000
DLM2<-DLM(y,sigma2,tau2,m0,C0)
pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)

estimatedx[i,3]<-RMSE(truex,DLM2$m)
estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
i=i+1

}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
```

```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='ht', fig.width=10, message=FALSE, warning=FALSE, results=F}
N1=100
N2=1000
N3=10000

Combined<-data.frame(DLM.df,df)
pf1<-PFfun(y,N1,m0,C0,tau,sigma)
pf2<-PFfun(y,N2,m0,C0,tau,sigma)
pf3<-PFfun(y,N3,m0,C0,tau,sigma)

Compareplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=m, col="KF", linetype="KF"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean, col="BPF", linetype="BPF"))+
  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
   #           fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("KF" = "red",
                               "True States" = "black",
                               "BPF"= "blue"))+
  scale_linetype_manual("",
                      values=c("KF" = 1,
                               "True States" = 3,
                               "BPF"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

p1<-Compareplot(Combined,pf1,"Filtered States")
p2<-Compareplot(Combined,pf2,"")
p3<-Compareplot(Combined,pf3,"")

CompareVarplot<-function(dataframe,fun,title){

Filt<-Filtervalues(fun)
mean<-Filt$mean
sd<-Filt$sd
dataframe<-data.frame(dataframe,mean,sd)

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sqrt(C), col="KF"))+
geom_line(aes(y=sd,col="BPF"))+
  scale_color_manual("",
                      values=c("KF" = "red",
                               "BPF"= "blue"))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30"))+
  theme(plot.title = element_text(hjust = 0.5))

}

c1<-CompareVarplot(Combined,pf1,"Standard Deviation")
c2<-CompareVarplot(Combined,pf2,"")
c3<-CompareVarplot(Combined,pf3,"")

ggarrange(p1,c1,p2,c2,p3,c3,nrow=3,ncol=2,labels=c("N=100","","N=1000","","N=10000",""))
```
\
\textbf{Implementation}
\
The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section XX can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$. In the Normal model we are considering, the optimal proposal will be a Normal density as well with mean and variance given by 
\begin{align*}
\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
\end{align*}
On the other hand, the incremental weights, using this importance transition density, are proportional to the conditional density of $y_{t}$ given $x_{t-1}=x_{t-1}^{(i)}$ , i.e $N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})$, evaluated at $y_{t}$. In other words the algorithm implemented in \R is:
\begin{algorithm} GPF for Random Walk plus Noise Model
\begin{itemize}
\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$. In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
\item Compute $\sigma_{opt}^{2}$ 
\item For $t=1,...,n$:
\begin{enumerate}
\item Compute $\mu_{opt}$
\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
\end{enumerate}
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
\end{itemize}
\end{algorithm}
The \code{GPF} function resume this passages.
\
\hrule
\hrule
\code{GPF(data,N,m0,C0,tau,sigma,r)}
\hrule
\textbf{Arguments}
\
\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
\code{N} \ \ number of particles generated at each step
\
\code{m0} \ \ central value of the normal prior state distribution
\
\code{C0} \ \ variance of the normal prior state distribution
\
\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
\
\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
\
\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
\hrule
\hrule
```{r}
GPFfun<-function(data,N,m0,C0,tau,sigma,r){
  if(missing(r)){r=2}else{}
  xs<-NULL
  ws<-NULL
  ess<-NULL
  x  = rnorm(N,m0,sqrt(C0))
  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
  predsd <- sqrt(sigma+tau)
  w  = rep(1/N,N)

  for(t in 1:length(data)){

    means<-x+(tau/(tau+sigma))*(data[t]-x)
    x<-rnorm(N,means,importancesd)
    w1<-w*dnorm(data[t],x,predsd)

    w = w1/sum(w1)
    ESS  = 1/sum(w^2)

    if(ESS<(N/r)){
      index<-sample(N,size=N,replace=T,prob=w)
      x<-x[index]
      w<-rep(1/N,N)
    }else{}

    xs = rbind(xs,x)
    ws = rbind(ws,w)
    ess =rbind(ess,ESS)
  }
  return(list(xs=xs,ws=ws,ess=ess))
}
```
Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF). As we can see from the Figure XX, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE. Moreover, also the variance is better, suggesting for higher precision.

```{r include=FALSE}
truex<-x
estimatedx<-matrix(NA,ncol=4,nrow=3)
colnames(estimatedx)<-c("N","Threshold","BPF","GPF")
estimatedx[,1]<-c(1000,1000,1000)
estimatedx[,2]<-c(0.5,0.25,0.1)

i=1
for(k in c(2,5,10)){
  N=1000

pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
gpf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)

estimatedx[i,3]<-RMSE(truex,Filtervalues(pf2)$mean)
estimatedx[i,4]<-RMSE(truex,Filtervalues(gpf2)$mean)
i=i+1

}

```

```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
kbl(estimatedx[,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
```

```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
N2=1000

pf2<-PFfun(y,N2,m0,C0,tau,sigma)
gpf2<-GPFfun(y,N2,m0,C0,tau,sigma)

Comparetwofunplot<-function(dataframe,fun1,fun2,title){

Filt1<-Filtervalues(fun1)
mean1<-Filt1$mean
sd1<-Filt1$sd
dataframe<-data.frame(dataframe,mean1,sd1)

Filt2<-Filtervalues(fun2)
mean2<-Filt2$mean
sd2<-Filt2$sd
dataframe<-data.frame(dataframe,mean2,sd2)

ggplot(dataframe,aes(x=timeframe))+
  geom_line(aes(y=mean1, col="GPF", linetype="GPF"))+
  geom_line(aes(y=x, col="True States", linetype="True States"))+
  geom_line(aes(y=mean2, col="BPF", linetype="BPF"))+
  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
   #           fill="red",alpha=0.16) +
    scale_color_manual("",
                      values=c("GPF" = "green",
                               "True States" = "black",
                               "BPF"= "blue"))+
  scale_linetype_manual("",
                      values=c("GPF" = 1,
                               "True States" = 3,
                               "BPF"=1))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30")) +
  theme(plot.title = element_text(hjust = 0.5))
}

p2<-Comparetwofunplot(df,pf2,gpf2,"Filtered States")

ComparetwoVarplot<-function(dataframe,fun1,fun2,title){

Filt1<-Filtervalues(fun1)
mean1<-Filt1$mean
sd1<-Filt1$sd
dataframe<-data.frame(dataframe,mean1,sd1)

Filt2<-Filtervalues(fun2)
mean2<-Filt2$mean
sd2<-Filt2$sd
dataframe<-data.frame(dataframe,mean2,sd2)

ggplot(dataframe,aes(x=timeframe))+
geom_line(aes(y=sd2, col="GPF"))+
geom_line(aes(y=sd1,col="BPF"))+
  scale_color_manual("",
                      values=c("GPF" = "green",
                               "BPF"= "blue"))+
  labs(x="Time",
       y="")+
  ggtitle(title)+
  theme_bw()+
    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
        legend.background = element_rect(fill = "white", colour = "gray30"))+
  theme(plot.title = element_text(hjust = 0.5))

}

c2<-ComparetwoVarplot(Combined,pf2,gpf2,"Standard Deviation")

ggarrange(p2,c2)
```

\textbf{Implementation}

### 2021-07-11 final draft.Rmd
Two drafts update

---
header-includes:
- \usepackage{amsmath}
- \usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
- \usepackage{graphicx}
- \usepackage{amssymb}
- \usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
- \usepackage{hyperref} %set the hyperlink to cite sections, equations...
- \urlstyle{same}
- \usepackage{comment}
- \usepackage{enumitem}
- \usepackage{ntheorem}
- \usepackage{lipsum}
- \usepackage{amssymb}
- \usepackage{xcolor}
- \theoremstyle{break}
- \newtheorem{thm}{Theorem}% theorem counter resets every \subsection
- \renewcommand{\thethm}{\arabic{thm}}
- \newtheorem{proposition}{Proposition}
- \newtheorem{assumption}{Assumption}
- \newtheorem{lemma}{Lemma}
- \theoremstyle{nonumberplain}
- \newtheorem{proof*}{Proof.}
- \newtheorem{definition}{Definition}
- \renewcommand{\P}{\mathbb{P}}
- \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
- \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
- \setcounter{MaxMatrixCols}{10}
- \setlength{\oddsidemargin}{.0in}
- \setlength{\textwidth}{6 in}
- \setlength{\topmargin}{-.1in}
- \setlength{\textheight}{8.1in}
- \renewcommand{\baselinestretch}{1.5}
- \usepackage{amssymb}
- \newcommand{\E}{\mathbb E}
- \newcommand{\Q}{\mathbb Q}
- \newcommand{\M}{\mathbb M}
- \newcommand{\R}{\mathbb R}
- \newcommand{\B}{\mathcal B}
- \newcommand{\X}{\mathcal X}
- \newcommand{\Pd}{\mathbb P}
- \def\code#1{\texttt{#1}}
- \newtheorem{algorithm}{Algorithm}[section]

title: "Final draft"

documentclass: book

output:
  pdf_document: 
   latex_engine: xelatex
   fig_width: 5
   fig_height: 3.5
   fig_caption: true
   extra_dependencies: ["float"]
   number_sections: true

---

\chapter*{Introduction}

\chapter{Markov Processes and State-Space Models}
\section{Notation}
Let us fix the basic notation that will be employed henceforth. Upper case letters, e.g. $X_t, Y_t,$ denote random variables, while lower case letters 
denote the realizations. Finite sequences are made compact using the semi-colon notation, e.g. $x_{0:t}=(x_0, x_1, \dots, x_t)$ for some $t \geq 0.$ The sets of outcomes are denoted by upper case calligraphic letters, e.g. $\mathcal{X}, \mathcal{Y}.$  The $\sigma$-algebra
of a set $\mathcal{X}$ reads $\mathcal{B}(\mathcal{X}).$ The pair $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ is a measurable space.
The densities are represented with lower case cursive letters. For example, the density of $X_t$ in $x_t$ is $p_t^\theta(x_t),$ where $\theta\in\Theta$ is some (eventually known) vector, while the density 
of $X_t$ in $x_t,$ given that $X_{t-1}$ has previously taken value $x_{t-1},$ is $p_t^\theta(x_t\mid x_{t-1}).$  Note that $p_t^\theta(x_t\mid x_{t-1})$ is a member of the parametric family with parameters $(\theta, x_{t-1})$.
The probability distribution of a random variable $X_t$ is represented by $\mathbb{P}_t(dx_t),$ while the probability distribution of a sequence $X_{0:t}$ is represented by $\mathbb{P}_t(dx_{0:t}).$ The Dirac measure, i.e., the measure that assigns probability $1$ to the singleton $\{x\},$ is denoted by $\delta_x(dy)$.The expectation operator is denoted by $\mathbb{E},$ where the section of $\mathbb{E}$ at $\mathbb{Q},$ i.e., $\mathbb{E}_{\mathbb{Q}}$ denotes the operator computed with respect to the probability distribution $\mathbb{Q}.$ 
Let $\varphi$ be a function defined on the random variable $X\propto\mathbb{Q}.$ Then, we denote the expected value of $\varphi(X)$ with 
$$
    \mathbb{E}_{\mathbb{Q}}[\varphi(X)]=
    \int_{\mathbb{X}}\varphi(x)\mathbb{Q}(dx).
$$

\section{Markov Processes}
Let $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ and $(\mathcal{Y}, \mathcal{B}(\mathcal{Y}))$ be two (eventually equal) measurable spaces. 
\begin{definition}
A function $P(x,dy): (\mathcal{X}, \mathcal{B}(\mathcal{X}))\rightarrow [0,1]$ such that 
\begin{itemize}
    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})$, and 
    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X})),$
\end{itemize}
is defined a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})).$
\end{definition}
Notice that for any random variables $X_1$ and $X_0=x_0,$ 
$$
    \mathbb{P}_1(dx_{0:1})=\mathbb{P}_0(dx_{0})P_1(x_0,dx_1),
$$
where $P_1(x_0,dx_1)$ is a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ This remark, provides some intuition about the link between probability kernels and conditional probabilities. As a matter of fact, it is possible to show that for given $X_{0:1}$ with probability distribution $\mathbb{P}_1(dx_{0:1}),$ 
\begin{equation}\label{prob_kern_condit_prob}
    \mathbb{P}_1(X_1\in dx_1\mid X_0=x_0)=P_1(x_0,dx_1).
\end{equation}

Probability kernels can be used to define Markov processes. Before moving to the definition, we will introduce some important concepts. Recall that given two measures $\mathbb{M}$ and $\mathbb{Q}$ defined on a measurable space $(\mathcal{X}, \mathcal{B}(\mathcal{X})),$ $\mathbb{Q}$ is absolutely continuous with respect to $\mathbb{M},$ that is $\mathbb{Q}\ll\mathbb{M},$ if for all sets $A\in\mathcal{B}(\mathcal{X})$ such that $\mathbb{M}(A)=0,$ then it must also hold that $\mathbb{Q}(A)=0.$ If $\mathbb{Q}(A)/\mathbb{M}(A)$ is well-defined for all $A,$ then the Radon-Nikodym theorem guarantees that there exists a measurable function $w(x)\geq0$ such that 
$$
    w(x)=\frac{\mathbb{Q}(dx)}{\mathbb{M}(dx)}, 
$$
meaning that, 
$$
    \mathbb{Q}(A)=\int_\mathcal{X} w(x) \mathbb{M}(dx).
$$

Let $P_1, P_2, \dots, P_T$ be a finite measure of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
\begin{equation}
    \P_T(X_{0:T}\in dx_{0:T})=\P_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},x_s).
\end{equation}
The set $\mathcal{X}$ is the state-space, $\P_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
\end{definition}
It is possible to show that for any $t\in T,$
$$
\P_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\P_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\P_t(x_{t-1},dx_t).
$$
The first equality implies conditional independence. The second equality implies that it is possible to identify conditional distributions with transition kernels. 
Assume $\P_t, t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
\begin{proposition}
For all $t\leq T,$ 
$$
\P_T(dx_{0:t})=\P_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\P_t(dx_{0:t}).
$$
\end{proposition}
This means that the marginal distribution of $X_{0:t}$ with respect to $\P_T$ is $\P_t.$

\section{State-Space Models}

\chapter{Linear-Gaussian case}

\chapter{Particle Filtering}

### 2021-07-11 final draft.Rmd
change font

@@
 header-includes:
+- \usepackage{txfonts}
 - \usepackage{amsmath}
@@
 - \usepackage{graphicx}
-- \usepackage{amssymb}
 - \usepackage{MnSymbol} %allows to insert the QED at the end of the proofs

### 2021-07-11 draft implementation 2.Rmd
Delete some drafts
modifica

--- 2021-07-11 draft implementation.Rmd
@@
 - \newtheorem{algorithm}{Algorithm}[section]
+
 title: "R Notebook"
@@
 \item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
 \end{enumerate}
-\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
 \end{itemize}
@@
 \item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
-\end{enumerate}
 \item if $ESS<N/2$ then
@@
 \item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
 \end{itemize}
@@
 \item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
-\end{enumerate}
 \item if $ESS<N/2$ then
@@
 ```
+
+```{r echo=FALSE}
+gpf<-GPFfun(y,N,m0,C0,tau,sigma)
+ESS.GPF<-ESSplot(df,gpf)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) GPF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.GPF<-Filtplot(df,gpf,"")
+ggarrange(FILT.GPF,ESS.GPF,labels=c("(a)","(b)"))
+```
+
 Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF).
@@
 pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
-gpf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+gpf2<-GPFfun(y,N,m0,C0,tau,sigma,r=k)
 
@@
 \textbf{Implementation}
+
+For illustration purposes, we are going implement an auxiliary particle filter with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.  
+\begin{algorithm} APF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item For $k=1,...,N$:
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+The \code{APF} function resume this passages.
+\
+\hrule
+\hrule
+\code{APF(data,N,m0,C0,tau,sigma,r)}
+\hrule
+\textbf{Arguments}
+\
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+APFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],x,sigma)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,x[k],tau)
+    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x1<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+```{r echo=FALSE}
+apf<-APFfun(y,N,m0,C0,tau,sigma)
+ESS.APF<-ESSplot(df,apf)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.APF<-Filtplot(df,apf,"")
+ggarrange(FILT.APF,ESS.APF,labels=c("(a)","(b)"))
+```
+
+Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap Particle Filter (BPF).
+
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+N2=1000
+
+pf2<-PFfun(y,N2,m0,C0,tau,sigma)
+apf2<-APFfun(y,N2,m0,C0,tau,sigma)
+
+Comparetwofunplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=mean1, col="APF", linetype="APF"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=mean2, col="BPF", linetype="BPF"))+
+  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+   #           fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("APF" = "orange",
+                               "True States" = "black",
+                               "BPF"= "blue"))+
+  scale_linetype_manual("",
+                      values=c("APF" = 1,
+                               "True States" = 3,
+                               "BPF"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+p2<-Comparetwofunplot(df,pf2,apf2,"Filtered States")
+
+ComparetwoVarplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+geom_line(aes(y=sd2, col="APF"))+
+geom_line(aes(y=sd1,col="BPF"))+
+  scale_color_manual("",
+                      values=c("APF" = "orange",
+                               "BPF"= "blue"))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30"))+
+  theme(plot.title = element_text(hjust = 0.5))
+
+}
+
+c2<-ComparetwoVarplot(Combined,pf2,apf2,"Standard Deviation")
+
+ggarrange(p2,c2)
+```
+
+```{r include=FALSE}
+truex<-x
+estimatedx<-matrix(NA,ncol=4,nrow=3)
+colnames(estimatedx)<-c("N","Threshold","BPF","APF")
+estimatedx[,1]<-c(1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.25,0.1)
+
+i=1
+for(k in c(2,5,10)){
+  N=1000
+
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
+
+estimatedx[i,3]<-RMSE(truex,Filtervalues(pf2)$mean)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(apf2)$mean)
+i=i+1
+
+}
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
+```

### 2021-07-11 draft implementation 2.Rmd
da finire**

@@
 ```
+
+\
+\textbf{Implementation}
+\
+Consider the linear Gaussian example of section XX, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
+Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
+\begin{align*}
+\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
+\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
+\end{align*}
+Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. The algorithm follows the following steps.
+\begin{algorithm} LWF for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
+$p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $$
+\end{enumerate}
+
+and $(\psi^{(1)},...,\psi^{(N)})$ from
+
+Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item For $k=1,...,N$:
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+After having drown the hyperparameters  indipendently form their priors and having set $w_{0}^{(i)}=N^{-1}$, $i=1,...,N$, and $\hat{\pi}_{0}=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$, for t=1,..T
+\begin{itemize}
+\item Compute $\hat{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=Var_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$ set
+\begin{align*}
+m(\psi^{(i)}) & =a\psi^{(i)}+(1-a)\overline{\psi}\\
+v(\psi^{(i)}) & =(1-a^2)\Sigma
+\end{align*}
+and
+\begin{align*}
+\alpha(\psi^{(i)}) & =\frac{m(\psi^{(i)})^2}{v(\psi^{(i)})} \\
+\beta(\psi^{(i)}) & =\frac{m(\psi^{(i)})}{v(\psi^{(i)})}
+\end{align*}
+For $k=1,...,N$
+\begin{itemize}
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1^{(i)}}f_{N}(y_{t}|g(x_{t}^{(i)}),m(\psi^{(i)}))$ where for simplicity $g(x_{t}^{(i)})=E(x_{t}|x_{t-1},m(\psi^{(i)}))$
+\item Draw $\psi^{(k)} \sim G(\alpha(\psi^{(I_{k})}),\beta(\psi^{(I_{k})}))$
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^{{2}^{(k)}})$
+\item Set $\tilde{w}_{t}^{k}=\frac{f_N(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_N(y_{t}|g(x_{t}^{(I_{k})}),\psi=m(\psi)^{(I_k)})}$
+\end{itemize}
+\item Normalize the weights
+\item Compute the effective sample size ($ESS$)
+\item If $ESS<N/2$, resample:
+\begin{itemize}
+\item Draw a sample of size N, $x_{t}^{(1)},...,x_{t}^{(N)}$,from the discrete distribution $P((x_{t},\psi)=(x_{t}^{(i)},\psi^{(i)}))=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{itemize}

### 2021-07-12 script di prova.R
Commit B

@@ -2,8 +2,8 @@ library(usethis)
 
 use_git_config(user.name = "edoardo-marcelli", 
                user.email = "edoardomarcelli2@gmail.com")
-use_git_config(user.name = "michelebolla", 
-               user.email = "michele.bolla@studbocconi.it")
+use_git_config(user.name = "nebedetta", 
+               user.email = "etta.bruni@gmail.com")
 
 
 
@@ -101,3 +101,4 @@ ggplot(DLM.df,aes(x=timeframe))+
 #ciao sono Michele
 # test (Alexei)
 # alexei teast from local
+# Test Benedetta

### 2021-07-12 script di prova.R
commit

@@ -2,12 +2,18 @@ library(usethis)
 
 use_git_config(user.name = "edoardo-marcelli", 
                user.email = "edoardomarcelli2@gmail.com")
+<<<<<<< HEAD
+use_git_config(user.name = "nebedetta", 
+               user.email = "etta.bruni@gmail.com")
+
+=======
 use_git_config(user.name = "michelebolla", 
                user.email = "michele.bolla@studbocconi.it")
 use_git_config(user.name = "RobertoColarieti", 
                user.email = "roberto.colarieti1998@gmail.com")
 use_git_config(user.name = "gabrieleromano111", 
                user.email = "gabrieleromano111@gmail.com")
+>>>>>>> 6f1752646ddcf3919f83af94f17e9231e557b3fd
 
 
 #@@@@@@@@@@@@@@@@@@@@##
@@ -104,5 +110,9 @@ ggplot(DLM.df,aes(x=timeframe))+
 #ciao sono Michele
 # test (Alexei)
 # alexei teast from local
+<<<<<<< HEAD
+# Test Benedetta
+=======
 #ciao 
 #Robbi
+>>>>>>> 6f1752646ddcf3919f83af94f17e9231e557b3fd

### 2021-07-12 Real Dataset SV.Rmd
commit

---
title: "SV application - real data"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2021-07-12 final draft.Rmd
i
Adding section on Comparing Constant and Stochastic Volatility Models of Equity Returns in final draft
piccole modifiche a header

@@
 - \newtheorem{lemma}{Lemma}
+- \newtheorem{definition}{Definition}
 - \theoremstyle{nonumberplain}
@@
     \mathbb{E}_{\mathbb{Q}}[\varphi(X)]=
-    \int_{\mathbb{X}}\varphi(x)\mathbb{Q}(dx).
+    \int_{\mathcal{X}}\varphi(x)\mathbb{Q}(dx).
 $$
@@
 where $P_1(x_0,dx_1)$ is a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ This remark, provides some intuition about the link between probability kernels and conditional probabilities.
-As a matter of fact, it is possible to show that for given $X_{0:1}$ with probability distribution $\mathbb{P}_1(dx_{0:1}),$
+As a matter of fact, it is possible to show that for given $X_{0:1}$ with joint probability distribution $\mathbb{P}_1(dx_{0:1}),$
 \begin{equation}\label{prob_kern_condit_prob}
@@
 $$
-    \mathbb{Q}(A)=\int_\mathcal{X} w(x) \mathbb{M}(dx).
+    \mathbb{Q}(A)=\int_A w(x) \mathbb{M}(dx).
 $$

### 2021-07-12 SV/SV_relative_ll.R
working on SV application

@@ -17,12 +17,12 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
   ws<-NULL
   ess<-NULL
   x  = rnorm(N,m0,sqrt(C0))
+  xp.1<-rnorm(N,a.v+b.v*x,s.v)
   w  = rep(1/N,N)
   
   for(t in 1:length(data)){
     
     x<-rnorm(N,a.v+b.v*x,s.v)
-    xp.1<-x
     w1<-w*dnorm(data[t],a,exp(x/2))
     
     w = w1/sum(w1)
@@ -34,9 +34,9 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
       w<-rep(1/N,N)
     }else{}
     
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
+    xs  = rbind(xs,x)
+    ws  = rbind(ws,w)
+    ess = rbind(ess,ESS)
   }
   return(list(xs=xs,ws=ws,ess=ess,xp.1=xp.1))
 }
@@ -72,18 +72,18 @@ SVPFfun<-function(data,N,m0,C0,a,a.v,b.v,s.v,r){
 #   return(list(xs=xs,xs.f=xs.f,ws=ws,ess=ess,x.f1=x.f1))
 # }
 ## data
-# y<-as.matrix(read.csv("sp500_17-21.csv", header=F))
-y<-as.matrix(rnorm(1000,mean=1,sd=1)) #test
-
-n<-dim(y)[1]
+data<-read.csv("sp500_17-21.csv", header=T)
+data$t<-as.Date(data$t)
+y<-data$r
+#y<-rnorm(996,mean=1,sd=1) #test
+n<-length(y)
 
 # find parameters for constant and stochastic vol models
 mean=sum(y)/n
 res<-(y-mean)^2
 lres<-log(res)
-l.lres<-c(NA,lres[1:n-1,1])
+l.lres<-c(NA,lres[1:n-1])
 resdf=data.frame(lres,l.lres)
-colnames(resdf)=c("lres","l.lres")
 lm_res<-lm(lres~1+l.lres,data=resdf)
 c_res<-lm_res[[1]]
 sigma_res<-sqrt(mean(summary(lm_res)$residuals^2))
[…]

### 2021-07-12 Real Dataset Script.R
No resampling

@@ -11,6 +11,37 @@
 
 #Particle Filter
 #---------------
+SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS<0){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Particle Filter
+#---------------
 SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
@@ -273,6 +304,7 @@ dfsv<-data.frame(timeframe,y,x)
 #---------
 set.seed(12345)
 N=5000
+svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
 svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
@@ -305,12 +337,10 @@ Filtplot<-function(dataframe,fun,title){
     geom_ribbon(aes(ymin = mean1-1.96*sd, ymax = mean1+1.96*sd),
                 fill="red",alpha=0.16) +
     scale_color_manual("",
-                       values=c("Observations" = "black",
-                                "True States" = "black",
+                       values=c("True States" = "black",
                                 "Filtered States"= "red"))+
     scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "True States" = 3,
+                          values=c("True States" = 3,
[…]

### 2021-07-12 SV/SV_relative_ll.R
SV, final draft

@@ -138,7 +138,6 @@ cll_diff.df=data.frame(data$t,cll_diff)
 colnames(cll_diff.df)<-c('t','cll_diff')
 
 cll_plot<-ggplot(data = cll_diff.df, aes(x = t, y = cll_diff))+
-geom_line()+
-labs(x="Time",y="")
+geom_line()
 
 save(cll_plot, file="cll_plot.Rda")

### 2021-07-12 final draft.Rmd

--- 2021-07-11 final draft.Rmd
@@
 ---
-
+```{r, results= FALSE, message=FALSE, warning=FALSE, echo=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(fig.pos = '')
+knitr::opts_chunk$set(out.extra='')
+```
+```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
+Sys.setlocale("LC_TIME", "English")
+```
+```{r include=FALSE}
+library(ggplot2)
+library(ggpubr)
+library(kableExtra)
+```
 \chapter*{Introduction}
@@
 \chapter{Particle Filtering}
+
+\chapter{Applications to Stochastic Volatility Models}
+\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}
+
+In this section, we apply the particle filter in a model-comparison
+exercise, ultimately underlining the importance of taking stochastic
+volatility into account in modeling equity returns.
+
+We consider a stochastic volatility model
+$$
+\mathcal{M}_{SV}:\begin{array}{lc}
+r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
+\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
+\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
+$$
+
+and a constant volatility iid model 
+$$
+\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
+$$
+
+We calibrate the model parameters using the whole sample, and proceed
+by setting the return mean of both models to equal the sample mean,
+$\sigma^{2}$ equal to the return variance, and the parameters of
+the volatility equation in $\mathcal{M}_{SV}$ by running an $AR(1)$
+regression of the log squared residuals.
+
+In order to compare the predictive performance of the two, we compare
+the cumulative log likelihood of the observed returns up to time $t$
+given the considered model. In formulas, the cumulative likelihood
+can be computed as
+$$
+p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
+$$
+
+adopting the convention $y_{1:0}=\emptyset$. This is trivial to obtain
+in the iid case; considering the stohastic volatility model, the predictive
+likelihoods can be derived as
+
+$$
+p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
+$$
+$$
+=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
+$$
+
+Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
+is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
+with weighted particles obtained with particle filtering.
+
+Figure XX reports the time series of the cumulative log-likelihood
+given the constant volatility model $\mathcal{M}_{CV}$ relative to
+the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
+Notice that the first difference of this series is the relative log-likelihood
+of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.
+
+```{r cll_plot, echo=F, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
+load("SV/cll_plot.Rda")
+cll_plot+
+  theme_bw()+
+  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
+  labs(x="Time (month/year)",y="")
+```
+Overall, this analysis shows how a model with stochastic
+volatility is better at describing and predicting returns. As can be seen from the graph and perhaps unsurprisingly,
+the density forecast given the stochastic volatility model represents
+a particularly significant improvement in the period of extreme values of returns in March 2020.

### 2021-07-13 Real Dataset Script.R
commit

@@ -303,7 +303,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=5000
+N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
@@ -357,7 +357,7 @@ library(ggplot2)
 library(ggpubr)
 plot1<-Filtplot(dfsv,svpf,"Particle Filter")
 plot2<-Filtplot(dfsv,svapf, "Auxiliary Particle Filter")
-plot3<-Filtplot(dfsv,svlw,"Liu e West Filter")
+plot3<-Filtplot(dfsv,svlw,"Liu and West Filter")
 plot4<-Filtplot(dfsv,svopt,"Opt Ker Particle Filter")
 plot5<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
 plot6<-Filtplot(dfsv,svsis,"No Resampling")

### 2021-07-13 Real Dataset Script.R
RMSE comparison

@@ -361,13 +361,27 @@ plot3<-Filtplot(dfsv,svlw,"Liu and West Filter")
 plot4<-Filtplot(dfsv,svopt,"Opt Ker Particle Filter")
 plot5<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
 plot6<-Filtplot(dfsv,svsis,"No Resampling")
-ggarrange(plot1,plot2,plot3,plot4,plot5)
+ggarrange(plot1,plot2)
 ggarrange(plot1)
 ggarrange(plot2)
 ggarrange(plot3)
 ggarrange(plot4)
 ggarrange(plot5)
 ggarrange(plot6,plot1)
+
+## RMSE comparison
+
+realisedx<-x
+RMSEvol<-matrix(NA,ncol=6,nrow=1)
+colnames(RMSEvol)<-c("N","PF", "PFOPT", "APF","LWF", "SIS")
+RMSEvol[,1]<-c(10000)
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+comparablevol<-function(fun){exp((Filtervalues(fun)$mean)/2)}
+RMSEvol[1,2]<-RMSE(realisedx,comparablevol(svpf))
+RMSEvol[1,3]<-RMSE(realisedx,comparablevol(svopt))
+RMSEvol[1,4]<-RMSE(realisedx,comparablevol(svapf))
+RMSEvol[1,5]<-RMSE(realisedx,comparablevol(svlw))
+RMSEvol[1,6]<-RMSE(realisedx,comparablevol(svsis))
 
 #Kalman filter
 m00=0
@@ -433,3 +447,6 @@ ggplot(DLM.df,aes(x=timeframeKF))+
   theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
         legend.background = element_rect(fill = "white", colour = "gray30")) +
   theme(plot.title = element_text(hjust = 0.5))
+
+
+

### 2021-07-13 Real Dataset Script.R
RMSE+MAE

@@ -303,7 +303,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=10000
+N=5000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
@@ -369,19 +369,26 @@ ggarrange(plot4)
 ggarrange(plot5)
 ggarrange(plot6,plot1)
 
-## RMSE comparison
+## RMSE MAE comparison
 
 realisedx<-x
-RMSEvol<-matrix(NA,ncol=6,nrow=1)
-colnames(RMSEvol)<-c("N","PF", "PFOPT", "APF","LWF", "SIS")
-RMSEvol[,1]<-c(10000)
+Errorvol<-matrix(NA,ncol=6,nrow=2)
+colnames(Errorvol)<-c("N","PF", "PFOPT", "APF","LWF", "SIS")
+rownames(Errorvol)<-c("RMSE","MAE")
+Errorvol[,1]<-c(10000)
 RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
 comparablevol<-function(fun){exp((Filtervalues(fun)$mean)/2)}
-RMSEvol[1,2]<-RMSE(realisedx,comparablevol(svpf))
-RMSEvol[1,3]<-RMSE(realisedx,comparablevol(svopt))
-RMSEvol[1,4]<-RMSE(realisedx,comparablevol(svapf))
-RMSEvol[1,5]<-RMSE(realisedx,comparablevol(svlw))
-RMSEvol[1,6]<-RMSE(realisedx,comparablevol(svsis))
+Errorvol[1,2]<-RMSE(realisedx,comparablevol(svpf))
+Errorvol[1,3]<-RMSE(realisedx,comparablevol(svopt))
+Errorvol[1,4]<-RMSE(realisedx,comparablevol(svapf))
+Errorvol[1,5]<-RMSE(realisedx,comparablevol(svlw))
+Errorvol[1,6]<-RMSE(realisedx,comparablevol(svsis))
+Errorvol[2,2]<-MAE(realisedx,comparablevol(svpf))
+Errorvol[2,3]<-MAE(realisedx,comparablevol(svopt))
[…]

### 2021-07-13 draft implementation 2.Rmd
ultime modifiche

--- 2021-07-11 draft implementation 2.Rmd
@@
 \textbf{Arguments}
-\
+
 \code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
@@
 \end{algorithm}
-Our \code{SIS} function implemented in \R replicate this steps.
-\
-\hrule
-\hrule
-\code{SIS(data,N,m0,C0,tau,sigma)}
+Our \code{SISfun} function implemented in \R replicate this steps.
+\
+\hrule
+\hrule
+\code{SISfun(data,N,m0,C0,tau,sigma)}
 \hrule
 \textbf{Arguments}
-\
+
 \code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
@@
 \end{algorithm}
-These steps are resumed in our \code{PF} function.
-\
-\hrule
-\hrule
-\code{PF(data,N,m0,C0,tau,sigma,r)}
+These steps are resumed in our \code{PFfun} function.
+\
+\hrule
+\hrule
+\code{PFfun(data,N,m0,C0,tau,sigma,r)}
 \hrule
 \textbf{Arguments}
-\
+
 \code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
@@
 \end{algorithm}
-The \code{GPF} function resume this passages.
-\
-\hrule
-\hrule
-\code{GPF(data,N,m0,C0,tau,sigma,r)}
+The \code{GPFfun} function resume this passages.
+\
+\hrule
+\hrule
+\code{GPFfun(data,N,m0,C0,tau,sigma,r)}
 \hrule
 \textbf{Arguments}
-\
+
 \code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
@@
 \end{enumerate}
-\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
 \end{enumerate}
@@
 \end{algorithm}
-The \code{APF} function resume this passages.
-\
-\hrule
-\hrule
-\code{APF(data,N,m0,C0,tau,sigma,r)}
+The \code{APFfun} function resume this passages.
+\
+\hrule
+\hrule
+\code{APFfun(data,N,m0,C0,tau,sigma,r)}
 \hrule
 \textbf{Arguments}
-\
+
 \code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
@@
 ```
+
+```{r}
+APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+    weight = w*dnorm(data[t],x,predsd)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    }else{
+    weight = rep(1/N,N)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    }
+
+    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
+    x1   = rnorm(N,means,importancesd)
+    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    x <- x1
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+```{r echo=FALSE}
+apfopt<-APFoptfun(y,N,m0,C0,tau,sigma)
+ESS.APFopt<-ESSplot(df,apfopt)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.APFopt<-Filtplot(df,apfopt,"")
+ggarrange(FILT.APFopt,ESS.APFopt,labels=c("(a)","(b)"))
+```
+
+\
+\textbf{Implementation}
+\
+Consider the linear Gaussian example of section XX, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
+Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
+\begin{align*}
+\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
+\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
+\end{align*}
+Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. The algorithm follows the following steps.
+\begin{algorithm} LWF for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
+$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:.
+\begin{enumerate}
+\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\begin{align*}
+m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
+\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\end{align*}
+\item For $k=1,...,N$:
+\begin{itemize}
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
+\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
+\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
+\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
+\end{itemize}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+Our \code{LWfun} function goes through the illustrated steps.
+\
+\hrule
+\hrule
+\
+\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alphav, betav} \ \  Gamma prior hyperparameters on $\sigma^{2}$
+\
+\code{alphaw,betaw} \ \  Gamma prior hyperparameters on $\tau^{2}$
+\
+\code{delta} \ \ hyperparameter delta value
+\
+\code{unif} \ \ if True then it sets a Uniform $(0,10)$ prior on $\sigma^{2}$ and $\tau^{2}$
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+```{r}
+LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
+  if(missing(r)){r=2}else{}
+  xs     = rnorm(N,m0,sqrt(C0))
+  if(unif==T){
+  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
+  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,2,n))
+  xss    = NULL
+  ws     = NULL
+  ess    = NULL
+  w      = rep(1/N,N)
+  for (t in 1:length(data)){
+    meanV = weighted.mean(pars[,1],w)
+    varV  = weighted.mean((pars[,1]-meanV)^2,w)
+    meanW = weighted.mean(pars[,2],w)
+    varW  = weighted.mean((pars[,2]-meanW)^2,w)
+
+    muV = a*pars[,1]+(1-a)*meanV
+    sigma2V = (1-a^2)*varV
+    alphaV = muV^2/sigma2V
+    betaV = muV/sigma2V
+
+    muW = a*pars[,2]+(1-a)*meanW
+    sigma2W = (1-a^2)*varW
+    alphaW = muW^2/sigma2W
+    betaW = muW/sigma2W
+
+    weight      = w*dnorm(data[t],xs,sqrt(muV))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+
+    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
+    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])
+
+    xsprevious<-xs[k]
+    xs = rnorm(N,xs[k],sqrt(pars[,2]))
+
+    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
+                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xs[index]
+      pars<-pars[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xs
+      pars<-pars
+    }
+
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+    ess         = rbind(ess,ESS)
+  }
+  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
+}
+```

### 2021-07-13 final draft.Rmd
commit Edo

@@
 - \newcommand{\Pd}{\mathbb P}
+- \newcommand{\N}{\mathcal N}
 - \def\code#1{\texttt{#1}}
@@
 ```
+
 \chapter*{Introduction}
@@
 
-\chapter{Particle Filtering}
+\chapter{Particle Filter}
+
+In the previous chapter we explained how the Kalman Filter is applied to the case of a linear model with Gaussian errors to find closed-form solution for the filtering and the predictive distributions.
+Nonetheless, the majority of time-series models have a non-linear structure and errors might follow non-Gaussian distributions.\
+In this chapter we describe \textit{particle filter} as an algorithm that allows to update sequentially the filtering and the predictive distributions in the non-linear and non-Gaussian case.
+Additionally, the Kalman Filter can be seen as a special case of the more general particle filter.\
+In section \ref{pf_intro} we introduce the fundamental notions of \textit{importance sampling}, \textit{sequential importance sampling} and \textit{resampling}.
+Then, in section \ref{pf_main} we describe three particle filtering algorithms: the \textit{bootstrap}, the \textit{guided} and the {auxiliary particle filters}.
+Instead, in section \ref{pf_lui_west} we describe the {Liu and West filter}, when particle filtering is used to estimate model's parameters.
+Finally, in section \ref{pf_converg} we establish some general convergence results for Particle Filter - Monte Carlo estimators.
+
+\section{Introductory notions: sampling and resampling}\label{pf_intro}
+
+Let $\varphi: \X \rightarrow \R$ be a measurable, integrable and bounded real-valued function, where $(\X,\B(\X),\Q)$ is a probability space.\footnote{$\B(\X)$ is the Borel sigma-algebra of $\X$.}
+Suppose that the probability measure $\Q$ admits a density $q(x)$ with respect to some measure $\nu(dx)$ and it is feasible to sample from that density.\
+To estimate the expected value of $\varphi$ with respect to the density $q$, the basic Monte Carlo (MC) method would require to sample $N$ times from the target density $q$ and to compute the average of the function $\varphi$ values.
+$\hat \varphi_{MC}$ is the standard MC estimator:
+\begin{equation*}
+   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \E_q(\varphi)~~~X^n\thicksim q(\cdot)
+\end{equation*}
+However, in many circumstances either it is impossible to sample from the target density or there exist another density - call it $m(\cdot)$ - different from $q$ that produces ``more efficient" estimates.
+In all the cases for which sampling takes place from a distribution different from the ``correct" (i.e., the true) one, importance sampling and resampling methods can be applied (as augmented versions of the standard MC algorithm).
+
+\subsection{Importance sampling}
+
+Let us consider two probability spaces: $(\X,\B(\X),\Q)$ and $(\X,\B(\X),\M)$ and let the measurable function $w:\X \rightarrow \R$ be proportional to the Radom-Nikodym derivative between $\Q$ and $\M$, i.e.,  $w(x)\propto\frac{\Q(dx)}{\M(dx)}$.\footnote{From the Radom-Nikodym theorem a sufficient condition for $w$ to exist is that $\Q$ is absolutely continuous with respect to $\M$: $\Q\ll \M$.} \
+The following result represents the starting point for the idea of importance sampling.
+
+\begin{lemma}\label{lemma_imp_sam_1}
+For any measurable, integrable and bounded function $\varphi: \X \rightarrow \R$, the following holds:
+\begin{equation}
+    \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+\end{equation}
+\end{lemma}
+
+\begin{proof*}
+Note that $\Q(dx)=\frac{w(x)}{\E_\M(w)}\M(dx)$. Hence:
+\begin{equation*}
+    \E_\Q(\varphi)=\int_\X \varphi(x)\Q(dx)=\int_\X \varphi(x)\frac{w(x)}{\E_\M(w)}\M(dx)=\frac{\E_\M(\varphi \cdot w)}{\E_\M(w)}
+\end{equation*}
+$\hfill\blacksquare$
+\end{proof*}
+
+Lemma \ref{lemma_imp_sam_1} represents the key theoretical foundation of \textbf{importance sampling} (IS), a numerical algorithm that allows to sample from a \textit{proposal density} to estimate moments that are computed with respect to the \textit{target density}, by means of re-weighting the values of the target function $\varphi$.\
+Adopting the previous notation, the target measure is $\Q$, while the proposal is $\M$.
+IS requires to sample $N$ values from $\M$, then to find the associated \textit{normalized}\footnote{Weights are normalized when $\E_\M(W)=1$.
+In this case $W$ is the Radom-Nikodym derivative between $\Q$ and $\M$.} weights $W(X^n)$, finally to compute a weighted average of the values taken by the function $\varphi$.
+The IS is implemented according to the following algorithm.
+\begin{enumerate}
+    \item Sample $N$ values from the proposal measure: $X^n\thicksim \M$ for $n=1,...,N$.
+    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
+    \item Compute the estimator:
+    \begin{equation*}
+        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
+    \end{equation*}
+\end{enumerate}
+Note that $\hat \varphi$ is an unbiased estimator for the target of inference $\E_\Q(\varphi)$. In fact:
+\begin{equation*}
+    \begin{split}
+        \E_\M(\hat \varphi)=\frac{1}{N} \sum_{n=1}^N \E_\M(\varphi \cdot W)=\E_\M(\varphi \cdot W)=\E_\Q(\varphi)
+    \end{split}
+\end{equation*}
+where the last equality follows from lemma \ref{lemma_imp_sam_1} and the fact that weights are normalized to 1.
+
+\subsubsection{Auto-normalized IS}
+
+Usually, the Radom-Nikodym derivative is known up to a normalizing constant and the IS algorithm has to be described in more general terms.
+In the previous section the \textit{normalized} weighting function $W(\cdot)$ was the Radom-Nikodym derivative of $\Q$ with respect to $\M$.
+More generally, consider the function $w(x)\propto\frac{\Q(dx)}{\M(dx)}$, denoted as \textit{importance function}, proportional to the derivative $W(\cdot)$.\
+In this case, weights have to be \textit{auto}-normalized to obtain the IS - Monte Carlo estimator. The auto-normalized IS is implemented according to the following procedure.
+\begin{enumerate}
+    \item Sample $N$ values from the proposal measure: $X^n\thicksim \M$ for $n=1,...,N$.
+    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
+    \item Normalize the weights, for $n=1,...,N$:
+    \begin{equation*}
+        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
+    \end{equation*}
+    \item Compute the estimator:
+    \begin{equation*}
+        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
+    \end{equation*}
+\end{enumerate}
+While the auto-normalized estimator is biased (since  $W^n$ involves a ratio of random variables), it is consistent (see \ref{pf_is_conv} for details). \ \
+It is useful to interpret IS as an algorithm that allows to find an approximation for the probability measure $\Q$ as a weighted average of Dirac measures for the sampled values ($X^n$): $\delta_{X^n}(dx)$ where the weights are the auto-normalized ones ($W^n$).
+In particular:
+\begin{equation}
+    \Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \M
+\end{equation}
+Furthermore, the IS estimator is the expected value of the function $\varphi$ with respect to the approximating probability measure $\Q^N$:
+\begin{equation*}
+    \begin{split}
+        \E_{\Q^N}(\varphi)=&\int_\X \varphi(x)\Q^N(dx)=\int_\X \varphi(x)\sum_{n=1}^NW^n\delta_{X^n}(dx)=\\
+       = &\sum_{n=1}^N\int_\X \varphi(x)W^n\delta_{X^n}(dx)=\sum_{n=1}^N\varphi (X^n)W^n=\hat \varphi_{AN}
+    \end{split}
+\end{equation*}
+
+\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}
+
+Consider the auto-normalized IS estimator $\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n$. We can write:
+\begin{equation*}
+   \hat \varphi_{AN} - \E_{\Q}(\varphi)=\E_{\Q^N}(\varphi)-\E_{\Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\E_{\Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
+\end{equation*}
+Let $\bar \varphi(X):=\varphi(X)-\E_{\Q}(\varphi)$. Applying the CLT to the numerator, assuming that $\E_{\M}(w^2\bar\varphi^2)<+\infty$, we get:
+\begin{equation}
+    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \N \big(0,\E_{\M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
+\end{equation}
+where $\implies$ denotes convergence in distribution. Instead, for the denominator the strong LLN implies that:
+\begin{equation}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \E_{\M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
+\end{equation}
+Applying Slutsky's theorem, together with \eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain:
+\begin{equation}
+   \sqrt{N}\big[  \hat \varphi_{AN} - \E_{\Q}(\varphi)\big] \implies \N \Bigg(0,\frac{\E_{\M}(w^2\bar\varphi^2)}{[\E_{\M}(w)]^2}\Bigg)
+\end{equation}
+Therefore, the IS (auto-normalized) estimator for $\varphi$ is both consistent and asymptotically normal.
+
+\subsubsection{Effective Sample Size}
+
+A common measure of efficiency for IS is the \textit{effective sample size} (ESS), defined as:
+\begin{equation*}
+    ESS(W^{1:N}):=\frac{1}{\sum_{n=1}^N(W^n)^2}=\frac{[\sum_{n=1}^N w(X^n)]^2}{\sum_{n=1}^N(w(X^n))^2}
+\end{equation*}
+Observe that the $ESS$ lies in the interval $[1,N]$. For instance, if $W^n=1/N$, as in the case of the standard MC method, $ESS=N$.\
+It is possible to relate the $ESS$ to the variance of the importance function.  In fact:
+\begin{equation*}
+    \frac{N}{ESS}=1+\frac{N^{-1}\sum_{n=1}^N(w(X^n))^2-\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}{\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}
+\end{equation*}
+The second term in the numerator of the previous formula is the square of the coefficient of variation (i.e., the ratio between the variance and the squared mean) of the un-normalized weights.
+It is clear that a lower $ESS$ is associated with a higher variance for the weights.
+
+\subsection{Sequential Importance Sampling}
+
+When dealing with state-space models, the importance sampling algorithm has to be applied dynamically and the importance weights are updated period-by-period.
+To illustrate the relevance of the \textit{sequential importance sampling} (SIS) algorithm, we consider a two-period stochastic process.\ \
+Let us consider the approximation of the measure $\Q_0$ obtained at time 0 through IS from the measure $\M_0$:
+\begin{equation*}
+    \Q_0^N(dx_0)=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0),~~\text{with } X_0^n\thicksim \M_0,~~\text{and }  W_0^n:=\frac{w_0(X_0^n)}{\sum_{m=1}^Nw_0(X_0^m)}.
+\end{equation*}
+Moreover, let $M_1(x_0,dx_1)$ be the transition kernel. We want to update, sequentially, the approximating measure $\Q_0$ to obtain an approximation of the measure:
+\begin{equation*}
+    \Q_{1}(dx_{0:1})=\Q_0(dx_0)M_1(x_0,dx_1)
+\end{equation*}
+SIS works according to the following algorithm.
+\begin{enumerate}
+    \item Sample $N$ values for $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
+    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^n,dx_1)$ for $n=1,...,N$.
+    \item Compute the auto-normalized weights.
+In this case, they are computed taking into account only time 0 sampled observations since $X_1^n$ is sampled from the correct distribution: $W_1^n=W_0^n$.
+\end{enumerate}
+Therefore, the particle approximation of $\Q_{1}(dx_{0:1})$ is:
+\begin{equation*}
+      \Q_{1}^N(dx_{0:1})=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0)M_1(X_0^n,dx_1)
+\end{equation*}
+
+\subsection{Sequential Importance Resampling}
+
+A second approach can be used when sampling has to be done sequentially: \textit{sequential importance resampling} (SIR).
+Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
+While in the SIS all the particles (i.e., the sampled values) ``survive", with the ``risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR.
+In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
+We describe the SIR more formally, focusing on the previous two-period case.
+\begin{enumerate}
+    \item Sample $N$ values $X_0$ from the proposal measure $\M_0$: $X_0^n\thicksim \M_0$ for $n=1,...,N$.
+    \item Compute the weights $W_0^n$ for $n=1,...,N$.
+    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
+    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
+    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
+\end{enumerate}
+Steps 3-4 constitute the actual resampling ones.
+At this point, let us define the new two-period sequence $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$.
+Then, the particle approximation for the joint distribution of $(X_0,X_1)$ becomes:
+\begin{equation*}
+     \Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
+\end{equation*}
+Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
+To understand the reason of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$.
+Consider the sequential updating for the weights:
+\begin{equation}
+    w^n_1=\frac{\Q_1(\tilde dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
+=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
+\end{equation}
+where $\M_1$ denotes the probability measure for the extraction of the particle $\tilde x_{0:1}^n$ according to SIR.
+$\M_1$ (second equality sign in the previous formula) can be rewritten as the product between the kernel for $x_1^n$ given $x_0^{A_1^n}$ and the probability of $x_0^{A_1^n}$ itself.
+This last probability can also be rewritten as the product between the proposal density $m_0$ from which $X_0$ is extracted and the probability that it is sampled again when resampling occurs.
+
+\subsection{SIS vs SIR}
+The choice between the two sequential sampling strategies (SIS and SIR) has significant implications in terms of computational cost and variance of the importance weights.
+In fact, it is clear that SIR is more costly computationally since it requires an intermediate resampling step.
+Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
+Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation} admit a density too).
+Then, weights, for every $t=1,...,T$, are updated according to:
+\begin{equation}
+    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto
+    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
+\end{equation}
+Equation \eqref{sis_update} suggests, intuitively, that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue.
+Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.
+
+\section{Particle Filter}\label{pf_main}
+
+\color{red}
+Let us consider the standard state-space model introduced in the previous chapters.
+\color{black}
+\textit{Particle filter} (PF) is a numerical algorithm that is used to obtain approximations for the filtering - $\Pd_t (dx_t|y_{1:t})$ - and the prediction - $\Pd_t (dx_t|y_{1:t-1})$ - distributions for inference in state-space models.\ \
+In a state-space model sampling sequentially from the transition kernels for the states is not enough to obtain good approximations since the observation sample ($y_{1:T}$) contains information about the underlying states through the likelihood function $f_t(y_t|x_t)$.
+Observations constitute signals for the model's states.\
+Therefore, the conditional distribution of the states, given the observed data - $\Pd_t(dx_{1:t}|y_{1:t})$ - changes over time as new data points become available.
+The PF algorithm allows to incorporate sequentially the new data to obtain new approximations for this distribution (and for the filtering and prediction ones).
+\
+In general, at each step of the algorithm, states are sampled sequentially from the proposal measure, call it $\M_t$, and importance auto-normalized weights are computed as the Radom-Nikodym derivative of the target $\Pd_t$ with respect to the proposal $\M_t$.
+Moreover, the Markov properties of the sequence of the states allows to update sequentially the importance weights.\ \
+We will analyse three different types of particle filtering procedures: \textit{bootstrap}, \textit{guided} and \textit{auxiliary particle filters}, that differ between them with respect to their scope and efficiency.
+
+\subsection{Degeneracy of the SIS in Particle Filter}
+
+Usually, PF requires a resampling step at each stage of the algorithm.
+Recall that, in general, resampling determines a trade-off between the increase in the variance at the resampling stage and a decrease in the variance when future states are sampled from the transition kernel.\
+``Dead" particles (i.e., those with small weights) are  likely to be removed when resampling takes place and past weights get ``reset" (usually, they are set to be equal to $1/N$).
+Otherwise, the \textit{curse of dimensionality} problem will emerge and the variance of the weights will diverge over time as the variance of past weights increases as new updates occur period-by-period.
+Figure ***** shows an example of a PF where SIS is used: note that the $ESS$ drops dramatically after few periods, meaning that the variance of the importance weights diverges.
+This motivates the use of a SIR structure for the PF algorithm.\ \
+Lastly, it is now common practice to use a mixed approach with respect to resampling (recall that resampling has an additional computational cost).
+Generally, resampling is done \textit{adaptively}, when some condition is met.
+A standard criterion that is used (and that we adopt in our implementations of the PF) is to resample when the $ESS$ drops below some threshold $ESS_{min}\in [1,N]$:
+\begin{equation}
+    ESS(W_{t-1}^{1:N})<ESS_{min}
+\end{equation}
+
+\subsection{Bootstrap Particle Filter}
+
+The bootstrap particle filter (BPF) is the easiest form of an adaptive PF algorithm.
+In fact, future states are sampled from the \textit{correct} transition kernel and the weights are updated sequentially using the likelihood $f_t(y_t|x_t)$.\
+We summarize the procedure with the following scheme.
+\begin{enumerate}
+    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and set the un-normalized weights $w_0^n=1$.\
+    Then, the normalized weights are: $W_0^n=\frac{1}{N}$.
+    \item \textbf{For every $t=1,...T$:}
+    \begin{enumerate}
+        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:\footnote{It is indifferent to implement this resampling step at the beginning or at the end of each iteration (as we do in our codes) as long as the initial stage $t=0$ generates weights that are constant and equal to $1/N$.
+In fact, in this case, resampling will never take place at the beginning of the first iteration since $ESS(W_0^n=1/N)=N$ that is larger than or equal to $ESS_{min}$.\
+        However, when we describe the algorithm, we chose to leave this resampling step at the beginning of each iteration.
+In fact, when considering the guided or auxiliary PF, we can, more generally, sample the initial values (time $t=0$) for the state from a proposal distribution different from the target one.
+In this case, the weights at $t=0$ might differ from $1/N$, hence resampling might take place already at the beginning of the first iteration.}
+\begin{enumerate}
+    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
+    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
+\end{enumerate}
+\textbf{Otherwise, } 
+\begin{enumerate}
+    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$.
+    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}^n$ for $n=1,...,N$.
+\end{enumerate}
+\item Sample the future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t)$  for $n=1,...,N$.
+\item Update the un-normalized weights through the likelihood function for $n=1,...,N$.:
+\begin{equation*}
+    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n).
+\end{equation*}
+\item Normalize the weights  for $n=1,...,N$.: $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
+    \end{enumerate}
+\end{enumerate}
+Therefore, at the initial step the pre-sample period states $x_0^n$ are extracted and every sampled state is assigned an equal weight since there is no corresponding observation $y_0$ that gives additional information on the likelihood of that extraction.
+\
+Resampling does never take place in $t=1$\footnote{Since $ESS(W_0^n=1/N)=N\geq ESS_{min}$.} and the state in $t=1$ is sampled from the correct transition kernel $P_1(x_0^n,dx_1)$.
+Now, the time 0 weights are updated incorporating the information from the data point $y_1$ in a very intuitive way: more \textit{likely} states (given $y_1$) are given higher weights: $w_1^n=f_1(y_1|x_1^n)$.
+This weights are used to approximate the time $1$ filtering distribution.\
+At a subsequent stage, the $ESS$ criterion should be checked and, in case it is satisfied, resampling takes place and time $1$ weights are reset to be equal to $1/N$.\ \
+The key characteristic of the PF is the fact that the importance weights are updated sequentially, according to the ``information" provided by the observed data (i.e., the sequence $y_{1:T}$).
+Recall that time $t$ weight is the Radom-Nikodym derivative of the target with respect to the proposal distribution.
+In the BPF, the time $t$ target is the measure $\Pd_t(dx_{0:t}|y_{1:t})$, while the proposal, denoted generically by $\M_t$, is $\M_t(dx_{0:t}|y_{1:t})=\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)$.
+Hence:
+\begin{equation}
+    \begin{split}
+        w_t\propto&\frac{\Pd_t(dx_{0:t}|y_{1:t})}{\M_t(dx_{0:t}|y_{1:t})}\propto \frac{\Pd_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f(y_t|x_t)}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)}=\\
+        =&\underbrace{\frac{\Pd_{t-1}(dx_{0:t-1}|y_{1:t-1})}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot f(y_t|x_t)=\hat w_{t-1}\cdot f(y_t|x_t)
+    \end{split}\label{weight_updated_bpf}
+\end{equation}
+Finally, the BPF is able to generate the two key objects for inference in state-space models:
+\begin{itemize}
+    \item the predictive distribution:
+    \begin{equation}
+    \begin{split}
+        \Pd_t^N (dx_t|y_{1:t-1})=&\frac{1}{\sum_{n=1}^N\hat w_t^n}\sum_{n=1}^N\hat w_t^n\delta_{X_t^n}(dx_t)=\\
+        =&\begin{cases}
+        \frac{1}{N}\sum_{n=1}^N\delta_{X_t^n}(dx_t)
+        ~~~~~~~~~~~~~~~~ ~\text{ if resampling occurs}\\
+        \frac{1}{\sum_{n=1}^N w_{t-1}^n}\sum_{n=1}^Nw_{t-1}^n\delta_{X_t^n}(dx_t)~~~~ \text{ otherwise}
+        \end{cases},
+        \end{split}
+    \end{equation}
+    \item the filtering distribution:
+    \begin{equation}
+        \Pd_t^N (dx_t|y_{1:t})=\sum_{n=1}^NW_t^n\delta_{X_t^n}(dx_t)
+    \end{equation}
+\end{itemize}
+
+\subsection{Guided Particle Filter}
+
+It is not always possible to sample sequentially from the correct transition kernel.
+Moreover, even if it were possible, there exist a proposal transition kernel that makes the algorithm more efficient.
+This motivates the \textit{guided particle filter} (GPF), that can be interpreted as a more general version of the GPF.
+In fact, with the GPF it is possible, at each stage, to sample new states from a transition kernel that differs from the correct one.\ \
+Let $P_t(x_{t-1},dx_t)$ be the correct kernel and consider a kernel $M_t(x_{t-1},dx_t)$ defined on the same measurable space of $P_t$.
+Moreover, let us assume that that $P_t$ is absolutely continuous with respect to $M_t$\footnote{This condition is stronger than what would be needed for the existence of the importance weights in GPF, i.e., that  $P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)\ll M_t(x_{t-1},dx_t)$.
+} (i.e., $P_t\ll M_t$).
+If sampling occurs sequentially from this kernel, the way in which weights are updated should differ from the BPF.
+Consider equation  \eqref{weight_updated_bpf} and let us reformulate it for the GPF case.
+Here, the proposal distribution is not $\M_t(dx_{0:t}|y_{1:t})=\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)$ anymore, but becomes $\M_t(dx_{0:t}|y_{1:t})=\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)$.
+Hence:
+\begin{equation}
+    \begin{split}
+        w_t\propto&\frac{\Pd_t(dx_{0:t}|y_{1:t})}{\M_t(dx_{0:t}|y_{1:t})}\propto \frac{\Pd_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)}=\\
+        =&\underbrace{\frac{\Pd_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)} =\hat w_{t-1}\cdot\underbrace{\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}}_{\text{updating factor}}
+    \end{split}\label{weight_updated_gpf}
+\end{equation}
+Note that the updating factor is more complex that the one of the BPF: in addition to the likelihood function $f_t(y_t|x_t)$, the fact that sampling occurs from a kernel different from the correct ones requires the additional term $P_t(x_{t-1},dx_t)/M_t(x_{t-1},dx_t)$.\ \
+We are now ready to establish, informally, the optimality result about the transition kernel.
+Let us denote by  $G_t=[P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)dy_t]/M_t(x_{t-1},dx_t)$ the \textit{incremental weights} (i.e., the updating factor) and by $M_t^*(x_{t-1},dx_t)$ the optimal kernel.
+It can be proved\footnote{For a formal proof see \cite{**}, th.
+10.1 p.
+142.} that the optimal kernel, i.e., the one that minimizes the variance of the incremental weights $G_t$ (and of the time $t$ weights themselves), is:
+\begin{equation}
+    M_t^*(x_{t-1},dx_t)=\frac{1}{\int_{\X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)} P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)
+\end{equation}
+Note that $M_t^*$ is the conditional distribution of $x_t$ given $x_{t-1}$ and $y_t$.\footnote{In fact, assuming that the $P_t$ admits a density $p_t$, applying Bayes rule, we have that $m_t(x_t|x_{t-1},y_t)\propto p_t(x_t|x_{t-1})\cdot f(y_t|x_t)$.}
+The intuition for this result is apparent when considering equation \eqref{weight_updated_gpf}: if $M_t=M_t^*$, the updating factor collapses to $\int_{\X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)$ and information from the observed data are incorporated efficiently.\
+As a matter of fact, it is usually impossible to sample from the optimal kernel and, in practice,  it is replaced by its linear Gaussian approximation. \ \
+The GPF can be resumed by the following steps.
+\begin{enumerate}
+    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\M_0(dx_0)$ and set the un-normalized weights: 
+    $$w_0^n=\frac{\Pd_0(dx_0)}{\M_0(dx_0)}.$$
+    Then, the normalized weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
+    \item \textbf{For every $t=1,...T$:}
+    \begin{enumerate}
+        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
+\begin{enumerate}
+    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_n$.
+    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
+\end{enumerate}
+\textbf{Otherwise, } 
+\begin{enumerate}
+    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$
+    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
+\end{enumerate}
+\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
+\item Update the un-normalized weights, for $n=1,...,N$:
+\begin{equation*}
+    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
+\end{equation*}
+\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
+    \end{enumerate}
+\end{enumerate}
+Note that $\M_0(dx_0)$ is the initial sampling distribution, that could differ from the correct one $\Pd_0(dx_0)$.\ \
+To conclude, one of the issues related to the BPF is the fact that new states in $t$ are sampled ignoring their likelihood given the future observation $y_t$.
+Hence, extractions from the ``correct" transition kernel may be associated to low weights (small value for the likelihood).
+Instead, the GPF allows to solve for this problem, defining a proposal kernel that depends on the likelihood itself.
+Therefore, the extraction of future states already incorporates the information given by the data resulting in a more efficient procedure.
+
+\subsection{Auxiliary Particle Filter}
+
+The \textit{auxiliary particle filter} (APF) constitutes a further extension of the BPF.
+The GPF allowed to sample from a transition kernel different from the correct ones.
+The APF, instead, allows to extract the ancestor variables from an arbitrary distribution at the resampling step.\ \
+Let $\eta_t:\X\rightarrow \R_+$ be a non-negative, real-valued positive function, called \textit{auxiliary function}.
+At time $t$, when the resampling stage occurs, the multinomial distribution for the indexes to be extracted can use probabilities that differ from the weights in $t-1$, $W_{t-1}^n$.
+The new weights, called \textit{auxiliary weights}, are computed as a transformation of the original ones, denoted as \textit{inferential weights}, through the auxiliary function:
+\begin{equation}
+    \tilde W_t^n=\frac{W_t^n\cdot \eta_t(X_t^n)}{\sum_{m=1}^NW_t^m\cdot \eta_t(X_t^n)}
+\end{equation}
+Note that the computation of the auxiliary weights is done from the auto-normalized inferential ones.
+Additionally, note that imposing $\eta_t(x_t)=1$ we recover the GPF and the BPF as special cases of the APF.\
+Clearly, inferential weights are updated by incorporating the fact that resampling occurs according to different weights (the auxiliary ones):
+\begin{equation}
+    w_t^n=\frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}}G_t \label{apf_weights_up_}
+\end{equation}
+where $G_t$ is the incremental weight defined in the section on GPF.\ \
+
+To understand the logic of this sequential weight updating, consider  \eqref{weight_updated_gpf}.
+In particular, suppose that multinomial resampling takes place at every stage and let $\tilde w_{t-1}:=w_{t-1}\cdot \eta_{t-1} (x_{t-1})$.
+Then:
+\begin{equation}
+    \begin{split}
+        w_t\propto&\frac{\Pd_t(dx_{0:t}|y_{1:t})}{\M_t(dx_{0:t}|y_{1:t})}\propto \frac{\Pd_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
+        =&\underbrace{\frac{\Pd_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{w_{t-1}}\cdot\frac{1}{\tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
+        =&\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\label{weight_updated_apf}
+    \end{split}
+\end{equation}
+It is possible to derive an optimality result for the auxiliary functions\footnote{Details in ***** th 10.2, p.
+148}.
+In the special case in which the transition kernel used for sampling corresponds with the optimal one $M_t^*$, the optimal auxiliary function (i.e., the auxiliary function that minimizes the variance of the inferential weights) becomes:
+\begin{equation}
+    \eta_{t-1}^*(x_{t-1})=\int_\X P_t(x_{t-1},dx_t)f_t(y_t|x_t)
+\end{equation}
+and is called \textit{perfectly adapted} auxiliary function.
+In fact, if $M_t=M_t^*$, then, expression \eqref{weight_updated_apf} becomes:
+\begin{equation*}
+    \begin{split}
+        w_t\propto\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t^*(x_{t-1},dx_t)}=\frac{1}{\eta_{t-1} (x_{t-1})}\cdot\int_\X P_t(x_{t-1},dx_t)f_t(y_t|x_t)
+    \end{split}
+\end{equation*}
+From the previous equation, it is clear that if $\eta_t=\eta_t^*$, the inferential weights are constant and their variance is zero.\
+We can summarize the APF algorithm.
+\begin{enumerate}
+    \item \textbf{Inital stage: }
+    \begin{enumerate}
+        \item For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\M_0(dx_0)$ and set the un-normalized inferential weights: 
+    $$w_0^n=\frac{\Pd_0(dx_0)}{\M_0(dx_0)}.$$
+    Then, the normalized inferential weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
+    \item Compute the un-normalized auxiliary weights:
+        $$\tilde w_0^n=w_0^n \cdot \eta_0(X_0^n)$$
+        Then the normalized auxiliary weights are:
+        $\tilde W_0^n:=\frac{\tilde w_0^n}{\sum_{m=1}^N\tilde w_0^m}$.
+    \end{enumerate}
+    \item \textbf{For every $t=1,...T$:}
+    \begin{enumerate}
+        \item \textbf{Resample} if $ESS( W_{t-1}^{1:N})<ESS_{min}$:
+\begin{enumerate}
+    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(\tilde W_{t-1}^n)_n$.
+    \item Update weights as: $\hat w_{t-1}^n = \frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}} $ for $n=1,...,N$.
+\end{enumerate}
+\textbf{Otherwise, } 
+\begin{enumerate}
+    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$.
+    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
+\end{enumerate}
+\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
+\item Update the un-normalized inferential weights, for $n=1,...,N$:
+\begin{equation*}
+    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
+\end{equation*}
+\item Normalize the inferential weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
+\item Compute the un-normalized auxiliary weights, for $n=1,...,N$:
+\begin{equation*}
+ \tilde w_t^n=w_t^n \cdot \eta_t(X_t^n)
+\end{equation*}
+\item Normalize the auxiliary weights, for $n=1,...,N$:  $\tilde W_t^n:=\frac{\tilde w_t^n}{\sum_{m=1}^N\tilde w_t^m}$.\footnote{It is worth mentioning that the original APF included a final resampling step at the end of each iteration.
+However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance.
+Se ***Johansen p.
+120}
+    \end{enumerate}
+\end{enumerate}
+However, it is worth noting that the optimality results both for the kernel in the GPF and of the auxiliary  in the APF, has  two main weaknesses:
+\begin{itemize}
+    \item it is not clear that what is optimal at time $t$ remains such at future stages.
+In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affect the efficiency of the PF estimators;
+    \item both the choice of $M_t$ and of $\eta_t$ affect the way in which weights are computed and, indirectly, the decision to resample in the future.
+It is difficult to take also this indirect effect into account when evaluating the  overall efficiency of the algorithm.
+\end{itemize}
+
+\section{Liu and West Filter}\label{pf_lui_west}
+
+Let us consider a more general state-space model, where the state vector includes a time-constant parameter vector $\psi \in \Psi$, where $\Psi$ is the parameter space.
+Since we can interpret the parameter as a state with the law of motion $\psi_t=\psi_{t-1}=\psi$, then it is possible to adopt a PF algorithm to estimate $\psi$.\
+However, being $\psi$ constant over time, it is meaningless to sample its values sequentially: the first extraction of $\psi$ at the initial stage 0 (i.e., $\psi_0^n$) generates a constant path of particles for the parameter ($\psi_t^n=\psi_0^n$).
+\footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\
+Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution.
+In this way, at every time, the support of $\psi$ is not limited to the initially sampled $N$ values.
+\
+Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter.
+Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\
+Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot) $ be the prior distribution for the parameter $\psi$.\
+The LWF can be described by the following algorithm.
+\begin{enumerate}
+    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.
+
+    Then, set the un-normalized weights $w_0^n=1$.
+
+    Hence, the normalized weights are: $W_0^n=\frac{1}{N}$.
+    \item \textbf{For every $t=1,...T$:}
+    \begin{enumerate}
+        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
+\begin{enumerate}
+    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
+    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
+\end{enumerate}
+\textbf{Otherwise, } 
+\begin{enumerate}
+    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$
+    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
+\end{enumerate}
+\item \textbf{Extracting new parameter values:}
+\begin{enumerate}
+    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)^2$.
+    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$.
+    \item Draw new parameters: $\psi^n\thicksim \N \big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.
+\end{enumerate}
+\item Sample future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)$ for $n=1,...,N$.
+\item Update the un-normalized weights through the likelihood function, for $n=1,...,N$:
+\begin{equation*}
+    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n;\psi^n).
+\end{equation*}
+\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
+    \end{enumerate}
+\end{enumerate}
+At the initial step, $N$ states and $N$ parameter values are sampled independently, respectively, from $\Pd_0$ and $\pi$ and the particle weights are set to be constant, as usual.\
+Then, at each iteration (time $t$), the decision to resample is done according to the $ESS$ criterion and indexes are extracted from a multinomial distribution.
+However, in the LWF both previous periods states and parameters are resampled.\
+Meanwhile, to extract the next period states and parameters it is necessary, firstly, to sample the new $N$ parameter values.
+Let $\bar \psi$ be the weighted average of the parameter values and $\Omega$ be their variance (the weights used to compute this mean and variance are the previous period inferential ones, i.e.
+$W_{t-1}^n$).\ \
+The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$.
+While it would be more intuitively immediate to sample new parameters from a Normal density centered at the previously sampled values (i.e., $\psi^n$), this would increase the variance of $\psi$ over time.\
+In fact, let $\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \N(\psi^{A_t^{n}},\Omega)$\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.} be the empirical distribution of the parameters.
+From the law of iterated expectation and total variance, we obtain:\footnote{For completeness, we specify that the first expected value in the law of iterated expectations and total variances is with respect to the distribution of the indexes and the second one with respect to each of the Normal distributions centered at $\psi^{A_t^{n}}$.}
+\begin{equation*}
+    \begin{split}
+        \E_{\pi_{t-1}^N} ( \psi)=& \E \big[\E( \psi|A_t^{n})\big]=\E \big[\psi^{A_t^{n}}\big]=\bar \psi\\
+        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\E\big[\Omega\big]=2\cdot\Omega
+    \end{split}
+\end{equation*}
+Hence, while the mean is preserved, the variance increases at each iteration.
+Instead, sampling from the distribution proposed by Liu and West $\N \big(m^{A_t^n},h^2\cdot \Omega\big)$ preserves both the mean and the variance of the parameters over time.\
+ Let $\tilde \pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \N \big(m^{A_t^n},h^2\cdot \Omega\big)$. Hence:
+\begin{equation*}
+    \begin{split}
+        \E_{\tilde \pi_{t-1}^N} ( \psi)=& \E \big[\E( \psi|A_t^{n})\big]=\E \big[m^{A_t^n}\big]=\E \big[a\cdot \psi^{A_t^n}+(1-a)\cdot \bar \psi\big]=\bar \psi\\
+        Var_{\tilde \pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[m^{A_t^n}\big]+h^2\cdot\E\big[\Omega\big]=(\underbrace{a^2+h^2}_{=1})\cdot \Omega =\Omega
+    \end{split}
+\end{equation*}
+Finally, after sampling the new $N$ parameter values, conditional on these values, new states are also sampled from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)$.
+Lastly, weights are updated through the incremental weights.\ \
+To conclude, it is worth noting that the LWF produces, together with the usual filtering and prediction distributions, a posterior distribution for the parameter:
+\begin{equation}
+    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(\psi)
+\end{equation}
+
+\section{Convergence}\label{pf_converg}
 

### 2021-07-14 final draft.Rmd

@@
 
-We consider a stochastic volatility model
+We analyze a dataset of log close-to-close returns of the S&P500, from 2017 to 2021. Consider a stochastic volatility model
 $$
@@
 Overall, this analysis shows how a model with stochastic
-volatility is better at describing and predicting returns. As can be seen from the graph and perhaps unsurprisingly,
+volatility is better at describing and predicting the considered time-frame of S&P500 returns. As can be seen from the graph and perhaps unsurprisingly,
 the density forecast given the stochastic volatility model represents

### 2021-07-14 final draft.Rmd
sto aggiungendo le implementazioni

@@
 
+```{r include=FALSE}
+#Quantiles functions
+#-------------------
+q025=function(x){quantile(x,0.025)}
+q975=function(x){quantile(x,0.975)}
+
+#Calcolo intevalli di credibilità
+#--------------------------------
+Coverage<-function(fun){
+
+  Filt<-Filtervalues(fun)
+  mean<-Filt$mean
+  sd<-Filt$sd
+  vec<-NULL
+  for(t in 1:length(mean)){
+  if(mean[t]-1.96*sd[t]<truex[t] & truex[t]<mean[t]+1.96*sd[t]){vec[t]<-1}else{vec[t]<-0}
+  }
+  return(mean(vec))
+}
+
+#RMSE
+#----
+RMSE<-function(x,xhat){sqrt(mean((x - xhat)^2))}
+```
+
 \chapter*{Introduction}
@@
 
+\section{Implementation}
+
+In this section we present a practical illustration of the algorithms discussed.
+For the sake of simplicity we use a random walk plus noise model, i.e.
+the most basic form of a linear Gaussian state-space model.
+\begin{align}
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+x_{0} & \sim N(m_{0},C_{0})
+\end{align}
+As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
+However, this toy example will be used also to illustrate more involved filtering strategies described in this work.
+We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
+\
+The filtering strategy is applied to 50 simulated data.
+Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
+```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+set.seed(1991)
+dlm.sim = function(n,sigma,tau,x0){
+  x    = rep(0,n)  #state process
+  y    = rep(0,n)  #obs process
+  x[1] = rnorm(1,x0,tau)
+  y[1] = rnorm(1,x[1],sigma)
+  for (t in 2:n){
+    x[t] = rnorm(1,x[t-1],tau)
+    y[t] = rnorm(1,x[t],sigma)
+  }
+  return(list(x=x,y=y))
+}
+
+n=50
+tau2=1
+sigma2=1
+sigma=sqrt(sigma2)
+tau=sqrt(tau2)
+x0=0
+
+sim<-dlm.sim(n,sigma,tau,x0)
+y<-sim$y
+x<-sim$x
+truex<-x
+
+timeframe<-c(1:length(y))
+df<-data.frame(timeframe,y,x)
+
+ggplot(df,aes(x=timeframe))+
+  geom_line(aes(y=y, linetype="Observations"))+
+  geom_line(aes(y=x, linetype="True States"))+
+  coord_cartesian(ylim = c(-4, +12)) +
+  scale_linetype_manual(name="",
+                      values=c("Observations" = 1,
+                               "True States" = 3))+
+  labs(x="Time",
+       y="")+
+  #ggtitle("States and Observations")+
+  theme_bw()+
+  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
+
+The Kalman Filter for this model is implemented as described below.
+\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
+\begin{align*}
+x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
+a_{t} & = m_{t-1}\\
+R_{t} & = C_{t-1}+\tau^2
+\end{align*}
+\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e.
+the product of the one-step-ahead state predictive distribution and the likelihood
+\begin{align*}
+x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
+m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
+C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
+\end{align*}
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+Our \code{DLM} function implement in R replicate this steps.
+\
+\hrule
+\hrule
+\code{DLM(data,sig2,tau2,m0,C0)}
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{sig2} \ \ the variance $\sigma^{2}$ in Equation (1)
+\
+\code{tau2} \ \ the variance $\tau^{2}$ in Equation (2)
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\hrule
+\hrule
+```{r}
+DLM<-function(data,sig2,tau2,m0,C0){
+  n  = length(data)
+  m  = rep(0,n)
+  C  = rep(0,n)
+  for (t in 1:n){
+    if (t==1){
+      a = m0
+      R = C0 + tau2
+    }else{
+      a = m[t-1]
+      R = C[t-1] + tau2
+    }
+    A = R/(R+sig2)
+    m[t] = (1-A)*a + A*y[t]
+    C[t] = A*sig2
+  }
+  return(list(m=m,C=C))
+}
+```
+In Figure XX below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
+Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
+\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
+```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+#Kalman Filter
+#-------------
+m0=0
+C0=100
+filtval<-DLM(y,sigma2,tau2,m0,C0)
+m<-filtval$m
+C<-filtval$C
+Low = m + qnorm(0.025)*sqrt(C)
+Up = m + qnorm(0.975)*sqrt(C)
+
+#Put everything in a data frate (for ggplot)
+timeframe<-c(1:length(y))
+DLM.df<-data.frame(timeframe,y,x,m,C,Low,Up)
+
+#Plots (observetions in black and filtered states in red)
+ggplot(DLM.df,aes(x=timeframe))+
+  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=m, col="Filtered States", linetype="Filtered States"))+
+  geom_ribbon(aes(ymin = Low, ymax = Up),
+              fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("Observations" = "black",
+                               "True States" = "black",
+                               "Filtered States"= "red"))+
+  scale_linetype_manual("",
+                      values=c("Observations" = 1,
+                               "True States" = 3,
+                               "Filtered States"=1))+
+  labs(x="Time",
+       y="")+
+  #ggtitle("Kalman Filter")+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
+The discussion of Kalman Filter will continue in Section XX where we compare it to the Particle Filter.
+
+\textbf{Implementation}
+
+In this section we present a practical illustration of the algorithms discussed.
+For the sake of simplicity we use a random walk plus noise model, i.e.
+the most basic form of a linear Gaussian state-space model.
+\begin{align}
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+x_{0} & \sim N(m_{0},C_{0})
+\end{align}
+As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
+However, this toy example will be used also to illustrate more involved filtering strategies described in this work.
+We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
+\
+The filtering strategy is applied to 50 simulated data.
+Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
+```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+set.seed(1991)
+dlm.sim = function(n,sigma,tau,x0){
+  x    = rep(0,n)  #state process
+  y    = rep(0,n)  #obs process
+  x[1] = rnorm(1,x0,tau)
+  y[1] = rnorm(1,x[1],sigma)
+  for (t in 2:n){
+    x[t] = rnorm(1,x[t-1],tau)
+    y[t] = rnorm(1,x[t],sigma)
+  }
+  return(list(x=x,y=y))
+}
+
+n=50
+tau2=1
+sigma2=1
+sigma=sqrt(sigma2)
+tau=sqrt(tau2)
+x0=0
+
+sim<-dlm.sim(n,sigma,tau,x0)
+y<-sim$y
+x<-sim$x
+truex<-x
+
+timeframe<-c(1:length(y))
+df<-data.frame(timeframe,y,x)
+
+ggplot(df,aes(x=timeframe))+
+  geom_line(aes(y=y, linetype="Observations"))+
+  geom_line(aes(y=x, linetype="True States"))+
+  coord_cartesian(ylim = c(-4, +12)) +
+  scale_linetype_manual(name="",
+                      values=c("Observations" = 1,
+                               "True States" = 3))+
+  labs(x="Time",
+       y="")+
+  #ggtitle("States and Observations")+
+  theme_bw()+
+  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
+The discussion of Kalman Filter will continue in Section XX where we compare it to the Particle Filter.
+
 \chapter{Particle Filter}
@@
 
+\section{Implementation}
+Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
+The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm.
+We indicate with $n$ the sample size of time observations and with $N$ the generated sample size for each step of the Sequenial Monte Carlo.
+\begin{algorithm} SIS filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+Our \code{SISfun} function implemented in R replicate this steps.
+\
+\hrule
+\hrule
+\code{SISfun(data,N,m0,C0,tau,sigma)}
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\hrule
+\hrule
+```{r}
+SISfun<-function(data,N,m0,C0,tau,sigma){
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  for(t in 1:length(data)){
+    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
+    w    = w*dnorm(data[t],x,sigma)         #update weight
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+
+    wnorm= w/sum(w)                         #normalized weight
+    ESS  = 1/sum(wnorm^2)                   #effective sample size
+
+    ess =rbind(ess,ESS)
+  }
+
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+We have already discussed the reasons why the SIS algorithm does not provide a good strategy in the filtering problem.
+We provide a graphical intuition of what happens when we use such filtering strategy on a simulated dataset.
+We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$.
+The results shown in the following two plots shows a clear degeneration of the effective sample size and bad fit of filtered states with respect to the true values.
+```{r echo=FALSE}
+N=1000
+sis<-SISfun(y,N,m0,C0,tau,sigma)
+sisxs<-sis$xs
+sisws<-sis$ws
+sisess<-sis$ess
+sis.data<-data.frame(df,sisxs,sisws,sisess)
+
+ESS.SIS<-ggplot(sis.data,aes(x=timeframe))+
+  geom_line(aes(y=sisess))+
+    labs(x="Time",
+       y="")+
+  ggtitle("Effective Sample size")+
+  theme_bw()+
+  theme(plot.title = element_text(hjust = 0.5))
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) SIS Filtered States with credible interval (in red).
+b) Effective sample size.", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+Filtervalues<-function(fun){
+  xhat<-sapply(1:n,function(i)
+    weighted.mean(fun$xs[i,],fun$ws[i,]))
+  sdhat<-sapply(1:n,function(i)
+    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))
+
+  return(list(mean=xhat,sd=sdhat))
+
+}
+
+Filtplot<-function(dataframe,fun,title){
+
+Filt<-Filtervalues(fun)
+mean<-Filt$mean
+sd<-Filt$sd
+dataframe<-data.frame(dataframe,mean,sd)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=y, col="Observations", linetype="Observations"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+              fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("Observations" = "black",
+                               "True States" = "black",
+                               "Filtered States"= "red"))+
+  scale_linetype_manual("",
+                      values=c("Observations" = 1,
+                               "True States" = 3,
+                               "Filtered States"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+PlotSIS<-Filtplot(df,sis,title="")
+
+ggarrange(PlotSIS,ESS.SIS,labels=c("(a)","(b)"))
+```
+
 \subsection{Bootstrap Particle Filter}
@@
 \end{itemize}
+
+\section{Implementation}
+In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX.
+The overall strategy replicate the SIS filter with the addition of a ESS-based resampling step that applies when the effective sample size is smaller than a predetermined threshold opportunely chosen (in our example we decide to follow a common rule of thumb consisting in setting the threshold at $N/2$).
+The steps are presented in the following algorithm.
+\begin{algorithm} BPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+These steps are resumed in our \code{PFfun} function.
+\
+\hrule
+\hrule
+\code{PFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+```{r}
+PFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,x,tau)
+    w1<-w*dnorm(data[t],x,sigma)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+The estimated states of the Boostrap Particle Filter togheter with the effective sample size are shown in Figure XX.
+We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$.
+Notice how the resampling step allows the effective sample size not to drop, improving results.
+```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+N=1000
+
+pf<-PFfun(y,N,m0,C0,tau,sigma)
+
+ESSplot<-function(dataframe,fun,threshold){
+if(missing(threshold)){threshold=T}else{}
+xs<-fun$xs
+ws<-fun$ws
+ess<-fun$ess
+dt<-data.frame(dataframe,xs,ws,ess)
+
+ggplot(dt,aes(x=timeframe))+
+  geom_line(aes(y=ess))+
+  geom_line(aes(y=500),col="orange")+
+  labs(x="Time",
+       y="")+
+  ggtitle("Effective Sample Size")+
+  theme_bw()+
+  theme(plot.title = element_text(hjust = 0.5))
+}
+ESS.PF<-ESSplot(df,pf)
+
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) PF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.PF <- Filtplot(df,pf,"")
+ggarrange(FILT.PF,ESS.PF,labels=c("(a)","(b)"))
+```
+
+Since the Random Walk plus Noise model allows for closed form solutions, we want to compare the results of Boostrap Particle Filter(BPF) with Kalman Filter(KF).
+As we can see from figure XX, when the number of particles generated at any interaction increases the results for both the estimated mean and the variance tend to converge [[SPIEGARE PERCHE']].
+Moreover, comparing the Root Mean Square Errors of the filtered states with respect to the true state values, when the sample size increases the BPF decreases and, for large N, it reaches the accuracy of the Kalman Filter.
+
+```{r include=FALSE}
+truex<-x
+estimatedx<-matrix(NA,ncol=4,nrow=6)
+colnames(estimatedx)<-c("N","Threshold","KF","BPF")
+estimatedx[,1]<-c(100,1000,10000,1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
+
+i=1
+for(N in c(100,1000,10000)){
+
+DLM1<-DLM(y,sigma2,tau2,m0,C0)
+pf1<-PFfun(y,N,m0,C0,tau,sigma)
+
+estimatedx[i,3]<-RMSE(truex,DLM1$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
+i=i+1
+
+}
+
+i=4
+for(k in c(2,5,10)){
+  N=1000
+DLM2<-DLM(y,sigma2,tau2,m0,C0)
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+
+estimatedx[i,3]<-RMSE(truex,DLM2$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
+i=i+1
+
+}
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[1:3,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='ht', fig.width=10, message=FALSE, warning=FALSE, results=F}
+N1=100
+N2=1000
+N3=10000
+
+Combined<-data.frame(DLM.df,df)
+pf1<-PFfun(y,N1,m0,C0,tau,sigma)
+pf2<-PFfun(y,N2,m0,C0,tau,sigma)
+pf3<-PFfun(y,N3,m0,C0,tau,sigma)
+
+Compareplot<-function(dataframe,fun,title){
+
+Filt<-Filtervalues(fun)
+mean<-Filt$mean
+sd<-Filt$sd
+dataframe<-data.frame(dataframe,mean,sd)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=m, col="KF", linetype="KF"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=mean, col="BPF", linetype="BPF"))+
+  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+   #           fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("KF" = "red",
+                               "True States" = "black",
+                               "BPF"= "blue"))+
+  scale_linetype_manual("",
+                      values=c("KF" = 1,
+                               "True States" = 3,
+                               "BPF"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+p1<-Compareplot(Combined,pf1,"Filtered States")
+p2<-Compareplot(Combined,pf2,"")
+p3<-Compareplot(Combined,pf3,"")
+
+CompareVarplot<-function(dataframe,fun,title){
+
+Filt<-Filtervalues(fun)
+mean<-Filt$mean
+sd<-Filt$sd
+dataframe<-data.frame(dataframe,mean,sd)
+
+ggplot(dataframe,aes(x=timeframe))+
+geom_line(aes(y=sqrt(C), col="KF"))+
+geom_line(aes(y=sd,col="BPF"))+
+  scale_color_manual("",
+                      values=c("KF" = "red",
+                               "BPF"= "blue"))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30"))+
+  theme(plot.title = element_text(hjust = 0.5))
+
+}
+
+c1<-CompareVarplot(Combined,pf1,"Standard Deviation")
+c2<-CompareVarplot(Combined,pf2,"")
+c3<-CompareVarplot(Combined,pf3,"")
+
+ggarrange(p1,c1,p2,c2,p3,c3,nrow=3,ncol=2,labels=c("N=100","","N=1000","","N=10000",""))
+```
+\
 
@@
 
+\section{Implementation}
+\
+The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section XX can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$.
+In the Normal model we are considering, the optimal proposal will be a Normal density as well with mean and variance given by
+\begin{align*}
+\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
+\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
+\end{align*}
+On the other hand, the incremental weights, using this importance transition density, are proportional to the conditional density of $y_{t}$ given $x_{t-1}=x_{t-1}^{(i)}$ , i.e $N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})$, evaluated at $y_{t}$.
+In other words the algorithm implemented in R is:
+\begin{algorithm} GPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $\mu_{opt}$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+The \code{GPFfun} function resume this passages.
+\
+\hrule
+\hrule
+\code{GPFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+```{r}
+GPFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    means<-x+(tau/(tau+sigma))*(data[t]-x)
+    x<-rnorm(N,means,importancesd)
+    w1<-w*dnorm(data[t],x,predsd)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+```{r echo=FALSE}
+gpf<-GPFfun(y,N,m0,C0,tau,sigma)
+ESS.GPF<-ESSplot(df,gpf)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) GPF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.GPF<-Filtplot(df,gpf,"")
+ggarrange(FILT.GPF,ESS.GPF,labels=c("(a)","(b)"))
+```
+Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF).
+As we can see from the Figure XX, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE.
+Moreover, also the variance is better, suggesting for higher precision.
+
+```{r include=FALSE}
+truex<-x
+estimatedx<-matrix(NA,ncol=4,nrow=3)
+colnames(estimatedx)<-c("N","Threshold","BPF","GPF")
+estimatedx[,1]<-c(1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.25,0.1)
+
+i=1
+for(k in c(2,5,10)){
+  N=1000
+
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+gpf2<-GPFfun(y,N,m0,C0,tau,sigma,r=k)
+
+estimatedx[i,3]<-RMSE(truex,Filtervalues(pf2)$mean)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(gpf2)$mean)
+i=i+1
+
+}
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+N2=1000
+
+pf2<-PFfun(y,N2,m0,C0,tau,sigma)
+gpf2<-GPFfun(y,N2,m0,C0,tau,sigma)
+
+Comparetwofunplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=mean1, col="GPF", linetype="GPF"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=mean2, col="BPF", linetype="BPF"))+
+  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+   #           fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("GPF" = "green",
+                               "True States" = "black",
+                               "BPF"= "blue"))+
+  scale_linetype_manual("",
+                      values=c("GPF" = 1,
+                               "True States" = 3,
+                               "BPF"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+p2<-Comparetwofunplot(df,pf2,gpf2,"Filtered States")
+
+ComparetwoVarplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+geom_line(aes(y=sd2, col="GPF"))+
+geom_line(aes(y=sd1,col="BPF"))+
+  scale_color_manual("",
+                      values=c("GPF" = "green",
+                               "BPF"= "blue"))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30"))+
+  theme(plot.title = element_text(hjust = 0.5))
+
+}
+
+c2<-ComparetwoVarplot(Combined,pf2,gpf2,"Standard Deviation")
+
+ggarrange(p2,c2)
+```
+
 \subsection{Auxiliary Particle Filter}
@@
 \end{itemize}
+
+For illustration purposes, we are going implement an auxiliary particle filter with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.  
+\begin{algorithm} APF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item For $k=1,...,N$:
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+The \code{APFfun} function resume this passages.
+\
+\hrule
+\hrule
+\code{APFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+APFfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],x,sigma)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,x[k],tau)
+    lw  = dnorm(data[t],x1,sigma,log=TRUE)-dnorm(data[t],x[k],sigma,log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x1<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+```{r echo=FALSE}
+apf<-APFfun(y,N,m0,C0,tau,sigma)
+ESS.APF<-ESSplot(df,apf)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.APF<-Filtplot(df,apf,"")
+ggarrange(FILT.APF,ESS.APF,labels=c("(a)","(b)"))
+```
+
+Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap Particle Filter (BPF).
+
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+N2=1000
+
+pf2<-PFfun(y,N2,m0,C0,tau,sigma)
+apf2<-APFfun(y,N2,m0,C0,tau,sigma)
+
+Comparetwofunplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=mean1, col="APF", linetype="APF"))+
+  geom_line(aes(y=x, col="True States", linetype="True States"))+
+  geom_line(aes(y=mean2, col="BPF", linetype="BPF"))+
+  #geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+   #           fill="red",alpha=0.16) +
+    scale_color_manual("",
+                      values=c("APF" = "orange",
+                               "True States" = "black",
+                               "BPF"= "blue"))+
+  scale_linetype_manual("",
+                      values=c("APF" = 1,
+                               "True States" = 3,
+                               "BPF"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+p2<-Comparetwofunplot(df,pf2,apf2,"Filtered States")
+
+ComparetwoVarplot<-function(dataframe,fun1,fun2,title){
+
+Filt1<-Filtervalues(fun1)
+mean1<-Filt1$mean
+sd1<-Filt1$sd
+dataframe<-data.frame(dataframe,mean1,sd1)
+
+Filt2<-Filtervalues(fun2)
+mean2<-Filt2$mean
+sd2<-Filt2$sd
+dataframe<-data.frame(dataframe,mean2,sd2)
+
+ggplot(dataframe,aes(x=timeframe))+
+geom_line(aes(y=sd2, col="APF"))+
+geom_line(aes(y=sd1,col="BPF"))+
+  scale_color_manual("",
+                      values=c("APF" = "orange",
+                               "BPF"= "blue"))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30"))+
+  theme(plot.title = element_text(hjust = 0.5))
+
+}
+
+c2<-ComparetwoVarplot(Combined,pf2,apf2,"Standard Deviation")
+
+ggarrange(p2,c2)
+```
+
+```{r include=FALSE}
+truex<-x
+estimatedx<-matrix(NA,ncol=4,nrow=3)
+colnames(estimatedx)<-c("N","Threshold","BPF","APF")
+estimatedx[,1]<-c(1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.25,0.1)
+
+i=1
+for(k in c(2,5,10)){
+  N=1000
+
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
+
+estimatedx[i,3]<-RMSE(truex,Filtervalues(pf2)$mean)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(apf2)$mean)
+i=i+1
+
+}
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx[,], align="c", longtable = T, booktabs = T, caption="Root Mean Square Errors",digits=3)
+```
+
+```{r}
+APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+    weight = w*dnorm(data[t],x,predsd)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    }else{
+    weight = rep(1/N,N)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    }
+
+    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
+    x1   = rnorm(N,means,importancesd)
+    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    x <- x1
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+```{r echo=FALSE}
+apfopt<-APFoptfun(y,N,m0,C0,tau,sigma)
+ESS.APFopt<-ESSplot(df,apfopt)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.APFopt<-Filtplot(df,apfopt,"")
+ggarrange(FILT.APFopt,ESS.APFopt,labels=c("(a)","(b)"))
+```
 

### 2021-07-14 draft implementation 2.Rmd
change in section

@@
 As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
-However, this toy example will be used also to illustrate more involving filtering strategies described in this work.
+However, this toy example will be used also to illustrate more involved filtering strategies described in this work.
 We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
@@
 \
-\textbf{Implementation}
+\section{Implementation}
 Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
@@
 
-\textbf{Implementation}
+\section{Implementation}
 In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX.
@@
 \
-\textbf{Implementation}
+\section{Implementation}
 \

### 2021-07-14 Real Dataset Script.R
rmd Benedetta-Michele

@@ -303,7 +303,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=5000
+N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)

### 2021-07-15 final draft.Rmd
sto aggiungendo le implementazioni * da finire

@@
 - \newtheorem{algorithm}{Algorithm}[section]
+- \usepackage{float}
+- \floatplacement{figure}{}
 
@@
 Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
-```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 
@@
 \[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
-```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 #Kalman Filter
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="a) SIS Filtered States with credible interval (in red).
-b) Effective sample size.", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+b) Effective sample size.", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 Filtervalues<-function(fun){
@@
 Notice how the resampling step allows the effective sample size not to drop, improving results.
-```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 N=1000
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="a) PF Filtered States with credible interval (in red).
-b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 FILT.PF <- Filtplot(df,pf,"")
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='ht', fig.width=10, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='H', fig.width=10, message=FALSE, warning=FALSE, results=F}
 N1=100
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="a) GPF Filtered States with credible interval (in red).
-b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 FILT.GPF<-Filtplot(df,gpf,"")
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 N2=1000
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
-b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 FILT.APF<-Filtplot(df,apf,"")
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="a) APF Filtered States with credible interval (in red).
-b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 FILT.APFopt<-Filtplot(df,apfopt,"")
@@
 
+\section{Implementation}
+\
+Consider the linear Gaussian example of section XX, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
+Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
+\begin{align*}
+\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
+\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
+\end{align*}
+Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. The algorithm follows the following steps.
+\begin{algorithm} LWF for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
+$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:.
+\begin{enumerate}
+\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\begin{align*}
+m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
+\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\end{align*}
+\item For $k=1,...,N$:
+\begin{itemize}
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
+\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
+\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
+\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
+\end{itemize}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+Our \code{LWfun} function goes through the illustrated steps.
+\
+\hrule
+\hrule
+\
+\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alphav, betav} \ \  Gamma prior hyperparameters on $\sigma^{2}$
+\
+\code{alphaw,betaw} \ \  Gamma prior hyperparameters on $\tau^{2}$
+\
+\code{delta} \ \ hyperparameter delta value
+\
+\code{unif} \ \ if True then it sets a Uniform $(0,10)$ prior on $\sigma^{2}$ and $\tau^{2}$
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+```{r}
+LWfun<-function(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r){
+  if(missing(r)){r=2}else{}
+  xs     = rnorm(N,m0,sqrt(C0))
+  if(unif==T){
+  pars   = cbind(runif(N,0,10),runif(N,0,10))}else{}
+  pars   = cbind(rgamma(N,shape=alphav,scale=betav),rgamma(N,shape=alphaw,scale=betaw))
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,2,n))
+  xss    = NULL
+  ws     = NULL
+  ess    = NULL
+  w      = rep(1/N,N)
+  for (t in 1:length(data)){
+    meanV = weighted.mean(pars[,1],w)
+    varV  = weighted.mean((pars[,1]-meanV)^2,w)
+    meanW = weighted.mean(pars[,2],w)
+    varW  = weighted.mean((pars[,2]-meanW)^2,w)
+
+    muV = a*pars[,1]+(1-a)*meanV
+    sigma2V = (1-a^2)*varV
+    alphaV = muV^2/sigma2V
+    betaV = muV/sigma2V
+
+    muW = a*pars[,2]+(1-a)*meanW
+    sigma2W = (1-a^2)*varW
+    alphaW = muW^2/sigma2W
+    betaW = muW/sigma2W
+
+    weight      = w*dnorm(data[t],xs,sqrt(muV))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+
+    pars[,1]<-rgamma(N,shape=alphaV[k],rate=betaV[k])
+    pars[,2]<-rgamma(N,shape=alphaW[k],rate=betaW[k])
+
+    xsprevious<-xs[k]
+    xs = rnorm(N,xs[k],sqrt(pars[,2]))
+
+    w           = exp(dnorm( data[t],xs,sqrt(pars[,1]),log=T)-
+                        dnorm( data[t],xsprevious,sqrt(muV[k]),log=T))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xs[index]
+      pars<-pars[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xs
+      pars<-pars
+    }
+
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+    ess         = rbind(ess,ESS)
+  }
+  return(list(xss=xss,parss=parss,ws=ws,ess=ess))
+}
+```
+
 \section{Convergence}\label{pf_converg}

### 2021-07-15 final draft.Rmd
wf

--- 2021-07-14 final draft.Rmd
@@
 \section{State-Space Models}
+We now define \textit{State-Space models} as Markov processes that are partially observed.
+Let $X_t\in\X$ and $Y_t\in\Y$ and let $\{(X_t,Y_t)\}$ be a stochastic process.
+To help intuition, suppose that $\{Y_t\}$ is observable, i.e., coincides with a sequence of observations, while $\{X_t\}$ is latent.
+Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
+$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
+\begin{definition}\label{state-space-def}
+The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
+\begin{align*}
+  \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \P_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
+  &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
+  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t)].
+\end{align*}
+\end{definition}
+From the second inequality, it is apparent that $\{(X_t,T_t\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$
+
+Assume now that there exist a dominating measures $\nu(dy)$ such that 
+$F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
+Then by the last equality of definition \ref{state-space-ref}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
+\begin{equation}\label{state-space-densities}
+   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
+\end{equation}
+Integrating away $X_{0:T},$ one gets the marginal density
+\begin{align*}\label{state-space-marginal-y}
+  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\vu(dy_s) \\
+                &= p_t(y_{0:t})\prod\limits_{s=0}^t\vu(dy_s),
+\end{align*}
+where $p_t(y_{0:t})$ is the likelihood function.
+Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
+\begin{equation}
+\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
+\end{equation}
 
 \chapter{Linear-Gaussian case}
+\textit{Linear Gaussian models} are a class of popular state-space models.
+A notable property of such models is that the filtering and predictive distributions are analytically tractable, which makes such models very convenient in applications.
+
+Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
+$X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
+\begin{align}
+X_t &= A_tX_{t-1}+U_t \\
+Y_t &= B_tX_t + V_t.
+\end{align}
+Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
+\begin{lemma}\label{kalman-filter}
+  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Then, it holds that
+  \begin{align}
+    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
+    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
+    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+  \end{align}
+\end{lemma}
+\begin{proof}
+
+\end{proof}
+
+Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
+Then, using the transition equation of the latent variable, 
+by the linearity of the Gaussian distribution, we have that the predictive distribution is $\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).$
+Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
+\begin{align*}
+  \P_{t}(dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
+\end{align*}
 

### 2021-07-15 final-draft.tex
wf

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{book}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Final draft},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{txfonts}
\usepackage{amsmath}
\usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\usepackage{graphicx}
\usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
\usepackage{hyperref} %set the hyperlink to cite sections, equations...
\urlstyle{same}
\usepackage{comment}
\usepackage{enumitem}
\usepackage{ntheorem}
\usepackage{lipsum}
\usepackage{amssymb}
\usepackage{xcolor}
\theoremstyle{break}
\newtheorem{thm}{Theorem}% theorem counter resets every \subsection
\renewcommand{\thethm}{\arabic{thm}}
\newtheorem{proposition}{Proposition}
\newtheorem{assumption}{Assumption}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\theoremstyle{nonumberplain}
\newtheorem{proof*}{Proof.}
\newtheorem{definition}{Definition}
\renewcommand{\P}{\mathbb{P}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\setcounter{MaxMatrixCols}{10}
\setlength{\oddsidemargin}{.0in}
\setlength{\textwidth}{6 in}
\setlength{\topmargin}{-.1in}
\setlength{\textheight}{8.1in}
\renewcommand{\baselinestretch}{1.5}
\usepackage{amssymb}
\newcommand{\E}{\mathbb E}
\newcommand{\Q}{\mathbb Q}
\newcommand{\M}{\mathbb M}
\newcommand{\R}{\mathbb R}
\newcommand{\B}{\mathcal B}
\newcommand{\X}{\mathcal X}
\newcommand{\Pd}{\mathbb P}
\newcommand{\N}{\mathcal N}
\def\code#1{\texttt{#1}}
\newtheorem{algorithm}{Algorithm}[section]
\usepackage{float}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Final draft}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\frontmatter
\maketitle

\mainmatter
\chapter*{Introduction}

\chapter{Markov Processes and State-Space Models}
\section{Notation}

Let us fix the basic notation that will be employed henceforth. Upper
case letters, e.g.~\(X_t, Y_t,\) denote random variables, while lower
case letters denote the realizations. Finite sequences are made compact
using the semi-colon notation, e.g.~\(x_{0:t}=(x_0, x_1, \dots, x_t)\)
for some \(t \geq 0.\) The sets of outcomes are denoted by upper case
calligraphic letters, e.g.~\(\mathcal{X}, \mathcal{Y}.\) The
\(\sigma\)-algebra of a set \(\mathcal{X}\) reads
\(\mathcal{B}(\mathcal{X}).\) The pair
\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) is a measurable space. The
densities are represented with lower case cursive letters. For example,
the density of \(X_t\) in \(x_t\) is \(p_t^\theta(x_t),\) where
\(\theta\in\Theta\) is some (eventually known) vector, while the density
of \(X_t\) in \(x_t,\) given that \(X_{t-1}\) has previously taken value
\(x_{t-1},\) is \(p_t^\theta(x_t\mid x_{t-1}).\) Note that
\(p_t^\theta(x_t\mid x_{t-1})\) is a member of the parametric family
with parameters \((\theta, x_{t-1})\). The probability distribution of a
random variable \(X_t\) is represented by \(\mathbb{P}_t(dx_t),\) while
the probability distribution of a sequence \(X_{0:t}\) is represented by
\(\mathbb{P}_t(dx_{0:t}).\) The Dirac measure, i.e., the measure that
assigns probability \(1\) to the singleton \(\{x\},\) is denoted by
\(\delta_x(dy)\).The expectation operator is denoted by \(\mathbb{E},\)
where the section of \(\mathbb{E}\) at \(\mathbb{Q},\) i.e.,
\(\mathbb{E}_{\mathbb{Q}}\) denotes the operator computed with respect
to the probability distribution \(\mathbb{Q}.\) Let \(\varphi\) be a
function defined on the random variable \(X\propto\mathbb{Q}.\) Then, we
denote the expected value of \(\varphi(X)\) with \[
    \mathbb{E}_{\mathbb{Q}}[\varphi(X)]=
    \int_{\mathcal{X}}\varphi(x)\mathbb{Q}(dx).
\]

\section{Markov Processes}

Let \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) and
\((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
measurable spaces.

\begin{definition}
A function $P(x,dy): (\mathcal{X}, \mathcal{B}(\mathcal{X}))\rightarrow [0,1]$ such that 
\begin{itemize}
    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})$, and 
    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X})),$
\end{itemize}
is defined a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})).$
\end{definition}

Notice that for any random variables \(X_1\) and \(X_0=x_0,\) \[
    \mathbb{P}_1(dx_{0:1})=\mathbb{P}_0(dx_{0})P_1(x_0,dx_1),
\] where \(P_1(x_0,dx_1)\) is a probability kernel from
\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) This remark, provides some
intuition about the link between probability kernels and conditional
probabilities. As a matter of fact, it is possible to show that for
given \(X_{0:1}\) with joint probability distribution
\(\mathbb{P}_1(dx_{0:1}),\)
\begin{equation}\label{prob_kern_condit_prob}
    \mathbb{P}_1(X_1\in dx_1\mid X_0=x_0)=P_1(x_0,dx_1).
\end{equation}

Probability kernels can be used to define Markov processes. Before
moving to the definition, we will introduce some important concepts.
Recall that given two measures \(\mathbb{M}\) and \(\mathbb{Q}\) defined
on a measurable space \((\mathcal{X}, \mathcal{B}(\mathcal{X})),\)
\(\mathbb{Q}\) is absolutely continuous with respect to \(\mathbb{M},\)
that is \(\mathbb{Q}\ll\mathbb{M},\) if for all sets
\(A\in\mathcal{B}(\mathcal{X})\) such that \(\mathbb{M}(A)=0,\) then it
must also hold that \(\mathbb{Q}(A)=0.\) If
\(\mathbb{Q}(A)/\mathbb{M}(A)\) is well-defined for all \(A,\) then the
Radon-Nikodym theorem guarantees that there exists a measurable function
\(w(x)\geq0\) such that \[
    w(x)=\frac{\mathbb{Q}(dx)}{\mathbb{M}(dx)}, 
\] meaning that, \[
    \mathbb{Q}(A)=\int_A w(x) \mathbb{M}(dx).
\]

Let \(P_1, P_2, \dots, P_T\) be a finite measure of probability kernels
from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)

\begin{definition}
A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
\begin{equation}
    \mathbb{P}_T(X_{0:T}\in dx_{0:T})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},x_s).
\end{equation}
The set $\mathcal{X}$ is the state-space, $\mathbb{P}_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
\end{definition}

It is possible to show that for any \(t\in T,\) \[
\mathbb{P}_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
\mathbb{P}_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\mathbb{P}_t(x_{t-1},dx_t).
\] The first equality implies conditional independence. The second
equality implies that it is possible to identify conditional
distributions with transition kernels. Assume \(\mathbb{P}_t, t\leq T,\)
is a sequence of probability measures. Then, the following proposition
holds.

\begin{proposition}
For all $t\leq T,$ 
$$
\mathbb{P}_T(dx_{0:t})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\mathbb{P}_t(dx_{0:t}).
$$
\end{proposition}

This means that the marginal distribution of \(X_{0:t}\) with respect to
\(\mathbb{P}_T\) is \(\mathbb{P}_t.\)

\section{State-Space Models}

We now show that State-Space models can be understood as Markov
processes that are partially observed. Let \(X_t\in\mathcal X\) and
\(Y_t\in\Y\) and let \(\{(X_t,T_t\}\) be a stochastic process. Consider
the initial distribution \(\mathbb{P}_0(dx_0)\) and the probability
kernels \(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where
\(t=1,2,\dots\).

\begin{definition}
The process $\{(X_t,T_t\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
\begin{align*}
  \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \mathbb{P}_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
  &= [\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
  &= \mathbb{P}_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t)].
\end{align*}
\end{definition}

From the second inequality, it is apparent that \(\{(X_t,T_t\}\) is a
Markov process with initial distribution
\([\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\) and transition kernel
\(P_t(x_{t-1},dx_t)F_t(x_t,dy_t).\) The third equality, instead, shows
that, \(\{X_t\}\) has a marginal distribution
\(\mathbb{P}_T(dx_{0:T})=\mathbb{P}_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),\)
meaning that \(\{X_t\}\) is marginally distributed as a Markov process
with initial distribution \(\mathbb{P}_T(dx_{0:T})\) and transition
kernel \(P_t(x_{t-1},dx_t).\)

\chapter{Linear-Gaussian case}

\chapter{Particle Filter}

In the previous chapter we explained how the Kalman Filter is applied to
the case of a linear model with Gaussian errors to find closed-form
solution for the filtering and the predictive distributions.
Nonetheless, the majority of time-series models have a non-linear
structure and errors might follow non-Gaussian distributions.\\
In this chapter we describe \textit{particle filter} as an algorithm
that allows to update sequentially the filtering and the predictive
distributions in the non-linear and non-Gaussian case. Additionally, the
Kalman Filter can be seen as a special case of the more general particle
filter.\\
In section \ref{pf_intro} we introduce the fundamental notions of
\textit{importance sampling}, \textit{sequential importance sampling}
and \textit{resampling}. Then, in section \ref{pf_main} we describe
three particle filtering algorithms: the \textit{bootstrap}, the
\textit{guided} and the \{auxiliary particle filters\}. Instead, in
section \ref{pf_lui_west} we describe the \{Liu and West filter\}, when
particle filtering is used to estimate model's parameters. Finally, in
section \ref{pf_converg} we establish some general convergence results
for Particle Filter - Monte Carlo estimators.

\section{Introductory notions: sampling and resampling}\label{pf_intro}

Let \(\varphi: \mathcal X\rightarrow \mathbb R\) be a measurable,
integrable and bounded real-valued function, where
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) is a probability
space.\footnote{$\mathcal B(\mathcal X)$ is the Borel sigma-algebra of $\mathcal X$.}
Suppose that the probability measure \(\mathbb Q\) admits a density
\(q(x)\) with respect to some measure \(\nu(dx)\) and it is feasible to
sample from that density.\\
To estimate the expected value of \(\varphi\) with respect to the
density \(q\), the basic Monte Carlo (MC) method would require to sample
\(N\) times from the target density \(q\) and to compute the average of
the function \(\varphi\) values. \(\hat \varphi_{MC}\) is the standard
MC estimator: \begin{equation*}
   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \mathbb E_q(\varphi)~~~X^n\thicksim q(\cdot)
\end{equation*} However, in many circumstances either it is impossible
to sample from the target density or there exist another density - call
it \(m(\cdot)\) - different from \(q\) that produces
\texttt{more\ efficient"\ estimates.\ \ In\ all\ the\ cases\ for\ which\ sampling\ takes\ place\ from\ a\ distribution\ different\ from\ the}correct"
(i.e., the true) one, importance sampling and resampling methods can be
applied (as augmented versions of the standard MC algorithm).

\subsection{Importance sampling}

Let us consider two probability spaces:
\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) and
\((\mathcal X,\mathcal B(\mathcal X),\mathbb M)\) and let the measurable
function \(w:\mathcal X\rightarrow \mathbb R\) be proportional to the
Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\footnote{From the Radom-Nikodym theorem a sufficient condition for $w$ to exist is that $\mathbb Q$ is absolutely continuous with respect to $\mathbb M$: $\mathbb Q\ll \mathbb M$.}\\
The following result represents the starting point for the idea of
importance sampling.

\begin{lemma}\label{lemma_imp_sam_1}
For any measurable, integrable and bounded function $\varphi: \mathcal X\rightarrow \mathbb R$, the following holds:
\begin{equation}
    \mathbb E_\mathbb M(\varphi \cdot w)=\mathbb E_\mathbb Q(\varphi)\cdot \mathbb E_\mathbb M(w)
\end{equation}
\end{lemma}

\begin{proof*}
Note that $\mathbb Q(dx)=\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)$. Hence:
\begin{equation*}
    \mathbb E_\mathbb Q(\varphi)=\int_\mathcal X\varphi(x)\mathbb Q(dx)=\int_\mathcal X\varphi(x)\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)=\frac{\mathbb E_\mathbb M(\varphi \cdot w)}{\mathbb E_\mathbb M(w)}
\end{equation*}
$\hfill\blacksquare$
\end{proof*}

Lemma \ref{lemma_imp_sam_1} represents the key theoretical foundation of
\textbf{importance sampling} (IS), a numerical algorithm that allows to
sample from a \textit{proposal density} to estimate moments that are
computed with respect to the \textit{target density}, by means of
re-weighting the values of the target function \(\varphi\).\\
Adopting the previous notation, the target measure is \(\mathbb Q\),
while the proposal is \(\mathbb M\). IS requires to sample \(N\) values
from \(\mathbb M\), then to find the associated
\textit{normalized}\footnote{Weights are normalized when $\mathbb E_\mathbb M(W)=1$. In this case $W$ is the Radom-Nikodym derivative between $\mathbb Q$ and $\mathbb M$.}
weights \(W(X^n)\), finally to compute a weighted average of the values
taken by the function \(\varphi\). The IS is implemented according to
the following algorithm.

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
    \end{equation*}
\end{enumerate}

Note that \(\hat \varphi\) is an unbiased estimator for the target of
inference \(\mathbb E_\mathbb Q(\varphi)\). In fact: \begin{equation*}
    \begin{split}
        \mathbb E_\mathbb M(\hat \varphi)=\frac{1}{N} \sum_{n=1}^N \mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb Q(\varphi)
    \end{split}
\end{equation*} where the last equality follows from lemma
\ref{lemma_imp_sam_1} and the fact that weights are normalized to 1.

\subsubsection{Auto-normalized IS}

Usually, the Radom-Nikodym derivative is known up to a normalizing
constant and the IS algorithm has to be described in more general terms.
In the previous section the \textit{normalized} weighting function
\(W(\cdot)\) was the Radom-Nikodym derivative of \(\mathbb Q\) with
respect to \(\mathbb M\). More generally, consider the function
\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\), denoted as
\textit{importance function}, proportional to the derivative
\(W(\cdot)\).\\
In this case, weights have to be \textit{auto}-normalized to obtain the
IS - Monte Carlo estimator. The auto-normalized IS is implemented
according to the following procedure.

\begin{enumerate}
    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
    \item Normalize the weights, for $n=1,...,N$:
    \begin{equation*}
        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
    \end{equation*}
    \item Compute the estimator:
    \begin{equation*}
        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
    \end{equation*}
\end{enumerate}

While the auto-normalized estimator is biased (since \(W^n\) involves a
ratio of random variables), it is consistent (see \ref{pf_is_conv} for
details). ~\\
It is useful to interpret IS as an algorithm that allows to find an
approximation for the probability measure \(\mathbb Q\) as a weighted
average of Dirac measures for the sampled values (\(X^n\)):
\(\delta_{X^n}(dx)\) where the weights are the auto-normalized ones
(\(W^n\)). In particular: \begin{equation}
    \mathbb Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \mathbb M
\end{equation} Furthermore, the IS estimator is the expected value of
the function \(\varphi\) with respect to the approximating probability
measure \(\mathbb Q^N\): \begin{equation*}
    \begin{split}
        \mathbb E_{\mathbb Q^N}(\varphi)=&\int_\mathcal X\varphi(x)\mathbb Q^N(dx)=\int_\mathcal X\varphi(x)\sum_{n=1}^NW^n\delta_{X^n}(dx)=\\
       = &\sum_{n=1}^N\int_\mathcal X\varphi(x)W^n\delta_{X^n}(dx)=\sum_{n=1}^N\varphi (X^n)W^n=\hat \varphi_{AN}
    \end{split}
\end{equation*}

\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}

Consider the auto-normalized IS estimator
\(\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n\). We can write:
\begin{equation*}
   \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)=\mathbb E_{\mathbb Q^N}(\varphi)-\mathbb E_{\mathbb Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\mathbb E_{\mathbb Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
\end{equation*} Let
\(\bar \varphi(X):=\varphi(X)-\mathbb E_{\mathbb Q}(\varphi)\). Applying
the CLT to the numerator, assuming that
\(\mathbb E_{\mathbb M}(w^2\bar\varphi^2)<+\infty\), we get:
\begin{equation}
    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \mathcal N\big(0,\mathbb E_{\mathbb M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
\end{equation} where \(\implies\) denotes convergence in distribution.
Instead, for the denominator the strong LLN implies that:
\begin{equation}
    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \mathbb E_{\mathbb M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
\end{equation} Applying Slutsky's theorem, together with
\eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain: \begin{equation}
   \sqrt{N}\big[  \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)\big] \implies \mathcal N\Bigg(0,\frac{\mathbb E_{\mathbb M}(w^2\bar\varphi^2)}{[\mathbb E_{\mathbb M}(w)]^2}\Bigg)
\end{equation} Therefore, the IS (auto-normalized) estimator for
\(\varphi\) is both consistent and asymptotically normal.

\subsubsection{Effective Sample Size}

A common measure of efficiency for IS is the
\textit{effective sample size} (ESS), defined as: \begin{equation*}
    ESS(W^{1:N}):=\frac{1}{\sum_{n=1}^N(W^n)^2}=\frac{[\sum_{n=1}^N w(X^n)]^2}{\sum_{n=1}^N(w(X^n))^2}
\end{equation*} Observe that the \(ESS\) lies in the interval \([1,N]\).
For instance, if \(W^n=1/N\), as in the case of the standard MC method,
\(ESS=N\).\\
It is possible to relate the \(ESS\) to the variance of the importance
function. In fact: \begin{equation*}
    \frac{N}{ESS}=1+\frac{N^{-1}\sum_{n=1}^N(w(X^n))^2-\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}{\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}
\end{equation*} The second term in the numerator of the previous formula
is the square of the coefficient of variation (i.e., the ratio between
the variance and the squared mean) of the un-normalized weights. It is
clear that a lower \(ESS\) is associated with a higher variance for the
weights.

\subsection{Sequential Importance Sampling}

When dealing with state-space models, the importance sampling algorithm
has to be applied dynamically and the importance weights are updated
period-by-period. To illustrate the relevance of the
\textit{sequential importance sampling} (SIS) algorithm, we consider a
two-period stochastic process.~\\
Let us consider the approximation of the measure \(\mathbb Q_0\)
obtained at time 0 through IS from the measure \(\mathbb M_0\):
\begin{equation*}
    \mathbb Q_0^N(dx_0)=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0),~~\text{with } X_0^n\thicksim \mathbb M_0,~~\text{and }  W_0^n:=\frac{w_0(X_0^n)}{\sum_{m=1}^Nw_0(X_0^m)}.
\end{equation*} Moreover, let \(M_1(x_0,dx_1)\) be the transition
kernel. We want to update, sequentially, the approximating measure
\(\mathbb Q_0\) to obtain an approximation of the measure:
\begin{equation*}
    \mathbb Q_{1}(dx_{0:1})=\mathbb Q_0(dx_0)M_1(x_0,dx_1)
\end{equation*} SIS works according to the following algorithm.

\begin{enumerate}
    \item Sample $N$ values for $X_0$ from the proposal measure $\mathbb M_0$: $X_0^n\thicksim \mathbb M_0$ for $n=1,...,N$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^n,dx_1)$ for $n=1,...,N$.
    \item Compute the auto-normalized weights. In this case, they are computed taking into account only time 0 sampled observations since $X_1^n$ is sampled from the correct distribution: $W_1^n=W_0^n$.
\end{enumerate}

Therefore, the particle approximation of \(\mathbb Q_{1}(dx_{0:1})\) is:
\begin{equation*}
      \mathbb Q_{1}^N(dx_{0:1})=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0)M_1(X_0^n,dx_1)
\end{equation*}

\subsection{Sequential Importance Resampling}

A second approach can be used when sampling has to be done sequentially:
\textit{sequential importance resampling} (SIR). Differently from SIS,
an intermediate resampling stage takes place, in which time 0 values are
sampled with replacement according to a given resampling strategy
(usually, sampling indexes from a multinomial distribution with
probabilities equal to the weights computed through IS at time 0).\\
While in the SIS all the particles (i.e., the sampled values)
\texttt{survive",\ with\ the}risk" that particles associated with small
weights are carried-over (in the sense that sampling in period 1 occurs
from the kernel \(M_1(X_0^n,dx_1)\)), the same does not occur in SIR. In
this second algorithm, particles with larger weights are more likely to
be resampled and be used to ``generate" the particles in the next
period.\\
We describe the SIR more formally, focusing on the previous two-period
case.

\begin{enumerate}
    \item Sample $N$ values $X_0$ from the proposal measure $\mathbb M_0$: $X_0^n\thicksim \mathbb M_0$ for $n=1,...,N$.
    \item Compute the weights $W_0^n$ for $n=1,...,N$.
    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
\end{enumerate}

Steps 3-4 constitute the actual resampling ones. At this point, let us
define the new two-period sequence
\(\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)\). Then, the particle
approximation for the joint distribution of \((X_0,X_1)\) becomes:
\begin{equation*}
     \mathbb Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
\end{equation*} Note that, after resampling occurred, each particle
\(\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)\) has the same probability
weight.\\
To understand the reason of the previous statement, let us assume that
both \(\mathbb Q_0\) and \(\mathbb M_0\) admit densities \(q_0\),
\(m_0\) with respect to the same dominating measure \(\nu(dx_0)\) and
that the transition kernel \(M_1\) also admits the density \(m_1\).
Consider the sequential updating for the weights: \begin{equation}
    w^n_1=\frac{\mathbb Q_1(\tilde dx_{0:1})}{\mathbb M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
\end{equation} where \(\mathbb M_1\) denotes the probability measure for
the extraction of the particle \(\tilde x_{0:1}^n\) according to SIR.
\(\mathbb M_1\) (second equality sign in the previous formula) can be
rewritten as the product between the kernel for \(x_1^n\) given
\(x_0^{A_1^n}\) and the probability of \(x_0^{A_1^n}\) itself. This last
probability can also be rewritten as the product between the proposal
density \(m_0\) from which \(X_0\) is extracted and the probability that
it is sampled again when resampling occurs.

\subsection{SIS vs SIR}

The choice between the two sequential sampling strategies (SIS and SIR)
has significant implications in terms of computational cost and variance
of the importance weights. In fact, it is clear that SIR is more costly
computationally since it requires an intermediate resampling step.
Nonetheless, SIS suffers from the problem knonw as
\textit{curse of dimensionality}, i.e., the fact that the \(ESS\)
collapses over time as the variance of the weights diverges.\\
Consider an extension of SIS to a \(T\)-period model and let us assume
that the target and the proposal kernels admit densities, respectively,
\(q_t\) and \(m_t\) with respect to the same dominating measures (this
implies that the joint
distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation}
admit a density too). Then, weights, for every \(t=1,...,T\), are
updated according to: \begin{equation}
    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto 
    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
\end{equation} Equation \eqref{sis_update} suggests, intuitively, that
over time new variance is added from the incremental weights (i.e., the
updating factor \(q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)\)) determining
the curse of dimensionality issue. Instead, when resampling takes place,
since the past weights \(w_{t-1}\) are reset to be equal across them all
the variance of \(w_t\) inherited from the past is shut-down and the
only source of variability at each stage comes from the incremental
weights.

\section{Particle Filter}\label{pf_main}

\color{red}

Let us consider the standard state-space model introduced in the
previous chapters. \color{black} \textit{Particle filter} (PF) is a
numerical algorithm that is used to obtain approximations for the
filtering - \(\mathbb P_t (dx_t|y_{1:t})\) - and the prediction -
\(\mathbb P_t (dx_t|y_{1:t-1})\) - distributions for inference in
state-space models.~\\
In a state-space model sampling sequentially from the transition kernels
for the states is not enough to obtain good approximations since the
observation sample (\(y_{1:T}\)) contains information about the
underlying states through the likelihood function \(f_t(y_t|x_t)\).
Observations constitute signals for the model's states.\\
Therefore, the conditional distribution of the states, given the
observed data - \(\mathbb P_t(dx_{1:t}|y_{1:t})\) - changes over time as
new data points become available. The PF algorithm allows to incorporate
sequentially the new data to obtain new approximations for this
distribution (and for the filtering and prediction ones).\\
In general, at each step of the algorithm, states are sampled
sequentially from the proposal measure, call it \(\mathbb M_t\), and
importance auto-normalized weights are computed as the Radom-Nikodym
derivative of the target \(\mathbb P_t\) with respect to the proposal
\(\mathbb M_t\). Moreover, the Markov properties of the sequence of the
states allows to update sequentially the importance weights.~\\
We will analyse three different types of particle filtering procedures:
\textit{bootstrap}, \textit{guided} and
\textit{auxiliary particle filters}, that differ between them with
respect to their scope and efficiency.

\subsection{Degeneracy of the SIS in Particle Filter}

Usually, PF requires a resampling step at each stage of the algorithm.
Recall that, in general, resampling determines a trade-off between the
increase in the variance at the resampling stage and a decrease in the
variance when future states are sampled from the transition kernel.\\
\texttt{Dead"\ particles\ (i.e.,\ those\ with\ small\ weights)\ are\ \ likely\ to\ be\ removed\ when\ resampling\ takes\ place\ and\ past\ weights\ get}reset"
(usually, they are set to be equal to \(1/N\)). Otherwise, the
\textit{curse of dimensionality} problem will emerge and the variance of
the weights will diverge over time as the variance of past weights
increases as new updates occur period-by-period. Figure ***** shows an
example of a PF where SIS is used: note that the \(ESS\) drops
dramatically after few periods, meaning that the variance of the
importance weights diverges. This motivates the use of a SIR structure
for the PF algorithm.~\\
Lastly, it is now common practice to use a mixed approach with respect
to resampling (recall that resampling has an additional computational
cost). Generally, resampling is done \textit{adaptively}, when some
condition is met. A standard criterion that is used (and that we adopt
in our implementations of the PF) is to resample when the \(ESS\) drops
below some threshold \(ESS_{min}\in [1,N]\): \begin{equation}
    ESS(W_{t-1}^{1:N})<ESS_{min}
\end{equation}

\subsection{Bootstrap Particle Filter}

The bootstrap particle filter (BPF) is the easiest form of an adaptive
PF algorithm. In fact, future states are sampled from the
\textit{correct} transition kernel and the weights are updated
sequentially using the likelihood \(f_t(y_t|x_t)\).\\
We summarize the procedure with the following scheme.

\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb P_0(dx_0)$ and set the un-normalized weights $w_0^n=1$.\
    Then, the normalized weights are: $W_0^n=\frac{1}{N}$.
    \item \textbf{For every $t=1,...T$:}
    \begin{enumerate}
        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:\footnote{It is indifferent to implement this resampling step at the beginning or at the end of each iteration (as we do in our codes) as long as the initial stage $t=0$ generates weights that are constant and equal to $1/N$. In fact, in this case, resampling will never take place at the beginning of the first iteration since $ESS(W_0^n=1/N)=N$ that is larger than or equal to $ESS_{min}$.\
        However, when we describe the algorithm, we chose to leave this resampling step at the beginning of each iteration. In fact, when considering the guided or auxiliary PF, we can, more generally, sample the initial values (time $t=0$) for the state from a proposal distribution different from the target one. In this case, the weights at $t=0$ might differ from $1/N$, hence resampling might take place already at the beginning of the first iteration.}
\begin{enumerate}
    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
\end{enumerate}
\textbf{Otherwise, } 
\begin{enumerate}
    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$.
    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}^n$ for $n=1,...,N$.
\end{enumerate}
\item Sample the future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t)$  for $n=1,...,N$.
\item Update the un-normalized weights through the likelihood function for $n=1,...,N$.:
\begin{equation*}
    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n).
\end{equation*}
\item Normalize the weights  for $n=1,...,N$.: $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
    \end{enumerate}
\end{enumerate}

Therefore, at the initial step the pre-sample period states \(x_0^n\)
are extracted and every sampled state is assigned an equal weight since
there is no corresponding observation \(y_0\) that gives additional
information on the likelihood of that extraction.\\
Resampling does never take place in
\(t=1\)\footnote{Since $ESS(W_0^n=1/N)=N\geq ESS_{min}$.} and the state
in \(t=1\) is sampled from the correct transition kernel
\(P_1(x_0^n,dx_1)\). Now, the time 0 weights are updated incorporating
the information from the data point \(y_1\) in a very intuitive way:
more \textit{likely} states (given \(y_1\)) are given higher weights:
\(w_1^n=f_1(y_1|x_1^n)\). This weights are used to approximate the time
\(1\) filtering distribution.\\
At a subsequent stage, the \(ESS\) criterion should be checked and, in
case it is satisfied, resampling takes place and time \(1\) weights are
reset to be equal to \(1/N\).~\\
The key characteristic of the PF is the fact that the importance weights
are updated sequentially, according to the ``information" provided by
the observed data (i.e., the sequence \(y_{1:T}\)). Recall that time
\(t\) weight is the Radom-Nikodym derivative of the target with respect
to the proposal distribution. In the BPF, the time \(t\) target is the
measure \(\mathbb P_t(dx_{0:t}|y_{1:t})\), while the proposal, denoted
generically by \(\mathbb M_t\), is
\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\).
Hence: \begin{equation}
    \begin{split}
        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f(y_t|x_t)}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)}=\\
        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot f(y_t|x_t)=\hat w_{t-1}\cdot f(y_t|x_t)
    \end{split}\label{weight_updated_bpf}
\end{equation} Finally, the BPF is able to generate the two key objects
for inference in state-space models:

\begin{itemize}
    \item the predictive distribution:
    \begin{equation}
    \begin{split}
        \mathbb P_t^N (dx_t|y_{1:t-1})=&\frac{1}{\sum_{n=1}^N\hat w_t^n}\sum_{n=1}^N\hat w_t^n\delta_{X_t^n}(dx_t)=\\
        =&\begin{cases}
        \frac{1}{N}\sum_{n=1}^N\delta_{X_t^n}(dx_t)
        ~~~~~~~~~~~~~~~~ ~\text{ if resampling occurs}\\
        \frac{1}{\sum_{n=1}^N w_{t-1}^n}\sum_{n=1}^Nw_{t-1}^n\delta_{X_t^n}(dx_t)~~~~ \text{ otherwise}
        \end{cases},
        \end{split}
    \end{equation}
    \item the filtering distribution:
    \begin{equation}
        \mathbb P_t^N (dx_t|y_{1:t})=\sum_{n=1}^NW_t^n\delta_{X_t^n}(dx_t)
    \end{equation}
\end{itemize}

\subsection{Guided Particle Filter}

It is not always possible to sample sequentially from the correct
transition kernel. Moreover, even if it were possible, there exist a
proposal transition kernel that makes the algorithm more efficient. This
motivates the \textit{guided particle filter} (GPF), that can be
interpreted as a more general version of the GPF. In fact, with the GPF
it is possible, at each stage, to sample new states from a transition
kernel that differs from the correct one.~\\
Let \(P_t(x_{t-1},dx_t)\) be the correct kernel and consider a kernel
\(M_t(x_{t-1},dx_t)\) defined on the same measurable space of \(P_t\).
Moreover, let us assume that that \(P_t\) is absolutely continuous with
respect to
\(M_t\)\footnote{This condition is stronger than what would be needed for the existence of the importance weights in GPF, i.e., that  $P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)\ll M_t(x_{t-1},dx_t)$. }
(i.e., \(P_t\ll M_t\)). If sampling occurs sequentially from this
kernel, the way in which weights are updated should differ from the BPF.
Consider equation \eqref{weight_updated_bpf} and let us reformulate it
for the GPF case. Here, the proposal distribution is not
\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\)
anymore, but becomes
\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)\).
Hence: \begin{equation}
    \begin{split}
        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)}=\\
        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)} =\hat w_{t-1}\cdot\underbrace{\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}}_{\text{updating factor}}
    \end{split}\label{weight_updated_gpf}
\end{equation} Note that the updating factor is more complex that the
one of the BPF: in addition to the likelihood function \(f_t(y_t|x_t)\),
the fact that sampling occurs from a kernel different from the correct
ones requires the additional term
\(P_t(x_{t-1},dx_t)/M_t(x_{t-1},dx_t)\).~\\
We are now ready to establish, informally, the optimality result about
the transition kernel. Let us denote by
\(G_t=[P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)dy_t]/M_t(x_{t-1},dx_t)\) the
\textit{incremental weights} (i.e., the updating factor) and by
\(M_t^*(x_{t-1},dx_t)\) the optimal kernel. It can be
proved\footnote{For a formal proof see \cite{**}, th. 10.1 p. 142.} that
the optimal kernel, i.e., the one that minimizes the variance of the
incremental weights \(G_t\) (and of the time \(t\) weights themselves),
is: \begin{equation}
    M_t^*(x_{t-1},dx_t)=\frac{1}{\int_{\mathcal X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)} P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)
\end{equation} Note that \(M_t^*\) is the conditional distribution of
\(x_t\) given \(x_{t-1}\) and
\(y_t\).\footnote{In fact, assuming that the $P_t$ admits a density $p_t$, applying Bayes rule, we have that $m_t(x_t|x_{t-1},y_t)\propto p_t(x_t|x_{t-1})\cdot f(y_t|x_t)$.}
The intuition for this result is apparent when considering equation
\eqref{weight_updated_gpf}: if \(M_t=M_t^*\), the updating factor
collapses to \(\int_{\mathcal X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)\)
and information from the observed data are incorporated efficiently.\\
As a matter of fact, it is usually impossible to sample from the optimal
kernel and, in practice, it is replaced by its linear Gaussian
approximation. ~\\
The GPF can be resumed by the following steps.

\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb M_0(dx_0)$ and set the un-normalized weights: 
    $$w_0^n=\frac{\mathbb P_0(dx_0)}{\mathbb M_0(dx_0)}.$$
    Then, the normalized weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
    \item \textbf{For every $t=1,...T$:}
    \begin{enumerate}
        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
\begin{enumerate}
    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_n$.
    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
\end{enumerate}
\textbf{Otherwise, } 
\begin{enumerate}
    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$
    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
\end{enumerate}
\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
\item Update the un-normalized weights, for $n=1,...,N$:
\begin{equation*}
    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
\end{equation*}
\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
    \end{enumerate}
\end{enumerate}

Note that \(\mathbb M_0(dx_0)\) is the initial sampling distribution,
that could differ from the correct one \(\mathbb P_0(dx_0)\).~\\
To conclude, one of the issues related to the BPF is the fact that new
states in \(t\) are sampled ignoring their likelihood given the future
observation \(y_t\). Hence, extractions from the ``correct" transition
kernel may be associated to low weights (small value for the
likelihood). Instead, the GPF allows to solve for this problem, defining
a proposal kernel that depends on the likelihood itself. Therefore, the
extraction of future states already incorporates the information given
by the data resulting in a more efficient procedure.

\subsection{Auxiliary Particle Filter}

The \textit{auxiliary particle filter} (APF) constitutes a further
extension of the BPF. The GPF allowed to sample from a transition kernel
different from the correct ones. The APF, instead, allows to extract the
ancestor variables from an arbitrary distribution at the resampling
step.~\\
Let \(\eta_t:\mathcal X\rightarrow \mathbb R_+\) be a non-negative,
real-valued positive function, called \textit{auxiliary function}. At
time \(t\), when the resampling stage occurs, the multinomial
distribution for the indexes to be extracted can use probabilities that
differ from the weights in \(t-1\), \(W_{t-1}^n\). The new weights,
called \textit{auxiliary weights}, are computed as a transformation of
the original ones, denoted as \textit{inferential weights}, through the
auxiliary function: \begin{equation}
    \tilde W_t^n=\frac{W_t^n\cdot \eta_t(X_t^n)}{\sum_{m=1}^NW_t^m\cdot \eta_t(X_t^n)}
\end{equation} Note that the computation of the auxiliary weights is
done from the auto-normalized inferential ones. Additionally, note that
imposing \(\eta_t(x_t)=1\) we recover the GPF and the BPF as special
cases of the APF.\\
Clearly, inferential weights are updated by incorporating the fact that
resampling occurs according to different weights (the auxiliary ones):
\begin{equation}
    w_t^n=\frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}}G_t \label{apf_weights_up_}
\end{equation} where \(G_t\) is the incremental weight defined in the
section on GPF.~\\

To understand the logic of this sequential weight updating, consider
\eqref{weight_updated_gpf}. In particular, suppose that multinomial
resampling takes place at every stage and let
\(\tilde w_{t-1}:=w_{t-1}\cdot \eta_{t-1} (x_{t-1})\). Then:
\begin{equation}
    \begin{split}
        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{w_{t-1}}\cdot\frac{1}{\tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
        =&\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\label{weight_updated_apf}
    \end{split}
\end{equation} It is possible to derive an optimality result for the
auxiliary functions\footnote{Details in ***** th 10.2, p. 148}. In the
special case in which the transition kernel used for sampling
corresponds with the optimal one \(M_t^*\), the optimal auxiliary
function (i.e., the auxiliary function that minimizes the variance of
the inferential weights) becomes: \begin{equation}
    \eta_{t-1}^*(x_{t-1})=\int_\mathcal XP_t(x_{t-1},dx_t)f_t(y_t|x_t)
\end{equation} and is called \textit{perfectly adapted} auxiliary
function. In fact, if \(M_t=M_t^*\), then, expression
\eqref{weight_updated_apf} becomes: \begin{equation*}
    \begin{split}
        w_t\propto\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t^*(x_{t-1},dx_t)}=\frac{1}{\eta_{t-1} (x_{t-1})}\cdot\int_\mathcal XP_t(x_{t-1},dx_t)f_t(y_t|x_t)
    \end{split}
\end{equation*} From the previous equation, it is clear that if
\(\eta_t=\eta_t^*\), the inferential weights are constant and their
variance is zero.\\
We can summarize the APF algorithm.

\begin{enumerate}
    \item \textbf{Inital stage: }
    \begin{enumerate}
        \item For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb M_0(dx_0)$ and set the un-normalized inferential weights: 
    $$w_0^n=\frac{\mathbb P_0(dx_0)}{\mathbb M_0(dx_0)}.$$
    Then, the normalized inferential weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
    \item Compute the un-normalized auxiliary weights:
        $$\tilde w_0^n=w_0^n \cdot \eta_0(X_0^n)$$
        Then the normalized auxiliary weights are:
        $\tilde W_0^n:=\frac{\tilde w_0^n}{\sum_{m=1}^N\tilde w_0^m}$.
    \end{enumerate}
    \item \textbf{For every $t=1,...T$:}
    \begin{enumerate}
        \item \textbf{Resample} if $ESS( W_{t-1}^{1:N})<ESS_{min}$:
\begin{enumerate}
    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(\tilde W_{t-1}^n)_n$.
    \item Update weights as: $\hat w_{t-1}^n = \frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}} $ for $n=1,...,N$.
\end{enumerate}
\textbf{Otherwise, } 
\begin{enumerate}
    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$.
    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
\end{enumerate}
\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
\item Update the un-normalized inferential weights, for $n=1,...,N$:
\begin{equation*}
    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
\end{equation*}
\item Normalize the inferential weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
\item Compute the un-normalized auxiliary weights, for $n=1,...,N$:
\begin{equation*}
 \tilde w_t^n=w_t^n \cdot \eta_t(X_t^n)
\end{equation*}
\item Normalize the auxiliary weights, for $n=1,...,N$:  $\tilde W_t^n:=\frac{\tilde w_t^n}{\sum_{m=1}^N\tilde w_t^m}$.\footnote{It is worth mentioning that the original APF included a final resampling step at the end of each iteration. However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance. Se ***Johansen p. 120}
    \end{enumerate}
\end{enumerate}

However, it is worth noting that the optimality results both for the
kernel in the GPF and of the auxiliary in the APF, has two main
weaknesses:

\begin{itemize}
    \item it is not clear that what is optimal at time $t$ remains such at future stages. In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affect the efficiency of the PF estimators;
    \item both the choice of $M_t$ and of $\eta_t$ affect the way in which weights are computed and, indirectly, the decision to resample in the future. It is difficult to take also this indirect effect into account when evaluating the  overall efficiency of the algorithm.
\end{itemize}

\section{Liu and West Filter}\label{pf_lui_west}

Let us consider a more general state-space model, where the state vector
includes a time-constant parameter vector \(\psi \in \Psi\), where
\(\Psi\) is the parameter space. Since we can interpret the parameter as
a state with the law of motion \(\psi_t=\psi_{t-1}=\psi\), then it is
possible to adopt a PF algorithm to estimate \(\psi\).\\
However, being \(\psi\) constant over time, it is meaningless to sample
its values sequentially: the first extraction of \(\psi\) at the initial
stage 0 (i.e., \(\psi_0^n\)) generates a constant path of particles for
the parameter (\(\psi_t^n=\psi_0^n\)).
\footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\\
Therefore, Liu and West (2001) proposed a modified particle filter,
called \textit{Liu and West filter} (LWF) that allows to resample the
parameter over time from a continuous distribution. In this way, at
every time, the support of \(\psi\) is not limited to the initially
sampled \(N\) values.\\
Here, we describe the LWF that makes use of the normal distribution to
sample sequentially the parameter. Moreover, consider the simpler
version of the bootstrap
filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\\
Let the transition kernel and the likelihood depend on the parameter
\(\psi\) (i.e., \(P_t(x_{t-1},dx_t;\psi)\) and \(f_t(y_t|x_t;\psi)\))
and let \$\pi(\cdot) \$ be the prior distribution for the parameter
\(\psi\).\\
The LWF can be described by the following algorithm.

\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb P_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.

    Then, set the un-normalized weights $w_0^n=1$.

    Hence, the normalized weights are: $W_0^n=\frac{1}{N}$.
    \item \textbf{For every $t=1,...T$:}
    \begin{enumerate}
        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
\begin{enumerate}
    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
\end{enumerate}
\textbf{Otherwise, } 
\begin{enumerate}
    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$
    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
\end{enumerate}
\item \textbf{Extracting new parameter values:}
\begin{enumerate}
    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)^2$.
    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$.
    \item Draw new parameters: $\psi^n\thicksim \mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.
\end{enumerate}
\item Sample future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)$ for $n=1,...,N$.
\item Update the un-normalized weights through the likelihood function, for $n=1,...,N$:
\begin{equation*}
    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n;\psi^n).
\end{equation*}
\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
    \end{enumerate}
\end{enumerate}

At the initial step, \(N\) states and \(N\) parameter values are sampled
independently, respectively, from \(\mathbb P_0\) and \(\pi\) and the
particle weights are set to be constant, as usual.\\
Then, at each iteration (time \(t\)), the decision to resample is done
according to the \(ESS\) criterion and indexes are extracted from a
multinomial distribution. However, in the LWF both previous periods
states and parameters are resampled.\\
Meanwhile, to extract the next period states and parameters it is
necessary, firstly, to sample the new \(N\) parameter values. Let
\(\bar \psi\) be the weighted average of the parameter values and
\(\Omega\) be their variance (the weights used to compute this mean and
variance are the previous period inferential ones,
i.e.~\(W_{t-1}^n\)).~\\
The algorithm suggests to sample new values for the parameters from a
continuous distribution, i.e., a Normal centered at the (resampled)
values \(m^{A_t^n}\) with variance \(h^2\cdot \Omega\) where \(m^n\) is
a convex combination of the sample mean \(\bar\psi\) and the actual
extraction of \(\psi^n\) with coefficient \(a\in (0,1)\) and \(h\) is a
scalar such that \(a^2+h^2=1\). While it would be more intuitively
immediate to sample new parameters from a Normal density centered at the
previously sampled values (i.e., \(\psi^n\)), this would increase the
variance of \(\psi\) over time.\\
In fact, let
\(\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N(\psi^{A_t^{n}},\Omega)\)\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
be the empirical distribution of the parameters. From the law of
iterated expectation and total variance, we
obtain:\footnote{For completeness, we specify that the first expected value in the law of iterated expectations and total variances is with respect to the distribution of the indexes and the second one with respect to each of the Normal distributions centered at $\psi^{A_t^{n}}$.}
\begin{equation*}
    \begin{split}
        \mathbb E_{\pi_{t-1}^N} ( \psi)=& \mathbb E\big[\mathbb E( \psi|A_t^{n})\big]=\mathbb E\big[\psi^{A_t^{n}}\big]=\bar \psi\\
        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\mathbb E( \psi|A_t^{n})\big]+\mathbb E\big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\mathbb E\big[\Omega\big]=2\cdot\Omega
    \end{split}
\end{equation*} Hence, while the mean is preserved, the variance
increases at each iteration. Instead, sampling from the distribution
proposed by Liu and West
\(\mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)\) preserves both the
mean and the variance of the parameters over time.\\
Let
\(\tilde \pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)\).
Hence: \begin{equation*}
    \begin{split}
        \mathbb E_{\tilde \pi_{t-1}^N} ( \psi)=& \mathbb E\big[\mathbb E( \psi|A_t^{n})\big]=\mathbb E\big[m^{A_t^n}\big]=\mathbb E\big[a\cdot \psi^{A_t^n}+(1-a)\cdot \bar \psi\big]=\bar \psi\\
        Var_{\tilde \pi_{t-1}^N} ( \psi)=&Var\big[\mathbb E( \psi|A_t^{n})\big]+\mathbb E\big[Var( \psi|A_t^{n})\big]=Var\big[m^{A_t^n}\big]+h^2\cdot\mathbb E\big[\Omega\big]=(\underbrace{a^2+h^2}_{=1})\cdot \Omega =\Omega
    \end{split}
\end{equation*} Finally, after sampling the new \(N\) parameter values,
conditional on these values, new states are also sampled from the
transition kernel: \(X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)\).
Lastly, weights are updated through the incremental weights.~\\
To conclude, it is worth noting that the LWF produces, together with the
usual filtering and prediction distributions, a posterior distribution
for the parameter: \begin{equation}
    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(\psi)
\end{equation}

\section{Convergence}\label{pf_converg}

\chapter{Applications to Stochastic Volatility Models}
\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}

In this section, we apply the particle filter in a model-comparison
exercise, ultimately underlining the importance of taking stochastic
volatility into account in modeling equity returns.

We analyze a dataset of log close-to-close returns of the S\&P500, from
2017 to 2021. Consider a stochastic volatility model \[
\mathcal{M}_{SV}:\begin{array}{lc}
r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
\]

and a constant volatility iid model \[
\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
\]

We calibrate the model parameters using the whole sample, and proceed by
setting the return mean of both models to equal the sample mean,
\(\sigma^{2}\) equal to the return variance, and the parameters of the
volatility equation in \(\mathcal{M}_{SV}\) by running an \(AR(1)\)
regression of the log squared residuals.

In order to compare the predictive performance of the two, we compare
the cumulative log likelihood of the observed returns up to time \(t\)
given the considered model. In formulas, the cumulative likelihood can
be computed as \[
p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
\]

adopting the convention \(y_{1:0}=\emptyset\). This is trivial to obtain
in the iid case; considering the stohastic volatility model, the
predictive likelihoods can be derived as

\[
p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
\] \[
=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
\]

Whereas the emission distribution
\(p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})\) is known, we approximate
the predictive state distribution
\(p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})\) with weighted particles
obtained with particle filtering.

Figure XX reports the time series of the cumulative log-likelihood given
the constant volatility model \(\mathcal{M}_{CV}\) relative to the one
given \(\mathcal{M}_{SV}\), obtained as
\(\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})\).
Notice that the first difference of this series is the relative
log-likelihood of the incoming observation,
\(p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})\).

\begin{figure}[H]

{\centering \includegraphics{final-draft_files/figure-latex/cll_plot-1}

}

\caption{Relative log-likelihood}\label{fig:cll_plot}
\end{figure}

Overall, this analysis shows how a model with stochastic volatility is
better at describing and predicting the considered time-frame of S\&P500
returns. As can be seen from the graph and perhaps unsurprisingly, the
density forecast given the stochastic volatility model represents a
particularly significant improvement in the period of extreme values of
returns in March 2020.

\backmatter
\end{document}

### 2021-07-15 Real Dataset SV.Rmd

@@
 ---
-title: "SV application - real data"
-output: pdf_document
+header-includes:
+- \usepackage{babel} 
+- \usepackage{amsmath, amssymb, amsfonts}
+- \usepackage{graphicx}
+- \usepackage{mathtools}
+
+title: "Draft_Application_Stochastic_Volatility"
+
+documentclass: article
+
+output:
+  pdf_document: 
+   latex_engine: xelatex
+   fig_width: 5
+   fig_height: 3.5
+   fig_caption: true
+   extra_dependencies: ["float"]
+   number_sections: true
+
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+\section{SV models + Literary Review}
+In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
+Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
 
-## R Markdown
+This specification - which we will employ in this section on a real dataset - can be deemed as the standard version of the SV model.
+Throughout the recent decades, several extensions have been proposed, mostly (but not exclusively) in the field of financial econometrics.
+On the one hand, for simplicity we will stick to the standard version, one that will suffice for the scope of this work on Sequential Monte Carlo methods.
+On the other hand, it is worth mentioning some of these extensions: indeed, one can rather safely assume that, through their higher degree of sophistication, such extensions might well improve on the performance of our standard SV model in the analyses that will follow.\
+\
+Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility.
+Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as
+\begin{align*}
+    y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\tau \eta_t\\
+    x_1|\sigma,\beta & \sim  N\Big(\mu,\frac{\sigma^2}{1-\beta^2}\Big)\\
+    \epsilon_t & \sim N(0,1)\\ 
+    \eta_t & \sim N(0,1)
+\end{align*}
+where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
+Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated.
+This latter assumption justifies another definition of this specification, namely "discrete SV model without leverage".\footnote{Note that in this review we do not focus on continuous time SV models.
+In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing.
+Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility.
+Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al.
+(2010), who investigate alternatives to the entrenched affine square root SV model.
+Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties.
+More recently, Gatheral et al.
+(2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\
+In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007),
+\begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\mathbf{\Sigma}), \ \mathbf{\Sigma}=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}
+\end{align*}
+where $\rho<0$ captures the negative correlation.
+Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g.
+Harvey and Shephard 1996, Mao et al.
+2020).
+\
+Note that so far the assumptions on $\epsilon_t$ and $\eta_t$ implied that the returns are conditionally normally distributed.
+The discrete time SV models can also be extended to allow for heavy-tailed or asymmetric conditional returns distributions: symmetric or skewed Student-$t$, Generalised Hyperbolic (GH) distribution, Generalised Error Distribution (GED) and scale mixtures of normals feature as popular choices (Kim and Stoffer 2008, Nakajima and Omori 2012, Mao et al.
+2020).
+In fact, SV with heavy tailed return distributions were shown to better meet empirical regularities like the leptokurtic distribution of the returns and slowly decaying autocorrelation functions of the squared returns (Liesenfeld and Jung 2000).\
+Assuming that $\epsilon_t$ follows a Student-$t$ distribution, and exploiting the fact that $\epsilon_t$ can then be written as $\lambda_t^{-1/2}\zeta_t$, where $\zeta_t \sim N(0,1)$ and $v\lambda_t \sim \chi^2_v$ (Harvey et al.
+1994, Chib et al.
+2002),  we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004),
+\begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
+    x_{t+1} & = \alpha + \beta x_t + \eta_t\\
+    \begin{pmatrix}
+       \zeta_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\mathbf{\Sigma}), \ \mathbf{\Sigma}=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}\\
+    v\lambda_t & \sim \chi^2_v
+\end{align*}
+To capture other elements of the behaviour of financial data, several other extensions have been proposed.
+For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al.
+2020), allowing better goodness of fit and out-of-sample forecasts.\footnote{Dimitrakopoulos and Kolossiatis (2020) note that "the moving average component, the leverage effect and the conditional heteroscedasticity in mean have been considered separately in the stochastic volatility literature" and provide two specifications, one featuring an MA component and leverage effects, the other an MA component and conditional heteroskedasticity in mean.} \footnote{Other extensions, though ones for which we do not report the specifications, feature modelling the latent volatility process $x_t$ as an ARFIMA process (Long Memory Stochastic Volatility model, Breidt et al.
+1998) or as governed by a first-order Markov process (Markov Switching Stochastic Volatility model, So et al.
+1998).
+Recently, Luo et al.
+(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as
+\begin{align*}
+    y_t & = \nu_t+\psi e^{\frac{x_t}{2}}\epsilon_t\\
+    \nu_t & = a+by_{t-1}+d\psi^2e^{x_t}\\
+    x_t & = \beta x_{t-1}+\tau \eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} & \overset{i.i.d.}{\sim}  N_2\Big(\mathbf{0},\begin{pmatrix}
+       1 & 0 \\
+       0 & 1
+    \end{pmatrix}\Big)
+\end{align*}
+while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads
+\begin{align*}
+    y_t & = \nu_t + \gamma_t\\
+    \gamma_t & =\phi_1\gamma_{t-1}+...+\phi_p\gamma_{t-p}+u_t+\varphi_1 u_{t-1}+...+\varphi_q u_{t-q}\\
+    u_t|x_t & \sim N(0,e^{x_t})\\
+    x_{t} & = x_{t-1}+ \eta_t\\
+    \eta_t|\tau & \sim N(0,\tau^2)
+\end{align*}
+where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
+\
+Interestingly enough, the path along which the SV models evolved coincides with that suggested by optimal portfolio findings.
+Johannes, Korteweg and Polson (2014) found that, in order to generate statistically significant portfolio improvements in a Bayesian learning problem, the model employed by the investor should incorporate both time-varying expected returns and stochastic volatility: indeed, either of these features alone did not lead to statistically significant gains with respect to employing models with time-constant expected returns and volatility.\footnote{One could then argue that caution is needed when employing basic specifications of stochastic volatility models.
+For instance, Poon and Granger (2005) found that historical volatility and ARCH models both achieved better volatility forecasting performance than SV models.
+Similarly, Allen and McAleer (2020, see also Allen, 2020) found that, using realised volatility as benchmark, neither the canonical SV model or a GARCH(1,1) specification could forecast better than a simple form of historical volatility model.}\
+\
+Finally, before moving to the application of Sequential Monte Carlo techniques to a SV model, we conclude by going through some references for the Bayesian analysis proposed for such models.
+Starting from the seminal work of Jacquier et al.
+(1994), the use of MCMC methods has become increasingly popular for parameter estimation and smoothing exercises in SV models (e.g.
+Kastner 2019, presenting the R package \textit{stochvol} for Bayesian parameter estimation, and Chopin and Papaspiliopoulos 2020, who use MCMC to sample from the smoothing distribution of a SV model).
+As regards filtering exercises, the adoption of particle filters was rather rapid:\footnote{Actually, SV models are now often used as straightforward applications of particle filters on non-linear state space models, see for example Andrieu et al.
+(2010), Douc et al.
+(2014) or Chopin and Papaspiliopoulos (2020).} indeed, latent volatilities in Kim et al.
+(1998) were already filtered by employing the particle filter suggested in Pitt and Shephard (1999), paving the way for subsequent applications (\textit{inter alia}) in Chib et al.
+(2006), Omori et al.
+(2007), Kim and Stoffer (2008) and
+Nakajima and Omori (2012).\
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+\section{Data}
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+As previously mentioned, we will analyze the behaviour of the described filtering tools associated to a simple stochastic volatility model.
+In particular, we are evaluating the performance of such model on real data.
+We consider three indices S&P500, DOW JONES and STOXX50, in a time interval from Jan 1st, 2019 to June 30th, 2020, and from these, estimate the volatility computed from the model.
 
-```{r cars}
-summary(cars)
-```
+The indices have been selected as representatives of the global economic trends in the US and EU markets.
+Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exhange).
+The DOW JONES, instead, is a price-weighted stock-market index and accounts for the 30 major companies listed on US stock exchanges, characterized for being "blue-chip".
+Also the EURO STOXX 50 follows bluce-chip stocks representing leading firms in regions of the Eurozone.\
 
-## Including Plots
+The three plots below represent the time series of the log returns calculated for the three indices during our period of interest.
+In particular, each series is a sequence of daily observations representing the logarithm og the ratio between the closing price  of the index and the opening price of the index for a given day.
 
-You can also embed plots, for example:
+At a first visual inspection, we can see a common behaviour in the volatility of all three indices.
+Especially, it is worth pointing out the similarity between the pattern of DJI and that of S&P500 (especially in terms of peaks), which both differ from the STOXX50E, presumably because the first two describe the US stock market, while the last index describes the European one.
+For example, a difference that can be seen from a first glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations.
+Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.
 
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
+\textbf{Bibliography}\
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+Allen, D. E., & McAleer, M. (2020). Do we need stochastic volatility and generalised autoregressive conditional heteroscedasticity? Comparing squared end-of-day returns on FTSE. Risks, 8(1), 12.\
+
+Allen, D. E. (2020). Stochastic Volatility and GARCH: Do Squared End-of-Day Returns Provide Similar Information?. Journal of Risk and Financial Management, 13(9), 202.\
+
+Andrieu, C., Doucet, A., & Holenstein, R. (2010). Particle markov chain monte carlo methods. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 72(3), 269-342.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2001).
+Non‐Gaussian Ornstein–Uhlenbeck‐based models and some of their uses in financial economics.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
+
+Bollerslev, T. (1986). Generalized autoregressive conditional heteroskedasticity. Journal of econometrics, 31(3), 307-327.\
+
+Breidt, F. J., Crato, N., & De Lima, P. (1998). The detection and estimation of long memory in stochastic volatility. Journal of econometrics, 83(1-2), 325-348.\
+
+Chan, J. C. (2013). Moving average stochastic volatility models with application to inflation forecast. Journal of Econometrics, 176(2), 162-172.\
+
+Chib, S., Nardari, F., & Shephard, N. (2002). Markov chain Monte Carlo methods for stochastic volatility models. Journal of Econometrics, 108(2), 281-316.\
+
+Chib, S., Nardari, F., & Shephard, N. (2006). Analysis of high dimensional multivariate stochastic volatility models. Journal of Econometrics, 134(2), 341-371.\
+
+Chopin, N., & Papaspiliopoulos, O. (2020). An introduction to sequential Monte Carlo (Vol. 4). Springer.\
+
+Comte, F., & Renault, E. (1998). Long memory in continuous‐time stochastic volatility models. Mathematical finance, 8(4), 291-323.\
+
+Christoffersen, P., Jacobs, K., & Mimouni, K.
+(2010).
+Volatility dynamics for the S&P500: Evidence from realized volatility, daily returns, and option prices.
+The Review of Financial Studies, 23(8), 3141-3189.\
+
+Dimitrakopoulos, S., & Kolossiatis, M.
+(2020).
+Bayesian analysis of moving average stochastic volatility models: modeling in-mean effects and leverage for financial time series.
+Econometric Reviews, 39(4), 319-343.\
+
+Douc, R., Moulines, E., & Stoffer, D. (2014). Nonlinear time series: Theory, methods and applications with R examples. CRC press.\
+
+Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with estimates of the variance of United Kingdom inflation. Econometrica: Journal of the econometric society, 987-1007.\
+
+Eraker, B. (2004). Do stock prices and volatility jump? Reconciling evidence from spot and option prices. The Journal of Finance, 59(3), 1367-1403.\
+
+Friz, P. K., Gassiat, P., & Pigato, P. (2021). Precise asymptotics: robust stochastic volatility models. The Annals of Applied Probability, 31(2), 896-940.\
+
+Gatheral, J., Jaisson, T., & Rosenbaum, M. (2018). Volatility is rough. Quantitative finance, 18(6), 933-949.\
+
+Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\
+
+Harvey, A., Ruiz, E., & Shephard, N. (1994). Multivariate stochastic variance models. The Review of Economic Studies, 61(2), 247-264.\
+
+Heston, S. L. (1993). A closed-form solution for options with stochastic volatility with applications to bond and currency options. The review of financial studies, 6(2), 327-343.\
+
+Hull, J., & White, A. (1987). The pricing of options on assets with stochastic volatilities. The journal of finance, 42(2), 281-300.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. (1994). Bayesian analysis of stochastic volatility models. Journal of Business and Economic Statistics, 12(4), 371-389.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. E. (2004). Bayesian analysis of stochastic volatility models with fat-tails and correlated errors. Journal of Econometrics, 122(1), 185-212.\
+
+Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
+
+Kastner, G. (2019). Dealing with stochastic volatility in time series using the R package stochvol. arXiv preprint arXiv:1906.12134.\
+
+Kim, S., Shephard, N., & Chib, S. (1998). Stochastic volatility: likelihood inference and comparison with ARCH models. The review of economic studies, 65(3), 361-393.\
+
+Kim, J., & Stoffer, D.
+S.
+(2008).
+Fitting stochastic volatility models in the presence of irregular sampling via particle methods and the EM algorithm.
+Journal of time series analysis, 29(5), 811-833.\
+
+Koopman, S. J., & Hol Uspensky, E. (2002). The stochastic volatility in mean model: empirical evidence from international stock markets. Journal of applied Econometrics, 17(6), 667-689.\
+
+Liesenfeld, R., & Jung, R. C. (2000). Stochastic volatility models: conditional normality versus heavy‐tailed distributions. Journal of applied Econometrics, 15(2), 137-160.\
+
+Luo, R., Zhang, W., Xu, X., & Wang, J. (2018, April). A neural stochastic volatility model. In Thirty-second AAAI conference on artificial intelligence.\
+
+Mao, X., Czellar, V., Ruiz, E., & Veiga, H.
+(2020).
+Asymmetric stochastic volatility models: Properties and particle filter-based simulated maximum likelihood estimation.
+Econometrics and Statistics, 13, 84-105.\
+
+Nakajima, J., & Omori, Y.
+(2012).
+Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student’s t-distribution.
+Computational Statistics & Data Analysis, 56(11), 3690-3704.\
+
+Omori, Y., Chib, S., Shephard, N., & Nakajima, J. (2007). Stochastic volatility with leverage: Fast and efficient likelihood inference. Journal of Econometrics, 140(2), 425-449.\
+
+Pitt, M. K., & Shephard, N. (1999). Filtering via simulation: Auxiliary particle filters. Journal of the American statistical association, 94(446), 590-599.\
+
+Poon, S. H., & Granger, C. (2005). Practical issues in forecasting volatility. Financial analysts journal, 61(1), 45-56.\
+
+So, M. E. P., Lam, K., & Li, W. K. (1998). A stochastic volatility model with Markov switching. Journal of Business & Economic Statistics, 16(2), 244-253.\
+
+Xu, X., & Chen, Y. (2021). Deep Stochastic Volatility Model. arXiv preprint arXiv:2102.12658.\
+
+Zhang, B., Chan, J. C., & Cross, J. L. (2020). Stochastic volatility models with ARMA innovations: An application to G7 inflation forecasts. International Journal of Forecasting, 36(4), 1318-1328.

### 2021-07-15 final draft.Rmd
space
SV, adding comment on density forecasts
piccole correzioni

@@
 - \newcommand{\X}{\mathcal X}
+- \newcommand{\Y}{\mathcal Y}
 - \newcommand{\Pd}{\mathbb P}
@@
 
-Let $P_1, P_2, \dots, P_T$ be a finite measure of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
+Let $P_1, P_2, \dots, P_T$ be a finite sequence of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
 A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
 \begin{equation}
-    \P_T(X_{0:T}\in dx_{0:T})=\P_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},x_s).
+    \P_T(X_{0:T}\in dx_{0:T})=\P_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},dx_s).
 \end{equation}
@@
 The first equality implies conditional independence. The second equality implies that it is possible to identify conditional distributions with transition kernels. 
-Assume $\P_t, t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
+Assume $\P_t,$ with $t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
 \begin{proposition}
@@
   &= [\P_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
-  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t)].
+  &= \P_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
 \end{align*}
 \end{definition}
-From the second inequality, it is apparent that $\{(X_t,T_t\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$
-
-Assume now that there exist a dominating measures $\nu(dy)$ such that 
+From the second inequality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$
+
+Assume now that there exists a dominating measures $\nu(dy)$ such that 
 $F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)$ for all $t.$
-Then by the last equality of definition \ref{state-space-ref}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
+Then by the last equality of definition \ref{state-space-def}, the joint distribution of $(X_{0:T},Y_{0:T})$ is 
 \begin{equation}\label{state-space-densities}
@@
 Integrating away $X_{0:T},$ one gets the marginal density
-\begin{align*}\label{state-space-marginal-y}
-  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\vu(dy_s) \\
-                &= p_t(y_{0:t})\prod\limits_{s=0}^t\vu(dy_s),
+\begin{equation}\label{state-space-marginal-y}
+\begin{align*}
+  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
+                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
 \end{align*}                
+\end{equation}
 where $p_t(y_{0:t})$ is the likelihood function.
@@
 \end{lemma}
-\begin{proof}
-
-\end{proof}
 
@@
 Then, using the transition equation of the latent variable, 
-by the linearity of the Gaussian distribution, we have that the predictive distribution is $\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).$
-Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
+by the linearity of the Gaussian distribution, we have that the predictive distribution is 
+\begin{equation}
+\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
+\end{equation}
+Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
 \begin{align*}
-  \P_{t}(dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  \P_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
   m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.

### 2021-07-15 final draft.Rmd
La footnote è inutile, Radon Nikodym è già illustrato

@@
 
-Let us consider two probability spaces: $(\X,\B(\X),\Q)$ and $(\X,\B(\X),\M)$ and let the measurable function $w:\X \rightarrow \R$ be proportional to the Radom-Nikodym derivative between $\Q$ and $\M$, i.e.,  $w(x)\propto\frac{\Q(dx)}{\M(dx)}$.\footnote{From the Radom-Nikodym theorem a sufficient condition for $w$ to exist is that $\Q$ is absolutely continuous with respect to $\M$: $\Q\ll \M$.} \
+Let us consider two probability spaces: $(\X,\B(\X),\Q)$ and $(\X,\B(\X),\M)$ and let the measurable function $w:\X \rightarrow \R$ be proportional to the Radom-Nikodym derivative between $\Q$ and $\M$, i.e.,  $w(x)\propto\frac{\Q(dx)}{\M(dx)}$.
+\
 The following result represents the starting point for the idea of importance sampling.
@@
 exercise, ultimately underlining the importance of taking stochastic
-volatility into account in modeling equity returns.
+volatility into account in modeling equity returns. We follow XX, comparing density forecasts through relative predictive and cumulative likelihoods.
 

### 2021-07-15 Real Dataset Script.R
Minor changes

@@ -7,9 +7,14 @@
 #   w(t) ~ N(0,1)
 #   x(0) ~ N(0,100)
 #
-#   PRIOR ON alpha,beta,tau2
-
-#Particle Filter
+#   For LW, we have a prior on alpha,beta,tau2
+#
+#   alpha ~ N(ealpha,valpha)
+#   beta  ~ N(ebeta,vbeta)
+#   tau2  ~ IG(nu/2,nu*lambda/2)
+#   
+#
+#Sequential Importance Sampling
 #---------------
 SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}

### 2021-07-15 Real Dataset Script.R
Different names

@@ -45,7 +45,7 @@ SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 
-#Particle Filter
+# Bootstrap Particle Filter
 #---------------
 SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
@@ -78,7 +78,7 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 
 #
 #-----------------------
-#Particle Filter Optimal Kernel
+#Guided Particle Filter Optimal Kernel
 #------------------------------
 SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}

### 2021-07-15 Real Dataset Script.R
Resampling at each step

@@ -45,6 +45,37 @@ SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   return(list(xs=xs,ws=ws,ess=ess))
 }
 
+# Basis Particle Filter (Resampling at each step)
+#---------------
+SVBAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+  
+  for(t in 1:length(data)){
+    
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+    
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+    
+    if(ESS>-1){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+    
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
 # Bootstrap Particle Filter
 #---------------
 SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
@@ -308,8 +339,9 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=10000
+N=1000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
+svbapf<-SVBAPFfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-SVAPFfun(y,N,m0,C0,alpha,beta,tau)
 svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
@@ -366,6 +398,7 @@ plot3<-Filtplot(dfsv,svlw,"Liu and West Filter")
 plot4<-Filtplot(dfsv,svopt,"Opt Ker Particle Filter")
 plot5<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
 plot6<-Filtplot(dfsv,svsis,"No Resampling")
[…]

### 2021-07-15 Real Dataset Script.R
minor changes

@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=1000
+N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-SVBAPFfun(y,N,m0,C0,alpha,beta,tau)
 svpf<-SVPFfun(y,N,m0,C0,alpha,beta,tau)
@@ -412,7 +412,7 @@ ggarrange(plot7)
 
 realisedx<-x
 Errorvol<-matrix(NA,ncol=6,nrow=2)
-colnames(Errorvol)<-c("N","PF", "PFOPT", "APF","LWF", "SIS")
+colnames(Errorvol)<-c("N","BPF", "GPFOPT", "APF","LWF", "SIS", "PF")
 rownames(Errorvol)<-c("RMSE","MAE")
 Errorvol[,1]<-c(10000)
 RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
@@ -423,11 +423,13 @@ Errorvol[1,3]<-RMSE(realisedx,comparablevol(svopt))
 Errorvol[1,4]<-RMSE(realisedx,comparablevol(svapf))
 Errorvol[1,5]<-RMSE(realisedx,comparablevol(svlw))
 Errorvol[1,6]<-RMSE(realisedx,comparablevol(svsis))
+Errorvol[1,7]<-RMSE(realisedx,comparablevol(svbapf))
 Errorvol[2,2]<-MAE(realisedx,comparablevol(svpf))
 Errorvol[2,3]<-MAE(realisedx,comparablevol(svopt))
 Errorvol[2,4]<-MAE(realisedx,comparablevol(svapf))
 Errorvol[2,5]<-MAE(realisedx,comparablevol(svlw))
 Errorvol[2,6]<-MAE(realisedx,comparablevol(svsis))
+Errorvol[2,7]<-MAE(realisedx,comparablevol(svbapf))
 
 #Kalman filter
 m00=0

### 2021-07-15 Real Dataset SV.Rmd
Some changes

@@
    number_sections: true 
-
 ---
@@
 \section{SV models + Literary Review}
+
 In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
-Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
-\begin{align*}
+Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows, \begin{align*}
     y_t|x_t & \sim N(0,e^{x_t})\\
@@
 Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility.
-Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as
-\begin{align*}
+Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as \begin{align*}
     y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
@@
     \eta_t & \sim N(0,1)
-\end{align*}
-where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
+\end{align*} where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
 Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated.
@@
 (2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\
-In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007),
-\begin{align*}
+In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007), \begin{align*}
     y_t & = e^{\frac{x_t}{2}}\epsilon_t\\
@@
     \end{pmatrix}
-\end{align*}
-where $\rho<0$ captures the negative correlation.
+\end{align*} where $\rho<0$ captures the negative correlation.
 Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g.
 Harvey and Shephard 1996, Mao et al.
-2020).
-\
+2020).\
 Note that so far the assumptions on $\epsilon_t$ and $\eta_t$ implied that the returns are conditionally normally distributed.
@@
 1994, Chib et al.
-2002),  we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004),
-\begin{align*}
+2002), we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004), \begin{align*}
     y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
@@
     v\lambda_t & \sim \chi^2_v
-\end{align*}
-To capture other elements of the behaviour of financial data, several other extensions have been proposed.
+\end{align*} To capture other elements of the behaviour of financial data, several other extensions have been proposed.
 For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al.
@@
 Recently, Luo et al.
-(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as
-\begin{align*}
+(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as \begin{align*}
     y_t & = \nu_t+\psi e^{\frac{x_t}{2}}\epsilon_t\\
@@
     \end{pmatrix}\Big)
-\end{align*}
-while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads
-\begin{align*}
+\end{align*} while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads \begin{align*}
     y_t & = \nu_t + \gamma_t\\
@@
     \eta_t|\tau & \sim N(0,\tau^2)
-\end{align*}
-where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
+\end{align*} where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
 \
@@
 (2006), Omori et al.
-(2007), Kim and Stoffer (2008) and
-Nakajima and Omori (2012).\
+(2007), Kim and Stoffer (2008) and Nakajima and Omori (2012).\
 
@@
 In particular, we are evaluating the performance of such model on real data.
-We consider three indices S&P500, DOW JONES and STOXX50, in a time interval from Jan 1st, 2019 to June 30th, 2020, and from these, estimate the volatility computed from the model.
+We consider the continuously compounded returns (also called logarithmic returns) of three indices, S&P500, DOW JONES and STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021.
+From these, we estimate the daily volatility, as computed by the model.
 
 The indices have been selected as representatives of the global economic trends in the US and EU markets.
-Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exhange).
+Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exchange).
 The DOW JONES, instead, is a price-weighted stock-market index and accounts for the 30 major companies listed on US stock exchanges, characterized for being "blue-chip".
@@
 The three plots below represent the time series of the log returns calculated for the three indices during our period of interest.
-In particular, each series is a sequence of daily observations representing the logarithm og the ratio between the closing price  of the index and the opening price of the index for a given day.
+In particular, each series is a sequence of daily observations representing the logarithm of the ratio between the closing price of the index for a given day and the closing price of the day before.
 
@@
 Especially, it is worth pointing out the similarity between the pattern of DJI and that of S&P500 (especially in terms of peaks), which both differ from the STOXX50E, presumably because the first two describe the US stock market, while the last index describes the European one.
-For example, a difference that can be seen from a first glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations.
+For example, a difference that can be seen by having a glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations.
 Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.
@@
 (2001).
-Non‐Gaussian Ornstein–Uhlenbeck‐based models and some of their uses in financial economics.
+Non‐Gaussian Ornstein--Uhlenbeck‐based models and some of their uses in financial economics.
 Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
@@
 (2012).
-Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student’s t-distribution.
+Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student's t-distribution.
 Computational Statistics & Data Analysis, 56(11), 3690-3704.\

### 2021-07-15 Real Dataset SV.Rmd
Changes

@@
 In particular, we are evaluating the performance of such model on real data.
-We consider the continuously compounded returns (also called logarithmic returns) of three indices, S&P500, DOW JONES and STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021.
+For the observable process, namely financial returns in the SV model, we consider the continuously compounded daily returns (also called logarithmic returns) of three indices, S&P500, DOW JONES and STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021.
 From these, we estimate the daily volatility, as computed by the model.
@@
 Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.
+
+As regards the series that we will employ in section (\textit{fill with the number of the section}), namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.
+We considered a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimated based on 5 min intraday return intervals.
+Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
 
@@
 
+Heber, Gerd, Asger Lunde, Neil Shephard and Kevin Sheppard (2009).
+Oxford-Man Institute's realized library.
+Oxford-Man Institute, University of Oxford, version 0.3.
+Available from: https://realized.oxford-man.ox.ac.uk/.
+\
+
 Heston, S. L. (1993). A closed-form solution for options with stochastic volatility with applications to bond and currency options. The review of financial studies, 6(2), 327-343.\

### 2021-07-15 Real Dataset SV.Rmd
changes
a

@@
 As regards the series that we will employ in section (\textit{fill with the number of the section}), namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.
-We considered a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimated based on 5 min intraday return intervals.
+We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
@@
 Oxford-Man Institute, University of Oxford, version 0.3.
-Available from: https://realized.oxford-man.ox.ac.uk/.
+Available at: https://realized.oxford-man.ox.ac.uk/.
 \

### 2021-07-16 Real Dataset SV.Rmd
layout

@@
 In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
-Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows, \begin{align*}
+Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
+\begin{align*}
     y_t|x_t & \sim N(0,e^{x_t})\\
@@
 Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility.
-Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as \begin{align*}
+Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as
+\begin{align*}
     y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
@@
     \eta_t & \sim N(0,1)
-\end{align*} where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
+\end{align*} 
+where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
 Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated.
@@
     \end{pmatrix}
-\end{align*} where $\rho<0$ captures the negative correlation.
+\end{align*} 
+where $\rho<0$ captures the negative correlation.
 Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g.
@@
 1994, Chib et al.
-2002), we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004), \begin{align*}
+2002), we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004),
+\begin{align*}
     y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
@@
     v\lambda_t & \sim \chi^2_v
-\end{align*} To capture other elements of the behaviour of financial data, several other extensions have been proposed.
+\end{align*} 
+To capture other elements of the behaviour of financial data, several other extensions have been proposed.
 For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al.
@@
 Recently, Luo et al.
-(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as \begin{align*}
+(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as
+\begin{align*}
     y_t & = \nu_t+\psi e^{\frac{x_t}{2}}\epsilon_t\\
@@
     \end{pmatrix}\Big)
-\end{align*} while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads \begin{align*}
+\end{align*} 
+while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads 
+\begin{align*}
     y_t & = \nu_t + \gamma_t\\

### 2021-07-16 Real Dataset SV.Rmd
introduction

@@
 ---
+\section{Overview}
+In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data.
+However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques.
+Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.
+This part of the work is structured as follows:
 
@@
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
+\textit{Includere grafico e commento del grafico della realised volatility}
 

### 2021-07-16 Real Dataset SV.Rmd
aggiunta introduzione

@@
 ---
+
 \section{Overview}
@@
 Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.
-This part of the work is structured as follows:
+This part of the work is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied on them.
+In section \textit{fill with the number of the section} several particle filters are compared in terms of computational cost in approximating the filtering distribution implied by our SV model.
+Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
+In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.
 
@@
 
-In section (\textit{fill with the number of the section}) a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
+In section \textit{fill with the number of the section} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
 Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
@@
     \eta_t|\tau & \sim N(0,\tau^2)
-\end{align*} where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
+\end{align*} 
+where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
 \
@@
 
-As regards the series that we will employ in section (\textit{fill with the number of the section}), namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.
+As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.
 We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
@@
 \textit{Includere grafico e commento del grafico della realised volatility}
+
+\subsection{Transformations}
+Inserire le trasformazioni effettuate (e il perchè)
 

### 2021-07-16 Real Dataset SV.Rmd
ok

@@
 Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.
-This part of the work is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied on them.
+This part of the work is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
 In section \textit{fill with the number of the section} several particle filters are compared in terms of computational cost in approximating the filtering distribution implied by our SV model.

### 2021-07-16 Real Dataset SV.Rmd
update

@@
 This part of the work is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
-In section \textit{fill with the number of the section} several particle filters are compared in terms of computational cost in approximating the filtering distribution implied by our SV model.
+In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
 Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.

### 2021-07-16 Real Dataset Script.R
Qualche nome cambiato per maggiore chiarezza

@@ -47,7 +47,7 @@ SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 
 # Basis Particle Filter (Resampling at each step)
 #---------------
-SVBAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+BAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
   ws<-NULL
@@ -78,7 +78,7 @@ SVBAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 
 # Bootstrap Particle Filter
 #---------------
-SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+BPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
   ws<-NULL
@@ -111,7 +111,7 @@ SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 #-----------------------
 #Guided Particle Filter Optimal Kernel
 #------------------------------
-SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+GPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
   ws<-NULL
@@ -143,7 +143,7 @@ SVPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
 
 #Auxiliary Particle Filter
 #-------------------------
-SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+APFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
   xs<-NULL
   ws<-NULL
@@ -213,7 +213,7 @@ SVAPFfun_2<-function(data,N,m0,C0,alpha,beta,tau,r){
 
 #Auxiliary Particle Filter Optimal Kernel
 #----------------------------------------
-SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+APFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
[…]

### 2021-07-16 Real Dataset Script.R
ok

@@ -412,7 +412,7 @@ ggarrange(plot7)
 
 realisedx<-x
 Errorvol<-matrix(NA,ncol=6,nrow=2)
-colnames(Errorvol)<-c("N","BPF", "GPFOPT", "APF","LWF", "SIS", "PF")
+colnames(Errorvol)<-c("N","BPF", "GPFOPT", "APF","LWF", "SIS", "BAPF")
 rownames(Errorvol)<-c("RMSE","MAE")
 Errorvol[,1]<-c(10000)
 RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}

### 2021-07-16 Real Dataset Script.R
correzione

@@ -411,7 +411,7 @@ ggarrange(plot7)
 ## RMSE MAE comparison
 
 realisedx<-x
-Errorvol<-matrix(NA,ncol=6,nrow=2)
+Errorvol<-matrix(NA,ncol=7,nrow=2)
 colnames(Errorvol)<-c("N","BPF", "GPFOPT", "APF","LWF", "SIS", "BAPF")
 rownames(Errorvol)<-c("RMSE","MAE")
 Errorvol[,1]<-c(10000)

### 2021-07-16 Real Dataset Script.R
minor changes

@@ -396,17 +396,17 @@ plot1<-Filtplot(dfsv,svbpf,"Bootstrap Particle Filter")
 plot2<-Filtplot(dfsv,svapf, "Auxiliary Particle Filter")
 plot3<-Filtplot(dfsv,svlw,"Liu and West Filter")
 plot4<-Filtplot(dfsv,svgpfopt,"Opt Ker Guided Particle Filter")
-plot5<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
-plot6<-Filtplot(dfsv,svsis,"No Resampling")
-plot7<-Filtplot(dfsv,svbapf,"Always Resampling")
-ggarrange(plot1,plot2)
-ggarrange(plot1)
-ggarrange(plot2)
-ggarrange(plot3)
-ggarrange(plot4)
-ggarrange(plot5)
-ggarrange(plot6,plot1)
-ggarrange(plot7)
+plot5<-Filtplot(dfsv,svsis,"No Resampling")
+plot6<-Filtplot(dfsv,svbapf,"Always Resampling")
+plot7<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
+ggarrange(plot1,plot2,plot3,plot4)
+plot1
+plot2
+plot3
+plot4
+plot5
+plot6
+plot7
 
 ## RMSE MAE comparison
 

### 2021-07-16 Real Dataset Script.R
minor changes

@@ -399,7 +399,9 @@ plot4<-Filtplot(dfsv,svgpfopt,"Opt Ker Guided Particle Filter")
 plot5<-Filtplot(dfsv,svsis,"No Resampling")
 plot6<-Filtplot(dfsv,svbapf,"Always Resampling")
 plot7<-Filtplot(dfsv,svapfopt,"Opt Ker Auxiliary Particle Filter")
-ggarrange(plot1,plot2,plot3,plot4)
+ggarrange(plot1,plot2)
+ggarrange(plot3,plot4)
+ggarrange(plot5,plot6)
 plot1
 plot2
 plot3

### 2021-07-16 final draft.Rmd
proof kalman

@@
 \end{lemma}
+\begin{proof*}
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, Q_0).$ It is easy to see that
+\begin{equation} \label{normal-distrib-rewrite}
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
+\end{equation}
+with $A=Q_0^{-1}$ and $B=Am_0.$
+Now, by Bayes' Theorem,
+
+    \begin{align*}
+        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
+                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
+                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
+                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
+    \end{align*}
+
+Hence, it follows that 
+\begin{equation*}
+    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
+\end{equation*}
+and 
+    \begin{align*}
+        m_1 &=\E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
+            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+    \end{align*}
+\end{proof*}
 

### 2021-07-16 Real Dataset Script.R
aggiunto commento

@@ -311,7 +311,7 @@ alpha = -0.0031
 beta  =  0.9951
 tau2  =  0.0074
 m0      = 0
-C0      = 100
+C0      = 100 #N(0,100) is an informative prior about the mean of the volatility. It is used by Jacquier et al. (2004), and it is calibrated on percentage returns
 sC0     = sqrt(C0)
 ealpha  = alpha
 valpha  = 0.01

### 2021-07-16 Real Dataset SV.Rmd
Aggiunte

@@
 \section{Overview}
+
 In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data.
@@
 
-As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.
-We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
+As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \textit{fill with the number of the section}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
@@
 
-\subsection{Transformations}
-Inserire le trasformazioni effettuate (e il perchè)
+\subsection{Rescaling the data}
+
+When specifying the parameters of a stochastic volatility model, one should clarify whether the series of observed financial returns is one of log returns or of percentage log returns.
+Indeed, let $y_t$ denote the returns at time $t$, so that a basic SV specification would prescribe $y_t|x_t\sim N(0,e^{x_t})$: clearly, for each period $t$, $e^{x_t}$ needs to be higher to replicate the observed \textit{percentage} log return rather than the observed log return.
+This implies that the latent volatility $x$ should fluctuate around higher values when using percentage log returns as $y_t$.
+This feature, when adopting the aforementioned canonical SV model as defined in Kim et al.
+(1998), would be represented by a higher $\mu$.
+For example, as noted in Kastner (2019), daily log returns often have a variance of 0.0001 or less, which implies that $\mu$ should lie somewhere around $\ln(0.0001)\approx -9$; instead, when considering daily percentage log returns, one then has a variance of 1, so that $\mu=\ln(1)=0$.
+Note that when $\mu=0$, the canonical model boils down to the basic specification we employed on the simulated dataset, so that we prefer using percentage log returns for the 3 indices.
+In other terms, one might end up with the basic specification of the SV model when (roughly) calibrating the parameters of the canonical model on a wider set of financial returns.
+
+Note that rescaling the log returns has analogous implications for the section in which we employ measures of realised volatility as a proxy for the latent volatility process.
+Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database.
+Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.
+Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula, che considera returns, non returns in percentuale.
+Spiegare qui perchè possa avere senso paragonare questa varianza a quella dei log returns} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
+Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.
+To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$ (\textit{In realtà detto così è poco preciso, perchè in realtà andiamo a confrontare con la trasformazione della media dello stato filtrato}).
 

### 2021-07-17 Real Dataset SV.Rmd
corrections

@@
        \eta_t
-    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\mathbf{\Sigma}), \ \mathbf{\Sigma}=
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
     \begin{pmatrix}
@@
 For example, as noted in Kastner (2019), daily log returns often have a variance of 0.0001 or less, which implies that $\mu$ should lie somewhere around $\ln(0.0001)\approx -9$; instead, when considering daily percentage log returns, one then has a variance of 1, so that $\mu=\ln(1)=0$.
-Note that when $\mu=0$, the canonical model boils down to the basic specification we employed on the simulated dataset, so that we prefer using percentage log returns for the 3 indices.
-In other terms, one might end up with the basic specification of the SV model when (roughly) calibrating the parameters of the canonical model on a wider set of financial returns.
+In other terms, one might end up with a basic specification of the SV model when (roughly) calibrating the parameters of the canonical model on a wider set of financial returns.
 
@@
 Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula, che considera returns, non returns in percentuale.
-Spiegare qui perchè possa avere senso paragonare questa varianza a quella dei log returns} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
+Spiegare qui perchè possa avere senso paragonare questa varianza a quella dei log returns, come fa Guidolin} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
 Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.

### 2021-07-17 Real Dataset Script.R

@@ -307,7 +307,7 @@ LWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
 #Set hyperparameters
 #-------------------
 #n     =  length(y)
-alpha = -0.0031
+alpha =  -0.0031
 beta  =  0.9951
 tau2  =  0.0074
 m0      = 0
@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=10000
+N=5000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)

### 2021-07-17 Real Dataset Script.R
commit

@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=5000
+N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)

### 2021-07-18 Real Dataset Script.R
small

@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=10000
+N=1000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
@@ -365,7 +365,7 @@ Filtplot<-function(dataframe,fun,title){
   Filt<-Filtervalues(fun)
   mean<-Filt$mean
   mean1=exp(mean/2)
-  sd<-Filt$sd
+  sd<-Filt$sd ##qua c'è il problema che la sd si riferisca allo stato non trasformato...
   dataframe<-data.frame(dataframe,mean1,sd)
   
   ggplot(dataframe,aes(x=timeframe))+
@@ -497,6 +497,14 @@ ggplot(DLM.df,aes(x=timeframeKF))+
   theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
         legend.background = element_rect(fill = "white", colour = "gray30")) +
   theme(plot.title = element_text(hjust = 0.5))
+
+## Save mean and sd series
+library(openxlsx)
+Filt<-Filtervalues(svbpf)
+mean<-Filt$mean
+sd<-Filt$sd
+write.xlsx(mean,file="svbpf50000p.xlsx",sheet=1)
+write.xlsx(sd,file="svbpf50000p.xlsx",sheet=2)
 
 
 

### 2021-07-18 Real Dataset Script.R
Comparison with Bootstrap 50000 particles

@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=1000
+N=50000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
@@ -348,7 +348,7 @@ svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 svgpfopt<-GPFoptfun(y,N,m0,C0,alpha,beta,tau)
 svapfopt<-APFoptfun(y,N,m0,C0,alpha,beta,tau)
 
-#Filtering Values and Plot
+#Filtering Values
 #-------------------------
 Filtervalues<-function(fun){
   xhat<-sapply(1:n,function(i)
@@ -359,6 +359,8 @@ Filtervalues<-function(fun){
   return(list(mean=xhat,sd=sdhat))
   
 }
+
+#Graphical comparison of filtered volatility
 
 Filtplot<-function(dataframe,fun,title){
   
@@ -433,6 +435,77 @@ Errorvol[2,5]<-MAE(realisedx,comparablevol(svlw))
 Errorvol[2,6]<-MAE(realisedx,comparablevol(svsis))
 Errorvol[2,7]<-MAE(realisedx,comparablevol(svbapf))
 
+# Save mean and sd series of bootstrap with 50000 particles
+library(openxlsx)
+Filt<-Filtervalues(svbpf)
+mean<-Filt$mean
+sd<-Filt$sd
+list1<-list('mean'=mean,'sd'=sd)
+write.xlsx(list1,file="svbpf50000p.xlsx")
+
+#RMSE and MAE comparison of particle filters (but not LW) vs Bootstrap PF with 50000 particles
+
+library(openxlsx)
+bpf50000mean<-read.xlsx("svbpf50000p.xlsx", sheet = 1)
+bpf50000sd<-read.xlsx("svbpf50000p.xlsx", sheet = 2)
+zmean<-bpf50000mean[,1]
+zsd<-bpf50000sd[,1]
+dfsv1<-data.frame(timeframe,y,zmean,zsd)
+Errorcomp<-matrix(NA,ncol=6,nrow=2)
[…]

### 2021-07-18 Real Dataset Script.R
commit

@@ -339,13 +339,13 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=50000
+N=10
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
 svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
-svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 svgpfopt<-GPFoptfun(y,N,m0,C0,alpha,beta,tau)
+svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 svapfopt<-APFoptfun(y,N,m0,C0,alpha,beta,tau)
 
 #Filtering Values
@@ -454,20 +454,20 @@ dfsv1<-data.frame(timeframe,y,zmean,zsd)
 Errorcomp<-matrix(NA,ncol=6,nrow=2)
 colnames(Errorcomp)<-c("N","BPF", "GPFOPT", "APF", "SIS", "BAPF")
 rownames(Errorcomp)<-c("RMSE","MAE")
-Errorvol[,1]<-c(10000)
+Errorcomp[,1]<-c(10000)
 RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
 MAE<-function(x,xhat){mean(abs(x-xhat))}
 comparablemean<-function(fun){Filtervalues(fun)$mean}
-Errorvol[1,2]<-RMSE(bpf50000mean,comparablemean(svbpf))
-Errorvol[1,3]<-RMSE(bpf50000mean,comparablemean(svgpfopt))
-Errorvol[1,4]<-RMSE(bpf50000mean,comparablemean(svapf))
-Errorvol[1,5]<-RMSE(bpf50000mean,comparablemean(svsis))
-Errorvol[1,6]<-RMSE(bpf50000mean,comparablemean(svbapf))
-Errorvol[2,2]<-MAE(bpf50000mean,comparablemean(svbpf))
-Errorvol[2,3]<-MAE(bpf50000mean,comparablemean(svgpfopt))
-Errorvol[2,4]<-MAE(bpf50000mean,comparablemean(svapf))
[…]

### 2021-07-18 Real Dataset Script.R
esperimenti 50000 particles

@@ -435,9 +435,9 @@ Errorvol[2,5]<-MAE(realisedx,comparablevol(svlw))
 Errorvol[2,6]<-MAE(realisedx,comparablevol(svsis))
 Errorvol[2,7]<-MAE(realisedx,comparablevol(svbapf))
 
-# Save mean and sd series of bootstrap with 50000 particles
+# Save mean and sd series of PF with 50000 particles
 library(openxlsx)
-Filt<-Filtervalues(svbpf)
+Filt<-Filtervalues(svapf)
 mean<-Filt$mean
 sd<-Filt$sd
 list1<-list('mean'=mean,'sd'=sd)
@@ -448,8 +448,12 @@ write.xlsx(list1,file="svbpf50000p.xlsx")
 library(openxlsx)
 bpf50000mean<-read.xlsx("svbpf50000p.xlsx", sheet = 1)
 bpf50000sd<-read.xlsx("svbpf50000p.xlsx", sheet = 2)
+apf50000mean<-read.xlsx("svapf50000p.xlsx", sheet = 1)
+apf50000sd<-read.xlsx("svapf50000p.xlsx", sheet = 2)
 zmean<-bpf50000mean[,1]
 zsd<-bpf50000sd[,1]
+wmean<-apf50000mean[,1]
+wsd<-apf50000mean[,1]
 dfsv1<-data.frame(timeframe,y,zmean,zsd)
 Errorcomp<-matrix(NA,ncol=6,nrow=2)
 colnames(Errorcomp)<-c("N","BPF", "GPFOPT", "APF", "SIS", "BAPF")
@@ -468,6 +472,8 @@ Errorcomp[2,3]<-MAE(zmean,comparablemean(svgpfopt))
 Errorcomp[2,4]<-MAE(zmean,comparablemean(svapf))
 Errorcomp[2,5]<-MAE(zmean,comparablemean(svsis))
 Errorcomp[2,6]<-MAE(zmean,comparablemean(svbapf))
+RDiff50000p<-RMSE(zmean,wmean)
+ADiff50000p<-MAE(zmean,wmean)
 
 # Graphical comparison of filtered states
 

### 2021-07-18 Real Dataset Script.R
Commit

@@ -453,8 +453,8 @@ apf50000sd<-read.xlsx("svapf50000p.xlsx", sheet = 2)
 zmean<-bpf50000mean[,1]
 zsd<-bpf50000sd[,1]
 wmean<-apf50000mean[,1]
-wsd<-apf50000mean[,1]
-dfsv1<-data.frame(timeframe,y,zmean,zsd)
+wsd<-apf50000sd[,1]
+dfsv1<-data.frame(timeframe,y,zmean,zsd,wmean,zsd)
 Errorcomp<-matrix(NA,ncol=6,nrow=2)
 colnames(Errorcomp)<-c("N","BPF", "GPFOPT", "APF", "SIS", "BAPF")
 rownames(Errorcomp)<-c("RMSE","MAE")
@@ -505,6 +505,7 @@ Filtplot1<-function(dataframe,fun,title){
           legend.background = element_rect(fill = "white", colour = "gray30")) +
     theme(plot.title = element_text(hjust = 0.5))
 }
+
 library(ggplot2)
 library(ggpubr)
 plot8<-Filtplot1(dfsv1,svbpf,"Bootstrap Particle Filter")
@@ -517,6 +518,35 @@ plot9
 plot10
 plot11
 plot12
+
+# Plot the two 50000 particles benchmarks
+Filtplot2<-function(dataframe,title){
+  
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=zmean, col="50000p BT Filtered States", linetype="50000p BT Filtered States"))+
+    geom_ribbon(aes(ymin = zmean-1.96*zsd, ymax = zmean+1.96*zsd),
+                fill="black",alpha=0.16) +
+    geom_line(aes(y=mean, col="50000p AP Filtered States", linetype="50000p AP Filtered States"))+
+    geom_ribbon(aes(ymin = wmean-1.96*wsd, ymax = wmean+1.96*wsd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("50000p BT Filtered States" = "black",
+                                "50000p AP Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("50000p BT Filtered States" = 1,
+                                   "50000p AP Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
[…]

### 2021-07-18 Real Dataset Script.R
N=10000

@@ -339,7 +339,7 @@ dfsv<-data.frame(timeframe,y,x)
 #Filtering
 #---------
 set.seed(12345)
-N=10
+N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)

### 2021-07-18 Real Dataset Script.R
commit

@@ -368,6 +368,7 @@ Filtplot<-function(dataframe,fun,title){
   mean<-Filt$mean
   mean1=exp(mean/2)
   sd<-Filt$sd ##qua c'è il problema che la sd si riferisca allo stato non trasformato...
+  ## un'idea potrebbe essere di trasformare prima tutte le particles, calcolando solo poi media e standard deviation
   dataframe<-data.frame(dataframe,mean1,sd)
   
   ggplot(dataframe,aes(x=timeframe))+

### 2021-07-18 Real Dataset Script.R
commit

@@ -368,7 +368,7 @@ Filtplot<-function(dataframe,fun,title){
   mean<-Filt$mean
   mean1=exp(mean/2)
   sd<-Filt$sd ##qua c'è il problema che la sd si riferisca allo stato non trasformato...
-  ## un'idea potrebbe essere di trasformare prima tutte le particles, calcolando solo poi media e standard deviation
+  ## un'idea potrebbe essere di trasformare prima tutte le particles, calcolando solo poi media e standard deviation...ma questo porta a una diversa media
   dataframe<-data.frame(dataframe,mean1,sd)
   
   ggplot(dataframe,aes(x=timeframe))+

### 2021-07-18 Real Dataset SV.Rmd
commit

@@
 Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.
-This part of the work is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
+This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
 In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
@@
 
-\section{Data}
+\section{Data Description}
 
@@
 
-\textbf{Bibliography}\
+\section{Visual Analysis after Filtering}
+
+In this section, we offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the volatility of the observations obtained through the application of the different filtering techniques explored in the theoretical analysis.
+Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows.
+For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
+Note that the number of particles was set at 10000 for each filter.} \
+
+First, a general point is that the sequence of percentage log returns are different across the three indices.
+The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents (SEE SECTION X).
+In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy.
+However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy.
+Aside from this note, the considerations expressed below on the sequences "filtered volatilities"\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.} for the various techniques hold almost uniformly for all the three indices.
+Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\
+
+Generally speaking, note that, the sequence of "filtered volatilities" obtained from any of the filtering techniques follows roughly similar paths across all the plots.
+Unsurprisingly, the filtered volatility appears wider in periods corresponding to wider fluctuations of the percentage log returns and lower in periods corresponding to smaller such fluctuations.
+For example, all sequences of filtered volatilities show peaks corresponding to the higher fluctuations due to the 2020 Covid-19 Crisis outbreak and the consequent increases in uncertainty for the later periods.\
+
+When focusing the attention on specific filters, first we can discuss the effect of resampling.
+Indeed, in the plot referring to the Sequential Importance Sampling procedure, the sequence of filtered volatilities is not able to reproduce the peaks as in with other filters, but also is characterized by collapsing confidence intervals.
+This is because ::::::: \
+Then, if we instead look at the filtered volatilities obtained with resampling, we can appreciate how the filtered sequences for the Basic,\footnote{By this name, we refer to a particle filter that features no adaptive resampling; rather, resampling is conducted at each step of the algorithm.} Bootstrap, Auxiliary and Optimal Kernel Guided particle filters give rather analogous results, with no differences appearing at a first visual inspection.\footnote{Except for the Auxiliary Particle filtered sequence of volatilities, showing a slightly higher peak corresponding to the greatest volatility fluctuation in 2020, if compared to other filters in this subset.
+This is still valid for all three indices.} \
+However, the sequence of filtered volatilities obtained by applying the Liu and West filter is clearly different from the others.
+Indeed, it shows a "more detailed" sequence of states (in the sense of the sequence of filtered volatilities being less smooth) and also a higher peak for the wide fluctuation corresponding to the 2020 Covid Crisis outbreak.
+Therefore, we may say that, compared to other filtering techniques, it offers a better representation of the data.
+This is because ...
+\textit{Potremmo parlare di maggiore sensibilità alle oscillazioni dei prezzi, o simili}
+
+\section{Bibliography}\
 

### 2021-07-19 new draft.Rmd
l'unico file importante è nuova final draft

--- 2021-07-16 final draft.Rmd
@@
 header-includes:
+- \def\code#1{\texttt{#1}}
+- \usepackage{amssymb}
 - \usepackage{txfonts}
@@
 - \usepackage{lipsum}
-- \usepackage{amssymb}
 - \usepackage{xcolor}
@@
 - \newtheorem{proof*}{Proof.}
-- \newtheorem{definition}{Definition}
 - \renewcommand{\P}{\mathbb{P}}
@@
 - \renewcommand{\baselinestretch}{1.5}
-- \usepackage{amssymb}
 - \newcommand{\E}{\mathbb E}
@@
 - \newcommand{\N}{\mathcal N}
-- \def\code#1{\texttt{#1}}
 - \newtheorem{algorithm}{Algorithm}[section]
@@
 $$
+Convergence in distribution is denoted with the symbol $\implies.$ So, $X\implies\N(0,1)$ means that the random variable $X$ converges in distribution to a normal Gaussian.
 
@@
 \begin{equation}\label{state-space-densities}
-   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
+   \P_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\P_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
 \end{equation}
 Integrating away $X_{0:T},$ one gets the marginal density
-\begin{equation}\label{state-space-marginal-y}
-\begin{align*}
+\begin{align*}\label{state-space-marginal-y}
   \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
@@
 \end{align*}                
-\end{equation}
 where $p_t(y_{0:t})$ is the likelihood function.
@@
 
-\textbf{Implementation}
-
-In this section we present a practical illustration of the algorithms discussed.
-For the sake of simplicity we use a random walk plus noise model, i.e.
-the most basic form of a linear Gaussian state-space model.
-\begin{align}
-y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
-x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
-x_{0} & \sim N(m_{0},C_{0})
-\end{align}
-As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
-However, this toy example will be used also to illustrate more involved filtering strategies described in this work.
-We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
-\
-The filtering strategy is applied to 50 simulated data.
-Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
-```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
-
-set.seed(1991)
-dlm.sim = function(n,sigma,tau,x0){
-  x    = rep(0,n)  #state process
-  y    = rep(0,n)  #obs process
-  x[1] = rnorm(1,x0,tau)
-  y[1] = rnorm(1,x[1],sigma)
-  for (t in 2:n){
-    x[t] = rnorm(1,x[t-1],tau)
-    y[t] = rnorm(1,x[t],sigma)
-  }
-  return(list(x=x,y=y))
-}
-
-n=50
-tau2=1
-sigma2=1
-sigma=sqrt(sigma2)
-tau=sqrt(tau2)
-x0=0
-
-sim<-dlm.sim(n,sigma,tau,x0)
-y<-sim$y
-x<-sim$x
-truex<-x
-
-timeframe<-c(1:length(y))
-df<-data.frame(timeframe,y,x)
-
-ggplot(df,aes(x=timeframe))+
-  geom_line(aes(y=y, linetype="Observations"))+
-  geom_line(aes(y=x, linetype="True States"))+
-  coord_cartesian(ylim = c(-4, +12)) +
-  scale_linetype_manual(name="",
-                      values=c("Observations" = 1,
-                               "True States" = 3))+
-  labs(x="Time",
-       y="")+
-  #ggtitle("States and Observations")+
-  theme_bw()+
-  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-        legend.background = element_rect(fill = "white", colour = "gray30")) +
-  theme(plot.title = element_text(hjust = 0.5))
-```
-The discussion of Kalman Filter will continue in Section XX where we compare it to the Particle Filter.
-
 \chapter{Particle Filter}
@@
 \hrule
+\
 \code{APFfun(data,N,m0,C0,tau,sigma,r)}
 \hrule
+\
 \textbf{Arguments}

### 2021-07-19 new-draft.tex
l'unico file importante è nuova final draft

--- 2021-07-15 final-draft.tex
@@
 \urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\usepackage{graphicx}
 \makeatletter
@@
 \setcounter{secnumdepth}{5}
+\def\code#1{\texttt{#1}}
+\usepackage{amssymb}
 \usepackage{txfonts}
@@
 \usepackage{lipsum}
-\usepackage{amssymb}
 \usepackage{xcolor}
@@
 \newtheorem{proof*}{Proof.}
-\newtheorem{definition}{Definition}
 \renewcommand{\P}{\mathbb{P}}
@@
 \renewcommand{\baselinestretch}{1.5}
-\usepackage{amssymb}
 \newcommand{\E}{\mathbb E}
@@
 \newcommand{\X}{\mathcal X}
+\newcommand{\Y}{\mathcal Y}
 \newcommand{\Pd}{\mathbb P}
 \newcommand{\N}{\mathcal N}
-\def\code#1{\texttt{#1}}
 \newtheorem{algorithm}{Algorithm}[section]
+\usepackage{float}
+\floatplacement{figure}{}
 \usepackage{float}
@@
 \usepackage{xcolor}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 
@@
     \int_{\mathcal{X}}\varphi(x)\mathbb{Q}(dx).
-\]
+\] Convergence in distribution is denoted with the symbol \(\implies.\)
+So, \(X\implies\mathcal N(0,1)\) means that the random variable \(X\)
+converges in distribution to a normal Gaussian.
 
@@
 
-Let \(P_1, P_2, \dots, P_T\) be a finite measure of probability kernels
+Let \(P_1, P_2, \dots, P_T\) be a finite sequence of probability kernels
 from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
@@
 \begin{equation}
-    \mathbb{P}_T(X_{0:T}\in dx_{0:T})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},x_s).
+    \mathbb{P}_T(X_{0:T}\in dx_{0:T})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},dx_s).
 \end{equation}
@@
 equality implies that it is possible to identify conditional
-distributions with transition kernels. Assume \(\mathbb{P}_t, t\leq T,\)
-is a sequence of probability measures. Then, the following proposition
-holds.
+distributions with transition kernels. Assume \(\mathbb{P}_t,\) with
+\(t\leq T,\) is a sequence of probability measures. Then, the following
+proposition holds.
 
@@
 
-We now show that State-Space models can be understood as Markov
-processes that are partially observed. Let \(X_t\in\mathcal X\) and
-\(Y_t\in\Y\) and let \(\{(X_t,T_t\}\) be a stochastic process. Consider
-the initial distribution \(\mathbb{P}_0(dx_0)\) and the probability
-kernels \(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where
-\(t=1,2,\dots\).
-
-\begin{definition}
-The process $\{(X_t,T_t\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
+We now define \textit{State-Space models} as Markov processes that are
+partially observed. Let \(X_t\in\mathcal X\) and \(Y_t\in\mathcal Y\)
+and let \(\{(X_t,Y_t)\}\) be a stochastic process. To help intuition,
+suppose that \(\{Y_t\}\) is observable, i.e., coincides with a sequence
+of observations, while \(\{X_t\}\) is latent. Consider the initial
+distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
+\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).
+
+\begin{definition}\label{state-space-def}
+The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
 \begin{align*}
@@
   &= [\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
-  &= \mathbb{P}_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t)].
+  &= \mathbb{P}_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
 \end{align*}
@@
 
-From the second inequality, it is apparent that \(\{(X_t,T_t\}\) is a
+From the second inequality, it is apparent that \(\{(X_t,Y_t)\}\) is a
 Markov process with initial distribution
@@
 that, \(\{X_t\}\) has a marginal distribution
-\(\mathbb{P}_T(dx_{0:T})=\mathbb{P}_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),\)
+\[\mathbb{P}_T(dx_{0:T})=\mathbb{P}_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),\]
 meaning that \(\{X_t\}\) is marginally distributed as a Markov process
@@
 
+Assume now that there exists a dominating measures \(\nu(dy)\) such that
+\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
+last equality of definition \ref{state-space-def}, the joint
+distribution of \((X_{0:T},Y_{0:T})\) is
+\begin{equation}\label{state-space-densities}
+   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\nu(dy_t).
+\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
+density \begin{align*}\label{state-space-marginal-y}
+  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
+                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
+\end{align*}\\
+where \(p_t(y_{0:t})\) is the likelihood function. Dividing the
+left-hand side of equation \ref{state-space-densities} by the left-hand
+side of equation \ref{state-space-marginal-y} and the right-hand side of
+equation \ref{state-space-densities} by the right-hand side of equation
+\ref{state-space-marginal-y}, one obtains the filtering distribution
+\begin{equation}
+\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
+\end{equation}
+
 \chapter{Linear-Gaussian case}
+
+\textit{Linear Gaussian models} are a class of popular state-space
+models. A notable property of such models is that the filtering and
+predictive distributions are analytically tractable, which makes such
+models very convenient in applications.
+
+Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
+\(\mathcal Y=\mathbb R^{d_y}.\) Let
+\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
+\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
+and assume the initial distribution
+\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
+if \begin{align}
+X_t &= A_tX_{t-1}+U_t \\
+Y_t &= B_tX_t + V_t.
+\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
+normally distributed.
+
+\begin{lemma}\label{kalman-filter}
+  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Then, it holds that
+  \begin{align}
+    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
+    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
+    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+  \end{align}
+\end{lemma}
+\begin{proof*}
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
+\begin{equation} \label{normal-distrib-rewrite}
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
+\end{equation}
+with $A=Q_0^{-1}$ and $B=Am_0.$
+Now, by Bayes' Theorem,
+
+    \begin{align*}
+        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
+                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
+                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
+                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
+    \end{align*}
+
+Hence, it follows that 
+\begin{equation*}
+    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
+\end{equation*}
+and 
+    \begin{align*}
+        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
+            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+    \end{align*}
+\end{proof*}
+
+Assume that
+\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
+Then, using the transition equation of the latent variable, by the
+linearity of the Gaussian distribution, we have that the predictive
+distribution is \begin{equation}
+\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
+\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
+lemma \ref{kalman-filter}, it follows that the filtering distribution is
+\begin{align*}
+  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
+\end{align*}
+
+\section{Implementation}
+
+In this section we present a practical illustration of the algorithms
+discussed. For the sake of simplicity we use a random walk plus noise
+model, i.e.~the most basic form of a linear Gaussian state-space
+model.\\
+\begin{align}
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+x_{0} & \sim N(m_{0},C_{0})
+\end{align} As already mentioned before, in this case the filtering
+distribution can be computed in closed form solutions using the Kalman
+filter. However, this toy example will be used also to illustrate more
+involved filtering strategies described in this work. We believe indeed
+that it represents a useful starting point to understand the logic of
+the algorithms which may be eventually replicated when dealing with more
+complex models.\\
+The filtering strategy is applied to 50 simulated data. Figure XX shows
+the simulated true states sequence assuming as data generating process
+the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
+process form Equation (1) with \(\sigma^{2}=1\).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-5-1}
+
+}
+
+\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
+\end{figure}
+
+The Kalman Filter for this model is implemented as described below.
+
+\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
+\begin{align*}
+x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
+a_{t} & = m_{t-1}\\
+R_{t} & = C_{t-1}+\tau^2
+\end{align*}
+\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e.
+the product of the one-step-ahead state predictive distribution and the likelihood
+\begin{align*}
+x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
+m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
+C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
+\end{align*}
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Our \texttt{DLM} function implement in R replicate this steps.\\
+
+\hrule
+\hrule
+\texttt{DLM(data,sig2,tau2,m0,C0)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{sig2} ~~the variance \(\sigma^{2}\) in Equation (1)\\
+\texttt{tau2} ~~the variance \(\tau^{2}\) in Equation (2)\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DLM}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
+\NormalTok{  n  }\OtherTok{=} \FunctionTok{length}\NormalTok{(data)}
+\NormalTok{  m  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+\NormalTok{  C  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{n)\{}
+    \ControlFlowTok{if}\NormalTok{ (t}\SpecialCharTok{==}\DecValTok{1}\NormalTok{)\{}
+\NormalTok{      a }\OtherTok{=}\NormalTok{ m0}
+\NormalTok{      R }\OtherTok{=}\NormalTok{ C0 }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      a }\OtherTok{=}\NormalTok{ m[t}\DecValTok{{-}1}\NormalTok{]}
+\NormalTok{      R }\OtherTok{=}\NormalTok{ C[t}\DecValTok{{-}1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{    \}}
+\NormalTok{    A }\OtherTok{=}\NormalTok{ R}\SpecialCharTok{/}\NormalTok{(R}\SpecialCharTok{+}\NormalTok{sig2)}
+\NormalTok{    m[t] }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{A)}\SpecialCharTok{*}\NormalTok{a }\SpecialCharTok{+}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{y[t]}
+\NormalTok{    C[t] }\OtherTok{=}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{sig2}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{m=}\NormalTok{m,}\AttributeTok{C=}\NormalTok{C))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+In Figure XX below filtered states estimated using Kalman Filter with
+\(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are compared to
+the true states values. Notice how closely the filtered states follow
+the observations and the goodness of the approximation of the true
+states. 95 percent credible intervals are computed as
+\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-7-1}
+
+}
+
+\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
+\end{figure}
+
+The discussion of Kalman Filter will continue in Section XX where we
+compare it to the Particle Filter.
 
@@
 Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
-\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\footnote{From the Radom-Nikodym theorem a sufficient condition for $w$ to exist is that $\mathbb Q$ is absolutely continuous with respect to $\mathbb M$: $\mathbb Q\ll \mathbb M$.}\\
+\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\\
 The following result represents the starting point for the idea of
@@
 \end{equation}
+
+\section{Implementation}
+
+Consider again the random walk plus noise model of section XX,the
+challenge here is to estimate the filtered states using a Sequential
+Importance Sampling algorithm. The main idea of the SIS applied to this
+univariate linear gaussian model is described in the following
+algorithm. We indicate with \(n\) the sample size of time observations
+and with \(N\) the generated sample size for each step of the Sequenial
+Monte Carlo.
+
+\begin{algorithm} SIS filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Our \texttt{SISfun} function implemented in R replicate this steps.\\
+
+\hrule
+\hrule
+\texttt{SISfun(data,N,m0,C0,tau,sigma)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SISfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }\CommentTok{\#sample from N(x\_\{t{-}1\},tau)}
+\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }\CommentTok{\#update weight}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+
+\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }\CommentTok{\#normalized weight}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }\CommentTok{\#effective sample size}
+
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+We have already discussed the reasons why the SIS algorithm does not
+provide a good strategy in the filtering problem. We provide a graphical
+intuition of what happens when we use such filtering strategy on a
+simulated dataset. We decide to set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\)
+and \(\tau=\sigma=1\). The results shown in the following two plots
+shows a clear degeneration of the effective sample size and bad fit of
+filtered states with respect to the true values.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-10-1}
+
+}
+
+\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-10}
+\end{figure}
 
@@
 \end{itemize}
+
+\section{Implementation}
+
+In this section, Bootstrap Particle Filter will be used to estimate
+filtered states of the Random Walk plus Noise introduced in section XX.
+The overall strategy replicate the SIS filter with the addition of a
+ESS-based resampling step that applies when the effective sample size is
+smaller than a predetermined threshold opportunely chosen (in our
+example we decide to follow a common rule of thumb consisting in setting
+the threshold at \(N/2\)). The steps are presented in the following
+algorithm.
+
+\begin{algorithm} BPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+These steps are resumed in our \texttt{PFfun} function.\\
+
+\hrule
+\hrule
+\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{PFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,x,tau)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+The estimated states of the Boostrap Particle Filter togheter with the
+effective sample size are shown in Figure XX. We decide to set
+\(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice how
+the resampling step allows the effective sample size not to drop,
+improving results.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-13-1}
+
+}
+
+\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-13}
+\end{figure}
+
+Since the Random Walk plus Noise model allows for closed form solutions,
+we want to compare the results of Boostrap Particle Filter(BPF) with
+Kalman Filter(KF). As we can see from figure XX, when the number of
+particles generated at any interaction increases the results for both
+the estimated mean and the variance tend to converge {[}{[}SPIEGARE
+PERCHE'{]}{]}. Moreover, comparing the Root Mean Square Errors of the
+filtered states with respect to the true state values, when the sample
+size increases the BPF decreases and, for large N, it reaches the
+accuracy of the Kalman Filter.
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-15}Root Mean Square Errors}\\
+\toprule
+N & Threshold & KF & BPF\\
+\midrule
+100 & 0.5 & 0.879 & 0.888\\
+1000 & 0.5 & 0.879 & 0.886\\
+10000 & 0.5 & 0.879 & 0.878\\
+\bottomrule
+\end{longtable}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-16-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-16}
+\end{figure}
+
+\hfill\break
 
@@
 
+\section{Implementation}
+
+\hfill\break
+The Bootstrap Particle Filter Approach for the Random Walk plus Noise
+Model described in Section XX can be improved accounting for the
+observations in the importance transition density and it consists of
+generating \(x_{t}\) from its conditional distribution given \(x_{t-1}\)
+and \(y_{t}\). In the Normal model we are considering, the optimal
+proposal will be a Normal density as well with mean and variance given
+by \begin{align*}
+\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
+\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
+\end{align*} On the other hand, the incremental weights, using this
+importance transition density, are proportional to the conditional
+density of \(y_{t}\) given \(x_{t-1}=x_{t-1}^{(i)}\) , i.e
+\(N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})\), evaluated at \(y_{t}\). In
+other words the algorithm implemented in R is:
+
+\begin{algorithm} GPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $\mu_{opt}$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+The \texttt{GPFfun} function resume this passages.\\
+
+\hrule
+\hrule
+\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{GPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
+\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x)}
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-19-1}
+
+}
+
+\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-19}
+\end{figure}
+
+Let's provide directly a brief comparison between the Bootstrap Particle
+Filter (BPF) and the Guided Particle Filter (GPF). As we can see from
+the Figure XX, the Guided Particle Filter provides point estimates that
+are slightly better with respect to the ones of the BPF, and this is
+confirmed by the Table showing the RMSE. Moreover, also the variance is
+better, suggesting for higher precision.
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-21}Root Mean Square Errors}\\
+\toprule
+N & Threshold & BPF & GPF\\
+\midrule
+1000 & 0.50 & 0.916 & 0.880\\
+1000 & 0.25 & 0.895 & 0.862\\
+1000 & 0.10 & 0.869 & 0.881\\
+\bottomrule
+\end{longtable}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-22-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-22}
+\end{figure}
+
 \subsection{Auxiliary Particle Filter}
@@
 \end{itemize}
+
+For illustration purposes, we are going implement an auxiliary particle
+filter with auxiliary function
+\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).\\
+
+\begin{algorithm} APF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item For $k=1,...,N$:
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+The \texttt{APFfun} function resume this passages.\\
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}
+
+\hrule
+
+\hfill\break
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{APFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x[k],tau)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x1}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-25-1}
+
+}
+
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-25}
+\end{figure}
+
+Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap
+Particle Filter (BPF).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-26-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-26}
+\end{figure}
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-28}Root Mean Square Errors}\\
+\toprule
+N & Threshold & BPF & APF\\
+\midrule
+1000 & 0.50 & 0.885 & 0.875\\
+1000 & 0.25 & 0.893 & 0.892\\
+1000 & 0.10 & 0.855 & 0.870\\
+\bottomrule
+\end{longtable}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
+\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{    weight }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    \}}
+
+\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-31-1}
+
+}
+
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-31}
+\end{figure}
 
@@
 
+\section{Implementation}
+
+\hfill\break
+Consider the linear Gaussian example of section XX, but this time with
+unknown variances \(\tau^{2}\) and \(\sigma^{2}\). Thus, let
+\(\psi=(\sigma^{2},\tau^{2})\) be the unknown parameter vector and
+assign a gamma prior for its components, \begin{align*}
+\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
+\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
+\end{align*} Alternatively, assign them a uniform prior if we have no
+knowledge of the hyperparameters. The algorithm follows the following
+steps.
+
+\begin{algorithm} LWF for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
+$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:.
+\begin{enumerate}
+\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\begin{align*}
+m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
+\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\end{align*}
+\item For $k=1,...,N$:
+\begin{itemize}
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
+\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
+\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
+\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
+\end{itemize}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Our \texttt{LWfun} function goes through the illustrated steps.\\
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alphav, betav} ~~ Gamma prior hyperparameters on
+\(\sigma^{2}\)\\
+\texttt{alphaw,betaw} ~~ Gamma prior hyperparameters on \(\tau^{2}\)\\
+\texttt{delta} ~~hyperparameter delta value\\
+\texttt{unif} ~~if True then it sets a Uniform \((0,10)\) prior on
+\(\sigma^{2}\) and \(\tau^{2}\)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{LWfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs     }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+  \ControlFlowTok{if}\NormalTok{(unif}\SpecialCharTok{==}\NormalTok{T)\{}
+\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{),}\FunctionTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{))\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphav,}\AttributeTok{scale=}\NormalTok{betav),}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaw,}\AttributeTok{scale=}\NormalTok{betaw))}
+\NormalTok{  a      }\OtherTok{=}\NormalTok{ (}\DecValTok{3}\SpecialCharTok{*}\NormalTok{delta}\DecValTok{{-}1}\NormalTok{)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{2}\SpecialCharTok{*}\NormalTok{delta)}
+\NormalTok{  h2     }\OtherTok{=} \DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}
+\NormalTok{  parss  }\OtherTok{=} \FunctionTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\FunctionTok{c}\NormalTok{(N,}\DecValTok{2}\NormalTok{,n))}
+\NormalTok{  xss    }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  ws     }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  ess    }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  w      }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    meanV }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{],w)}
+\NormalTok{    varV  }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{{-}}\NormalTok{meanV)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,w)}
+\NormalTok{    meanW }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{],w)}
+\NormalTok{    varW  }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{{-}}\NormalTok{meanW)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,w)}
+
+\NormalTok{    muV }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\NormalTok{meanV}
+\NormalTok{    sigma2V }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{varV}
+\NormalTok{    alphaV }\OtherTok{=}\NormalTok{ muV}\SpecialCharTok{\^{}}\DecValTok{2}\SpecialCharTok{/}\NormalTok{sigma2V}
+\NormalTok{    betaV }\OtherTok{=}\NormalTok{ muV}\SpecialCharTok{/}\NormalTok{sigma2V}
+
+\NormalTok{    muW }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\NormalTok{meanW}
+\NormalTok{    sigma2W }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{varW}
+\NormalTok{    alphaW }\OtherTok{=}\NormalTok{ muW}\SpecialCharTok{\^{}}\DecValTok{2}\SpecialCharTok{/}\NormalTok{sigma2W}
+\NormalTok{    betaW }\OtherTok{=}\NormalTok{ muW}\SpecialCharTok{/}\NormalTok{sigma2W}
+
+\NormalTok{    weight      }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],xs,}\FunctionTok{sqrt}\NormalTok{(muV))}
+\NormalTok{    k           }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{weight)}
+
+\NormalTok{    pars[,}\DecValTok{1}\NormalTok{]}\OtherTok{\textless{}{-}}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaV[k],}\AttributeTok{rate=}\NormalTok{betaV[k])}
+\NormalTok{    pars[,}\DecValTok{2}\NormalTok{]}\OtherTok{\textless{}{-}}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaW[k],}\AttributeTok{rate=}\NormalTok{betaW[k])}
+
+\NormalTok{    xsprevious}\OtherTok{\textless{}{-}}\NormalTok{xs[k]}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,xs[k],}\FunctionTok{sqrt}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{]))}
+
+\NormalTok{    w           }\OtherTok{=} \FunctionTok{exp}\NormalTok{(}\FunctionTok{dnorm}\NormalTok{( data[t],xs,}\FunctionTok{sqrt}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{]),}\AttributeTok{log=}\NormalTok{T)}\SpecialCharTok{{-}}
+                        \FunctionTok{dnorm}\NormalTok{( data[t],xsprevious,}\FunctionTok{sqrt}\NormalTok{(muV[k]),}\AttributeTok{log=}\NormalTok{T))}
+\NormalTok{    w           }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS         }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xs[index]}
+\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{pars[index,]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xs}
+\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{pars}
+\NormalTok{    \}}
+
+\NormalTok{    xss         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xss,xs)}
+\NormalTok{    parss[,,t]  }\OtherTok{=}\NormalTok{ pars }
+\NormalTok{    ws          }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xss=}\NormalTok{xss,}\AttributeTok{parss=}\NormalTok{parss,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \section{Convergence}\label{pf_converg}
@@
 exercise, ultimately underlining the importance of taking stochastic
-volatility into account in modeling equity returns.
+volatility into account in modeling equity returns. We follow XX,
+comparing density forecasts through relative predictive and cumulative
+likelihoods.
 
@@
 
-{\centering \includegraphics{final-draft_files/figure-latex/cll_plot-1}
+{\centering \includegraphics{new-draft_files/figure-latex/cll_plot-1}
 

### 2021-07-19 nuova final draft.Rmd
l'unico file importante è nuova final draft

--- 2021-07-19 new draft.Rmd
@@
 ---
+title: "Final draft"
+
+documentclass: book
+
+output:
+  pdf_document: 
+   latex_engine: pdflatex
+   extra_dependencies: ["float"]
+   number_sections: true
+
 header-includes:
@@
 
-title: "Final draft"
-
-documentclass: book
-
-output:
-  pdf_document: 
-   latex_engine: xelatex
-   fig_width: 5
-   fig_height: 3.5
-   fig_caption: true
-   extra_dependencies: ["float"]
-   number_sections: true
-
 ---
@@
 \end{equation*}
-However, in many circumstances either it is impossible to sample from the target density or there exist another density - call it $m(\cdot)$ - different from $q$ that produces ``more efficient" estimates.
-In all the cases for which sampling takes place from a distribution different from the ``correct" (i.e., the true) one, importance sampling and resampling methods can be applied (as augmented versions of the standard MC algorithm).
+However, in many circumstances either it is impossible to sample from the target density or there exist another density - call it $m(\cdot)$ - different from $q$ that produces "more efficient" estimates.
+In all the cases for which sampling takes place from a distribution different from the "correct" (i.e., the true) one, importance sampling and resampling methods can be applied (as augmented versions of the standard MC algorithm).
 
@@
 Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\
-Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot) $ be the prior distribution for the parameter $\psi$.\
+Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
 The LWF can be described by the following algorithm.
@@
 
-```{r cll_plot, echo=F, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
+```{r cll_plot, eval=FALSE, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, include=FALSE, results=F}
 load("SV/cll_plot.Rda")

### 2021-07-19 Real Dataset SV.Rmd
commit

@@
 
-\section{Visual Analysis after Filtering}
+\section{Comparing the filters}
 
-In this section, we offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the volatility of the observations obtained through the application of the different filtering techniques explored in the theoretical analysis.
+In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset.
+First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions.
+Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases.
+In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.
+Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model.
+Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
+\begin{align*}
+y_t|x_t & \sim N(0,e^{x_t})\\
+x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
+x_0 & \sim N(0,100)\\
+\alpha & \sim N(\gamma,\zeta)\\
+\beta & \sim N(\pi,\phi)\\
+\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
+\end{align*}
+where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\
+
+\subsection{Visual Analysis after Filtering}
+
+We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis.
 Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows.
 For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
-Note that the number of particles was set at 10000 for each filter.} \
+Note that the number of particles was set at 10000 for each filter: the next section will come back to such a choice.} \
 
@@
 \textit{Potremmo parlare di maggiore sensibilità alle oscillazioni dei prezzi, o simili}
+
+\subsection{Quality of approximation}
+
+Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
+Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza.
+Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target.
+With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
+We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark.
+Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.
+
+We start by setting N=50000 and using the Bootstrap particle filter as the benchmark.
+In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+By "Basic PF we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when ESS is non negative (\textit{i.e.} always) or when ESS<0 (\textit{i.e.} never).
+In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
 

### 2021-07-19 final-draft.tex
commit

@@
 ]{book}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
@@
 \newcommand{\X}{\mathcal X}
+\newcommand{\Y}{\mathcal Y}
 \newcommand{\Pd}{\mathbb P}
@@
 \newtheorem{algorithm}{Algorithm}[section]
+\usepackage{float}
+\floatplacement{figure}{}
 \usepackage{float}
@@
 
+Assume now that there exists a dominating measures \(\nu(dy)\) such that
+\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
+last equality of definition \ref{state-space-def}, the joint
+distribution of \((X_{0:T},Y_{0:T})\) is
+\begin{equation}\label{state-space-densities}
+   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
+\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
+density \begin{equation}\label{state-space-marginal-y}
+\begin{align*}
+  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
+                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
+\end{align*}                
+\end{equation} where \(p_t(y_{0:t})\) is the likelihood function.
+Dividing the left-hand side of equation \ref{state-space-densities} by
+the left-hand side of equation \ref{state-space-marginal-y} and the
+right-hand side of equation \ref{state-space-densities} by the
+right-hand side of equation \ref{state-space-marginal-y}, one obtains
+the filtering distribution \begin{equation}
+\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
+\end{equation}
+
 \chapter{Linear-Gaussian case}
+
+\textit{Linear Gaussian models} are a class of popular state-space
+models. A notable property of such models is that the filtering and
+predictive distributions are analytically tractable, which makes such
+models very convenient in applications.
+
+Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
+\(\mathcal Y=\mathbb R^{d_y}.\) Let
+\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
+\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
+and assume the initial distribution
+\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
+if \begin{align}
+X_t &= A_tX_{t-1}+U_t \\
+Y_t &= B_tX_t + V_t.
+\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
+normally distributed.
+
+\begin{lemma}\label{kalman-filter}
+  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Then, it holds that
+  \begin{align}
+    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
+    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
+    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+  \end{align}
+\end{lemma}
+\begin{proof*}
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
+\begin{equation} \label{normal-distrib-rewrite}
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
+\end{equation}
+with $A=Q_0^{-1}$ and $B=Am_0.$
+Now, by Bayes' Theorem,
+
+    \begin{align*}
+        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
+                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
+                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
+                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
+    \end{align*}
+
+Hence, it follows that 
+\begin{equation*}
+    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
+\end{equation*}
+and 
+    \begin{align*}
+        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
+            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+    \end{align*}
+\end{proof*}
+
+Assume that
+\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
+Then, using the transition equation of the latent variable, by the
+linearity of the Gaussian distribution, we have that the predictive
+distribution is \begin{equation}
+\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
+\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
+lemma \ref{kalman-filter}, it follows that the filtering distribution is
+\begin{align*}
+  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
+\end{align*}
+
+\section{Implementation}
+
+In this section we present a practical illustration of the algorithms
+discussed. For the sake of simplicity we use a random walk plus noise
+model, i.e.~the most basic form of a linear Gaussian state-space
+model.\\
+\begin{align}
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+x_{0} & \sim N(m_{0},C_{0})
+\end{align} As already mentioned before, in this case the filtering
+distribution can be computed in closed form solutions using the Kalman
+filter. However, this toy example will be used also to illustrate more
+involved filtering strategies described in this work. We believe indeed
+that it represents a useful starting point to understand the logic of
+the algorithms which may be eventually replicated when dealing with more
+complex models.\\
+The filtering strategy is applied to 50 simulated data. Figure XX shows
+the simulated true states sequence assuming as data generating process
+the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
+process form Equation (1) with \(\sigma^{2}=1\).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-5-1}
+
+}
+
+\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
+\end{figure}
+
+The Kalman Filter for this model is implemented as described below.
+
+\begin{algorithm} Kalman Filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute the one-step-ahead state predictive distribution at time $t-1$, notice that if $t=1$ no data have been observed yet and therefore $x_{1}|x_{0}\sim N(a_{1},R_{1})$ otherwise
+\begin{align*}
+x_{t}|y_{1:t-1} & \sim N(a_{t},R_{t})\\
+a_{t} & = m_{t-1}\\
+R_{t} & = C_{t-1}+\tau^2
+\end{align*}
+\item Compute the filtering distribution at time $t$ as $p(x_{t}|y_{1:t}) \propto p(x_{t}|y_{1:t-1})p(y_{t}|x_{t})$, i.e.
+the product of the one-step-ahead state predictive distribution and the likelihood
+\begin{align*}
+x_{t}|y_{1:t} & \sim N(m_{t},C_{t}) \\
+m_{t} & = \big(1-\frac{R_{t}}{R_{t}+\sigma^2}\big)a_{t}+\frac{R_{t}}{R_{t}+\sigma^2}y_{t} \\
+C_{t} & = \frac{R_{t}}{R_{t}+\sigma^2}\sigma^2
+\end{align*}
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Our \texttt{DLM} function implement in R replicate this steps.\\
+
+\hrule
+\hrule
+\texttt{DLM(data,sig2,tau2,m0,C0)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{sig2} ~~the variance \(\sigma^{2}\) in Equation (1)\\
+\texttt{tau2} ~~the variance \(\tau^{2}\) in Equation (2)\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DLM}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
+\NormalTok{  n  }\OtherTok{=} \FunctionTok{length}\NormalTok{(data)}
+\NormalTok{  m  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+\NormalTok{  C  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{n)\{}
+    \ControlFlowTok{if}\NormalTok{ (t}\SpecialCharTok{==}\DecValTok{1}\NormalTok{)\{}
+\NormalTok{      a }\OtherTok{=}\NormalTok{ m0}
+\NormalTok{      R }\OtherTok{=}\NormalTok{ C0 }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      a }\OtherTok{=}\NormalTok{ m[t}\DecValTok{{-}1}\NormalTok{]}
+\NormalTok{      R }\OtherTok{=}\NormalTok{ C[t}\DecValTok{{-}1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{    \}}
+\NormalTok{    A }\OtherTok{=}\NormalTok{ R}\SpecialCharTok{/}\NormalTok{(R}\SpecialCharTok{+}\NormalTok{sig2)}
+\NormalTok{    m[t] }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{A)}\SpecialCharTok{*}\NormalTok{a }\SpecialCharTok{+}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{y[t]}
+\NormalTok{    C[t] }\OtherTok{=}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{sig2}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{m=}\NormalTok{m,}\AttributeTok{C=}\NormalTok{C))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+In Figure XX below filtered states estimated using Kalman Filter with
+\(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are compared to
+the true states values. Notice how closely the filtered states follow
+the observations and the goodness of the approximation of the true
+states. 95 percent credible intervals are computed as
+\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-7-1}
+
+}
+
+\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
+\end{figure}
+
+The discussion of Kalman Filter will continue in Section XX where we
+compare it to the Particle Filter.
+
+\textbf{Implementation}
+
+In this section we present a practical illustration of the algorithms
+discussed. For the sake of simplicity we use a random walk plus noise
+model, i.e.~the most basic form of a linear Gaussian state-space model.
+\begin{align}
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+x_{0} & \sim N(m_{0},C_{0})
+\end{align} As already mentioned before, in this case the filtering
+distribution can be computed in closed form solutions using the Kalman
+filter. However, this toy example will be used also to illustrate more
+involved filtering strategies described in this work. We believe indeed
+that it represents a useful starting point to understand the logic of
+the algorithms which may be eventually replicated when dealing with more
+complex models.\\
+The filtering strategy is applied to 50 simulated data. Figure XX shows
+the simulated true states sequence assuming as data generating process
+the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
+process form Equation (1) with \(\sigma^{2}=1\).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-8-1}
+
+}
+
+\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-8}
+\end{figure}
+
+The discussion of Kalman Filter will continue in Section XX where we
+compare it to the Particle Filter.
 
@@
 \end{equation}
+
+\section{Implementation}
+
+Consider again the random walk plus noise model of section XX,the
+challenge here is to estimate the filtered states using a Sequential
+Importance Sampling algorithm. The main idea of the SIS applied to this
+univariate linear gaussian model is described in the following
+algorithm. We indicate with \(n\) the sample size of time observations
+and with \(N\) the generated sample size for each step of the Sequenial
+Monte Carlo.
+
+\begin{algorithm} SIS filter for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Our \texttt{SISfun} function implemented in R replicate this steps.\\
+
+\hrule
+\hrule
+\texttt{SISfun(data,N,m0,C0,tau,sigma)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SISfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }\CommentTok{\#sample from N(x\_\{t{-}1\},tau)}
+\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }\CommentTok{\#update weight}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+
+\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }\CommentTok{\#normalized weight}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }\CommentTok{\#effective sample size}
+
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+We have already discussed the reasons why the SIS algorithm does not
+provide a good strategy in the filtering problem. We provide a graphical
+intuition of what happens when we use such filtering strategy on a
+simulated dataset. We decide to set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\)
+and \(\tau=\sigma=1\). The results shown in the following two plots
+shows a clear degeneration of the effective sample size and bad fit of
+filtered states with respect to the true values.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-11-1}
+
+}
+
+\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-11}
+\end{figure}
 
@@
 \end{itemize}
+
+\section{Implementation}
+
+In this section, Bootstrap Particle Filter will be used to estimate
+filtered states of the Random Walk plus Noise introduced in section XX.
+The overall strategy replicate the SIS filter with the addition of a
+ESS-based resampling step that applies when the effective sample size is
+smaller than a predetermined threshold opportunely chosen (in our
+example we decide to follow a common rule of thumb consisting in setting
+the threshold at \(N/2\)). The steps are presented in the following
+algorithm.
+
+\begin{algorithm} BPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+These steps are resumed in our \texttt{PFfun} function.\\
+
+\hrule
+\hrule
+\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{PFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,x,tau)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+The estimated states of the Boostrap Particle Filter togheter with the
+effective sample size are shown in Figure XX. We decide to set
+\(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice how
+the resampling step allows the effective sample size not to drop,
+improving results.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-14-1}
+
+}
+
+\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-14}
+\end{figure}
+
+Since the Random Walk plus Noise model allows for closed form solutions,
+we want to compare the results of Boostrap Particle Filter(BPF) with
+Kalman Filter(KF). As we can see from figure XX, when the number of
+particles generated at any interaction increases the results for both
+the estimated mean and the variance tend to converge {[}{[}SPIEGARE
+PERCHE'{]}{]}. Moreover, comparing the Root Mean Square Errors of the
+filtered states with respect to the true state values, when the sample
+size increases the BPF decreases and, for large N, it reaches the
+accuracy of the Kalman Filter.
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-16}Root Mean Square Errors}\\
+\toprule
+N & Threshold & KF & BPF\\
+\midrule
+100 & 0.5 & 0.879 & 0.888\\
+1000 & 0.5 & 0.879 & 0.886\\
+10000 & 0.5 & 0.879 & 0.878\\
+\bottomrule
+\end{longtable}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-17-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-17}
+\end{figure}
+
+\hfill\break
 
@@
 
+\section{Implementation}
+
+\hfill\break
+The Bootstrap Particle Filter Approach for the Random Walk plus Noise
+Model described in Section XX can be improved accounting for the
+observations in the importance transition density and it consists of
+generating \(x_{t}\) from its conditional distribution given \(x_{t-1}\)
+and \(y_{t}\). In the Normal model we are considering, the optimal
+proposal will be a Normal density as well with mean and variance given
+by \begin{align*}
+\mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
+\sigma_{opt}^{2}=V(x_{t}|x_{t-1},y_{t})&=\frac{\tau^{2}\sigma^{2}}{\tau^{2}+\sigma^{2}}
+\end{align*} On the other hand, the incremental weights, using this
+importance transition density, are proportional to the conditional
+density of \(y_{t}\) given \(x_{t-1}=x_{t-1}^{(i)}\) , i.e
+\(N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})\), evaluated at \(y_{t}\). In
+other words the algorithm implemented in R is:
+
+\begin{algorithm} GPF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $\mu_{opt}$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+The \texttt{GPFfun} function resume this passages.\\
+
+\hrule
+\hrule
+\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{GPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
+\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x)}
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-20-1}
+
+}
+
+\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-20}
+\end{figure}
+
+Let's provide directly a brief comparison between the Bootstrap Particle
+Filter (BPF) and the Guided Particle Filter (GPF). As we can see from
+the Figure XX, the Guided Particle Filter provides point estimates that
+are slightly better with respect to the ones of the BPF, and this is
+confirmed by the Table showing the RMSE. Moreover, also the variance is
+better, suggesting for higher precision.
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-22}Root Mean Square Errors}\\
+\toprule
+N & Threshold & BPF & GPF\\
+\midrule
+1000 & 0.50 & 0.916 & 0.880\\
+1000 & 0.25 & 0.895 & 0.862\\
+1000 & 0.10 & 0.869 & 0.881\\
+\bottomrule
+\end{longtable}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-23-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-23}
+\end{figure}
+
 \subsection{Auxiliary Particle Filter}
@@
 \end{itemize}
+
+For illustration purposes, we are going implement an auxiliary particle
+filter with auxiliary function
+\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).\\
+
+\begin{algorithm} APF for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item For $k=1,...,N$:
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})}))}$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+The \texttt{APFfun} function resume this passages.\\
+
+\hrule
+\hrule
+\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{APFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x[k],tau)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x1}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-26-1}
+
+}
+
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-26}
+\end{figure}
+
+Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap
+Particle Filter (BPF).
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-27-1}
+
+}
+
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-27}
+\end{figure}
+
+\begin{longtable}[t]{cccc}
+\caption{\label{tab:unnamed-chunk-29}Root Mean Square Errors}\\
+\toprule
+N & Threshold & BPF & APF\\
+\midrule
+1000 & 0.50 & 0.885 & 0.875\\
+1000 & 0.25 & 0.893 & 0.892\\
+1000 & 0.10 & 0.855 & 0.870\\
+\bottomrule
+\end{longtable}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
+\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{    weight }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    \}}
+
+\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-32-1}
+
+}
+
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
+\end{figure}
 

### 2021-07-19 nuova final draft.Rmd

@@
 Let $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ and $(\mathcal{Y}, \mathcal{B}(\mathcal{Y}))$ be two (eventually equal) measurable spaces.
+
 \begin{definition}
@@
 
-Let $P_1, P_2, \dots, P_T$ be a finite sequence of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ \begin{definition}
+Let $P_1, P_2, \dots, P_T$ be a finite sequence of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$
+
+\begin{definition}
 A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
@@
 The Kalman Filter for this model is implemented as described below.
+
 \begin{algorithm} Kalman Filter for Random Walk plus Noise Model
@@
 \hrule
+\
 \code{DLM(data,sig2,tau2,m0,C0)}
+\
 \hrule
@@
 \hrule
+\
 \code{SISfun(data,N,m0,C0,tau,sigma)}
+\
 \hrule
@@
 The steps are presented in the following algorithm.
+
 \begin{algorithm} BPF for Random Walk plus Noise Model
@@
 \hrule
+\
 \code{PFfun(data,N,m0,C0,tau,sigma,r)}
+\
 \hrule
@@
 In other words the algorithm implemented in R is:
+
 \begin{algorithm} GPF for Random Walk plus Noise Model
@@
 \hrule
+\
 \code{GPFfun(data,N,m0,C0,tau,sigma,r)}
+\
 \hrule
@@
 For illustration purposes, we are going implement an auxiliary particle filter with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.
+
 \begin{algorithm} APF for Random Walk plus Noise Model
@@
 \code{APFfun(data,N,m0,C0,tau,sigma,r)}
+\
 \hrule

### 2021-07-19 Real Dataset SV.Rmd
commit

@@
 
+```{r include=FALSE}
+Errorcomp<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp)<-c("N","BPF", "GPFOPT", "APF", "SIS", "BAPF")
+rownames(Errorcomp)<-c("RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE")
+Errorcomp[,1]<-c(10,10,100,100,1000,1000,5000,5000,10000,10000)
+Errorcomp[,2]<-c(0.3780316,0.2581917,0.13740936,0.08692049,0.06682471,0.03859926,0.02801391,0.01508258,0.04292242,0.01831794)
+Errorcomp[,3]<-c(0.3888922,0.2655844,0.13385416,0.08092348,0.05924883,0.03055189,0.05669256,0.02719531,0.04300498,0.02021422)
+Errorcomp[,4]<-c(0.4952102,0.3869999,0.228981,0.1717977,0.10257034,0.07240061,0.04911414,0.02690089,0.0280415,0.01949516)
+Errorcomp[,5]<-c(0.4601913,0.3651146,0.2447648,0.1838898,0.10985867,0.06507043,0.06133971,0.02966254,0.03314042,0.01913677)
+Errorcomp[,6]<-c(0.9103835,0.7211173,1.101557,0.914675,0.6150409,0.4895908,0.6117618,0.4419522,0.5054665,0.4022079)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Auxiliary PF 50000 particles",digits=5)
+```
+
 \section{Bibliography}\

### 2021-07-19 Real Dataset SV.Rmd
update

--- 2021-07-19 Real Dataset SV.Rmd
@@
 
-We start by setting N=50000 and using the Bootstrap particle filter as the benchmark.
+We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark.
 In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
@@
 In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
+
+```{r include=FALSE}
+Errorcomp1<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp1)<-c("N","BPF", "GPFOPT", "APF", "BAPF","SIS")
+rownames(Errorcomp1)<-c("RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE")
+Errorcomp1[,1]<-c(10,10,100,100,1000,1000,5000,5000,10000,10000)
+Errorcomp1[,2]<-c(0.4007204,0.2688306,0.15268841,0.09413161,0.08482937,0.04579035,0.04275829,0.02021174,0.02931792,0.01369827)
+Errorcomp1[,3]<-c(0.4123198,0.2757273,0.15152308,0.09076196,0.06776933,0.03282099,0.06018919,0.02622987,0.03052592,0.01399114)
+Errorcomp1[,4]<-c(0.5002458,0.3905036,0.2493751,0.1812517,0.122905,0.08131335,0.06863069,0.03401982,0.04399511,0.02529053)
+Errorcomp1[,5]<-c(0.4687276,0.3711974,0.2698023,0.1945553,0.12737504,0.07408736,0.08064348,0.03614264,0.0513611,0.02515361)
+Errorcomp1[,6]<-c(0.9168639,0.7275327,1.1055065,0.9179852,0.6250516,0.4971839,0.6200359,0.4475265,0.5118927,0.406363)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp1[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Bootstrap PF 50000 particles",digits=5)
+```
+
+APF
+
+```{r include=FALSE}
+Errorcomp2<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp2)<-c("N","BPF", "GPFOPT", "APF", "BAPF", "SIS")
+rownames(Errorcomp2)<-c("RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE","RMSE","MAE")
+Errorcomp2[,1]<-c(10,10,100,100,1000,1000,5000,5000,10000,10000)
+Errorcomp2[,2]<-c(0.3780316,0.2581917,0.13740936,0.08692049,0.06682471,0.03859926,0.02801391,0.01508258,0.04292242,0.01831794)
+Errorcomp2[,3]<-c(0.3888922,0.2655844,0.13385416,0.08092348,0.05924883,0.03055189,0.05669256,0.02719531,0.04300498,0.02021422)
+Errorcomp2[,4]<-c(0.4952102,0.3869999,0.228981,0.1717977,0.10257034,0.07240061,0.04911414,0.02690089,0.0280415,0.01949516)
+Errorcomp2[,5]<-c(0.4601913,0.3651146,0.2447648,0.1838898,0.10985867,0.06507043,0.06133971,0.02966254,0.03314042,0.01913677)
+Errorcomp2[,6]<-c(0.9103835,0.7211173,1.101557,0.914675,0.6150409,0.4895908,0.6117618,0.4419522,0.5054665,0.4022079)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp2[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Auxiliary PF 50000 particles",digits=5)
+```
+
+APF vs Bootstrap
+
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=2,nrow=1,byrow=TRUE)
+colnames(Errorcomp3)<-c("RMSE", "MAE")
+rownames(Errorcomp3)<-("Discrepancy")
+Errorcomp3[1,]<-c(0.04407704,0.01667898)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3[,], align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
+```
 

### 2021-07-19 Real Dataset SV.Rmd
Commit

--- 2021-07-19 Real Dataset SV.Rmd
@@
 In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
+
+```{r include=FALSE}
+Errorcomp1<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp1)<-c("N","BPF", "GPFOPT", "APF", "BAPF","SIS")
+rownames(Errorcomp1)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
+Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
+Errorcomp1[,2]<-c(0.4007204,0.15268841,0.08482937,0.04275829,0.02931792,0.01369827,0.2688306,0.09413161,0.04579035,0.02021174)
+Errorcomp1[,3]<-c(0.4123198,0.15152308,0.06776933,0.06018919,0.03052592,0.2757273,0.09076196,0.03282099,0.02622987,0.01399114)
+Errorcomp1[,4]<-c(0.5002458,0.2493751,0.122905,0.06863069,0.04399511,0.3905036,0.1812517,0.08131335,0.03401982,0.02529053)
+Errorcomp1[,5]<-c(0.4687276,0.2698023,0.12737504,0.08064348,0.0513611,0.3711974,0.1945553,0.07408736,0.03614264,0.02515361)
+Errorcomp1[,6]<-c(0.9168639,1.1055065,0.6250516,0.6200359,0.5118927,0.7275327,0.9179852,0.4971839,0.4475265,0.406363)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp1[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Bootstrap PF 50000 particles",digits=5)
+```
+
+APF
+
+```{r include=FALSE}
+Errorcomp2<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp2)<-c("N","BPF", "GPFOPT", "APF", "BAPF", "SIS")
+rownames(Errorcomp2)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
+Errorcomp2[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
+Errorcomp2[,2]<-c(0.3780316,0.13740936,0.06682471,0.02801391,0.04292242,0.2581917,0.08692049,0.03859926,0.01508258,0.01831794)
+Errorcomp2[,3]<-c(0.3888922,0.13385416,0.05924883,0.05669256,0.04300498,0.2655844,0.08092348,0.03055189,0.02719531,0.02021422)
+Errorcomp2[,4]<-c(0.4952102,0.228981,0.10257034,0.04911414,0.0280415,0.3869999,0.1717977,0.07240061,0.02690089,0.01949516)
+Errorcomp2[,5]<-c(0.4601913,0.2447648,0.10985867,0.06133971,0.03314042,0.3651146,0.1838898,0.06507043,0.02966254,0.01913677)
+Errorcomp2[,6]<-c(0.9103835,1.101557,0.6150409,0.6117618,0.5054665,0.7211173,0.914675,0.4895908,0.4419522,0.4022079)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp2[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Auxiliary PF 50000 particles",digits=5)
+```
+
+APF vs Bootstrap
+
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=2,nrow=1,byrow=TRUE)
+colnames(Errorcomp3)<-c("RMSE", "MAE")
+rownames(Errorcomp3)<-("Discrepancy")
+Errorcomp3[1,]<-c(0.04407704,0.01667898)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3[,], align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
+```
 

### 2021-07-19 Real Dataset SV.Rmd
commit

@@
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
-\textit{Includere grafico e commento del grafico della realised volatility}
+The plots below represent the realized volatility in the period of interest for all the three indices.
+As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time).
+Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.
 
@@
 In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
-By "Basic PF we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when ESS is non negative (\textit{i.e.} always) or when ESS<0 (\textit{i.e.} never).
+By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
 In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
@@
 Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
-Errorcomp1[,2]<-c(0.4007204,0.15268841,0.08482937,0.04275829,0.02931792,0.01369827,0.2688306,0.09413161,0.04579035,0.02021174)
+Errorcomp1[,2]<-c(0.4007204,0.15268841,0.08482937,0.04275829,0.02931792,0.2688306,0.09413161,0.04579035,0.02021174,0.01369827)
 Errorcomp1[,3]<-c(0.4123198,0.15152308,0.06776933,0.06018919,0.03052592,0.2757273,0.09076196,0.03282099,0.02622987,0.01399114)
@@
 ```
+
+As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse.
+With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.
 

### 2021-07-19 Real Dataset SV.Rmd
commit

@@
 We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark.
-Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.
+Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we also conduct this analysis only on one of the three indices, the SP500.
+Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}
 
@@
 With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.
+As it turns out, such suggestion is confirmed by running the algorithms with N=1000.
+Passing from 100 to 1000 particles nearly halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a sizable hedge over the Auxiliary PF.
+Such a hedge almost vanishes when increasing the number of particles to 5000: according to the RMSE, the Bootstrap PF performs slightly better than the Guided PF, which in turn yields very similar results to the Auxiliary PF.
+We would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others.
 

### 2021-07-19 Real Dataset SV.Rmd
commit

--- 2021-07-19 Real Dataset SV.Rmd
@@
 For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
-Note that the number of particles was set at 10000 for each filter: the next section will come back to such a choice.} \
+Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \
 
@@
 We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark.
-Note that the Liu and West filter can not be included in such a comparison: trivially, the latter assumes that the competing algorithms are approximating the same target distribution.
+Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we also conduct this analysis only on one of the three indices, the SP500.
+Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}
 
 We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark.
-In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles.
+Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
+Unsurprisingly, it will be apparent that including a resampling step is essential to avoid degeneracy of the weights on the particles.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
 By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
@@
 As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse.
-With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.
+With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.\
+As it turns out, such suggestion is confirmed by running the algorithms with N=1000.
+Passing from 100 to 1000 particles nearly halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a sizable hedge over the Auxiliary PF.
+Such a hedge almost vanishes when increasing the number of particles to 5000: according to the RMSE, the Bootstrap PF performs slightly better than the Guided PF, which in turn yields very similar results to the Auxiliary PF.\
+We would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm.
+Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (as the RMSE and MAE criteria only prefer the Basic PF to the APF on a couple of occasions), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
+Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.
 
-APF
+Now, one might argue that the relative performance of the filters is influenced by the choice of the Bootstrap filter as the benchmark algorithm when run with N=50000.
+Intuitively, if the convergence phase were not reached, the Bootstrap filter would clearly find it easier to replicate the benchmark we set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.} Such criticism should definitely be taken into account, even more so when noticing that the BT filter did in fact come out on top for several values of N.\
+As a robustness check, we thus conduct the same analysis employing the Auxiliary Particle Filter as the benchmark, keeping the number of particles set to 50000 as for the benchmark BT filter.
 
@@
 
-APF vs Bootstrap
+Using the APF as the benchmark does not deliver a different picture than the one we previously had.
+Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
+Its relative performance ameliorates dramatically at N=5000, when the halving of its RMSE and MAE allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF.
+At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
+Interestingly, this time the performance of the Bootstrap filter worsens when N is increased from 5000 to 10000 (both according to RMSEs and MAEs, whose disagreements, once again, are overall infrequent).
+Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\
+Indeed, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=5000 is closer to the benchmark APF than the benchmark BT is.\footnote{According to the RMSE, also the Bootstrap with N=10000 is closer than the Bootstrap benchmark; however, in this occasion the MAE disagrees.
+Similarly, while according to the RMSE the APF with N=10000 is closer to the benchmark BT than the benchmark APF is, the MAE is lowest for the benchmark APF.}
 
@@
 library(kableExtra)
-kbl(Errorcomp3[,], align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
 ```
+
+Nevertheless, we should not overplay tiny differences. Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
+First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
+Second, when the number of particles is set at 5000 or more, neither benchmark could single out an algorithm that clearly outperformed the others, when restricting the analysis to procedures that feature adaptive resampling rules.\
+Third, when the number of particles was set to 10, 100 or 1000, the Bootstrap particle filter and the Guided particle filter with optimal proposal kernel significantly emerged as the best performing algorithms for our SV model, suggesting that their computational cost is lower than that of the Auxiliary particle filter, which needed many more particles to "get rolling".\
+Finally, some discrepancy between different approximations persisted even when the number of particles was set to 50000.
+However, the narrow extent of the RMSEs and MAEs with respect to both benchmarks supports the idea that at N=10000 the best-performing algorithms have already entered the phase of convergence.
+Even in light of these markedly more computationally expensive benchmarks, we then choose N=10000 as the number of particles which we routinely employ in this empirical application.
 

### 2021-07-19 Real Dataset SV.Rmd
Aggiunto qualche grafico

@@
 
+```{r echo=FALSE, fig.align='center', fig.cap="Close to close percentage log returns, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+library(openxlsx)
+library(ggplot2)
+library(ggpubr)
+data_SP500 <- read.xlsx("Dataset 17-21.xlsx", sheet = "SP500")
+data_STOXX50E <- read.xlsx("Dataset 17-21.xlsx", sheet = "STOXX50E")
+data_DJI <- read.xlsx("Dataset 17-21.xlsx", sheet = "DJI")
+
+# DOW JONES
+
+timestamp_DJI <- as.POSIXlt(data_DJI$Date, tz = "GMT", "%Y-%m-%d")
+data_DJI<-cbind(data_DJI, timestamp=timestamp_DJI)
+
+plot_DJI <- ggplot(data_DJI, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="DJI")+
+  theme_bw()
+
+DJI_mean <- mean(data_DJI$log_close_to_close_return)
+DJI_sd <- sd(data_DJI$log_close_to_close_return)
+
+# S&P500
+
+timestamp_SP <- as.POSIXlt(data_SP500$Date, tz = "GMT", "%Y-%m-%d")
+data_SP500<-cbind(data_SP500, timestamp=timestamp_SP)
+
+plot_SP <- ggplot(data_SP500, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="S&P500")+
+  theme_bw()
+
+SP_mean <- mean(data_SP500$log_close_to_close_return)
+SP_sd <- sd(data_SP500$log_close_to_close_return)
+
+# STOXX50E
+
+timestamp_STOXX <- as.POSIXlt(data_STOXX50E$Date, tz = "GMT", "%Y-%m-%d")
+data_STOXX50E<-cbind(data_STOXX50E, timestamp=timestamp_STOXX)
+
+plot_STOXX <- ggplot(data_STOXX50E, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="STOXX50E")+
+  theme_bw()
+
+STOXX_mean <- mean(data_STOXX50E$log_close_to_close_return)
+STOXX_sd <- sd(data_STOXX50E$log_close_to_close_return)
+
+ggarrange(plot_DJI, plot_SP, plot_STOXX, nrow = 3, ncol = 1)
+
+SC<-matrix(nrow=2,ncol=3)
+rownames(SC)<-c("Mean","Standard Deviation")
+colnames(SC) <- c("Dow Jones", "S&P500", "STOXX50E")
+
+SC[1,1] = SP_mean
+SC[1,2] = DJI_mean
+SC[1,3] = STOXX_mean
+SC[2,1] = SP_sd
+SC[2,2] = DJI_sd
+SC[2,3] = STOXX_sd
+
+```
+
 At a first visual inspection, we can see a common behaviour in the volatility of all three indices.
@@
 Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(SC, align="c", longtable = T, booktabs = T, caption="Summary statistics, close to close percentage log returns",digits=5)
+```
 
@@
 Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.
+
+```{r echo=FALSE, fig.align='center', fig.cap="Close to close percentage log returns, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+# Realized Volatility Plots
+
+plotvolsp500<-ggplot(data_SP500, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="SP")+
+  theme_bw()
+
+plotvoldji<-ggplot(data_DJI, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="DJI")+
+  theme_bw()
+
+plotvolstoxx<-ggplot(data_STOXX50E, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="STOXX")+
+  theme_bw()
+ggarrange(plotvolsp500,plotvoldji,plotvolstoxx,nrow = 3, ncol = 1)
+
+```
 
@@
 
+\section{Realised volatility as benchmark}
+
+...
+
 \section{Bibliography}\

### 2021-07-19 Real Dataset SV.Rmd
Corretta la didascalia

@@
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
-The plots below represent the realized volatility in the period of interest for all the three indices.
-As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time).
+The plots below represent the realized volatility in the period of interest for all the three indices.\footnote{In the graph, the series $rv5$ is rescaled as specified in the upcoming section: we took the square root of each value, which is a daily variance measure, and multiplied it by 100.} As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time).
 Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.
 
-```{r echo=FALSE, fig.align='center', fig.cap="Close to close percentage log returns, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Rescaled realised daily volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 

### 2021-07-19 Real Dataset SV.Rmd
Corrette tabelle rmd, risolto problema del seed per sequenza di particle filters

@@
 Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
-Errorcomp1[,2]<-c(0.4007204,0.15268841,0.08482937,0.04275829,0.02931792,0.2688306,0.09413161,0.04579035,0.02021174,0.01369827)
-Errorcomp1[,3]<-c(0.4123198,0.15152308,0.06776933,0.06018919,0.03052592,0.2757273,0.09076196,0.03282099,0.02622987,0.01399114)
-Errorcomp1[,4]<-c(0.5002458,0.2493751,0.122905,0.06863069,0.04399511,0.3905036,0.1812517,0.08131335,0.03401982,0.02529053)
-Errorcomp1[,5]<-c(0.4687276,0.2698023,0.12737504,0.08064348,0.0513611,0.3711974,0.1945553,0.07408736,0.03614264,0.02515361)
+Errorcomp1[,2]<-c(0.4233094,0.15689352,0.06901099,0.03817009,0.03045272,0.2807802,0.09658088,0.03312132,0.01889245,0.01455814)
+Errorcomp1[,3]<-c(0.3685259,0.15094281,0.07669050,0.03568058,0.02709059,0.2796141,0.09623692,0.03696833,0.01664805,0.01422964)
+Errorcomp1[,4]<-c(0.5478362,0.2404408,0.08878446,0.05742351,0.04295610,0.4310335,0.1756985,0.05538968,0.03186140,0.02448565)
+Errorcomp1[,5]<-c(0.5595929,0.2348842,0.10295053,0.06109947,0.05151115,0.4377886,0.1839363,0.06726034,0.03448388,0.02726281)
 Errorcomp1[,6]<-c(0.9168639,1.1055065,0.6250516,0.6200359,0.5118927,0.7275327,0.9179852,0.4971839,0.4475265,0.406363)
@@
 Errorcomp2[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
-Errorcomp2[,2]<-c(0.3780316,0.13740936,0.06682471,0.02801391,0.04292242,0.2581917,0.08692049,0.03859926,0.01508258,0.01831794)
-Errorcomp2[,3]<-c(0.3888922,0.13385416,0.05924883,0.05669256,0.04300498,0.2655844,0.08092348,0.03055189,0.02719531,0.02021422)
-Errorcomp2[,4]<-c(0.4952102,0.228981,0.10257034,0.04911414,0.0280415,0.3869999,0.1717977,0.07240061,0.02690089,0.01949516)
-Errorcomp2[,5]<-c(0.4601913,0.2447648,0.10985867,0.06133971,0.03314042,0.3651146,0.1838898,0.06507043,0.02966254,0.01913677)
+Errorcomp2[,2]<-c(0.4054868,0.13510456,0.04145471,0.02847469,0.03048093,0.2726597,0.08539766,0.02328878,0.01755582,0.01417144)
+Errorcomp2[,3]<-c(0.3495018,0.14208597,0.06873795,0.04921716,0.03336987,0.2694089,0.09033775,0.03531985,0.02409677,0.01717600)
+Errorcomp2[,4]<-c(0.5323702,0.2224267,0.07796291,0.03931358,0.02647522,0.4215135,0.1669123,0.05105029,0.02662432,0.01973858)
+Errorcomp2[,5]<-c(0.5459719,0.2239766,0.07692431,0.05229897,0.02410626,0.4290946,0.1748428,0.05753199,0.03275473,0.01741627)
 Errorcomp2[,6]<-c(0.9103835,1.101557,0.6150409,0.6117618,0.5054665,0.7211173,0.914675,0.4895908,0.4419522,0.4022079)

### 2021-07-19 Real Dataset Script.R

@@ -341,11 +341,23 @@ dfsv<-data.frame(timeframe,y,x)
 set.seed(12345)
 N=10000
 svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=10000
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=10000
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=10000
 svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=10000
 svgpfopt<-GPFoptfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=10000
 svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+set.seed(12345)
+N=10000
 svapfopt<-APFoptfun(y,N,m0,C0,alpha,beta,tau)
 
 #Filtering Values

### 2021-07-20 Real Dataset SV.Rmd
realised volatility RMSE MAE

@@
 
-...
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=7,nrow=6)
+colnames(Errorcomp3)<-c("N","BPF", "GPFOPT", "APF", "BAPF","LW", "SIS")
+rownames(Errorcomp3)<-c("RMSE SP500","MAE SP500","RMSE STOXX50E","MAE STOXX50E","RMSE DJI","MAE DJI")
+Errorcomp3[,1]<-c(10000,10000,10000,10000,10000,10000)
+Errorcomp3[,2]<-c(0.508642137861919,0.353767519133916,0.463584874517798,0.301898325469172,0.507624385776376,0.337543646258033)
+Errorcomp3[,3]<-c(0.509256032254369,0.35314674168209,0.463446053320389,0.304178601158929,0.506222076244991,0.338268970683459)
+Errorcomp3[,4]<-c(0.509943715852212,0.351738603428736,0.475708272485282,0.30422270387141,0.509343139732461,0.337449244130176)
+Errorcomp3[,5]<-c(0.510334461444389,0.35305864294958,0.46783471248187,0.304213895457104,0.505360139532104,0.3361502165818)
+Errorcomp3[,6]<-c(0.428716693039206,0.291557534456417,0.344065170120452,0.237159281665533,0.440832171188778,0.280952286978564)
+Errorcomp3[,7]<-c(0.602883459334495,0.431972177956328,0.652533564914959,0.440767074642285,0.604672122750974,0.415631855223678)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark",digits=5)
+```
+
+```{r include=FALSE}
+Errorcomp4<-matrix(NA,ncol=3,nrow=2)
+colnames(Errorcomp4)<-c("N","BPF", "APF")
+rownames(Errorcomp4)<-c("RMSE","MAE")
+Errorcomp4[,1]<-c(50000,50000)
+Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001)
+Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771)
+```
+
+Il fatto che con 50000 particles si performi peggio conferma che il problema sia il SV sotto, che è chiaramente battuto dal modello che sta sotto Liu West
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp4, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SP500)",digits=5)
+```
 

### 2021-07-20 Real Dataset SV.Rmd
commit

--- 2021-07-19 Real Dataset SV.Rmd
@@
 Indeed, it shows a "more detailed" sequence of states (in the sense of the sequence of filtered volatilities being less smooth) and also a higher peak for the wide fluctuation corresponding to the 2020 Covid Crisis outbreak.
-Therefore, we may say that, compared to other filtering techniques, it offers a better representation of the data.
+Therefore, we may expect that, compared to other filtering techniques, it can provide a better representation of the data.
 This is because ...
@@
 
-...
+In section XX we compared such distributions already, pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters; however, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
+We will now use a realised volatility measure as one such metric.
+
+\subsection{Comparing the filters}
+
+In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots.
+For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} With respect to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy as to determine the relative performance of the competing filters.\
+
+Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy).
+The SIS algorithm appears completely inadequate to describe the "true" volatility: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}.
+On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
+On the other hand, the Liu and West filter offers the best performance among all filters, at least visually.
+This suggests that, since the Liu and West particle filter is built on a different underlying model than the other filters considered, this model may offer a better representation of the data analyzed.
+Once more, these considerations hold for all of the three indices.\
+
+GRAFICI
+
+The above intuitions are confirmed by looking at Table XX, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark.
+Indeed, for any index, most of the filters give rather similar results in terms of RMSE and MAE.
+For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter.
+(Metterei qua l'intuizione del Liu West)\
+
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=7,nrow=6)
+colnames(Errorcomp3)<-c("N","BPF", "GPFOPT", "APF", "BAPF","LW", "SIS")
+rownames(Errorcomp3)<-c("RMSE SP500","MAE SP500","RMSE STOXX","MAE STOXX","RMSE DJI","MAE DJI")
+Errorcomp3[,1]<-c(10000,10000,10000,10000,10000,10000)
+Errorcomp3[,2]<-c(0.508642137861919,0.353767519133916,0.463584874517798,0.301898325469172,0.507624385776376,0.337543646258033)
+Errorcomp3[,3]<-c(0.509256032254369,0.35314674168209,0.463446053320389,0.304178601158929,0.506222076244991,0.338268970683459)
+Errorcomp3[,4]<-c(0.509943715852212,0.351738603428736,0.475708272485282,0.30422270387141,0.509343139732461,0.337449244130176)
+Errorcomp3[,5]<-c(0.510334461444389,0.35305864294958,0.46783471248187,0.304213895457104,0.505360139532104,0.3361502165818)
+Errorcomp3[,6]<-c(0.428716693039206,0.291557534456417,0.344065170120452,0.237159281665533,0.440832171188778,0.280952286978564)
+Errorcomp3[,7]<-c(0.602883459334495,0.431972177956328,0.652533564914959,0.440767074642285,0.604672122750974,0.415631855223678)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark",digits=4)
+```
+
+Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000 for the SP500 index (\textit{we will consider whether it is worth it to include other indexes: for the moment, I would stick to the SP500, as when considering N=10000 we are already in the convergence phase}).
+In this case, although the number of particles increases, the RMSE and MAE increase (although by very small amounts).
+This is evidence supporting the idea that the model we have selected underlying for the first two filters is inappropriate for describing this specific data.
+Indeed, as widely discussed in SECTION XX, it is obsolete and surpassed by more advanced and precise models.
+
+```{r include=FALSE}
+Errorcomp4<-matrix(NA,ncol=4,nrow=2)
+colnames(Errorcomp4)<-c("N","BPF", "APF","LWF")
+rownames(Errorcomp4)<-c("RMSE","MAE")
+Errorcomp4[,1]<-c(50000,50000)
+Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001)
+Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771)
+Errorcomp4[,4]<-c(0.4242305,0.2906932)
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp4, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SP500)",digits=5)
+```
 

### 2021-07-20 Real Dataset SV.Rmd
commit

@@
 
-Nevertheless, we should not overplay tiny differences. Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
+Nevertheless, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison.
+For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
 First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
@@
 
-Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000 for the SP500 index (\textit{we will consider whether it is worth it to include other indexes: for the moment, I would stick to the SP500, as when considering N=10000 we are already in the convergence phase}).
+Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000 for the SP500 index and the EuroStoxx50.
 In this case, although the number of particles increases, the RMSE and MAE increase (although by very small amounts).
@@
 ```{r include=FALSE}
-Errorcomp4<-matrix(NA,ncol=4,nrow=2)
+Errorcomp4<-matrix(NA,ncol=4,nrow=4)
 colnames(Errorcomp4)<-c("N","BPF", "APF","LWF")
-rownames(Errorcomp4)<-c("RMSE","MAE")
-Errorcomp4[,1]<-c(50000,50000)
-Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001)
-Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771)
-Errorcomp4[,4]<-c(0.4242305,0.2906932)
+rownames(Errorcomp4)<-c("RMSE SP500","MAE SP500","RMSE STOXX","MAE STOXX")
+Errorcomp4[,1]<-c(50000,50000,50000,50000)
+Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001,0.4650284,0.3034359)
+Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771,0.4670426,0.3040741)
+Errorcomp4[,4]<-c(0.4242305,0.2906932,0.3422955,0.2395650)
+
 ```
@@
 library(kableExtra)
-kbl(Errorcomp4, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SP500)",digits=5)
+kbl(Errorcomp4, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SP500 and STOXX50E)",digits=5)
 ```

### 2021-07-20 Real Dataset SV.Rmd
commit

@@
 In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots.
-For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} With respect to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy as to determine the relative performance of the competing filters.\
+For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy to determine the relative performance of the competing filters.\
 
 Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy).
-The SIS algorithm appears completely inadequate to describe the "true" volatility: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}.
+The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}.
 On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
@@
 
-Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000 for the SP500 index and the EuroStoxx50.
-In this case, although the number of particles increases, the RMSE and MAE increase (although by very small amounts).
-This is evidence supporting the idea that the model we have selected underlying for the first two filters is inappropriate for describing this specific data.
-Indeed, as widely discussed in SECTION XX, it is obsolete and surpassed by more advanced and precise models.
+Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices.
+We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
+In the case of the SP500 index, although the number of particles increases, the RMSE and MAE increase (although by very small amounts) for both the particle filters based on our usual specification of the SV model.
+Instead, for the Liu and West filter we see some improvements, although the small extent suggests that also the LW filter was already in the convergence phase with N=10000.\
+In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter.
 
@@
 ```
+
+Nevertheless, as we previously argued, we should not let tiny differences drive our conclusions.
+The idea behind employing also such 50000 particles approximations (\textit{i.e} more precise and computationally expensive ones) is to check whether the relative performance that we assessed with N=10000 might be significantly influenced by issues of convergence among the filters: possibly, by further approaching the exact filtering distribution implied by the basic SV model, the filters built on such specification would improve their relative performance.
+This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\
+Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.
 

### 2021-07-20 nuova final draft.Rmd
aggiunto impl Liu west

@@
 \chapter{Markov Processes and State-Space Models}
+
 \section{Notation}
@@
 Let $P_1, P_2, \dots, P_T$ be a finite sequence of probability kernels from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$ Fix some $\mathbb{P}_0(dx)$ on $(\mathcal{X}, \mathcal{B}(\mathcal{X})).$
-
+\
 \begin{definition}
@@
 Assume $\P_t,$ with $t\leq T,$ is a sequence of probability measures. Then, the following proposition holds.
+\
 \begin{proposition}
@@
 $P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
+\
 \begin{definition}\label{state-space-def}
@@
 \chapter{Linear-Gaussian case}
+
 \textit{Linear Gaussian models} are a class of popular state-space models.
@@
 Note that  $X_t\mid X_{t-1}$ and $Y_t\mid X_t$ are both normally distributed. 
+\
 \begin{lemma}\label{kalman-filter}
@@
 
-\section{Implementation}
+\subsection{Implementation}
 
@@
 Recall that, in general, resampling determines a trade-off between the increase in the variance at the resampling stage and a decrease in the variance when future states are sampled from the transition kernel.\
-``Dead" particles (i.e., those with small weights) are  likely to be removed when resampling takes place and past weights get ``reset" (usually, they are set to be equal to $1/N$).
+"Dead" particles (i.e., those with small weights) are  likely to be removed when resampling takes place and past weights get ``reset" (usually, they are set to be equal to $1/N$).
 Otherwise, the \textit{curse of dimensionality} problem will emerge and the variance of the weights will diverge over time as the variance of past weights increases as new updates occur period-by-period.
@@
 
-\section{Implementation}
+\subsection{Implementation}
 Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
@@
 
-\section{Implementation}
+\subsection{Implementation}
 In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX.
@@
 
-\section{Implementation}
+\subsection{Implementation}
 \
@@
 
-For illustration purposes, we are going implement an auxiliary particle filter with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.
+\subsection{Implementation}
+
+For illustration purposes, we are going implement an auxiliary particle filter for the linear Gaussian model with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.
 
@@
 
+We also implement the auxiliary particle filter with optimal transition kernel. Therfore the algorithm is similiar to the one discussed above with the difference that:
+\
+[[[[Inserire l'algoritmo]]]]
+
 ```{r}
@@
 
+```{r}
+###################
+## Per roberto:::::: cambia molto se usiamo l'uno o l'altro!!!!!! perchè??
+APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],x,predsd)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
+    x1   = rnorm(N,means,importancesd)
+    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x1<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
 ```{r echo=FALSE}
@@
 
+We provide an example of Liu and West filter fixing $\delta=0.7$ and drawing $\alpha_{v}$, $\beta_{v}$, $\alpha_{w}$ and $\beta_{w}$ from a Uniform $U(0,10)$.
+The results are shown in the following figure.
+
+```{r echo=FALSE}
+delta=0.7
+lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
+ESS.lwf<-ESSplot(df,lwf)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="a) LW Filtered States with credible interval (in red).
+b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+FILT.lwf<-Filtplot(df,lwf,"")
+ggarrange(FILT.lwf,ESS.lwf,labels=c("(a)","(b)"))
+```
+
 \section{Convergence}\label{pf_converg}

### 2021-07-20 Real Dataset SV.Rmd
100000 particles update

@@
 ```{r include=FALSE}
-Errorcomp3<-matrix(NA,ncol=2,nrow=1,byrow=TRUE)
+Errorcomp3<-matrix(NA,ncol=2,nrow=2)
 colnames(Errorcomp3)<-c("RMSE", "MAE")
-rownames(Errorcomp3)<-("Discrepancy")
+rownames(Errorcomp3)<-c("N=50000", "N=100000")
 Errorcomp3[1,]<-c(0.04407704,0.01667898)
-
+Errorcomp3[2,]<-c(0.03575915,0.01476220)
 ```
@@
 library(kableExtra)
-kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="Bootstrap vs Auxiliary PF, 50000 particles",digits=5)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="Discrepancy between Bootstrap and Auxiliary PFs",digits=5)
 ```
@@
 
-In section XX we compared such distributions already, pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters; however, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
-We will now use a realised volatility measure as one such metric.
+In section XX we compared several filtered distributions, pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
+However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
+In this section, we will employ a realised volatility measure as one such metric.
 

### 2021-07-20 nuova final draft.Rmd

@@
 
-```{r cll_plot, eval=FALSE, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, include=FALSE, results=F}
+```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
 load("SV/cll_plot.Rda")

### 2021-07-20 nuova final draft.Rmd
titolo aggiustato + settati i seed

@@
 ---
-title: "Final draft"
+title: | 
+ \Huge State-Space Models and Particle Filters
+
+author:
+  - Edoardo Marcelli
+  - Roberto Colarieti
+  - Alexei Verkhovtsev
+  - Michele Bolla
+  - Gabriele Romano
+  - Benedetta Bruni
 
 documentclass: book
+
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 
@@
 ```{r echo=FALSE}
+set.seed(1991)
 N=1000
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="Effective Sample Size", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+set.seed(1991)
 N=1000
@@
 ```{r include=FALSE}
+set.seed(1991)
 truex<-x
@@
 for(N in c(100,1000,10000)){
-
+set.seed(1991)
 DLM1<-DLM(y,sigma2,tau2,m0,C0)
@@
   N=1000
+  set.seed(1991)
 DLM2<-DLM(y,sigma2,tau2,m0,C0)
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='H', fig.width=10, message=FALSE, warning=FALSE, results=F}
+set.seed(1991)
 N1=100
@@
 ```{r echo=FALSE}
+set.seed(1991)
 gpf<-GPFfun(y,N,m0,C0,tau,sigma)
@@
   N=1000
-
+set.seed(1991)
 pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
@@
 ```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+set.seed(1991)
 N2=1000
@@
 ```{r echo=FALSE}
+set.seed(1991)
 apf<-APFfun(y,N,m0,C0,tau,sigma)
@@
 ```{r echo=FALSE}
+set.seed(1991)
 apfopt<-APFoptfun(y,N,m0,C0,tau,sigma)
@@
 delta=0.7
+set.seed(1991)
 lwf<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
@@
 ggarrange(FILT.lwf,ESS.lwf,labels=c("(a)","(b)"))
+```
+
+Finally, we compare the performances of the Liu and West filter and the other filtering strategies discussed so far.
+As we expect, in the linear Gaussian case, Kalman filter outperforms the other strategies.
+On the other hand, the Liu and West, with hyperparameters setted as described above, provides the worst estimates in terms of RMSE.
+This is not necessarily due to an overall weakness of the LWF with respect to the other filters, but simply to a not accurate choice of the hyperparameters.
+Moreover, while increasing the number of particles improves the accuracy, notice that the threshold for ESS-based resampling is not strictly related to better performances.
+
+```{r include=FALSE}
+set.seed(1991)
+truex<-x
+estimatedx<-matrix(NA,ncol=6,nrow=6)
+colnames(estimatedx)<-c("N","Threshold","KF","BPF","APF","LWF")
+estimatedx[,1]<-c(100,1000,10000,1000,1000,1000)
+estimatedx[,2]<-c(0.5,0.5,0.5,0.5,0.25,0.1)
+
+i=1
+for(N in c(100,1000,10000)){
+set.seed(1991)
+DLM1<-DLM(y,sigma2,tau2,m0,C0)
+pf1<-PFfun(y,N,m0,C0,tau,sigma)
+apf1<-APFfun(y,N,m0,C0,tau,sigma)
+lwf1<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T)
+
+estimatedx[i,3]<-RMSE(truex,DLM1$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf1)$mean)
+estimatedx[i,5]<-RMSE(truex,Filtervalues(apf1)$mean)
+estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf1)$mean)
+i=i+1
+
+}
+
+i=4
+for(k in c(2,5,10)){
+  set.seed(1991)
+  N=10
+DLM2<-DLM(y,sigma2,tau2,m0,C0)
+pf2<-PFfun(y,N,m0,C0,tau,sigma,r=k)
+apf2<-APFfun(y,N,m0,C0,tau,sigma,r=k)
+lwf2<-LWfun(y,N,m0,C0,1,1,1,1,delta,unif=T,r=k)
+
+estimatedx[i,3]<-RMSE(truex,DLM2$m)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(pf2)$mean)
+estimatedx[i,5]<-RMSE(truex,Filtervalues(apf2)$mean)
+estimatedx[i,6]<-RMSE(truex,Filtervalues(lwf2)$mean)
+i=i+1
+
+}
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
 ```

### 2021-07-20 Real Dataset SV.Rmd

@@
 Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database.
-Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.
-Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula, che considera returns, non returns in percentuale.
-Spiegare qui perchè possa avere senso paragonare questa varianza a quella dei log returns, come fa Guidolin} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
+Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula di rv5, che considera returns, non returns in percentuale.} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
 Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.
-To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$ (\textit{In realtà detto così è poco preciso, perchè in realtà andiamo a confrontare con la trasformazione della media dello stato filtrato}).
+To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$.
 
@@
 
-Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.
-Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Bisognerebbe fare riferimento alla sezione sulla convergenza.
-Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare già di convergenza quando N è molto grande) } Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target.
+Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
+Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
+Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare di convergenza quando N è molto grande).} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target.\
 With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
@@
 
+Guidolin, M., & Pedio, M. (2021). Media Attention vs. Sentiment as Drivers of Conditional Volatility Predictions: An Application to Brexit. Finance Research Letters, 101943.\
+
 Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\

### 2021-07-21 Real Dataset SV.Rmd
update

@@
 However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques.
-Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.
+Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
 This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
@@
 
-\section{SV models + Literary Review}
+\section{A review of Stochastic Volatility models}
 
@@
 Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database.
-Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$,\footnote{Nota sulla formula di rv5, che considera returns, non returns in percentuale.} the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.
-Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.
+Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$, the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.\footnote{Indeed, this estimator is based on the sum of 5-minute intra-day squared returns: such returns are computed as the difference between the opening and closure price in an interval.} Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.
 To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$.
@@
 Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases.
-In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.
+In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.\
 Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model.
@@
 
-In section XX we compared several filtered distributions, pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
+In section XX we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
 However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
-In this section, we will employ a realised volatility measure as one such metric.
+In this section, we will employ a realised volatility measure as one such metric.\
+A rescaled realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices we presented in the previous sections.
 

### 2021-07-21 Real Dataset SV.Rmd
commit

@@
 Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
-Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare di convergenza quando N è molto grande).} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the target.\
+Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare di convergenza quando N è molto grande).} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
 With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
 We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark.
-Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we also conduct this analysis only on one of the three indices, the SP500.
+Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500.
 Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}
@@
 Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
-Unsurprisingly, it will be apparent that including a resampling step is essential to avoid degeneracy of the weights on the particles.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
 By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
@@
 As it turns out, such suggestion is confirmed by running the algorithms with N=1000.
-Passing from 100 to 1000 particles nearly halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a sizable hedge over the Auxiliary PF.
-Such a hedge almost vanishes when increasing the number of particles to 5000: according to the RMSE, the Bootstrap PF performs slightly better than the Guided PF, which in turn yields very similar results to the Auxiliary PF.\
-We would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm.
-Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (as the RMSE and MAE criteria only prefer the Basic PF to the APF on a couple of occasions), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
+Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles.
+Further sizable gains for every filter, except for the SIS algorithm, are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
+Nevertheless, we would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm.
+Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only once, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
 Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.
 
-Now, one might argue that the relative performance of the filters is influenced by the choice of the Bootstrap filter as the benchmark algorithm when run with N=50000.
-Intuitively, if the convergence phase were not reached, the Bootstrap filter would clearly find it easier to replicate the benchmark we set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.} Such criticism should definitely be taken into account, even more so when noticing that the BT filter did in fact come out on top for several values of N.\
+Now, one might argue that the relative performance of the filters is influenced by the choice of the Bootstrap filter as the benchmark algorithm when run with N=50000: intuitively, if the convergence phase were not reached, the Bootstrap filter (which did in fact come out as one of the best performing algorithms) would clearly find it easier to replicate the benchmark we set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.}\
 As a robustness check, we thus conduct the same analysis employing the Auxiliary Particle Filter as the benchmark, keeping the number of particles set to 50000 as for the benchmark BT filter.
@@
 Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
-Its relative performance ameliorates dramatically at N=5000, when the halving of its RMSE and MAE allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF.
+Its performance ameliorates dramatically at N=1000, while at N=5000 the halving of its RMSE and MAE even allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF.
 At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
@@
 Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\
-Indeed, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=5000 is closer to the benchmark APF than the benchmark BT is.\footnote{According to the RMSE, also the Bootstrap with N=10000 is closer than the Bootstrap benchmark; however, in this occasion the MAE disagrees.
-Similarly, while according to the RMSE the APF with N=10000 is closer to the benchmark BT than the benchmark APF is, the MAE is lowest for the benchmark APF.}
 
@@
 
-Nevertheless, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison.
+In fact, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=10000 is closer to the benchmark APF than the benchmark BT is, or that the RMSE ranks the APF with N=10000 as closer to the the benchmark BT than the APF with N=50000, mildly pointing at a theoretically-unsound difference between the limiting distributions of the filters.
+Nevertheless, now the RMSE and MAE do not deliver unanimous verdicts, nor the differences seem large enough to back extreme claims.
+And, conveniently, running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
+
+Overall, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison, implying different target distributions.
 For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
@@
 Second, when the number of particles is set at 5000 or more, neither benchmark could single out an algorithm that clearly outperformed the others, when restricting the analysis to procedures that feature adaptive resampling rules.\
-Third, when the number of particles was set to 10, 100 or 1000, the Bootstrap particle filter and the Guided particle filter with optimal proposal kernel significantly emerged as the best performing algorithms for our SV model, suggesting that their computational cost is lower than that of the Auxiliary particle filter, which needed many more particles to "get rolling".\
-Finally, some discrepancy between different approximations persisted even when the number of particles was set to 50000.
+Third, when the number of particles was set to 10, 100 or 1000, the Bootstrap particle filter and the Guided particle filter with optimal proposal kernel significantly emerged as the best performing algorithms for our SV model, suggesting that their computational cost is lower than that of the Auxiliary particle filter, which needed more particles to "get rolling".\
+Finally, some discrepancy between different approximations persisted even when the number of particles was set to 50000 or 100000.
 However, the narrow extent of the RMSEs and MAEs with respect to both benchmarks supports the idea that at N=10000 the best-performing algorithms have already entered the phase of convergence.
@@
 In this section, we will employ a realised volatility measure as one such metric.\
-A rescaled realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices we presented in the previous sections.
+A realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices presented in the earlier sections.
+As we previously mentioned, a transformation is needed as to compare filtered states and realised volatility, namely one turning both into estimates of the standard deviation of the percentage log returns.
+For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by computing the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
+Throughout this section, we will deem as the best performing algorithm that coming closest to the proxy, either "literally" in a graphical framework or in a Root Mean Squared Error or Mean Absolute Error sense.
+Clearly enough, this approach builds on the assumption that realised variance is indeed a good proxy for latent volatility in real financial series.\
+
+A considerable amount of research in the 1990s and 2000s has focused on the properties of realised volatility measures, let them be realised variances or realised kernels.
+A thorough review of such literature can be found on the Oxford-Man Institute Realized volatility website, as their database specifically focuses these two measures.\footnote{Here we would mention the work by Barndorff‐Nielsen and Shephard (2002), who focus on the asymptotic properties of the realised volatility error in stochastic volatility models, namely the difference between realized volatility and the "discretised integrated volatility", also called "actual volatility".}\
+As Shephard and Sheppard (2010) themselves put it, in a paper that also presents the methodology that backs their Library, "such statistics are based on a variety of theoretically sound non-parametric estimators of the daily variation of prices"
+
+Clear assumptions
+Uses as proxy
 
@@
 Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy).
-The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}.
+The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}).
 On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
@@
 The idea behind employing also such 50000 particles approximations (\textit{i.e} more precise and computationally expensive ones) is to check whether the relative performance that we assessed with N=10000 might be significantly influenced by issues of convergence among the filters: possibly, by further approaching the exact filtering distribution implied by the basic SV model, the filters built on such specification would improve their relative performance.
-This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\
+This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Nota che riporta il fatto che in certi casi si ottengano MAE e RMSE più bassi usando N più bassi di 10000: in altre parole, approssimare un po' peggio la distribuzione implicata dal modello SV può portare a rispettare meglio la proxy.}\
 Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.
@@
 Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2002).
+Econometric analysis of realized volatility and its use in estimating stochastic volatility models.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 64(2), 253-280.\
 
@@
 
+Shephard, N., & Sheppard, K. (2010). Realising the future: forecasting with high‐frequency‐based volatility (HEAVY) models. Journal of Applied Econometrics, 25(2), 197-231.\
+
 So, M. E. P., Lam, K., & Li, W. K. (1998). A stochastic volatility model with Markov switching. Journal of Business & Economic Statistics, 16(2), 244-253.\

### 2021-07-21 Real Dataset SV.Rmd
update

@@
 However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
-In this section, we will employ a realised volatility measure as one such metric.\
+In this section, we will employ realised volatility as one such metric.\
 A realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices presented in the earlier sections.
 As we previously mentioned, a transformation is needed as to compare filtered states and realised volatility, namely one turning both into estimates of the standard deviation of the percentage log returns.
-For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by computing the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
+For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by taking the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
 Throughout this section, we will deem as the best performing algorithm that coming closest to the proxy, either "literally" in a graphical framework or in a Root Mean Squared Error or Mean Absolute Error sense.
@@
 A thorough review of such literature can be found on the Oxford-Man Institute Realized volatility website, as their database specifically focuses these two measures.\footnote{Here we would mention the work by Barndorff‐Nielsen and Shephard (2002), who focus on the asymptotic properties of the realised volatility error in stochastic volatility models, namely the difference between realized volatility and the "discretised integrated volatility", also called "actual volatility".}\
-As Shephard and Sheppard (2010) themselves put it, in a paper that also presents the methodology that backs their Library, "such statistics are based on a variety of theoretically sound non-parametric estimators of the daily variation of prices"
-
-Clear assumptions
-Uses as proxy
+As Shephard and Sheppard (2010) themselves put it, in a paper that also presents the methodology that backs their Library, "such statistics are based on a variety of theoretically sound non-parametric estimators of the daily variation of prices".
+Thanks to their theoretical justification and their being model-free estimates, in recent years realised volatility measures have been frequently used in financial applications as proxies for different types of latent volatility, providing a valuable (and arguably) preferable alternative to using squared returns (Hansen and Lunde, 2006).\footnote{See for example Kambouroudis et al.
+(2016), Buncic and Gisler (2016), Gatheral et al.(2018), Allen and McAleer (2020), Guidolin and Pedio (2021).}\
+Nevertheless, we shall not indulge into forgetting the inherent limitations of using a proxy for a latent process.
+In particular, realised volatility should not be seen as a perfect proxy of latent volatility (as, for example, realised measures ignore overnight price variations), but rather as a comparatively solid one.
+Therefore, in this section conclusions will be drawn and put forward only when supported by large differences between models and filters, while we will keep a neutral stance when such support turns out to be mild.
 
@@
 
+Buncic, D., & Gisler, K. I. (2016). Global equity market volatility spillovers: A broader role for the United States. International Journal of Forecasting, 32(4), 1317-1339.\
+
 Chan, J. C. (2013). Moving average stochastic volatility models with application to inflation forecast. Journal of Econometrics, 176(2), 162-172.\
@@
 
+Hansen, P. R., & Lunde, A. (2006). Consistent ranking of volatility models. Journal of Econometrics, 131(1-2), 97-121.\
+
 Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\
@@
 Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
+
+Kambouroudis, D.
+S., McMillan, D.
+G., & Tsakou, K.
+(2016).
+Forecasting stock return volatility: a comparison of GARCH, Implied volatility, and realized volatility models.
+Journal of Futures Markets, 36(12), 1127-1163.\
 

### 2021-07-21 prova_knit_finale.Rmd
prova draft completo

--- 2021-07-20 nuova final draft.Rmd
@@
 - \def\code#1{\texttt{#1}}
+- \usepackage{babel} 
 - \usepackage{amssymb}
+- \usepackage{amsfonts}
 - \usepackage{txfonts}
 - \usepackage{amsmath}
+- \usepackage{mathtools}
 - \usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
@@
 \chapter{Applications to Stochastic Volatility Models}
+
+\section{Overview}
+
+In the following sections, we will try to replicate some of the analyses which were previously conducted on synthetic data, generated by a basic stochastic volatility model; this time, the same model specification and filters will be used on real data.
+However, as we will see, real data require different strategies to evaluate the effectiveness of several Sequential Monte Carlo techniques.
+Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
+This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
+In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
+Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
+In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.
+
+\section{A review of Stochastic Volatility models}
+
+In section \textit{fill with the number of the section} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
+Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
+
+This specification - which we will employ in this section on a real dataset - can be deemed as the standard version of the SV model.
+Throughout the recent decades, several extensions have been proposed, mostly (but not exclusively) in the field of financial econometrics.
+On the one hand, for simplicity we will stick to the standard version, one that will suffice for the scope of this work on Sequential Monte Carlo methods.
+On the other hand, it is worth mentioning some of these extensions: indeed, one can rather safely assume that, through their higher degree of sophistication, such extensions might well improve on the performance of our standard SV model in the analyses that will follow.\
+\
+Stochastic Volatility Models, whose early formulation is commonly attributed to Taylor (1982, 1986), allow to account for time-varying and autoregressive volatility in financial returns, posing themselves as a valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models in dealing with non-constant volatility.
+Kim, Shephard and Chib (1998) define the canonical model for regularly spaced data as
+\begin{align*}
+    y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\tau \eta_t\\
+    x_1|\sigma,\beta & \sim  N\Big(\mu,\frac{\sigma^2}{1-\beta^2}\Big)\\
+    \epsilon_t & \sim N(0,1)\\ 
+    \eta_t & \sim N(0,1)
+\end{align*} 
+where the log volatility $x_t$ is assumed to follow a stationary process ($|\beta|<1$), $\psi$ is a constant scaling factor, $\beta$ is the persistence in the volatility and $\tau$ is the volatility of the log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$.
+Kim, Shephard and Chib (1998) prefer $\psi=1$.} Most importantly, $\epsilon_t$ and $\eta_t$, the Gaussian white noise processes that drive the canonical model, are assumed to be uncorrelated.
+This latter assumption justifies another definition of this specification, namely "discrete SV model without leverage".\footnote{Note that in this review we do not focus on continuous time SV models.
+In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing.
+Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility.
+Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al.
+(2010), who investigate alternatives to the entrenched affine square root SV model.
+Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties.
+More recently, Gatheral et al.
+(2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\
+In order to accommodate for leverage effects, namely capture the increase in volatility that follows a drop in the returns, the model can extended as in Omori et al. (2007), \begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}
+\end{align*} 
+where $\rho<0$ captures the negative correlation.
+Such specification, which can be referred to as "discrete SV model with leverage", captures the asymmetric response of volatility to returns of different signs, so that similar specifications are sometimes also deemed as "asymmetric SV models" (e.g.
+Harvey and Shephard 1996, Mao et al.
+2020).\
+Note that so far the assumptions on $\epsilon_t$ and $\eta_t$ implied that the returns are conditionally normally distributed.
+The discrete time SV models can also be extended to allow for heavy-tailed or asymmetric conditional returns distributions: symmetric or skewed Student-$t$, Generalised Hyperbolic (GH) distribution, Generalised Error Distribution (GED) and scale mixtures of normals feature as popular choices (Kim and Stoffer 2008, Nakajima and Omori 2012, Mao et al.
+2020).
+In fact, SV with heavy tailed return distributions were shown to better meet empirical regularities like the leptokurtic distribution of the returns and slowly decaying autocorrelation functions of the squared returns (Liesenfeld and Jung 2000).\
+Assuming that $\epsilon_t$ follows a Student-$t$ distribution, and exploiting the fact that $\epsilon_t$ can then be written as $\lambda_t^{-1/2}\zeta_t$, where $\zeta_t \sim N(0,1)$ and $v\lambda_t \sim \chi^2_v$ (Harvey et al.
+1994, Chib et al.
+2002), we have the following SV model with both fat tails and leverage effect (Jacquier et al., 2004),
+\begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
+    x_{t+1} & = \alpha + \beta x_t + \eta_t\\
+    \begin{pmatrix}
+       \zeta_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}\\
+    v\lambda_t & \sim \chi^2_v
+\end{align*}
+To capture other elements of the behaviour of financial data, several other extensions have been proposed.
+For example, SV models have been extended to include conditional heteroskedasticity in the mean returns (Koopman and Hol Uspensky 2002) to capture potential volatility feedback effects, or to feature autoregressive moving average innovations (Chan 2013, Zhang et al.
+2020), allowing better goodness of fit and out-of-sample forecasts.\footnote{Dimitrakopoulos and Kolossiatis (2020) note that "the moving average component, the leverage effect and the conditional heteroscedasticity in mean have been considered separately in the stochastic volatility literature" and provide two specifications, one featuring an MA component and leverage effects, the other an MA component and conditional heteroskedasticity in mean.} \footnote{Other extensions, though ones for which we do not report the specifications, feature modelling the latent volatility process $x_t$ as an ARFIMA process (Long Memory Stochastic Volatility model, Breidt et al.
+1998) or as governed by a first-order Markov process (Markov Switching Stochastic Volatility model, So et al.
+1998).
+Recently, Luo et al.
+(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).} The model proposed by Koopman and Hol Uspensky is specified as
+\begin{align*}
+    y_t & = \nu_t+\psi e^{\frac{x_t}{2}}\epsilon_t\\
+    \nu_t & = a+by_{t-1}+d\psi^2e^{x_t}\\
+    x_t & = \beta x_{t-1}+\tau \eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} & \overset{i.i.d.}{\sim}  N_2\Big(\mathbf{0},\begin{pmatrix}
+       1 & 0 \\
+       0 & 1
+    \end{pmatrix}\Big)
+\end{align*} 
+while the state space representation of the ARMA(p,q)-SV framework, as in Zhang et al. (2020), reads 
+\begin{align*}
+    y_t & = \nu_t + \gamma_t\\
+    \gamma_t & =\phi_1\gamma_{t-1}+...+\phi_p\gamma_{t-p}+u_t+\varphi_1 u_{t-1}+...+\varphi_q u_{t-q}\\
+    u_t|x_t & \sim N(0,e^{x_t})\\
+    x_{t} & = x_{t-1}+ \eta_t\\
+    \eta_t|\tau & \sim N(0,\tau^2)
+\end{align*} 
+where the error terms $u_t$ and $\eta_t$ are independent across all leads and lags, while $\nu_t$ follows an unspecified time-varying process.\
+\
+Interestingly enough, the path along which the SV models evolved coincides with that suggested by optimal portfolio findings.
+Johannes, Korteweg and Polson (2014) found that, in order to generate statistically significant portfolio improvements in a Bayesian learning problem, the model employed by the investor should incorporate both time-varying expected returns and stochastic volatility: indeed, either of these features alone did not lead to statistically significant gains with respect to employing models with time-constant expected returns and volatility.\footnote{One could then argue that caution is needed when employing basic specifications of stochastic volatility models.
+For instance, Poon and Granger (2005) found that historical volatility and ARCH models both achieved better volatility forecasting performance than SV models.
+Similarly, Allen and McAleer (2020, see also Allen, 2020) found that, using realised volatility as benchmark, neither the canonical SV model or a GARCH(1,1) specification could forecast better than a simple form of historical volatility model.}\
+\
+Finally, before moving to the application of Sequential Monte Carlo techniques to a SV model, we conclude by going through some references for the Bayesian analysis proposed for such models.
+Starting from the seminal work of Jacquier et al.
+(1994), the use of MCMC methods has become increasingly popular for parameter estimation and smoothing exercises in SV models (e.g.
+Kastner 2019, presenting the R package \textit{stochvol} for Bayesian parameter estimation, and Chopin and Papaspiliopoulos 2020, who use MCMC to sample from the smoothing distribution of a SV model).
+As regards filtering exercises, the adoption of particle filters was rather rapid:\footnote{Actually, SV models are now often used as straightforward applications of particle filters on non-linear state space models, see for example Andrieu et al.
+(2010), Douc et al.
+(2014) or Chopin and Papaspiliopoulos (2020).} indeed, latent volatilities in Kim et al.
+(1998) were already filtered by employing the particle filter suggested in Pitt and Shephard (1999), paving the way for subsequent applications (\textit{inter alia}) in Chib et al.
+(2006), Omori et al.
+(2007), Kim and Stoffer (2008) and Nakajima and Omori (2012).\
+
+\section{Data Description}
+
+As previously mentioned, we will analyze the behaviour of the described filtering tools associated to a simple stochastic volatility model.
+In particular, we are evaluating the performance of such model on real data.
+For the observable process, namely financial returns in the SV model, we consider the continuously compounded daily returns (also called logarithmic returns) of three indices, S&P500, DOW JONES and STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021.
+From these, we estimate the daily volatility, as computed by the model.
+
+The indices have been selected as representatives of the global economic trends in the US and EU markets.
+Specifically, the S&P500 is a market-capitalization-weighted stock-price index tracing the performance of the 500 largest companies listed on US stock exchanges (NYSE and Nasdaq Exchange).
+The DOW JONES, instead, is a price-weighted stock-market index and accounts for the 30 major companies listed on US stock exchanges, characterized for being "blue-chip".
+Also the EURO STOXX 50 follows bluce-chip stocks representing leading firms in regions of the Eurozone.\
+
+The three plots below represent the time series of the log returns calculated for the three indices during our period of interest.
+In particular, each series is a sequence of daily observations representing the logarithm of the ratio between the closing price of the index for a given day and the closing price of the day before.
+
+```{r echo=FALSE, fig.align='center', fig.cap="Close to close percentage log returns, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+library(openxlsx)
+library(ggplot2)
+library(ggpubr)
+data_SP500 <- read.xlsx("Dataset 17-21.xlsx", sheet = "SP500")
+data_STOXX50E <- read.xlsx("Dataset 17-21.xlsx", sheet = "STOXX50E")
+data_DJI <- read.xlsx("Dataset 17-21.xlsx", sheet = "DJI")
+
+# DOW JONES
+
+timestamp_DJI <- as.POSIXlt(data_DJI$Date, tz = "GMT", "%Y-%m-%d")
+data_DJI<-cbind(data_DJI, timestamp=timestamp_DJI)
+
+plot_DJI <- ggplot(data_DJI, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="DJI")+
+  theme_bw()
+
+DJI_mean <- mean(data_DJI$log_close_to_close_return)
+DJI_sd <- sd(data_DJI$log_close_to_close_return)
+
+# S&P500
+
+timestamp_SP <- as.POSIXlt(data_SP500$Date, tz = "GMT", "%Y-%m-%d")
+data_SP500<-cbind(data_SP500, timestamp=timestamp_SP)
+
+plot_SP <- ggplot(data_SP500, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="S&P500")+
+  theme_bw()
+
+SP_mean <- mean(data_SP500$log_close_to_close_return)
+SP_sd <- sd(data_SP500$log_close_to_close_return)
+
+# STOXX50E
+
+timestamp_STOXX <- as.POSIXlt(data_STOXX50E$Date, tz = "GMT", "%Y-%m-%d")
+data_STOXX50E<-cbind(data_STOXX50E, timestamp=timestamp_STOXX)
+
+plot_STOXX <- ggplot(data_STOXX50E, aes(x=timestamp,y=log_close_to_close_return))+
+  geom_line()+
+  labs(x="Date",y="STOXX50E")+
+  theme_bw()
+
+STOXX_mean <- mean(data_STOXX50E$log_close_to_close_return)
+STOXX_sd <- sd(data_STOXX50E$log_close_to_close_return)
+
+ggarrange(plot_DJI, plot_SP, plot_STOXX, nrow = 3, ncol = 1)
+
+SC<-matrix(nrow=2,ncol=3)
+rownames(SC)<-c("Mean","Standard Deviation")
+colnames(SC) <- c("Dow Jones", "S&P500", "STOXX50E")
+
+SC[1,1] = SP_mean
+SC[1,2] = DJI_mean
+SC[1,3] = STOXX_mean
+SC[2,1] = SP_sd
+SC[2,2] = DJI_sd
+SC[2,3] = STOXX_sd
+
+```
+
+At a first visual inspection, we can see a common behaviour in the volatility of all three indices.
+Especially, it is worth pointing out the similarity between the pattern of DJI and that of S&P500 (especially in terms of peaks), which both differ from the STOXX50E, presumably because the first two describe the US stock market, while the last index describes the European one.
+For example, a difference that can be seen by having a glance at the plots is that the US indices log returns displays differences in fluctuation magnitude across different periods that are more marked than for the EU index, for which log returns display, in general, wider fluctuations.
+Anyways, what stands out the most in all three plots is the very wide fluctuations present from February - March 2020 to around September 2020, which, very intuitively, are connected with the Covid-19 Pandemic Crisis.
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(SC, align="c", longtable = T, booktabs = T, caption="Summary statistics, close to close percentage log returns",digits=5)
+```
+
+As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \textit{fill with the number of the section}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
+Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
+The plots below represent the realized volatility in the period of interest for all the three indices.\footnote{In the graph, the series $rv5$ is rescaled as specified in the upcoming section: we took the square root of each value, which is a daily variance measure, and multiplied it by 100.} As usual, they show a rather similar pattern across all three indices, identifying peaks corresponding to periods of higher uncertainty (in particular referring to the 2020 Covid-19 Crisis outbreak and its impact over time).
+Once more, there is great similarity between the series for the DJI and the one for the S&P500, while the series of the STOXX50E appears slightly different, in terms of peaks.
+
+```{r echo=FALSE, fig.align='center', fig.cap="Rescaled realised daily volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+# Realized Volatility Plots
+
+plotvolsp500<-ggplot(data_SP500, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="SP")+
+  theme_bw()
+
+plotvoldji<-ggplot(data_DJI, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="DJI")+
+  theme_bw()
+
+plotvolstoxx<-ggplot(data_STOXX50E, aes(x= timestamp, y=rv5))+
+  geom_line()+
+  labs(x="Date",y="STOXX")+
+  theme_bw()
+ggarrange(plotvolsp500,plotvoldji,plotvolstoxx,nrow = 3, ncol = 1)
+
+```
+
+\subsection{Rescaling the data}
+
+When specifying the parameters of a stochastic volatility model, one should clarify whether the series of observed financial returns is one of log returns or of percentage log returns.
+Indeed, let $y_t$ denote the returns at time $t$, so that a basic SV specification would prescribe $y_t|x_t\sim N(0,e^{x_t})$: clearly, for each period $t$, $e^{x_t}$ needs to be higher to replicate the observed \textit{percentage} log return rather than the observed log return.
+This implies that the latent volatility $x$ should fluctuate around higher values when using percentage log returns as $y_t$.
+This feature, when adopting the aforementioned canonical SV model as defined in Kim et al.
+(1998), would be represented by a higher $\mu$.
+For example, as noted in Kastner (2019), daily log returns often have a variance of 0.0001 or less, which implies that $\mu$ should lie somewhere around $\ln(0.0001)\approx -9$; instead, when considering daily percentage log returns, one then has a variance of 1, so that $\mu=\ln(1)=0$.
+In other terms, one might end up with a basic specification of the SV model when (roughly) calibrating the parameters of the canonical model on a wider set of financial returns.
+
+Note that rescaling the log returns has analogous implications for the section in which we employ measures of realised volatility as a proxy for the latent volatility process.
+Let us denote by $rv5_t$ the daily realised variance at time $t$, which we obtain from the \textit{rv5} series provided in the Oxford-Man database.
+Now, note that in a SV framework one should not directly compare $rv5_t$ to $x_t$, as $rv5_t$ can be rather seen as an estimate of the variance of the log returns at time $t$, \textit{i.e.} of $e^{x_t}$ in the SV model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.} Moreover, by construction of $rv5_t$, the comparison should feature a SV model where the log returns did \textit{not} undergo the percentage transformation.\footnote{Indeed, this estimator is based on the sum of 5-minute intra-day squared returns: such returns are computed as the difference between the opening and closure price in an interval.} Thus, in order to achieve a common scale with the percentage log returns, we need to multiply $rv5$ by $100^2$.
+To ease the interpretation and representation of the comparison between differently approximated filtering distributions, we finally take the square root of the rescaled $rv5$, so that we compare it to $e^{x_t/2}$.
+
+\section{Comparing the filters}
+
+In this section the different particle filter techniques we discussed in earlier sections will be employed to approximate the filtering distribution that springs from a simple SV model applied to a real dataset.
+First, we provide a qualitative comparison between the filters, basing on visual analysis of the (approximated) filtering distributions.
+Subsequently, we move to a more quantitative approach, in order to assess the quality of the approximations as the number of particles increases.
+In a way, while in the first part most filters can be rather confidently assumed to be close to convergence, the second subsection will hint at possibly different computational costs across the competing filters.\
+Before moving to such analyses, it is worth mentioning that not all the filters we employ approximate the same filtering distribution $p(x_t|y_{1:t},\psi)$, where $\psi$ is a vector containing the parameters of the model.
+Indeed, the Liu and West filter actually approximates the distribution of a slightly different SV model, one that assumes an additional hierarchical level,
+\begin{align*}
+y_t|x_t & \sim N(0,e^{x_t})\\
+x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
+x_0 & \sim N(0,100)\\
+\alpha & \sim N(\gamma,\zeta)\\
+\beta & \sim N(\pi,\phi)\\
+\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
+\end{align*}
+where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\
+
+\subsection{Visual Analysis after Filtering}
+
+We now offer a preliminary comparison of the various filtering methods, based on a visual analysis of graphs representing, for each index, the time-varying volatility of the observations, obtained through the application of the different filtering techniques explored in the theoretical analysis.
+Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows.
+For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
+Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \
+
+First, a general point is that the sequence of percentage log returns are different across the three indices.
+The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents (SEE SECTION X).
+In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy.
+However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy.
+Aside from this note, the considerations expressed below on the sequences "filtered volatilities"\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.} for the various techniques hold almost uniformly for all the three indices.
+Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\
+
+Generally speaking, note that, the sequence of "filtered volatilities" obtained from any of the filtering techniques follows roughly similar paths across all the plots.
+Unsurprisingly, the filtered volatility appears wider in periods corresponding to wider fluctuations of the percentage log returns and lower in periods corresponding to smaller such fluctuations.
+For example, all sequences of filtered volatilities show peaks corresponding to the higher fluctuations due to the 2020 Covid-19 Crisis outbreak and the consequent increases in uncertainty for the later periods.\
+
+When focusing the attention on specific filters, first we can discuss the effect of resampling.
+Indeed, in the plot referring to the Sequential Importance Sampling procedure, the sequence of filtered volatilities is not able to reproduce the peaks as in with other filters, but also is characterized by collapsing confidence intervals.
+This is because ::::::: \
+Then, if we instead look at the filtered volatilities obtained with resampling, we can appreciate how the filtered sequences for the Basic,\footnote{By this name, we refer to a particle filter that features no adaptive resampling; rather, resampling is conducted at each step of the algorithm.} Bootstrap, Auxiliary and Optimal Kernel Guided particle filters give rather analogous results, with no differences appearing at a first visual inspection.\footnote{Except for the Auxiliary Particle filtered sequence of volatilities, showing a slightly higher peak corresponding to the greatest volatility fluctuation in 2020, if compared to other filters in this subset.
+This is still valid for all three indices.} \
+However, the sequence of filtered volatilities obtained by applying the Liu and West filter is clearly different from the others.
+Indeed, it shows a "more detailed" sequence of states (in the sense of the sequence of filtered volatilities being less smooth) and also a higher peak for the wide fluctuation corresponding to the 2020 Covid Crisis outbreak.
+Therefore, we may expect that, compared to other filtering techniques, it can provide a better representation of the data.
+This is because ...
+\textit{Potremmo parlare di maggiore sensibilità alle oscillazioni dei prezzi, o simili}
+
+\subsection{Quality of approximation}
+
+Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
+Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
+Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare di convergenza quando N è molto grande).} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
+With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
+We employ the Root Mean Squared Error and the Mean Absolute Error to measure discrepancies from the benchmark.
+Note that the Liu and West filter can not be included in such an analysis: trivially, the comparison assumes that the competing algorithms are approximating the same target distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500.
+Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}
+
+We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark.
+In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles.
+Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
+Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
+In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
+
+```{r include=FALSE}
+Errorcomp1<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp1)<-c("N","BPF", "GPFOPT", "APF", "BAPF","SIS")
+rownames(Errorcomp1)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
+Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
+Errorcomp1[,2]<-c(0.4233094,0.15689352,0.06901099,0.03817009,0.03045272,0.2807802,0.09658088,0.03312132,0.01889245,0.01455814)
+Errorcomp1[,3]<-c(0.3685259,0.15094281,0.07669050,0.03568058,0.02709059,0.2796141,0.09623692,0.03696833,0.01664805,0.01422964)
+Errorcomp1[,4]<-c(0.5478362,0.2404408,0.08878446,0.05742351,0.04295610,0.4310335,0.1756985,0.05538968,0.03186140,0.02448565)
+Errorcomp1[,5]<-c(0.5595929,0.2348842,0.10295053,0.06109947,0.05151115,0.4377886,0.1839363,0.06726034,0.03448388,0.02726281)
+Errorcomp1[,6]<-c(0.9168639,1.1055065,0.6250516,0.6200359,0.5118927,0.7275327,0.9179852,0.4971839,0.4475265,0.406363)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp1[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Bootstrap PF 50000 particles",digits=5)
+```
+
+As we can see, almost every filter shows dramatic improvement already when passing from 10 particles to 100 particles, while the performance of the SIS algorithm actually gets worse.
+With N=10 or N=100, the Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge quite clearly as the best performing filters; however, no conclusion should be drawn at this stage, as the large errors seem to suggest quite a wide margin from improvement even for the best performing filters.\
+As it turns out, such suggestion is confirmed by running the algorithms with N=1000.
+Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles.
+Further sizable gains for every filter, except for the SIS algorithm, are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
+Nevertheless, we would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm.
+Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only once, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
+Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.
+
+Now, one might argue that the relative performance of the filters is influenced by the choice of the Bootstrap filter as the benchmark algorithm when run with N=50000: intuitively, if the convergence phase were not reached, the Bootstrap filter (which did in fact come out as one of the best performing algorithms) would clearly find it easier to replicate the benchmark we set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.}\
+As a robustness check, we thus conduct the same analysis employing the Auxiliary Particle Filter as the benchmark, keeping the number of particles set to 50000 as for the benchmark BT filter.
+
+```{r include=FALSE}
+Errorcomp2<-matrix(NA,ncol=6,nrow=10)
+colnames(Errorcomp2)<-c("N","BPF", "GPFOPT", "APF", "BAPF", "SIS")
+rownames(Errorcomp2)<-c("RMSE","RMSE","RMSE","RMSE","RMSE","MAE","MAE","MAE","MAE","MAE")
+Errorcomp2[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
+Errorcomp2[,2]<-c(0.4054868,0.13510456,0.04145471,0.02847469,0.03048093,0.2726597,0.08539766,0.02328878,0.01755582,0.01417144)
+Errorcomp2[,3]<-c(0.3495018,0.14208597,0.06873795,0.04921716,0.03336987,0.2694089,0.09033775,0.03531985,0.02409677,0.01717600)
+Errorcomp2[,4]<-c(0.5323702,0.2224267,0.07796291,0.03931358,0.02647522,0.4215135,0.1669123,0.05105029,0.02662432,0.01973858)
+Errorcomp2[,5]<-c(0.5459719,0.2239766,0.07692431,0.05229897,0.02410626,0.4290946,0.1748428,0.05753199,0.03275473,0.01741627)
+Errorcomp2[,6]<-c(0.9103835,1.101557,0.6150409,0.6117618,0.5054665,0.7211173,0.914675,0.4895908,0.4419522,0.4022079)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp2[,], align="c", longtable = T, booktabs = T, caption="RMSE and MAE, Auxiliary PF 50000 particles",digits=5)
+```
+
+Using the APF as the benchmark does not deliver a different picture than the one we previously had.
+Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
+Its performance ameliorates dramatically at N=1000, while at N=5000 the halving of its RMSE and MAE even allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF.
+At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
+Interestingly, this time the performance of the Bootstrap filter worsens when N is increased from 5000 to 10000 (both according to RMSEs and MAEs, whose disagreements, once again, are overall infrequent).
+Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\
+
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=2,nrow=2)
+colnames(Errorcomp3)<-c("RMSE", "MAE")
+rownames(Errorcomp3)<-c("N=50000", "N=100000")
+Errorcomp3[1,]<-c(0.04407704,0.01667898)
+Errorcomp3[2,]<-c(0.03575915,0.01476220)
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="Discrepancy between Bootstrap and Auxiliary PFs",digits=5)
+```
+
+In fact, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=10000 is closer to the benchmark APF than the benchmark BT is, or that the RMSE ranks the APF with N=10000 as closer to the the benchmark BT than the APF with N=50000, mildly pointing at a theoretically-unsound difference between the limiting distributions of the filters.
+Nevertheless, now the RMSE and MAE do not deliver unanimous verdicts, nor the differences seem large enough to back extreme claims.
+And, conveniently, running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
+
+Overall, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison, implying different target distributions.
+For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
+First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
+Second, when the number of particles is set at 5000 or more, neither benchmark could single out an algorithm that clearly outperformed the others, when restricting the analysis to procedures that feature adaptive resampling rules.\
+Third, when the number of particles was set to 10, 100 or 1000, the Bootstrap particle filter and the Guided particle filter with optimal proposal kernel significantly emerged as the best performing algorithms for our SV model, suggesting that their computational cost is lower than that of the Auxiliary particle filter, which needed more particles to "get rolling".\
+Finally, some discrepancy between different approximations persisted even when the number of particles was set to 50000 or 100000.
+However, the narrow extent of the RMSEs and MAEs with respect to both benchmarks supports the idea that at N=10000 the best-performing algorithms have already entered the phase of convergence.
+Even in light of these markedly more computationally expensive benchmarks, we then choose N=10000 as the number of particles which we routinely employ in this empirical application.
+
 \section{Comparing Constant and Stochastic Volatility Models of Equity Returns}
@@
 a particularly significant improvement in the period of extreme values of returns in March 2020.
+
+\section{Realised volatility as benchmark}
+
+In section XX we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
+However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
+In this section, we will employ realised volatility as one such metric.\
+A realised volatility measure - the realised variance based on 5 min intraday return intervals ($rv5$ series in the Oxford-Man Realized Library) - will be used as a proxy for the latent volatility process driving the log returns on the 3 indices presented in the earlier sections.
+As we previously mentioned, a transformation is needed as to compare filtered states and realised volatility, namely one turning both into estimates of the standard deviation of the percentage log returns.
+For each $t$, $rv5_t$ is then turned into $rv5_t^{0.5}*100$, while $e^{x_t/2}$ (where $x_t$ is the filtered state) is computed by taking the mean of the generated particles and then applying the $exp(\cdot /2)$ transformation.\
+Throughout this section, we will deem as the best performing algorithm that coming closest to the proxy, either "literally" in a graphical framework or in a Root Mean Squared Error or Mean Absolute Error sense.
+Clearly enough, this approach builds on the assumption that realised variance is indeed a good proxy for latent volatility in real financial series.\
+
+A considerable amount of research in the 1990s and 2000s has focused on the properties of realised volatility measures, let them be realised variances or realised kernels.
+A thorough review of such literature can be found on the Oxford-Man Institute Realized volatility website, as their database specifically focuses these two measures.\footnote{Here we would mention the work by Barndorff‐Nielsen and Shephard (2002), who focus on the asymptotic properties of the realised volatility error in stochastic volatility models, namely the difference between realized volatility and the "discretised integrated volatility", also called "actual volatility".}\
+As Shephard and Sheppard (2010) themselves put it, in a paper that also presents the methodology that backs their Library, "such statistics are based on a variety of theoretically sound non-parametric estimators of the daily variation of prices".
+Thanks to their theoretical justification and their being model-free estimates, in recent years realised volatility measures have been frequently used in financial applications as proxies for different types of latent volatility, providing a valuable (and arguably) preferable alternative to using squared returns (Hansen and Lunde, 2006).\footnote{See for example Kambouroudis et al.
+(2016), Buncic and Gisler (2016), Gatheral et al.(2018), Allen and McAleer (2020), Guidolin and Pedio (2021).}\
+Nevertheless, we shall not indulge into forgetting the inherent limitations of using a proxy for a latent process.
+In particular, realised volatility should not be seen as a perfect proxy of latent volatility (as, for example, realised measures ignore overnight price variations), but rather as a comparatively solid one.
+Therefore, in this section conclusions will be drawn and put forward only when supported by large differences between models and filters, while we will keep a neutral stance when such support turns out to be mild.
+
+\subsection{Comparing the filters}
+
+In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots.
+For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy to determine the relative performance of the competing filters.\
+
+Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy).
+The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}).
+On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\
+On the other hand, the Liu and West filter offers the best performance among all filters, at least visually.
+This suggests that, since the Liu and West particle filter is built on a different underlying model than the other filters considered, this model may offer a better representation of the data analyzed.
+Once more, these considerations hold for all of the three indices.\
+
+GRAFICI
+
+The above intuitions are confirmed by looking at Table XX, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark.
+Indeed, for any index, most of the filters give rather similar results in terms of RMSE and MAE.
+For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter.
+(Metterei qua l'intuizione del Liu West)\
+
+```{r include=FALSE}
+Errorcomp3<-matrix(NA,ncol=7,nrow=6)
+colnames(Errorcomp3)<-c("N","BPF", "GPFOPT", "APF", "BAPF","LW", "SIS")
+rownames(Errorcomp3)<-c("RMSE SP500","MAE SP500","RMSE STOXX","MAE STOXX","RMSE DJI","MAE DJI")
+Errorcomp3[,1]<-c(10000,10000,10000,10000,10000,10000)
+Errorcomp3[,2]<-c(0.508642137861919,0.353767519133916,0.463584874517798,0.301898325469172,0.507624385776376,0.337543646258033)
+Errorcomp3[,3]<-c(0.509256032254369,0.35314674168209,0.463446053320389,0.304178601158929,0.506222076244991,0.338268970683459)
+Errorcomp3[,4]<-c(0.509943715852212,0.351738603428736,0.475708272485282,0.30422270387141,0.509343139732461,0.337449244130176)
+Errorcomp3[,5]<-c(0.510334461444389,0.35305864294958,0.46783471248187,0.304213895457104,0.505360139532104,0.3361502165818)
+Errorcomp3[,6]<-c(0.428716693039206,0.291557534456417,0.344065170120452,0.237159281665533,0.440832171188778,0.280952286978564)
+Errorcomp3[,7]<-c(0.602883459334495,0.431972177956328,0.652533564914959,0.440767074642285,0.604672122750974,0.415631855223678)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp3, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark",digits=4)
+```
+
+Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices.
+We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
+In the case of the SP500 index, although the number of particles increases, the RMSE and MAE increase (although by very small amounts) for both the particle filters based on our usual specification of the SV model.
+Instead, for the Liu and West filter we see some improvements, although the small extent suggests that also the LW filter was already in the convergence phase with N=10000.\
+In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter.
+
+```{r include=FALSE}
+Errorcomp4<-matrix(NA,ncol=4,nrow=4)
+colnames(Errorcomp4)<-c("N","BPF", "APF","LWF")
+rownames(Errorcomp4)<-c("RMSE SP500","MAE SP500","RMSE STOXX","MAE STOXX")
+Errorcomp4[,1]<-c(50000,50000,50000,50000)
+Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001,0.4650284,0.3034359)
+Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771,0.4670426,0.3040741)
+Errorcomp4[,4]<-c(0.4242305,0.2906932,0.3422955,0.2395650)
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(Errorcomp4, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SP500 and STOXX50E)",digits=5)
+```
+
+Nevertheless, as we previously argued, we should not let tiny differences drive our conclusions.
+The idea behind employing also such 50000 particles approximations (\textit{i.e} more precise and computationally expensive ones) is to check whether the relative performance that we assessed with N=10000 might be significantly influenced by issues of convergence among the filters: possibly, by further approaching the exact filtering distribution implied by the basic SV model, the filters built on such specification would improve their relative performance.
+This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Nota che riporta il fatto che in certi casi si ottengano MAE e RMSE più bassi usando N più bassi di 10000: in altre parole, approssimare un po' peggio la distribuzione implicata dal modello SV può portare a rispettare meglio la proxy.}\
+Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.
+
+\section{Bibliography}\
+
+Allen, D. E., & McAleer, M. (2020). Do we need stochastic volatility and generalised autoregressive conditional heteroscedasticity? Comparing squared end-of-day returns on FTSE. Risks, 8(1), 12.\
+
+Allen, D. E. (2020). Stochastic Volatility and GARCH: Do Squared End-of-Day Returns Provide Similar Information?. Journal of Risk and Financial Management, 13(9), 202.\
+
+Andrieu, C., Doucet, A., & Holenstein, R. (2010). Particle markov chain monte carlo methods. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 72(3), 269-342.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2001).
+Non‐Gaussian Ornstein--Uhlenbeck‐based models and some of their uses in financial economics.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2002).
+Econometric analysis of realized volatility and its use in estimating stochastic volatility models.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 64(2), 253-280.\
+
+Bollerslev, T. (1986). Generalized autoregressive conditional heteroskedasticity. Journal of econometrics, 31(3), 307-327.\
+
+Breidt, F. J., Crato, N., & De Lima, P. (1998). The detection and estimation of long memory in stochastic volatility. Journal of econometrics, 83(1-2), 325-348.\
+
+Buncic, D., & Gisler, K. I. (2016). Global equity market volatility spillovers: A broader role for the United States. International Journal of Forecasting, 32(4), 1317-1339.\
+
+Chan, J. C. (2013). Moving average stochastic volatility models with application to inflation forecast. Journal of Econometrics, 176(2), 162-172.\
+
+Chib, S., Nardari, F., & Shephard, N. (2002). Markov chain Monte Carlo methods for stochastic volatility models. Journal of Econometrics, 108(2), 281-316.\
+
+Chib, S., Nardari, F., & Shephard, N. (2006). Analysis of high dimensional multivariate stochastic volatility models. Journal of Econometrics, 134(2), 341-371.\
+
+Chopin, N., & Papaspiliopoulos, O. (2020). An introduction to sequential Monte Carlo (Vol. 4). Springer.\
+
+Comte, F., & Renault, E. (1998). Long memory in continuous‐time stochastic volatility models. Mathematical finance, 8(4), 291-323.\
+
+Christoffersen, P., Jacobs, K., & Mimouni, K.
+(2010).
+Volatility dynamics for the S&P500: Evidence from realized volatility, daily returns, and option prices.
+The Review of Financial Studies, 23(8), 3141-3189.\
+
+Dimitrakopoulos, S., & Kolossiatis, M.
+(2020).
+Bayesian analysis of moving average stochastic volatility models: modeling in-mean effects and leverage for financial time series.
+Econometric Reviews, 39(4), 319-343.\
+
+Douc, R., Moulines, E., & Stoffer, D. (2014). Nonlinear time series: Theory, methods and applications with R examples. CRC press.\
+
+Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with estimates of the variance of United Kingdom inflation. Econometrica: Journal of the econometric society, 987-1007.\
+
+Eraker, B. (2004). Do stock prices and volatility jump? Reconciling evidence from spot and option prices. The Journal of Finance, 59(3), 1367-1403.\
+
+Friz, P. K., Gassiat, P., & Pigato, P. (2021). Precise asymptotics: robust stochastic volatility models. The Annals of Applied Probability, 31(2), 896-940.\
+
+Gatheral, J., Jaisson, T., & Rosenbaum, M. (2018). Volatility is rough. Quantitative finance, 18(6), 933-949.\
+
+Guidolin, M., & Pedio, M. (2021). Media Attention vs. Sentiment as Drivers of Conditional Volatility Predictions: An Application to Brexit. Finance Research Letters, 101943.\
+
+Hansen, P. R., & Lunde, A. (2006). Consistent ranking of volatility models. Journal of Econometrics, 131(1-2), 97-121.\
+
+Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\
+
+Harvey, A., Ruiz, E., & Shephard, N. (1994). Multivariate stochastic variance models. The Review of Economic Studies, 61(2), 247-264.\
+
+Heber, Gerd, Asger Lunde, Neil Shephard and Kevin Sheppard (2009).
+Oxford-Man Institute's realized library.
+Oxford-Man Institute, University of Oxford, version 0.3.
+Available at: https://realized.oxford-man.ox.ac.uk/.
+\
+
+Heston, S. L. (1993). A closed-form solution for options with stochastic volatility with applications to bond and currency options. The review of financial studies, 6(2), 327-343.\
+
+Hull, J., & White, A. (1987). The pricing of options on assets with stochastic volatilities. The journal of finance, 42(2), 281-300.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. (1994). Bayesian analysis of stochastic volatility models. Journal of Business and Economic Statistics, 12(4), 371-389.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. E. (2004). Bayesian analysis of stochastic volatility models with fat-tails and correlated errors. Journal of Econometrics, 122(1), 185-212.\
+
+Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
+
+Kambouroudis, D.
+S., McMillan, D.
+G., & Tsakou, K.
+(2016).
+Forecasting stock return volatility: a comparison of GARCH, Implied volatility, and realized volatility models.
+Journal of Futures Markets, 36(12), 1127-1163.\
+
+Kastner, G. (2019). Dealing with stochastic volatility in time series using the R package stochvol. arXiv preprint arXiv:1906.12134.\
+
+Kim, S., Shephard, N., & Chib, S. (1998). Stochastic volatility: likelihood inference and comparison with ARCH models. The review of economic studies, 65(3), 361-393.\
+
+Kim, J., & Stoffer, D.
+S.
+(2008).
+Fitting stochastic volatility models in the presence of irregular sampling via particle methods and the EM algorithm.
+Journal of time series analysis, 29(5), 811-833.\
+
+Koopman, S. J., & Hol Uspensky, E. (2002). The stochastic volatility in mean model: empirical evidence from international stock markets. Journal of applied Econometrics, 17(6), 667-689.\
+
+Liesenfeld, R., & Jung, R. C. (2000). Stochastic volatility models: conditional normality versus heavy‐tailed distributions. Journal of applied Econometrics, 15(2), 137-160.\
+
+Luo, R., Zhang, W., Xu, X., & Wang, J. (2018, April). A neural stochastic volatility model. In Thirty-second AAAI conference on artificial intelligence.\
+
+Mao, X., Czellar, V., Ruiz, E., & Veiga, H.
+(2020).
+Asymmetric stochastic volatility models: Properties and particle filter-based simulated maximum likelihood estimation.
+Econometrics and Statistics, 13, 84-105.\
+
+Nakajima, J., & Omori, Y.
+(2012).
+Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student's t-distribution.
+Computational Statistics & Data Analysis, 56(11), 3690-3704.\
+
+Omori, Y., Chib, S., Shephard, N., & Nakajima, J. (2007). Stochastic volatility with leverage: Fast and efficient likelihood inference. Journal of Econometrics, 140(2), 425-449.\
+
+Pitt, M. K., & Shephard, N. (1999). Filtering via simulation: Auxiliary particle filters. Journal of the American statistical association, 94(446), 590-599.\
+
+Poon, S. H., & Granger, C. (2005). Practical issues in forecasting volatility. Financial analysts journal, 61(1), 45-56.\
+
+Shephard, N., & Sheppard, K. (2010). Realising the future: forecasting with high‐frequency‐based volatility (HEAVY) models. Journal of Applied Econometrics, 25(2), 197-231.\
+
+So, M. E. P., Lam, K., & Li, W. K. (1998). A stochastic volatility model with Markov switching. Journal of Business & Economic Statistics, 16(2), 244-253.\
+
+Xu, X., & Chen, Y. (2021). Deep Stochastic Volatility Model. arXiv preprint arXiv:2102.12658.\
+
+Zhang, B., Chan, J. C., & Cross, J. L. (2020). Stochastic volatility models with ARMA innovations: An application to G7 inflation forecasts. International Journal of Forecasting, 36(4), 1318-1328.

### 2021-07-21 prova_knit_finale.Rmd
commit

@@
 Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
-Inoltre, va chiarito il punto della "doppia convergenza" per T e per N (ossia, se il nostro T sia sufficientemente alto da parlare di convergenza quando N è molto grande).} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
+Inoltre, va chiarito il punto della "doppia convergenza".} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
 With such a benchmark, we can then compare how different filters approach it as the number of particles increases.

### 2021-07-21 prova_knit_finale.Rmd
ho levato le XX nella prima parte

--- 2021-07-21 prova_knit_finale.Rmd
@@
 
-\subsection{Implementation}
-
-In this section we present a practical illustration of the algorithms discussed.
-For the sake of simplicity we use a random walk plus noise model, i.e.
-the most basic form of a linear Gaussian state-space model.
+\subsection{Implementation}\label{implementation}
+
+The aim of this subsection is to provide a practical illustration of the algorithms discussed in the main chapters.
+For the sake of simplicity we use for implementations a random walk plus noise model, i.e.
+the most basic form of linear Gaussian state-space model.
 \begin{align}
-y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
-x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \label{eq:1} \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \label{eq:2} \\
 x_{0} & \sim N(m_{0},C_{0})
@@
 The filtering strategy is applied to 50 simulated data.
-Figure XX shows the simulated true states sequence assuming as data generating process the Equation (2) with $\tau^{2}=1$ and simulated observed sequence process form Equation (1) with $\sigma^{2}=1$.
-```{r echo=FALSE, fig.align='center', fig.cap="Simulated random walk plus noise model", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+Figure \ref{fig:myfig} shows the simulated true states sequence assuming as data generating process the Equation \eqref{eq:2} with $\tau^{2}=1$ and simulated observed sequence process form Equation \eqref{eq:1} with $\sigma^{2}=1$.
+```{r myfig, echo=FALSE, fig.align='center', fig.cap=" Simulated random walk plus noise model", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 
@@
 
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.\
-\code{sig2} \ \ the variance $\sigma^{2}$ in Equation (1)
-\
-\code{tau2} \ \ the variance $\tau^{2}$ in Equation (2)
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{sig2} \ \ the variance $\sigma^{2}$ in observation equation
+\
+\code{tau2} \ \ the variance $\tau^{2}$ in state equation
 \
@@
 ```
-In Figure XX below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
+In Figure \ref{fig:myfig2} below filtered states estimated using Kalman Filter with $x_{0} \sim N(0,100)$ and $\sigma^{2}=\tau^{2}=1$ are compared to the true states values.
 Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
 \[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
-```{r echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r myfig2, echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 #Kalman Filter
@@
 ```
-The discussion of Kalman Filter will continue in Section XX where we compare it to the Particle Filter.
+The discussion of Kalman Filter will continue in the next sections where we compare it to the Particle Filter.
 
@@
 \subsection{Implementation}
-Consider again the random walk plus noise model of section XX,the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
+Consider again the random walk plus noise model of section \ref{implementation},the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
 The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm.
@@
 \
-\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
-\
-\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\code{tau} \ \ the standard deviation $\tau$ in state equation
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in observation equation
 \hrule
@@
 \subsection{Implementation}
-In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section XX.
+In this section, Bootstrap Particle Filter will be used to estimate filtered states of the Random Walk plus Noise introduced in section \ref{implementation}.
 The overall strategy replicate the SIS filter with the addition of a ESS-based resampling step that applies when the effective sample size is smaller than a predetermined threshold opportunely chosen (in our example we decide to follow a common rule of thumb consisting in setting the threshold at $N/2$).
@@
 \
-\code{tau} \ \ the standard deviation $\tau$ in Equation (2)
-\
-\code{sigma} \ \ the standard deviation $\sigma$ in Equation (1)
+\code{tau} \ \ the standard deviation $\tau$ in state equation
+\
+\code{sigma} \ \ the standard deviation $\sigma$ in observation equation
 \
@@
 ```
-The estimated states of the Boostrap Particle Filter togheter with the effective sample size are shown in Figure XX.
+The estimated states of the Boostrap Particle Filter togheter with the effective sample size are shown in Figure \ref{fig:myfig3}.
 We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$.
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="a) PF Filtered States with credible interval (in red).
+```{r myfig3, echo=FALSE, fig.align='center', fig.cap="a) PF Filtered States with credible interval (in red).
 b) Effective sample size (in black) with threshold (in yellow).", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
@@
 Since the Random Walk plus Noise model allows for closed form solutions, we want to compare the results of Boostrap Particle Filter(BPF) with Kalman Filter(KF).
-As we can see from figure XX, when the number of particles generated at any interaction increases the results for both the estimated mean and the variance tend to converge [[SPIEGARE PERCHE']].
+As we can see from figure \ref{fig:myfig4}, when the number of particles generated at any interaction increases the results for both the estimated mean and the variance tend to converge [[SPIEGARE PERCHE']].
 Moreover, comparing the Root Mean Square Errors of the filtered states with respect to the true state values, when the sample size increases the BPF decreases and, for large N, it reaches the accuracy of the Kalman Filter.
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='H', fig.width=10, message=FALSE, warning=FALSE, results=F}
+```{r myfig4, echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)", fig.height=13, fig.pos='H', fig.width=10, message=FALSE, warning=FALSE, results=F}
 set.seed(1991)
@@
 \
-The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section XX can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$.
+The Bootstrap Particle Filter Approach for the Random Walk plus Noise Model described in Section \ref{implementation} can be improved accounting for the observations in the importance transition density and it consists of generating $x_{t}$ from its conditional distribution given $x_{t-1}$ and $y_{t}$.
 In the Normal model we are considering, the optimal proposal will be a Normal density as well with mean and variance given by
@@
 Let's provide directly a brief comparison between the Bootstrap Particle Filter (BPF) and the Guided Particle Filter (GPF).
-As we can see from the Figure XX, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE.
+As we can see from the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE.
 Moreover, also the variance is better, suggesting for higher precision.
@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r myfig5, echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 set.seed(1991)
@@
 \
-Consider the linear Gaussian example of section XX, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
+Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
 Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
@@
 \section{Convergence}\label{pf_converg}
+
+\chapter{Implementations for Stochastic Volatility Models}
+
+\section{}
 

### 2021-07-22 final draft.Rmd
aggiunta a notazione
piccolo fix

@@
 The probability distribution of a random variable $X_t$ is represented by $\mathbb{P}_t(dx_t),$ while the probability distribution of a sequence $X_{0:t}$ is represented by $\mathbb{P}_t(dx_{0:t}).$ The Dirac measure, i.e., the measure that assigns probability $1$ to the singleton $\{x\},$ is denoted by $\delta_x(dy)$.The expectation operator is denoted by $\mathbb{E},$ where the section of $\mathbb{E}$ at $\mathbb{Q},$ i.e., $\mathbb{E}_{\mathbb{Q}}$ denotes the operator computed with respect to the probability distribution $\mathbb{Q}.$
-Let $\varphi$ be a function defined on the random variable $X\propto\mathbb{Q}.$ Then, we denote the expected value of $\varphi(X)$ with 
+Let $\varphi$ be a function defined on the random variable $X\thicksim\mathbb{Q}.$ Then, we denote the expected value of $\varphi(X)$ with 
 $$

### 2021-07-22 prova_knit_finale.Rmd

@@
 - \urlstyle{same}
+- \usepackage[toc,page]{appendix}
 - \usepackage{comment}
@@
   for(t in 1:length(data)){
-    x    = rnorm(N,x,tau)                   #sample from N(x_{t-1},tau)
-    w    = w*dnorm(data[t],x,sigma)         #update weight
+    x    = rnorm(N,x,tau)                   
+    w    = w*dnorm(data[t],x,sigma)         
     xs = rbind(xs,x)
@@
 
-    wnorm= w/sum(w)                         #normalized weight
-    ESS  = 1/sum(wnorm^2)                   #effective sample size
+    wnorm= w/sum(w)                         
+    ESS  = 1/sum(wnorm^2)
 
@@
 Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap Particle Filter (BPF).
-
-```{r echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+From Figure \ref{fig:myfig6} it is difficult to say which of the two is better and the tables does not help much since the APF apparently outperforms the BPF in two cases over three.
+For this reason we can conclude that for this model the Auxiliary Particle Filtering strategy does not provide any serious improvement to a the basic one.
+
+```{r myfig6, echo=FALSE, fig.align='center', fig.cap="Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
 set.seed(1991)
@@
 
-\chapter{Implementations for Stochastic Volatility Models}
-
-\section{}
+\section{Implementations for Stochastic Volatility Models}
 
@@
 Zhang, B., Chan, J. C., & Cross, J. L. (2020). Stochastic volatility models with ARMA innovations: An application to G7 inflation forecasts. International Journal of Forecasting, 36(4), 1318-1328.
+
+\chapter{Appendix}
+
+\section{The Stochastic Volatility Model}
+
+We decide to organise this Appendix as an R package. All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model.
+
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
+
+All the plots and tables of this work concerning stochastic volatility models use this functions.
+
+\hrule
+\hrule
+\
+\code{SVPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+\
+```{r message=FALSE, warning=FALSE, paged.print=FALSE}
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+\
+\hrule
+\hrule
+\
+\code{SVGPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVGPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+\
+\hrule
+\hrule
+\
+\code{SVAPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],0,exp(x/2))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k],tau)
+    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\hrule
+\hrule
+\
+\code{SVAPFoptfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+    x_prev=x
+    x_hat = alpha+beta*x #means of the transition from the previous particle 
+    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
+    for(k in 1:N){
+      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
+      w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
+    }
+    w = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\hrule
+\hrule
+\
+\code{SVLWfun(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+  xs = rnorm(N,m0,sqrt(C0))
+  pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
+                 log(1/rgamma(N,nu/2,nu*lambda/2)))
+  delta  = 0.75
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,3,n))
+  xss    = NULL
+  ws     = NULL
+  ESS    = NULL
+  w      = rep(1/N,N)
+  vpars   = NULL
+  for (t in 1:length(data)){
+
+    #mpar        = apply(pars,2,mean) #alternatively use this 
+    mpar<-c()
+    for(i in 1:3){
+      mpar[i]    = weighted.mean(pars[,i],w)
+    }
+
+    vpar     = var(pars)
+    ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
+    mus         = pars[,1]+pars[,2]*xs 
+    weight      = w*dnorm(data[t],0,exp(mus/2))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+    ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
+    xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
+    w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/2)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xt[index]
+      pars<-ms1[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xt
+      pars<-ms1
+    }
+
+    vpars        = rbind(vpars,vpar)
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+  }
+  return(list(xs=xss,pars=parss,ws=ws,vpars=vpars))
+}
+```

### 2021-07-22 prova_knit_finale.Rmd
auxiliry fixed

@@
 \begin{itemize}
-\item Initialize $\theta_{0} \sim N(m_{0},C_{0})$
+\item Initialize $x_{0} \sim N(m_{0},C_{0})$
 \item For $t=1,...,n$:
@@
 Notice how closely the filtered states follow the observations and the goodness of the approximation of the true states. 95 percent credible intervals are computed as
-\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
+\[[E(x_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})},E(x_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})}]\]
 ```{r myfig2, echo=FALSE, fig.align='center', fig.cap="Kalman Filtered States with credible interval (in red)", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
@@
 As we can see from the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point estimates that are slightly better with respect to the ones of the BPF, and this is confirmed by the Table showing the RMSE.
-Moreover, also the variance is better, suggesting for higher precision.
+Moreover, the variance is lower, suggesting for higher precision.
 
@@
     k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
+    x1   = rnorm(N,means,importancesd)
+    w   = rep(1/N,N)
     }else{
     weight = rep(1/N,N)
-    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-    }
-
+    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
     means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
     x1   = rnorm(N,means,importancesd)
-    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
-    w   = exp(lw)
+    w   <-  w*dnorm(data[t],x1,predsd)
+    }
+
     w   = w/sum(w)
@@
 
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-```
-
-```{r}
-###################
-## Per roberto:::::: cambia molto se usiamo l'uno o l'altro!!!!!! perchè??
-APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
-  if(missing(r)){r=2}else{}
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
-  predsd <- sqrt(sigma+tau)
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    weight = w*dnorm(data[t],x,predsd)
-    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
-    x1   = rnorm(N,means,importancesd)
-    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)
-    w   = exp(lw)
-    w   = w/sum(w)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x1<-x1[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    x <- x1
     xs = rbind(xs,x)
@@
 \section{Implementations for Stochastic Volatility Models}
+
+So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution
 
@@
 We decide to organise this Appendix as an R package. All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model. 
-
 \begin{align*}
@@
 \end{align*}
-
 All the plots and tables of this work concerning stochastic volatility models use this functions.
@@
 ```
+
+```{r}
+###################
+## Per roberto:::::: cambia molto se usiamo l'uno o l'altro!!!!!! perchè??
+APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
+  predsd <- sqrt(sigma+tau)
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],x,predsd)
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
+    x1   = rnorm(N,means,importancesd)
+    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)+log(w)-log(weight)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x1<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```

### 2021-07-22 prova_knit_finale.Rmd
small change

@@
 In section \ref{pf_intro} we introduce the fundamental notions of \textit{importance sampling}, \textit{sequential importance sampling} and \textit{resampling}.
-Then, in section \ref{pf_main} we describe three particle filtering algorithms: the \textit{bootstrap}, the \textit{guided} and the {auxiliary particle filters}.
-Instead, in section \ref{pf_lui_west} we describe the {Liu and West filter}, when particle filtering is used to estimate model's parameters.
+Then, in section \ref{pf_main} we describe three particle filtering algorithms: the \textit{bootstrap}, the \textit{guided} and the \textit{auxiliary particle filters}.
+Instead, in section \ref{pf_lui_west} we describe the \textit{Liu and West filter}, when particle filtering is used to estimate model's parameters.
 Finally, in section \ref{pf_converg} we establish some general convergence results for Particle Filter - Monte Carlo estimators.

### 2021-07-22 prova_knit_finale.Rmd
notazione

--- 2021-07-22 prova_knit_finale.Rmd
@@
 \begin{equation}
-\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
+\P_t(X_{0:t}\in dx_{0:t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
 \end{equation}
@@
 \begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
   Then, it holds that
   \begin{align}
-    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
-    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
-    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+    X\mid Y=y&\thicksim\N(m_1,C_1),\\
+    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ R)^{-1} B C_0) \\
+    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
   \end{align}
@@
 \begin{proof*}
-   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, Q_0).$ It is easy to see that
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, C_0).$ It is easy to see that
 \begin{equation} \label{normal-distrib-rewrite}
-        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'C_0^{-1}X + X'C_0^{-1}m_0\Big).
 \end{equation}
-with $A=Q_0^{-1}$ and $B=Am_0.$
 Now, by Bayes' Theorem,
@@
         p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
-                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
-                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
-                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
-                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
+                    &\propto \text{exp}\Big(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
+                    &\propto \text{exp}(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}X'(C_0^{-1}+B' R^{-1}B)X+
+                    X'(C_0^{-1}m_0+B' R^{-1}y)\Big).
     \end{align*}
@@
 \begin{equation*}
-    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
+    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' R^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+R)^{-1}B C_0
 \end{equation*}
@@
     \begin{align*}
-        m_1 &=\E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
-            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+        m_1 &=\E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' R^{-1}y)\\
+            &= (C_0-C_0B'(B C_0B'+R)^{-1}B C_0)(C_0^{-1}m_0+B' R^{-1}y)\\
+            &= m_0-C_0B'(B C_0B'+R)^{-1}B m_0 + C_0 B' R^{-1}y - C_0B'(B C_0B'+R)^{-1}B C_0 B' R^{-1}y \\
+            &=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
     \end{align*}
@@
 
-Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
+Assume that $\P_{t-1}(dx_{t-1}\mid Y_{1:t-1}=y_{1:t-1})=\N(m_{t-1},C_{t-1}).$
 Then, using the transition equation of the latent variable, 
@@
 \begin{equation}
-\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
+\P_{t-1}(dx_{t}\mid Y_{1:t-1}=y_{1:t-1})=\N(A_tm_{t-1},A_tC_{t-1}A_t'+\Sigma_t).
 \end{equation}
-Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
+Letting $E_T=A_tC_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
 \begin{align*}
-  \P_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{where} \\ C_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
   m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
 \end{align*}
+
+In some situations, it is possible to linearize otherwise non-linear models in order to employ this filtering method.
+There are, however, cases where this is not possible or where the linearization is not satisfactory.
+One such example is the stochastic volatility model.
+In the following chapters, we will introduce powerful algorithms that can be used in much more general settings.
 

### 2021-07-22 prova_knit_finale.Rmd
aggiunto algoritmo
correzioni

@@
 Integrating away $X_{0:T},$ one gets the marginal density
-\begin{align*}\label{state-space-marginal-y}
-  \P_T(dy_{0:T}) &= \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
-                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
-\end{align*}                
+\begin{equation}\label{state-space-marginal-y}
+  \P_T(dy_{0:T}) = \E_{\P_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s)
+                = p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
+\end{equation}                
 where $p_t(y_{0:t})$ is the likelihood function.
-Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the filtering distribution
+Dividing the left-hand side of equation \ref{state-space-densities} by the left-hand side of equation \ref{state-space-marginal-y} and the right-hand side of equation \ref{state-space-densities} by the right-hand side of equation \ref{state-space-marginal-y}, one obtains the distribution
 \begin{equation}
@@
 
-Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,R_t)$ be two independent white-noise errors and assume the initial distribution 
+Suppose that $\X=\R^{d_x}$ and $\Y=\R^{d_y}.$ Let $U_t\thicksim\N(0,\Sigma_t)$ and $V_t\thicksim\N(0,\Lambda_t)$ be two independent white-noise errors and assume the initial distribution 
 $X_0\thicksim\N(0,\Sigma_0).$ We have a linear Gaussian model if 
@@
 \begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,\Lambda)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
   Then, it holds that
@@
     X\mid Y=y&\thicksim\N(m_1,C_1),\\
-    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ R)^{-1} B C_0) \\
-    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
+    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ \Lambda)^{-1} B C_0) \\
+    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
   \end{align}
@@
 \begin{equation} \label{normal-distrib-rewrite}
-        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'C_0^{-1}X + X'C_0^{-1}m_0\Big).
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}x'C_0^{-1}x + x'C_0^{-1}m_0\Big).
 \end{equation}
@@
         p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
-                    &\propto \text{exp}\Big(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
-                    &\propto \text{exp}(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
-                    &=\text{exp}(-\frac{1}{2}X'(C_0^{-1}+B' R^{-1}B)X+
-                    X'(C_0^{-1}m_0+B' R^{-1}y)\Big).
+                    &\propto \text{exp}\Big(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}(y-Bx)'\Lambda^{-1}(y-Bx)) \\
+                    &\propto \text{exp}(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}x' B' \Lambda^{-1}Bx+x' B' \Lambda^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}x'(C_0^{-1}+B' \Lambda^{-1}B)x+
+                    x'(C_0^{-1}m_0+B' \Lambda^{-1}y)\Big).
     \end{align*}
@@
 \begin{equation*}
-    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' R^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+R)^{-1}B C_0
+    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' \Lambda^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+\Lambda)^{-1}B C_0
 \end{equation*}
 and 
-    \begin{align*}
-        m_1 &=\E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' R^{-1}y)\\
-            &= (C_0-C_0B'(B C_0B'+R)^{-1}B C_0)(C_0^{-1}m_0+B' R^{-1}y)\\
-            &= m_0-C_0B'(B C_0B'+R)^{-1}B m_0 + C_0 B' R^{-1}y - C_0B'(B C_0B'+R)^{-1}B C_0 B' R^{-1}y \\
-            &=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
-    \end{align*}
+\begin{equation*}
+\begin{split}
+        m_1 &=\E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
+            &= (C_0-C_0B'(B C_0B'+\Lambda)^{-1}B C_0)(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
+            &= m_0-C_0B'(B C_0B'+\Lambda)^{-1}B m_0 + C_0 B' \Lambda^{-1}y - C_0B'(B C_0B'+\Lambda)^{-1}B C_0 B' \Lambda^{-1}y \\
+            &=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
+\end{split}
+\end{equation*}$\hfill\blacksquare$
 \end{proof*}
@@
 \end{equation}
-Letting $E_T=A_tC_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
+Letting $R_T=A_tC_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
 \begin{align*}
-  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{where} \\ C_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
-  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
+  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{ where} \\ C_t&=R_t[I_{d_x} - B_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_tR_t] \\
+  m_t &= [I_{d_x} - R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_t]A_tm_{t-1}+R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1}y_t.
 \end{align*}

### 2021-07-24 prova_knit_finale.Rmd
qualche modifica

--- 2021-07-22 prova_knit_finale.Rmd
@@
 
-\section{State-Space Models}
+\section{State-Space Models}\label{sec_state_space}
 We now define \textit{State-Space models} as Markov processes that are partially observed.
@@
 \begin{equation}
-\P_t(X_{0:t}\in dx_{0:t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
+\P_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\P_t(dx_{0:t}).
 \end{equation}
@@
 \begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Let $X\thicksim\N(m_0,Q_0)$ and $Y\mid X\thicksim\N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
   Then, it holds that
   \begin{align}
-    X\mid Y=y&\thicksim\N(m_1,C_1),\\
-    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ R)^{-1} B C_0) \\
-    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
+    X\mid Y=y&\thicksim\N(m_1,Q_1),\\
+    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
+    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
   \end{align}
@@
 \begin{proof*}
-   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, C_0).$ It is easy to see that
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\N(m_0, Q_0).$ It is easy to see that
 \begin{equation} \label{normal-distrib-rewrite}
-        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'C_0^{-1}X + X'C_0^{-1}m_0\Big).
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
 \end{equation}
+with $A=Q_0^{-1}$ and $B=Am_0.$
 Now, by Bayes' Theorem,
@@
         p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
-                    &\propto \text{exp}\Big(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
-                    &\propto \text{exp}(-\frac{1}{2}X' C_0^{-1}X + X' C_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
-                    &=\text{exp}(-\frac{1}{2}X'(C_0^{-1}+B' R^{-1}B)X+
-                    X'(C_0^{-1}m_0+B' R^{-1}y)\Big).
+                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
+                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
+                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
     \end{align*}
@@
 \begin{equation*}
-    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' R^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+R)^{-1}B C_0
+    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
 \end{equation*}
@@
     \begin{align*}
-        m_1 &=\E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' R^{-1}y)\\
-            &= (C_0-C_0B'(B C_0B'+R)^{-1}B C_0)(C_0^{-1}m_0+B' R^{-1}y)\\
-            &= m_0-C_0B'(B C_0B'+R)^{-1}B m_0 + C_0 B' R^{-1}y - C_0B'(B C_0B'+R)^{-1}B C_0 B' R^{-1}y \\
-            &=[I_{d_x} - C_0 B'(B C_0 B'+ R)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ R)^{-1}y.
+        m_1 &=\E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
+            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
+            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
     \end{align*}
@@
 
-Assume that $\P_{t-1}(dx_{t-1}\mid Y_{1:t-1}=y_{1:t-1})=\N(m_{t-1},C_{t-1}).$
+Assume that $\P_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\N(m_{t-1},Q_{t-1}).$
 Then, using the transition equation of the latent variable, 
@@
 \begin{equation}
-\P_{t-1}(dx_{t}\mid Y_{1:t-1}=y_{1:t-1})=\N(A_tm_{t-1},A_tC_{t-1}A_t'+\Sigma_t).
+\P_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
 \end{equation}
-Letting $E_T=A_tC_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
+Letting $E_T=A_tQ_{t-1}A_t'+\Sigma_t,$ and applying lemma \ref{kalman-filter}, it follows that the filtering distribution is 
 \begin{align*}
-  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{where} \\ C_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
+  \P_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
   m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
 \end{align*}
-
-In some situations, it is possible to linearize otherwise non-linear models in order to employ this filtering method.
-There are, however, cases where this is not possible or where the linearization is not satisfactory.
-One such example is the stochastic volatility model.
-In the following chapters, we will introduce powerful algorithms that can be used in much more general settings.
 
@@
 The following result represents the starting point for the idea of importance sampling.
-
+\bigskip
 \begin{lemma}\label{lemma_imp_sam_1}
@@
 Differently from SIS, an intermediate resampling stage takes place, in which time 0 values are sampled with replacement according to a given resampling strategy (usually, sampling indexes from a multinomial distribution with probabilities equal to the weights computed through IS at time 0).\
-While in the SIS all the particles (i.e., the sampled values) ``survive", with the ``risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR.
+While in the SIS all the particles (i.e., the sampled values) "survive", with the "risk" that particles associated with small weights are carried-over (in the sense that sampling in period 1 occurs from the kernel $M_1(X_0^n,dx_1)$), the same does not occur in SIR.
 In this second algorithm, particles with larger weights are more likely to be resampled and be used to ``generate" the particles in the next period.\
@@
 \begin{equation}
-    w^n_1=\frac{\Q_1(\tilde dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
+    w^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
 =\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
@@
 Nonetheless, SIS suffers from the problem knonw as \textit{curse of dimensionality}, i.e., the fact that the $ESS$ collapses over time as the variance of the weights diverges.\
-Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation} admit a density too).
+Consider an extension of SIS to a $T$-period model and let us assume that the target and the proposal kernels admit densities, respectively, $q_t$ and $m_t$ with respect to the same dominating measures (this implies that the joint distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation.} admit a density too).
 Then, weights, for every $t=1,...,T$, are updated according to:
@@
 
-\color{red}
-Let us consider the standard state-space model introduced in the previous chapters.
-\color{black}
+Let us consider the standard state-space model introduced previously (see \ref{sec_state_space}).
 \textit{Particle filter} (PF) is a numerical algorithm that is used to obtain approximations for the filtering - $\Pd_t (dx_t|y_{1:t})$ - and the prediction - $\Pd_t (dx_t|y_{1:t-1})$ - distributions for inference in state-space models.\ \
@@
 \begin{itemize}
-\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
 In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
@@
 We provide a graphical intuition of what happens when we use such filtering strategy on a simulated dataset.
-We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$.
-The results shown in the following two plots shows a clear degeneration of the effective sample size and bad fit of filtered states with respect to the true values.
+We decide to set $N=1000$, $m_{0}=0$, $C_{0}=100$ and $\tau=\sigma=1$.
+The following two plots show a clear degeneration of the effective sample size and bad fit of filtered states with respect to the true values.
 ```{r echo=FALSE}
@@
 The estimated states of the Boostrap Particle Filter togheter with the effective sample size are shown in Figure \ref{fig:myfig3}.
-We decide to set $N=1000$,$m_{0}=0$,$C_{0}=100$ and $\tau=\sigma=1$.
+We decide to set $N=1000$, $m_{0}=0$, $C_{0}=100$ and $\tau=\sigma=1$.
 Notice how the resampling step allows the effective sample size not to drop, improving results.
@@
 Moreover, even if it were possible, there exist a proposal transition kernel that makes the algorithm more efficient.
-This motivates the \textit{guided particle filter} (GPF), that can be interpreted as a more general version of the GPF.
+This motivates the \textit{guided particle filter} (GPF), that can be interpreted as a more general version of the BPF.
 In fact, with the GPF it is possible, at each stage, to sample new states from a transition kernel that differs from the correct one.\ \
@@
 \begin{itemize}
-\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
 In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.

### 2021-07-26 prova_knit_finale.Rmd
Aggiunte funzioni

--- 2021-07-22 prova_knit_finale.Rmd
@@
 \begin{enumerate}
-\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Draw $x_{t}^{(i)} \sim N(x_{t-1}^{(i)},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
 \item if $ESS<N/2$ then
@@
 \item Compute $\mu_{opt}$
-\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{t})$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma_{opt}^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};x_{t}^{(i)},\sigma^2+\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
 \item if $ESS<N/2$ then
@@
 \begin{enumerate}
-\item Draw $I_{k}$ with $P(I_{k}) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f(y_{t}|g(x_{t-1}^{(i)})) $
 \item Draw $x_{t}^{(k)} \sim N(x_{t-1}^{(I_{k})},\tau^2)$
@@
 We also implement the auxiliary particle filter with optimal transition kernel. Therfore the algorithm is similiar to the one discussed above with the difference that:
-\
-[[[[Inserire l'algoritmo]]]]
+
+\begin{algorithm}APF with optimal transition kernel for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\sigma^{2}+\tau^{2}) $
+\item Compute $\mu_{opt}=\mu(x_{t-1}^{(I_{k})})$
+\item Draw $x_{t}^{(k)} \sim N(\mu_{opt},\sigma^{2}_{opt})$
+\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
+\end{enumerate}
+\item otherwise
+\begin{enumerate}
+\item Compute $\mu_{opt}=\mu(x_{t-1})$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma^{2}_{opt})$ for $i=1,...N$
+\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};x_{t}^{(i)},\sigma^{2}+\tau^{2})$ for $i=1,...,N$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
 
@@
 ```
+\
+\hrule
+\hrule
+\
+\code{SISfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+\
+```{r message=FALSE, warning=FALSE, paged.print=FALSE}
+SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<0){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+\
+\hrule
+\hrule
+\
+\code{BAPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+\
+```{r message=FALSE, warning=FALSE, paged.print=FALSE}
+BAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS>-1){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```

### 2021-07-27 prova_knit_finale.Rmd
small change

--- 2021-07-24 prova_knit_finale.Rmd
@@
     \item it is not clear that what is optimal at time $t$ remains such at future stages.
-In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affect the efficiency of the PF estimators;
+In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affects the efficiency of the PF estimators;
     \item both the choice of $M_t$ and of $\eta_t$ affect the way in which weights are computed and, indirectly, the decision to resample in the future.

### 2021-07-27 prova_knit_finale.Rmd
small changes

@@
 Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter.
-Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\
+Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.
+For instance, in our implementation with the random walk model of section \ref{implementation}, we use an auxiliary particle LWF.}\
 Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
@@
 $W_{t-1}^n$).\ \
-The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$.
+The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal\footnote{More generally, distributions whose support coincides with the relevant parameter space.} centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$.
 While it would be more intuitively immediate to sample new parameters from a Normal density centered at the previously sampled values (i.e., $\psi^n$), this would increase the variance of $\psi$ over time.\
@@
 \end{align*}
-Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. The algorithm follows the following steps.
+Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters.
+Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values.
+The algorithm follows the following steps.
 \begin{algorithm} LWF for Random Walk plus Noise Model

### 2021-07-27 prova_knit_finale.Rmd
aggiunta a header

@@
 - \usepackage{xcolor}
+- \usepackage{mathtools}
 - \theoremstyle{break}
@@
 
-\subsection{Sequential Importance Sampling}
+\subsection{Sequential Importance Sampling}\label{seq_impt_sampling}
 

### 2021-07-29 prova_knit_finale.Rmd
corretto errore tabella
added cite

--- 2021-07-27 prova_knit_finale.Rmd
@@
 Otherwise, the \textit{curse of dimensionality} problem will emerge and the variance of the weights will diverge over time as the variance of past weights increases as new updates occur period-by-period.
-Figure ***** shows an example of a PF where SIS is used: note that the $ESS$ drops dramatically after few periods, meaning that the variance of the importance weights diverges.
+Next, we show an example of a PF where SIS is used: note that the $ESS$ drops dramatically after few periods, meaning that the variance of the importance weights diverges.
 This motivates the use of a SIR structure for the PF algorithm.\ \
@@
 Let us denote by  $G_t=[P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)dy_t]/M_t(x_{t-1},dx_t)$ the \textit{incremental weights} (i.e., the updating factor) and by $M_t^*(x_{t-1},dx_t)$ the optimal kernel.
-It can be proved\footnote{For a formal proof see \cite{**}, th.
+It can be proved\footnote{For a formal proof see Chopin and Papaspiliopoulos (2020), th.
 10.1 p.
@@
 \end{equation}
-It is possible to derive an optimality result for the auxiliary functions\footnote{Details in ***** th 10.2, p.
+It is possible to derive an optimality result for the auxiliary functions\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p.
 148}.
@@
 However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance.
-Se ***Johansen p.
+See Johansen and Evers (2007), p.
 120}
@@
 
+Johannes, M., Evers, L. (2007). Monte Carlo Methods. Lecture Notes. University of Bristol, Department of Mathematics.\
+
 Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
@@
 Omori, Y., Chib, S., Shephard, N., & Nakajima, J. (2007). Stochastic volatility with leverage: Fast and efficient likelihood inference. Journal of Econometrics, 140(2), 425-449.\
+
+Petris, G., Petrone, S., Campagnali, P. (2010). Dynamic Linear Models with R. Springer.\
 

### 2021-07-29 prova_knit_finale.Rmd

--- 2021-07-27 prova_knit_finale.Rmd
@@
 rownames(SC)<-c("Mean","Standard Deviation")
-colnames(SC) <- c("Dow Jones", "S&P500", "STOXX50E")
+colnames(SC) <- c("S&P500", "Dow Jones", "STOXX50E")
 

### 2021-07-30 prova_knit_finale.Rmd
added intro

@@
 \chapter*{Introduction}
-
-\chapter{Markov Processes and State-Space Models}
+State-Space models represent an extremely interesting class of models and they are used extensively, for instance, for inference and forecasting in time-series and macroeconometric analysis.
+In fact, they allow to model observational data as function of underlying states that cannot be observed.
+For instance, in macroeconomic models GDP is usually observed while the TFP sequence is the hidden state determining output.
+In this work we describe techniques, such as Kalman and Particle filter, that allows to approximate probability distributions for the states from the observed data.
+For example, from the sequence of observed GDP, it is possible to "filter" the TFP factor.\
+Furthermore, it is possible to impose a Markov structure to State-Space models, assuming that the states are Markov, i.e., the probability of the current state depends on the entire sequence of past states only through its first lag.
+The combination of State-Space and Markov structures constitutes a powerful modeling assumption for inference and forecasting.\ \
+Withing the Monte Carlo (MC) methods, where inference is based on repeated simulations that allow to approximate the target distributions (needed to compute the estimator of interest), Particle Filter (PF) Algorithms represent the most important class of procedures for State-Space-Markov models.
+PF allows to sample states period-by-period, updating sequentially the filtering distributions (i.e., the distribution of current state conditional on all the sequence of observations - up to the current period one -), incorporating the information on the likelihood of the extracted states given the sequence of observations, by means of re-weighting.\
+Moreover, the PF is a very general algorithm, since it can be applied to non-linear models with non-Gaussian errors, while it does not allow for a closed form solution.
+On the contrary, the Kalman Filter, is a filtering method that can be applied in the case of linear-Gaussian  State-Space-Markov models, that remains analytically tractable.\
+In this work we will describe both these algorithms, while our main focus remains on Particle Filter.\ \
+In chapter \ref{MP_SSM_chapter} we define precisely Markov processes and State-Space Models.
+In chapter \ref{KF_chapter} we describe the Kalman Filter and in chapter \ref{PF_chapter} we analyze extensively the Particle Filter.
+Finally, in chapter \ref{application_chapter}, we apply these methods to stochastic volatility models.\
+Finally, we find it useful to integrate each section where an algorithm is described and analyzed from a theoretical perspective with an implementation subsection, using as an example a simple Random-Walk State-Space model.
+In this way, the reader can visualize the codes and the practical meaning of the Kalman and Particle Filter algorithms, immediately after their theoretical description.
+
+\chapter{Markov Processes and State-Space Models}\label{MP_SSM_chapter}
 
@@
 
-\chapter{Linear-Gaussian case}
+\chapter{Linear-Gaussian case}\label{KF_chapter}
 
@@
 
-\chapter{Particle Filter}
+\chapter{Particle Filter}\label{PF_chapter}
 
@@
 
-\chapter{Applications to Stochastic Volatility Models}
+\chapter{Applications to Stochastic Volatility Models}\label{application_chapter}
 

### 2021-07-30 backup1.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

--- 2021-07-19 new-draft.tex
@@
 \hypersetup{
-  pdftitle={Final draft},
+  pdftitle={State-Space Models and Particle Filters},
+  pdfauthor={Edoardo Marcelli; Roberto Colarieti; Alexei Verkhovtsev; Michele Bolla; Gabriele Romano; Benedetta Bruni},
   hidelinks,
@@
 
-\title{Final draft}
-\author{}
-\date{\vspace{-2.5em}}
+\title{\Huge State-Space Models and Particle Filters}
+\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei
+Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta
+Bruni}
+\date{21 luglio, 2021}
 
@@
 \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
-\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)
+\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)\\
 
@@
 \(t\leq T,\) is a sequence of probability measures. Then, the following
-proposition holds.
+proposition holds.\\
 
@@
 distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
-\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).
+\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).\\
 
@@
 \end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
-normally distributed.
+normally distributed.\\
 
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-5-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-5-1}
 
@@
 \hrule
-\texttt{DLM(data,sig2,tau2,m0,C0)}
+
+\hfill\break
+\texttt{DLM(data,sig2,tau2,m0,C0)}\\
+
 \hrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-7-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-7-1}
 
@@
 to sample from the target density or there exist another density - call
-it \(m(\cdot)\) - different from \(q\) that produces
-\texttt{more\ efficient"\ estimates.\ \ In\ all\ the\ cases\ for\ which\ sampling\ takes\ place\ from\ a\ distribution\ different\ from\ the}correct"
-(i.e., the true) one, importance sampling and resampling methods can be
-applied (as augmented versions of the standard MC algorithm).
+it \(m(\cdot)\) - different from \(q\) that produces ``more efficient''
+estimates. In all the cases for which sampling takes place from a
+distribution different from the ``correct'' (i.e., the true) one,
+importance sampling and resampling methods can be applied (as augmented
+versions of the standard MC algorithm).
 
@@
 variance when future states are sampled from the transition kernel.\\
-\texttt{Dead"\ particles\ (i.e.,\ those\ with\ small\ weights)\ are\ \ likely\ to\ be\ removed\ when\ resampling\ takes\ place\ and\ past\ weights\ get}reset"
+``Dead'' particles (i.e., those with small weights) are likely to be
+removed when resampling takes place and past weights get ``reset"
 (usually, they are set to be equal to \(1/N\)). Otherwise, the
@@
 \hrule
-\texttt{SISfun(data,N,m0,C0,tau,sigma)}
+
+\hfill\break
+\texttt{SISfun(data,N,m0,C0,tau,sigma)}\\
+
 \hrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-10-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-10-1}
 
@@
 \hrule
-\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}
+
+\hfill\break
+\texttt{PFfun(data,N,m0,C0,tau,sigma,r)}\\
+
 \hrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-13-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-13-1}
 
@@
 \midrule
-100 & 0.5 & 0.879 & 0.888\\
-1000 & 0.5 & 0.879 & 0.886\\
-10000 & 0.5 & 0.879 & 0.878\\
+100 & 0.5 & 0.879 & 0.916\\
+1000 & 0.5 & 0.879 & 0.882\\
+10000 & 0.5 & 0.879 & 0.885\\
 \bottomrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-16-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-16-1}
 
@@
 \hrule
-\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}
+
+\hfill\break
+\texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}\\
+
 \hrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-19-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-19-1}
 
@@
 \midrule
-1000 & 0.50 & 0.916 & 0.880\\
-1000 & 0.25 & 0.895 & 0.862\\
-1000 & 0.10 & 0.869 & 0.881\\
+1000 & 0.50 & 0.882 & 0.874\\
+1000 & 0.25 & 0.889 & 0.874\\
+1000 & 0.10 & 0.885 & 0.879\\
 \bottomrule
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-22-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-22-1}
 
@@
 
+\subsection{Implementation}
+
 For illustration purposes, we are going implement an auxiliary particle
-filter with auxiliary function
-\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).\\
+filter for the linear Gaussian model with auxiliary function
+\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).
 
@@
 \hfill\break
-\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}
+\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}\\
 
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-25-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-25-1}
 
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-26-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-26-1}
 
@@
 \midrule
-1000 & 0.50 & 0.885 & 0.875\\
-1000 & 0.25 & 0.893 & 0.892\\
-1000 & 0.10 & 0.855 & 0.870\\
+1000 & 0.50 & 0.882 & 0.899\\
+1000 & 0.25 & 0.889 & 0.883\\
+1000 & 0.10 & 0.885 & 0.877\\
 \bottomrule
 \end{longtable}
+
+We also implement the auxiliary particle filter with optimal transition
+kernel. Therfore the algorithm is similiar to the one discussed above
+with the difference that:\\
+{[}{[}{[}{[}Inserire l'algoritmo{]}{]}{]}{]}
 
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
+\DocumentationTok{\#\# Per roberto:::::: cambia molto se usiamo l\textquotesingle{}uno o l\textquotesingle{}altro!!!!!! perchè??}
+\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
+\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x1}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \begin{figure}[H]
 
-{\centering \includegraphics{new-draft_files/figure-latex/unnamed-chunk-31-1}
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-32-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-31}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
 \end{figure}
@@
 \(\psi\) (i.e., \(P_t(x_{t-1},dx_t;\psi)\) and \(f_t(y_t|x_t;\psi)\))
-and let \$\pi(\cdot) \$ be the prior distribution for the parameter
+and let \(\pi(\cdot)\) be the prior distribution for the parameter
 \(\psi\).\\
@@
 
+We provide an example of Liu and West filter fixing \(\delta=0.7\) and
+drawing \(\alpha_{v}\), \(\beta_{v}\), \(\alpha_{w}\) and \(\beta_{w}\)
+from a Uniform \(U(0,10)\). The results are shown in the following
+figure.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-35-1}
+
+}
+
+\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-35}
+\end{figure}
+
+Finally, we compare the performances of the Liu and West filter and the
+other filtering strategies discussed so far. As we expect, in the linear
+Gaussian case, Kalman filter outperforms the other strategies. On the
+other hand, the Liu and West, with hyperparameters setted as described
+above, provides the worst estimates in terms of RMSE. This is not
+necessarily due to an overall weakness of the LWF with respect to the
+other filters, but simply to a not accurate choice of the
+hyperparameters. Moreover, while increasing the number of particles
+improves the accuracy, notice that the threshold for ESS-based
+resampling is not strictly related to better performances.
+
+\begin{longtable}[t]{cccccc}
+\caption{\label{tab:unnamed-chunk-37}RMSE}\\
+\toprule
+N & Threshold & KF & BPF & APF & LWF\\
+\midrule
+100 & 0.50 & 0.879 & 0.916 & 0.881 & 0.883\\
+1000 & 0.50 & 0.879 & 0.882 & 0.899 & 0.977\\
+10000 & 0.50 & 0.879 & 0.885 & 0.874 & 0.945\\
+1000 & 0.50 & 0.879 & 1.172 & 1.677 & 1.140\\
+1000 & 0.25 & 0.879 & 1.085 & 0.974 & 3.419\\
+\addlinespace
+1000 & 0.10 & 0.879 & 4.285 & 0.994 & 1.142\\
+\bottomrule
+\end{longtable}
+
 \section{Convergence}\label{pf_converg}
@@
 
-{\centering \includegraphics{new-draft_files/figure-latex/cll_plot-1}
+{\centering \includegraphics{backup1_files/figure-latex/cll_plot-1}
 

### 2021-07-30 draft-implementation-2.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

--- 2021-07-19 final-draft.tex
@@
 \documentclass[
-]{book}
-\usepackage{amsmath,amssymb}
+]{article}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
@@
 \hypersetup{
-  pdftitle={Final draft},
+  pdftitle={R Notebook},
   hidelinks,
@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage[margin=1in]{geometry}
 \usepackage{color}
@@
 \setcounter{secnumdepth}{5}
-\usepackage{txfonts}
-\usepackage{amsmath}
-\usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
-\usepackage{graphicx}
-\usepackage{MnSymbol} %allows to insert the QED at the end of the proofs
-\usepackage{hyperref} %set the hyperlink to cite sections, equations...
-\urlstyle{same}
-\usepackage{comment}
-\usepackage{enumitem}
-\usepackage{ntheorem}
-\usepackage{lipsum}
-\usepackage{amssymb}
-\usepackage{xcolor}
-\theoremstyle{break}
-\newtheorem{thm}{Theorem}% theorem counter resets every \subsection
-\renewcommand{\thethm}{\arabic{thm}}
-\newtheorem{proposition}{Proposition}
-\newtheorem{assumption}{Assumption}
-\newtheorem{lemma}{Lemma}
-\newtheorem{definition}{Definition}
-\theoremstyle{nonumberplain}
-\newtheorem{proof*}{Proof.}
-\newtheorem{definition}{Definition}
-\renewcommand{\P}{\mathbb{P}}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\setcounter{MaxMatrixCols}{10}
-\setlength{\oddsidemargin}{.0in}
-\setlength{\textwidth}{6 in}
-\setlength{\topmargin}{-.1in}
-\setlength{\textheight}{8.1in}
-\renewcommand{\baselinestretch}{1.5}
-\usepackage{amssymb}
-\newcommand{\E}{\mathbb E}
-\newcommand{\Q}{\mathbb Q}
-\newcommand{\M}{\mathbb M}
-\newcommand{\R}{\mathbb R}
-\newcommand{\B}{\mathcal B}
-\newcommand{\X}{\mathcal X}
-\newcommand{\Y}{\mathcal Y}
-\newcommand{\Pd}{\mathbb P}
-\newcommand{\N}{\mathcal N}
+\newcommand{\R}{\textnormal{\sffamily\bfseries R}}
 \def\code#1{\texttt{#1}}
 \newtheorem{algorithm}{Algorithm}[section]
-\usepackage{float}
-\floatplacement{figure}{}
 \usepackage{float}
@@
 
-\title{Final draft}
+\title{R Notebook}
 \author{}
@@
 \begin{document}
-\frontmatter
 \maketitle
 
-\mainmatter
-\chapter*{Introduction}
-
-\chapter{Markov Processes and State-Space Models}
-\section{Notation}
-
-Let us fix the basic notation that will be employed henceforth. Upper
-case letters, e.g.~\(X_t, Y_t,\) denote random variables, while lower
-case letters denote the realizations. Finite sequences are made compact
-using the semi-colon notation, e.g.~\(x_{0:t}=(x_0, x_1, \dots, x_t)\)
-for some \(t \geq 0.\) The sets of outcomes are denoted by upper case
-calligraphic letters, e.g.~\(\mathcal{X}, \mathcal{Y}.\) The
-\(\sigma\)-algebra of a set \(\mathcal{X}\) reads
-\(\mathcal{B}(\mathcal{X}).\) The pair
-\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) is a measurable space. The
-densities are represented with lower case cursive letters. For example,
-the density of \(X_t\) in \(x_t\) is \(p_t^\theta(x_t),\) where
-\(\theta\in\Theta\) is some (eventually known) vector, while the density
-of \(X_t\) in \(x_t,\) given that \(X_{t-1}\) has previously taken value
-\(x_{t-1},\) is \(p_t^\theta(x_t\mid x_{t-1}).\) Note that
-\(p_t^\theta(x_t\mid x_{t-1})\) is a member of the parametric family
-with parameters \((\theta, x_{t-1})\). The probability distribution of a
-random variable \(X_t\) is represented by \(\mathbb{P}_t(dx_t),\) while
-the probability distribution of a sequence \(X_{0:t}\) is represented by
-\(\mathbb{P}_t(dx_{0:t}).\) The Dirac measure, i.e., the measure that
-assigns probability \(1\) to the singleton \(\{x\},\) is denoted by
-\(\delta_x(dy)\).The expectation operator is denoted by \(\mathbb{E},\)
-where the section of \(\mathbb{E}\) at \(\mathbb{Q},\) i.e.,
-\(\mathbb{E}_{\mathbb{Q}}\) denotes the operator computed with respect
-to the probability distribution \(\mathbb{Q}.\) Let \(\varphi\) be a
-function defined on the random variable \(X\propto\mathbb{Q}.\) Then, we
-denote the expected value of \(\varphi(X)\) with \[
-    \mathbb{E}_{\mathbb{Q}}[\varphi(X)]=
-    \int_{\mathcal{X}}\varphi(x)\mathbb{Q}(dx).
-\] Convergence in distribution is denoted with the symbol \(\implies.\)
-So, \(X\implies\mathcal N(0,1)\) means that the random variable \(X\)
-converges in distribution to a normal Gaussian.
-
-\section{Markov Processes}
-
-Let \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) and
-\((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
-measurable spaces.
-
-\begin{definition}
-A function $P(x,dy): (\mathcal{X}, \mathcal{B}(\mathcal{X}))\rightarrow [0,1]$ such that 
-\begin{itemize}
-    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})$, and 
-    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X})),$
-\end{itemize}
-is defined a probability kernel from $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ to $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})).$
-\end{definition}
-
-Notice that for any random variables \(X_1\) and \(X_0=x_0,\) \[
-    \mathbb{P}_1(dx_{0:1})=\mathbb{P}_0(dx_{0})P_1(x_0,dx_1),
-\] where \(P_1(x_0,dx_1)\) is a probability kernel from
-\((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
-\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) This remark, provides some
-intuition about the link between probability kernels and conditional
-probabilities. As a matter of fact, it is possible to show that for
-given \(X_{0:1}\) with joint probability distribution
-\(\mathbb{P}_1(dx_{0:1}),\)
-\begin{equation}\label{prob_kern_condit_prob}
-    \mathbb{P}_1(X_1\in dx_1\mid X_0=x_0)=P_1(x_0,dx_1).
-\end{equation}
-
-Probability kernels can be used to define Markov processes. Before
-moving to the definition, we will introduce some important concepts.
-Recall that given two measures \(\mathbb{M}\) and \(\mathbb{Q}\) defined
-on a measurable space \((\mathcal{X}, \mathcal{B}(\mathcal{X})),\)
-\(\mathbb{Q}\) is absolutely continuous with respect to \(\mathbb{M},\)
-that is \(\mathbb{Q}\ll\mathbb{M},\) if for all sets
-\(A\in\mathcal{B}(\mathcal{X})\) such that \(\mathbb{M}(A)=0,\) then it
-must also hold that \(\mathbb{Q}(A)=0.\) If
-\(\mathbb{Q}(A)/\mathbb{M}(A)\) is well-defined for all \(A,\) then the
-Radon-Nikodym theorem guarantees that there exists a measurable function
-\(w(x)\geq0\) such that \[
-    w(x)=\frac{\mathbb{Q}(dx)}{\mathbb{M}(dx)}, 
-\] meaning that, \[
-    \mathbb{Q}(A)=\int_A w(x) \mathbb{M}(dx).
-\]
-
-Let \(P_1, P_2, \dots, P_T\) be a finite sequence of probability kernels
-from \((\mathcal{X}, \mathcal{B}(\mathcal{X}))\) to
-\((\mathcal{X}, \mathcal{B}(\mathcal{X})).\) Fix some
-\(\mathbb{P}_0(dx)\) on \((\mathcal{X}, \mathcal{B}(\mathcal{X})).\)
-
-\begin{definition}
-A discrete-time Markov process is a sequence $X_{0:T}$ of random variables whose joint distribution can be written as
-\begin{equation}
-    \mathbb{P}_T(X_{0:T}\in dx_{0:T})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^T P_s(x_{s-1},dx_s).
-\end{equation}
-The set $\mathcal{X}$ is the state-space, $\mathbb{P}_0$ is the initial distribution and the probability kernel $P_t$ is the transition kernel.
-\end{definition}
-
-It is possible to show that for any \(t\in T,\) \[
-\mathbb{P}_T(X_t\in dx_t\mid X_{0:t-1}\in dx_{0:t-1})=
-\mathbb{P}_T(X_t\in dx_t\mid X_{t-1}=x_{t-1})=\mathbb{P}_t(x_{t-1},dx_t).
-\] The first equality implies conditional independence. The second
-equality implies that it is possible to identify conditional
-distributions with transition kernels. Assume \(\mathbb{P}_t,\) with
-\(t\leq T,\) is a sequence of probability measures. Then, the following
-proposition holds.
-
-\begin{proposition}
-For all $t\leq T,$ 
-$$
-\mathbb{P}_T(dx_{0:t})=\mathbb{P}_0(dx_0)\prod\limits_{s=1}^t P_s(x_{s-1},dx_s)=\mathbb{P}_t(dx_{0:t}).
-$$
-\end{proposition}
-
-This means that the marginal distribution of \(X_{0:t}\) with respect to
-\(\mathbb{P}_T\) is \(\mathbb{P}_t.\)
-
-\section{State-Space Models}
-
-We now define \textit{State-Space models} as Markov processes that are
-partially observed. Let \(X_t\in\mathcal X\) and \(Y_t\in\mathcal Y\)
-and let \(\{(X_t,Y_t)\}\) be a stochastic process. To help intuition,
-suppose that \(\{Y_t\}\) is observable, i.e., coincides with a sequence
-of observations, while \(\{X_t\}\) is latent. Consider the initial
-distribution \(\mathbb{P}_0(dx_0)\) and the probability kernels
-\(P_t(x_{t-1},dx_t)\) and \(F_t(x_t,d_t)\), where \(t=1,2,\dots\).
-
-\begin{definition}\label{state-space-def}
-The process $\{(X_t,Y_t)\}$ is a State-Space model if the joint distribution of $(X_{0:T},Y_{0:T})$ is 
-\begin{align*}
-  \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T}) &= \mathbb{P}_0(dx_0)\prod\limits_{t=1}^T P_t(x_{t-1},dx_t)\prod\limits_{t=1}^T F_t(x_t,dy_t)\\
-  &= [\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\prod\limits_{t=1}^T[P_t(x_{t-1},dx_t)F_t(x_t,dy_t)] \\
-  &= \mathbb{P}_T(dx_{0:T})\prod \limits_{t=1}^T F_t(x_t,dy_t).
-\end{align*}
-\end{definition}
-
-From the second inequality, it is apparent that \(\{(X_t,Y_t)\}\) is a
-Markov process with initial distribution
-\([\mathbb{P}_0(dx_0)F_0(x_0,y_0)]\) and transition kernel
-\(P_t(x_{t-1},dx_t)F_t(x_t,dy_t).\) The third equality, instead, shows
-that, \(\{X_t\}\) has a marginal distribution
-\[\mathbb{P}_T(dx_{0:T})=\mathbb{P}_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),\]
-meaning that \(\{X_t\}\) is marginally distributed as a Markov process
-with initial distribution \(\mathbb{P}_T(dx_{0:T})\) and transition
-kernel \(P_t(x_{t-1},dx_t).\)
-
-Assume now that there exists a dominating measures \(\nu(dy)\) such that
-\(F_t(x_t,dy_t)=f_t(y_t\mid x_t)\nu(dy_t)\) for all \(t.\) Then by the
-last equality of definition \ref{state-space-def}, the joint
-distribution of \((X_{0:T},Y_{0:T})\) is
-\begin{equation}\label{state-space-densities}
-   \mathbb{P}_T(X_{0:T}\in dx_{0:T},Y_{0:T}\in dy_{0:T})=\mathbb{P}_T(dx_{0:T})\prod\limits_{t=0}^T f_t(y_t\mid x_t)\prod\limits_{t=0}^T\vu(dy_t).
-\end{equation} Integrating away \(X_{0:T},\) one gets the marginal
-density \begin{equation}\label{state-space-marginal-y}
-\begin{align*}
-  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
-                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
-\end{align*}                
-\end{equation} where \(p_t(y_{0:t})\) is the likelihood function.
-Dividing the left-hand side of equation \ref{state-space-densities} by
-the left-hand side of equation \ref{state-space-marginal-y} and the
-right-hand side of equation \ref{state-space-densities} by the
-right-hand side of equation \ref{state-space-marginal-y}, one obtains
-the filtering distribution \begin{equation}
-\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
-\end{equation}
-
-\chapter{Linear-Gaussian case}
-
-\textit{Linear Gaussian models} are a class of popular state-space
-models. A notable property of such models is that the filtering and
-predictive distributions are analytically tractable, which makes such
-models very convenient in applications.
-
-Suppose that \(\mathcal X=\mathbb R^{d_x}\) and
-\(\mathcal Y=\mathbb R^{d_y}.\) Let
-\(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
-\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
-and assume the initial distribution
-\(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
-if \begin{align}
-X_t &= A_tX_{t-1}+U_t \\
-Y_t &= B_tX_t + V_t.
-\end{align} Note that \(X_t\mid X_{t-1}\) and \(Y_t\mid X_t\) are both
-normally distributed.
-
-\begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
-  Then, it holds that
-  \begin{align}
-    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
-    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
-    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
-  \end{align}
-\end{lemma}
-\begin{proof*}
-   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
-\begin{equation} \label{normal-distrib-rewrite}
-        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
-\end{equation}
-with $A=Q_0^{-1}$ and $B=Am_0.$
-Now, by Bayes' Theorem,
-
-    \begin{align*}
-        p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
-                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
-                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
-                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
-                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
-    \end{align*}
-
-Hence, it follows that 
-\begin{equation*}
-    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
-\end{equation*}
-and 
-    \begin{align*}
-        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
-            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
-    \end{align*}
-\end{proof*}
-
-Assume that
-\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
-Then, using the transition equation of the latent variable, by the
-linearity of the Gaussian distribution, we have that the predictive
-distribution is \begin{equation}
-\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
-\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
-lemma \ref{kalman-filter}, it follows that the filtering distribution is
-\begin{align*}
-  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
-  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
-\end{align*}
-
-\section{Implementation}
+\textbf{Implementation}
 
@@
 discussed. For the sake of simplicity we use a random walk plus noise
-model, i.e.~the most basic form of a linear Gaussian state-space
-model.\\
+model, i.e.~the most basic form of a linear Gaussian state-space model.
 \begin{align}
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-5-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-3-1}
 
@@
 
-\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
+\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-3}
 \end{figure}
@@
 
-Our \texttt{DLM} function implement in R replicate this steps.\\
-
-\hrule
-\hrule
-\texttt{DLM(data,sig2,tau2,m0,C0)}
+Our \texttt{DLM} function implement in
+\textnormal{\sffamily\bfseries R} replicate this steps.\\
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{DLM(data,sig2,tau2,m0,C0)}\\
+
 \hrule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-7-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-5-1}
 
@@
 
-\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
+\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-5}
 \end{figure}
@@
 
-\textbf{Implementation}
-
-In this section we present a practical illustration of the algorithms
-discussed. For the sake of simplicity we use a random walk plus noise
-model, i.e.~the most basic form of a linear Gaussian state-space model.
-\begin{align}
-y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
-x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
-x_{0} & \sim N(m_{0},C_{0})
-\end{align} As already mentioned before, in this case the filtering
-distribution can be computed in closed form solutions using the Kalman
-filter. However, this toy example will be used also to illustrate more
-involved filtering strategies described in this work. We believe indeed
-that it represents a useful starting point to understand the logic of
-the algorithms which may be eventually replicated when dealing with more
-complex models.\\
-The filtering strategy is applied to 50 simulated data. Figure XX shows
-the simulated true states sequence assuming as data generating process
-the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
-process form Equation (1) with \(\sigma^{2}=1\).
-
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-8-1}
-
-}
-
-\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-8}
-\end{figure}
-
-The discussion of Kalman Filter will continue in Section XX where we
-compare it to the Particle Filter.
-
-\chapter{Particle Filter}
-
-In the previous chapter we explained how the Kalman Filter is applied to
-the case of a linear model with Gaussian errors to find closed-form
-solution for the filtering and the predictive distributions.
-Nonetheless, the majority of time-series models have a non-linear
-structure and errors might follow non-Gaussian distributions.\\
-In this chapter we describe \textit{particle filter} as an algorithm
-that allows to update sequentially the filtering and the predictive
-distributions in the non-linear and non-Gaussian case. Additionally, the
-Kalman Filter can be seen as a special case of the more general particle
-filter.\\
-In section \ref{pf_intro} we introduce the fundamental notions of
-\textit{importance sampling}, \textit{sequential importance sampling}
-and \textit{resampling}. Then, in section \ref{pf_main} we describe
-three particle filtering algorithms: the \textit{bootstrap}, the
-\textit{guided} and the \{auxiliary particle filters\}. Instead, in
-section \ref{pf_lui_west} we describe the \{Liu and West filter\}, when
-particle filtering is used to estimate model's parameters. Finally, in
-section \ref{pf_converg} we establish some general convergence results
-for Particle Filter - Monte Carlo estimators.
-
-\section{Introductory notions: sampling and resampling}\label{pf_intro}
-
-Let \(\varphi: \mathcal X\rightarrow \mathbb R\) be a measurable,
-integrable and bounded real-valued function, where
-\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) is a probability
-space.\footnote{$\mathcal B(\mathcal X)$ is the Borel sigma-algebra of $\mathcal X$.}
-Suppose that the probability measure \(\mathbb Q\) admits a density
-\(q(x)\) with respect to some measure \(\nu(dx)\) and it is feasible to
-sample from that density.\\
-To estimate the expected value of \(\varphi\) with respect to the
-density \(q\), the basic Monte Carlo (MC) method would require to sample
-\(N\) times from the target density \(q\) and to compute the average of
-the function \(\varphi\) values. \(\hat \varphi_{MC}\) is the standard
-MC estimator: \begin{equation*}
-   \hat \varphi_{MC}=\frac{1}{N} \sum_{n=1}^N\varphi (X^n)\approx \mathbb E_q(\varphi)~~~X^n\thicksim q(\cdot)
-\end{equation*} However, in many circumstances either it is impossible
-to sample from the target density or there exist another density - call
-it \(m(\cdot)\) - different from \(q\) that produces
-\texttt{more\ efficient"\ estimates.\ \ In\ all\ the\ cases\ for\ which\ sampling\ takes\ place\ from\ a\ distribution\ different\ from\ the}correct"
-(i.e., the true) one, importance sampling and resampling methods can be
-applied (as augmented versions of the standard MC algorithm).
-
-\subsection{Importance sampling}
-
-Let us consider two probability spaces:
-\((\mathcal X,\mathcal B(\mathcal X),\mathbb Q)\) and
-\((\mathcal X,\mathcal B(\mathcal X),\mathbb M)\) and let the measurable
-function \(w:\mathcal X\rightarrow \mathbb R\) be proportional to the
-Radom-Nikodym derivative between \(\mathbb Q\) and \(\mathbb M\), i.e.,
-\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\).\\
-The following result represents the starting point for the idea of
-importance sampling.
-
-\begin{lemma}\label{lemma_imp_sam_1}
-For any measurable, integrable and bounded function $\varphi: \mathcal X\rightarrow \mathbb R$, the following holds:
-\begin{equation}
-    \mathbb E_\mathbb M(\varphi \cdot w)=\mathbb E_\mathbb Q(\varphi)\cdot \mathbb E_\mathbb M(w)
-\end{equation}
-\end{lemma}
-
-\begin{proof*}
-Note that $\mathbb Q(dx)=\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)$. Hence:
-\begin{equation*}
-    \mathbb E_\mathbb Q(\varphi)=\int_\mathcal X\varphi(x)\mathbb Q(dx)=\int_\mathcal X\varphi(x)\frac{w(x)}{\mathbb E_\mathbb M(w)}\mathbb M(dx)=\frac{\mathbb E_\mathbb M(\varphi \cdot w)}{\mathbb E_\mathbb M(w)}
-\end{equation*}
-$\hfill\blacksquare$
-\end{proof*}
-
-Lemma \ref{lemma_imp_sam_1} represents the key theoretical foundation of
-\textbf{importance sampling} (IS), a numerical algorithm that allows to
-sample from a \textit{proposal density} to estimate moments that are
-computed with respect to the \textit{target density}, by means of
-re-weighting the values of the target function \(\varphi\).\\
-Adopting the previous notation, the target measure is \(\mathbb Q\),
-while the proposal is \(\mathbb M\). IS requires to sample \(N\) values
-from \(\mathbb M\), then to find the associated
-\textit{normalized}\footnote{Weights are normalized when $\mathbb E_\mathbb M(W)=1$. In this case $W$ is the Radom-Nikodym derivative between $\mathbb Q$ and $\mathbb M$.}
-weights \(W(X^n)\), finally to compute a weighted average of the values
-taken by the function \(\varphi\). The IS is implemented according to
-the following algorithm.
-
-\begin{enumerate}
-    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
-    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
-    \item Compute the estimator:
-    \begin{equation*}
-        \hat \varphi= \frac{1}{N} \sum_{n=1}^N\varphi (X^n)W(X^n).
-    \end{equation*}
-\end{enumerate}
-
-Note that \(\hat \varphi\) is an unbiased estimator for the target of
-inference \(\mathbb E_\mathbb Q(\varphi)\). In fact: \begin{equation*}
-    \begin{split}
-        \mathbb E_\mathbb M(\hat \varphi)=\frac{1}{N} \sum_{n=1}^N \mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb M(\varphi \cdot W)=\mathbb E_\mathbb Q(\varphi)
-    \end{split}
-\end{equation*} where the last equality follows from lemma
-\ref{lemma_imp_sam_1} and the fact that weights are normalized to 1.
-
-\subsubsection{Auto-normalized IS}
-
-Usually, the Radom-Nikodym derivative is known up to a normalizing
-constant and the IS algorithm has to be described in more general terms.
-In the previous section the \textit{normalized} weighting function
-\(W(\cdot)\) was the Radom-Nikodym derivative of \(\mathbb Q\) with
-respect to \(\mathbb M\). More generally, consider the function
-\(w(x)\propto\frac{\mathbb Q(dx)}{\mathbb M(dx)}\), denoted as
-\textit{importance function}, proportional to the derivative
-\(W(\cdot)\).\\
-In this case, weights have to be \textit{auto}-normalized to obtain the
-IS - Monte Carlo estimator. The auto-normalized IS is implemented
-according to the following procedure.
-
-\begin{enumerate}
-    \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
-    \item Compute the associated \textit{un-normalized} weights: $w(X^n)$ for $n=1,...,N$.
-    \item Normalize the weights, for $n=1,...,N$:
-    \begin{equation*}
-        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
-    \end{equation*}
-    \item Compute the estimator:
-    \begin{equation*}
-        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
-    \end{equation*}
-\end{enumerate}
-
-While the auto-normalized estimator is biased (since \(W^n\) involves a
-ratio of random variables), it is consistent (see \ref{pf_is_conv} for
-details). ~\\
-It is useful to interpret IS as an algorithm that allows to find an
-approximation for the probability measure \(\mathbb Q\) as a weighted
-average of Dirac measures for the sampled values (\(X^n\)):
-\(\delta_{X^n}(dx)\) where the weights are the auto-normalized ones
-(\(W^n\)). In particular: \begin{equation}
-    \mathbb Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \mathbb M
-\end{equation} Furthermore, the IS estimator is the expected value of
-the function \(\varphi\) with respect to the approximating probability
-measure \(\mathbb Q^N\): \begin{equation*}
-    \begin{split}
-        \mathbb E_{\mathbb Q^N}(\varphi)=&\int_\mathcal X\varphi(x)\mathbb Q^N(dx)=\int_\mathcal X\varphi(x)\sum_{n=1}^NW^n\delta_{X^n}(dx)=\\
-       = &\sum_{n=1}^N\int_\mathcal X\varphi(x)W^n\delta_{X^n}(dx)=\sum_{n=1}^N\varphi (X^n)W^n=\hat \varphi_{AN}
-    \end{split}
-\end{equation*}
-
-\subsubsection{Convergence of the IS estimator}\label{pf_is_conv}
-
-Consider the auto-normalized IS estimator
-\(\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n\). We can write:
-\begin{equation*}
-   \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)=\mathbb E_{\mathbb Q^N}(\varphi)-\mathbb E_{\mathbb Q}(\varphi)=\frac{N^{-1}\sum_{n=1}^Nw(X^n)[\varphi(X^n)-\mathbb E_{\mathbb Q}(\varphi)]}{N^{-1}\sum_{n=1}^Nw(X^n)}
-\end{equation*} Let
-\(\bar \varphi(X):=\varphi(X)-\mathbb E_{\mathbb Q}(\varphi)\). Applying
-the CLT to the numerator, assuming that
-\(\mathbb E_{\mathbb M}(w^2\bar\varphi^2)<+\infty\), we get:
-\begin{equation}
-    \sqrt{N}\Bigg\{\frac{1}{N}\sum_{n=1}^Nw(X^n)\bar \varphi (X^n)\Bigg\}\implies \mathcal N\big(0,\mathbb E_{\mathbb M}(w^2\bar\varphi^2)\big)\label{is_conv_clt}
-\end{equation} where \(\implies\) denotes convergence in distribution.
-Instead, for the denominator the strong LLN implies that:
-\begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \mathbb E_{\mathbb M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
-\end{equation} Applying Slutsky's theorem, together with
-\eqref{is_conv_clt} and \eqref{is_conv_lln}, we obtain: \begin{equation}
-   \sqrt{N}\big[  \hat \varphi_{AN} - \mathbb E_{\mathbb Q}(\varphi)\big] \implies \mathcal N\Bigg(0,\frac{\mathbb E_{\mathbb M}(w^2\bar\varphi^2)}{[\mathbb E_{\mathbb M}(w)]^2}\Bigg)
-\end{equation} Therefore, the IS (auto-normalized) estimator for
-\(\varphi\) is both consistent and asymptotically normal.
-
-\subsubsection{Effective Sample Size}
-
-A common measure of efficiency for IS is the
-\textit{effective sample size} (ESS), defined as: \begin{equation*}
-    ESS(W^{1:N}):=\frac{1}{\sum_{n=1}^N(W^n)^2}=\frac{[\sum_{n=1}^N w(X^n)]^2}{\sum_{n=1}^N(w(X^n))^2}
-\end{equation*} Observe that the \(ESS\) lies in the interval \([1,N]\).
-For instance, if \(W^n=1/N\), as in the case of the standard MC method,
-\(ESS=N\).\\
-It is possible to relate the \(ESS\) to the variance of the importance
-function. In fact: \begin{equation*}
-    \frac{N}{ESS}=1+\frac{N^{-1}\sum_{n=1}^N(w(X^n))^2-\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}{\big[N^{-1}\sum_{n=1}^N w(X^n)\big]^2}
-\end{equation*} The second term in the numerator of the previous formula
-is the square of the coefficient of variation (i.e., the ratio between
-the variance and the squared mean) of the un-normalized weights. It is
-clear that a lower \(ESS\) is associated with a higher variance for the
-weights.
-
-\subsection{Sequential Importance Sampling}
-
-When dealing with state-space models, the importance sampling algorithm
-has to be applied dynamically and the importance weights are updated
-period-by-period. To illustrate the relevance of the
-\textit{sequential importance sampling} (SIS) algorithm, we consider a
-two-period stochastic process.~\\
-Let us consider the approximation of the measure \(\mathbb Q_0\)
-obtained at time 0 through IS from the measure \(\mathbb M_0\):
-\begin{equation*}
-    \mathbb Q_0^N(dx_0)=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0),~~\text{with } X_0^n\thicksim \mathbb M_0,~~\text{and }  W_0^n:=\frac{w_0(X_0^n)}{\sum_{m=1}^Nw_0(X_0^m)}.
-\end{equation*} Moreover, let \(M_1(x_0,dx_1)\) be the transition
-kernel. We want to update, sequentially, the approximating measure
-\(\mathbb Q_0\) to obtain an approximation of the measure:
-\begin{equation*}
-    \mathbb Q_{1}(dx_{0:1})=\mathbb Q_0(dx_0)M_1(x_0,dx_1)
-\end{equation*} SIS works according to the following algorithm.
-
-\begin{enumerate}
-    \item Sample $N$ values for $X_0$ from the proposal measure $\mathbb M_0$: $X_0^n\thicksim \mathbb M_0$ for $n=1,...,N$.
-    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^n,dx_1)$ for $n=1,...,N$.
-    \item Compute the auto-normalized weights.
-In this case, they are computed taking into account only time 0 sampled observations since $X_1^n$ is sampled from the correct distribution: $W_1^n=W_0^n$.
-\end{enumerate}
-
-Therefore, the particle approximation of \(\mathbb Q_{1}(dx_{0:1})\) is:
-\begin{equation*}
-      \mathbb Q_{1}^N(dx_{0:1})=\sum_{n=1}^NW_0^n\delta_{X_0^n}(dx_0)M_1(X_0^n,dx_1)
-\end{equation*}
-
-\subsection{Sequential Importance Resampling}
-
-A second approach can be used when sampling has to be done sequentially:
-\textit{sequential importance resampling} (SIR). Differently from SIS,
-an intermediate resampling stage takes place, in which time 0 values are
-sampled with replacement according to a given resampling strategy
-(usually, sampling indexes from a multinomial distribution with
-probabilities equal to the weights computed through IS at time 0).\\
-While in the SIS all the particles (i.e., the sampled values)
-\texttt{survive",\ with\ the}risk" that particles associated with small
-weights are carried-over (in the sense that sampling in period 1 occurs
-from the kernel \(M_1(X_0^n,dx_1)\)), the same does not occur in SIR. In
-this second algorithm, particles with larger weights are more likely to
-be resampled and be used to ``generate" the particles in the next
-period.\\
-We describe the SIR more formally, focusing on the previous two-period
-case.
-
-\begin{enumerate}
-    \item Sample $N$ values $X_0$ from the proposal measure $\mathbb M_0$: $X_0^n\thicksim \mathbb M_0$ for $n=1,...,N$.
-    \item Compute the weights $W_0^n$ for $n=1,...,N$.
-    \item Sample $N$ indexes $(A_1^1,...A_1^N)$ from a multinomial distribution such that for all $j=1,...N$, $Pr(\{A_1^j=n\})=W_0^n$, for all $n=1,...N$.
-    \item Build a new time 0 sample, using the values for $X_0$ associated to the previously extracted indexes $(A_1^n)_{n}$: $(X_0^{A_1^1},...,X_0^{A_1^N})$.
-    \item Sample $N$ values for $X_1$ from the transition kernel $M_1$: $X_1^n\thicksim M_1(X_0^{A_1^n},dx_1)$ for $n=1,...,N$.
-\end{enumerate}
-
-Steps 3-4 constitute the actual resampling ones. At this point, let us
-define the new two-period sequence
-\(\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)\). Then, the particle
-approximation for the joint distribution of \((X_0,X_1)\) becomes:
-\begin{equation*}
-     \mathbb Q_{1}^N(dx_{0:1})=\frac{1}{N}\sum_{n=1}^N\delta_{\tilde X_{0:1}^n}(dx_{0:1})
-\end{equation*} Note that, after resampling occurred, each particle
-\(\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)\) has the same probability
-weight.\\
-To understand the reason of the previous statement, let us assume that
-both \(\mathbb Q_0\) and \(\mathbb M_0\) admit densities \(q_0\),
-\(m_0\) with respect to the same dominating measure \(\nu(dx_0)\) and
-that the transition kernel \(M_1\) also admits the density \(m_1\).
-Consider the sequential updating for the weights: \begin{equation}
-    w^n_1=\frac{\mathbb Q_1(\tilde dx_{0:1})}{\mathbb M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
-=\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
-\end{equation} where \(\mathbb M_1\) denotes the probability measure for
-the extraction of the particle \(\tilde x_{0:1}^n\) according to SIR.
-\(\mathbb M_1\) (second equality sign in the previous formula) can be
-rewritten as the product between the kernel for \(x_1^n\) given
-\(x_0^{A_1^n}\) and the probability of \(x_0^{A_1^n}\) itself. This last
-probability can also be rewritten as the product between the proposal
-density \(m_0\) from which \(X_0\) is extracted and the probability that
-it is sampled again when resampling occurs.
-
-\subsection{SIS vs SIR}
-
-The choice between the two sequential sampling strategies (SIS and SIR)
-has significant implications in terms of computational cost and variance
-of the importance weights. In fact, it is clear that SIR is more costly
-computationally since it requires an intermediate resampling step.
-Nonetheless, SIS suffers from the problem knonw as
-\textit{curse of dimensionality}, i.e., the fact that the \(ESS\)
-collapses over time as the variance of the weights diverges.\\
-Consider an extension of SIS to a \(T\)-period model and let us assume
-that the target and the proposal kernels admit densities, respectively,
-\(q_t\) and \(m_t\) with respect to the same dominating measures (this
-implies that the joint
-distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation}
-admit a density too). Then, weights, for every \(t=1,...,T\), are
-updated according to: \begin{equation}
-    w_t^n\propto \frac{q_t(x_{0:t}^n)}{m_t(x_{0:t}^n)}\propto \underbrace{\frac{q_{t-1}( x_{0:t-1}^n)}{m_{t-1}(x_{0:t-1}^n)}}_{\propto w_{t-1}^n}\cdot \frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\propto
-    w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
-\end{equation} Equation \eqref{sis_update} suggests, intuitively, that
-over time new variance is added from the incremental weights (i.e., the
-updating factor \(q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)\)) determining
-the curse of dimensionality issue. Instead, when resampling takes place,
-since the past weights \(w_{t-1}\) are reset to be equal across them all
-the variance of \(w_t\) inherited from the past is shut-down and the
-only source of variability at each stage comes from the incremental
-weights.
-
-\section{Particle Filter}\label{pf_main}
-
-\color{red}
-
-Let us consider the standard state-space model introduced in the
-previous chapters. \color{black} \textit{Particle filter} (PF) is a
-numerical algorithm that is used to obtain approximations for the
-filtering - \(\mathbb P_t (dx_t|y_{1:t})\) - and the prediction -
-\(\mathbb P_t (dx_t|y_{1:t-1})\) - distributions for inference in
-state-space models.~\\
-In a state-space model sampling sequentially from the transition kernels
-for the states is not enough to obtain good approximations since the
-observation sample (\(y_{1:T}\)) contains information about the
-underlying states through the likelihood function \(f_t(y_t|x_t)\).
-Observations constitute signals for the model's states.\\
-Therefore, the conditional distribution of the states, given the
-observed data - \(\mathbb P_t(dx_{1:t}|y_{1:t})\) - changes over time as
-new data points become available. The PF algorithm allows to incorporate
-sequentially the new data to obtain new approximations for this
-distribution (and for the filtering and prediction ones).\\
-In general, at each step of the algorithm, states are sampled
-sequentially from the proposal measure, call it \(\mathbb M_t\), and
-importance auto-normalized weights are computed as the Radom-Nikodym
-derivative of the target \(\mathbb P_t\) with respect to the proposal
-\(\mathbb M_t\). Moreover, the Markov properties of the sequence of the
-states allows to update sequentially the importance weights.~\\
-We will analyse three different types of particle filtering procedures:
-\textit{bootstrap}, \textit{guided} and
-\textit{auxiliary particle filters}, that differ between them with
-respect to their scope and efficiency.
-
-\subsection{Degeneracy of the SIS in Particle Filter}
-
-Usually, PF requires a resampling step at each stage of the algorithm.
-Recall that, in general, resampling determines a trade-off between the
-increase in the variance at the resampling stage and a decrease in the
-variance when future states are sampled from the transition kernel.\\
-\texttt{Dead"\ particles\ (i.e.,\ those\ with\ small\ weights)\ are\ \ likely\ to\ be\ removed\ when\ resampling\ takes\ place\ and\ past\ weights\ get}reset"
-(usually, they are set to be equal to \(1/N\)). Otherwise, the
-\textit{curse of dimensionality} problem will emerge and the variance of
-the weights will diverge over time as the variance of past weights
-increases as new updates occur period-by-period. Figure ***** shows an
-example of a PF where SIS is used: note that the \(ESS\) drops
-dramatically after few periods, meaning that the variance of the
-importance weights diverges. This motivates the use of a SIR structure
-for the PF algorithm.~\\
-Lastly, it is now common practice to use a mixed approach with respect
-to resampling (recall that resampling has an additional computational
-cost). Generally, resampling is done \textit{adaptively}, when some
-condition is met. A standard criterion that is used (and that we adopt
-in our implementations of the PF) is to resample when the \(ESS\) drops
-below some threshold \(ESS_{min}\in [1,N]\): \begin{equation}
-    ESS(W_{t-1}^{1:N})<ESS_{min}
-\end{equation}
+\hfill\break
 
@@
 
-Our \texttt{SISfun} function implemented in R replicate this steps.\\
-
-\hrule
-\hrule
-\texttt{SISfun(data,N,m0,C0,tau,sigma)}
+Our \texttt{SISfun} function implemented in
+\textnormal{\sffamily\bfseries R} replicate this steps.\\
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SISfun(data,N,m0,C0,tau,sigma)}\\
+
 \hrule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-11-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-8-1}
 
@@
 
-\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-11}
+\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-8}
 \end{figure}
-
-\subsection{Bootstrap Particle Filter}
-
-The bootstrap particle filter (BPF) is the easiest form of an adaptive
-PF algorithm. In fact, future states are sampled from the
-\textit{correct} transition kernel and the weights are updated
-sequentially using the likelihood \(f_t(y_t|x_t)\).\\
-We summarize the procedure with the following scheme.
-
-\begin{enumerate}
-    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb P_0(dx_0)$ and set the un-normalized weights $w_0^n=1$.\
-    Then, the normalized weights are: $W_0^n=\frac{1}{N}$.
-    \item \textbf{For every $t=1,...T$:}
-    \begin{enumerate}
-        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:\footnote{It is indifferent to implement this resampling step at the beginning or at the end of each iteration (as we do in our codes) as long as the initial stage $t=0$ generates weights that are constant and equal to $1/N$.
-In fact, in this case, resampling will never take place at the beginning of the first iteration since $ESS(W_0^n=1/N)=N$ that is larger than or equal to $ESS_{min}$.\
-        However, when we describe the algorithm, we chose to leave this resampling step at the beginning of each iteration.
-In fact, when considering the guided or auxiliary PF, we can, more generally, sample the initial values (time $t=0$) for the state from a proposal distribution different from the target one.
-In this case, the weights at $t=0$ might differ from $1/N$, hence resampling might take place already at the beginning of the first iteration.}
-\begin{enumerate}
-    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
-    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
-\end{enumerate}
-\textbf{Otherwise, } 
-\begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$.
-    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}^n$ for $n=1,...,N$.
-\end{enumerate}
-\item Sample the future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t)$  for $n=1,...,N$.
-\item Update the un-normalized weights through the likelihood function for $n=1,...,N$.:
-\begin{equation*}
-    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n).
-\end{equation*}
-\item Normalize the weights  for $n=1,...,N$.: $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
-    \end{enumerate}
-\end{enumerate}
-
-Therefore, at the initial step the pre-sample period states \(x_0^n\)
-are extracted and every sampled state is assigned an equal weight since
-there is no corresponding observation \(y_0\) that gives additional
-information on the likelihood of that extraction.\\
-Resampling does never take place in
-\(t=1\)\footnote{Since $ESS(W_0^n=1/N)=N\geq ESS_{min}$.} and the state
-in \(t=1\) is sampled from the correct transition kernel
-\(P_1(x_0^n,dx_1)\). Now, the time 0 weights are updated incorporating
-the information from the data point \(y_1\) in a very intuitive way:
-more \textit{likely} states (given \(y_1\)) are given higher weights:
-\(w_1^n=f_1(y_1|x_1^n)\). This weights are used to approximate the time
-\(1\) filtering distribution.\\
-At a subsequent stage, the \(ESS\) criterion should be checked and, in
-case it is satisfied, resampling takes place and time \(1\) weights are
-reset to be equal to \(1/N\).~\\
-The key characteristic of the PF is the fact that the importance weights
-are updated sequentially, according to the ``information" provided by
-the observed data (i.e., the sequence \(y_{1:T}\)). Recall that time
-\(t\) weight is the Radom-Nikodym derivative of the target with respect
-to the proposal distribution. In the BPF, the time \(t\) target is the
-measure \(\mathbb P_t(dx_{0:t}|y_{1:t})\), while the proposal, denoted
-generically by \(\mathbb M_t\), is
-\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\).
-Hence: \begin{equation}
-    \begin{split}
-        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f(y_t|x_t)}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)}=\\
-        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot f(y_t|x_t)=\hat w_{t-1}\cdot f(y_t|x_t)
-    \end{split}\label{weight_updated_bpf}
-\end{equation} Finally, the BPF is able to generate the two key objects
-for inference in state-space models:
-
-\begin{itemize}
-    \item the predictive distribution:
-    \begin{equation}
-    \begin{split}
-        \mathbb P_t^N (dx_t|y_{1:t-1})=&\frac{1}{\sum_{n=1}^N\hat w_t^n}\sum_{n=1}^N\hat w_t^n\delta_{X_t^n}(dx_t)=\\
-        =&\begin{cases}
-        \frac{1}{N}\sum_{n=1}^N\delta_{X_t^n}(dx_t)
-        ~~~~~~~~~~~~~~~~ ~\text{ if resampling occurs}\\
-        \frac{1}{\sum_{n=1}^N w_{t-1}^n}\sum_{n=1}^Nw_{t-1}^n\delta_{X_t^n}(dx_t)~~~~ \text{ otherwise}
-        \end{cases},
-        \end{split}
-    \end{equation}
-    \item the filtering distribution:
-    \begin{equation}
-        \mathbb P_t^N (dx_t|y_{1:t})=\sum_{n=1}^NW_t^n\delta_{X_t^n}(dx_t)
-    \end{equation}
-\end{itemize}
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-14-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-11-1}
 
@@
 
-\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-14}
+\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-11}
 \end{figure}
@@
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-16}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-13}Root Mean Square Errors}\\
 \toprule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-17-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-14-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-17}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-14}
 \end{figure}
@@
 \hfill\break
-
-\subsection{Guided Particle Filter}
-
-It is not always possible to sample sequentially from the correct
-transition kernel. Moreover, even if it were possible, there exist a
-proposal transition kernel that makes the algorithm more efficient. This
-motivates the \textit{guided particle filter} (GPF), that can be
-interpreted as a more general version of the GPF. In fact, with the GPF
-it is possible, at each stage, to sample new states from a transition
-kernel that differs from the correct one.~\\
-Let \(P_t(x_{t-1},dx_t)\) be the correct kernel and consider a kernel
-\(M_t(x_{t-1},dx_t)\) defined on the same measurable space of \(P_t\).
-Moreover, let us assume that that \(P_t\) is absolutely continuous with
-respect to
-\(M_t\)\footnote{This condition is stronger than what would be needed for the existence of the importance weights in GPF, i.e., that  $P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)\ll M_t(x_{t-1},dx_t)$. }
-(i.e., \(P_t\ll M_t\)). If sampling occurs sequentially from this
-kernel, the way in which weights are updated should differ from the BPF.
-Consider equation \eqref{weight_updated_bpf} and let us reformulate it
-for the GPF case. Here, the proposal distribution is not
-\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\)
-anymore, but becomes
-\(\mathbb M_t(dx_{0:t}|y_{1:t})=\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)\).
-Hence: \begin{equation}
-    \begin{split}
-        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot M_t(x_{t-1},dx_t)}=\\
-        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{\hat w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)} =\hat w_{t-1}\cdot\underbrace{\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}}_{\text{updating factor}}
-    \end{split}\label{weight_updated_gpf}
-\end{equation} Note that the updating factor is more complex that the
-one of the BPF: in addition to the likelihood function \(f_t(y_t|x_t)\),
-the fact that sampling occurs from a kernel different from the correct
-ones requires the additional term
-\(P_t(x_{t-1},dx_t)/M_t(x_{t-1},dx_t)\).~\\
-We are now ready to establish, informally, the optimality result about
-the transition kernel. Let us denote by
-\(G_t=[P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)dy_t]/M_t(x_{t-1},dx_t)\) the
-\textit{incremental weights} (i.e., the updating factor) and by
-\(M_t^*(x_{t-1},dx_t)\) the optimal kernel. It can be
-proved\footnote{For a formal proof see \cite{**}, th. 10.1 p. 142.} that
-the optimal kernel, i.e., the one that minimizes the variance of the
-incremental weights \(G_t\) (and of the time \(t\) weights themselves),
-is: \begin{equation}
-    M_t^*(x_{t-1},dx_t)=\frac{1}{\int_{\mathcal X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)} P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)
-\end{equation} Note that \(M_t^*\) is the conditional distribution of
-\(x_t\) given \(x_{t-1}\) and
-\(y_t\).\footnote{In fact, assuming that the $P_t$ admits a density $p_t$, applying Bayes rule, we have that $m_t(x_t|x_{t-1},y_t)\propto p_t(x_t|x_{t-1})\cdot f(y_t|x_t)$.}
-The intuition for this result is apparent when considering equation
-\eqref{weight_updated_gpf}: if \(M_t=M_t^*\), the updating factor
-collapses to \(\int_{\mathcal X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)\)
-and information from the observed data are incorporated efficiently.\\
-As a matter of fact, it is usually impossible to sample from the optimal
-kernel and, in practice, it is replaced by its linear Gaussian
-approximation. ~\\
-The GPF can be resumed by the following steps.
-
-\begin{enumerate}
-    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb M_0(dx_0)$ and set the un-normalized weights: 
-    $$w_0^n=\frac{\mathbb P_0(dx_0)}{\mathbb M_0(dx_0)}.$$
-    Then, the normalized weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
-    \item \textbf{For every $t=1,...T$:}
-    \begin{enumerate}
-        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
-\begin{enumerate}
-    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_n$.
-    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
-\end{enumerate}
-\textbf{Otherwise, } 
-\begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$
-    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
-\end{enumerate}
-\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
-\item Update the un-normalized weights, for $n=1,...,N$:
-\begin{equation*}
-    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
-\end{equation*}
-\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
-    \end{enumerate}
-\end{enumerate}
-
-Note that \(\mathbb M_0(dx_0)\) is the initial sampling distribution,
-that could differ from the correct one \(\mathbb P_0(dx_0)\).~\\
-To conclude, one of the issues related to the BPF is the fact that new
-states in \(t\) are sampled ignoring their likelihood given the future
-observation \(y_t\). Hence, extractions from the ``correct" transition
-kernel may be associated to low weights (small value for the
-likelihood). Instead, the GPF allows to solve for this problem, defining
-a proposal kernel that depends on the likelihood itself. Therefore, the
-extraction of future states already incorporates the information given
-by the data resulting in a more efficient procedure.
 
@@
 \(N(x_{t-1}^{(i)},\tau^{2}+\sigma^{2})\), evaluated at \(y_{t}\). In
-other words the algorithm implemented in R is:
+other words the algorithm implemented in
+\textnormal{\sffamily\bfseries R} is:
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-20-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-17-1}
 
@@
 
-\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-20}
+\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-17}
 \end{figure}
@@
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-22}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-19}Root Mean Square Errors}\\
 \toprule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-23-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-20-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-23}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-20}
 \end{figure}
 
-\subsection{Auxiliary Particle Filter}
-
-The \textit{auxiliary particle filter} (APF) constitutes a further
-extension of the BPF. The GPF allowed to sample from a transition kernel
-different from the correct ones. The APF, instead, allows to extract the
-ancestor variables from an arbitrary distribution at the resampling
-step.~\\
-Let \(\eta_t:\mathcal X\rightarrow \mathbb R_+\) be a non-negative,
-real-valued positive function, called \textit{auxiliary function}. At
-time \(t\), when the resampling stage occurs, the multinomial
-distribution for the indexes to be extracted can use probabilities that
-differ from the weights in \(t-1\), \(W_{t-1}^n\). The new weights,
-called \textit{auxiliary weights}, are computed as a transformation of
-the original ones, denoted as \textit{inferential weights}, through the
-auxiliary function: \begin{equation}
-    \tilde W_t^n=\frac{W_t^n\cdot \eta_t(X_t^n)}{\sum_{m=1}^NW_t^m\cdot \eta_t(X_t^n)}
-\end{equation} Note that the computation of the auxiliary weights is
-done from the auto-normalized inferential ones. Additionally, note that
-imposing \(\eta_t(x_t)=1\) we recover the GPF and the BPF as special
-cases of the APF.\\
-Clearly, inferential weights are updated by incorporating the fact that
-resampling occurs according to different weights (the auxiliary ones):
-\begin{equation}
-    w_t^n=\frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}}G_t \label{apf_weights_up_}
-\end{equation} where \(G_t\) is the incremental weight defined in the
-section on GPF.~\\
-
-To understand the logic of this sequential weight updating, consider
-\eqref{weight_updated_gpf}. In particular, suppose that multinomial
-resampling takes place at every stage and let
-\(\tilde w_{t-1}:=w_{t-1}\cdot \eta_{t-1} (x_{t-1})\). Then:
-\begin{equation}
-    \begin{split}
-        w_t\propto&\frac{\mathbb P_t(dx_{0:t}|y_{1:t})}{\mathbb M_t(dx_{0:t}|y_{1:t})}\propto \frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})\cdot \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
-        =&\underbrace{\frac{\mathbb P_{t-1}(dx_{0:t-1}|y_{0:t-1})}{\mathbb M_{t-1}(dx_{0:t-1}|y_{1:t-1})}}_{w_{t-1}}\cdot\frac{1}{\tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\\
-        =&\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t(x_{t-1},dx_t)}\label{weight_updated_apf}
-    \end{split}
-\end{equation} It is possible to derive an optimality result for the
-auxiliary functions\footnote{Details in ***** th 10.2, p. 148}. In the
-special case in which the transition kernel used for sampling
-corresponds with the optimal one \(M_t^*\), the optimal auxiliary
-function (i.e., the auxiliary function that minimizes the variance of
-the inferential weights) becomes: \begin{equation}
-    \eta_{t-1}^*(x_{t-1})=\int_\mathcal XP_t(x_{t-1},dx_t)f_t(y_t|x_t)
-\end{equation} and is called \textit{perfectly adapted} auxiliary
-function. In fact, if \(M_t=M_t^*\), then, expression
-\eqref{weight_updated_apf} becomes: \begin{equation*}
-    \begin{split}
-        w_t\propto\frac{w_{t-1}}{ \tilde w_{t-1}}\cdot\frac{P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)}{ M_t^*(x_{t-1},dx_t)}=\frac{1}{\eta_{t-1} (x_{t-1})}\cdot\int_\mathcal XP_t(x_{t-1},dx_t)f_t(y_t|x_t)
-    \end{split}
-\end{equation*} From the previous equation, it is clear that if
-\(\eta_t=\eta_t^*\), the inferential weights are constant and their
-variance is zero.\\
-We can summarize the APF algorithm.
-
-\begin{enumerate}
-    \item \textbf{Inital stage: }
-    \begin{enumerate}
-        \item For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb M_0(dx_0)$ and set the un-normalized inferential weights: 
-    $$w_0^n=\frac{\mathbb P_0(dx_0)}{\mathbb M_0(dx_0)}.$$
-    Then, the normalized inferential weights are: $W_0^n:=\frac{w_0^n}{\sum_{m=1}^Nw_0^m}$.
-    \item Compute the un-normalized auxiliary weights:
-        $$\tilde w_0^n=w_0^n \cdot \eta_0(X_0^n)$$
-        Then the normalized auxiliary weights are:
-        $\tilde W_0^n:=\frac{\tilde w_0^n}{\sum_{m=1}^N\tilde w_0^m}$.
-    \end{enumerate}
-    \item \textbf{For every $t=1,...T$:}
-    \begin{enumerate}
-        \item \textbf{Resample} if $ESS( W_{t-1}^{1:N})<ESS_{min}$:
-\begin{enumerate}
-    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(\tilde W_{t-1}^n)_n$.
-    \item Update weights as: $\hat w_{t-1}^n = \frac{W_{t-1}^{A_t^n}}{\tilde W_{t-1}^{A_t^n}} $ for $n=1,...,N$.
-\end{enumerate}
-\textbf{Otherwise, } 
-\begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$.
-    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
-\end{enumerate}
-\item Sample the future states from the transition kernel: $X_t^n\thicksim M_t(X_{t-1}^{A_t^n},dx_t)$ for $n=1,...,N$.
-\item Update the un-normalized inferential weights, for $n=1,...,N$:
-\begin{equation*}
-    w_t^n=\hat w_{t-1}^n\cdot \frac{P_t(X_{t-1}^{A_t^n},dx_t)\cdot f_t(y_t|X_t^n)}{ M_t(X_{t-1}^{A_t^n},dx_t)}.
-\end{equation*}
-\item Normalize the inferential weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
-\item Compute the un-normalized auxiliary weights, for $n=1,...,N$:
-\begin{equation*}
- \tilde w_t^n=w_t^n \cdot \eta_t(X_t^n)
-\end{equation*}
-\item Normalize the auxiliary weights, for $n=1,...,N$:  $\tilde W_t^n:=\frac{\tilde w_t^n}{\sum_{m=1}^N\tilde w_t^m}$.\footnote{It is worth mentioning that the original APF included a final resampling step at the end of each iteration.
-However, our formulation is more efficient since having two resampling steps might introduce additional Monte Carlo variance.
-Se ***Johansen p.
-120}
-    \end{enumerate}
-\end{enumerate}
-
-However, it is worth noting that the optimality results both for the
-kernel in the GPF and of the auxiliary in the APF, has two main
-weaknesses:
-
-\begin{itemize}
-    \item it is not clear that what is optimal at time $t$ remains such at future stages.
-In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affect the efficiency of the PF estimators;
-    \item both the choice of $M_t$ and of $\eta_t$ affect the way in which weights are computed and, indirectly, the decision to resample in the future.
-It is difficult to take also this indirect effect into account when evaluating the  overall efficiency of the algorithm.
-\end{itemize}
+\textbf{Implementation}
 
 For illustration purposes, we are going implement an auxiliary particle
-filter with auxiliary function
-\(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).\\
+filter with auxiliary function \(g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}\).
 
@@
 \hrule
-\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}
+
+\hfill\break
+\texttt{APFfun(data,N,m0,C0,tau,sigma,r)}\\
+
 \hrule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-26-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-23-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-26}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-23}
 \end{figure}
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-27-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-24-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-27}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-24}
 \end{figure}
@@
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-29}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-26}Root Mean Square Errors}\\
 \toprule
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/unnamed-chunk-32-1}
+\begin{figure}[ht]
+
+{\centering \includegraphics{draft-implementation-2_files/figure-latex/unnamed-chunk-29-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-29}
 \end{figure}
 
-\section{Liu and West Filter}\label{pf_lui_west}
-
-Let us consider a more general state-space model, where the state vector
-includes a time-constant parameter vector \(\psi \in \Psi\), where
-\(\Psi\) is the parameter space. Since we can interpret the parameter as
-a state with the law of motion \(\psi_t=\psi_{t-1}=\psi\), then it is
-possible to adopt a PF algorithm to estimate \(\psi\).\\
-However, being \(\psi\) constant over time, it is meaningless to sample
-its values sequentially: the first extraction of \(\psi\) at the initial
-stage 0 (i.e., \(\psi_0^n\)) generates a constant path of particles for
-the parameter (\(\psi_t^n=\psi_0^n\)).
-\footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\\
-Therefore, Liu and West (2001) proposed a modified particle filter,
-called \textit{Liu and West filter} (LWF) that allows to resample the
-parameter over time from a continuous distribution. In this way, at
-every time, the support of \(\psi\) is not limited to the initially
-sampled \(N\) values.\\
-Here, we describe the LWF that makes use of the normal distribution to
-sample sequentially the parameter. Moreover, consider the simpler
-version of the bootstrap
-filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\\
-Let the transition kernel and the likelihood depend on the parameter
-\(\psi\) (i.e., \(P_t(x_{t-1},dx_t;\psi)\) and \(f_t(y_t|x_t;\psi)\))
-and let \$\pi(\cdot) \$ be the prior distribution for the parameter
-\(\psi\).\\
-The LWF can be described by the following algorithm.
-
-\begin{enumerate}
-    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\mathbb P_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.
-
-    Then, set the un-normalized weights $w_0^n=1$.
-
-    Hence, the normalized weights are: $W_0^n=\frac{1}{N}$.
-    \item \textbf{For every $t=1,...T$:}
-    \begin{enumerate}
-        \item \textbf{Resample} if $ESS(W_{t-1}^{1:N})<ESS_{min}$:
-\begin{enumerate}
-    \item Multinomial sampling for the indexes $(A_t^n)_n$ with probabilities given by $(W_{t-1}^n)_{n}$.
-    \item Update weights as: $\hat w_{t-1}^n = 1$ for $n=1,...,N$.
-\end{enumerate}
-\textbf{Otherwise, } 
-\begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$
-    \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
-\end{enumerate}
-\item \textbf{Extracting new parameter values:}
-\begin{enumerate}
-    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)^2$.
-    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$.
-    \item Draw new parameters: $\psi^n\thicksim \mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.
-\end{enumerate}
-\item Sample future states from the transition kernel: $X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)$ for $n=1,...,N$.
-\item Update the un-normalized weights through the likelihood function, for $n=1,...,N$:
-\begin{equation*}
-    w_t^n=\hat w_{t-1}^n\cdot f_t(y_t|X_t^n;\psi^n).
-\end{equation*}
-\item Normalize the weights, for $n=1,...,N$:  $W_t^n:=\frac{w_t^n}{\sum_{m=1}^Nw_t^m}$.
-    \end{enumerate}
-\end{enumerate}
-
-At the initial step, \(N\) states and \(N\) parameter values are sampled
-independently, respectively, from \(\mathbb P_0\) and \(\pi\) and the
-particle weights are set to be constant, as usual.\\
-Then, at each iteration (time \(t\)), the decision to resample is done
-according to the \(ESS\) criterion and indexes are extracted from a
-multinomial distribution. However, in the LWF both previous periods
-states and parameters are resampled.\\
-Meanwhile, to extract the next period states and parameters it is
-necessary, firstly, to sample the new \(N\) parameter values. Let
-\(\bar \psi\) be the weighted average of the parameter values and
-\(\Omega\) be their variance (the weights used to compute this mean and
-variance are the previous period inferential ones,
-i.e.~\(W_{t-1}^n\)).~\\
-The algorithm suggests to sample new values for the parameters from a
-continuous distribution, i.e., a Normal centered at the (resampled)
-values \(m^{A_t^n}\) with variance \(h^2\cdot \Omega\) where \(m^n\) is
-a convex combination of the sample mean \(\bar\psi\) and the actual
-extraction of \(\psi^n\) with coefficient \(a\in (0,1)\) and \(h\) is a
-scalar such that \(a^2+h^2=1\). While it would be more intuitively
-immediate to sample new parameters from a Normal density centered at the
-previously sampled values (i.e., \(\psi^n\)), this would increase the
-variance of \(\psi\) over time.\\
-In fact, let
-\(\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N(\psi^{A_t^{n}},\Omega)\)\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
-be the empirical distribution of the parameters. From the law of
-iterated expectation and total variance, we
-obtain:\footnote{For completeness, we specify that the first expected value in the law of iterated expectations and total variances is with respect to the distribution of the indexes and the second one with respect to each of the Normal distributions centered at $\psi^{A_t^{n}}$.}
-\begin{equation*}
-    \begin{split}
-        \mathbb E_{\pi_{t-1}^N} ( \psi)=& \mathbb E\big[\mathbb E( \psi|A_t^{n})\big]=\mathbb E\big[\psi^{A_t^{n}}\big]=\bar \psi\\
-        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\mathbb E( \psi|A_t^{n})\big]+\mathbb E\big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\mathbb E\big[\Omega\big]=2\cdot\Omega
-    \end{split}
-\end{equation*} Hence, while the mean is preserved, the variance
-increases at each iteration. Instead, sampling from the distribution
-proposed by Liu and West
-\(\mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)\) preserves both the
-mean and the variance of the parameters over time.\\
-Let
-\(\tilde \pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)\).
-Hence: \begin{equation*}
-    \begin{split}
-        \mathbb E_{\tilde \pi_{t-1}^N} ( \psi)=& \mathbb E\big[\mathbb E( \psi|A_t^{n})\big]=\mathbb E\big[m^{A_t^n}\big]=\mathbb E\big[a\cdot \psi^{A_t^n}+(1-a)\cdot \bar \psi\big]=\bar \psi\\
-        Var_{\tilde \pi_{t-1}^N} ( \psi)=&Var\big[\mathbb E( \psi|A_t^{n})\big]+\mathbb E\big[Var( \psi|A_t^{n})\big]=Var\big[m^{A_t^n}\big]+h^2\cdot\mathbb E\big[\Omega\big]=(\underbrace{a^2+h^2}_{=1})\cdot \Omega =\Omega
-    \end{split}
-\end{equation*} Finally, after sampling the new \(N\) parameter values,
-conditional on these values, new states are also sampled from the
-transition kernel: \(X_t^n\thicksim P_t(X_{t-1}^{A_t^n},dx_t; \psi^n)\).
-Lastly, weights are updated through the incremental weights.~\\
-To conclude, it is worth noting that the LWF produces, together with the
-usual filtering and prediction distributions, a posterior distribution
-for the parameter: \begin{equation}
-    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(\psi)
-\end{equation}
+\hfill\break
 
@@
 
-\section{Convergence}\label{pf_converg}
-
-\chapter{Applications to Stochastic Volatility Models}
-\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}
-
-In this section, we apply the particle filter in a model-comparison
-exercise, ultimately underlining the importance of taking stochastic
-volatility into account in modeling equity returns. We follow XX,
-comparing density forecasts through relative predictive and cumulative
-likelihoods.
-
-We analyze a dataset of log close-to-close returns of the S\&P500, from
-2017 to 2021. Consider a stochastic volatility model \[
-\mathcal{M}_{SV}:\begin{array}{lc}
-r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
-\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
-\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
-\]
-
-and a constant volatility iid model \[
-\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
-\]
-
-We calibrate the model parameters using the whole sample, and proceed by
-setting the return mean of both models to equal the sample mean,
-\(\sigma^{2}\) equal to the return variance, and the parameters of the
-volatility equation in \(\mathcal{M}_{SV}\) by running an \(AR(1)\)
-regression of the log squared residuals.
-
-In order to compare the predictive performance of the two, we compare
-the cumulative log likelihood of the observed returns up to time \(t\)
-given the considered model. In formulas, the cumulative likelihood can
-be computed as \[
-p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
-\]
-
-adopting the convention \(y_{1:0}=\emptyset\). This is trivial to obtain
-in the iid case; considering the stohastic volatility model, the
-predictive likelihoods can be derived as
-
-\[
-p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
-\] \[
-=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
-\]
-
-Whereas the emission distribution
-\(p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})\) is known, we approximate
-the predictive state distribution
-\(p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})\) with weighted particles
-obtained with particle filtering.
-
-Figure XX reports the time series of the cumulative log-likelihood given
-the constant volatility model \(\mathcal{M}_{CV}\) relative to the one
-given \(\mathcal{M}_{SV}\), obtained as
-\(\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})\).
-Notice that the first difference of this series is the relative
-log-likelihood of the incoming observation,
-\(p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})\).
-
-\begin{figure}[H]
-
-{\centering \includegraphics{final-draft_files/figure-latex/cll_plot-1}
-
-}
-
-\caption{Relative log-likelihood}\label{fig:cll_plot}
-\end{figure}
-
-Overall, this analysis shows how a model with stochastic volatility is
-better at describing and predicting the considered time-frame of S\&P500
-returns. As can be seen from the graph and perhaps unsurprisingly, the
-density forecast given the stochastic volatility model represents a
-particularly significant improvement in the period of extreme values of
-returns in March 2020.
-
-\backmatter
 \end{document}

### 2021-07-30 new-draft.Rmd
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

--- 2021-07-19 new draft.Rmd
@@
 library(kableExtra)
+tinytex::tlmgr_install() 
 ```

### 2021-07-30 nuova-final-draft.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

--- 2021-07-30 backup1.tex
@@
 \hypersetup{
-  pdftitle={State-Space Models and Particle Filters},
-  pdfauthor={Edoardo Marcelli; Roberto Colarieti; Alexei Verkhovtsev; Michele Bolla; Gabriele Romano; Benedetta Bruni},
+  pdftitle={State-Space Models and Sequential Monte Carlo methods},
+  pdfauthor={Edoardo Marcelli; Roberto Colarieti; Alex; Michele Bolla; Gabriele Romano; Benedetta Bruni},
   hidelinks,
@@
 
-\title{\Huge State-Space Models and Particle Filters}
-\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei
-Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta
-Bruni}
+\title{\Huge State-Space Models and Sequential Monte Carlo methods}
+\author{Edoardo Marcelli \and Roberto Colarieti \and Alex \and Michele
+Bolla \and Gabriele Romano \and Benedetta Bruni}
 \date{21 luglio, 2021}
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-5-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-5-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-7-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-7-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-10-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-10-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-13-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-13-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-16-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-16-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-19-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-19-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-22-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-22-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-25-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-25-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-26-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-26-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-32-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-32-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-35-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/unnamed-chunk-35-1}
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/cll_plot-1}
+{\centering \includegraphics{nuova-final-draft_files/figure-latex/cll_plot-1}
 

### 2021-07-30 prova_knit_finale.tex
merged
Merge branch 'main' of https://github.com/edoardo-marcelli/Progetto-Bayesian

# Conflicts:
#	prova_knit_finale.log
#	prova_knit_finale.pdf

--- 2021-07-30 backup1.tex
@@
 \def\code#1{\texttt{#1}}
+\usepackage{babel}
 \usepackage{amssymb}
+\usepackage{amsfonts}
 \usepackage{txfonts}
 \usepackage{amsmath}
+\usepackage{mathtools}
 \usepackage[bindingoffset=1.5cm, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
@@
 \urlstyle{same}
+\usepackage[toc,page]{appendix}
 \usepackage{comment}
@@
 
-\subsection{Implementation}
-
-In this section we present a practical illustration of the algorithms
-discussed. For the sake of simplicity we use a random walk plus noise
-model, i.e.~the most basic form of a linear Gaussian state-space
-model.\\
+\subsection{Implementation}\label{implementation}
+
+The aim of this subsection is to provide a practical illustration of the
+algorithms discussed in the main chapters. For the sake of simplicity we
+use for implementations a random walk plus noise model, i.e.~the most
+basic form of linear Gaussian state-space model.\\
 \begin{align}
-y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \\
-x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \\
+y_{t}|x_{t} & \sim N(x_{t},\sigma^{2}) \label{eq:1} \\
+x_{t}|x_{t-1} & \sim N(x_{t-1},\tau^{2}) \label{eq:2} \\
 x_{0} & \sim N(m_{0},C_{0})
@@
 complex models.\\
-The filtering strategy is applied to 50 simulated data. Figure XX shows
-the simulated true states sequence assuming as data generating process
-the Equation (2) with \(\tau^{2}=1\) and simulated observed sequence
-process form Equation (1) with \(\sigma^{2}=1\).
+The filtering strategy is applied to 50 simulated data. Figure
+\ref{fig:myfig} shows the simulated true states sequence assuming as
+data generating process the Equation \eqref{eq:2} with \(\tau^{2}=1\)
+and simulated observed sequence process form Equation \eqref{eq:1} with
+\(\sigma^{2}=1\).
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-5-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig-1}
 
@@
 
-\caption{Simulated random walk plus noise model}\label{fig:unnamed-chunk-5}
+\caption{ Simulated random walk plus noise model}\label{fig:myfig}
 \end{figure}
@@
 univariate time series.\\
-\texttt{sig2} ~~the variance \(\sigma^{2}\) in Equation (1)\\
-\texttt{tau2} ~~the variance \(\tau^{2}\) in Equation (2)\\
+\texttt{sig2} ~~the variance \(\sigma^{2}\) in observation equation\\
+\texttt{tau2} ~~the variance \(\tau^{2}\) in state equation\\
 \texttt{m0} ~~central value of the normal prior state distribution\\
@@
 
-In Figure XX below filtered states estimated using Kalman Filter with
-\(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are compared to
-the true states values. Notice how closely the filtered states follow
-the observations and the goodness of the approximation of the true
-states. 95 percent credible intervals are computed as
+In Figure \ref{fig:myfig2} below filtered states estimated using Kalman
+Filter with \(x_{0} \sim N(0,100)\) and \(\sigma^{2}=\tau^{2}=1\) are
+compared to the true states values. Notice how closely the filtered
+states follow the observations and the goodness of the approximation of
+the true states. 95 percent credible intervals are computed as
 \[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-7-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig2-1}
 
@@
 
-\caption{Kalman Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-7}
+\caption{Kalman Filtered States with credible interval (in red)}\label{fig:myfig2}
 \end{figure}
 
-The discussion of Kalman Filter will continue in Section XX where we
-compare it to the Particle Filter.
+The discussion of Kalman Filter will continue in the next sections where
+we compare it to the Particle Filter.
 
@@
 
-Consider again the random walk plus noise model of section XX,the
-challenge here is to estimate the filtered states using a Sequential
-Importance Sampling algorithm. The main idea of the SIS applied to this
-univariate linear gaussian model is described in the following
-algorithm. We indicate with \(n\) the sample size of time observations
-and with \(N\) the generated sample size for each step of the Sequenial
-Monte Carlo.
+Consider again the random walk plus noise model of section
+\ref{implementation},the challenge here is to estimate the filtered
+states using a Sequential Importance Sampling algorithm. The main idea
+of the SIS applied to this univariate linear gaussian model is described
+in the following algorithm. We indicate with \(n\) the sample size of
+time observations and with \(N\) the generated sample size for each step
+of the Sequenial Monte Carlo.
 
@@
 \texttt{C0} ~~variance of the normal prior state distribution\\
-\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
-\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)
+\texttt{tau} ~~the standard deviation \(\tau\) in state equation\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in observation
+equation
 
@@
   \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }\CommentTok{\#sample from N(x\_\{t{-}1\},tau)}
-\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }\CommentTok{\#update weight}
+\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }
+\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }
 \NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
@@
 
-\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }\CommentTok{\#normalized weight}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }\CommentTok{\#effective sample size}
+\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-10-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-8-1}
 
@@
 
-\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-10}
+\caption{a) SIS Filtered States with credible interval (in red). b) Effective sample size.}\label{fig:unnamed-chunk-8}
 \end{figure}
@@
 In this section, Bootstrap Particle Filter will be used to estimate
-filtered states of the Random Walk plus Noise introduced in section XX.
-The overall strategy replicate the SIS filter with the addition of a
-ESS-based resampling step that applies when the effective sample size is
-smaller than a predetermined threshold opportunely chosen (in our
-example we decide to follow a common rule of thumb consisting in setting
-the threshold at \(N/2\)). The steps are presented in the following
-algorithm.
+filtered states of the Random Walk plus Noise introduced in section
+\ref{implementation}. The overall strategy replicate the SIS filter with
+the addition of a ESS-based resampling step that applies when the
+effective sample size is smaller than a predetermined threshold
+opportunely chosen (in our example we decide to follow a common rule of
+thumb consisting in setting the threshold at \(N/2\)). The steps are
+presented in the following algorithm.
 
@@
 \texttt{C0} ~~variance of the normal prior state distribution\\
-\texttt{tau} ~~the standard deviation \(\tau\) in Equation (2)\\
-\texttt{sigma} ~~the standard deviation \(\sigma\) in Equation (1)\\
+\texttt{tau} ~~the standard deviation \(\tau\) in state equation\\
+\texttt{sigma} ~~the standard deviation \(\sigma\) in observation
+equation\\
 \texttt{r} ~~ if present the threshold is set equal to \(N/r\)
@@
 The estimated states of the Boostrap Particle Filter togheter with the
-effective sample size are shown in Figure XX. We decide to set
-\(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice how
-the resampling step allows the effective sample size not to drop,
+effective sample size are shown in Figure \ref{fig:myfig3}. We decide to
+set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice
+how the resampling step allows the effective sample size not to drop,
 improving results.
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-13-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig3-1}
 
@@
 
-\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-13}
+\caption{a) PF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:myfig3}
 \end{figure}
@@
 we want to compare the results of Boostrap Particle Filter(BPF) with
-Kalman Filter(KF). As we can see from figure XX, when the number of
-particles generated at any interaction increases the results for both
-the estimated mean and the variance tend to converge {[}{[}SPIEGARE
-PERCHE'{]}{]}. Moreover, comparing the Root Mean Square Errors of the
-filtered states with respect to the true state values, when the sample
-size increases the BPF decreases and, for large N, it reaches the
-accuracy of the Kalman Filter.
+Kalman Filter(KF). As we can see from figure \ref{fig:myfig4}, when the
+number of particles generated at any interaction increases the results
+for both the estimated mean and the variance tend to converge
+{[}{[}SPIEGARE PERCHE'{]}{]}. Moreover, comparing the Root Mean Square
+Errors of the filtered states with respect to the true state values,
+when the sample size increases the BPF decreases and, for large N, it
+reaches the accuracy of the Kalman Filter.
 
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-15}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-12}Root Mean Square Errors}\\
 \toprule
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-16-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig4-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:unnamed-chunk-16}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Kalman Filter (KF) for increasing number of generated particles (N)}\label{fig:myfig4}
 \end{figure}
@@
 The Bootstrap Particle Filter Approach for the Random Walk plus Noise
-Model described in Section XX can be improved accounting for the
-observations in the importance transition density and it consists of
-generating \(x_{t}\) from its conditional distribution given \(x_{t-1}\)
-and \(y_{t}\). In the Normal model we are considering, the optimal
-proposal will be a Normal density as well with mean and variance given
-by \begin{align*}
+Model described in Section \ref{implementation} can be improved
+accounting for the observations in the importance transition density and
+it consists of generating \(x_{t}\) from its conditional distribution
+given \(x_{t-1}\) and \(y_{t}\). In the Normal model we are considering,
+the optimal proposal will be a Normal density as well with mean and
+variance given by \begin{align*}
 \mu_{opt}=E(x_{t}|x_{t-1},y_{t})&=x_{t-1}+\frac{\tau^{2}}{\tau^{2}+\sigma^{2}}(y_{t}-x_{t-1})\\
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-19-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-15-1}
 
@@
 
-\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-19}
+\caption{a) GPF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-15}
 \end{figure}
@@
 Filter (BPF) and the Guided Particle Filter (GPF). As we can see from
-the Figure XX, the Guided Particle Filter provides point estimates that
-are slightly better with respect to the ones of the BPF, and this is
-confirmed by the Table showing the RMSE. Moreover, also the variance is
-better, suggesting for higher precision.
+the Figure \ref{fig:myfig5}, the Guided Particle Filter provides point
+estimates that are slightly better with respect to the ones of the BPF,
+and this is confirmed by the Table showing the RMSE. Moreover, also the
+variance is better, suggesting for higher precision.
 
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-21}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-17}Root Mean Square Errors}\\
 \toprule
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-22-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig5-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-22}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:myfig5}
 \end{figure}
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-25-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-20-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-25}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-20}
 \end{figure}
@@
 Let's compare the Auxiliary Particle Filter (APF) and the Bootstrap
-Particle Filter (BPF).
+Particle Filter (BPF). From Figure \ref{fig:myfig6} it is difficult to
+say which of the two is better and the tables does not help much since
+the APF apparently outperforms the BPF in two cases over three. For this
+reason we can conclude that for this model the Auxiliary Particle
+Filtering strategy does not provide any serious improvement to a the
+basic one.
 
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-26-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/myfig6-1}
 
@@
 
-\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:unnamed-chunk-26}
+\caption{Comparison Bootstrap Particle Filter(BPF) and Guided Particle Filter (GPF), number of generated particles N=1000}\label{fig:myfig6}
 \end{figure}
@@
 \begin{longtable}[t]{cccc}
-\caption{\label{tab:unnamed-chunk-28}Root Mean Square Errors}\\
+\caption{\label{tab:unnamed-chunk-22}Root Mean Square Errors}\\
 \toprule
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-32-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-26-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-32}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-26}
 \end{figure}
@@
 \hfill\break
-Consider the linear Gaussian example of section XX, but this time with
-unknown variances \(\tau^{2}\) and \(\sigma^{2}\). Thus, let
-\(\psi=(\sigma^{2},\tau^{2})\) be the unknown parameter vector and
-assign a gamma prior for its components, \begin{align*}
+Consider the linear Gaussian example of section \ref{implementation},
+but this time with unknown variances \(\tau^{2}\) and \(\sigma^{2}\).
+Thus, let \(\psi=(\sigma^{2},\tau^{2})\) be the unknown parameter vector
+and assign a gamma prior for its components, \begin{align*}
 \sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/unnamed-chunk-35-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-29-1}
 
@@
 
-\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-35}
+\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-29}
 \end{figure}
@@
 \begin{longtable}[t]{cccccc}
-\caption{\label{tab:unnamed-chunk-37}RMSE}\\
+\caption{\label{tab:unnamed-chunk-31}RMSE}\\
 \toprule
@@
 
+\section{Implementations for Stochastic Volatility Models}
+
 \chapter{Applications to Stochastic Volatility Models}
+
+\section{Overview}
+
+In the following sections, we will try to replicate some of the analyses
+which were previously conducted on synthetic data, generated by a basic
+stochastic volatility model; this time, the same model specification and
+filters will be used on real data. However, as we will see, real data
+require different strategies to evaluate the effectiveness of several
+Sequential Monte Carlo techniques. Intuitively, such differences are due
+to the fact that now the latent process driving the Hidden Markov Model
+is not observed, nor do we know exactly its law of motion.\\
+This part of the analysis is structured as follows: section
+\textit{fill with the number of the section} provides a review of recent
+research in the stochastic volatility literature, including extensions
+of the basic specification and the use of particle filters; section
+\textit{fill with the number of the section} describes the data which we
+employ and the transformations that were applied to them. In section
+\textit{fill with the number of the section} we briefly compare the
+series of filtered states that are obtained from different particle
+filters; moreover, several PFs are compared in terms of computational
+cost in approximating the exact filtering distribution implied by our SV
+model. Section \textit{fill with the number of the section} features a
+forecast comparison between two SV specifications and a model that
+assumes constant mean and variance for the financial returns, namely a
+forecasting exercise that would not be possible were it not for particle
+filtering techniques. In section
+\textit{fill with the number of the section} we conclude by employing a
+non-parametric measure of realised volatility as a proxy for the latent
+volatility process, allowing a RMSE and MAE comparison between filters
+which is similar to that of section
+\textit{fill with the number of the section} for synthetic data.
+
+\section{A review of Stochastic Volatility models}
+
+In section \textit{fill with the number of the section} a simple
+specification of a Stochastic Volatility model was presented and used as
+the data generating process of a simulated dataset. Denoting by \(y_t\)
+the log returns at time \(t\) and by \(x_t\) the latent volatility
+stochastic process, such state-space model was presented as follows,
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
+
+This specification - which we will employ in this section on a real
+dataset - can be deemed as the standard version of the SV model.
+Throughout the recent decades, several extensions have been proposed,
+mostly (but not exclusively) in the field of financial econometrics. On
+the one hand, for simplicity we will stick to the standard version, one
+that will suffice for the scope of this work on Sequential Monte Carlo
+methods. On the other hand, it is worth mentioning some of these
+extensions: indeed, one can rather safely assume that, through their
+higher degree of sophistication, such extensions might well improve on
+the performance of our standard SV model in the analyses that will
+follow.\\
+~\\
+Stochastic Volatility Models, whose early formulation is commonly
+attributed to Taylor (1982, 1986), allow to account for time-varying and
+autoregressive volatility in financial returns, posing themselves as a
+valid alternative to ARCH (Engle 1982) or GARCH (Bollerslev 1986) models
+in dealing with non-constant volatility. Kim, Shephard and Chib (1998)
+define the canonical model for regularly spaced data as \begin{align*}
+    y_t & = \psi e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\tau \eta_t\\
+    x_1|\sigma,\beta & \sim  N\Big(\mu,\frac{\sigma^2}{1-\beta^2}\Big)\\
+    \epsilon_t & \sim N(0,1)\\ 
+    \eta_t & \sim N(0,1)
+\end{align*} where the log volatility \(x_t\) is assumed to follow a
+stationary process (\(|\beta|<1\)), \(\psi\) is a constant scaling
+factor, \(\beta\) is the persistence in the volatility and \(\tau\) is
+the volatility of the
+log-volatility.\footnote{For identifiability reasons, either $\psi=1$ or $\mu=0$. Kim, Shephard and Chib (1998) prefer $\psi=1$.}
+Most importantly, \(\epsilon_t\) and \(\eta_t\), the Gaussian white
+noise processes that drive the canonical model, are assumed to be
+uncorrelated. This latter assumption justifies another definition of
+this specification, namely ``discrete SV model without
+leverage''.\footnote{Note that in this review we do not focus on continuous time SV models.
+In fact, such models have attracted a considerable amount of research in financial econometrics and mathematical finance, especially after Hull and White (1987) considered stochastic volatility for option pricing.
+Arguably, the most influential model was then proposed for option pricing by Heston (1993), a SV model with leverage effects and square root diffusion driving volatility.
+Diffusion-based SV models enjoyed increasing popularity, see for example Barndorff-Nielsen and Shephard (2001), or Christoffersen et al.
+(2010), who investigate alternatives to the entrenched affine square root SV model.
+Eraker (2004) proposed a SV model with correlated jumps in prices and volatility, extending Heston's model, while Comte and Renault (1998) extended Hull and White's model as to feature long memory properties.
+More recently, Gatheral et al.
+(2018) further built on Comte and Renault's fractional SV model to propose the popular "rough volatility models" (see also Friz et al., 2021).}\\
+In order to accommodate for leverage effects, namely capture the
+increase in volatility that follows a drop in the returns, the model can
+extended as in Omori et al.~(2007), \begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\epsilon_t\\
+    x_{t+1} & = \mu+\beta(x_t-\mu)+\eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}
+\end{align*} where \(\rho<0\) captures the negative correlation. Such
+specification, which can be referred to as ``discrete SV model with
+leverage'', captures the asymmetric response of volatility to returns of
+different signs, so that similar specifications are sometimes also
+deemed as ``asymmetric SV models'' (e.g.~Harvey and Shephard 1996, Mao
+et al.~2020).\\
+Note that so far the assumptions on \(\epsilon_t\) and \(\eta_t\)
+implied that the returns are conditionally normally distributed. The
+discrete time SV models can also be extended to allow for heavy-tailed
+or asymmetric conditional returns distributions: symmetric or skewed
+Student-\(t\), Generalised Hyperbolic (GH) distribution, Generalised
+Error Distribution (GED) and scale mixtures of normals feature as
+popular choices (Kim and Stoffer 2008, Nakajima and Omori 2012, Mao et
+al.~2020). In fact, SV with heavy tailed return distributions were shown
+to better meet empirical regularities like the leptokurtic distribution
+of the returns and slowly decaying autocorrelation functions of the
+squared returns (Liesenfeld and Jung 2000).\\
+Assuming that \(\epsilon_t\) follows a Student-\(t\) distribution, and
+exploiting the fact that \(\epsilon_t\) can then be written as
+\(\lambda_t^{-1/2}\zeta_t\), where \(\zeta_t \sim N(0,1)\) and
+\(v\lambda_t \sim \chi^2_v\) (Harvey et al.~1994, Chib et al.~2002), we
+have the following SV model with both fat tails and leverage effect
+(Jacquier et al., 2004), \begin{align*}
+    y_t & = e^{\frac{x_t}{2}}\lambda_t^{-1/2}\zeta_t\\
+    x_{t+1} & = \alpha + \beta x_t + \eta_t\\
+    \begin{pmatrix}
+       \zeta_t \\
+       \eta_t
+    \end{pmatrix} \Big|\rho,\tau & \overset{i.i.d.}{\sim}  N_2(\mathbf{0},\Sigma), \ \Sigma=
+    \begin{pmatrix}
+       1 & \rho\tau \\
+       \rho\tau & \tau^2
+    \end{pmatrix}\\
+    v\lambda_t & \sim \chi^2_v
+\end{align*} To capture other elements of the behaviour of financial
+data, several other extensions have been proposed. For example, SV
+models have been extended to include conditional heteroskedasticity in
+the mean returns (Koopman and Hol Uspensky 2002) to capture potential
+volatility feedback effects, or to feature autoregressive moving average
+innovations (Chan 2013, Zhang et al.~2020), allowing better goodness of
+fit and out-of-sample
+forecasts.\footnote{Dimitrakopoulos and Kolossiatis (2020) note that "the moving average component, the leverage effect and the conditional heteroscedasticity in mean have been considered separately in the stochastic volatility literature" and provide two specifications, one featuring an MA component and leverage effects, the other an MA component and conditional heteroskedasticity in mean.}
+\footnote{Other extensions, though ones for which we do not report the specifications, feature modelling the latent volatility process $x_t$ as an ARFIMA process (Long Memory Stochastic Volatility model, Breidt et al.
+1998) or as governed by a first-order Markov process (Markov Switching Stochastic Volatility model, So et al.
+1998).
+Recently, Luo et al.
+(2018) incorporated neural networks in the stochastic volatility model (Neural Stochastic Volatility Model), while Xu and Chen (2021) employ deep learning models (Deep Stochastic Volatility Model).}
+The model proposed by Koopman and Hol Uspensky is specified as
+\begin{align*}
+    y_t & = \nu_t+\psi e^{\frac{x_t}{2}}\epsilon_t\\
+    \nu_t & = a+by_{t-1}+d\psi^2e^{x_t}\\
+    x_t & = \beta x_{t-1}+\tau \eta_t\\
+    \begin{pmatrix}
+       \epsilon_t \\
+       \eta_t
+    \end{pmatrix} & \overset{i.i.d.}{\sim}  N_2\Big(\mathbf{0},\begin{pmatrix}
+       1 & 0 \\
+       0 & 1
+    \end{pmatrix}\Big)
+\end{align*} while the state space representation of the ARMA(p,q)-SV
+framework, as in Zhang et al.~(2020), reads \begin{align*}
+    y_t & = \nu_t + \gamma_t\\
+    \gamma_t & =\phi_1\gamma_{t-1}+...+\phi_p\gamma_{t-p}+u_t+\varphi_1 u_{t-1}+...+\varphi_q u_{t-q}\\
+    u_t|x_t & \sim N(0,e^{x_t})\\
+    x_{t} & = x_{t-1}+ \eta_t\\
+    \eta_t|\tau & \sim N(0,\tau^2)
+\end{align*} where the error terms \(u_t\) and \(\eta_t\) are
+independent across all leads and lags, while \(\nu_t\) follows an
+unspecified time-varying process.\\
+~\\
+Interestingly enough, the path along which the SV models evolved
+coincides with that suggested by optimal portfolio findings. Johannes,
+Korteweg and Polson (2014) found that, in order to generate
+statistically significant portfolio improvements in a Bayesian learning
+problem, the model employed by the investor should incorporate both
+time-varying expected returns and stochastic volatility: indeed, either
+of these features alone did not lead to statistically significant gains
+with respect to employing models with time-constant expected returns and
+volatility.\footnote{One could then argue that caution is needed when employing basic specifications of stochastic volatility models.
+For instance, Poon and Granger (2005) found that historical volatility and ARCH models both achieved better volatility forecasting performance than SV models.
+Similarly, Allen and McAleer (2020, see also Allen, 2020) found that, using realised volatility as benchmark, neither the canonical SV model or a GARCH(1,1) specification could forecast better than a simple form of historical volatility model.}\\
+~\\
+Finally, before moving to the application of Sequential Monte Carlo
+techniques to a SV model, we conclude by going through some references
+for the Bayesian analysis proposed for such models. Starting from the
+seminal work of Jacquier et al.~(1994), the use of MCMC methods has
+become increasingly popular for parameter estimation and smoothing
+exercises in SV models (e.g.~Kastner 2019, presenting the R package
+\textit{stochvol} for Bayesian parameter estimation, and Chopin and
+Papaspiliopoulos 2020, who use MCMC to sample from the smoothing
+distribution of a SV model). As regards filtering exercises, the
+adoption of particle filters was rather
+rapid:\footnote{Actually, SV models are now often used as straightforward applications of particle filters on non-linear state space models, see for example Andrieu et al.
+(2010), Douc et al.
+(2014) or Chopin and Papaspiliopoulos (2020).}
+indeed, latent volatilities in Kim et al.~(1998) were already filtered
+by employing the particle filter suggested in Pitt and Shephard (1999),
+paving the way for subsequent applications (\textit{inter alia}) in Chib
+et al.~(2006), Omori et al.~(2007), Kim and Stoffer (2008) and Nakajima
+and Omori (2012).\\
+
+\section{Data Description}
+
+As previously mentioned, we will analyze the behaviour of the described
+filtering tools associated to a simple stochastic volatility model. In
+particular, we are evaluating the performance of such model on real
+data. For the observable process, namely financial returns in the SV
+model, we consider the continuously compounded daily returns (also
+called logarithmic returns) of three indices, S\&P500, DOW JONES and
+STOXX50, in a time interval from June 1st, 2017 to May 30th, 2021. From
+these, we estimate the daily volatility, as computed by the model.
+
+The indices have been selected as representatives of the global economic
+trends in the US and EU markets. Specifically, the S\&P500 is a
+market-capitalization-weighted stock-price index tracing the performance
+of the 500 largest companies listed on US stock exchanges (NYSE and
+Nasdaq Exchange). The DOW JONES, instead, is a price-weighted
+stock-market index and accounts for the 30 major companies listed on US
+stock exchanges, characterized for being ``blue-chip''. Also the EURO
+STOXX 50 follows bluce-chip stocks representing leading firms in regions
+of the Eurozone.\\
+
+The three plots below represent the time series of the log returns
+calculated for the three indices during our period of interest. In
+particular, each series is a sequence of daily observations representing
+the logarithm of the ratio between the closing price of the index for a
+given day and the closing price of the day before.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-32-1}
+
+}
+
+\caption{Close to close percentage log returns, 3 indices}\label{fig:unnamed-chunk-32}
+\end{figure}
+
+At a first visual inspection, we can see a common behaviour in the
+volatility of all three indices. Especially, it is worth pointing out
+the similarity between the pattern of DJI and that of S\&P500
+(especially in terms of peaks), which both differ from the STOXX50E,
+presumably because the first two describe the US stock market, while the
+last index describes the European one. For example, a difference that
+can be seen by having a glance at the plots is that the US indices log
+returns displays differences in fluctuation magnitude across different
+periods that are more marked than for the EU index, for which log
+returns display, in general, wider fluctuations. Anyways, what stands
+out the most in all three plots is the very wide fluctuations present
+from February - March 2020 to around September 2020, which, very
+intuitively, are connected with the Covid-19 Pandemic Crisis.
+
+\begin{longtable}[t]{lccc}
+\caption{\label{tab:unnamed-chunk-33}Summary statistics, close to close percentage log returns}\\
+\toprule
+  & Dow Jones & S\&P500 & STOXX50E\\
+\midrule
+Mean & 0.05581 & 0.05002 & 0.01229\\
+Standard Deviation & 1.32564 & 1.38510 & 1.22951\\
+\bottomrule
+\end{longtable}
+
+As regards the series that we will employ in section
+\textit{fill with the number of the section}, namely the series of
+realised volatilities which we will employ as a proxy for the true
+latent volatilities, the data were retrieved from the Oxford-Man
+Institute's Realized Library, which provides several different daily
+non-parametric measures of past
+volatility.\footnote{Such class of measures will be better presented in section \textit{fill with the number of the section}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.}
+We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.}
+model-free daily volatility estimates based on 5 min intraday return
+intervals. Also in this case, the chosen interval spanned from June 1st,
+2017 to May 30th, 2021. The plots below represent the realized
+volatility in the period of interest for all the three
+indices.\footnote{In the graph, the series $rv5$ is rescaled as specified in the upcoming section: we took the square root of each value, which is a daily variance measure, and multiplied it by 100.}
+As usual, they show a rather similar pattern across all three indices,
+identifying peaks corresponding to periods of higher uncertainty (in
+particular referring to the 2020 Covid-19 Crisis outbreak and its impact
+over time). Once more, there is great similarity between the series for
+the DJI and the one for the S\&P500, while the series of the STOXX50E
+appears slightly different, in terms of peaks.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-34-1}
+
+}
+
+\caption{Rescaled realised daily volatility, 3 indices}\label{fig:unnamed-chunk-34}
+\end{figure}
+
+\subsection{Rescaling the data}
+
+When specifying the parameters of a stochastic volatility model, one
+should clarify whether the series of observed financial returns is one
+of log returns or of percentage log returns. Indeed, let \(y_t\) denote
+the returns at time \(t\), so that a basic SV specification would
+prescribe \(y_t|x_t\sim N(0,e^{x_t})\): clearly, for each period \(t\),
+\(e^{x_t}\) needs to be higher to replicate the observed
+\textit{percentage} log return rather than the observed log return. This
+implies that the latent volatility \(x\) should fluctuate around higher
+values when using percentage log returns as \(y_t\). This feature, when
+adopting the aforementioned canonical SV model as defined in Kim et
+al.~(1998), would be represented by a higher \(\mu\). For example, as
+noted in Kastner (2019), daily log returns often have a variance of
+0.0001 or less, which implies that \(\mu\) should lie somewhere around
+\(\ln(0.0001)\approx -9\); instead, when considering daily percentage
+log returns, one then has a variance of 1, so that \(\mu=\ln(1)=0\). In
+other terms, one might end up with a basic specification of the SV model
+when (roughly) calibrating the parameters of the canonical model on a
+wider set of financial returns.
+
+Note that rescaling the log returns has analogous implications for the
+section in which we employ measures of realised volatility as a proxy
+for the latent volatility process. Let us denote by \(rv5_t\) the daily
+realised variance at time \(t\), which we obtain from the \textit{rv5}
+series provided in the Oxford-Man database. Now, note that in a SV
+framework one should not directly compare \(rv5_t\) to \(x_t\), as
+\(rv5_t\) can be rather seen as an estimate of the variance of the log
+returns at time \(t\), \textit{i.e.} of \(e^{x_t}\) in the SV
+model.\footnote{See for example Guidolin and Pedio (2021), who employ $rv10$, another measure of realised volatility provided in the Oxford-Man Library, as a proxy for the latent conditional variance of the daily log returns on the FTSE100 index.}
+Moreover, by construction of \(rv5_t\), the comparison should feature a
+SV model where the log returns did \textit{not} undergo the percentage
+transformation.\footnote{Indeed, this estimator is based on the sum of 5-minute intra-day squared returns: such returns are computed as the difference between the opening and closure price in an interval.}
+Thus, in order to achieve a common scale with the percentage log
+returns, we need to multiply \(rv5\) by \(100^2\). To ease the
+interpretation and representation of the comparison between differently
+approximated filtering distributions, we finally take the square root of
+the rescaled \(rv5\), so that we compare it to \(e^{x_t/2}\).
+
+\section{Comparing the filters}
+
+In this section the different particle filter techniques we discussed in
+earlier sections will be employed to approximate the filtering
+distribution that springs from a simple SV model applied to a real
+dataset. First, we provide a qualitative comparison between the filters,
+basing on visual analysis of the (approximated) filtering distributions.
+Subsequently, we move to a more quantitative approach, in order to
+assess the quality of the approximations as the number of particles
+increases. In a way, while in the first part most filters can be rather
+confidently assumed to be close to convergence, the second subsection
+will hint at possibly different computational costs across the competing
+filters.\\
+Before moving to such analyses, it is worth mentioning that not all the
+filters we employ approximate the same filtering distribution
+\(p(x_t|y_{1:t},\psi)\), where \(\psi\) is a vector containing the
+parameters of the model. Indeed, the Liu and West filter actually
+approximates the distribution of a slightly different SV model, one that
+assumes an additional hierarchical level, \begin{align*}
+y_t|x_t & \sim N(0,e^{x_t})\\
+x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)\\
+x_0 & \sim N(0,100)\\
+\alpha & \sim N(\gamma,\zeta)\\
+\beta & \sim N(\pi,\phi)\\
+\tau^2 & \sim IG(\nu/2,\lambda \nu/2)
+\end{align*} where \(\gamma\),\(\zeta\),\(\pi\),\(\phi\),\(\nu\) and
+\(\lambda\) are assumed to be
+known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\\
+
+\subsection{Visual Analysis after Filtering}
+
+We now offer a preliminary comparison of the various filtering methods,
+based on a visual analysis of graphs representing, for each index, the
+time-varying volatility of the observations, obtained through the
+application of the different filtering techniques explored in the
+theoretical analysis. Plots that are compared represent the sequence of
+percentage log returns for a given index (the observations, in gray) and
+the sequence of filtered volatilities calculated using filtered states
+(in red) obtained through a specific filtering technique indicated in
+the caption of the
+plot.\footnote{Such filtered volatilities are obtained as follows.
+For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
+Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.}\\
+
+First, a general point is that the sequence of percentage log returns
+are different across the three indices. The reasons why this holds can
+be associated with the index-construction method and the characteristics
+of the economy and industrial setting it represents (SEE SECTION X). In
+particular, as previously noticed, there are extensive similarities
+between the DOW JONES Index and the S\&P500, which can be visualized
+when looking at the percentage of log returns in the corresponding
+plots, since both are describing the US Economy. However, the STOXX50E
+differs quite extensively, especially in the different sensitivity to
+the Covid-19 Crisis and its consequences, representing the European
+Economy. Aside from this note, the considerations expressed below on the
+sequences ``filtered
+volatilities''\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.}
+for the various techniques hold almost uniformly for all the three
+indices. Therefore, in the next paragraph we will be referring to the
+set of plots pertaining to one specific index (which could be any of the
+three).\\
+
+Generally speaking, note that, the sequence of ``filtered volatilities''
+obtained from any of the filtering techniques follows roughly similar
+paths across all the plots. Unsurprisingly, the filtered volatility
+appears wider in periods corresponding to wider fluctuations of the
+percentage log returns and lower in periods corresponding to smaller
+such fluctuations. For example, all sequences of filtered volatilities
+show peaks corresponding to the higher fluctuations due to the 2020
+Covid-19 Crisis outbreak and the consequent increases in uncertainty for
+the later periods.\\
+
+When focusing the attention on specific filters, first we can discuss
+the effect of resampling. Indeed, in the plot referring to the
+Sequential Importance Sampling procedure, the sequence of filtered
+volatilities is not able to reproduce the peaks as in with other
+filters, but also is characterized by collapsing confidence intervals.
+This is because :::::::\\
+Then, if we instead look at the filtered volatilities obtained with
+resampling, we can appreciate how the filtered sequences for the
+Basic,\footnote{By this name, we refer to a particle filter that features no adaptive resampling; rather, resampling is conducted at each step of the algorithm.}
+Bootstrap, Auxiliary and Optimal Kernel Guided particle filters give
+rather analogous results, with no differences appearing at a first
+visual
+inspection.\footnote{Except for the Auxiliary Particle filtered sequence of volatilities, showing a slightly higher peak corresponding to the greatest volatility fluctuation in 2020, if compared to other filters in this subset.
+This is still valid for all three indices.}\\
+However, the sequence of filtered volatilities obtained by applying the
+Liu and West filter is clearly different from the others. Indeed, it
+shows a ``more detailed'' sequence of states (in the sense of the
+sequence of filtered volatilities being less smooth) and also a higher
+peak for the wide fluctuation corresponding to the 2020 Covid Crisis
+outbreak. Therefore, we may expect that, compared to other filtering
+techniques, it can provide a better representation of the data. This is
+because \ldots{}
+\textit{Potremmo parlare di maggiore sensibilità alle oscillazioni dei prezzi, o simili}
+
+\subsection{Quality of approximation}
+
+Since we do not have access to the exact filtering distribution of the
+SV model, when assessing how rapidly different particle filters converge
+to it, we actually need to approximate even such target distribution as
+to have a benchmark.\\
+Fortunately, convergence results about particle filters ensure that when
+the number of particles diverges, the approximation converges to the
+target
+distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
+Inoltre, va chiarito il punto della "doppia convergenza".}
+Thus, we run one of the particle filters with a very large number of
+particles, and assume that such approximation is close enough to the
+target to be used itself as the distribution the filters should tend
+to.\\
+With such a benchmark, we can then compare how different filters
+approach it as the number of particles increases. We employ the Root
+Mean Squared Error and the Mean Absolute Error to measure discrepancies
+from the benchmark. Note that the Liu and West filter can not be
+included in such an analysis: trivially, the comparison assumes that the
+competing algorithms are approximating the same target
+distribution.\footnote{For an analogous reason, we do not need to conduct this analysis on all the indices, but we can rather focus on one, the SP500.
+Indeed, what matters in this section is not how well the model fits the data, but rather how different algorithms approach the same distribution yielded by the same model on a given set of data.}
+
+We start by running the Bootstrap particle filter with N=50000, setting
+it as the benchmark. In table XX we report the RMSE and MAE measures one
+obtains by comparing the mean of the particles generated at each step by
+the Bootstrap PF with N=50000 and the
+mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles.
+Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
+Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.}
+of the particles generated by other particles filters with N equal to
+10, 100, 1000, 5000 or
+10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
+In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
+
+\begin{longtable}[t]{lcccccc}
+\caption{\label{tab:unnamed-chunk-36}RMSE and MAE, Bootstrap PF 50000 particles}\\
+\toprule
+  & N & BPF & GPFOPT & APF & BAPF & SIS\\
+\midrule
+RMSE & 10 & 0.42331 & 0.36853 & 0.54784 & 0.55959 & 0.91686\\
+RMSE & 100 & 0.15689 & 0.15094 & 0.24044 & 0.23488 & 1.10551\\
+RMSE & 1000 & 0.06901 & 0.07669 & 0.08878 & 0.10295 & 0.62505\\
+RMSE & 5000 & 0.03817 & 0.03568 & 0.05742 & 0.06110 & 0.62004\\
+RMSE & 10000 & 0.03045 & 0.02709 & 0.04296 & 0.05151 & 0.51189\\
+\addlinespace
+MAE & 10 & 0.28078 & 0.27961 & 0.43103 & 0.43779 & 0.72753\\
+MAE & 100 & 0.09658 & 0.09624 & 0.17570 & 0.18394 & 0.91799\\
+MAE & 1000 & 0.03312 & 0.03697 & 0.05539 & 0.06726 & 0.49718\\
+MAE & 5000 & 0.01889 & 0.01665 & 0.03186 & 0.03448 & 0.44753\\
+MAE & 10000 & 0.01456 & 0.01423 & 0.02449 & 0.02726 & 0.40636\\
+\bottomrule
+\end{longtable}
+
+As we can see, almost every filter shows dramatic improvement already
+when passing from 10 particles to 100 particles, while the performance
+of the SIS algorithm actually gets worse. With N=10 or N=100, the
+Bootstrap PF and the Guided PF (with optimal proposal kernel) emerge
+quite clearly as the best performing filters; however, no conclusion
+should be drawn at this stage, as the large errors seem to suggest quite
+a wide margin from improvement even for the best performing filters.\\
+As it turns out, such suggestion is confirmed by running the algorithms
+with N=1000. Passing from 100 to 1000 particles more than halves the
+RMSEs across the 5 competing filters, with even more pronounced gains in
+the MAEs; at N=1000, the Bootstrap and the Guided particle filters still
+hold their lead over the rest of the field, keeping a shrinking hedge
+over the Auxiliary PF, which shows most impressive gains when passing
+from 100 to 1000 particles. Further sizable gains for every filter,
+except for the SIS algorithm, are observed when increasing the number of
+particles to 5000, a level at which both the RMSE and the MAE agree on a
+slight hedge of the Guided and the Bootstrap filters over the Auxiliary
+and the Basic PFs.\\
+Nevertheless, we would argue that already for N=5000, but even more
+clearly for N=10000, the difference between the Bootstrap, the Guided
+and the Auxiliary particle filters is so small that none of these
+filters can be straightforwardly chosen over the others, a figure which
+visual inspection of the graphs below can confirm. Interestingly enough,
+no gains seem to arise from resampling at each step with respect to
+adaptive resampling (note that the Basic PF is preferred to the APF only
+once, by the RMSE), whereas the need of some degree of resampling is
+apparent, as shown by the awful performance of the SIS algorithm with
+any number of particles. Moreover, the RMSE and the MAE show infrequent
+disagreement, delivering a rather clear picture of the comparison.
+
+Now, one might argue that the relative performance of the filters is
+influenced by the choice of the Bootstrap filter as the benchmark
+algorithm when run with N=50000: intuitively, if the convergence phase
+were not reached, the Bootstrap filter (which did in fact come out as
+one of the best performing algorithms) would clearly find it easier to
+replicate the benchmark we
+set.\footnote{Note that all the filters, including the BT filter run with 50000 particles, shared the same seed in the $R$ software, ruling randomness out of these considerations.}\\
+As a robustness check, we thus conduct the same analysis employing the
+Auxiliary Particle Filter as the benchmark, keeping the number of
+particles set to 50000 as for the benchmark BT filter.
+
+\begin{longtable}[t]{lcccccc}
+\caption{\label{tab:unnamed-chunk-38}RMSE and MAE, Auxiliary PF 50000 particles}\\
+\toprule
+  & N & BPF & GPFOPT & APF & BAPF & SIS\\
+\midrule
+RMSE & 10 & 0.40549 & 0.34950 & 0.53237 & 0.54597 & 0.91038\\
+RMSE & 100 & 0.13510 & 0.14209 & 0.22243 & 0.22398 & 1.10156\\
+RMSE & 1000 & 0.04145 & 0.06874 & 0.07796 & 0.07692 & 0.61504\\
+RMSE & 5000 & 0.02847 & 0.04922 & 0.03931 & 0.05230 & 0.61176\\
+RMSE & 10000 & 0.03048 & 0.03337 & 0.02648 & 0.02411 & 0.50547\\
+\addlinespace
+MAE & 10 & 0.27266 & 0.26941 & 0.42151 & 0.42909 & 0.72112\\
+MAE & 100 & 0.08540 & 0.09034 & 0.16691 & 0.17484 & 0.91468\\
+MAE & 1000 & 0.02329 & 0.03532 & 0.05105 & 0.05753 & 0.48959\\
+MAE & 5000 & 0.01756 & 0.02410 & 0.02662 & 0.03275 & 0.44195\\
+MAE & 10000 & 0.01417 & 0.01718 & 0.01974 & 0.01742 & 0.40221\\
+\bottomrule
+\end{longtable}
+
+Using the APF as the benchmark does not deliver a different picture than
+the one we previously had. Leaving the SIS algorithm aside (whose
+analogous performance comes as no surprise), it is interesting to note
+how once again the Auxiliary particle filter actually fares worse than
+the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\\
+Its performance ameliorates dramatically at N=1000, while at N=5000 the
+halving of its RMSE and MAE even allows the APF to overtake the Guided
+PF, though falling short of the Bootstrap PF. At N=10000 both the Basic
+PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF,
+even though the MAE criterion still selects the Bootstrap as the closest
+to the target distribution.\\
+Interestingly, this time the performance of the Bootstrap filter worsens
+when N is increased from 5000 to 10000 (both according to RMSEs and
+MAEs, whose disagreements, once again, are overall infrequent). Such
+evidence suggests that, even at very large (yet finite) values of N,
+some difference between the approximated filtering distributions still
+persists across our particle filters.\\
+
+\begin{longtable}[t]{lcc}
+\caption{\label{tab:unnamed-chunk-40}Discrepancy between Bootstrap and Auxiliary PFs}\\
+\toprule
+  & RMSE & MAE\\
+\midrule
+N=50000 & 0.04408 & 0.01668\\
+N=100000 & 0.03576 & 0.01476\\
+\bottomrule
+\end{longtable}
+
+In fact, by comparing the two benchmark approximations with N=50000, we
+can see that, for example, the Bootstrap PF run with N=10000 is closer
+to the benchmark APF than the benchmark BT is, or that the RMSE ranks
+the APF with N=10000 as closer to the the benchmark BT than the APF with
+N=50000, mildly pointing at a theoretically-unsound difference between
+the limiting distributions of the filters. Nevertheless, now the RMSE
+and MAE do not deliver unanimous verdicts, nor the differences seem
+large enough to back extreme claims. And, conveniently, running the
+Auxiliary and the Bootstrap filters with 100000 particles testifies a
+shrinking difference between the approximations - at increasingly large
+computational costs, though.\\
+
+Overall, we should not overplay tiny
+differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison, implying different target distributions.
+For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.}
+Rather, we prefer to stress the points that emerged from both benchmark
+analyses and we can therefore support quite confidently.\\
+First, if on the one hand it is clear that some degree of resampling is
+essential, on the other we found no gains from resampling at each
+period.\\
+Second, when the number of particles is set at 5000 or more, neither
+benchmark could single out an algorithm that clearly outperformed the
+others, when restricting the analysis to procedures that feature
+adaptive resampling rules.\\
+Third, when the number of particles was set to 10, 100 or 1000, the
+Bootstrap particle filter and the Guided particle filter with optimal
+proposal kernel significantly emerged as the best performing algorithms
+for our SV model, suggesting that their computational cost is lower than
+that of the Auxiliary particle filter, which needed more particles to
+``get rolling''.\\
+Finally, some discrepancy between different approximations persisted
+even when the number of particles was set to 50000 or 100000. However,
+the narrow extent of the RMSEs and MAEs with respect to both benchmarks
+supports the idea that at N=10000 the best-performing algorithms have
+already entered the phase of convergence. Even in light of these
+markedly more computationally expensive benchmarks, we then choose
+N=10000 as the number of particles which we routinely employ in this
+empirical application.
+
 \section{Comparing Constant and Stochastic Volatility Models of Equity Returns}
@@
 
-{\centering \includegraphics{backup1_files/figure-latex/cll_plot-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/cll_plot-1}
 
@@
 
+\section{Realised volatility as benchmark}
+
+In section XX we compared several filtered distributions, briefly
+pointing out three main behaviours, namely that of the SIS algorithm,
+that of the Liu and West particle filter and that of the remaining
+filters. However, we did not have metrics that could help us clearly
+point at one behaviour as the closest to reality. In this section, we
+will employ realised volatility as one such metric.\\
+A realised volatility measure - the realised variance based on 5 min
+intraday return intervals (\(rv5\) series in the Oxford-Man Realized
+Library) - will be used as a proxy for the latent volatility process
+driving the log returns on the 3 indices presented in the earlier
+sections. As we previously mentioned, a transformation is needed as to
+compare filtered states and realised volatility, namely one turning both
+into estimates of the standard deviation of the percentage log returns.
+For each \(t\), \(rv5_t\) is then turned into \(rv5_t^{0.5}*100\), while
+\(e^{x_t/2}\) (where \(x_t\) is the filtered state) is computed by
+taking the mean of the generated particles and then applying the
+\(exp(\cdot /2)\) transformation.\\
+Throughout this section, we will deem as the best performing algorithm
+that coming closest to the proxy, either ``literally'' in a graphical
+framework or in a Root Mean Squared Error or Mean Absolute Error sense.
+Clearly enough, this approach builds on the assumption that realised
+variance is indeed a good proxy for latent volatility in real financial
+series.\\
+
+A considerable amount of research in the 1990s and 2000s has focused on
+the properties of realised volatility measures, let them be realised
+variances or realised kernels. A thorough review of such literature can
+be found on the Oxford-Man Institute Realized volatility website, as
+their database specifically focuses these two
+measures.\footnote{Here we would mention the work by Barndorff‐Nielsen and Shephard (2002), who focus on the asymptotic properties of the realised volatility error in stochastic volatility models, namely the difference between realized volatility and the "discretised integrated volatility", also called "actual volatility".}\\
+As Shephard and Sheppard (2010) themselves put it, in a paper that also
+presents the methodology that backs their Library, ``such statistics are
+based on a variety of theoretically sound non-parametric estimators of
+the daily variation of prices''. Thanks to their theoretical
+justification and their being model-free estimates, in recent years
+realised volatility measures have been frequently used in financial
+applications as proxies for different types of latent volatility,
+providing a valuable (and arguably) preferable alternative to using
+squared returns (Hansen and Lunde,
+2006).\footnote{See for example Kambouroudis et al. (2016), Buncic and Gisler (2016), Gatheral et al.(2018), Allen and McAleer (2020), Guidolin and Pedio (2021).}\\
+Nevertheless, we shall not indulge into forgetting the inherent
+limitations of using a proxy for a latent process. In particular,
+realised volatility should not be seen as a perfect proxy of latent
+volatility (as, for example, realised measures ignore overnight price
+variations), but rather as a comparatively solid one. Therefore, in this
+section conclusions will be drawn and put forward only when supported by
+large differences between models and filters, while we will keep a
+neutral stance when such support turns out to be mild.
+
+\subsection{Comparing the filters}
+
+In order to evaluate the different filters according to the evaluation
+tool represented by the realized volatility, we start once more from the
+visual inspection of the plots. For any of the three indices, below
+graphs are shown including both the filtered volatility obtained from
+the filtered states (for any of the Sequential MC methods described in
+the theoretical section) and the corresponding rescaled realised
+volatility
+series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.}
+As compared to the visual analysis we conducted in section
+\textit{fill with the number of the section}, we can now use such a
+proxy to determine the relative performance of the competing filters.\\
+
+Once more, most algorithms used (aside from SIS and Liu-West) seem to
+provide indistinguishable results and, generally speaking, to
+underestimate the volatility (as indicated by the realised volatility
+proxy). The SIS algorithm appears completely inadequate to describe the
+``true'' volatility process: as we argued in section XX, the degeneracy
+of its weights on the particles prevents it from delivering a good
+enough approximation of the target distribution
+(\textit{check whether this is the correct intuition}). On top of that,
+it does not seem the case that the SIS algorithm, by falling short of
+the SV target distribution, ends up fitting the proxy
+better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\\
+On the other hand, the Liu and West filter offers the best performance
+among all filters, at least visually. This suggests that, since the Liu
+and West particle filter is built on a different underlying model than
+the other filters considered, this model may offer a better
+representation of the data analyzed. Once more, these considerations
+hold for all of the three indices.\\
+
+GRAFICI
+
+The above intuitions are confirmed by looking at Table XX, where the
+RMSE and MAE are shown for each filter (run with a number of particles
+equal to 10000), considering the realised volatility as benchmark.
+Indeed, for any index, most of the filters give rather similar results
+in terms of RMSE and MAE. For any index and for any measure, the worst
+performing is the SIS algorithm, while the best performing is the Liu
+and West particle filter. (Metterei qua l'intuizione del Liu West)\\
+
+\begin{longtable}[t]{lccccccc}
+\caption{\label{tab:unnamed-chunk-42}RMSE and MAE, realised volatility as benchmark}\\
+\toprule
+  & N & BPF & GPFOPT & APF & BAPF & LW & SIS\\
+\midrule
+RMSE SP500 & 10000 & 0.5086 & 0.5093 & 0.5099 & 0.5103 & 0.4287 & 0.6029\\
+MAE SP500 & 10000 & 0.3538 & 0.3531 & 0.3517 & 0.3531 & 0.2916 & 0.4320\\
+RMSE STOXX & 10000 & 0.4636 & 0.4634 & 0.4757 & 0.4678 & 0.3441 & 0.6525\\
+MAE STOXX & 10000 & 0.3019 & 0.3042 & 0.3042 & 0.3042 & 0.2372 & 0.4408\\
+RMSE DJI & 10000 & 0.5076 & 0.5062 & 0.5093 & 0.5054 & 0.4408 & 0.6047\\
+\addlinespace
+MAE DJI & 10000 & 0.3375 & 0.3383 & 0.3374 & 0.3362 & 0.2810 & 0.4156\\
+\bottomrule
+\end{longtable}
+
+Finally, we can see from Table XX the RMSE and MAE obtained with a
+number of particles equal to 50000, for the SP500 and the EuroStoxx50
+indices. We consider the Bootstrap particle filter and the Auxiliary
+particle filter, which we employed in section XX; in addition to these,
+we add the Liu and West filter, which did not feature in that section of
+this work.\\
+In the case of the SP500 index, although the number of particles
+increases, the RMSE and MAE increase (although by very small amounts)
+for both the particle filters based on our usual specification of the SV
+model. Instead, for the Liu and West filter we see some improvements,
+although the small extent suggests that also the LW filter was already
+in the convergence phase with N=10000.\\
+In the case of the EuroStoxx50 index, while the Bootstrap filter
+increases both RMSE and MAE when N is raised to 50000, the APF shows
+improvements in its performance, according to both criteria, which
+instead disagree for the Liu and West filter.
+
+\begin{longtable}[t]{lcccc}
+\caption{\label{tab:unnamed-chunk-44}RMSE and MAE, realised volatility as benchmark (SP500 and STOXX50E)}\\
+\toprule
+  & N & BPF & APF & LWF\\
+\midrule
+RMSE SP500 & 50000 & 0.51170 & 0.51489 & 0.42423\\
+MAE SP500 & 50000 & 0.35419 & 0.35534 & 0.29069\\
+RMSE STOXX & 50000 & 0.46503 & 0.46704 & 0.34230\\
+MAE STOXX & 50000 & 0.30344 & 0.30407 & 0.23956\\
+\bottomrule
+\end{longtable}
+
+Nevertheless, as we previously argued, we should not let tiny
+differences drive our conclusions. The idea behind employing also such
+50000 particles approximations (\textit{i.e} more precise and
+computationally expensive ones) is to check whether the relative
+performance that we assessed with N=10000 might be significantly
+influenced by issues of convergence among the filters: possibly, by
+further approaching the exact filtering distribution implied by the
+basic SV model, the filters built on such specification would improve
+their relative performance. This does not seem to be the case, even
+though the use of a proxy of the latent volatility process leaves room
+for discussion, as such conclusions rely on the assumption that such
+proxy is indeed a good
+one.\footnote{Nota che riporta il fatto che in certi casi si ottengano MAE e RMSE più bassi usando N più bassi di 10000: in altre parole, approssimare un po' peggio la distribuzione implicata dal modello SV può portare a rispettare meglio la proxy.}\\
+Rather, what this exercise highlights (under its assumptions) is a need
+for more flexibility and variability in the basic SV model: the
+underlying SV model of the LW filter meets this request, but clearly
+other solutions are possible, and have been argued for in the field of
+financial econometrics and mathematical finance, as we noted in the
+literary review section.
+
+\section{Bibliography}
+
+\hfill\break
+
+Allen, D. E., \& McAleer, M. (2020). Do we need stochastic volatility
+and generalised autoregressive conditional heteroscedasticity? Comparing
+squared end-of-day returns on FTSE. Risks, 8(1), 12.\\
+
+Allen, D. E. (2020). Stochastic Volatility and GARCH: Do Squared
+End-of-Day Returns Provide Similar Information?. Journal of Risk and
+Financial Management, 13(9), 202.\\
+
+Andrieu, C., Doucet, A., \& Holenstein, R. (2010). Particle markov chain
+monte carlo methods. Journal of the Royal Statistical Society: Series B
+(Statistical Methodology), 72(3), 269-342.\\
+
+Barndorff‐Nielsen, O. E., \& Shephard, N. (2001). Non‐Gaussian
+Ornstein--Uhlenbeck‐based models and some of their uses in financial
+economics. Journal of the Royal Statistical Society: Series B
+(Statistical Methodology), 63(2), 167-241.\\
+
+Barndorff‐Nielsen, O. E., \& Shephard, N. (2002). Econometric analysis
+of realized volatility and its use in estimating stochastic volatility
+models. Journal of the Royal Statistical Society: Series B (Statistical
+Methodology), 64(2), 253-280.\\
+
+Bollerslev, T. (1986). Generalized autoregressive conditional
+heteroskedasticity. Journal of econometrics, 31(3), 307-327.\\
+
+Breidt, F. J., Crato, N., \& De Lima, P. (1998). The detection and
+estimation of long memory in stochastic volatility. Journal of
+econometrics, 83(1-2), 325-348.\\
+
+Buncic, D., \& Gisler, K. I. (2016). Global equity market volatility
+spillovers: A broader role for the United States. International Journal
+of Forecasting, 32(4), 1317-1339.\\
+
+Chan, J. C. (2013). Moving average stochastic volatility models with
+application to inflation forecast. Journal of Econometrics, 176(2),
+162-172.\\
+
+Chib, S., Nardari, F., \& Shephard, N. (2002). Markov chain Monte Carlo
+methods for stochastic volatility models. Journal of Econometrics,
+108(2), 281-316.\\
+
+Chib, S., Nardari, F., \& Shephard, N. (2006). Analysis of high
+dimensional multivariate stochastic volatility models. Journal of
+Econometrics, 134(2), 341-371.\\
+
+Chopin, N., \& Papaspiliopoulos, O. (2020). An introduction to
+sequential Monte Carlo (Vol. 4). Springer.\\
+
+Comte, F., \& Renault, E. (1998). Long memory in continuous‐time
+stochastic volatility models. Mathematical finance, 8(4), 291-323.\\
+
+Christoffersen, P., Jacobs, K., \& Mimouni, K. (2010). Volatility
+dynamics for the S\&P500: Evidence from realized volatility, daily
+returns, and option prices. The Review of Financial Studies, 23(8),
+3141-3189.\\
+
+Dimitrakopoulos, S., \& Kolossiatis, M. (2020). Bayesian analysis of
+moving average stochastic volatility models: modeling in-mean effects
+and leverage for financial time series. Econometric Reviews, 39(4),
+319-343.\\
+
+Douc, R., Moulines, E., \& Stoffer, D. (2014). Nonlinear time series:
+Theory, methods and applications with R examples. CRC press.\\
+
+Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with
+estimates of the variance of United Kingdom inflation. Econometrica:
+Journal of the econometric society, 987-1007.\\
+
+Eraker, B. (2004). Do stock prices and volatility jump? Reconciling
+evidence from spot and option prices. The Journal of Finance, 59(3),
+1367-1403.\\
+
+Friz, P. K., Gassiat, P., \& Pigato, P. (2021). Precise asymptotics:
+robust stochastic volatility models. The Annals of Applied Probability,
+31(2), 896-940.\\
+
+Gatheral, J., Jaisson, T., \& Rosenbaum, M. (2018). Volatility is rough.
+Quantitative finance, 18(6), 933-949.\\
+
+Guidolin, M., \& Pedio, M. (2021). Media Attention vs.~Sentiment as
+Drivers of Conditional Volatility Predictions: An Application to Brexit.
+Finance Research Letters, 101943.\\
+
+Hansen, P. R., \& Lunde, A. (2006). Consistent ranking of volatility
+models. Journal of Econometrics, 131(1-2), 97-121.\\
+
+Harvey, A. C., \& Shephard, N. (1996). Estimation of an asymmetric
+stochastic volatility model for asset returns. Journal of Business \&
+Economic Statistics, 14(4), 429-434.\\
+
+Harvey, A., Ruiz, E., \& Shephard, N. (1994). Multivariate stochastic
+variance models. The Review of Economic Studies, 61(2), 247-264.\\
+
+Heber, Gerd, Asger Lunde, Neil Shephard and Kevin Sheppard (2009).
+Oxford-Man Institute's realized library. Oxford-Man Institute,
+University of Oxford, version 0.3. Available at:
+\url{https://realized.oxford-man.ox.ac.uk/}.\\
+
+Heston, S. L. (1993). A closed-form solution for options with stochastic
+volatility with applications to bond and currency options. The review of
+financial studies, 6(2), 327-343.\\
+
+Hull, J., \& White, A. (1987). The pricing of options on assets with
+stochastic volatilities. The journal of finance, 42(2), 281-300.\\
+
+Jacquier, E., Polson, N. G., \& Rossi, P. (1994). Bayesian analysis of
+stochastic volatility models. Journal of Business and Economic
+Statistics, 12(4), 371-389.\\
+
+Jacquier, E., Polson, N. G., \& Rossi, P. E. (2004). Bayesian analysis
+of stochastic volatility models with fat-tails and correlated errors.
+Journal of Econometrics, 122(1), 185-212.\\
+
+Johannes, M., Korteweg, A., \& Polson, N. (2014). Sequential learning,
+predictability, and optimal portfolio returns. The Journal of Finance,
+69(2), 611-644.\\
+
+Kambouroudis, D. S., McMillan, D. G., \& Tsakou, K. (2016). Forecasting
+stock return volatility: a comparison of GARCH, Implied volatility, and
+realized volatility models. Journal of Futures Markets, 36(12),
+1127-1163.\\
+
+Kastner, G. (2019). Dealing with stochastic volatility in time series
+using the R package stochvol. arXiv preprint arXiv:1906.12134.\\
+
+Kim, S., Shephard, N., \& Chib, S. (1998). Stochastic volatility:
+likelihood inference and comparison with ARCH models. The review of
+economic studies, 65(3), 361-393.\\
+
+Kim, J., \& Stoffer, D. S. (2008). Fitting stochastic volatility models
+in the presence of irregular sampling via particle methods and the EM
+algorithm. Journal of time series analysis, 29(5), 811-833.\\
+
+Koopman, S. J., \& Hol Uspensky, E. (2002). The stochastic volatility in
+mean model: empirical evidence from international stock markets. Journal
+of applied Econometrics, 17(6), 667-689.\\
+
+Liesenfeld, R., \& Jung, R. C. (2000). Stochastic volatility models:
+conditional normality versus heavy‐tailed distributions. Journal of
+applied Econometrics, 15(2), 137-160.\\
+
+Luo, R., Zhang, W., Xu, X., \& Wang, J. (2018, April). A neural
+stochastic volatility model. In Thirty-second AAAI conference on
+artificial intelligence.\\
+
+Mao, X., Czellar, V., Ruiz, E., \& Veiga, H. (2020). Asymmetric
+stochastic volatility models: Properties and particle filter-based
+simulated maximum likelihood estimation. Econometrics and Statistics,
+13, 84-105.\\
+
+Nakajima, J., \& Omori, Y. (2012). Stochastic volatility model with
+leverage and asymmetrically heavy-tailed error using GH skew Student's
+t-distribution. Computational Statistics \& Data Analysis, 56(11),
+3690-3704.\\
+
+Omori, Y., Chib, S., Shephard, N., \& Nakajima, J. (2007). Stochastic
+volatility with leverage: Fast and efficient likelihood inference.
+Journal of Econometrics, 140(2), 425-449.\\
+
+Pitt, M. K., \& Shephard, N. (1999). Filtering via simulation: Auxiliary
+particle filters. Journal of the American statistical association,
+94(446), 590-599.\\
+
+Poon, S. H., \& Granger, C. (2005). Practical issues in forecasting
+volatility. Financial analysts journal, 61(1), 45-56.\\
+
+Shephard, N., \& Sheppard, K. (2010). Realising the future: forecasting
+with high‐frequency‐based volatility (HEAVY) models. Journal of Applied
+Econometrics, 25(2), 197-231.\\
+
+So, M. E. P., Lam, K., \& Li, W. K. (1998). A stochastic volatility
+model with Markov switching. Journal of Business \& Economic Statistics,
+16(2), 244-253.\\
+
+Xu, X., \& Chen, Y. (2021). Deep Stochastic Volatility Model. arXiv
+preprint arXiv:2102.12658.\\
+
+Zhang, B., Chan, J. C., \& Cross, J. L. (2020). Stochastic volatility
+models with ARMA innovations: An application to G7 inflation forecasts.
+International Journal of Forecasting, 36(4), 1318-1328.
+
+\chapter{Appendix}
+
+\section{The Stochastic Volatility Model}
+
+We decide to organise this Appendix as an R package. All the R functions
+described in this Appendix take as reference a basic specification of
+Stochastic Volatility Model.
+
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
+
+All the plots and tables of this work concerning stochastic volatility
+models use this functions.
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SVPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\hfill\break
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x,tau)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hfill\break
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SVGPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVGPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    xprev}\OtherTok{\textless{}{-}}\NormalTok{x}
+\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{/}\DecValTok{4}\SpecialCharTok{*}\NormalTok{((data[t]}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{(alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x))}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{),tau)}
+\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(x,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev,tau)}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(x,}\AttributeTok{mean=}\NormalTok{alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{/}\DecValTok{4}\SpecialCharTok{*}\NormalTok{((data[t]}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{(alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev))}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{),tau)}
+
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hfill\break
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SVAPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVAPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x[k],tau)}
+\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x1}\SpecialCharTok{/}\DecValTok{2}\NormalTok{),}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x[k]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{),}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
+\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SVAPFoptfun(data,N,m0,C0,alpha,beta,tau,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVAPFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
+\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+\NormalTok{    x\_prev}\OtherTok{=}\NormalTok{x}
+\NormalTok{    x\_hat }\OtherTok{=}\NormalTok{ alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x }\CommentTok{\#means of the transition from the previous particle }
+\NormalTok{    I\_k }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{prob=}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x\_hat)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))))}
+    \ControlFlowTok{for}\NormalTok{(k }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{N)\{}
+\NormalTok{      x[k]}\OtherTok{=}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{size=}\DecValTok{1}\NormalTok{,}\AttributeTok{mean=}\NormalTok{alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x\_prev[I\_k],}\AttributeTok{sd=}\NormalTok{tau)}
+\NormalTok{      w[k]}\OtherTok{=}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x[k])}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x\_hat)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}
+\NormalTok{    \}}
+\NormalTok{    w }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hrule
+\hrule
+
+\hfill\break
+\texttt{SVLWfun(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVLWfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)\{}
+
+\NormalTok{  xs }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(N,ealpha,}\FunctionTok{sqrt}\NormalTok{(valpha)),}\FunctionTok{rnorm}\NormalTok{(N,ebeta,}\FunctionTok{sqrt}\NormalTok{(vbeta)),}
+                 \FunctionTok{log}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\FunctionTok{rgamma}\NormalTok{(N,nu}\SpecialCharTok{/}\DecValTok{2}\NormalTok{,nu}\SpecialCharTok{*}\NormalTok{lambda}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}
+\NormalTok{  delta  }\OtherTok{=} \FloatTok{0.75}
+\NormalTok{  a      }\OtherTok{=}\NormalTok{ (}\DecValTok{3}\SpecialCharTok{*}\NormalTok{delta}\DecValTok{{-}1}\NormalTok{)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{2}\SpecialCharTok{*}\NormalTok{delta)}
+\NormalTok{  h2     }\OtherTok{=} \DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}
+\NormalTok{  parss  }\OtherTok{=} \FunctionTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\FunctionTok{c}\NormalTok{(N,}\DecValTok{3}\NormalTok{,n))}
+\NormalTok{  xss    }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  ws     }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  ESS    }\OtherTok{=} \ConstantTok{NULL}
+\NormalTok{  w      }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{  vpars   }\OtherTok{=} \ConstantTok{NULL}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
+
+    \CommentTok{\#mpar        = apply(pars,2,mean) \#alternatively use this }
+\NormalTok{    mpar}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{()}
+    \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{3}\NormalTok{)\{}
+\NormalTok{      mpar[i]    }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,i],w)}
+\NormalTok{    \}}
+
+\NormalTok{    vpar     }\OtherTok{=} \FunctionTok{var}\NormalTok{(pars)}
+\NormalTok{    ms          }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\FunctionTok{matrix}\NormalTok{(mpar,N,}\DecValTok{3}\NormalTok{,}\AttributeTok{byrow=}\NormalTok{T)}
+\NormalTok{    mus         }\OtherTok{=}\NormalTok{ pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{xs }
+\NormalTok{    weight      }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(mus}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    k           }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    ms1         }\OtherTok{=}\NormalTok{ ms[k,] }\SpecialCharTok{+} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{3}\SpecialCharTok{*}\NormalTok{N),N,}\DecValTok{3}\NormalTok{)}\SpecialCharTok{\%*\%}\FunctionTok{chol}\NormalTok{(h2}\SpecialCharTok{*}\NormalTok{vpar)}
+\NormalTok{    xt          }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,ms1[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{ms1[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{xs[k],}\FunctionTok{exp}\NormalTok{(ms1[,}\DecValTok{3}\NormalTok{]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    w           }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(xt}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(mus[k]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    w           }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS         }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))\{}
+\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
+\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xt[index]}
+\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{ms1[index,]}
+\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xt}
+\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{ms1}
+\NormalTok{    \}}
+
+\NormalTok{    vpars        }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(vpars,vpar)}
+\NormalTok{    xss         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xss,xs)}
+\NormalTok{    parss[,,t]  }\OtherTok{=}\NormalTok{ pars }
+\NormalTok{    ws          }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{  \}}
+  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xss,}\AttributeTok{pars=}\NormalTok{parss,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{vpars=}\NormalTok{vpars))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \backmatter

### 2021-07-30 prova_knit_finale.Rmd
intro added
SVAPFopt è errata

@@
 Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values.
-The algorithm follows the following steps.
+Next, we descrive the algorithm.
 \begin{algorithm} LWF for Random Walk plus Noise Model
@@
 
-So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution
+So far we have seen Sequential Monte Carlo methods applied to the linear Gaussian case, however, as already mentioned, for this class of state space models, Kalman Filter provides a closed form solution.
+Here, we want to provide an illustration of Particle Filters for a state space model for which no analytical solutions are available.
+More precisely, we deal with a stochastic volatility model.
+For the sake of simplicity we use a basic specification of stochastic volatility model
+
+\begin{align}
+y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
+x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align}
+
+We start describing the main four particle filtering strategies discussed in this work and then we are going to compare their performances.
+The R code of these and other algorithms used in the Chapter application can be found in Appendix 1.
+
+\begin{algorithm} BPF for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{x_{t}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+\begin{algorithm} GPF for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+\begin{algorithm}APF with optimal transition kernel for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\sigma^{2}+\tau^{2}) $
+\item Compute $\mu_{opt}=\mu(x_{t-1}^{(I_{k})})$
+\item Draw $x_{t}^{(k)} \sim N(\mu_{opt},\sigma^{2}_{opt})$
+\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
+\end{enumerate}
+\item otherwise
+\begin{enumerate}
+\item Compute $\mu_{opt}=\mu(x_{t-1})$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma^{2}_{opt})$ for $i=1,...N$
+\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};x_{t}^{(i)},\sigma^{2}+\tau^{2})$ for $i=1,...,N$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
 
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
+      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
 
@@
 
-```{r}
-###################
-## Per roberto:::::: cambia molto se usiamo l'uno o l'altro!!!!!! perchè??
-APFoptfun<-function(data,N,m0,C0,tau,sigma,r){
-  if(missing(r)){r=2}else{}
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  importancesd<-sqrt(tau - tau^2 /(tau + sigma))
-  predsd <- sqrt(sigma+tau)
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    weight = w*dnorm(data[t],x,predsd)
-    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-    means<-x[k]+(tau/(tau+sigma))*(data[t]-x[k])
-    x1   = rnorm(N,means,importancesd)
-    lw  = dnorm(data[t],x1,predsd,log=TRUE)-dnorm(data[t],x[k],predsd,log=TRUE)+log(w)-log(weight)
-    w   = exp(lw)
-    w   = w/sum(w)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x1<-x1[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    x <- x1
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-```
 \

### 2021-07-31 prova_knit_finale.Rmd
small changes

@@
 
+\tableofcontents
+
 \chapter*{Introduction}
@@
 In fact, they allow to model observational data as function of underlying states that cannot be observed.
-For instance, in macroeconomic models GDP is usually observed while the TFP sequence is the hidden state determining output.
-In this work we describe techniques, such as Kalman and Particle filter, that allows to approximate probability distributions for the states from the observed data.
-For example, from the sequence of observed GDP, it is possible to "filter" the TFP factor.\
+For instance, in macroeconomic models GDP is usually observed while the TFP sequence is the hidden state determining output.\
+In this work we describe techniques, such as Kalman and Particle filter, that allow to approximate probability distributions for the states from the observed data.
+For example, from the sequence of observed GDP, it is possible to "filter" the TFP factor.
+\
 Furthermore, it is possible to impose a Markov structure to State-Space models, assuming that the states are Markov, i.e., the probability of the current state depends on the entire sequence of past states only through its first lag.
-The combination of State-Space and Markov structures constitutes a powerful modeling assumption for inference and forecasting.\ \
-Withing the Monte Carlo (MC) methods, where inference is based on repeated simulations that allow to approximate the target distributions (needed to compute the estimator of interest), Particle Filter (PF) Algorithms represent the most important class of procedures for State-Space-Markov models.
+The combination of State-Space and Markov structures constitutes a powerful modeling assumption for inference and forecasting.
+\
+
+Within the Monte Carlo (MC) methods, where inference is based on repeated simulations that allow to approximate the target distributions (needed to compute the estimator of interest), Particle Filter (PF) Algorithms represent the most important class of procedures for State-Space-Markov models.
 PF allows to sample states period-by-period, updating sequentially the filtering distributions (i.e., the distribution of current state conditional on all the sequence of observations - up to the current period one -), incorporating the information on the likelihood of the extracted states given the sequence of observations, by means of re-weighting.\
 Moreover, the PF is a very general algorithm, since it can be applied to non-linear models with non-Gaussian errors, while it does not allow for a closed form solution.
-On the contrary, the Kalman Filter, is a filtering method that can be applied in the case of linear-Gaussian  State-Space-Markov models, that remains analytically tractable.\
-In this work we will describe both these algorithms, while our main focus remains on Particle Filter.\ \
+On the contrary, the Kalman Filter is a filtering method that can be applied in the case of linear-Gaussian State-Space-Markov models, and remains analytically tractable.\
+In this work we will describe both these algorithms, while our main focus remains on Particle Filter. \
+
 In chapter \ref{MP_SSM_chapter} we define precisely Markov processes and State-Space Models.
@@
 Let $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ and $(\mathcal{Y}, \mathcal{B}(\mathcal{Y}))$ be two (eventually equal) measurable spaces. 
+\
 
@@
 \
+
 \begin{definition}
@@
 Then, in section \ref{pf_main} we describe three particle filtering algorithms: the \textit{bootstrap}, the \textit{guided} and the \textit{auxiliary particle filters}.
-Instead, in section \ref{pf_lui_west} we describe the \textit{Liu and West filter}, when particle filtering is used to estimate model's parameters.
+Instead, in section \ref{pf_liu_west} we describe the \textit{Liu and West filter}, when particle filtering is used to estimate model's parameters.
 Finally, in section \ref{pf_converg} we establish some general convergence results for Particle Filter - Monte Carlo estimators.
@@
 
-\section{Liu and West Filter}\label{pf_lui_west}
+\section{Liu and West Filter}\label{pf_liu_west}
 
@@
 Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution.
-In this way, at every time, the support of $\psi$ is not limited to the initially sampled $N$ values.
+In this way, at every time, the support of $\psi$ is not limited to the $N$ initially sampled values.
 \
@@
 Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.
-For instance, in our implementation with the random walk model of section \ref{implementation}, we use an auxiliary particle LWF.}\
+For instance, in our following implementation with the random walk model we use an auxiliary particle LWF.}\
 Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
@@
 \begin{enumerate}
-    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)^2$.
-    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$.
+    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)(\psi^n-\bar \psi)'$.
+    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.
     \item Draw new parameters: $\psi^n\thicksim \N \big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.

### 2021-07-31 prova_knit_finale.Rmd
fixe LW

--- 2021-07-30 prova_knit_finale.Rmd
@@
 \begin{enumerate}
-\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\item Compute
 \begin{align*}
-m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
-\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\overline{\sigma^{2}}& =E_{\hat{\pi}_{t-1}}(\sigma^{2})\\
+\overline{\tau^{2}}& =E_{\hat{\pi}_{t-1}}(\tau^{2}) \\
+\Sigma_{v} & =V_{\hat{\pi}_{t-1}}(\sigma^{2}) \\
+\Sigma_{w} & =V_{\hat{\pi}_{t-1}}(\tau^{2})
+\end{align*}
+then set
+\begin{align*}
+m_{v}^{(i)} & = a{\sigma^{2}}^{(i)}+(1-a)\overline{\sigma^{2}} \\
+{s_{v}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\sigma^{2}} \\
+\alpha_{v}^{(i)} & =\frac{(m_{v}^{(i)})^{2}}{{s_{v}^{2}}^{(i)}} \\
+\beta_{v}^{(i)} & =\frac{(m_{v}^{(i)})}{{s_{v}^{2}}^{(i)}}\\
+m_{w}^{(i)} & = a{\tau^{2}}^{(i)}+(1-a)\overline{\tau^{2}} \\
+{s_{w}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\tau^{2}}\\
+\alpha_{w}^{(i)} & =\frac{(m_{w}^{(i)})^{2}}{{s_{w}^{2}}^{(i)}} \\
+\beta_{w}^{(i)} & =\frac{(m_{w}^{(i)})}{{s_{w}^{2}}^{(i)}}
 \end{align*}
@@
 \begin{itemize}
-\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\psi=m^{(i)}) $
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),m_{v}) $
 \item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
 \item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
-\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},\psi=\psi^{(k)})$
-\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},\psi=\psi^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},\psi=m^{(I_k)}))}$
+\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},{\tau^2}^{(k)})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},{\sigma^2}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},m_{v}^{(I_k)}))}$
 \end{itemize}
@@
 \begin{align}
-y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
+y_{t}|x_{t} & \sim N(0,e^{\frac{x_{t}}{2}}) \\
 x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
@@
 
+In the Bootstrap Particle filter we basically use the model transition kernel and 
+do resampling any time the effective sample size is smaller than a predetermined threshold.
 \begin{algorithm} BPF for Stochastic Volatility Model
@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{x_{t}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{\frac{x_{t}}{2}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
 \item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
@@
 \end{algorithm}
+
+If we instead want to sample from the optimal transition kernel, things get more complicated. 
+It can be shown that the optimal transition kernel is
+\[ m_{t}^{opt}(x_{t}|x_{t-1}) \propto f_{t}(y_{t}|x_{t})p(x_{t}|x_{t-1}) \propto exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg]\]
+and it is not straightforward to sample from it.
 
@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{\frac{x_{t}}{2}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
 \item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
@@
 In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
-\item Compute $\sigma_{opt}^{2}$ 
 \item For $t=1,...,n$:
@@
 \item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+\begin{algorithm} LWF for Random Walk plus Noise Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $(\alpha^{(1)},...,\alpha^{(N)})$ from $N(e_{\alpha},v_\alpha)$ and
+$(\beta^{(1)},...,\beta^{(N)})$ from $N(e_{\beta},v_\beta)$ $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $IG(\nu_{0}/2,\nu_{0}\lambda/2)$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$ and chose the appropriate value for $\delta$.
+Therefore $\psi^{(i)}=({\alpha}^{(i)},{\beta}^{(i)},{\tau^{2}}^{(i)})$, and
+$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:.
+\begin{enumerate}
+\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\begin{align*}
+m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
+\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\end{align*}
+\item For $k=1,...,N$:
+\begin{itemize}
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|0,e^{\frac{\hat{x}_{t}^{(i)}}{2}}) $
+\item Draw $\psi^{(k)}$ from $N(m^{(I_{k})},h^{2}\Sigma)$
+\item Draw $x_{t}^{(k)}$ from $N(\alpha^{(k)}+\beta^{(k)}x_{t-1}^{(I_{k})},\tau^{2})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|0,e^{\frac{x_{t}^{(k)}}{2}})}{f_{N}(y_{t}|0,e^{\frac{\hat{x}_{t}^{(I_{k})}}{2}})}$
+\end{itemize}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
 \end{enumerate}

### 2021-07-31 prova_knit_finale.Rmd
small changes to LWF

--- 2021-07-31 prova_knit_finale.Rmd
@@
 \begin{enumerate}
-    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)(\psi^n-\bar \psi)'$.
+    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}\psi^{A_t^n}$, $\Omega = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}(\psi^{A_t^n}-\bar \psi)(\psi^{A_t^n}-\bar \psi)'$.\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
     \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.
@@
 Meanwhile, to extract the next period states and parameters it is necessary, firstly, to sample the new $N$ parameter values.
-Let $\bar \psi$ be the weighted average of the parameter values and $\Omega$ be their variance (the weights used to compute this mean and variance are the previous period inferential ones, i.e.
-$W_{t-1}^n$).\ \
+Let $\bar \psi$ be the weighted average of the parameter values and $\Omega$ be their variance (using the normalized weights in $t-1$ after the resampling step).\ \
 The algorithm suggests to sample new values for the parameters from a continuous distribution, i.e., a Normal\footnote{More generally, distributions whose support coincides with the relevant parameter space.} centered at the (resampled) values $m^{A_t^n}$ with variance $h^2\cdot \Omega$ where $m^n$ is a convex combination of the sample mean $\bar\psi$ and the actual extraction of $\psi^n$ with coefficient $a\in (0,1)$ and $h$ is a scalar such that $a^2+h^2=1$.
 While it would be more intuitively immediate to sample new parameters from a Normal density centered at the previously sampled values (i.e., $\psi^n$), this would increase the variance of $\psi$ over time.\
-In fact, let $\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \N(\psi^{A_t^{n}},\Omega)$\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.} be the empirical distribution of the parameters.
+In fact, let $\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \N(\psi^{A_t^{n}},\Omega)$ be the empirical distribution of the parameters.
 From the law of iterated expectation and total variance, we obtain:\footnote{For completeness, we specify that the first expected value in the law of iterated expectations and total variances is with respect to the distribution of the indexes and the second one with respect to each of the Normal distributions centered at $\psi^{A_t^{n}}$.}

### 2021-07-31 prova_knit_finale.Rmd
SV modified

--- 2021-07-31 prova_knit_finale.Rmd
@@
 \begin{align}
-y_{t}|x_{t} & \sim N(0,e^{\frac{x_{t}}{2}}) \\
+y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
 x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{\frac{x_{t}}{2}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{x_{t}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
 \item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
@@
 and it is not straightforward to sample from it.
+Therefore, we follow a common praxis consisting in approximating $m_{t}^{opt}$ by a Gaussian distribution obtained by linearising $e^{-x_{t}}$, i.e.
+$e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)$.
+Taking $x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}$, i.e the mean of the Gaussian transition density $p(x_{t}|x_{t-1})$.
+After some calculations we would end up with a Gaussian distribution $N(\xi(x_{t-1}),\sigma^{2})$ with
+\[ \xi(x_{t-1}) = \mu^{\star}(x_{t-1}) + \frac{\sigma^{2}}{4}(y_{t}^{2}e^{-\mu^{\star}(x_{t-1})}-2)\].
+With this transition kernel we can implement the Guided Particle Filter in the following way
 
@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
-\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{\frac{x_{t}}{2}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
 \item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
@@
 \end{algorithm}
+
+The Auxiliary Particle Filter with optimal transition kernel uses the same transition kernel of the Guided Particle Filter and we decided to use an auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=\alpha+\beta x_{t-1}$.
 
@@
 \begin{enumerate}
-\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\sigma^{2}+\tau^{2}) $
-\item Compute $\mu_{opt}=\mu(x_{t-1}^{(I_{k})})$
-\item Draw $x_{t}^{(k)} \sim N(\mu_{opt},\sigma^{2}_{opt})$
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|0,e^{\alpha+\beta x_{t-1}}) $
+\item $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}^{(I_{k})}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1}^{(I_{k})})}-2\big],\tau^{2})$ for $i=1,...N$
 \item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
@@
 \begin{enumerate}
-\item Compute $\mu_{opt}=\mu(x_{t-1})$
-\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma^{2}_{opt})$ for $i=1,...N$
-\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};x_{t}^{(i)},\sigma^{2}+\tau^{2})$ for $i=1,...,N$
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
+\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};0,\alpha+\beta x_{t})$ for $i=1,...,N$
 \end{enumerate}
@@
 
+Finally we 
 \begin{algorithm} LWF for Random Walk plus Noise Model
@@
     x<-rnorm(N,alpha+beta*x,tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))
+    w1<-w*dnorm(data[t],0,exp(x))
 
@@
 
-    weight = w*dnorm(data[t],0,exp(x/2))
+    weight = w*dnorm(data[t],0,exp(x))
     k   = sample(1:N,size=N,replace=TRUE,prob=weight)
     x1   = rnorm(N,alpha+beta*x[k],tau)
-    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
+    lw  = dnorm(data[t],0,exp(x1),log=TRUE)-dnorm(data[t],0,exp(x[k]),log=TRUE)
     w   = exp(lw)
@@
       x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
-      w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
+      w[k]=dnorm(data[t],0,sqrt(exp((x[k]))))/dnorm(data[t],0,sqrt(exp((x_hat))))
     }
@@
     mus         = pars[,1]+pars[,2]*xs 
-    weight      = w*dnorm(data[t],0,exp(mus/2))
+    weight      = w*dnorm(data[t],0,exp(mus))
     k           = sample(1:N,size=N,replace=T,prob=weight)
@@
     xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
-    w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
+    w           = dnorm(data[t],0,exp(xt))/dnorm(data[t],0,exp(mus[k]))
     w           = w/sum(w)

### 2021-07-31 prova_knit_finale.Rmd
sto facendo implementations SV model

@@
 
-Finally we 
-\begin{algorithm} LWF for Random Walk plus Noise Model
+Finally we present a Liu and West Filter for the case of parameters $\alpha$, $\beta$ and $\tau^{2}$ unknown.
+
+\begin{algorithm} LWF for Stochastic Volatility Model
 \begin{itemize}
@@
 \begin{align*}
-m^{(i)} & = a\psi^{(i)}+(1-a)\hat{\psi} \\
+m^{(i)} & = a\psi^{(i)}+(1-a)\overline{\psi} \\
 \hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
@@
 \end{algorithm}
+
+We are now ready to compare the performances of these four algorithms.
+We simulate a Stochastic Volatility process setting arbitrarily the following parameters: $\alpha = -0.1$, $\beta = 0.9$ and $\tau^{2}=0.01$.
+The resulting observation and state process is shown in the figure below.
+
+```{r include=FALSE}
+# Simulating the data
+# -------------------
+set.seed(12345)
+n     =  50
+alpha = -0.1
+beta  =  0.9
+tau2  =  0.01
+tau   = sqrt(tau2)
+y     = rep(0,n)
+x     = rep(0,n)
+x[1]  = alpha/(1-beta)
+y[1]  = rnorm(1,0,exp(x[1]))
+for (t in 2:n){
+  x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
+  y[t] = rnorm(1,0,exp(x[t]))
+}
+alpha.true = alpha
+beta.true  = beta
+tau2.true  = tau2
+
+timeframe<-c(1:length(y))
+dfsv<-data.frame(timeframe,y,x)
+```
+
+```{r echo=FALSE, fig.align='center', fig.cap="Simulated Stochastic Volatility Model", fig.height=4, fig.pos='ht', fig.width=8, message=FALSE, warning=FALSE, results=F}
+ggplot(dfsv,aes(x=timeframe))+
+  geom_line(aes(y=y, linetype="Observations"))+
+  geom_line(aes(y=x, linetype="True States"))+
+  scale_linetype_manual(name="",
+                      values=c("Observations" = 1,
+                               "True States" = 3))+
+  labs(x="Time",
+       y="")+
+  theme_bw()+
+  theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+```
 

### 2021-08-01 prova_knit_finale.Rmd
change
non va SVAPFopt

@@
     \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}\psi^{A_t^n}$, $\Omega = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}(\psi^{A_t^n}-\bar \psi)(\psi^{A_t^n}-\bar \psi)'$.\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
-    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.
+    \item Compute: $m^{A_t^{n}}=a \cdot  \psi^{{A_t}^{n}}+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.
     \item Draw new parameters: $\psi^n\thicksim \N \big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.
@@
 Therefore, we follow a common praxis consisting in approximating $m_{t}^{opt}$ by a Gaussian distribution obtained by linearising $e^{-x_{t}}$, i.e.
-$e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)$.
-Taking $x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}$, i.e the mean of the Gaussian transition density $p(x_{t}|x_{t-1})$.
+$e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)$, taking $x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}$, i.e the mean of the Gaussian transition density $p(x_{t}|x_{t-1})$.
 After some calculations we would end up with a Gaussian distribution $N(\xi(x_{t-1}),\sigma^{2})$ with
@@
 
-The Auxiliary Particle Filter with optimal transition kernel uses the same transition kernel of the Guided Particle Filter and we decided to use an auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=\alpha+\beta x_{t-1}$.
+The Auxiliary Particle Filter with optimal transition kernel uses the same transition kernel of the Guided Particle Filter and while the perfectly adapted auxiliary function is
+\[g(x_{t-1})=\frac{1}{2\pi\sigma^{2}}\int_{\mathcal{X}}exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg] dx_{t-1}\]
+The relevant literature suggest to approximate this integral by linearising the exponential term as in the GPF case. The resulting auxiliary function is
+\[ g(x_{t-1})=exp \bigg[ \frac{1}{2\sigma^{2}} \big\{\xi(x_{t-1})^{2}-\mu^{\star}(x_{t-1})^{2} \big\} - \frac{y_{t}^{2}}{2}e^{-\mu^{\star}(x_{t-1})} \big\{ 1+\mu^{\star}(x_{t-1}) \big\}  \bigg]\]
+where $\xi(x_{t-1})$ and $\mu^{\star}$ are defined as in the case of GPF.
 
@@
 \begin{enumerate}
-\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|0,e^{\alpha+\beta x_{t-1}}) $
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}g(x_{t-1}) $
 \item $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}^{(I_{k})}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1}^{(I_{k})})}-2\big],\tau^{2})$ for $i=1,...N$
@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
-\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};0,\alpha+\beta x_{t})$ for $i=1,...,N$
+\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}g(x_{t-1})$ for $i=1,...,N$
 \end{enumerate}
@@
   theme(plot.title = element_text(hjust = 0.5))
+```
+
+```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+SVGPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
+      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+#-------------------------------------------------------------------------
+
+SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
+    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    w   = rep(1/N,N)
+    }else{
+    weight = rep(1/N,N)
+    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+    }
+
+    w   = w/sum(w)
+    x <- x1
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+#----------------------------------------------------------------------------------
+
+SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+  xs = rnorm(N,m0,sqrt(C0))
+  pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
+                 log(1/rgamma(N,nu/2,nu*lambda/2)))
+  delta  = 0.75
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,3,n))
+  xss    = NULL
+  ws     = NULL
+  ESS    = NULL
+  w      = rep(1/N,N)
+  vpars   = NULL
+  for (t in 1:length(data)){
+
+    #mpar        = apply(pars,2,mean) #alternatively use this 
+    mpar<-c()
+    for(i in 1:3){
+      mpar[i]    = weighted.mean(pars[,i],w)
+    }
+
+    vpar     = var(pars)
+    ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
+    mus         = pars[,1]+pars[,2]*xs 
+    weight      = w*dnorm(data[t],0,exp(mus))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+    ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
+    xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
+    w           = dnorm(data[t],0,exp(xt))/dnorm(data[t],0,exp(mus[k]))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/2)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xt[index]
+      pars<-ms1[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xt
+      pars<-ms1
+    }
+
+    vpars        = rbind(vpars,vpar)
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+  }
+  return(list(xs=xss,pars=parss,ws=ws,vpars=vpars))
+}
+```
+
+```{r include=FALSE}
+N <- 5000
+alpha <- -0.1
+beta  <-  0.9
+tau2  =  0.01
+tau   = sqrt(tau2)
+m0      = 0
+C0      = 100
+ealpha  = alpha
+valpha  = 0.01
+ebeta    = beta
+vbeta    = 0.01
+nu      = 3
+lambda  = tau2
+
+svpf<-SVPFfun(y,N,m0,C0, alpha,beta,tau)
+
+svgpf<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
+
+svapf<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)
+
+svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+
+SVFiltplot<-function(dataframe,fun,title,ub,lb){
+
+Filt<-Filtervalues(fun)
+mean<-Filt$mean
+sd<-Filt$sd
+dataframe<-data.frame(dataframe,mean,sd)
+
+ggplot(dataframe,aes(x=timeframe))+
+  geom_line(aes(y=y), col="black", linetype=1)+
+  geom_line(aes(y=x), col="black", linetype=3)+
+  geom_line(aes(y=mean), col="red", linetype=1)+
+  geom_ribbon(aes(ymin = mean-1.96*sd, ymax = mean+1.96*sd),
+              fill="red",alpha=0.16) +
+  #  scale_color_manual("",
+   #                  values=c("Observations" = "black",
+    #                         "True States" = "black",
+     #                       "Filtered States"= "red"))+
+  #scale_linetype_manual("",
+   #                  values=c("Observations" = 1,
+    #                         "True States" = 3,
+     #                       "Filtered States"=1))+
+  labs(x="Time",
+       y="")+
+  ggtitle(title)+
+  coord_cartesian(ylim = c(lb, ub)) +
+  theme_bw()+
+#    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+#        legend.background = element_rect(fill = "white", colour = "gray30")) +
+  theme(plot.title = element_text(hjust = 0.5))
+}
+
+```
+
+We provide both a graphical and a numerical comparison and
+
+```{r echo=FALSE, fig.align='center', fig.cap="Comparison of filtering strategies for Stochastic Volatility Model.
+Filtered States with credible interval (in red)", fig.height=6, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+plotone<-SVFiltplot(dfsv,svpf,"",1.5,-2.5)
+plottwo<-SVFiltplot(dfsv,svgpf,"",1.5,-2.5)
+plotthree<-SVFiltplot(dfsv,svapf,"",1.5,-2.5)
+plotfour<-SVFiltplot(dfsv,svlw,"",1.5,-2.5)
+
+ggarrange(plotone, plottwo, plotthree, plotfour, nrow = 2, ncol = 2)
+```
+
+```{r}
+set.seed(123)
+truex<-x
+estimatedx<-matrix(NA,ncol=5,nrow=4)
+colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
+estimatedx[,1]<-c(100,1000,10000,100000)
+
+i=1
+for(N in c(100,1000,10000,100000)){
+set.seed(123)
+
+svpf_1<-SVPFfun(y,N,m0,C0, alpha,beta,tau)
+svgpf_1<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
+svapf_1<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)
+svlw_1<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+
+estimatedx[i,2]<-RMSE(truex,Filtervalues(svpf_1)$mean)
+estimatedx[i,3]<-RMSE(truex,Filtervalues(svgpf_1)$mean)
+estimatedx[i,4]<-RMSE(truex,Filtervalues(svapf_1)$mean)
+estimatedx[i,5]<-RMSE(truex,Filtervalues(svlw_1)$mean)
+i=i+1
+
+}
+
+```
+
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
 ```
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
+    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
       dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
@@
   for(t in 1:length(data)){
-    x_prev=x
-    x_hat = alpha+beta*x #means of the transition from the previous particle 
-    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
-    for(k in 1:N){
-      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
-      w[k]=dnorm(data[t],0,sqrt(exp((x[k]))))/dnorm(data[t],0,sqrt(exp((x_hat))))
-    }
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
+    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    w   = rep(1/N,N)
+    }else{
+    weight = rep(1/N,N)
+    k   = sample(1:N,size=N,replace=FALSE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+    }
+
     w   = w/sum(w)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x[index]
-      w<-rep(1/N,N)
-    }else{}
+    x <- x1
 
@@
 }
+
 ```

### 2021-08-01 prova_knit_finale.Rmd
belli rega APFoptfun funziona!

@@
 \item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
-\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}g(x_{t-1})$ for $i=1,...,N$
+\item Set $\tilde{w}_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})}$ for $i=1,...,N$
 \end{enumerate}
@@
 n     =  50
-alpha = -0.1
+alpha =  0
 beta  =  0.9
@@
    for(t in 1:length(data)){
+
       ESS  = 1/sum(w^2)
@@
       epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
-    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
     k   = sample(1:N,size=N,replace=TRUE,prob=weight)
@@
     x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+    w   <-  w*dnorm(data[t],0,exp(x1))*dnorm(x1,alpha+beta*x[k],tau)/
+      dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
     }
@@
 N <- 5000
-alpha <- -0.1
+alpha <- 0
 beta  <-  0.9
@@
 
-svapf<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)
+svapf<-SVAPFoptfun(y,N,m0,C0,alpha,beta,tau)
 
@@
 Filtered States with credible interval (in red)", fig.height=6, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
-plotone<-SVFiltplot(dfsv,svpf,"",1.5,-2.5)
-plottwo<-SVFiltplot(dfsv,svgpf,"",1.5,-2.5)
-plotthree<-SVFiltplot(dfsv,svapf,"",1.5,-2.5)
-plotfour<-SVFiltplot(dfsv,svlw,"",1.5,-2.5)
+plotone<-SVFiltplot(dfsv,svpf,"",2,-2)
+plottwo<-SVFiltplot(dfsv,svgpf,"",2,-2)
+plotthree<-SVFiltplot(dfsv,svapf,"",2,-2)
+plotfour<-SVFiltplot(dfsv,svlw,"",2,-2)
 
@@
 truex<-x
-estimatedx<-matrix(NA,ncol=5,nrow=4)
+estimatedx<-matrix(NA,ncol=5,nrow=5)
 colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
-estimatedx[,1]<-c(100,1000,10000,100000)
+estimatedx[,1]<-c(10,100,1000,10000,100000)
 

### 2021-08-01 prova_knit_finale.Rmd
Come viaggia sto APFoptfun

@@
 
-We provide both a graphical and a numerical comparison and
+We provide both a graphical and a numerical comparison.
+From the plots it is hard to say which particle filter provides the best performances.
+By and large, BPF, GPF and APF behave similarly while the Liu and West filter presents larger credible intervals but at the same time a higher coverage of the true states.
+On the other hand, by a RMSE comparison, the Auxiliary Particle filter provides the best performances followed bt the Liu and West filter.
+Moreover, notice that when the number of particles increases the accuracy improves, however not significant improvements are detected for $N>10000$.
 
@@
 
-```{r}
+```{r include=FALSE}
 set.seed(123)
 truex<-x
-estimatedx<-matrix(NA,ncol=5,nrow=5)
+estimatedx<-matrix(NA,ncol=5,nrow=4)
 colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
-estimatedx[,1]<-c(10,100,1000,10000,100000)
+estimatedx[,1]<-c(100,1000,10000,100000)
 
@@
 for(N in c(100,1000,10000,100000)){
-set.seed(123)
+set.seed(12345)
 
@@
 svgpf_1<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
-svapf_1<-SVAPFoptfun(y,N,m0,C0,alpha,beta=0,tau)
+svapf_1<-SVAPFoptfun(y,N,m0,C0,alpha,beta,tau)
 svlw_1<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
-kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE",digits=3)
+kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=3)
 ```

### 2021-08-07 prova_knit_finale.Rmd
Sistemato x/2

@@
 Since the Random Walk plus Noise model allows for closed form solutions, we want to compare the results of Boostrap Particle Filter(BPF) with Kalman Filter(KF).
-As we can see from figure \ref{fig:myfig4}, when the number of particles generated at any interaction increases the results for both the estimated mean and the variance tend to converge [[SPIEGARE PERCHE']].
+As we can see from figure \ref{fig:myfig4}, as the number of particles N increases, both the state level and the standard deviation estimated with particle filter converge to the ones obtained by the recursive Kalman filter formulas, confirming Monte Carlo approximations to be asymptotically consistent.
 Moreover, comparing the Root Mean Square Errors of the filtered states with respect to the true state values, when the sample size increases the BPF decreases and, for large N, it reaches the accuracy of the Kalman Filter.
@@
     x<-rnorm(N,alpha+beta*x,tau)
-    w1<-w*dnorm(data[t],0,exp(x))
+    w1<-w*dnorm(data[t],0,exp(x/2))
 
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
       dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
@@
 
-    weight = w*dnorm(data[t],0,exp(x))
+    weight = w*dnorm(data[t],0,exp(x/2))
     k   = sample(1:N,size=N,replace=TRUE,prob=weight)
     x1   = rnorm(N,alpha+beta*x[k],tau)
-    lw  = dnorm(data[t],0,exp(x1),log=TRUE)-dnorm(data[t],0,exp(x[k]),log=TRUE)
+    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
     w   = exp(lw)
@@
     mus         = pars[,1]+pars[,2]*xs 
-    weight      = w*dnorm(data[t],0,exp(mus))
+    weight      = w*dnorm(data[t],0,exp(mus/2))
     k           = sample(1:N,size=N,replace=T,prob=weight)
@@
     xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
-    w           = dnorm(data[t],0,exp(xt))/dnorm(data[t],0,exp(mus[k]))
+    w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
     w           = w/sum(w)

### 2021-08-07 Real Dataset Script.R
Nuovi parametri, definitivi

@@ -309,7 +309,7 @@ LWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
 #n     =  length(y)
 alpha =  0
 beta  =  0.99
-tau2  =  0.075
+tau2  =  0.05
 m0      = 0
 C0      = 100 #N(0,100) is an informative prior about the mean of the volatility. It is used by Jacquier et al. (2004), and it is calibrated on percentage returns
 sC0     = sqrt(C0)
@@ -317,8 +317,8 @@ ealpha  = alpha
 valpha  = 0.01
 ebeta    = beta
 vbeta    = 0.01
-nu      = 1
-lambda  = tau2/3
+nu      = 5
+lambda  = tau2*3/5
 tau   = sqrt(tau2)
 
 #Real dataset

### 2021-08-07 prova_knit_finale.Rmd
typo
Funzioni aggiornate (x/2)

@@
 As already mentioned before, in this case the filtering distribution can be computed in closed form solutions using the Kalman filter.
-However, this toy example will be used also to illustrate more involved filtering strategies described in this work.
+However, this toy example will be used also to illustrate more involving filtering strategies described in this work.
 We believe indeed that it represents a useful starting point to understand the logic of the algorithms which may be eventually replicated when dealing with more complex models.
@@
 
-We decide to organise this Appendix as an R package. All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model. 
+We decide to organise this Appendix as an R package.
+All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model.\footnote{Actually, the function SVLWfun considers a SV model with an additional hierarchical level, as specified in section XX.}
 \begin{align*}
@@
 \code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
 \
@@
 \
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
+\
 \code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/
-      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
 
@@
 \
-\code{alpha} \ \ intercept $\alpha$ in the state equation
-\
-\code{beta} \ \ coefficient $\beta$ in the state equation
+\code{ealpha} \ \ expected value of the intercept $\alpha$ in the state equation
+\
+\code{valpha} \ \ variance of the intercept $\alpha$ in the state equation
+\
+\code{ebeta} \ \ expected value of the coefficient $\beta$ in the state equation
+\
+\code{vbeta} \ \ variance of the coefficient $\beta$ in the state equation
+\
+\code{nu} \ \ when divided by 2, first hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
+\
+\code{lambda} \ \ when multiplied by $\nu/2$, second hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
 \

### 2021-08-07 Real Dataset Script.R

@@ -222,25 +222,25 @@ APFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   w  = rep(1/N,N)
   
   for(t in 1:length(data)){
-    
-    xprev<-x
-    
-    weight = w*dnorm(data[t],0,exp(x/2))
-    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-    x1 <-rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
-    w   = exp(lw)
-    w   = w/sum(w)
     ESS  = 1/sum(w^2)
     
     if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x1[index]
-      w<-rep(1/N,N)
-    }else{}
+      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
+      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+      k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+      w   = rep(1/N,N)
+    }else{
+      weight = rep(1/N,N)
+      k   = sample(1:N,size=N,replace=FALSE,prob=weight)
+      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+      w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+    }
     
[…]

### 2021-08-07 prova_knit_finale.Rmd
Aggiornato con x/2

@@
 We are now ready to compare the performances of these four algorithms.
-We simulate a Stochastic Volatility process setting arbitrarily the following parameters: $\alpha = -0.1$, $\beta = 0.9$ and $\tau^{2}=0.01$.
+We simulate a Stochastic Volatility process setting arbitrarily the following parameters: $\alpha = 0$, $\beta = 0.99$ and $\tau^{2}=0.05$.
 The resulting observation and state process is shown in the figure below.
@@
 alpha =  0
-beta  =  0.9
-tau2  =  0.01
+beta  =  0.99
+tau2  =  0.05
 tau   = sqrt(tau2)
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x))*dnorm(x,alpha+beta*xprev,tau)/
-      dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
 
@@
 alpha <- 0
-beta  <-  0.9
-tau2  =  0.01
+beta  <-  0.99
+tau2  =  0.05
 tau   = sqrt(tau2)
@@
 vbeta    = 0.01
-nu      = 3
-lambda  = tau2
+nu      = 5
+lambda  = tau2*3/5
 

### 2021-08-07 prova_knit_finale.Rmd
altri x/2

@@
 x[1]  = alpha/(1-beta)
-y[1]  = rnorm(1,0,exp(x[1]))
+y[1]  = rnorm(1,0,exp(x[1]/2))
 for (t in 2:n){
   x[t] = alpha + beta*x[t-1] + rnorm(1,0,tau)
-  y[t] = rnorm(1,0,exp(x[t]))
+  y[t] = rnorm(1,0,exp(x[t]/2))
 }

### 2021-08-07 prova_knit_finale.Rmd
Knit

@@
     x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-    w   <-  w*dnorm(data[t],0,exp(x1))*dnorm(x1,alpha+beta*x[k],tau)/
-      dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    w   <-  w*dnorm(data[t],0,exp(x1))*dnorm(x1,alpha+beta*x[k],tau)/dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
     }
@@
 }
+
 #----------------------------------------------------------------------------------
@@
 ```{r include=FALSE}
-N <- 5000
+N <- 10000
 alpha <- 0
@@
 Filtered States with credible interval (in red)", fig.height=6, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
-plotone<-SVFiltplot(dfsv,svpf,"",2,-2)
-plottwo<-SVFiltplot(dfsv,svgpf,"",2,-2)
-plotthree<-SVFiltplot(dfsv,svapf,"",2,-2)
-plotfour<-SVFiltplot(dfsv,svlw,"",2,-2)
+plotone<-SVFiltplot(dfsv,svpf,"",3,-3)
+plottwo<-SVFiltplot(dfsv,svgpf,"",3,-3)
+plotthree<-SVFiltplot(dfsv,svapf,"",3,-3)
+plotfour<-SVFiltplot(dfsv,svlw,"",3,-3)
 
@@
   for(t in 1:length(data)){
+
     ESS  = 1/sum(w^2)
@@
       epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
-    weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
       k   = sample(1:N,size=N,replace=TRUE,prob=weight)
@@
       x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-    w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+      w   <-  w*dnorm(data[t],0,exp(x1/2))*dnorm(x1,alpha+beta*x[k],tau)/
+        dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
     }

### 2021-08-07 Real Dataset Script.R

@@ -222,11 +222,12 @@ APFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   w  = rep(1/N,N)
   
   for(t in 1:length(data)){
+    
     ESS  = 1/sum(w^2)
     
     if(ESS<(N/r)){
       epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
-      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x))-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
       k   = sample(1:N,size=N,replace=TRUE,prob=weight)
       x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
       w   = rep(1/N,N)
@@ -234,7 +235,8 @@ APFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
       weight = rep(1/N,N)
       k   = sample(1:N,size=N,replace=FALSE,prob=weight)
       x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-      w   <-  w*exp((1/(2*tau^2))*((alpha+beta*x[k]+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x[k])-2))^2-(alpha+beta*x[k]))-((data[t]^2)/2)*exp(-alpha-beta*x[k])*(1+alpha+beta*x[k]))
+      w   <-  w*dnorm(data[t],0,exp(x1/2))*dnorm(x1,alpha+beta*x[k],tau)/
+        dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
     }
     
     w   = w/sum(w)
@@ -248,7 +250,6 @@ APFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   }
   return(list(xs=xs,ws=ws,ess=ess))
 }
-
 
 #Liu West
 #--------

### 2021-08-07 prova_knit_finale.Rmd
knit e risolto bug seed

@@
 set.seed(12345)
-n     =  50
+n     =  100
 alpha =  0
@@
 
-svpf<-SVPFfun(y,N,m0,C0, alpha,beta,tau)
-
-svgpf<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
-
-svapf<-SVAPFoptfun(y,N,m0,C0,alpha,beta,tau)
-
-svlw<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+set.seed(12345)
+N1=N
+svpf<-SVPFfun(y,N1,m0,C0, alpha,beta,tau)
+set.seed(12345)
+N1=N
+svgpf<-SVGPFfun(y,N1,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N1=N
+svapf<-SVAPFoptfun(y,N1,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N1=N
+svlw<-SVLWfun(y,N1,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 
@@
 Filtered States with credible interval (in red)", fig.height=6, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
-plotone<-SVFiltplot(dfsv,svpf,"",3,-3)
-plottwo<-SVFiltplot(dfsv,svgpf,"",3,-3)
-plotthree<-SVFiltplot(dfsv,svapf,"",3,-3)
-plotfour<-SVFiltplot(dfsv,svlw,"",3,-3)
+plotone<-SVFiltplot(dfsv,svpf,"",5,-5)
+plottwo<-SVFiltplot(dfsv,svgpf,"",5,-5)
+plotthree<-SVFiltplot(dfsv,svapf,"",5,-5)
+plotfour<-SVFiltplot(dfsv,svlw,"",5,-5)
 
@@
 ```{r include=FALSE}
-set.seed(123)
 truex<-x
-estimatedx<-matrix(NA,ncol=5,nrow=4)
+estimatedx<-matrix(NA,ncol=5,nrow=5)
 colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
-estimatedx[,1]<-c(100,1000,10000,100000)
+estimatedx[,1]<-c(20,100,1000,10000,100000)
 
 i=1
-for(N in c(100,1000,10000,100000)){
+for(N in c(20,100,1000,10000,100000)){
+
 set.seed(12345)
-
-svpf_1<-SVPFfun(y,N,m0,C0, alpha,beta,tau)
-svgpf_1<-SVGPFfun(y,N,m0,C0,alpha,beta,tau)
-svapf_1<-SVAPFoptfun(y,N,m0,C0,alpha,beta,tau)
-svlw_1<-SVLWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+N1=N
+svpf_1<-SVPFfun(y,N1,m0,C0, alpha,beta,tau)
+set.seed(12345)
+N1=N
+svgpf_1<-SVGPFfun(y,N1,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N1=N
+svapf_1<-SVAPFoptfun(y,N1,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N1=N
+svlw_1<-SVLWfun(y,N1,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
-kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=3)
+kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=6)
 ```

### 2021-08-07 prova_knit_finale.Rmd
Aggiornamento

@@
 
-We provide both a graphical and a numerical comparison.
-From the plots it is hard to say which particle filter provides the best performances.
-By and large, BPF, GPF and APF behave similarly while the Liu and West filter presents larger credible intervals but at the same time a higher coverage of the true states.
+We provide both a graphical and a numerical comparison.\footnote{The sequences of filtered states shown in the graphs are obtained by using filters with 10000 particles.} From the plots it is hard to say which particle filter provides the best performances.
+By and large, BPF, GPF and APF behave similarly, while the Liu and West filter presents larger credible intervals but at the same time a higher coverage of the true states.
 On the other hand, by a RMSE comparison, the Auxiliary Particle filter provides the best performances followed bt the Liu and West filter.
@@
 colnames(estimatedx)<-c("N","BPF","GPF", "APF","LWF")
-estimatedx[,1]<-c(20,100,1000,10000,100000)
+estimatedx[,1]<-c("20","100","1000","10000","100000")
 
@@
 ```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
-kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=6)
+kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=4)
 ```

### 2021-08-08 Real Dataset Script.R
Nuovi benchmark real dataset

@@ -451,7 +451,7 @@ Errorvol[2,7]<-MAE(realisedx,comparablevol(svbapf))
 
 # Save mean and sd series of PF with 50000 particles
 library(openxlsx)
-Filt<-Filtervalues(svapf)
+Filt<-Filtervalues(svbpf)
 mean<-Filt$mean
 sd<-Filt$sd
 list1<-list('mean'=mean,'sd'=sd)
@@ -462,13 +462,21 @@ write.xlsx(list1,file="svbpf50000p.xlsx")
 library(openxlsx)
 bpf50000mean<-read.xlsx("svbpf50000p.xlsx", sheet = 1)
 bpf50000sd<-read.xlsx("svbpf50000p.xlsx", sheet = 2)
+bpf100000mean<-read.xlsx("svbpf100000p.xlsx", sheet = 1)
+bpf100000sd<-read.xlsx("svbpf100000p.xlsx", sheet = 2)
 apf50000mean<-read.xlsx("svapf50000p.xlsx", sheet = 1)
 apf50000sd<-read.xlsx("svapf50000p.xlsx", sheet = 2)
-zmean<-bpf50000mean[,1]
-zsd<-bpf50000sd[,1]
-wmean<-apf50000mean[,1]
-wsd<-apf50000sd[,1]
-dfsv1<-data.frame(timeframe,y,zmean,zsd,wmean,zsd)
+apf100000mean<-read.xlsx("svapf100000p.xlsx", sheet = 1)
+apf100000sd<-read.xlsx("svapf100000p.xlsx", sheet = 2)
+z50mean<-bpf50000mean[,1]
+z50sd<-bpf50000sd[,1]
+z100mean<-bpf100000mean[,1]
+z100sd<-bpf100000sd[,1]
+w50mean<-apf50000mean[,1]
+w100mean<-apf100000mean[,1]
+w50sd<-apf50000sd[,1]
+w100sd<-apf100000sd[,1]
+dfsv1<-data.frame(timeframe,y,z50mean,z50sd,w50mean,z50sd)
 Errorcomp<-matrix(NA,ncol=6,nrow=2)
 colnames(Errorcomp)<-c("N","BPF", "GPFOPT", "APF", "SIS", "BAPF")
 rownames(Errorcomp)<-c("RMSE","MAE")
@@ -476,18 +484,18 @@ Errorcomp[,1]<-c(10000)
 RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
[…]

### 2021-08-08 Real Dataset Script.R
commit

@@ -460,6 +460,8 @@ write.xlsx(list1,file="svbpf50000p.xlsx")
 #RMSE and MAE comparison of particle filters (but not LW) vs Bootstrap PF with 50000 particles
 
 library(openxlsx)
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
 bpf50000mean<-read.xlsx("svbpf50000p.xlsx", sheet = 1)
 bpf50000sd<-read.xlsx("svbpf50000p.xlsx", sheet = 2)
 bpf100000mean<-read.xlsx("svbpf100000p.xlsx", sheet = 1)
@@ -496,6 +498,10 @@ Errorcomp[2,5]<-MAE(z50mean,comparablemean(svsis))
 Errorcomp[2,6]<-MAE(z50mean,comparablemean(svbapf))
 RMSEbench50000p<-RMSE(z50mean,w50mean)
 MAEDbench50000p<-MAE(z50mean,w50mean)
+difference50000p<-c(RMSEbench50000p,MAEDbench50000p)
+RMSEbench100000p<-RMSE(z100mean,w100mean)
+MAEDbench100000p<-MAE(z100mean,w100mean)
+difference100000p<-c(RMSEbench100000p,MAEDbench100000p)
 
 # Graphical comparison of filtered states
 

### 2021-08-08 Real Dataset Script.R
aggiornamento tabelle

@@ -325,10 +325,10 @@ tau   = sqrt(tau2)
 #Real dataset
 
 library(openxlsx)
-realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
 y<-realdatasetSPX[,2]
 n = length(y)
-volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
 x<-volatilitySPX[,3]
 alpha.true = alpha
 beta.true  = beta
@@ -346,10 +346,10 @@ set.seed(12345)
 N=10000
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
-N=10000
+N=50000
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
-N=10000
+N=50000
 svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
 N=10000

### 2021-08-08 Real Dataset Script.R
commit

@@ -325,10 +325,10 @@ tau   = sqrt(tau2)
 #Real dataset
 
 library(openxlsx)
-realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
 y<-realdatasetSPX[,2]
 n = length(y)
-volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
 x<-volatilitySPX[,3]
 alpha.true = alpha
 beta.true  = beta
@@ -346,10 +346,10 @@ set.seed(12345)
 N=10000
 svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
-N=50000
+N=10000
 svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
-N=50000
+N=10000
 svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
 set.seed(12345)
 N=10000

### 2021-08-09 Real Dataset SV.Rmd
Aggiornamento tabelle e commenti

@@
 rownames(SC)<-c("Mean","Standard Deviation")
-colnames(SC) <- c("Dow Jones", "S&P500", "STOXX50E")
+colnames(SC) <- c("S&P500","Dow Jones","STOXX50E")
 
@@
 rownames(Errorcomp3)<-c("N=50000", "N=100000")
-Errorcomp3[1,]<-c(0.04407704,0.01667898)
-Errorcomp3[2,]<-c(0.03575915,0.01476220)
+Errorcomp3[1,]<-c(0.015181843,0.008679359)
+Errorcomp3[2,]<-c(0.013735482,0.006467285) #updated
 ```
@@
 Overall, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison, implying different target distributions.
-For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
+For example, running the same experiment using the STOXX50E dataset, we have similar differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.016121989 and mean absolute difference equal to 0.009306925.
+The extent of the difference clearly depends on the chosen parameters too: for example, at N=50000, choosing $\alpha=-0.0031$, $\beta=0.9951$ and $\tau^2=0.0074$, using SP500 data yields RMS difference equal to 0.04408 and MA difference equal to 0.01668, while using STOXX50E data we have 0.01236279 RMS difference and 0.008028818 mean absolute difference.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
 First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
@@
 ```{r include=FALSE}
+# RMSE MAE comparison
+
+library(openxlsx)
 Errorcomp3<-matrix(NA,ncol=7,nrow=6)
@@
 Errorcomp3[,1]<-c(10000,10000,10000,10000,10000,10000)
-Errorcomp3[,2]<-c(0.508642137861919,0.353767519133916,0.463584874517798,0.301898325469172,0.507624385776376,0.337543646258033)
-Errorcomp3[,3]<-c(0.509256032254369,0.35314674168209,0.463446053320389,0.304178601158929,0.506222076244991,0.338268970683459)
-Errorcomp3[,4]<-c(0.509943715852212,0.351738603428736,0.475708272485282,0.30422270387141,0.509343139732461,0.337449244130176)
-Errorcomp3[,5]<-c(0.510334461444389,0.35305864294958,0.46783471248187,0.304213895457104,0.505360139532104,0.3361502165818)
-Errorcomp3[,6]<-c(0.428716693039206,0.291557534456417,0.344065170120452,0.237159281665533,0.440832171188778,0.280952286978564)
-Errorcomp3[,7]<-c(0.602883459334495,0.431972177956328,0.652533564914959,0.440767074642285,0.604672122750974,0.415631855223678)
+
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x1<-volatilitySPX[,3]
+realisedx1<-x1
+sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp(fun/2)}
+Errorcomp3[1,2]<-RMSE(realisedx1,comparablevol(sv_bpf1))
+Errorcomp3[1,3]<-RMSE(realisedx1,comparablevol(sv_gpfopt1))
+Errorcomp3[1,4]<-RMSE(realisedx1,comparablevol(sv_apf1))
+Errorcomp3[1,5]<-RMSE(realisedx1,comparablevol(sv_bapf1))
+Errorcomp3[1,6]<-RMSE(realisedx1,comparablevol(sv_lw1))
+Errorcomp3[1,7]<-RMSE(realisedx1,comparablevol(sv_sis1))
+Errorcomp3[2,2]<-MAE(realisedx1,comparablevol(sv_bpf1))
+Errorcomp3[2,3]<-MAE(realisedx1,comparablevol(sv_gpfopt1))
+Errorcomp3[2,4]<-MAE(realisedx1,comparablevol(sv_apf1))
+Errorcomp3[2,5]<-MAE(realisedx1,comparablevol(sv_bapf1))
+Errorcomp3[2,6]<-MAE(realisedx1,comparablevol(sv_lw1))
+Errorcomp3[2,7]<-MAE(realisedx1,comparablevol(sv_sis1))
+
+volatilitySTOXX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+x2<-volatilitySTOXX[,3]
+realisedx2<-x2
+sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp(fun/2)}
+Errorcomp3[3,2]<-RMSE(realisedx2,comparablevol(sv_bpf2))
+Errorcomp3[3,3]<-RMSE(realisedx2,comparablevol(sv_gpfopt2))
+Errorcomp3[3,4]<-RMSE(realisedx2,comparablevol(sv_apf2))
+Errorcomp3[3,5]<-RMSE(realisedx2,comparablevol(sv_bapf2))
+Errorcomp3[3,6]<-RMSE(realisedx2,comparablevol(sv_lw2))
+Errorcomp3[3,7]<-RMSE(realisedx2,comparablevol(sv_sis2))
+Errorcomp3[4,2]<-MAE(realisedx2,comparablevol(sv_bpf2))
+Errorcomp3[4,3]<-MAE(realisedx2,comparablevol(sv_gpfopt2))
+Errorcomp3[4,4]<-MAE(realisedx2,comparablevol(sv_apf2))
+Errorcomp3[4,5]<-MAE(realisedx2,comparablevol(sv_bapf2))
+Errorcomp3[4,6]<-MAE(realisedx2,comparablevol(sv_lw2))
+Errorcomp3[4,7]<-MAE(realisedx2,comparablevol(sv_sis2))
+
+volatilityDJI<-read.xlsx("Dataset 17-21.xlsx", sheet = 3)
+x3<-volatilityDJI[,3]
+realisedx3<-x3
+sv_bpf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 6))[,1]
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp(fun/2)}
+Errorcomp3[5,2]<-RMSE(realisedx3,comparablevol(sv_bpf3))
+Errorcomp3[5,3]<-RMSE(realisedx3,comparablevol(sv_gpfopt3))
+Errorcomp3[5,4]<-RMSE(realisedx3,comparablevol(sv_apf3))
+Errorcomp3[5,5]<-RMSE(realisedx3,comparablevol(sv_bapf3))
+Errorcomp3[5,6]<-RMSE(realisedx3,comparablevol(sv_lw3))
+Errorcomp3[5,7]<-RMSE(realisedx3,comparablevol(sv_sis3))
+Errorcomp3[6,2]<-MAE(realisedx3,comparablevol(sv_bpf3))
+Errorcomp3[6,3]<-MAE(realisedx3,comparablevol(sv_gpfopt3))
+Errorcomp3[6,4]<-MAE(realisedx3,comparablevol(sv_apf3))
+Errorcomp3[6,5]<-MAE(realisedx3,comparablevol(sv_bapf3))
+Errorcomp3[6,6]<-MAE(realisedx3,comparablevol(sv_lw3))
+Errorcomp3[6,7]<-MAE(realisedx3,comparablevol(sv_sis3))
 
@@
 In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter.
+COMMENTI DA AGGIORNARE
 
 ```{r include=FALSE}
+
+library(openxlsx)
 Errorcomp4<-matrix(NA,ncol=4,nrow=4)
@@
 Errorcomp4[,1]<-c(50000,50000,50000,50000)
-Errorcomp4[,2]<-c(0.511700619467058,0.354186384429001,0.4650284,0.3034359)
-Errorcomp4[,3]<-c(0.514888164962505,0.355336840861771,0.4670426,0.3040741)
-Errorcomp4[,4]<-c(0.4242305,0.2906932,0.3422955,0.2395650)
+
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x1<-volatilitySPX[,3]
+realisedx1<-x1
+sv_bpf1_5<-(read.xlsx("svbpf50000p.xlsx", sheet = 1))[,1]
+sv_apf1_5<-(read.xlsx("svapf50000p.xlsx", sheet = 1))[,1]
+sv_lw1_5<-(read.xlsx("svlw50000p.xlsx", sheet = 1))[,1]
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp(fun/2)}
+Errorcomp4[1,2]<-RMSE(realisedx1,comparablevol(sv_bpf1_5))
+Errorcomp4[1,3]<-RMSE(realisedx1,comparablevol(sv_apf1_5))
+Errorcomp4[1,4]<-RMSE(realisedx1,comparablevol(sv_lw1_5))
+Errorcomp4[2,2]<-MAE(realisedx1,comparablevol(sv_bpf1_5))
+Errorcomp4[2,3]<-MAE(realisedx1,comparablevol(sv_apf1_5))
+Errorcomp4[2,4]<-MAE(realisedx1,comparablevol(sv_lw1_5))
+
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+x2<-volatilitySPX[,3]
+realisedx2<-x2
+sv_bpf2_5<-(read.xlsx("svbpf50000p.xlsx", sheet = 2))[,1]
+sv_apf2_5<-(read.xlsx("svapf50000p.xlsx", sheet = 2))[,1]
+sv_lw2_5<-(read.xlsx("svlw50000p.xlsx", sheet = 2))[,1]
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp(fun/2)}
+Errorcomp4[3,2]<-RMSE(realisedx2,comparablevol(sv_bpf2_5))
+Errorcomp4[3,3]<-RMSE(realisedx2,comparablevol(sv_apf2_5))
+Errorcomp4[3,4]<-RMSE(realisedx2,comparablevol(sv_lw2_5))
+Errorcomp4[4,2]<-MAE(realisedx2,comparablevol(sv_bpf2_5))
+Errorcomp4[4,3]<-MAE(realisedx2,comparablevol(sv_apf2_5))
+Errorcomp4[4,4]<-MAE(realisedx2,comparablevol(sv_lw2_5))
 

### 2021-08-09 prova_knit_finale.Rmd
Aggiornamento tabelle e commenti

@@
 We provide both a graphical and a numerical comparison.\footnote{The sequences of filtered states shown in the graphs are obtained by using filters with 10000 particles.} From the plots it is hard to say which particle filter provides the best performances.
-By and large, BPF, GPF and APF behave similarly, while the Liu and West filter presents larger credible intervals but at the same time a higher coverage of the true states.
-On the other hand, by a RMSE comparison, the Auxiliary Particle filter provides the best performances followed bt the Liu and West filter.
-Moreover, notice that when the number of particles increases the accuracy improves, however not significant improvements are detected for $N>10000$.
+By and large, BPF, GPF and APF behave similarly, while the Liu and West filter presents larger credible intervals but at the same time a higher coverage of the true states.\
+On the other hand, by a RMSE comparison, we see several differences between the filters.
+Note that the Guided Particle Filter yields the lowest RMSE at N=100, and does not improve much even when raising the number of particles to 100000.
+Instead, the APF achieves the best performance in the field when N rises to 10000 and 100000, despite lagging well behind the BPF and the GPF for the three initial N thresholds.
+For the Bootstrap PF, the most significant improvements is that from N=100 to N=1000.
+We could then point at the APF as the best performing and more computationally expensive filter, while the GPF as the best one at low thresholds (\textit{i.e.} least computationally expensive, in a way).
+However, fine margins suggest caution in such comparative assessments.\
+Rather, we would stress the performance of the Liu and West filter: indeed, it is apparent that, at least for the first four thresholds, such filter delivers the worst performance.
+Even at N=100000, the APF, the GPF and the BPF are much closer to each other than the LWF, which still yields the highest RMSE.
+Such results, in fact, do not come as a surprise: indeed, note that, while the three best performing filters are built on a model that perfectly coincides with the true DGP, the Liu and West filter considers a somewhat misspecified model, which features additional uncertainty on several parameters.
+Intuitively, even though the LWF features some degree of parameter learning, and would then converge to the correct figures of such parameters, the other filters "have already learnt" the true parameters from the beginning.\footnote{Such difference can be represented as follows.
+When trying to pin down real valued parameters, the probability that a simple initial guess for each value is correct can be deemed as null.
+So, in contexts where parameters are unknown, we expect the additional uncertainty featured in the LWF to prove useful, as it can adjust the initial parameters to get closer to the true unknown values.
+However, in synthetic contexts where we know the DGP, such uncertainty is almost useless.}\
+Finally, notice that in general when the number of particles increases the accuracy improves; however, significant improvements are not detected for $N>10000$, suggesting that such threshold may be a telling one as regards the convergence of our filters.
 

### 2021-08-09 prova_knit_finale.Rmd
convergence

--- 2021-08-07 prova_knit_finale.Rmd
@@
 \section{Convergence}\label{pf_converg}
+In this section, we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest.
+Firstly, we present two convergence results for the estimators of the predictive and filtering distributions.
+As we will discuss, these results hold under reasonable conditions.
+Then, we will show that the error of the particle estimators are asymptotically distributed according to a Gaussian distribution with known variance.
+Henceforth, we will maintain the assumption that (multinomial) resampling is repeated at each iteration.
+\subsection{MSE convergence}
+As already seen in subsection \ref{seq_impt_sampling}, the particle approximation of $\Q_{t}(dx_{0:t})$ is:
+\begin{equation*}
+      \sum_{n=1}^NW_{t-1}^nM_t(X_{t-1}^n,dx_t).
+\end{equation*}
+The following lemma states that the (mean squared) error related to sampling from this conditional distribution is bounded.
+\bigskip
+\begin{lemma}
+For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
+$$
+\E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg\}^2 \Bigg]\leq \frac{\|\varphi\|_{\infty}^2}{N},
+$$
+where $\|\cdot\|_{\infty}$ is the supnorm and $M_t\varphi$ is the map 
+$x_{t-1}\mapsto \int_{\X}M_t(x_{t-1},dx_t)\varphi(x_t).$
+\end{lemma}
+This result can be used to prove the convergence of the estimator $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$. To see this, note that 
+\begin{equation}
+    \begin{split}
+        \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)=\bigg(&\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg) \\
+        & + \bigg(\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n)- \E_{\Q_{t-1}M_t}(\varphi)\bigg).
+    \end{split}
+\end{equation}
+where $\E_{\Q_{t-1}M_t}(\varphi)=\int_{\X^{t+1}}\varphi(x_t)  \Q_{t-1}(dx_{0:t-1})M_t(x_{t-1},dx_t)$.\
+By the lemma, the first component of the right-hand side has a bounded MSE.
+It can be proved that also the second component of the right-hand side has a bounded MSE.
+Using the fact that $\E\{(X+Y)^2\}\leq2\E\{\E(X^2)+\E(Y^2)\},$ it follows that also the MSE of the left-hand side is bounded.
+A simple induction argument proves that the result holds for all $t$.\
+Using a similar argument, one can prove that the MSE of the estimator $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ is bounded as well. These ideas are formalized in the following proposition.
+\bigskip
+\begin{proposition}
+For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$ $\forall t\geq0$ there exist constants $c_t,c_t'$ such that:
+$$
+\E \Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg\}^2 \Bigg]\leq c_t \frac{\|\varphi\|_{\infty}^2}{N}
+$$
+and 
+$$
+    \E \Bigg[ \bigg\{\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_t}(\varphi) \bigg\}^2 \Bigg]\leq c_t' \frac{\|\varphi\|_{\infty}^2}{N}.
+$$
+Therefore, taking the limit for $N\rightarrow \infty,$ we have that 
+$$
+\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_{t-1}M_t}(\varphi)
+$$
+and 
+$$
+\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_t}(\varphi)
+$$
+\end{proposition}
+In the next subsection, we establish a much stronger property of the particle filter algorithms.
+
+\subsection{Almost sure convergence}
+The following proposition states that estimators $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$ and $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ converge to their true values, up to negligible sets, i.e., the sets such that the estimators do not converge to $\E_{\Q_{t-1}M_t}(\varphi)$ and $\E_{\Q_t}(\varphi)$ have zero probability.
+Additionally, recall that almost sure convergence implies convergence in probability.
+\bigskip
+\begin{proposition}
+For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
+$$
+\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{a.s}}\E_{\Q_{t-1}M_t}(\varphi)
+$$
+and 
+$$
+\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{a.s}}\E_{\Q_{t}}(\varphi).
+$$
+\end{proposition}
+
+\subsection{Central limit theorem}
+The last asymptotic result of PF estimators concerns convergence in law: the error of a particle estimate is asymptotically normally distributed. 
+\bigskip
+\begin{proposition}
+For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
+$$
+\sqrt{N}\bigg( \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg)\implies \N(0,\hat{\mathcal{V}}_t(\varphi))
+$$
+and 
+$$
+\sqrt{N}\bigg(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_{t}}(\varphi)\bigg)\implies \N(0,\mathcal{V}_t(\varphi)),
+$$
+where the asymptotic variances are defined recursively, with $ \hat{\mathcal{V}}_0(\varphi):=Var_{\M_0}(\varphi) $,
+\begin{equation*}
+    \begin{split}
+        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\E_{M_t}(\varphi))+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
+        \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\E_{\Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
+    \end{split}
+\end{equation*}
+Here, $\bar{G}_t=G_t/\E_{\Q_{t-1}M_t}(G_t)$ is the normalized incremental weight.
+\end{proposition}
+This version of the CLT for PF suggest that PF estimators with respect both to the predictive and the filtering distributions are approximately normal for a large number of particles. \
+In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{asympt_IS}.
+In fact, recall that the estimator with respect to the filtering distribution is a more sophisticated form of IS estimator, where the target distribution differs from the proposal one.
+\
+Instead, to get an intuition of the variance $\hat{\mathcal{V}}_t(\varphi),$ note that by the law of total variance, 
+\begin{equation*}
+    \begin{split}
+        Var\Bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n) \Bigg] = &Var\Bigg[ \E\bigg[\frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)\Big| \sigma(X_{t-1}^{1:N}) \bigg] \Bigg] \\
+        &+ \E \Bigg[ Var\bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)    \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg] \\
+        &= Var\bigg[\sum\limits_{n=1}^N W_{t-1}^n (M_t\varphi)(X_{t-1}^n)\bigg] \\
+        &+\frac{1}{N}\E\Bigg[Var\bigg[\varphi(X_t^1) \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg].
+    \end{split}
+\end{equation*}
+The notation $\mid \sigma(X_{t-1}^{1:N})$ means conditioning on having sampled (from the multinomial distribution) $X_{t-1}^{1:N}$.
+Conditional on multinomial sampling, the $X_{t-1}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$.
+So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$.
+These results are of practical use because they provide a way to explicitly compute the variance of the PF estimates.
 

### 2021-08-09 CI/CI.symmetric.R
Copied from before_commit branch

@@ -1,44 +1,0 @@
-F_order<-function(xs,ws){
-  t=dim(xs)[1]
-  n=dim(xs)[2]
-  F_order=matrix(NA,t,n)
-  for (i_t in 1:t){
-    order=order(xs[i_t,])
-    ws.o=ws[i_t,][order]
-    F_order[i_t,1]=ws.o[1]
-    for (i_n in 1:(n-1)){
-      F_order[i_t,i_n+1]=F_order[i_t,i_n]+ws.o[i_n+1]
-    }
-  }
-  return(F_order)
-}
-
-CI.symmetric<-function(xs,ws,p=0.05){
-  t=dim(xs)[1]
-  n=dim(xs)[2]
-  CI=matrix(NA,t,2)
-  F_order=F_order(xs,ws)
-  lower.mat=(F_order>p/2)
-  higher.mat=(F_order>1-p/2)
-  for (i_t in 1:t){
-    xs.o=xs[i_t,][order(xs[i_t,])]
-    if (lower.mat[i_t,1]==T){
-      lower=xs.o[1]
-    }else{
-      for (i_n in 1:(n-1)){
-        if (lower.mat[i_t,i_n+1]==T & lower.mat[i_t,i_n]==F){
-          lower=(xs.o[i_n]+xs.o[i_n+1])/2
-        }
-      }
-    }
-    
-    for (i_n in (n-1):1){
-      if (higher.mat[i_t,i_n+1]==T & higher.mat[i_t,i_n]==F){
-        higher=(xs.o[i_n+1]+xs.o[i_n])/2
-      }
-    }
-
-    CI[i_t,]=c(lower,higher)
-  }
-  return(CI)
-}

### 2021-08-09 Real Dataset SV.Rmd
grafici realised volatility

--- 2021-08-09 Real Dataset SV.Rmd
@@
 
-Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results and, generally speaking, to underestimate the volatility (as indicated by the realised volatility proxy).
+Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results.
 The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}).
@@
 
-GRAFICI
+```{r echo=FALSE, fig.align='center', fig.cap="Filtered volatility and realised volatility", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+# Realized Volatility Plots
+
+library(openxlsx)
+library(ggplot2)
+library(ggpubr)
+
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y<-realdatasetSPX[,2]
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x1<-volatilitySPX[,3]
+realisedx1<-x1
+timeframe<-c(1:length(y))
+
+sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
+dfsv1<-data.frame(timeframe,y,realisedx1,sv_bpf1)
+dfsv2<-data.frame(timeframe,y,realisedx1,sv_gpfopt1)
+dfsv3<-data.frame(timeframe,y,realisedx1,sv_apf1)
+dfsv4<-data.frame(timeframe,y,realisedx1,sv_lw1)
+dfsv5<-data.frame(timeframe,y,realisedx1,sv_bapf1)
+dfsv6<-data.frame(timeframe,y,realisedx1,sv_sis1)
+
+Filtplot1<-function(dataframe,title){
+  mean=sv_bpf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot2<-function(dataframe,title){
+  mean=sv_gpfopt1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot3<-function(dataframe,title){
+  mean=sv_apf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot4<-function(dataframe,title){
+  mean=sv_lw1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot5<-function(dataframe,title){
+  mean=sv_bapf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot6<-function(dataframe,title){
+  mean=sv_sis1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotproxy1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
+plotproxy2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
+plotproxy3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
+plotproxy4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
+plotproxy5<-Filtplot5(dfsv5,"Always Resampling SP500")
+plotproxy6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+plotproxy1
+plotproxy2
+plotproxy3
+plotproxy4
+plotproxy5
+plotproxy6
+
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+y<-realdatasetSPX[,2]
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+x2<-volatilitySPX[,3]
+realisedx2<-x2
+timeframe<-c(1:length(y))
+
+sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
+dfsv7<-data.frame(timeframe,y,realisedx2,sv_bpf2)
+dfsv8<-data.frame(timeframe,y,realisedx2,sv_gpfopt2)
+dfsv9<-data.frame(timeframe,y,realisedx2,sv_apf2)
+dfsv10<-data.frame(timeframe,y,realisedx2,sv_lw2)
+dfsv11<-data.frame(timeframe,y,realisedx2,sv_bapf2)
+dfsv12<-data.frame(timeframe,y,realisedx2,sv_sis2)
+
+Filtplot7<-function(dataframe,title){
+  mean=sv_bpf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot8<-function(dataframe,title){
+  mean=sv_gpfopt2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot9<-function(dataframe,title){
+  mean=sv_apf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot10<-function(dataframe,title){
+  mean=sv_lw2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot11<-function(dataframe,title){
+  mean=sv_bapf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot12<-function(dataframe,title){
+  mean=sv_sis2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Proxy" = "black",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Proxy" = 3,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotproxy7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX")
+plotproxy8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX")
+plotproxy9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX")
+plotproxy10<-Filtplot10(dfsv10,"Liu and West Filter STOXX")
+plotproxy11<-Filtplot11(dfsv11,"Always Resampling STOXX")
+plotproxy12<-Filtplot12(dfsv12,"No Resampling STOXX")
+
+plotproxy7
+plotproxy8
+plotproxy9
+plotproxy10
+plotproxy11
+plotproxy12
+
+```
 
@@
 We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
-In the case of the SP500 index, although the number of particles increases, the RMSE and MAE increase (although by very small amounts) for both the particle filters based on our usual specification of the SV model.
-Instead, for the Liu and West filter we see some improvements, although the small extent suggests that also the LW filter was already in the convergence phase with N=10000.\
-In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter.
-COMMENTI DA AGGIORNARE
+In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF.
+Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a RMSE and a MAE.
+In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500.
+A possible interpretation goes, once again, as that reported in the main text.}\
+In the case of the EuroStoxx50 index, while the Bootstrap filter slightly increases both RMSE and MAE when N is raised to 50000, the APF shows mild improvements in its MAE performance, while the RMSE criterion points in the opposite direction.
+The criteria disagree also in the case of the LW filter, as the MAE mildly prefers the performance of N=10000, while the RMSE detects minor improvements for N=50000.
 

### 2021-08-09 CI/CI.symmetric.R
adding CI folder with function to estimate a symmetric confidence interval from empirical distribution of weighted particles just in case and to have it since it's done

@@ -1,0 +1,44 @@
+F_order<-function(xs,ws){
+  t=dim(xs)[1]
+  n=dim(xs)[2]
+  F_order=matrix(NA,t,n)
+  for (i_t in 1:t){
+    order=order(xs[i_t,])
+    ws.o=ws[i_t,][order]
+    F_order[i_t,1]=ws.o[1]
+    for (i_n in 1:(n-1)){
+      F_order[i_t,i_n+1]=F_order[i_t,i_n]+ws.o[i_n+1]
+    }
+  }
+  return(F_order)
+}
+
+CI.symmetric<-function(xs,ws,p=0.05){
+  t=dim(xs)[1]
+  n=dim(xs)[2]
+  CI=matrix(NA,t,2)
+  F_order=F_order(xs,ws)
+  lower.mat=(F_order>p/2)
+  higher.mat=(F_order>1-p/2)
+  for (i_t in 1:t){
+    xs.o=xs[i_t,][order(xs[i_t,])]
+    if (lower.mat[i_t,1]==T){
+      lower=xs.o[1]
+    }else{
+      for (i_n in 1:(n-1)){
+        if (lower.mat[i_t,i_n+1]==T & lower.mat[i_t,i_n]==F){
+          lower=(xs.o[i_n]+xs.o[i_n+1])/2
+        }
+      }
+    }
+    
+    for (i_n in (n-1):1){
+      if (higher.mat[i_t,i_n+1]==T & higher.mat[i_t,i_n]==F){
+        higher=(xs.o[i_n+1]+xs.o[i_n])/2
+      }
+    }
+
+    CI[i_t,]=c(lower,higher)
+  }
+  return(CI)
+}

### 2021-08-10 prova_knit_finale.Rmd
small change
changes

@@
 - \usepackage{mathtools}
+- \usepackage{appendix}
 - \theoremstyle{break}
@@
     \item Sample $N$ values from the proposal measure: $X^n\thicksim \M$ for $n=1,...,N$.
-    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
+    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$.
     \item Compute the estimator:
@@
     \begin{equation*}
-        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}
+        W^n:=\frac{w(X^n)}{\sum_{m=1}^Nw(X^m)}.
     \end{equation*}
@@
     \begin{equation*}
-        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n
+        \hat \varphi_{AN}=\sum_{n=1}^N\varphi (X^n)W^n.
     \end{equation*}
@@
 \end{equation}
-Furthermore, the IS estimator is the expected value of the function $\varphi$ with respect to the approximating probability measure $\Q^N$:
+Then, the auto-normalized IS estimator $\hat\varphi_{AN}$ is the expected value of the function $\varphi$ with respect to the approximating probability measure $\Q^N$:
 \begin{equation*}
@@
 \subsubsection{Convergence of the IS estimator}\label{pf_is_conv}
-
+We can show that the auto-normalized IS estimator is both consistent and asymptotically normal.\
 Consider the auto-normalized IS estimator $\hat \varphi= \sum_{n=1}^N\varphi (X^n)W^n$. We can write:
@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \E_{\M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow^{\text{a.s.}} \E_{\M}(w) \label{is_conv_lln}
 \end{equation}
@@
 \end{equation}
-Therefore, the IS (auto-normalized) estimator for $\varphi$ is both consistent and asymptotically normal.
+Therefore, the IS (auto-normalized) estimator for $\varphi$ is asymptotically normal. Consistency can be proved in an analogous way, applying the LLN and lemma \ref{lemma_imp_sam_1}:
+\begin{equation}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+\end{equation}
+Together with \eqref{is_conv_lln}, we get that:
+\begin{equation}
+   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\Q(\varphi)
+\end{equation}
 
@@
 This version of the CLT for PF suggest that PF estimators with respect both to the predictive and the filtering distributions are approximately normal for a large number of particles. \
-In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{asympt_IS}.
+In particular, the logic behind the expression for $\mathcal{V}_t(\varphi)$ is the same as described in section \ref{pf_is_conv}.
 In fact, recall that the estimator with respect to the filtering distribution is a more sophisticated form of IS estimator, where the target distribution differs from the proposal one.
@@
 
-\chapter{Appendix}
-
-\section{The Stochastic Volatility Model}
+\appendix
+\chapter{The Stochastic Volatility Model}
 

### 2021-08-10 prova_knit_finale.Rmd
change

@@
 \end{equation}
-Therefore, the IS (auto-normalized) estimator for $\varphi$ is asymptotically normal. Consistency can be proved in an analogous way, applying the LLN and lemma \ref{lemma_imp_sam_1}:
+Therefore, the IS (auto-normalized) estimator for $\varphi$ is asymptotically normal. Consistency can be proved in an analogous way, applying the strong LLN and lemma \ref{lemma_imp_sam_1}:
 \begin{equation}
@@
 \end{equation}
-Together with \eqref{is_conv_lln}, we get that:
+Together with \eqref{is_conv_lln}, we get strong convergence of the estimator (that implies consistency).
 \begin{equation}

### 2021-08-10 prova_knit_finale.Rmd
change

@@
 Note that, after resampling occurred, each particle $\tilde X_{0:1}^n:=(X_0^{A_1^n},X_1^n)$ has the same probability weight.\
-To understand the reason of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$.
+To understand the logic of the previous statement, let us assume that both $\Q_0$ and $\M_0$ admit densities $q_0$, $m_0$ with respect to the same dominating measure $\nu(dx_0)$ and that the transition kernel $M_1$ also admits the density $m_1$.
 Consider the sequential updating for the weights:
 \begin{equation}
-    w^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
+    W^n_1=\frac{\Q_1(dx_{0:1})}{\M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
 =\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
@@
 \end{equation}
-Equation \eqref{sis_update} suggests, intuitively, that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue.
+Intuitively, equation \eqref{sis_update} suggests that over time new variance is added from the incremental weights (i.e., the updating factor $q_t( x_t|x_{t-1}^n)/m_t( x_t|x_{t-1}^n)$) determining the curse of dimensionality issue.
 Instead, when resampling takes place, since the past weights $w_{t-1}$ are reset to be equal across them all the variance of $w_t$ inherited from the past is shut-down and the only source of variability at each stage comes from the incremental weights.

### 2021-08-10 prova_knit_finale.Rmd
remove appendix

@@
 - \usepackage{mathtools}
-- \usepackage{appendix}
 - \theoremstyle{break}
@@
 
-\appendix
-\chapter{The Stochastic Volatility Model}
+\chapter*{Appendix}
+\section{The Stochastic Volatility Model}
 

### 2021-08-10 prova_knit_finale.Rmd
correct

@@
 "Dead" particles (i.e., those with small weights) are  likely to be removed when resampling takes place and past weights get ``reset" (usually, they are set to be equal to $1/N$).
-Otherwise, the \textit{curse of dimensionality} problem will emerge and the variance of the weights will diverge over time as the variance of past weights increases as new updates occur period-by-period.
+Otherwise, the \textit{curse of dimensionality} problem will emerge and the variance of the weights will diverge over time as new updates occur period-by-period.
 Next, we show an example of a PF where SIS is used: note that the $ESS$ drops dramatically after few periods, meaning that the variance of the importance weights diverges.

### 2021-08-10 prova_knit_finale.Rmd
elimanato 1000000 particles da rimettere

@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow^{\text{a.s.}} \E_{\M}(w) \label{is_conv_lln}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n)\xrightarrow{\text{a.s.}} \E_{\M}(w) \label{is_conv_lln}
 \end{equation}
@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
 \end{equation}
@@
 \begin{equation}
-   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\Q(\varphi)
+   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\Q(\varphi)
 \end{equation}
@@
 
-```{r include=FALSE}
+```{r eval=FALSE, include=FALSE}
 truex<-x
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, results="asis"}
 kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=4)

### 2021-08-10 prova_knit_finale.Rmd
arrow

--- 2021-08-10 prova_knit_finale.Rmd
@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrigtharrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
 \end{equation}
@@
 \begin{equation}
-   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\rigtharrow^{\text{a.s.}}  \E_\Q(\varphi)
+   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\xrigtharrow{\text{a.s.}}  \E_\Q(\varphi)
 \end{equation}
@@
 \subsection{Implementation}
-Consider again the random walk plus noise model of section \ref{implementation},the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
+Consider again the random walk plus noise model of section \ref{implementation}, the challenge here is to estimate the filtered states using a Sequential Importance Sampling algorithm.
 The main idea of the SIS applied to this univariate linear gaussian model is described in the following algorithm.

### 2021-08-10 prova_knit_finale.Rmd
change

--- 2021-08-10 prova_knit_finale.Rmd
@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrigtharrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
 \end{equation}

### 2021-08-10 prova_knit_finale.tex
brutti file

@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \usepackage{xcolor}
+\usepackage{mathtools}
 \theoremstyle{break}
@@
 \usepackage{xcolor}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 
 \title{\Huge State-Space Models and Particle Filters}
-\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei
-Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta
-Bruni}
-\date{21 luglio, 2021}
+\author{Edoardo Marcelli \and Roberto Colarieti \and Alexei Verkhovtsev \and Michele Bolla \and Gabriele Romano \and Benedetta Bruni}
+\date{10 Agosto, 2021}
 
@@
 \mainmatter
+\tableofcontents
+
 \chapter*{Introduction}
 
-\chapter{Markov Processes and State-Space Models}
+State-Space models represent an extremely interesting class of models
+and they are used extensively, for instance, for inference and
+forecasting in time-series and macroeconometric analysis. In fact, they
+allow to model observational data as function of underlying states that
+cannot be observed. For instance, in macroeconomic models GDP is usually
+observed while the TFP sequence is the hidden state determining
+output.\\
+In this work we describe techniques, such as Kalman and Particle filter,
+that allow to approximate probability distributions for the states from
+the observed data. For example, from the sequence of observed GDP, it is
+possible to ``filter'' the TFP factor.\\
+Furthermore, it is possible to impose a Markov structure to State-Space
+models, assuming that the states are Markov, i.e., the probability of
+the current state depends on the entire sequence of past states only
+through its first lag. The combination of State-Space and Markov
+structures constitutes a powerful modeling assumption for inference and
+forecasting. ~
+
+Within the Monte Carlo (MC) methods, where inference is based on
+repeated simulations that allow to approximate the target distributions
+(needed to compute the estimator of interest), Particle Filter (PF)
+Algorithms represent the most important class of procedures for
+State-Space-Markov models. PF allows to sample states period-by-period,
+updating sequentially the filtering distributions (i.e., the
+distribution of current state conditional on all the sequence of
+observations - up to the current period one -), incorporating the
+information on the likelihood of the extracted states given the sequence
+of observations, by means of re-weighting.\\
+Moreover, the PF is a very general algorithm, since it can be applied to
+non-linear models with non-Gaussian errors, while it does not allow for
+a closed form solution. On the contrary, the Kalman Filter is a
+filtering method that can be applied in the case of linear-Gaussian
+State-Space-Markov models, and remains analytically tractable.~ In this
+work we will describe both these algorithms, while our main focus
+remains on Particle Filter. ~
+
+In chapter \ref{MP_SSM_chapter} we define precisely Markov processes and
+State-Space Models. In chapter \ref{KF_chapter} we describe the Kalman
+Filter and in chapter \ref{PF_chapter} we analyze extensively the
+Particle Filter. Finally, in chapter \ref{application_chapter}, we apply
+these methods to stochastic volatility models.\\
+Finally, we find it useful to integrate each section where an algorithm
+is described and analyzed from a theoretical perspective with an
+implementation subsection, using as an example a simple Random-Walk
+State-Space model. In this way, the reader can visualize the codes and
+the practical meaning of the Kalman and Particle Filter algorithms,
+immediately after their theoretical description.
+
+\chapter{Markov Processes and State-Space Models}\label{MP_SSM_chapter}
 
@@
 \((\mathcal{Y}, \mathcal{B}(\mathcal{Y}))\) be two (eventually equal)
-measurable spaces.
+measurable spaces.\\
 
@@
 
-\section{State-Space Models}
+\section{State-Space Models}\label{sec_state_space}
 
@@
 \end{equation} Integrating away \(X_{0:T},\) one gets the marginal
-density \begin{align*}\label{state-space-marginal-y}
-  \mathbb{P}_T(dy_{0:T}) &= \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s) \\
-                &= p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
-\end{align*}\\
+density \begin{equation}\label{state-space-marginal-y}
+  \mathbb{P}_T(dy_{0:T}) = \mathbb E_{\mathbb{P}_{t}}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big] \prod\limits_{s=0}^t\nu(dy_s)
+                = p_t(y_{0:t})\prod\limits_{s=0}^t\nu(dy_s),
+\end{equation}\\
 where \(p_t(y_{0:t})\) is the likelihood function. Dividing the
@@
 equation \ref{state-space-densities} by the right-hand side of equation
-\ref{state-space-marginal-y}, one obtains the filtering distribution
+\ref{state-space-marginal-y}, one obtains the distribution
 \begin{equation}
-\mathbb{P}_t(X_{t}\in dx_{t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
+\mathbb{P}_t(X_{0:t}\in dx_{0:t}\mid Y_{0:t}=y_{0:t})=\frac{1}{p_t(y_{0:t})}\big[\prod\limits_{s=0}^t f_s(y_s\mid X_s) \big]\mathbb{P}_t(dx_{0:t}).
 \end{equation}
 
-\chapter{Linear-Gaussian case}
+\chapter{Linear-Gaussian case}\label{KF_chapter}
 
@@
 \(U_t\thicksim\mathcal N(0,\Sigma_t)\) and
-\(V_t\thicksim\mathcal N(0,R_t)\) be two independent white-noise errors
-and assume the initial distribution
+\(V_t\thicksim\mathcal N(0,\Lambda_t)\) be two independent white-noise
+errors and assume the initial distribution
 \(X_0\thicksim\mathcal N(0,\Sigma_0).\) We have a linear Gaussian model
@@
 \begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\mathcal N(m_0,Q_0)$ and $Y\mid X\thicksim\mathcal N(BX,R)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Let $X\thicksim\mathcal N(m_0,C_0)$ and $Y\mid X\thicksim\mathcal N(BX,\Lambda)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
   Then, it holds that
   \begin{align}
-    X\mid Y=y&\thicksim\mathcal N(m_1,Q_1),\\
-    Q_1&= Q_0 (I_{d_x} - B'(B Q_0 B'+ R)^{-1} B Q_0) \\
-    m_1&=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
+    X\mid Y=y&\thicksim\mathcal N(m_1,C_1),\\
+    C_1&= C_0 (I_{d_x} - B'(B C_0 B'+ \Lambda)^{-1} B C_0) \\
+    m_1&=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
   \end{align}
@@
 \begin{proof*}
-   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, Q_0).$ It is easy to see that
+   The lemma can be proved using Bayes' Theorem. Suppose $X\thicksim\mathcal N(m_0, C_0).$ It is easy to see that
 \begin{equation} \label{normal-distrib-rewrite}
-        p(x)\propto \text{exp}\Big(-\frac{1}{2}X'AX + X'B\Big),
+        p(x)\propto \text{exp}\Big(-\frac{1}{2}x'C_0^{-1}x + x'C_0^{-1}m_0\Big).
 \end{equation}
-with $A=Q_0^{-1}$ and $B=Am_0.$
 Now, by Bayes' Theorem,
@@
         p(x\mid Y=y)  &\propto p(x) p(y\mid x) \\
-                    &\propto \text{exp}\Big(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}(y-BX)'R^{-1}(y-BX)) \\
-                    &\propto \text{exp}(-\frac{1}{2}X' Q_0^{-1}X + X' Q_0^{-1}m_0-\frac{1}{2}X' B' R^{-1}BX+X' B' R^{-1}y)\\
-                    &=\text{exp}(-\frac{1}{2}X'(Q_0^{-1}+B' R^{-1}B)X+
-                    X'(Q_0^{-1}m_0+B' R^{-1}y)\Big).
+                    &\propto \text{exp}\Big(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}(y-Bx)'\Lambda^{-1}(y-Bx)) \\
+                    &\propto \text{exp}(-\frac{1}{2}x' C_0^{-1}x + x' C_0^{-1}m_0-\frac{1}{2}x' B' \Lambda^{-1}Bx+x' B' \Lambda^{-1}y)\\
+                    &=\text{exp}(-\frac{1}{2}x'(C_0^{-1}+B' \Lambda^{-1}B)x+
+                    x'(C_0^{-1}m_0+B' \Lambda^{-1}y)\Big).
     \end{align*}
@@
 \begin{equation*}
-    Q_1=Var(X\mid Y=y)= (Q_0^{-1}+ B' R^{-1} B)^{-1}=Q_0-Q_0 B'(B Q_0 B'+R)^{-1}B Q_0
+    C_1=Var(X\mid Y=y)= (C_0^{-1}+ B' \Lambda^{-1} B)^{-1}=C_0-C_0 B'(B C_0 B'+\Lambda)^{-1}B C_0
 \end{equation*}
 and 
-    \begin{align*}
-        m_1 &=\mathbb E(X\mid Y=y) = Q_1^{-1}(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= (Q_0-Q_0B'(B Q_0B'+R)^{-1}B Q_0)(Q_0^{-1}m_0+B' R^{-1}y)\\
-            &= m_0-Q_0B'(B Q_0B'+R)^{-1}B m_0 + Q_0 B' R^{-1}y - Q_0B'(B Q_0B'+R)^{-1}B Q_0 B' R^{-1}y \\
-            &=[I_{d_x} - Q_0 B'(B Q_0 B'+ R)^{-1} B ]m_0 + Q_0 B'(B Q_0 B'+ R)^{-1}y.
-    \end{align*}
+\begin{equation*}
+\begin{split}
+        m_1 &=\mathbb E(X\mid Y=y) = C_1^{-1}(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
+            &= (C_0-C_0B'(B C_0B'+\Lambda)^{-1}B C_0)(C_0^{-1}m_0+B' \Lambda^{-1}y)\\
+            &= m_0-C_0B'(B C_0B'+\Lambda)^{-1}B m_0 + C_0 B' \Lambda^{-1}y - C_0B'(B C_0B'+\Lambda)^{-1}B C_0 B' \Lambda^{-1}y \\
+            &=[I_{d_x} - C_0 B'(B C_0 B'+ \Lambda)^{-1} B ]m_0 + C_0 B'(B C_0 B'+ \Lambda)^{-1}y.
+\end{split}
+\end{equation*}$\hfill\blacksquare$
 \end{proof*}
@@
 Assume that
-\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(m_{t-1},Q_{t-1}).\)
+\(\mathbb{P}_{t-1}(dx_{t-1}\mid Y_{1:t-1}=y_{1:t-1})=\mathcal N(m_{t-1},C_{t-1}).\)
 Then, using the transition equation of the latent variable, by the
@@
 distribution is \begin{equation}
-\mathbb{P}_{t-1}(dx_{t}\mid Y_{0:t-1}=y_{0:t-1})=\mathcal N(A_tm_{t-1},A_tQ_{t-1}A_t'+\Sigma_t).
-\end{equation} Letting \(E_T=A_tQ_{t-1}A_t'+\Sigma_t,\) and applying
+\mathbb{P}_{t-1}(dx_{t}\mid Y_{1:t-1}=y_{1:t-1})=\mathcal N(A_tm_{t-1},A_tC_{t-1}A_t'+\Sigma_t).
+\end{equation} Letting \(R_T=A_tC_{t-1}A_t'+\Sigma_t,\) and applying
 lemma \ref{kalman-filter}, it follows that the filtering distribution is
 \begin{align*}
-  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{0:t-1}&=y_{0:t-1})=\mathcal N(m_t,Q_t), \text{where} \\ Q_t&=E_t[I_{d_x} - B_t'(B_tE_tB_t'+R_t)^{-1} B_tE_t] \\
-  m_t &= [I_{d_x} - E_tB_t'(B_tE_tB_t'+R_t)^{-1} B_t]A_tm_{t-1}+E_tB_t'(B_tE_tB_t'+R_t)^{-1}y_t.
+  \mathbb{P}_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\mathcal N(m_t,C_t), \text{ where} \\ C_t&=R_t[I_{d_x} - B_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_tR_t] \\
+  m_t &= [I_{d_x} - R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_t]A_tm_{t-1}+R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1}y_t.
 \end{align*}
+
+In some situations, it is possible to linearize otherwise non-linear
+models in order to employ this filtering method. There are, however,
+cases where this is not possible or where the linearization is not
+satisfactory. One such example is the stochastic volatility model. In
+the following chapters, we will introduce powerful algorithms that can
+be used in much more general settings.
 
@@
 filter. However, this toy example will be used also to illustrate more
-involved filtering strategies described in this work. We believe indeed
+involving filtering strategies described in this work. We believe indeed
 that it represents a useful starting point to understand the logic of
@@
 
-\hfill\break
 \texttt{DLM(data,sig2,tau2,m0,C0)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{DLM}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
-\NormalTok{  n  }\OtherTok{=} \FunctionTok{length}\NormalTok{(data)}
-\NormalTok{  m  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
-\NormalTok{  C  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
-  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{n)\{}
-    \ControlFlowTok{if}\NormalTok{ (t}\SpecialCharTok{==}\DecValTok{1}\NormalTok{)\{}
-\NormalTok{      a }\OtherTok{=}\NormalTok{ m0}
-\NormalTok{      R }\OtherTok{=}\NormalTok{ C0 }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{DLM<-}\ControlFlowTok{function}\NormalTok{(data,sig2,tau2,m0,C0)\{}
+\NormalTok{  n  =}\StringTok{ }\KeywordTok{length}\NormalTok{(data)}
+\NormalTok{  m  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+\NormalTok{  C  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,n)}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\NormalTok{n)\{}
+    \ControlFlowTok{if}\NormalTok{ (t}\OperatorTok{==}\DecValTok{1}\NormalTok{)\{}
+\NormalTok{      a =}\StringTok{ }\NormalTok{m0}
+\NormalTok{      R =}\StringTok{ }\NormalTok{C0 }\OperatorTok{+}\StringTok{ }\NormalTok{tau2}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{      a }\OtherTok{=}\NormalTok{ m[t}\DecValTok{{-}1}\NormalTok{]}
-\NormalTok{      R }\OtherTok{=}\NormalTok{ C[t}\DecValTok{{-}1}\NormalTok{] }\SpecialCharTok{+}\NormalTok{ tau2}
+\NormalTok{      a =}\StringTok{ }\NormalTok{m[t}\DecValTok{-1}\NormalTok{]}
+\NormalTok{      R =}\StringTok{ }\NormalTok{C[t}\DecValTok{-1}\NormalTok{] }\OperatorTok{+}\StringTok{ }\NormalTok{tau2}
 \NormalTok{    \}}
-\NormalTok{    A }\OtherTok{=}\NormalTok{ R}\SpecialCharTok{/}\NormalTok{(R}\SpecialCharTok{+}\NormalTok{sig2)}
-\NormalTok{    m[t] }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{A)}\SpecialCharTok{*}\NormalTok{a }\SpecialCharTok{+}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{y[t]}
-\NormalTok{    C[t] }\OtherTok{=}\NormalTok{ A}\SpecialCharTok{*}\NormalTok{sig2}
+\NormalTok{    A =}\StringTok{ }\NormalTok{R}\OperatorTok{/}\NormalTok{(R}\OperatorTok{+}\NormalTok{sig2)}
+\NormalTok{    m[t] =}\StringTok{ }\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{A)}\OperatorTok{*}\NormalTok{a }\OperatorTok{+}\StringTok{ }\NormalTok{A}\OperatorTok{*}\NormalTok{y[t]}
+\NormalTok{    C[t] =}\StringTok{ }\NormalTok{A}\OperatorTok{*}\NormalTok{sig2}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{m=}\NormalTok{m,}\AttributeTok{C=}\NormalTok{C))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{m=}\NormalTok{m,}\DataTypeTok{C=}\NormalTok{C))}
 \NormalTok{\}}
@@
 the true states. 95 percent credible intervals are computed as
-\[[E(\theta_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})},E(\theta_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(\theta_{t}|y_{1:t})}]\]
+\[[E(x_{t}|y_{1:t})-z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})},E(x_{t}|y_{1:t})+z_{1-\alpha/2}\sqrt{V(x_{t}|y_{1:t})}]\]
 
@@
 three particle filtering algorithms: the \textit{bootstrap}, the
-\textit{guided} and the \{auxiliary particle filters\}. Instead, in
-section \ref{pf_lui_west} we describe the \{Liu and West filter\}, when
-particle filtering is used to estimate model's parameters. Finally, in
-section \ref{pf_converg} we establish some general convergence results
-for Particle Filter - Monte Carlo estimators.
+\textit{guided} and the \textit{auxiliary particle filters}. Instead, in
+section \ref{pf_liu_west} we describe the \textit{Liu and West filter},
+when particle filtering is used to estimate model's parameters. Finally,
+in section \ref{pf_converg} we establish some general convergence
+results for Particle Filter - Monte Carlo estimators.
 
@@
 The following result represents the starting point for the idea of
-importance sampling.
+importance sampling. \bigskip
 
@@
     \item Sample $N$ values from the proposal measure: $X^n\thicksim \mathbb M$ for $n=1,...,N$.
-    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$
+    \item Compute the associated normalized weights: $W(X^n)$ for $n=1,...,N$.
     \item Compute the estimator:
@@
     \mathbb Q^N(dx):=\sum_{n=1}^NW^n\delta_{X^n}(dx), ~~~X^n\thicksim \mathbb M
-\end{equation} Furthermore, the IS estimator is the expected value of
-the function \(\varphi\) with respect to the approximating probability
-measure \(\mathbb Q^N\): \begin{equation*}
+\end{equation} Then, the auto-normalized IS estimator
+\(\hat\varphi_{AN}\) is the expected value of the function \(\varphi\)
+with respect to the approximating probability measure \(\mathbb Q^N\):
+\begin{equation*}
     \begin{split}
@@
 
+We can show that the auto-normalized IS estimator is both consistent and
+asymptotically normal.\\
 Consider the auto-normalized IS estimator
@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n)\rightarrow \mathbb E_{\mathbb M}(w) ~~~\text{ a.s.}\label{is_conv_lln}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n)\xrightarrow{\text{a.s.}} \mathbb E_{\mathbb M}(w) \label{is_conv_lln}
 \end{equation} Applying Slutsky's theorem, together with
@@
 \end{equation} Therefore, the IS (auto-normalized) estimator for
-\(\varphi\) is both consistent and asymptotically normal.
+\(\varphi\) is asymptotically normal. Consistency can be proved in an
+analogous way, applying the strong LLN and lemma \ref{lemma_imp_sam_1}:
+\begin{equation}
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrightarrow{\text{a.s.}}  \mathbb E_\mathbb M(\varphi \cdot w)=\mathbb E_\mathbb Q(\varphi)\cdot \mathbb E_\mathbb M(w)
+\end{equation} Together with \eqref{is_conv_lln}, we get strong
+convergence of the estimator (that implies consistency).
+\begin{equation}
+   \hat \varphi_{AN}=\sum_{n=1}^NW^n \varphi (X^n)\xrightarrow{\text{a.s.}}  \mathbb E_\mathbb Q(\varphi)
+\end{equation}
 
@@
 While in the SIS all the particles (i.e., the sampled values)
-\texttt{survive",\ with\ the}risk" that particles associated with small
+``survive'', with the ``risk'' that particles associated with small
 weights are carried-over (in the sense that sampling in period 1 occurs
@@
 weight.\\
-To understand the reason of the previous statement, let us assume that
+To understand the logic of the previous statement, let us assume that
 both \(\mathbb Q_0\) and \(\mathbb M_0\) admit densities \(q_0\),
@@
 Consider the sequential updating for the weights: \begin{equation}
-    w^n_1=\frac{\mathbb Q_1(\tilde dx_{0:1})}{\mathbb M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
+    W^n_1=\frac{\mathbb Q_1(dx_{0:1})}{\mathbb M_1(dx_{0:1})}\propto \frac{q_0(x_0^{A_1^n})}{Pr(A_1^n)\cdot m_0(x_0^{A_1^n})}\cdot \frac{m_1(x_1^n|x_0^{A_1^n})}{m_1(x_1^n|x_0^{A_1^n})}
 =\underbrace{\frac{q_0(x_0^{A_1^n})}{ m_0(x_0^{A_1^n})}}_{W_0^n}\cdot \underbrace{\frac{1}{Pr(A_1^n)}}_{1/W_0^n}=1\label{sis_update}
@@
 implies that the joint
-distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation}
+distributions\footnote{Let us denote the joint distributions with $q_t$ and $m_t$ too, with some abuse of notation.}
 admit a density too). Then, weights, for every \(t=1,...,T\), are
@@
     w_{t-1}^n\cdot\frac{q_t( x_t|x_{t-1}^n)}{m_t(x_t|x_{t-1}^n)}\label{sis_update}
-\end{equation} Equation \eqref{sis_update} suggests, intuitively, that
+\end{equation} Intuitively, equation \eqref{sis_update} suggests that
 over time new variance is added from the incremental weights (i.e., the
@@
 
-\color{red}
-
-Let us consider the standard state-space model introduced in the
-previous chapters. \color{black} \textit{Particle filter} (PF) is a
+Let us consider the standard state-space model introduced previously
+(see \ref{sec_state_space}). \textit{Particle filter} (PF) is a
 numerical algorithm that is used to obtain approximations for the
@@
 \textit{curse of dimensionality} problem will emerge and the variance of
-the weights will diverge over time as the variance of past weights
-increases as new updates occur period-by-period. Figure ***** shows an
-example of a PF where SIS is used: note that the \(ESS\) drops
-dramatically after few periods, meaning that the variance of the
-importance weights diverges. This motivates the use of a SIR structure
-for the PF algorithm.~\\
+the weights will diverge over time as new updates occur
+period-by-period. Next, we show an example of a PF where SIS is used:
+note that the \(ESS\) drops dramatically after few periods, meaning that
+the variance of the importance weights diverges. This motivates the use
+of a SIR structure for the PF algorithm.~\\
 Lastly, it is now common practice to use a mixed approach with respect
@@
 Consider again the random walk plus noise model of section
-\ref{implementation},the challenge here is to estimate the filtered
+\ref{implementation}, the challenge here is to estimate the filtered
 states using a Sequential Importance Sampling algorithm. The main idea
@@
 
-\hfill\break
 \texttt{SISfun(data,N,m0,C0,tau,sigma)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{SISfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-\NormalTok{    x    }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x,tau)                   }
-\NormalTok{    w    }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)         }
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-
-\NormalTok{    wnorm}\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)                         }
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(wnorm}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)                   }
-
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{SISfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma)\{}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+\NormalTok{    x    =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,x,tau)                   }
+\NormalTok{    w    =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x,sigma)         }
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+
+\NormalTok{    wnorm=}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)                         }
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(wnorm}\OperatorTok{^}\DecValTok{2}\NormalTok{)                   }
+
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
 
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 intuition of what happens when we use such filtering strategy on a
-simulated dataset. We decide to set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\)
-and \(\tau=\sigma=1\). The results shown in the following two plots
-shows a clear degeneration of the effective sample size and bad fit of
-filtered states with respect to the true values.
+simulated dataset. We decide to set \(N=1000\), \(m_{0}=0\),
+\(C_{0}=100\) and \(\tau=\sigma=1\). The following two plots show a
+clear degeneration of the effective sample size and bad fit of filtered
+states with respect to the true values.
 
@@
 
-\hfill\break
 \texttt{PFfun(data,N,m0,C0,tau,sigma,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{PFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,x,tau)}
-\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
-
-\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{PFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,x,tau)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x,sigma)}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 effective sample size are shown in Figure \ref{fig:myfig3}. We decide to
-set \(N=1000\),\(m_{0}=0\),\(C_{0}=100\) and \(\tau=\sigma=1\). Notice
+set \(N=1000\), \(m_{0}=0\), \(C_{0}=100\) and \(\tau=\sigma=1\). Notice
 how the resampling step allows the effective sample size not to drop,
@@
 we want to compare the results of Boostrap Particle Filter(BPF) with
-Kalman Filter(KF). As we can see from figure \ref{fig:myfig4}, when the
-number of particles generated at any interaction increases the results
-for both the estimated mean and the variance tend to converge
-{[}{[}SPIEGARE PERCHE'{]}{]}. Moreover, comparing the Root Mean Square
-Errors of the filtered states with respect to the true state values,
-when the sample size increases the BPF decreases and, for large N, it
-reaches the accuracy of the Kalman Filter.
+Kalman Filter(KF). As we can see from figure \ref{fig:myfig4}, as the
+number of particles N increases, both the state level and the standard
+deviation estimated with particle filter converge to the ones obtained
+by the recursive Kalman filter formulas, confirming Monte Carlo
+approximations to be asymptotically consistent. Moreover, comparing the
+Root Mean Square Errors of the filtered states with respect to the true
+state values, when the sample size increases the BPF decreases and, for
+large N, it reaches the accuracy of the Kalman Filter.
 
@@
 
-\hfill\break
-
 \subsection{Guided Particle Filter}
@@
 motivates the \textit{guided particle filter} (GPF), that can be
-interpreted as a more general version of the GPF. In fact, with the GPF
+interpreted as a more general version of the BPF. In fact, with the GPF
 it is possible, at each stage, to sample new states from a transition
@@
 \(M_t^*(x_{t-1},dx_t)\) the optimal kernel. It can be
-proved\footnote{For a formal proof see \cite{**}, th. 10.1 p. 142.} that
-the optimal kernel, i.e., the one that minimizes the variance of the
-incremental weights \(G_t\) (and of the time \(t\) weights themselves),
-is: \begin{equation}
+proved\footnote{For a formal proof see Chopin and Papaspiliopoulos (2020), th. 10.1 p. 142.}
+that the optimal kernel, i.e., the one that minimizes the variance of
+the incremental weights \(G_t\) (and of the time \(t\) weights
+themselves), is: \begin{equation}
     M_t^*(x_{t-1},dx_t)=\frac{1}{\int_{\mathcal X}P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)} P_t(x_{t-1},dx_t)\cdot f_t(y_t|x_t)
@@
 
-\hfill\break
 The Bootstrap Particle Filter Approach for the Random Walk plus Noise
@@
 
-\hfill\break
 \texttt{GPFfun(data,N,m0,C0,tau,sigma,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{GPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
-\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x)}
-\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
-\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
-
-\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{GPFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd<-}\KeywordTok{sqrt}\NormalTok{(tau }\OperatorTok{-}\StringTok{ }\NormalTok{tau}\OperatorTok{^}\DecValTok{2} \OperatorTok{/}\NormalTok{(tau }\OperatorTok{+}\StringTok{ }\NormalTok{sigma))}
+\NormalTok{  predsd <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(sigma}\OperatorTok{+}\NormalTok{tau)}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    means<-x}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{/}\NormalTok{(tau}\OperatorTok{+}\NormalTok{sigma))}\OperatorTok{*}\NormalTok{(data[t]}\OperatorTok{-}\NormalTok{x)}
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x,predsd)}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 estimates that are slightly better with respect to the ones of the BPF,
-and this is confirmed by the Table showing the RMSE. Moreover, also the
-variance is better, suggesting for higher precision.
+and this is confirmed by the Table showing the RMSE. Moreover, the
+variance is lower, suggesting for higher precision.
 
@@
 \end{equation} It is possible to derive an optimality result for the
-auxiliary functions\footnote{Details in ***** th 10.2, p. 148}. In the
-special case in which the transition kernel used for sampling
+auxiliary
+functions\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p. 148}.
+In the special case in which the transition kernel used for sampling
 corresponds with the optimal one \(M_t^*\), the optimal auxiliary
@@
 
-\hfill\break
 \texttt{APFfun(data,N,m0,C0,tau,sigma,r)}\\
@@
 
-\hfill\break
 \textbf{Arguments}
@@
 \begin{Highlighting}[]
-\NormalTok{APFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,sigma)}
-\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
-\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,x[k],tau)}
-\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],sigma,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
-\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x1}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{APFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x,sigma)}
+\NormalTok{    k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{    x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,x[k],tau)}
+\NormalTok{    lw  =}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(data[t],x1,sigma,}\DataTypeTok{log=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{-}\KeywordTok{dnorm}\NormalTok{(data[t],x[k],sigma,}\DataTypeTok{log=}\OtherTok{TRUE}\NormalTok{)}
+\NormalTok{    w   =}\StringTok{ }\KeywordTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x1<-x1[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    x <-}\StringTok{ }\NormalTok{x1}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 kernel. Therfore the algorithm is similiar to the one discussed above
-with the difference that:\\
-{[}{[}{[}{[}Inserire l'algoritmo{]}{]}{]}{]}
+with the difference that:
+
+\begin{algorithm}APF with optimal transition kernel for Random Walk plus Noise Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item Compute $\sigma_{opt}^{2}$ 
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),\sigma^{2}+\tau^{2}) $
+\item Compute $\mu_{opt}=\mu(x_{t-1}^{(I_{k})})$
+\item Draw $x_{t}^{(k)} \sim N(\mu_{opt},\sigma^{2}_{opt})$
+\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
+\end{enumerate}
+\item otherwise
+\begin{enumerate}
+\item Compute $\mu_{opt}=\mu(x_{t-1})$
+\item Draw $x_{t}^{(i)} \sim N(\mu_{opt},\sigma^{2}_{opt})$ for $i=1,...N$
+\item Set $\tilde{w}^{(i)}_{t}=w^{(i)}_{t-1}f_{N}(y_{t};x_{t}^{(i)},\sigma^{2}+\tau^{2})$ for $i=1,...,N$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
 
@@
 \begin{Highlighting}[]
-\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
-\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
-\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{APFoptfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  importancesd<-}\KeywordTok{sqrt}\NormalTok{(tau }\OperatorTok{-}\StringTok{ }\NormalTok{tau}\OperatorTok{^}\DecValTok{2} \OperatorTok{/}\NormalTok{(tau }\OperatorTok{+}\StringTok{ }\NormalTok{sigma))}
+\NormalTok{  predsd <-}\StringTok{ }\KeywordTok{sqrt}\NormalTok{(sigma}\OperatorTok{+}\NormalTok{tau)}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{    weight =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x,predsd)}
+\NormalTok{    k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{    means<-x[k]}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{/}\NormalTok{(tau}\OperatorTok{+}\NormalTok{sigma))}\OperatorTok{*}\NormalTok{(data[t]}\OperatorTok{-}\NormalTok{x[k])}
+\NormalTok{    x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    w   =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{    weight }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
+\NormalTok{    weight =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{    k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{    means<-x[k]}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{/}\NormalTok{(tau}\OperatorTok{+}\NormalTok{sigma))}\OperatorTok{*}\NormalTok{(data[t]}\OperatorTok{-}\NormalTok{x[k])}
+\NormalTok{    x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,means,importancesd)}
+\NormalTok{    w   <-}\StringTok{  }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],x1,predsd)}
 \NormalTok{    \}}
 
-\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
-\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
-\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
-\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
-
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    w   =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    x <-}\StringTok{ }\NormalTok{x1}
+
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#\#}
-\DocumentationTok{\#\# Per roberto:::::: cambia molto se usiamo l\textquotesingle{}uno o l\textquotesingle{}altro!!!!!! perchè??}
-\NormalTok{APFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,tau,sigma,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  importancesd}\OtherTok{\textless{}{-}}\FunctionTok{sqrt}\NormalTok{(tau }\SpecialCharTok{{-}}\NormalTok{ tau}\SpecialCharTok{\^{}}\DecValTok{2} \SpecialCharTok{/}\NormalTok{(tau }\SpecialCharTok{+}\NormalTok{ sigma))}
-\NormalTok{  predsd }\OtherTok{\textless{}{-}} \FunctionTok{sqrt}\NormalTok{(sigma}\SpecialCharTok{+}\NormalTok{tau)}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],x,predsd)}
-\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
-\NormalTok{    means}\OtherTok{\textless{}{-}}\NormalTok{x[k]}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{/}\NormalTok{(tau}\SpecialCharTok{+}\NormalTok{sigma))}\SpecialCharTok{*}\NormalTok{(data[t]}\SpecialCharTok{{-}}\NormalTok{x[k])}
-\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,means,importancesd)}
-\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],x1,predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],x[k],predsd,}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
-\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x1}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
-
-\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
-
-\NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{figure}[H]
 
-{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-26-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-25-1}
 
@@
 
-\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-26}
+\caption{a) APF Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-25}
 \end{figure}
 
-\section{Liu and West Filter}\label{pf_lui_west}
+\section{Liu and West Filter}\label{pf_liu_west}
 
@@
 parameter over time from a continuous distribution. In this way, at
-every time, the support of \(\psi\) is not limited to the initially
-sampled \(N\) values.\\
+every time, the support of \(\psi\) is not limited to the \(N\)
+initially sampled values.\\
 Here, we describe the LWF that makes use of the normal distribution to
@@
 version of the bootstrap
-filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.}\\
+filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF.
+For instance, in our following implementation with the random walk model we use an auxiliary particle LWF.}\\
 Let the transition kernel and the likelihood depend on the parameter
@@
 \begin{enumerate}
-    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N W_{t-1}^n\psi^n$, $\Omega = \sum_{n=1}^N W_{t-1}^n(\psi^n-\bar \psi)^2$.
-    \item Compute: $m^n=a\cdot  \psi^n+(1-a)\cdot \bar \psi$ for $n=1,...,N$.
+    \item Compute the weighted average and the variance for the parameters, respectively: $\bar \psi = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}\psi^{A_t^n}$, $\Omega = \sum_{n=1}^N \hat W_{t-1}^{A_t^n}(\psi^{A_t^n}-\bar \psi)(\psi^{A_t^n}-\bar \psi)'$.\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
+    \item Compute: $m^{A_t^{n}}=a \cdot  \psi^{{A_t}^{n}}+(1-a)\cdot \bar \psi$ for $n=1,...,N$, with $a\in (0,1)$ and $h^2=1-a^2$.
     \item Draw new parameters: $\psi^n\thicksim \mathcal N\big(m^{A_t^n},h^2\cdot \Omega\big)$ for $n=1,...,N$.
@@
 \(\bar \psi\) be the weighted average of the parameter values and
-\(\Omega\) be their variance (the weights used to compute this mean and
-variance are the previous period inferential ones,
-i.e.~\(W_{t-1}^n\)).~\\
+\(\Omega\) be their variance (using the normalized weights in \(t-1\)
+after the resampling step).~\\
 The algorithm suggests to sample new values for the parameters from a
-continuous distribution, i.e., a Normal centered at the (resampled)
-values \(m^{A_t^n}\) with variance \(h^2\cdot \Omega\) where \(m^n\) is
-a convex combination of the sample mean \(\bar\psi\) and the actual
-extraction of \(\psi^n\) with coefficient \(a\in (0,1)\) and \(h\) is a
-scalar such that \(a^2+h^2=1\). While it would be more intuitively
-immediate to sample new parameters from a Normal density centered at the
-previously sampled values (i.e., \(\psi^n\)), this would increase the
-variance of \(\psi\) over time.\\
+continuous distribution, i.e., a
+Normal\footnote{More generally, distributions whose support coincides with the relevant parameter space.}
+centered at the (resampled) values \(m^{A_t^n}\) with variance
+\(h^2\cdot \Omega\) where \(m^n\) is a convex combination of the sample
+mean \(\bar\psi\) and the actual extraction of \(\psi^n\) with
+coefficient \(a\in (0,1)\) and \(h\) is a scalar such that
+\(a^2+h^2=1\). While it would be more intuitively immediate to sample
+new parameters from a Normal density centered at the previously sampled
+values (i.e., \(\psi^n\)), this would increase the variance of \(\psi\)
+over time.\\
 In fact, let
-\(\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N(\psi^{A_t^{n}},\Omega)\)\footnote{Big hat-weights $ \hat W_{t-1}^{A_t^{n}}$ are the normalized weights after the resampling stage, equal to $1/N$ if resampling takes place, $W_{t-1}^n$ otherwise.}
+\(\pi_{t-1}^N(\psi)=\sum_{n=1}^N \hat W_{t-1}^{A_t^{n}}\cdot \mathcal N(\psi^{A_t^{n}},\Omega)\)
 be the empirical distribution of the parameters. From the law of
@@
 
-\hfill\break
 Consider the linear Gaussian example of section \ref{implementation},
@@
 \end{align*} Alternatively, assign them a uniform prior if we have no
-knowledge of the hyperparameters. The algorithm follows the following
-steps.
+knowledge of the hyperparameters. Note that in this model we sample from
+a Gamma distribution (instead of the Normal one, as in the previous
+description of the LWF) since the parameters of interest are variances,
+that can take only positive values. Next, we descrive the algorithm.
 
@@
 
-\hfill\break
 \texttt{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{LWfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs     }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-  \ControlFlowTok{if}\NormalTok{(unif}\SpecialCharTok{==}\NormalTok{T)\{}
-\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{),}\FunctionTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{))\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphav,}\AttributeTok{scale=}\NormalTok{betav),}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaw,}\AttributeTok{scale=}\NormalTok{betaw))}
-\NormalTok{  a      }\OtherTok{=}\NormalTok{ (}\DecValTok{3}\SpecialCharTok{*}\NormalTok{delta}\DecValTok{{-}1}\NormalTok{)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{2}\SpecialCharTok{*}\NormalTok{delta)}
-\NormalTok{  h2     }\OtherTok{=} \DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}
-\NormalTok{  parss  }\OtherTok{=} \FunctionTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\FunctionTok{c}\NormalTok{(N,}\DecValTok{2}\NormalTok{,n))}
-\NormalTok{  xss    }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  ws     }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  ess    }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  w      }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-\NormalTok{    meanV }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{],w)}
-\NormalTok{    varV  }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{{-}}\NormalTok{meanV)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,w)}
-\NormalTok{    meanW }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{],w)}
-\NormalTok{    varW  }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{{-}}\NormalTok{meanW)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{,w)}
-
-\NormalTok{    muV }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\NormalTok{meanV}
-\NormalTok{    sigma2V }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{varV}
-\NormalTok{    alphaV }\OtherTok{=}\NormalTok{ muV}\SpecialCharTok{\^{}}\DecValTok{2}\SpecialCharTok{/}\NormalTok{sigma2V}
-\NormalTok{    betaV }\OtherTok{=}\NormalTok{ muV}\SpecialCharTok{/}\NormalTok{sigma2V}
-
-\NormalTok{    muW }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\NormalTok{meanW}
-\NormalTok{    sigma2W }\OtherTok{=}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\NormalTok{varW}
-\NormalTok{    alphaW }\OtherTok{=}\NormalTok{ muW}\SpecialCharTok{\^{}}\DecValTok{2}\SpecialCharTok{/}\NormalTok{sigma2W}
-\NormalTok{    betaW }\OtherTok{=}\NormalTok{ muW}\SpecialCharTok{/}\NormalTok{sigma2W}
-
-\NormalTok{    weight      }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],xs,}\FunctionTok{sqrt}\NormalTok{(muV))}
-\NormalTok{    k           }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{weight)}
-
-\NormalTok{    pars[,}\DecValTok{1}\NormalTok{]}\OtherTok{\textless{}{-}}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaV[k],}\AttributeTok{rate=}\NormalTok{betaV[k])}
-\NormalTok{    pars[,}\DecValTok{2}\NormalTok{]}\OtherTok{\textless{}{-}}\FunctionTok{rgamma}\NormalTok{(N,}\AttributeTok{shape=}\NormalTok{alphaW[k],}\AttributeTok{rate=}\NormalTok{betaW[k])}
-
-\NormalTok{    xsprevious}\OtherTok{\textless{}{-}}\NormalTok{xs[k]}
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,xs[k],}\FunctionTok{sqrt}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{]))}
-
-\NormalTok{    w           }\OtherTok{=} \FunctionTok{exp}\NormalTok{(}\FunctionTok{dnorm}\NormalTok{( data[t],xs,}\FunctionTok{sqrt}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{]),}\AttributeTok{log=}\NormalTok{T)}\SpecialCharTok{{-}}
-                        \FunctionTok{dnorm}\NormalTok{( data[t],xsprevious,}\FunctionTok{sqrt}\NormalTok{(muV[k]),}\AttributeTok{log=}\NormalTok{T))}
-\NormalTok{    w           }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS         }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xs[index]}
-\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{pars[index,]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{LWfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs     =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+  \ControlFlowTok{if}\NormalTok{(unif}\OperatorTok{==}\NormalTok{T)\{}
+\NormalTok{  pars   =}\StringTok{ }\KeywordTok{cbind}\NormalTok{(}\KeywordTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{),}\KeywordTok{runif}\NormalTok{(N,}\DecValTok{0}\NormalTok{,}\DecValTok{10}\NormalTok{))\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  pars   =}\StringTok{ }\KeywordTok{cbind}\NormalTok{(}\KeywordTok{rgamma}\NormalTok{(N,}\DataTypeTok{shape=}\NormalTok{alphav,}\DataTypeTok{scale=}\NormalTok{betav),}\KeywordTok{rgamma}\NormalTok{(N,}\DataTypeTok{shape=}\NormalTok{alphaw,}\DataTypeTok{scale=}\NormalTok{betaw))}
+\NormalTok{  a      =}\StringTok{ }\NormalTok{(}\DecValTok{3}\OperatorTok{*}\NormalTok{delta}\DecValTok{-1}\NormalTok{)}\OperatorTok{/}\NormalTok{(}\DecValTok{2}\OperatorTok{*}\NormalTok{delta)}
+\NormalTok{  h2     =}\StringTok{ }\DecValTok{1}\OperatorTok{-}\NormalTok{a}\OperatorTok{^}\DecValTok{2}
+\NormalTok{  parss  =}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\KeywordTok{c}\NormalTok{(N,}\DecValTok{2}\NormalTok{,n))}
+\NormalTok{  xss    =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  ws     =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  ess    =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  w      =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+\NormalTok{    meanV =}\StringTok{ }\KeywordTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{],w)}
+\NormalTok{    varV  =}\StringTok{ }\KeywordTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{1}\NormalTok{]}\OperatorTok{-}\NormalTok{meanV)}\OperatorTok{^}\DecValTok{2}\NormalTok{,w)}
+\NormalTok{    meanW =}\StringTok{ }\KeywordTok{weighted.mean}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{],w)}
+\NormalTok{    varW  =}\StringTok{ }\KeywordTok{weighted.mean}\NormalTok{((pars[,}\DecValTok{2}\NormalTok{]}\OperatorTok{-}\NormalTok{meanW)}\OperatorTok{^}\DecValTok{2}\NormalTok{,w)}
+
+\NormalTok{    muV =}\StringTok{ }\NormalTok{a}\OperatorTok{*}\NormalTok{pars[,}\DecValTok{1}\NormalTok{]}\OperatorTok{+}\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{a)}\OperatorTok{*}\NormalTok{meanV}
+\NormalTok{    sigma2V =}\StringTok{ }\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{a}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\NormalTok{varV}
+\NormalTok{    alphaV =}\StringTok{ }\NormalTok{muV}\OperatorTok{^}\DecValTok{2}\OperatorTok{/}\NormalTok{sigma2V}
+\NormalTok{    betaV =}\StringTok{ }\NormalTok{muV}\OperatorTok{/}\NormalTok{sigma2V}
+
+\NormalTok{    muW =}\StringTok{ }\NormalTok{a}\OperatorTok{*}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\OperatorTok{+}\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{a)}\OperatorTok{*}\NormalTok{meanW}
+\NormalTok{    sigma2W =}\StringTok{ }\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{a}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\NormalTok{varW}
+\NormalTok{    alphaW =}\StringTok{ }\NormalTok{muW}\OperatorTok{^}\DecValTok{2}\OperatorTok{/}\NormalTok{sigma2W}
+\NormalTok{    betaW =}\StringTok{ }\NormalTok{muW}\OperatorTok{/}\NormalTok{sigma2W}
+
+\NormalTok{    weight      =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],xs,}\KeywordTok{sqrt}\NormalTok{(muV))}
+\NormalTok{    k           =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{weight)}
+
+\NormalTok{    pars[,}\DecValTok{1}\NormalTok{]<-}\KeywordTok{rgamma}\NormalTok{(N,}\DataTypeTok{shape=}\NormalTok{alphaV[k],}\DataTypeTok{rate=}\NormalTok{betaV[k])}
+\NormalTok{    pars[,}\DecValTok{2}\NormalTok{]<-}\KeywordTok{rgamma}\NormalTok{(N,}\DataTypeTok{shape=}\NormalTok{alphaW[k],}\DataTypeTok{rate=}\NormalTok{betaW[k])}
+
+\NormalTok{    xsprevious<-xs[k]}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,xs[k],}\KeywordTok{sqrt}\NormalTok{(pars[,}\DecValTok{2}\NormalTok{]))}
+
+\NormalTok{    w           =}\StringTok{ }\KeywordTok{exp}\NormalTok{(}\KeywordTok{dnorm}\NormalTok{( data[t],xs,}\KeywordTok{sqrt}\NormalTok{(pars[,}\DecValTok{1}\NormalTok{]),}\DataTypeTok{log=}\NormalTok{T)}\OperatorTok{-}
+\StringTok{                        }\KeywordTok{dnorm}\NormalTok{( data[t],xsprevious,}\KeywordTok{sqrt}\NormalTok{(muV[k]),}\DataTypeTok{log=}\NormalTok{T))}
+\NormalTok{    w           =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS         =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      xs<-xs[index]}
+\NormalTok{      pars<-pars[index,]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xs}
-\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{pars}
+\NormalTok{      xs<-xs}
+\NormalTok{      pars<-pars}
 \NormalTok{    \}}
 
-\NormalTok{    xss         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xss,xs)}
-\NormalTok{    parss[,,t]  }\OtherTok{=}\NormalTok{ pars }
-\NormalTok{    ws          }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    xss         =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xss,xs)}
+\NormalTok{    parss[,,t]  =}\StringTok{ }\NormalTok{pars }
+\NormalTok{    ws          =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess         =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xss=}\NormalTok{xss,}\AttributeTok{parss=}\NormalTok{parss,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xss=}\NormalTok{xss,}\DataTypeTok{parss=}\NormalTok{parss,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-29-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-28-1}
 
@@
 
-\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-29}
+\caption{a) LW Filtered States with credible interval (in red). b) Effective sample size (in black) with threshold (in yellow).}\label{fig:unnamed-chunk-28}
 \end{figure}
@@
 \begin{longtable}[t]{cccccc}
-\caption{\label{tab:unnamed-chunk-31}RMSE}\\
+\caption{\label{tab:unnamed-chunk-30}RMSE}\\
 \toprule
@@
 
+In this section, we cover three propositions that justify the use of the
+PF algorithms to estimate quantities of interest. Firstly, we present
+two convergence results for the estimators of the predictive and
+filtering distributions. As we will discuss, these results hold under
+reasonable conditions. Then, we will show that the error of the particle
+estimators are asymptotically distributed according to a Gaussian
+distribution with known variance. Henceforth, we will maintain the
+assumption that (multinomial) resampling is repeated at each iteration.
+
+\subsection{MSE convergence}
+
+As already seen in subsection \ref{seq_impt_sampling}, the particle
+approximation of \(\mathbb Q_{t}(dx_{0:t})\) is: \begin{equation*}
+      \sum_{n=1}^NW_{t-1}^nM_t(X_{t-1}^n,dx_t).
+\end{equation*} The following lemma states that the (mean squared) error
+related to sampling from this conditional distribution is bounded.
+\bigskip
+
+\begin{lemma}
+For any measurable, integrable, and bounded function $\varphi:\mathcal X\rightarrow\mathbb R,$
+$$
+\mathbb E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg\}^2 \Bigg]\leq \frac{\|\varphi\|_{\infty}^2}{N},
+$$
+where $\|\cdot\|_{\infty}$ is the supnorm and $M_t\varphi$ is the map 
+$x_{t-1}\mapsto \int_{\mathcal X}M_t(x_{t-1},dx_t)\varphi(x_t).$
+\end{lemma}
+
+This result can be used to prove the convergence of the estimator
+\(\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\). To see this, note
+that \begin{equation}
+    \begin{split}
+        \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)=\bigg(&\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg) \\
+        & + \bigg(\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n)- \mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)\bigg).
+    \end{split}
+\end{equation} where
+\(\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)=\int_{\mathcal X^{t+1}}\varphi(x_t) \mathbb Q_{t-1}(dx_{0:t-1})M_t(x_{t-1},dx_t)\).\\
+By the lemma, the first component of the right-hand side has a bounded
+MSE. It can be proved that also the second component of the right-hand
+side has a bounded MSE. Using the fact that
+\(\mathbb E\{(X+Y)^2\}\leq2\mathbb E\{\mathbb E(X^2)+\mathbb E(Y^2)\},\)
+it follows that also the MSE of the left-hand side is bounded. A simple
+induction argument proves that the result holds for all \(t\).\\
+Using a similar argument, one can prove that the MSE of the estimator
+\(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\) is bounded as well. These
+ideas are formalized in the following proposition. \bigskip
+
+\begin{proposition}
+For any measurable, integrable, and bounded function $\varphi:\mathcal X\rightarrow\mathbb R,$ $\forall t\geq0$ there exist constants $c_t,c_t'$ such that:
+$$
+\mathbb E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)\bigg\}^2 \Bigg]\leq c_t \frac{\|\varphi\|_{\infty}^2}{N}
+$$
+and 
+$$
+    \mathbb E\Bigg[ \bigg\{\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\mathbb E_{\mathbb Q_t}(\varphi) \bigg\}^2 \Bigg]\leq c_t' \frac{\|\varphi\|_{\infty}^2}{N}.
+$$
+Therefore, taking the limit for $N\rightarrow \infty,$ we have that 
+$$
+\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{MSE}}\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)
+$$
+and 
+$$
+\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{MSE}}\mathbb E_{\mathbb Q_t}(\varphi)
+$$
+\end{proposition}
+
+In the next subsection, we establish a much stronger property of the
+particle filter algorithms.
+
+\subsection{Almost sure convergence}
+
+The following proposition states that estimators
+\(\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\) and
+\(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\) converge to their true
+values, up to negligible sets, i.e., the sets such that the estimators
+do not converge to \(\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)\) and
+\(\mathbb E_{\mathbb Q_t}(\varphi)\) have zero probability.
+Additionally, recall that almost sure convergence implies convergence in
+probability. \bigskip
+
+\begin{proposition}
+For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\mathcal X\rightarrow\mathbb R,$
+$$
+\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{a.s}}\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)
+$$
+and 
+$$
+\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{a.s}}\mathbb E_{\mathbb Q_{t}}(\varphi).
+$$
+\end{proposition}
+
+\subsection{Central limit theorem}
+
+The last asymptotic result of PF estimators concerns convergence in law:
+the error of a particle estimate is asymptotically normally distributed.
+\bigskip
+
+\begin{proposition}
+For any $t\geq0$ and any measurable, integrable, and bounded function $\varphi:\mathcal X\rightarrow\mathbb R,$
+$$
+\sqrt{N}\bigg( \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\mathbb E_{\mathbb Q_{t-1}M_t}(\varphi)\bigg)\implies \mathcal N(0,\hat{\mathcal{V}}_t(\varphi))
+$$
+and 
+$$
+\sqrt{N}\bigg(\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\mathbb E_{\mathbb Q_{t}}(\varphi)\bigg)\implies \mathcal N(0,\mathcal{V}_t(\varphi)),
+$$
+where the asymptotic variances are defined recursively, with $ \hat{\mathcal{V}}_0(\varphi):=Var_{\mathbb M_0}(\varphi) $,
+\begin{equation*}
+    \begin{split}
+        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\mathbb E_{M_t}(\varphi))+Var_{\mathbb Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
+        \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\mathbb E_{\mathbb Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
+    \end{split}
+\end{equation*}
+Here, $\bar{G}_t=G_t/\mathbb E_{\mathbb Q_{t-1}M_t}(G_t)$ is the normalized incremental weight.
+\end{proposition}
+
+This version of the CLT for PF suggest that PF estimators with respect
+both to the predictive and the filtering distributions are approximately
+normal for a large number of particles.\\
+In particular, the logic behind the expression for
+\(\mathcal{V}_t(\varphi)\) is the same as described in section
+\ref{pf_is_conv}. In fact, recall that the estimator with respect to the
+filtering distribution is a more sophisticated form of IS estimator,
+where the target distribution differs from the proposal one. ~ Instead,
+to get an intuition of the variance \(\hat{\mathcal{V}}_t(\varphi),\)
+note that by the law of total variance, \begin{equation*}
+    \begin{split}
+        Var\Bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n) \Bigg] = &Var\Bigg[ \mathbb E\bigg[\frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)\Big| \sigma(X_{t-1}^{1:N}) \bigg] \Bigg] \\
+        &+ \mathbb E\Bigg[ Var\bigg[ \frac{1}{N}  \sum\limits_{n=1}^N \varphi(X_t^n)    \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg] \\
+        &= Var\bigg[\sum\limits_{n=1}^N W_{t-1}^n (M_t\varphi)(X_{t-1}^n)\bigg] \\
+        &+\frac{1}{N}\mathbb E\Bigg[Var\bigg[\varphi(X_t^1) \Big| \sigma(X_{t-1}^{1:N})\bigg]\Bigg].
+    \end{split}
+\end{equation*} The notation \(\mid \sigma(X_{t-1}^{1:N})\) means
+conditioning on having sampled (from the multinomial distribution)
+\(X_{t-1}^{1:N}\). Conditional on multinomial sampling, the
+\(X_{t-1}^n\) are i.i.d., with distribution
+\(\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)\), that should
+converge to \(\mathbb Q_{t-1}M_t\). So, it is natural to think that the
+second element of the sum converges to the variance of \(\varphi\) with
+respect to the distribution \(\mathbb Q_{t-1}M_t\). These results are of
+practical use because they provide a way to explicitly compute the
+variance of the PF estimates.
+
 \section{Implementations for Stochastic Volatility Models}
 
-\chapter{Applications to Stochastic Volatility Models}
+So far we have seen Sequential Monte Carlo methods applied to the linear
+Gaussian case, however, as already mentioned, for this class of state
+space models, Kalman Filter provides a closed form solution. Here, we
+want to provide an illustration of Particle Filters for a state space
+model for which no analytical solutions are available. More precisely,
+we deal with a stochastic volatility model. For the sake of simplicity
+we use a basic specification of stochastic volatility model
+
+\begin{align}
+y_{t}|x_{t} & \sim N(0,e^{x_{t}}) \\
+x_{t}|x_{t-1} & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align}
+
+We start describing the main four particle filtering strategies
+discussed in this work and then we are going to compare their
+performances. The R code of these and other algorithms used in the
+Chapter application can be found in Appendix 1.
+
+In the Bootstrap Particle filter we basically use the model transition
+kernel and do resampling any time the effective sample size is smaller
+than a predetermined threshold.
+
+\begin{algorithm} BPF for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ form $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1},\tau^2) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}f_{N}(y_{t};0,e^{x_{t}}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+If we instead want to sample from the optimal transition kernel, things
+get more complicated. It can be shown that the optimal transition kernel
+is
+\[ m_{t}^{opt}(x_{t}|x_{t-1}) \propto f_{t}(y_{t}|x_{t})p(x_{t}|x_{t-1}) \propto exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg]\]
+and it is not straightforward to sample from it. Therefore, we follow a
+common praxis consisting in approximating \(m_{t}^{opt}\) by a Gaussian
+distribution obtained by linearising \(e^{-x_{t}}\),
+i.e.~\(e^{-x_{t}}=e^{-x_{0}}(1+x_{0}-x+...)\), taking
+\(x_{0}=\mu^{\star}(x_{t-1}) := \alpha + \beta x_{t-1}\), i.e the mean
+of the Gaussian transition density \(p(x_{t}|x_{t-1})\). After some
+calculations we would end up with a Gaussian distribution
+\(N(\xi(x_{t-1}),\sigma^{2})\) with
+\[ \xi(x_{t-1}) = \mu^{\star}(x_{t-1}) + \frac{\sigma^{2}}{4}(y_{t}^{2}e^{-\mu^{\star}(x_{t-1})}-2)\].
+With this transition kernel we can implement the Guided Particle Filter
+in the following way
+
+\begin{algorithm} GPF for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarize $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ such that $\{(x_{t},w_{t-1})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t-1})$
+\item Set $w_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})} \ \ i=1,...,N$ such that $\{(x_{t},w_{t})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{t}|y_{1:t})$
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{x_{t}^{(i)}}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+The Auxiliary Particle Filter with optimal transition kernel uses the
+same transition kernel of the Guided Particle Filter and while the
+perfectly adapted auxiliary function is
+\[g(x_{t-1})=\frac{1}{2\pi\sigma^{2}}\int_{\mathcal{X}}exp\bigg[-\frac{1}{2\sigma^{2}}(x_{t}-\alpha-\beta x_{t-1})^{2}-\frac{e^{-x_{t}}}{2}y_{t}^{2}-\frac{x_{t}}{2}\bigg] dx_{t-1}\]
+The relevant literature suggest to approximate this integral by
+linearising the exponential term as in the GPF case. The resulting
+auxiliary function is
+\[ g(x_{t-1})=exp \bigg[ \frac{1}{2\sigma^{2}} \big\{\xi(x_{t-1})^{2}-\mu^{\star}(x_{t-1})^{2} \big\} - \frac{y_{t}^{2}}{2}e^{-\mu^{\star}(x_{t-1})} \big\{ 1+\mu^{\star}(x_{t-1}) \big\}  \bigg]\]
+where \(\xi(x_{t-1})\) and \(\mu^{\star}\) are defined as in the case of
+GPF.
+
+\begin{algorithm}APF with optimal transition kernel for Stochastic Volatility Model
+\begin{itemize}
+\item Let $\{(x_{0},w_{0})^{(i)}\}_{i=1}^{N}$ summarizes $p(x_{0}|y_{0})$ such that, for example, $E(g(x_{0})|y_{0}) \approx \sum_{i=1}^{N}w_{0}^{(i)}g(x_{0}^{(i)})$.
+In particular, initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$ and set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$.
+\item For $t=1,...,n$:
+\begin{enumerate}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw $I_{k}$ with $P(I_{k}=i) \propto w_{t-1}^{(i)}g(x_{t-1}) $
+\item $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}^{(I_{k})}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1}^{(I_{k})})}-2\big],\tau^{2})$ for $i=1,...N$
+\item Set  $\tilde{w}^{(i)}_{t} = N^{-1}$ for $i=1,...,N$
+\end{enumerate}
+\item otherwise
+\begin{enumerate}
+\item Draw $x_{t}^{(i)} \sim N(\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2}) \ \ i=1,...,N$ for $i=1,...N$
+\item Set $\tilde{w}_{t}^{(i)} = w_{t-1}^{(i)}\frac{f_{N}(y_{t};0,e^{x_{t}})f_{N}(x_{t};\alpha+\beta x_{t-1},\tau^{2})}{f_{N}(x_{t};\alpha+\beta x_{t-1}+\frac{\tau^2}{4}\big[y_{t}^2e^{-(\alpha+\beta x_{t-1})}-2\big],\tau^{2})}$ for $i=1,...,N$
+\end{enumerate}
+\item Normalize the weights: $w_{t}^{(i)}=\frac{\tilde{w}_{t}^{(i)}}{\sum_{j=1}^{N}(\tilde{w}_{t}^{(j)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+Finally we present a Liu and West Filter for the case of parameters
+\(\alpha\), \(\beta\) and \(\tau^{2}\) unknown.
+
+\begin{algorithm} LWF for Stochastic Volatility Model
+\begin{itemize}
+\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $(\alpha^{(1)},...,\alpha^{(N)})$ from $N(e_{\alpha},v_\alpha)$ and
+$(\beta^{(1)},...,\beta^{(N)})$ from $N(e_{\beta},v_\beta)$ $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $IG(\nu_{0}/2,\nu_{0}\lambda/2)$.
+Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$ and chose the appropriate value for $\delta$.
+Therefore $\psi^{(i)}=({\alpha}^{(i)},{\beta}^{(i)},{\tau^{2}}^{(i)})$, and
+$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
+\item For $t=1,...,n$:.
+\begin{enumerate}
+\item Compute $\overline{\psi}=E_{\hat{\pi}_{t-1}}(\psi)$ and $\Sigma=V_{\hat{\pi}_{t-1}}(\psi)$. For $i=1,...,N$, set
+\begin{align*}
+m^{(i)} & = a\psi^{(i)}+(1-a)\overline{\psi} \\
+\hat{x}_{t}^{(i)} & = E(x_{t}|x_{t-1}=x_{t-1}^{(i)},\psi=\psi^{(k)})
+\end{align*}
+\item For $k=1,...,N$:
+\begin{itemize}
+\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|0,e^{\frac{\hat{x}_{t}^{(i)}}{2}}) $
+\item Draw $\psi^{(k)}$ from $N(m^{(I_{k})},h^{2}\Sigma)$
+\item Draw $x_{t}^{(k)}$ from $N(\alpha^{(k)}+\beta^{(k)}x_{t-1}^{(I_{k})},\tau^{2})$
+\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|0,e^{\frac{x_{t}^{(k)}}{2}})}{f_{N}(y_{t}|0,e^{\frac{\hat{x}_{t}^{(I_{k})}}{2}})}$
+\end{itemize}
+\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
+\item if $ESS<N/2$ then
+\begin{enumerate}
+\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
+\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
+\end{enumerate}
+\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
+\end{enumerate}
+\end{itemize}
+\end{algorithm}
+
+We are now ready to compare the performances of these four algorithms.
+We simulate a Stochastic Volatility process setting arbitrarily the
+following parameters: \(\alpha = 0\), \(\beta = 0.99\) and
+\(\tau^{2}=0.05\). The resulting observation and state process is shown
+in the figure below.
+
+\begin{figure}[ht]
+
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-32-1}
+
+}
+
+\caption{Simulated Stochastic Volatility Model}\label{fig:unnamed-chunk-32}
+\end{figure}
+
+We provide both a graphical and a numerical
+comparison.\footnote{The sequences of filtered states shown in the graphs are obtained by using filters with 10000 particles.}
+From the plots it is hard to say which particle filter provides the best
+performances. By and large, BPF, GPF and APF behave similarly, while the
+Liu and West filter presents larger credible intervals but at the same
+time a higher coverage of the true states.\\
+On the other hand, by a RMSE comparison, we see several differences
+between the filters. Note that the Guided Particle Filter yields the
+lowest RMSE at N=100, and does not improve much even when raising the
+number of particles to 100000. Instead, the APF achieves the best
+performance in the field when N rises to 10000 and 100000, despite
+lagging well behind the BPF and the GPF for the three initial N
+thresholds. For the Bootstrap PF, the most significant improvements is
+that from N=100 to N=1000. We could then point at the APF as the best
+performing and more computationally expensive filter, while the GPF as
+the best one at low thresholds (\textit{i.e.} least computationally
+expensive, in a way). However, fine margins suggest caution in such
+comparative assessments.\\
+Rather, we would stress the performance of the Liu and West filter:
+indeed, it is apparent that, at least for the first four thresholds,
+such filter delivers the worst performance. Even at N=100000, the APF,
+the GPF and the BPF are much closer to each other than the LWF, which
+still yields the highest RMSE. Such results, in fact, do not come as a
+surprise: indeed, note that, while the three best performing filters are
+built on a model that perfectly coincides with the true DGP, the Liu and
+West filter considers a somewhat misspecified model, which features
+additional uncertainty on several parameters. Intuitively, even though
+the LWF features some degree of parameter learning, and would then
+converge to the correct figures of such parameters, the other filters
+``have already learnt'' the true parameters from the
+beginning.\footnote{Such difference can be represented as follows.
+When trying to pin down real valued parameters, the probability that a simple initial guess for each value is correct can be deemed as null.
+So, in contexts where parameters are unknown, we expect the additional uncertainty featured in the LWF to prove useful, as it can adjust the initial parameters to get closer to the true unknown values.
+However, in synthetic contexts where we know the DGP, such uncertainty is almost useless.}\\
+Finally, notice that in general when the number of particles increases
+the accuracy improves; however, significant improvements are not
+detected for \(N>10000\), suggesting that such threshold may be a
+telling one as regards the convergence of our filters.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-35-1}
+
+}
+
+\caption{Comparison of filtering strategies for Stochastic Volatility Model. Filtered States with credible interval (in red)}\label{fig:unnamed-chunk-35}
+\end{figure}
+
+\chapter{Applications to Stochastic Volatility Models}\label{application_chapter}
 
@@
 
-{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-32-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-38-1}
 
@@
 
-\caption{Close to close percentage log returns, 3 indices}\label{fig:unnamed-chunk-32}
+\caption{Close to close percentage log returns, 3 indices}\label{fig:unnamed-chunk-38}
 \end{figure}
@@
 \begin{longtable}[t]{lccc}
-\caption{\label{tab:unnamed-chunk-33}Summary statistics, close to close percentage log returns}\\
+\caption{\label{tab:unnamed-chunk-39}Summary statistics, close to close percentage log returns}\\
 \toprule
-  & Dow Jones & S\&P500 & STOXX50E\\
+  & S\&P500 & Dow Jones & STOXX50E\\
 \midrule
@@
 
-{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-34-1}
+{\centering \includegraphics{prova_knit_finale_files/figure-latex/unnamed-chunk-40-1}
 
@@
 
-\caption{Rescaled realised daily volatility, 3 indices}\label{fig:unnamed-chunk-34}
+\caption{Rescaled realised daily volatility, 3 indices}\label{fig:unnamed-chunk-40}
 \end{figure}
@@
 \begin{longtable}[t]{lcccccc}
-\caption{\label{tab:unnamed-chunk-36}RMSE and MAE, Bootstrap PF 50000 particles}\\
+\caption{\label{tab:unnamed-chunk-42}RMSE and MAE, Bootstrap PF 50000 particles}\\
 \toprule
@@
 \begin{longtable}[t]{lcccccc}
-\caption{\label{tab:unnamed-chunk-38}RMSE and MAE, Auxiliary PF 50000 particles}\\
+\caption{\label{tab:unnamed-chunk-44}RMSE and MAE, Auxiliary PF 50000 particles}\\
 \toprule
@@
 \begin{longtable}[t]{lcc}
-\caption{\label{tab:unnamed-chunk-40}Discrepancy between Bootstrap and Auxiliary PFs}\\
+\caption{\label{tab:unnamed-chunk-46}Discrepancy between Bootstrap and Auxiliary PFs}\\
 \toprule
@@
 \begin{longtable}[t]{lccccccc}
-\caption{\label{tab:unnamed-chunk-42}RMSE and MAE, realised volatility as benchmark}\\
+\caption{\label{tab:unnamed-chunk-48}RMSE and MAE, realised volatility as benchmark}\\
 \toprule
@@
 \begin{longtable}[t]{lcccc}
-\caption{\label{tab:unnamed-chunk-44}RMSE and MAE, realised volatility as benchmark (SP500 and STOXX50E)}\\
+\caption{\label{tab:unnamed-chunk-50}RMSE and MAE, realised volatility as benchmark (SP500 and STOXX50E)}\\
 \toprule
@@
 
-\hfill\break
-
 Allen, D. E., \& McAleer, M. (2020). Do we need stochastic volatility
@@
 
+Johannes, M., Evers, L. (2007). Monte Carlo Methods. Lecture Notes.
+University of Bristol, Department of Mathematics.\\
+
 Johannes, M., Korteweg, A., \& Polson, N. (2014). Sequential learning,
@@
 
+Petris, G., Petrone, S., Campagnali, P. (2010). Dynamic Linear Models
+with R. Springer.\\
+
 Pitt, M. K., \& Shephard, N. (1999). Filtering via simulation: Auxiliary
@@
 
-\chapter{Appendix}
-
+\chapter*{Appendix}
 \section{The Stochastic Volatility Model}
@@
 described in this Appendix take as reference a basic specification of
-Stochastic Volatility Model.
-
+Stochastic Volatility
+Model.\footnote{Actually, the function SVLWfun considers a SV model with an additional hierarchical level, as specified in section XX.}
 \begin{align*}
@@
     x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
-\end{align*}
-
-All the plots and tables of this work concerning stochastic volatility
-models use this functions.
-
-\hrule
-\hrule
-
-\hfill\break
+\end{align*} All the plots and tables of this work concerning stochastic
+volatility models use this functions.
+
+\hrule
+\hrule
+
 \texttt{SVPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
@@
 \texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{tau} ~~square root of \(\tau^2\), the conditional variance of
+the state\\
 \texttt{r} ~~ if present the threshold is set equal to \(N/r\)
@@
 
-\hfill\break
-
 \begin{Shaded}
 \begin{Highlighting}[]
-\NormalTok{SVPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x,tau)}
-\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
-
-\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{SVPFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x,tau)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-\hfill\break
-
-\hrule
-\hrule
-
-\hfill\break
+\hrule
+\hrule
+
 \texttt{SVGPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{SVGPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    xprev}\OtherTok{\textless{}{-}}\NormalTok{x}
-\NormalTok{    x}\OtherTok{\textless{}{-}}\FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{/}\DecValTok{4}\SpecialCharTok{*}\NormalTok{((data[t]}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{(alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x))}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{),tau)}
-\NormalTok{    w1}\OtherTok{\textless{}{-}}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(x,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev,tau)}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(x,}\AttributeTok{mean=}\NormalTok{alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev}\SpecialCharTok{+}\NormalTok{(tau}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{/}\DecValTok{4}\SpecialCharTok{*}\NormalTok{((data[t]}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{(alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{xprev))}\SpecialCharTok{{-}}\DecValTok{2}\NormalTok{),tau)}
-
-\NormalTok{    w }\OtherTok{=}\NormalTok{ w1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w1)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{SVGPFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    xprev<-x}
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x))}\OperatorTok{-}\DecValTok{2}\NormalTok{),tau)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x}\OperatorTok{/}\DecValTok{2}\NormalTok{))}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(x,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{xprev,tau)}\OperatorTok{/}\KeywordTok{dnorm}\NormalTok{(x,}\DataTypeTok{mean=}\NormalTok{alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{xprev}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{xprev))}\OperatorTok{-}\DecValTok{2}\NormalTok{),tau)}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-\hfill\break
-
-\hrule
-\hrule
-
-\hfill\break
+\hrule
+\hrule
+
 \texttt{SVAPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{SVAPFfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-\NormalTok{    weight }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
-\NormalTok{    k   }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\ConstantTok{TRUE}\NormalTok{,}\AttributeTok{prob=}\NormalTok{weight)}
-\NormalTok{    x1   }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x[k],tau)}
-\NormalTok{    lw  }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x1}\SpecialCharTok{/}\DecValTok{2}\NormalTok{),}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}\SpecialCharTok{{-}}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(x[k]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{),}\AttributeTok{log=}\ConstantTok{TRUE}\NormalTok{)}
-\NormalTok{    w   }\OtherTok{=} \FunctionTok{exp}\NormalTok{(lw)}
-\NormalTok{    w   }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x1[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
+\NormalTok{SVAPFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    weight =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{    x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k],tau)}
+\NormalTok{    lw  =}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x1}\OperatorTok{/}\DecValTok{2}\NormalTok{),}\DataTypeTok{log=}\OtherTok{TRUE}\NormalTok{)}\OperatorTok{-}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x[k]}\OperatorTok{/}\DecValTok{2}\NormalTok{),}\DataTypeTok{log=}\OtherTok{TRUE}\NormalTok{)}
+\NormalTok{    w   =}\StringTok{ }\KeywordTok{exp}\NormalTok{(lw)}
+\NormalTok{    w   =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x1[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
 \NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
 
-\NormalTok{    x }\OtherTok{\textless{}{-}}\NormalTok{ x1}
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{    x <-}\StringTok{ }\NormalTok{x1}
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-\hfill\break
 \texttt{SVAPFoptfun(data,N,m0,C0,alpha,beta,tau,r)}\\
@@
 \begin{Highlighting}[]
-\NormalTok{SVAPFoptfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{missing}\NormalTok{(r))\{r}\OtherTok{=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
-\NormalTok{  xs}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ws}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  ess}\OtherTok{\textless{}{-}}\ConstantTok{NULL}
-\NormalTok{  x  }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  w  }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-
-  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-\NormalTok{    x\_prev}\OtherTok{=}\NormalTok{x}
-\NormalTok{    x\_hat }\OtherTok{=}\NormalTok{ alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x }\CommentTok{\#means of the transition from the previous particle }
-\NormalTok{    I\_k }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{prob=}\NormalTok{w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x\_hat)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))))}
-    \ControlFlowTok{for}\NormalTok{(k }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\NormalTok{N)\{}
-\NormalTok{      x[k]}\OtherTok{=}\FunctionTok{rnorm}\NormalTok{(}\AttributeTok{size=}\DecValTok{1}\NormalTok{,}\AttributeTok{mean=}\NormalTok{alpha}\SpecialCharTok{+}\NormalTok{beta}\SpecialCharTok{*}\NormalTok{x\_prev[I\_k],}\AttributeTok{sd=}\NormalTok{tau)}
-\NormalTok{      w[k]}\OtherTok{=}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x[k])}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{sqrt}\NormalTok{(}\FunctionTok{exp}\NormalTok{((x\_hat)}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}
+\NormalTok{SVAPFoptfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\NormalTok{r))\{}
+\NormalTok{      epsilon=alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x}\OperatorTok{+}\NormalTok{((tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\NormalTok{)}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{alpha}\OperatorTok{-}\NormalTok{beta}\OperatorTok{*}\NormalTok{x)}\OperatorTok{-}\DecValTok{2}\NormalTok{)}
+\NormalTok{      weight =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{exp}\NormalTok{((}\DecValTok{1}\OperatorTok{/}\NormalTok{(}\DecValTok{2}\OperatorTok{*}\NormalTok{tau}\OperatorTok{^}\DecValTok{2}\NormalTok{))}\OperatorTok{*}\NormalTok{((epsilon)}\OperatorTok{^}\DecValTok{2}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x)}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{-}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{alpha}\OperatorTok{-}\NormalTok{beta}\OperatorTok{*}\NormalTok{x)}\OperatorTok{*}\NormalTok{(}\DecValTok{1}\OperatorTok{+}\NormalTok{alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x))}
+\NormalTok{      k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{TRUE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{      x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]))}\OperatorTok{-}\DecValTok{2}\NormalTok{),tau)}
+\NormalTok{      w   =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      weight =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{      k   =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\OtherTok{FALSE}\NormalTok{,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{      x1   =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]))}\OperatorTok{-}\DecValTok{2}\NormalTok{),tau)}
+\NormalTok{      w   <-}\StringTok{  }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x1}\OperatorTok{/}\DecValTok{2}\NormalTok{))}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(x1,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k],tau)}\OperatorTok{/}
+\StringTok{        }\KeywordTok{dnorm}\NormalTok{(x1,}\DataTypeTok{mean=}\NormalTok{alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]}\OperatorTok{+}\NormalTok{(tau}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{/}\DecValTok{4}\OperatorTok{*}\NormalTok{((data[t]}\OperatorTok{^}\DecValTok{2}\NormalTok{)}\OperatorTok{*}\KeywordTok{exp}\NormalTok{(}\OperatorTok{-}\NormalTok{(alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x[k]))}\OperatorTok{-}\DecValTok{2}\NormalTok{),tau)}
 \NormalTok{    \}}
-\NormalTok{    w }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS  }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\NormalTok{r))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      x}\OtherTok{\textless{}{-}}\NormalTok{x[index]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
-
-\NormalTok{    xs }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xs,x)}
-\NormalTok{    ws }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{    ess }\OtherTok{=}\FunctionTok{rbind}\NormalTok{(ess,ESS)}
+
+\NormalTok{    w   =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    x <-}\StringTok{ }\NormalTok{x1}
+
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 
 \NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xs,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{ess=}\NormalTok{ess))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
-\hfill\break
 \texttt{SVLWfun(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{ealpha} ~~expected value of the intercept \(\alpha\) in the
+state equation\\
+\texttt{valpha} ~~variance of the intercept \(\alpha\) in the state
+equation\\
+\texttt{ebeta} ~~expected value of the coefficient \(\beta\) in the
+state equation\\
+\texttt{vbeta} ~~variance of the coefficient \(\beta\) in the state
+equation\\
+\texttt{nu} ~~when divided by 2, first hyperparameter of the Inverse
+Gamma distribution on \(\tau^2\), the conditional variance of the
+state\\
+\texttt{lambda} ~~when multiplied by \(\nu/2\), second hyperparameter of
+the Inverse Gamma distribution on \(\tau^2\), the conditional variance
+of the state\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{SVLWfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)\{}
+
+\NormalTok{  xs =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  pars   =}\StringTok{ }\KeywordTok{cbind}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(N,ealpha,}\KeywordTok{sqrt}\NormalTok{(valpha)),}\KeywordTok{rnorm}\NormalTok{(N,ebeta,}\KeywordTok{sqrt}\NormalTok{(vbeta)),}
+                 \KeywordTok{log}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\KeywordTok{rgamma}\NormalTok{(N,nu}\OperatorTok{/}\DecValTok{2}\NormalTok{,nu}\OperatorTok{*}\NormalTok{lambda}\OperatorTok{/}\DecValTok{2}\NormalTok{)))}
+\NormalTok{  delta  =}\StringTok{ }\FloatTok{0.75}
+\NormalTok{  a      =}\StringTok{ }\NormalTok{(}\DecValTok{3}\OperatorTok{*}\NormalTok{delta}\DecValTok{-1}\NormalTok{)}\OperatorTok{/}\NormalTok{(}\DecValTok{2}\OperatorTok{*}\NormalTok{delta)}
+\NormalTok{  h2     =}\StringTok{ }\DecValTok{1}\OperatorTok{-}\NormalTok{a}\OperatorTok{^}\DecValTok{2}
+\NormalTok{  parss  =}\StringTok{ }\KeywordTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\KeywordTok{c}\NormalTok{(N,}\DecValTok{3}\NormalTok{,n))}
+\NormalTok{  xss    =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  ws     =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  ESS    =}\StringTok{ }\OtherTok{NULL}
+\NormalTok{  w      =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{  vpars   =}\StringTok{ }\OtherTok{NULL}
+  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+    \CommentTok{#mpar        = apply(pars,2,mean) #alternatively use this }
+\NormalTok{    mpar<-}\KeywordTok{c}\NormalTok{()}
+    \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{)\{}
+\NormalTok{      mpar[i]    =}\StringTok{ }\KeywordTok{weighted.mean}\NormalTok{(pars[,i],w)}
+\NormalTok{    \}}
+
+\NormalTok{    vpar     =}\StringTok{ }\KeywordTok{var}\NormalTok{(pars)}
+\NormalTok{    ms          =}\StringTok{ }\NormalTok{a}\OperatorTok{*}\NormalTok{pars}\OperatorTok{+}\NormalTok{(}\DecValTok{1}\OperatorTok{-}\NormalTok{a)}\OperatorTok{*}\KeywordTok{matrix}\NormalTok{(mpar,N,}\DecValTok{3}\NormalTok{,}\DataTypeTok{byrow=}\NormalTok{T)}
+\NormalTok{    mus         =}\StringTok{ }\NormalTok{pars[,}\DecValTok{1}\NormalTok{]}\OperatorTok{+}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\OperatorTok{*}\NormalTok{xs }
+\NormalTok{    weight      =}\StringTok{ }\NormalTok{w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(mus}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    k           =}\StringTok{ }\KeywordTok{sample}\NormalTok{(}\DecValTok{1}\OperatorTok{:}\NormalTok{N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{weight)}
+\NormalTok{    ms1         =}\StringTok{ }\NormalTok{ms[k,] }\OperatorTok{+}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\KeywordTok{rnorm}\NormalTok{(}\DecValTok{3}\OperatorTok{*}\NormalTok{N),N,}\DecValTok{3}\NormalTok{)}\OperatorTok{%*%}\KeywordTok{chol}\NormalTok{(h2}\OperatorTok{*}\NormalTok{vpar)}
+\NormalTok{    xt          =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,ms1[,}\DecValTok{1}\NormalTok{]}\OperatorTok{+}\NormalTok{ms1[,}\DecValTok{2}\NormalTok{]}\OperatorTok{*}\NormalTok{xs[k],}\KeywordTok{exp}\NormalTok{(ms1[,}\DecValTok{3}\NormalTok{]}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    w           =}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(xt}\OperatorTok{/}\DecValTok{2}\NormalTok{))}\OperatorTok{/}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(mus[k]}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+\NormalTok{    w           =}\StringTok{ }\NormalTok{w}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w)}
+\NormalTok{    ESS         =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\NormalTok{(N}\OperatorTok{/}\DecValTok{2}\NormalTok{))\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      xs<-xt[index]}
+\NormalTok{      pars<-ms1[index,]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
+\NormalTok{      xs<-xt}
+\NormalTok{      pars<-ms1}
+\NormalTok{    \}}
+
+\NormalTok{    vpars        =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(vpars,vpar)}
+\NormalTok{    xss         =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xss,xs)}
+\NormalTok{    parss[,,t]  =}\StringTok{ }\NormalTok{pars }
+\NormalTok{    ws          =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{  \}}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xss,}\DataTypeTok{pars=}\NormalTok{parss,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{vpars=}\NormalTok{vpars))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hrule
+\hrule
+
+\texttt{SISfun(data,N,m0,C0,alpha,beta,tau,r)}\\
 
@@
 \begin{Highlighting}[]
-\NormalTok{SVLWfun}\OtherTok{\textless{}{-}}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)\{}
-
-\NormalTok{  xs }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,m0,}\FunctionTok{sqrt}\NormalTok{(C0))}
-\NormalTok{  pars   }\OtherTok{=} \FunctionTok{cbind}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(N,ealpha,}\FunctionTok{sqrt}\NormalTok{(valpha)),}\FunctionTok{rnorm}\NormalTok{(N,ebeta,}\FunctionTok{sqrt}\NormalTok{(vbeta)),}
-                 \FunctionTok{log}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\FunctionTok{rgamma}\NormalTok{(N,nu}\SpecialCharTok{/}\DecValTok{2}\NormalTok{,nu}\SpecialCharTok{*}\NormalTok{lambda}\SpecialCharTok{/}\DecValTok{2}\NormalTok{)))}
-\NormalTok{  delta  }\OtherTok{=} \FloatTok{0.75}
-\NormalTok{  a      }\OtherTok{=}\NormalTok{ (}\DecValTok{3}\SpecialCharTok{*}\NormalTok{delta}\DecValTok{{-}1}\NormalTok{)}\SpecialCharTok{/}\NormalTok{(}\DecValTok{2}\SpecialCharTok{*}\NormalTok{delta)}
-\NormalTok{  h2     }\OtherTok{=} \DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a}\SpecialCharTok{\^{}}\DecValTok{2}
-\NormalTok{  parss  }\OtherTok{=} \FunctionTok{array}\NormalTok{(}\DecValTok{0}\NormalTok{,}\FunctionTok{c}\NormalTok{(N,}\DecValTok{3}\NormalTok{,n))}
-\NormalTok{  xss    }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  ws     }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  ESS    }\OtherTok{=} \ConstantTok{NULL}
-\NormalTok{  w      }\OtherTok{=} \FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-\NormalTok{  vpars   }\OtherTok{=} \ConstantTok{NULL}
-  \ControlFlowTok{for}\NormalTok{ (t }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(data))\{}
-
-    \CommentTok{\#mpar        = apply(pars,2,mean) \#alternatively use this }
-\NormalTok{    mpar}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{()}
-    \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{3}\NormalTok{)\{}
-\NormalTok{      mpar[i]    }\OtherTok{=} \FunctionTok{weighted.mean}\NormalTok{(pars[,i],w)}
+\NormalTok{SISfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x,tau)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{<}\DecValTok{0}\NormalTok{)\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
 \NormalTok{  \}}
-
-\NormalTok{    vpar     }\OtherTok{=} \FunctionTok{var}\NormalTok{(pars)}
-\NormalTok{    ms          }\OtherTok{=}\NormalTok{ a}\SpecialCharTok{*}\NormalTok{pars}\SpecialCharTok{+}\NormalTok{(}\DecValTok{1}\SpecialCharTok{{-}}\NormalTok{a)}\SpecialCharTok{*}\FunctionTok{matrix}\NormalTok{(mpar,N,}\DecValTok{3}\NormalTok{,}\AttributeTok{byrow=}\NormalTok{T)}
-\NormalTok{    mus         }\OtherTok{=}\NormalTok{ pars[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{pars[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{xs }
-\NormalTok{    weight      }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{*}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(mus}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
-\NormalTok{    k           }\OtherTok{=} \FunctionTok{sample}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\NormalTok{N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{weight)}
-\NormalTok{    ms1         }\OtherTok{=}\NormalTok{ ms[k,] }\SpecialCharTok{+} \FunctionTok{matrix}\NormalTok{(}\FunctionTok{rnorm}\NormalTok{(}\DecValTok{3}\SpecialCharTok{*}\NormalTok{N),N,}\DecValTok{3}\NormalTok{)}\SpecialCharTok{\%*\%}\FunctionTok{chol}\NormalTok{(h2}\SpecialCharTok{*}\NormalTok{vpar)}
-\NormalTok{    xt          }\OtherTok{=} \FunctionTok{rnorm}\NormalTok{(N,ms1[,}\DecValTok{1}\NormalTok{]}\SpecialCharTok{+}\NormalTok{ms1[,}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{xs[k],}\FunctionTok{exp}\NormalTok{(ms1[,}\DecValTok{3}\NormalTok{]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
-\NormalTok{    w           }\OtherTok{=} \FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(xt}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}\SpecialCharTok{/}\FunctionTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\FunctionTok{exp}\NormalTok{(mus[k]}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))}
-\NormalTok{    w           }\OtherTok{=}\NormalTok{ w}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w)}
-\NormalTok{    ESS         }\OtherTok{=} \DecValTok{1}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(w}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}
-
-    \ControlFlowTok{if}\NormalTok{(ESS}\SpecialCharTok{\textless{}}\NormalTok{(N}\SpecialCharTok{/}\DecValTok{2}\NormalTok{))\{}
-\NormalTok{      index}\OtherTok{\textless{}{-}}\FunctionTok{sample}\NormalTok{(N,}\AttributeTok{size=}\NormalTok{N,}\AttributeTok{replace=}\NormalTok{T,}\AttributeTok{prob=}\NormalTok{w)}
-\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xt[index]}
-\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{ms1[index,]}
-\NormalTok{      w}\OtherTok{\textless{}{-}}\FunctionTok{rep}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{N,N)}
-\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{}
-\NormalTok{      xs}\OtherTok{\textless{}{-}}\NormalTok{xt}
-\NormalTok{      pars}\OtherTok{\textless{}{-}}\NormalTok{ms1}
-\NormalTok{    \}}
-
-\NormalTok{    vpars        }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(vpars,vpar)}
-\NormalTok{    xss         }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(xss,xs)}
-\NormalTok{    parss[,,t]  }\OtherTok{=}\NormalTok{ pars }
-\NormalTok{    ws          }\OtherTok{=} \FunctionTok{rbind}\NormalTok{(ws,w)}
-\NormalTok{  \}}
-  \FunctionTok{return}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{xs=}\NormalTok{xss,}\AttributeTok{pars=}\NormalTok{parss,}\AttributeTok{ws=}\NormalTok{ws,}\AttributeTok{vpars=}\NormalTok{vpars))}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
 \NormalTok{\}}
@@
 
+\hrule
+\hrule
+
+\texttt{BAPFfun(data,N,m0,C0,alpha,beta,tau,r)}\\
+
+\hrule
+
+\textbf{Arguments}
+
+\texttt{data} ~~the observed process. It has to be a vector or a
+univariate time series.\\
+\texttt{N} ~~number of particles generated at each step\\
+\texttt{m0} ~~central value of the normal prior state distribution\\
+\texttt{C0} ~~variance of the normal prior state distribution\\
+\texttt{alpha} ~~intercept \(\alpha\) in the state equation\\
+\texttt{beta} ~~coefficient \(\beta\) in the state equation\\
+\texttt{tau} ~~square root of \(\tau^2\), the conditional variance of
+the state\\
+\texttt{r} ~~ if present the threshold is set equal to \(N/r\)
+otherwise, if missing, the threshold is set equal to \(N/2\)
+
+\hrule
+\hrule
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{BAPFfun<-}\ControlFlowTok{function}\NormalTok{(data,N,m0,C0,alpha,beta,tau,r)\{}
+  \ControlFlowTok{if}\NormalTok{(}\KeywordTok{missing}\NormalTok{(r))\{r=}\DecValTok{2}\NormalTok{\}}\ControlFlowTok{else}\NormalTok{\{\}}
+\NormalTok{  xs<-}\OtherTok{NULL}
+\NormalTok{  ws<-}\OtherTok{NULL}
+\NormalTok{  ess<-}\OtherTok{NULL}
+\NormalTok{  x  =}\StringTok{ }\KeywordTok{rnorm}\NormalTok{(N,m0,}\KeywordTok{sqrt}\NormalTok{(C0))}
+\NormalTok{  w  =}\StringTok{ }\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+
+  \ControlFlowTok{for}\NormalTok{(t }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\KeywordTok{length}\NormalTok{(data))\{}
+
+\NormalTok{    x<-}\KeywordTok{rnorm}\NormalTok{(N,alpha}\OperatorTok{+}\NormalTok{beta}\OperatorTok{*}\NormalTok{x,tau)}
+\NormalTok{    w1<-w}\OperatorTok{*}\KeywordTok{dnorm}\NormalTok{(data[t],}\DecValTok{0}\NormalTok{,}\KeywordTok{exp}\NormalTok{(x}\OperatorTok{/}\DecValTok{2}\NormalTok{))}
+
+\NormalTok{    w =}\StringTok{ }\NormalTok{w1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w1)}
+\NormalTok{    ESS  =}\StringTok{ }\DecValTok{1}\OperatorTok{/}\KeywordTok{sum}\NormalTok{(w}\OperatorTok{^}\DecValTok{2}\NormalTok{)}
+
+    \ControlFlowTok{if}\NormalTok{(ESS}\OperatorTok{>-}\DecValTok{1}\NormalTok{)\{}
+\NormalTok{      index<-}\KeywordTok{sample}\NormalTok{(N,}\DataTypeTok{size=}\NormalTok{N,}\DataTypeTok{replace=}\NormalTok{T,}\DataTypeTok{prob=}\NormalTok{w)}
+\NormalTok{      x<-x[index]}
+\NormalTok{      w<-}\KeywordTok{rep}\NormalTok{(}\DecValTok{1}\OperatorTok{/}\NormalTok{N,N)}
+\NormalTok{    \}}\ControlFlowTok{else}\NormalTok{\{\}}
+
+\NormalTok{    xs =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(xs,x)}
+\NormalTok{    ws =}\StringTok{ }\KeywordTok{rbind}\NormalTok{(ws,w)}
+\NormalTok{    ess =}\KeywordTok{rbind}\NormalTok{(ess,ESS)}
+\NormalTok{  \}}
+  \KeywordTok{return}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{xs=}\NormalTok{xs,}\DataTypeTok{ws=}\NormalTok{ws,}\DataTypeTok{ess=}\NormalTok{ess))}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \backmatter

### 2021-08-10 prova_knit_finale.Rmd
little change
cambio APF

@@
 \begin{equation}
-    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrigtharrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
+    \frac{1}{N}\sum_{n=1}^Nw(X^n) \varphi (X^n)\xrightarrow{\text{a.s.}}  \E_\M(\varphi \cdot w)=\E_\Q(\varphi)\cdot \E_\M(w)
 \end{equation}
@@
 \begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$
+    \item Set the indexes: $A_t^{n}=n$ for $n=1,...,N$.
     \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
@@
 The GPF allowed to sample from a transition kernel different from the correct ones.
-The APF, instead, allows to extract the ancestor variables from an arbitrary distribution at the resampling step.\ \
+Instead, the APF allows to extract the ancestor variables from an arbitrary distribution at the resampling step.\ \
 Let $\eta_t:\X\rightarrow \R_+$ be a non-negative, real-valued positive function, called \textit{auxiliary function}.
@@
 \end{equation}
-It is possible to derive an optimality result for the auxiliary functions\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p.
-148}.
-In the special case in which the transition kernel used for sampling corresponds with the optimal one $M_t^*$, the optimal auxiliary function (i.e., the auxiliary function that minimizes the variance of the inferential weights) becomes:
+It is possible to derive an optimality result for the auxiliary functions.\footnote{Details in Chopin and Papaspiliopoulos (2020) th 10.2, p.
+148.} In the special case in which the transition kernel used for sampling corresponds with the optimal one $M_t^*$, the optimal auxiliary function (i.e., the auxiliary function that minimizes the variance of the inferential weights) becomes:
 \begin{equation}
@@
 See Johansen and Evers (2007), p.
-120}
-    \end{enumerate}
-\end{enumerate}
-However, it is worth noting that the optimality results both for the kernel in the GPF and of the auxiliary  in the APF, has  two main weaknesses:
+120.\label{footnote_apf}}
+    \end{enumerate}
+\end{enumerate}
+However, it is worth noting that the optimality results both for the kernel in the GPF and of the auxiliary function in the APF, have  two main weaknesses:
 \begin{itemize}
     \item it is not clear that what is optimal at time $t$ remains such at future stages.
-In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affects the efficiency of the PF estimators;
+In other words, it is not truly clear how the minimization of the variance of the inferential weights at each stage affects the overall efficiency of the PF estimators;
     \item both the choice of $M_t$ and of $\eta_t$ affect the way in which weights are computed and, indirectly, the decision to resample in the future.
@@
 
-For illustration purposes, we are going implement an auxiliary particle filter for the linear Gaussian model with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.
+For illustration purposes, we are going implement an auxiliary particle filter for the linear Gaussian model with auxiliary function $g(x_{t-1})=E(x_{t}|x_{t-1})=x_{t-1}$.  \
+
+Firstly, states are drawn from the true transition kernel, while a second resampling step is done at the end of each iteration, according to the original formulation of the algorithm (see footnote \ref{footnote_apf}).
 
@@
 
-We also implement the auxiliary particle filter with optimal transition kernel. Therfore the algorithm is similiar to the one discussed above with the difference that:
+We also implement the auxiliary particle filter with optimal transition kernel.
+Note that we are also using the perfectly adapted auxiliary function and that we resample only once, at the beginning of each iteration.
+Therefore the algorithm is similar to the one discussed above with the difference that:
 

### 2021-08-10 Real Dataset SV.Rmd

@@
 For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter.
-(Metterei qua l'intuizione del Liu West)\
+Although the latter algorithm was outperformed in the analysis we conducted on synthetic data, in fact we expected such reversal of fortunes: intuitively, while the rest of the field is built on a model with fixed guesses for the unknown parameters, the Liu and West filters enjoys more flexibility thanks to the higher variability implied by its additional hierarchical level and by its parameter learning, two features that should both pay off in real data analyses.\
 
@@
 In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF.
-Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a RMSE and a MAE.
+Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a 0.3794359 RMSE and a 0.2687336 MAE.
 In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500.
@@
 The idea behind employing also such 50000 particles approximations (\textit{i.e} more precise and computationally expensive ones) is to check whether the relative performance that we assessed with N=10000 might be significantly influenced by issues of convergence among the filters: possibly, by further approaching the exact filtering distribution implied by the basic SV model, the filters built on such specification would improve their relative performance.
-This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Nota che riporta il fatto che in certi casi si ottengano MAE e RMSE più bassi usando N più bassi di 10000: in altre parole, approssimare un po' peggio la distribuzione implicata dal modello SV può portare a rispettare meglio la proxy.}\
+This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Indeed, it can be argued that the issue is almost the opposite one, as in some cases the filters replicate the proxy better when they are further from convergence.
+For example, note how comparing the following table for the SP500 dataset with that of the main text reveals that some filters perform better when N=1000 as compared to N=10000, even though we know that at N=10000 they replicate their target distribution better.
+```{r include=FALSE}
+
+SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<0){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+# Basis Particle Filter (Resampling at each step)
+#---------------
+BAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS>-1){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+# Bootstrap Particle Filter
+#---------------
+BPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#
+#-----------------------
+#Guided Particle Filter Optimal Kernel
+#------------------------------
+GPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Auxiliary Particle Filter
+#-------------------------
+APFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],0,exp(x/2))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k],tau)
+    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Auxiliary Particle Filter (PETRONE p 218)
+#-------------------------
+SVAPFfun_2<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+    x_prev=x
+    x_hat = alpha+beta*x #means of the transition from the previous particle 
+    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
+    for(k in 1:N){
+      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
+      w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
+    }
+    w = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Liu West
+#--------
+LWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+  xs = rnorm(N,m0,sqrt(C0))
+  pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
+                 log(1/rgamma(N,nu/2,nu*lambda/2)))
+  delta  = 0.75
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,3,n))
+  xss    = NULL
+  ws     = NULL
+  ESS    = NULL
+  w      = rep(1/N,N)
+  vpars   = NULL
+  for (t in 1:length(data)){
+
+    #mpar        = apply(pars,2,mean) #alternatively use this 
+    mpar<-c()
+    for(i in 1:3){
+      mpar[i]    = weighted.mean(pars[,i],w)
+    }
+
+    vpar     = var(pars)
+    ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
+    mus         = pars[,1]+pars[,2]*xs 
+    weight      = w*dnorm(data[t],0,exp(mus/2))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+    ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
+    xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
+    w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/2)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xt[index]
+      pars<-ms1[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xt
+      pars<-ms1
+    }
+
+    vpars        = rbind(vpars,vpar)
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+  }
+  return(list(xs=xss,pars=parss,ws=ws,vpars=vpars))
+}
+#Set hyperparameters
+#-------------------
+#n     =  length(y)
+alpha =  0
+beta  =  0.99
+tau2  =  0.05
+m0      = 0
+C0      = 100 #N(0,100) is an informative prior about the mean of the volatility. It is used by Jacquier et al. (2004), and it is calibrated on percentage returns
+sC0     = sqrt(C0)
+ealpha  = alpha
+valpha  = 0.01
+ebeta    = beta
+vbeta    = 0.01
+nu      = 5
+lambda  = tau2*3/5
+tau   = sqrt(tau2)
+
+#Real dataset
+
+library(openxlsx)
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y<-realdatasetSPX[,2]
+n = length(y)
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x<-volatilitySPX[,3]
+alpha.true = alpha
+beta.true  = beta
+tau2.true  = tau2
+timeframe<-c(1:length(y))
+dfsv<-data.frame(timeframe,y,x)
+
+#Filtering
+#---------
+set.seed(12345)
+N=1000
+svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svgpfopt<-GPFoptfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+
+Filtervalues<-function(fun){
+  xhat<-sapply(1:n,function(i)
+    weighted.mean(fun$xs[i,],fun$ws[i,]))
+  sdhat<-sapply(1:n,function(i)
+    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))
+
+  return(list(mean=xhat,sd=sdhat))
+
+}
+
+## RMSE MAE comparison
+
+realisedx<-x
+ErrorvolSPX<-matrix(NA,ncol=7,nrow=2)
+colnames(ErrorvolSPX)<-c("N","BPF", "GPFOPT", "APF","BAPF","LWF", "SIS")
+rownames(ErrorvolSPX)<-c("RMSE SP500","MAE SP500")
+ErrorvolSPX[,1]<-c(1000)
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp((Filtervalues(fun)$mean)/2)}
+ErrorvolSPX[1,2]<-RMSE(realisedx,comparablevol(svbpf))
+ErrorvolSPX[1,3]<-RMSE(realisedx,comparablevol(svgpfopt))
+ErrorvolSPX[1,4]<-RMSE(realisedx,comparablevol(svapf))
+ErrorvolSPX[1,5]<-RMSE(realisedx,comparablevol(svbapf))
+ErrorvolSPX[1,6]<-RMSE(realisedx,comparablevol(svlw))
+ErrorvolSPX[1,7]<-RMSE(realisedx,comparablevol(svsis))
+ErrorvolSPX[2,2]<-MAE(realisedx,comparablevol(svbpf))
+ErrorvolSPX[2,3]<-MAE(realisedx,comparablevol(svgpfopt))
+ErrorvolSPX[2,4]<-MAE(realisedx,comparablevol(svapf))
+ErrorvolSPX[2,5]<-MAE(realisedx,comparablevol(svbapf))
+ErrorvolSPX[2,6]<-MAE(realisedx,comparablevol(svlw))
+ErrorvolSPX[2,7]<-MAE(realisedx,comparablevol(svsis))
+```
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(ErrorvolSPX, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SPX)",digits=5)
+```
+}\
 Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.

### 2021-08-10 prova_knit_finale.Rmd
altre modifiche

@@
 For instance, in our following implementation with the random walk model we use an auxiliary particle LWF.}\
-Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
+Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi_0(\cdot)$ be the prior distribution for the parameter $\psi$.\
 The LWF can be described by the following algorithm.
 \begin{enumerate}
-    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.
+    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi^n\thicksim \pi_0(\psi)$.
 
@@
 \begin{enumerate}
-    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$
+    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$.
     \item Set the weights: $\hat w_{t-1}^n = w_{t-1}$ for $n=1,...,N$.
@@
 \end{enumerate}
-At the initial step, $N$ states and $N$ parameter values are sampled independently, respectively, from $\Pd_0$ and $\pi$ and the particle weights are set to be constant, as usual.\
+At the initial step, $N$ states and $N$ parameter values are sampled independently, respectively, from $\Pd_0$ and $\pi_0$ and the particle weights are set to be constant, as usual.\
 Then, at each iteration (time $t$), the decision to resample is done according to the $ESS$ criterion and indexes are extracted from a multinomial distribution.
@@
         \E_{\pi_{t-1}^N} ( \psi)=& \E \big[\E( \psi|A_t^{n})\big]=\E \big[\psi^{A_t^{n}}\big]=\bar \psi\\
-        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\E\big[\Omega\big]=2\cdot\Omega
+        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\E\big[\Omega\big]>\Omega
     \end{split}
@@
 \begin{equation}
-    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(\psi)
+    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(d\psi)
 \end{equation}
@@
 \subsection{Implementation}
-\
 Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$.
@@
 Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values.
-Next, we descrive the algorithm.
+Note that we are applying the LWF in the framework of APF with two resampling stage and the correct transition kernel.
+Next, we describe the algorithm.
 \begin{algorithm} LWF for Random Walk plus Noise Model
@@
 \section{Convergence}\label{pf_converg}
-In this section, we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest.
+In this section,\footnote{This section has the purpose to give only a very general idea about convergence of PF estimators.} we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest.
 Firstly, we present two convergence results for the estimators of the predictive and filtering distributions.
@@
 Then, we will show that the error of the particle estimators are asymptotically distributed according to a Gaussian distribution with known variance.
+All the convergence results that we present here involve the limit for the number of particles $N$ at each iteration.\
 Henceforth, we will maintain the assumption that (multinomial) resampling is repeated at each iteration.

### 2021-08-10 Real Dataset SV.Rmd
commit

@@
 Errorcomp1[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
-Errorcomp1[,2]<-c(0.4233094,0.15689352,0.06901099,0.03817009,0.03045272,0.2807802,0.09658088,0.03312132,0.01889245,0.01455814)
-Errorcomp1[,3]<-c(0.3685259,0.15094281,0.07669050,0.03568058,0.02709059,0.2796141,0.09623692,0.03696833,0.01664805,0.01422964)
-Errorcomp1[,4]<-c(0.5478362,0.2404408,0.08878446,0.05742351,0.04295610,0.4310335,0.1756985,0.05538968,0.03186140,0.02448565)
-Errorcomp1[,5]<-c(0.5595929,0.2348842,0.10295053,0.06109947,0.05151115,0.4377886,0.1839363,0.06726034,0.03448388,0.02726281)
-Errorcomp1[,6]<-c(0.9168639,1.1055065,0.6250516,0.6200359,0.5118927,0.7275327,0.9179852,0.4971839,0.4475265,0.406363)
+Errorcomp1[,2]<-c(0.3017986,0.11645976,0.02906876,0.01511962,0.010462488,0.2342816,0.08340162,0.02145381,0.01110398,0.008013466)
+Errorcomp1[,3]<-c(0.3367199,0.09173166,0.03356642,0.01461339,0.009572002,0.2539500,0.06764348,0.02447722,0.01026689,0.007198043)
+Errorcomp1[,4]<-c(0.4822315,0.1647978,0.05644653,0.02657352,0.02006171,0.3648792,0.1159503,0.03722911,0.01879724,0.01420517)
+Errorcomp1[,5]<-c(0.4561433,0.1740267,0.04786161,0.02741662,0.02085289,0.3643665,0.1352035,0.03833600,0.01939683,0.01481767)
+Errorcomp1[,6]<-c(1.566291,1.770554,1.1788696,0.9680993,0.9301619,1.256867,1.396812,0.8683638,0.7212099,0.7182857)
 
@@
 Errorcomp2[,1]<-c(10,100,1000,5000,10000,10,100,1000,5000,10000)
-Errorcomp2[,2]<-c(0.4054868,0.13510456,0.04145471,0.02847469,0.03048093,0.2726597,0.08539766,0.02328878,0.01755582,0.01417144)
-Errorcomp2[,3]<-c(0.3495018,0.14208597,0.06873795,0.04921716,0.03336987,0.2694089,0.09033775,0.03531985,0.02409677,0.01717600)
-Errorcomp2[,4]<-c(0.5323702,0.2224267,0.07796291,0.03931358,0.02647522,0.4215135,0.1669123,0.05105029,0.02662432,0.01973858)
-Errorcomp2[,5]<-c(0.5459719,0.2239766,0.07692431,0.05229897,0.02410626,0.4290946,0.1748428,0.05753199,0.03275473,0.01741627)
-Errorcomp2[,6]<-c(0.9103835,1.101557,0.6150409,0.6117618,0.5054665,0.7211173,0.914675,0.4895908,0.4419522,0.4022079)
+Errorcomp2[,2]<-c(0.2998373,0.11719984,0.03339627,0.02213888,0.01879835,0.2334205,0.08306614,0.02414570,0.01444042,0.01184698)
+Errorcomp2[,3]<-c(0.3365006,0.09527023,0.03979288,0.02073930,0.01744916,0.2540961,0.06926543,0.02793461,0.01360319,0.01149824)
+Errorcomp2[,4]<-c(0.4828496,0.1665464,0.05696147,0.03056349,0.01929485,0.3651631,0.1174350,0.03692383,0.02031451,0.01437991)
+Errorcomp2[,5]<-c(0.4555563,0.1715253,0.04847040,0.02916413,0.02211777,0.3634416,0.1343189,0.03877838,0.02075878,0.01495394)
+Errorcomp2[,6]<-c(1.565179,1.770452,1.1782202,0.9681921,0.9298538,1.254832,1.397196,0.8666459,0.7203661,0.7181958)
 
@@
 
-#Auxiliary Particle Filter (PETRONE p 218)
-#-------------------------
-SVAPFfun_2<-function(data,N,m0,C0,alpha,beta,tau,r){
-  if(missing(r)){r=2}else{}
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-    x_prev=x
-    x_hat = alpha+beta*x #means of the transition from the previous particle 
-    I_k = sample(1:N,prob=w*dnorm(data[t],0,sqrt(exp((x_hat)/2))))
-    for(k in 1:N){
-      x[k]=rnorm(size=1,mean=alpha+beta*x_prev[I_k],sd=tau)
-      w[k]=dnorm(data[t],0,sqrt(exp((x[k])/2)))/dnorm(data[t],0,sqrt(exp((x_hat)/2)))
-    }
-    w = w/sum(w)
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/r)){
-      index<-sample(N,size=N,replace=T,prob=w)
-      x<-x[index]
-      w<-rep(1/N,N)
-    }else{}
-
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-
 #Liu West

### 2021-08-10 Real Dataset SV.Rmd
Aggiunta grafici

@@
 Plots that are compared represent the sequence of percentage log returns for a given index (the observations, in gray) and the sequence of filtered volatilities calculated using filtered states (in red) obtained through a specific filtering technique indicated in the caption of the plot.\footnote{Such filtered volatilities are obtained as follows.
-For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t}$.
+For each $t$, let us denote by $\mu_t$ the mean of the generated particles; the filtered volatility is then defined as $e^{\mu_t/2}$.
 Note that the number of particles was set at 10000 for each filter: the conclusion of the next section will come back to such a choice.} \
 
+```{r echo=FALSE, fig.align='center', fig.cap="Observations and filtered volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+# Filtered Volatility Plots
+
+library(openxlsx)
+library(ggplot2)
+library(ggpubr)
+
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y1<-realdatasetSPX[,2]
+timeframe<-c(1:length(y1))
+
+sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
+dfsv1<-data.frame(timeframe,y1,sv_bpf1)
+dfsv2<-data.frame(timeframe,y1,sv_gpfopt1)
+dfsv3<-data.frame(timeframe,y1,sv_apf1)
+dfsv4<-data.frame(timeframe,y1,sv_lw1)
+dfsv5<-data.frame(timeframe,y1,sv_bapf1)
+dfsv6<-data.frame(timeframe,y1,sv_sis1)
+
+Filtplot1<-function(dataframe,title){
+  mean=sv_bpf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot2<-function(dataframe,title){
+  mean=sv_gpfopt1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot3<-function(dataframe,title){
+  mean=sv_apf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot4<-function(dataframe,title){
+  mean=sv_lw1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot5<-function(dataframe,title){
+  mean=sv_bapf1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot6<-function(dataframe,title){
+  mean=sv_sis1
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotobs1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
+plotobs2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
+plotobs3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
+plotobs4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
+plotobs5<-Filtplot5(dfsv5,"Always Resampling SP500")
+plotobs6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+ggarrange(plotobs1,plotobs2)
+ggarrange(plotobs3,plotobs4)
+ggarrange(plotobs5,plotobs6)
+
+realdatasetSTOXX<-read.xlsx("Dataset 17-21.xlsx", sheet = 2)
+y2<-realdatasetSTOXX[,2]
+timeframe<-c(1:length(y2))
+
+sv_bpf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis2<-(read.xlsx("STOXX_filters_10000p.xlsx", sheet = 6))[,1]
+dfsv7<-data.frame(timeframe,y2,sv_bpf2)
+dfsv8<-data.frame(timeframe,y2,sv_gpfopt2)
+dfsv9<-data.frame(timeframe,y2,sv_apf2)
+dfsv10<-data.frame(timeframe,y2,sv_lw2)
+dfsv11<-data.frame(timeframe,y2,sv_bapf2)
+dfsv12<-data.frame(timeframe,y2,sv_sis2)
+
+Filtplot7<-function(dataframe,title){
+  mean=sv_bpf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot8<-function(dataframe,title){
+  mean=sv_gpfopt2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot9<-function(dataframe,title){
+  mean=sv_apf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot10<-function(dataframe,title){
+  mean=sv_lw2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot11<-function(dataframe,title){
+  mean=sv_bapf2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot12<-function(dataframe,title){
+  mean=sv_sis2
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotobs7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX50E")
+plotobs8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX50E")
+plotobs9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX50E")
+plotobs10<-Filtplot10(dfsv10,"Liu and West Filter STOXX50E")
+plotobs11<-Filtplot11(dfsv11,"Always Resampling STOXX50E")
+plotobs12<-Filtplot12(dfsv12,"No Resampling STOXX50E")
+
+ggarrange(plotobs7,plotobs8)
+ggarrange(plotobs9,plotobs10)
+ggarrange(plotobs11,plotobs12)
+
+realdatasetDJI<-read.xlsx("Dataset 17-21.xlsx", sheet = 3)
+y3<-realdatasetDJI[,2]
+timeframe<-c(1:length(y3))
+
+sv_bpf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 3))[,1]
+sv_lw3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 4))[,1]
+sv_bapf3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis3<-(read.xlsx("DJI_filters_10000p.xlsx", sheet = 6))[,1]
+dfsv13<-data.frame(timeframe,y3,sv_bpf3)
+dfsv14<-data.frame(timeframe,y3,sv_gpfopt3)
+dfsv15<-data.frame(timeframe,y3,sv_apf3)
+dfsv16<-data.frame(timeframe,y3,sv_lw3)
+dfsv17<-data.frame(timeframe,y3,sv_bapf3)
+dfsv18<-data.frame(timeframe,y3,sv_sis3)
+
+Filtplot13<-function(dataframe,title){
+  mean=sv_bpf3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot14<-function(dataframe,title){
+  mean=sv_gpfopt3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot15<-function(dataframe,title){
+  mean=sv_apf3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot16<-function(dataframe,title){
+  mean=sv_lw3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot17<-function(dataframe,title){
+  mean=sv_bapf3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot18<-function(dataframe,title){
+  mean=sv_sis3
+  mean1=exp(mean/2)
+  dataframe<-data.frame(dataframe,mean1)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
+    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
+    scale_color_manual("",
+                       values=c("Observations" = "gray",
+                                "Filtered Volatility"= "red"))+
+    scale_linetype_manual("",
+                          values=c("Observations" = 1,
+                                   "Filtered Volatility"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotobs13<-Filtplot13(dfsv13,"Bootstrap Particle Filter DJI")
+plotobs14<-Filtplot14(dfsv14,"Opt Ker Guided Particle Filter DJI")
+plotobs15<-Filtplot15(dfsv15,"Auxiliary Particle Filter DJI")
+plotobs16<-Filtplot16(dfsv16,"Liu and West Filter DJI")
+plotobs17<-Filtplot17(dfsv17,"Always Resampling DJI")
+plotobs18<-Filtplot18(dfsv18,"No Resampling DJI")
+
+ggarrange(plotobs13,plotobs14)
+ggarrange(plotobs15,plotobs16)
+ggarrange(plotobs17,plotobs18)
+
+```
+
 First, a general point is that the sequence of percentage log returns are different across the three indices.
-The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents (SEE SECTION X).
+The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents.
 In particular, as previously noticed, there are extensive similarities between the DOW JONES Index and the S&P500, which can be visualized when looking at the percentage of log returns in the corresponding plots, since both are describing the US Economy.
 However, the STOXX50E differs quite extensively, especially in the different sensitivity to the Covid-19 Crisis and its consequences,  representing the European Economy.
-Aside from this note, the considerations expressed below on the sequences "filtered volatilities"\footnote{Per ora nei grafici c'è scritto "Filtered States", verrà corretto.} for the various techniques hold almost uniformly for all the three indices.
+Aside from this note, the considerations expressed below on the sequences "filtered volatilities" for the various techniques hold almost uniformly for all the three indices.
 Therefore, in the next paragraph we will be referring to the set of plots pertaining to one specific index (which could be any of the three).\
@@
 Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
-Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Molto probabilmente la frase formulata così non è formalmente corretta...bisognerebbe fare riferimento alla sezione sulla convergenza.
-Inoltre, va chiarito il punto della "doppia convergenza".} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
+Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Riferimento alla sezione di convergenza?.} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
 With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
@@
 Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
-Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling alogrithm.
+Unsurprisingly, it will be apparent that including a resampling step is essential to avoid a collapsing effective sample size.} of the particles generated by other particles filters with N equal to 10, 100, 1000, 5000 or 10000.\footnote{We consider 5 algorithms: a Bootstrap PF, a Guided PF (with the optimal proposal kernel), an Auxiliary PF, a "Basic PF" and a Sequential Importance Sampling algorithm.
 By "Basic PF" we refer to a Bootstrap PF that features resampling at each step, whereas resampling never happens in the SIS algorithm: one can then consider these 2 algorithms as extreme cases of an adaptive resampling algorithm, namely one that features resampling when $ESS$ is non negative (\textit{i.e.} always) or when $ESS<0$ (\textit{i.e.} never).
-In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.}
+In this sense, we use the shorthand "particle filters" when referring to all the 5 algorithms.
+Finally, note that the Auxiliary Particle Filter employed here does not coincide with that of the section with synthetic data, which employed a different kernel implied by optimality results.}
 
@@
 As it turns out, such suggestion is confirmed by running the algorithms with N=1000.
-Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with even more pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles.
-Further sizable gains for every filter, except for the SIS algorithm, are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
+Passing from 100 to 1000 particles more than halves the RMSEs across the 5 competing filters, with similarly pronounced gains in the MAEs; at N=1000, the Bootstrap and the Guided particle filters still hold their lead over the rest of the field, keeping a shrinking hedge over the Auxiliary PF, which shows most impressive gains when passing from 100 to 1000 particles.
+Further sizable gains for every filter are observed when increasing the number of particles to 5000, a level at which both the RMSE and the MAE agree on a slight hedge of the Guided and the Bootstrap filters over the Auxiliary and the Basic PFs.\
 Nevertheless, we would argue that already for N=5000, but even more clearly for N=10000, the difference between the Bootstrap, the Guided and the Auxiliary particle filters is so small that none of these filters can be straightforwardly chosen over the others, a figure which visual inspection of the graphs below can confirm.
-Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only once, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
+Interestingly enough, no gains seem to arise from resampling at each step with respect to adaptive resampling (note that the Basic PF is preferred to the APF only twice, by the RMSE), whereas the need of some degree of resampling is apparent, as shown by the awful performance of the SIS algorithm with any number of particles.
 Moreover, the RMSE and the MAE show infrequent disagreement, delivering a rather clear picture of the comparison.
+
+```{r echo=FALSE, fig.align='center', fig.cap="Approximations at N=10000, Bootstrap benchmark", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+
+# Benchmark comparison Plots
+
+library(openxlsx)
+library(ggplot2)
+library(ggpubr)
+
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y<-realdatasetSPX[,2]
+btbenchmark<-(read.xlsx("svbpf50000p.xlsx", sheet = 1))[,1]
+timeframe<-c(1:length(y))
+
+sv_bpf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,1]
+sv_gpfopt1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,1]
+sv_apf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,1]
+sv_bapf1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,1]
+sv_sis1<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,1]
+sv_bpf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 1))[,2]
+sv_gpfopt1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 2))[,2]
+sv_apf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 3))[,2]
+sv_bapf1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 5))[,2]
+sv_sis1sd<-(read.xlsx("SP500_filters_10000p.xlsx", sheet = 6))[,2]
+
+dfsv1<-data.frame(timeframe,y,btbenchmark,sv_bpf1,sv_bpf1sd)
+dfsv2<-data.frame(timeframe,y,btbenchmark,sv_gpfopt1,sv_gpfopt1sd)
+dfsv3<-data.frame(timeframe,y,btbenchmark,sv_apf1,sv_apf1sd)
+dfsv5<-data.frame(timeframe,y,btbenchmark,sv_bapf1,sv_bapf1sd)
+dfsv6<-data.frame(timeframe,y,btbenchmark,sv_sis1,sv_sis1sd)
+
+Filtplot1<-function(dataframe,title){
+  mean=sv_bpf1
+  sd=sv_bpf1sd
+  dataframe<-data.frame(dataframe,mean)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
+    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("BT benchmark" = "black",
+                                "Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("BT benchmark" = 1,
+                                   "Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot2<-function(dataframe,title){
+  mean=sv_gpfopt1
+  sd=sv_gpfopt1sd
+  dataframe<-data.frame(dataframe,mean)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
+    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("BT benchmark" = "black",
+                                "Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("BT benchmark" = 1,
+                                   "Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot3<-function(dataframe,title){
+  mean=sv_apf1
+  sd=sv_apf1sd
+  dataframe<-data.frame(dataframe,mean)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
+    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("BT benchmark" = "black",
+                                "Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("BT benchmark" = 1,
+                                   "Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+Filtplot5<-function(dataframe,title){
+  mean=sv_bapf1
+  sd=sv_bapf1sd
+  dataframe<-data.frame(dataframe,mean)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
+    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("BT benchmark" = "black",
+                                "Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("BT benchmark" = 1,
+                                   "Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+Filtplot6<-function(dataframe,title){
+  mean=sv_sis1
+  sd=sv_sis1sd
+  dataframe<-data.frame(dataframe,mean)
+
+  ggplot(dataframe,aes(x=timeframe))+
+    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
+    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
+    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
+                fill="red",alpha=0.16) +
+    scale_color_manual("",
+                       values=c("BT benchmark" = "black",
+                                "Filtered States"= "red"))+
+    scale_linetype_manual("",
+                          values=c("BT benchmark" = 1,
+                                   "Filtered States"=1))+
+    labs(x="Time",
+         y="")+
+    ggtitle(title)+
+    theme_bw()+
+    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
+          legend.background = element_rect(fill = "white", colour = "gray30")) +
+    theme(plot.title = element_text(hjust = 0.5))
+}
+
+plotcomp1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
+plotcomp2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
+plotcomp3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
+plotcomp5<-Filtplot5(dfsv5,"Always Resampling SP500")
+plotcomp6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+plotcomp1
+plotcomp2
+plotcomp3
+plotcomp5
+plotcomp6
+
+```
 
@@
 Using the APF as the benchmark does not deliver a different picture than the one we previously had.
-Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at 10, 100 and 1000 particles.\
-Its performance ameliorates dramatically at N=1000, while at N=5000 the halving of its RMSE and MAE even allows the APF to overtake the Guided PF, though falling short of the Bootstrap PF.
-At N=10000 both the Basic PF and the APF achieve lower RMSEs than the Bootstrap and the Guided PF, even though the MAE criterion still selects the Bootstrap as the closest to the target distribution.\
-Interestingly, this time the performance of the Bootstrap filter worsens when N is increased from 5000 to 10000 (both according to RMSEs and MAEs, whose disagreements, once again, are overall infrequent).
-Such evidence suggests that, even at very large (yet finite) values of N, some difference between the approximated filtering distributions still persists across our particle filters.\
+Leaving the SIS algorithm aside (whose analogous performance comes as no surprise), it is interesting to note how once again the Auxiliary particle filter actually fares worse than the Bootstrap and the Guided PFs at each threshold (though very slightly so for the higher ones).
+Its performance ameliorates dramatically at N=1000, while it still shows some sizable improvements even when passing from N=5000 to N=10000, almost halving its errors.
+Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N$>>10000$.\
 
@@
 
-In fact, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=10000 is closer to the benchmark APF than the benchmark BT is, or that the RMSE ranks the APF with N=10000 as closer to the the benchmark BT than the APF with N=50000, mildly pointing at a theoretically-unsound difference between the limiting distributions of the filters.
-Nevertheless, now the RMSE and MAE do not deliver unanimous verdicts, nor the differences seem large enough to back extreme claims.
-And, conveniently, running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
+By comparing the two benchmark approximations with N=50000, we can see that some discrepancy between competing approximations still persists at high numbers of particles.
+Conveniently - as a different outcome would point at a theoretically-unsound difference between the limiting distributions of the filters - running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
 

### 2021-08-10 Real Dataset SV.Rmd
Aggiunta parte SV aggiornata

@@
 When focusing the attention on specific filters, first we can discuss the effect of resampling.
-Indeed, in the plot referring to the Sequential Importance Sampling procedure, the sequence of filtered volatilities is not able to reproduce the peaks as in with other filters, but also is characterized by collapsing confidence intervals.
-This is because ::::::: \
-Then, if we instead look at the filtered volatilities obtained with resampling, we can appreciate how the filtered sequences for the Basic,\footnote{By this name, we refer to a particle filter that features no adaptive resampling; rather, resampling is conducted at each step of the algorithm.} Bootstrap, Auxiliary and Optimal Kernel Guided particle filters give rather analogous results, with no differences appearing at a first visual inspection.\footnote{Except for the Auxiliary Particle filtered sequence of volatilities, showing a slightly higher peak corresponding to the greatest volatility fluctuation in 2020, if compared to other filters in this subset.
-This is still valid for all three indices.} \
-However, the sequence of filtered volatilities obtained by applying the Liu and West filter is clearly different from the others.
-Indeed, it shows a "more detailed" sequence of states (in the sense of the sequence of filtered volatilities being less smooth) and also a higher peak for the wide fluctuation corresponding to the 2020 Covid Crisis outbreak.
+Indeed, in the plot referring to the Sequential Importance Sampling procedure, the sequence of filtered volatilities is not able to reproduce the peaks detected by all the other filters, often pointing at completely different timing and extents of such peaks.
+Then, if we instead look at the filtered volatilities obtained with resampling, we can appreciate how the filtered sequences for the Basic,\footnote{By this name, we refer to a particle filter that features no adaptive resampling; rather, resampling is conducted at each step of the algorithm.} Bootstrap, Auxiliary and Optimal Kernel Guided particle filters give rather analogous results, with no differences appearing at a first visual inspection.\
+However, the sequence of filtered volatilities obtained by applying the Liu and West filter is somewhat different from the others.
+Indeed, it shows a "more detailed" sequence of states, meaning that the sequence of filtered volatilities is less smooth and features wider oscillations, with seemingly higher responsiveness to sudden changes in the observations.
 Therefore, we may expect that, compared to other filtering techniques, it can provide a better representation of the data.
-This is because ...
-\textit{Potremmo parlare di maggiore sensibilità alle oscillazioni dei prezzi, o simili}
 

### 2021-08-10 prova_knit_finale.Rmd
Aggiunta parte SV aggiornata

@@
 rownames(SC)<-c("Mean","Standard Deviation")
-colnames(SC) <- c("S&P500", "Dow Jones", "STOXX50E")
+colnames(SC) <- c("S&P500","Dow Jones","STOXX50E")
 
@@
 
-In fact, by comparing the two benchmark approximations with N=50000, we can see that, for example, the Bootstrap PF run with N=10000 is closer to the benchmark APF than the benchmark BT is, or that the RMSE ranks the APF with N=10000 as closer to the the benchmark BT than the APF with N=50000, mildly pointing at a theoretically-unsound difference between the limiting distributions of the filters.
-Nevertheless, now the RMSE and MAE do not deliver unanimous verdicts, nor the differences seem large enough to back extreme claims.
-And, conveniently, running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
+By comparing the two benchmark approximations with N=50000, we can see that some discrepancy between competing approximations still persists at high numbers of particles.
+Conveniently - as a different outcome would point at a theoretically-unsound difference between the limiting distributions of the filters - running the Auxiliary and the Bootstrap filters with 100000 particles testifies a shrinking difference between the approximations - at increasingly large computational costs, though.\
 
 Overall, we should not overplay tiny differences.\footnote{Indeed, the mere extent of such difference also depends on the real dataset which is used for the comparison, implying different target distributions.
-For example, running the same experiment using the STOXX50E dataset, we have even smaller differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.01236279 and mean absolute difference equal to 0.008028818.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
+For example, running the same experiment using the STOXX50E dataset, we have similar differences between the filters at N=50000, with RMS difference between the Bootstrap and the Auxiliary equal to 0.016121989 and mean absolute difference equal to 0.009306925.
+The extent of the difference clearly depends on the chosen parameters too: for example, at N=50000, choosing $\alpha=-0.0031$, $\beta=0.9951$ and $\tau^2=0.0074$, using SP500 data yields RMS difference equal to 0.04408 and MA difference equal to 0.01668, while using STOXX50E data we have 0.01236279 RMS difference and 0.008028818 mean absolute difference.} Rather, we prefer to stress the points that emerged from both benchmark analyses and we can therefore support quite confidently.\
 First, if on the one hand it is clear that some degree of resampling is essential, on the other we found no gains from resampling at each period.\
@@
 For any index and for any measure, the worst performing is the SIS algorithm, while the best performing is the Liu and West particle filter.
-(Metterei qua l'intuizione del Liu West)\
+Although the latter algorithm was outperformed in the analysis we conducted on synthetic data, in fact we expected such reversal of fortunes: intuitively, while the rest of the field is built on a model with fixed guesses for the unknown parameters, the Liu and West filters enjoys more flexibility thanks to the higher variability implied by its additional hierarchical level and by its parameter learning, two features that should both pay off in real data analyses.\
 
 ```{r include=FALSE}
+# RMSE MAE comparison
+
+library(openxlsx)
 Errorcomp3<-matrix(NA,ncol=7,nrow=6)
@@
 We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
-In the case of the SP500 index, although the number of particles increases, the RMSE and MAE increase (although by very small amounts) for both the particle filters based on our usual specification of the SV model.
-Instead, for the Liu and West filter we see some improvements, although the small extent suggests that also the LW filter was already in the convergence phase with N=10000.\
-In the case of the EuroStoxx50 index, while the Bootstrap filter increases both RMSE and MAE when N is raised to 50000, the APF shows improvements in its performance, according to both criteria, which instead disagree for the Liu and West filter.
+In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF.
+Instead, for the Liu and West filter we see a sizable deterioration in the performance: this suggests that the parameter learning process embedded in the LWF is actually pointing in a different direction than that of the proxy, suggesting that either the proxy does not capture well the latent volatility, or a SV model may not be the best tool to disclose such latent process.\footnote{Note that at N=1000, the LWF yields a 0.3794359 RMSE and a 0.2687336 MAE.
+In words, worse approximations of the target distribution of the LWF are actually best at replicating the proxy for the SP500.
+A possible interpretation goes, once again, as that reported in the main text.}\
+In the case of the EuroStoxx50 index, while the Bootstrap filter slightly increases both RMSE and MAE when N is raised to 50000, the APF shows mild improvements in its MAE performance, while the RMSE criterion points in the opposite direction.
+The criteria disagree also in the case of the LW filter, as the MAE mildly prefers the performance of N=10000, while the RMSE detects minor improvements for N=50000.
 
 ```{r include=FALSE}
+
+library(openxlsx)
 Errorcomp4<-matrix(NA,ncol=4,nrow=4)
@@
 The idea behind employing also such 50000 particles approximations (\textit{i.e} more precise and computationally expensive ones) is to check whether the relative performance that we assessed with N=10000 might be significantly influenced by issues of convergence among the filters: possibly, by further approaching the exact filtering distribution implied by the basic SV model, the filters built on such specification would improve their relative performance.
-This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Nota che riporta il fatto che in certi casi si ottengano MAE e RMSE più bassi usando N più bassi di 10000: in altre parole, approssimare un po' peggio la distribuzione implicata dal modello SV può portare a rispettare meglio la proxy.}\
-Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.
-
-\section{Bibliography}\
-
-Allen, D. E., & McAleer, M. (2020). Do we need stochastic volatility and generalised autoregressive conditional heteroscedasticity? Comparing squared end-of-day returns on FTSE. Risks, 8(1), 12.\
-
-Allen, D. E. (2020). Stochastic Volatility and GARCH: Do Squared End-of-Day Returns Provide Similar Information?. Journal of Risk and Financial Management, 13(9), 202.\
-
-Andrieu, C., Doucet, A., & Holenstein, R. (2010). Particle markov chain monte carlo methods. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 72(3), 269-342.\
-
-Barndorff‐Nielsen, O.
-E., & Shephard, N.
-(2001).
-Non‐Gaussian Ornstein--Uhlenbeck‐based models and some of their uses in financial economics.
-Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
-
-Barndorff‐Nielsen, O.
-E., & Shephard, N.
-(2002).
-Econometric analysis of realized volatility and its use in estimating stochastic volatility models.
-Journal of the Royal Statistical Society: Series B (Statistical Methodology), 64(2), 253-280.\
-
-Bollerslev, T. (1986). Generalized autoregressive conditional heteroskedasticity. Journal of econometrics, 31(3), 307-327.\
-
-Breidt, F. J., Crato, N., & De Lima, P. (1998). The detection and estimation of long memory in stochastic volatility. Journal of econometrics, 83(1-2), 325-348.\
-
-Buncic, D., & Gisler, K. I. (2016). Global equity market volatility spillovers: A broader role for the United States. International Journal of Forecasting, 32(4), 1317-1339.\
-
-Chan, J. C. (2013). Moving average stochastic volatility models with application to inflation forecast. Journal of Econometrics, 176(2), 162-172.\
-
-Chib, S., Nardari, F., & Shephard, N. (2002). Markov chain Monte Carlo methods for stochastic volatility models. Journal of Econometrics, 108(2), 281-316.\
-
-Chib, S., Nardari, F., & Shephard, N. (2006). Analysis of high dimensional multivariate stochastic volatility models. Journal of Econometrics, 134(2), 341-371.\
-
-Chopin, N., & Papaspiliopoulos, O. (2020). An introduction to sequential Monte Carlo (Vol. 4). Springer.\
-
-Comte, F., & Renault, E. (1998). Long memory in continuous‐time stochastic volatility models. Mathematical finance, 8(4), 291-323.\
-
-Christoffersen, P., Jacobs, K., & Mimouni, K.
-(2010).
-Volatility dynamics for the S&P500: Evidence from realized volatility, daily returns, and option prices.
-The Review of Financial Studies, 23(8), 3141-3189.\
-
-Dimitrakopoulos, S., & Kolossiatis, M.
-(2020).
-Bayesian analysis of moving average stochastic volatility models: modeling in-mean effects and leverage for financial time series.
-Econometric Reviews, 39(4), 319-343.\
-
-Douc, R., Moulines, E., & Stoffer, D. (2014). Nonlinear time series: Theory, methods and applications with R examples. CRC press.\
-
-Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with estimates of the variance of United Kingdom inflation. Econometrica: Journal of the econometric society, 987-1007.\
-
-Eraker, B. (2004). Do stock prices and volatility jump? Reconciling evidence from spot and option prices. The Journal of Finance, 59(3), 1367-1403.\
-
-Friz, P. K., Gassiat, P., & Pigato, P. (2021). Precise asymptotics: robust stochastic volatility models. The Annals of Applied Probability, 31(2), 896-940.\
-
-Gatheral, J., Jaisson, T., & Rosenbaum, M. (2018). Volatility is rough. Quantitative finance, 18(6), 933-949.\
-
-Guidolin, M., & Pedio, M. (2021). Media Attention vs. Sentiment as Drivers of Conditional Volatility Predictions: An Application to Brexit. Finance Research Letters, 101943.\
-
-Hansen, P. R., & Lunde, A. (2006). Consistent ranking of volatility models. Journal of Econometrics, 131(1-2), 97-121.\
-
-Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\
-
-Harvey, A., Ruiz, E., & Shephard, N. (1994). Multivariate stochastic variance models. The Review of Economic Studies, 61(2), 247-264.\
-
-Heber, Gerd, Asger Lunde, Neil Shephard and Kevin Sheppard (2009).
-Oxford-Man Institute's realized library.
-Oxford-Man Institute, University of Oxford, version 0.3.
-Available at: https://realized.oxford-man.ox.ac.uk/.
-\
-
-Heston, S. L. (1993). A closed-form solution for options with stochastic volatility with applications to bond and currency options. The review of financial studies, 6(2), 327-343.\
-
-Hull, J., & White, A. (1987). The pricing of options on assets with stochastic volatilities. The journal of finance, 42(2), 281-300.\
-
-Jacquier, E., Polson, N. G., & Rossi, P. (1994). Bayesian analysis of stochastic volatility models. Journal of Business and Economic Statistics, 12(4), 371-389.\
-
-Jacquier, E., Polson, N. G., & Rossi, P. E. (2004). Bayesian analysis of stochastic volatility models with fat-tails and correlated errors. Journal of Econometrics, 122(1), 185-212.\
-
-Johannes, M., Evers, L. (2007). Monte Carlo Methods. Lecture Notes. University of Bristol, Department of Mathematics.\
-
-Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
-
-Kambouroudis, D.
-S., McMillan, D.
-G., & Tsakou, K.
-(2016).
-Forecasting stock return volatility: a comparison of GARCH, Implied volatility, and realized volatility models.
-Journal of Futures Markets, 36(12), 1127-1163.\
-
-Kastner, G. (2019). Dealing with stochastic volatility in time series using the R package stochvol. arXiv preprint arXiv:1906.12134.\
-
-Kim, S., Shephard, N., & Chib, S. (1998). Stochastic volatility: likelihood inference and comparison with ARCH models. The review of economic studies, 65(3), 361-393.\
-
-Kim, J., & Stoffer, D.
-S.
-(2008).
-Fitting stochastic volatility models in the presence of irregular sampling via particle methods and the EM algorithm.
-Journal of time series analysis, 29(5), 811-833.\
-
-Koopman, S. J., & Hol Uspensky, E. (2002). The stochastic volatility in mean model: empirical evidence from international stock markets. Journal of applied Econometrics, 17(6), 667-689.\
-
-Liesenfeld, R., & Jung, R. C. (2000). Stochastic volatility models: conditional normality versus heavy‐tailed distributions. Journal of applied Econometrics, 15(2), 137-160.\
-
-Luo, R., Zhang, W., Xu, X., & Wang, J. (2018, April). A neural stochastic volatility model. In Thirty-second AAAI conference on artificial intelligence.\
-
-Mao, X., Czellar, V., Ruiz, E., & Veiga, H.
-(2020).
-Asymmetric stochastic volatility models: Properties and particle filter-based simulated maximum likelihood estimation.
-Econometrics and Statistics, 13, 84-105.\
-
-Nakajima, J., & Omori, Y.
-(2012).
-Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student's t-distribution.
-Computational Statistics & Data Analysis, 56(11), 3690-3704.\
-
-Omori, Y., Chib, S., Shephard, N., & Nakajima, J. (2007). Stochastic volatility with leverage: Fast and efficient likelihood inference. Journal of Econometrics, 140(2), 425-449.\
-
-Petris, G., Petrone, S., Campagnali, P. (2010). Dynamic Linear Models with R. Springer.\
-
-Pitt, M. K., & Shephard, N. (1999). Filtering via simulation: Auxiliary particle filters. Journal of the American statistical association, 94(446), 590-599.\
-
-Poon, S. H., & Granger, C. (2005). Practical issues in forecasting volatility. Financial analysts journal, 61(1), 45-56.\
-
-Shephard, N., & Sheppard, K. (2010). Realising the future: forecasting with high‐frequency‐based volatility (HEAVY) models. Journal of Applied Econometrics, 25(2), 197-231.\
-
-So, M. E. P., Lam, K., & Li, W. K. (1998). A stochastic volatility model with Markov switching. Journal of Business & Economic Statistics, 16(2), 244-253.\
-
-Xu, X., & Chen, Y. (2021). Deep Stochastic Volatility Model. arXiv preprint arXiv:2102.12658.\
-
-Zhang, B., Chan, J. C., & Cross, J. L. (2020). Stochastic volatility models with ARMA innovations: An application to G7 inflation forecasts. International Journal of Forecasting, 36(4), 1318-1328.
-
-\chapter*{Appendix}
-\section{The Stochastic Volatility Model}
-
-We decide to organise this Appendix as an R package.
-All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model.\footnote{Actually, the function SVLWfun considers a SV model with an additional hierarchical level, as specified in section XX.}
-\begin{align*}
-    y_t|x_t & \sim N(0,e^{x_t})\\
-    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
-\end{align*}
-All the plots and tables of this work concerning stochastic volatility models use this functions.
-
-\hrule
-\hrule
-\
-\code{SVPFfun(data,N,m0,C0,alpha,beta,tau,r)}
-\
-\hrule
-\textbf{Arguments}
-
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
-\
-\code{N} \ \ number of particles generated at each step
-\
-\code{m0} \ \ central value of the normal prior state distribution
-\
-\code{C0} \ \ variance of the normal prior state distribution
-\
-\code{alpha} \ \ intercept $\alpha$ in the state equation
-\
-\code{beta} \ \ coefficient $\beta$ in the state equation
-\
-\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
-\
-\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
-\hrule
-\hrule
-\
-```{r message=FALSE, warning=FALSE, paged.print=FALSE}
-SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+This does not seem to be the case, even though the use of a proxy of the latent volatility process leaves room for discussion, as such conclusions rely on the assumption that such proxy is indeed a good one.\footnote{Indeed, it can be argued that the issue is almost the opposite one, as in some cases the filters replicate the proxy better when they are further from convergence.
+For example, note how comparing the following table for the SP500 dataset with that of the main text reveals that some filters perform better when N=1000 as compared to N=10000, even though we know that at N=10000 they replicate their target distribution better.
+```{r include=FALSE}
+
+SISfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
@@
 
-    if(ESS<(N/r)){
+    if(ESS<0){
       index<-sample(N,size=N,replace=T,prob=w)
@@
 }
-```
-\
-\hrule
-\hrule
-\
-\code{SVGPFfun(data,N,m0,C0,alpha,beta,tau,r)}
-\
-\hrule
-\textbf{Arguments}
-
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
-\
-\code{N} \ \ number of particles generated at each step
-\
-\code{m0} \ \ central value of the normal prior state distribution
-\
-\code{C0} \ \ variance of the normal prior state distribution
-\
-\code{alpha} \ \ intercept $\alpha$ in the state equation
-\
-\code{beta} \ \ coefficient $\beta$ in the state equation
-\
-\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
-\
-\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
-\hrule
-\hrule
-
-```{r}
-SVGPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+
+# Basis Particle Filter (Resampling at each step)
+#---------------
+BAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
@@
 
-    xprev<-x
-    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS>-1){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+# Bootstrap Particle Filter
+#---------------
+BPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
 
@@
 }
-```
-\
-\hrule
-\hrule
-\
-\code{SVAPFfun(data,N,m0,C0,alpha,beta,tau,r)}
-\
-\hrule
-\textbf{Arguments}
-
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
-\
-\code{N} \ \ number of particles generated at each step
-\
-\code{m0} \ \ central value of the normal prior state distribution
-\
-\code{C0} \ \ variance of the normal prior state distribution
-\
-\code{alpha} \ \ intercept $\alpha$ in the state equation
-\
-\code{beta} \ \ coefficient $\beta$ in the state equation
-\
-\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
-\
-\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
-\hrule
-\hrule
-
-```{r}
-SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+
+#
+#-----------------------
+#Guided Particle Filter Optimal Kernel
+#------------------------------
+GPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+#Auxiliary Particle Filter
+#-------------------------
+APFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
   if(missing(r)){r=2}else{}
@@
 }
-```
-
-\hrule
-\hrule
-\
-\code{SVAPFoptfun(data,N,m0,C0,alpha,beta,tau,r)}
-\
-\hrule
-\textbf{Arguments}
-
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
-\
-\code{N} \ \ number of particles generated at each step
-\
-\code{m0} \ \ central value of the normal prior state distribution
-\
-\code{C0} \ \ variance of the normal prior state distribution
-\
-\code{alpha} \ \ intercept $\alpha$ in the state equation
-\
-\code{beta} \ \ coefficient $\beta$ in the state equation
-\
-\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
-\
-\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
-\hrule
-\hrule
-
-```{r}
-SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
-  if(missing(r)){r=2}else{}
-  xs<-NULL
-  ws<-NULL
-  ess<-NULL
-  x  = rnorm(N,m0,sqrt(C0))
-  w  = rep(1/N,N)
-
-  for(t in 1:length(data)){
-
-    ESS  = 1/sum(w^2)
-
-    if(ESS<(N/r)){
-      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
-      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
-      k   = sample(1:N,size=N,replace=TRUE,prob=weight)
-      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-      w   = rep(1/N,N)
-    }else{
-      weight = rep(1/N,N)
-      k   = sample(1:N,size=N,replace=FALSE,prob=weight)
-      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-      w   <-  w*dnorm(data[t],0,exp(x1/2))*dnorm(x1,alpha+beta*x[k],tau)/
-        dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
-    }
-
-    w   = w/sum(w)
-    x <- x1
-
-    xs = rbind(xs,x)
-    ws = rbind(ws,w)
-    ess =rbind(ess,ESS)
-
-  }
-  return(list(xs=xs,ws=ws,ess=ess))
-}
-
-```
-
-\hrule
-\hrule
-\
-\code{SVLWfun(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)}
-\
-\hrule
-\textbf{Arguments}
-
-\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
-\
-\code{N} \ \ number of particles generated at each step
-\
-\code{m0} \ \ central value of the normal prior state distribution
-\
-\code{C0} \ \ variance of the normal prior state distribution
-\
-\code{ealpha} \ \ expected value of the intercept $\alpha$ in the state equation
-\
-\code{valpha} \ \ variance of the intercept $\alpha$ in the state equation
-\
-\code{ebeta} \ \ expected value of the coefficient $\beta$ in the state equation
-\
-\code{vbeta} \ \ variance of the coefficient $\beta$ in the state equation
-\
-\code{nu} \ \ when divided by 2, first hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
-\
-\code{lambda} \ \ when multiplied by $\nu/2$, second hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
-\
-\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
-\hrule
-\hrule
-
-```{r}
-SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+#Liu West
+#--------
+LWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
 
@@
 }
+#Set hyperparameters
+#-------------------
+#n     =  length(y)
+alpha =  0
+beta  =  0.99
+tau2  =  0.05
+m0      = 0
+C0      = 100 #N(0,100) is an informative prior about the mean of the volatility. It is used by Jacquier et al. (2004), and it is calibrated on percentage returns
+sC0     = sqrt(C0)
+ealpha  = alpha
+valpha  = 0.01
+ebeta    = beta
+vbeta    = 0.01
+nu      = 5
+lambda  = tau2*3/5
+tau   = sqrt(tau2)
+
+#Real dataset
+
+library(openxlsx)
+realdatasetSPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+y<-realdatasetSPX[,2]
+n = length(y)
+volatilitySPX<-read.xlsx("Dataset 17-21.xlsx", sheet = 1)
+x<-volatilitySPX[,3]
+alpha.true = alpha
+beta.true  = beta
+tau2.true  = tau2
+timeframe<-c(1:length(y))
+dfsv<-data.frame(timeframe,y,x)
+
+#Filtering
+#---------
+set.seed(12345)
+N=1000
+svsis<-SISfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svbapf<-BAPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svbpf<-BPFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svapf<-APFfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svgpfopt<-GPFoptfun(y,N,m0,C0,alpha,beta,tau)
+set.seed(12345)
+N=1000
+svlw<-LWfun(y,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)
+
+Filtervalues<-function(fun){
+  xhat<-sapply(1:n,function(i)
+    weighted.mean(fun$xs[i,],fun$ws[i,]))
+  sdhat<-sapply(1:n,function(i)
+    sqrt(weighted.mean((fun$xs[i,]-xhat[i])^2,fun$ws[i,])))
+
+  return(list(mean=xhat,sd=sdhat))
+
+}
+
+## RMSE MAE comparison
+
+realisedx<-x
+ErrorvolSPX<-matrix(NA,ncol=7,nrow=2)
+colnames(ErrorvolSPX)<-c("N","BPF", "GPFOPT", "APF","BAPF","LWF", "SIS")
+rownames(ErrorvolSPX)<-c("RMSE SP500","MAE SP500")
+ErrorvolSPX[,1]<-c(1000)
+RMSE<-function(x,xhat){sqrt(mean((x-xhat)^2))}
+MAE<-function(x,xhat){mean(abs(x-xhat))}
+comparablevol<-function(fun){exp((Filtervalues(fun)$mean)/2)}
+ErrorvolSPX[1,2]<-RMSE(realisedx,comparablevol(svbpf))
+ErrorvolSPX[1,3]<-RMSE(realisedx,comparablevol(svgpfopt))
+ErrorvolSPX[1,4]<-RMSE(realisedx,comparablevol(svapf))
+ErrorvolSPX[1,5]<-RMSE(realisedx,comparablevol(svbapf))
+ErrorvolSPX[1,6]<-RMSE(realisedx,comparablevol(svlw))
+ErrorvolSPX[1,7]<-RMSE(realisedx,comparablevol(svsis))
+ErrorvolSPX[2,2]<-MAE(realisedx,comparablevol(svbpf))
+ErrorvolSPX[2,3]<-MAE(realisedx,comparablevol(svgpfopt))
+ErrorvolSPX[2,4]<-MAE(realisedx,comparablevol(svapf))
+ErrorvolSPX[2,5]<-MAE(realisedx,comparablevol(svbapf))
+ErrorvolSPX[2,6]<-MAE(realisedx,comparablevol(svlw))
+ErrorvolSPX[2,7]<-MAE(realisedx,comparablevol(svsis))
+```
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
+library(kableExtra)
+kbl(ErrorvolSPX, align="c", longtable = T, booktabs = T, caption="RMSE and MAE, realised volatility as benchmark (SPX)",digits=5)
+```
+}\
+Rather, what this exercise highlights (under its assumptions) is a need for more flexibility and variability in the basic SV model: the underlying SV model of the LW filter meets this request, but clearly other solutions are possible, and have been argued for in the field of financial econometrics and mathematical finance, as we noted in the literary review section.
+
+\section{Bibliography}\
+
+Allen, D. E., & McAleer, M. (2020). Do we need stochastic volatility and generalised autoregressive conditional heteroscedasticity? Comparing squared end-of-day returns on FTSE. Risks, 8(1), 12.\
+
+Allen, D. E. (2020). Stochastic Volatility and GARCH: Do Squared End-of-Day Returns Provide Similar Information?. Journal of Risk and Financial Management, 13(9), 202.\
+
+Andrieu, C., Doucet, A., & Holenstein, R. (2010). Particle markov chain monte carlo methods. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 72(3), 269-342.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2001).
+Non‐Gaussian Ornstein--Uhlenbeck‐based models and some of their uses in financial economics.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 63(2), 167-241.\
+
+Barndorff‐Nielsen, O.
+E., & Shephard, N.
+(2002).
+Econometric analysis of realized volatility and its use in estimating stochastic volatility models.
+Journal of the Royal Statistical Society: Series B (Statistical Methodology), 64(2), 253-280.\
+
+Bollerslev, T. (1986). Generalized autoregressive conditional heteroskedasticity. Journal of econometrics, 31(3), 307-327.\
+
+Breidt, F. J., Crato, N., & De Lima, P. (1998). The detection and estimation of long memory in stochastic volatility. Journal of econometrics, 83(1-2), 325-348.\
+
+Buncic, D., & Gisler, K. I. (2016). Global equity market volatility spillovers: A broader role for the United States. International Journal of Forecasting, 32(4), 1317-1339.\
+
+Chan, J. C. (2013). Moving average stochastic volatility models with application to inflation forecast. Journal of Econometrics, 176(2), 162-172.\
+
+Chib, S., Nardari, F., & Shephard, N. (2002). Markov chain Monte Carlo methods for stochastic volatility models. Journal of Econometrics, 108(2), 281-316.\
+
+Chib, S., Nardari, F., & Shephard, N. (2006). Analysis of high dimensional multivariate stochastic volatility models. Journal of Econometrics, 134(2), 341-371.\
+
+Chopin, N., & Papaspiliopoulos, O. (2020). An introduction to sequential Monte Carlo (Vol. 4). Springer.\
+
+Comte, F., & Renault, E. (1998). Long memory in continuous‐time stochastic volatility models. Mathematical finance, 8(4), 291-323.\
+
+Christoffersen, P., Jacobs, K., & Mimouni, K.
+(2010).
+Volatility dynamics for the S&P500: Evidence from realized volatility, daily returns, and option prices.
+The Review of Financial Studies, 23(8), 3141-3189.\
+
+Dimitrakopoulos, S., & Kolossiatis, M.
+(2020).
+Bayesian analysis of moving average stochastic volatility models: modeling in-mean effects and leverage for financial time series.
+Econometric Reviews, 39(4), 319-343.\
+
+Douc, R., Moulines, E., & Stoffer, D. (2014). Nonlinear time series: Theory, methods and applications with R examples. CRC press.\
+
+Engle, R. F. (1982). Autoregressive conditional heteroscedasticity with estimates of the variance of United Kingdom inflation. Econometrica: Journal of the econometric society, 987-1007.\
+
+Eraker, B. (2004). Do stock prices and volatility jump? Reconciling evidence from spot and option prices. The Journal of Finance, 59(3), 1367-1403.\
+
+Friz, P. K., Gassiat, P., & Pigato, P. (2021). Precise asymptotics: robust stochastic volatility models. The Annals of Applied Probability, 31(2), 896-940.\
+
+Gatheral, J., Jaisson, T., & Rosenbaum, M. (2018). Volatility is rough. Quantitative finance, 18(6), 933-949.\
+
+Guidolin, M., & Pedio, M. (2021). Media Attention vs. Sentiment as Drivers of Conditional Volatility Predictions: An Application to Brexit. Finance Research Letters, 101943.\
+
+Hansen, P. R., & Lunde, A. (2006). Consistent ranking of volatility models. Journal of Econometrics, 131(1-2), 97-121.\
+
+Harvey, A. C., & Shephard, N. (1996). Estimation of an asymmetric stochastic volatility model for asset returns. Journal of Business & Economic Statistics, 14(4), 429-434.\
+
+Harvey, A., Ruiz, E., & Shephard, N. (1994). Multivariate stochastic variance models. The Review of Economic Studies, 61(2), 247-264.\
+
+Heber, Gerd, Asger Lunde, Neil Shephard and Kevin Sheppard (2009).
+Oxford-Man Institute's realized library.
+Oxford-Man Institute, University of Oxford, version 0.3.
+Available at: https://realized.oxford-man.ox.ac.uk/.
+\
+
+Heston, S. L. (1993). A closed-form solution for options with stochastic volatility with applications to bond and currency options. The review of financial studies, 6(2), 327-343.\
+
+Hull, J., & White, A. (1987). The pricing of options on assets with stochastic volatilities. The journal of finance, 42(2), 281-300.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. (1994). Bayesian analysis of stochastic volatility models. Journal of Business and Economic Statistics, 12(4), 371-389.\
+
+Jacquier, E., Polson, N. G., & Rossi, P. E. (2004). Bayesian analysis of stochastic volatility models with fat-tails and correlated errors. Journal of Econometrics, 122(1), 185-212.\
+
+Johannes, M., Korteweg, A., & Polson, N. (2014). Sequential learning, predictability, and optimal portfolio returns. The Journal of Finance, 69(2), 611-644.\
+
+Kambouroudis, D.
+S., McMillan, D.
+G., & Tsakou, K.
+(2016).
+Forecasting stock return volatility: a comparison of GARCH, Implied volatility, and realized volatility models.
+Journal of Futures Markets, 36(12), 1127-1163.\
+
+Kastner, G. (2019). Dealing with stochastic volatility in time series using the R package stochvol. arXiv preprint arXiv:1906.12134.\
+
+Kim, S., Shephard, N., & Chib, S. (1998). Stochastic volatility: likelihood inference and comparison with ARCH models. The review of economic studies, 65(3), 361-393.\
+
+Kim, J., & Stoffer, D.
+S.
+(2008).
+Fitting stochastic volatility models in the presence of irregular sampling via particle methods and the EM algorithm.
+Journal of time series analysis, 29(5), 811-833.\
+
+Koopman, S. J., & Hol Uspensky, E. (2002). The stochastic volatility in mean model: empirical evidence from international stock markets. Journal of applied Econometrics, 17(6), 667-689.\
+
+Liesenfeld, R., & Jung, R. C. (2000). Stochastic volatility models: conditional normality versus heavy‐tailed distributions. Journal of applied Econometrics, 15(2), 137-160.\
+
+Luo, R., Zhang, W., Xu, X., & Wang, J. (2018, April). A neural stochastic volatility model. In Thirty-second AAAI conference on artificial intelligence.\
+
+Mao, X., Czellar, V., Ruiz, E., & Veiga, H.
+(2020).
+Asymmetric stochastic volatility models: Properties and particle filter-based simulated maximum likelihood estimation.
+Econometrics and Statistics, 13, 84-105.\
+
+Nakajima, J., & Omori, Y.
+(2012).
+Stochastic volatility model with leverage and asymmetrically heavy-tailed error using GH skew Student's t-distribution.
+Computational Statistics & Data Analysis, 56(11), 3690-3704.\
+
+Omori, Y., Chib, S., Shephard, N., & Nakajima, J. (2007). Stochastic volatility with leverage: Fast and efficient likelihood inference. Journal of Econometrics, 140(2), 425-449.\
+
+Pitt, M. K., & Shephard, N. (1999). Filtering via simulation: Auxiliary particle filters. Journal of the American statistical association, 94(446), 590-599.\
+
+Poon, S. H., & Granger, C. (2005). Practical issues in forecasting volatility. Financial analysts journal, 61(1), 45-56.\
+
+Shephard, N., & Sheppard, K. (2010). Realising the future: forecasting with high‐frequency‐based volatility (HEAVY) models. Journal of Applied Econometrics, 25(2), 197-231.\
+
+So, M. E. P., Lam, K., & Li, W. K. (1998). A stochastic volatility model with Markov switching. Journal of Business & Economic Statistics, 16(2), 244-253.\
+
+Xu, X., & Chen, Y. (2021). Deep Stochastic Volatility Model. arXiv preprint arXiv:2102.12658.\
+
+Zhang, B., Chan, J. C., & Cross, J. L. (2020). Stochastic volatility models with ARMA innovations: An application to G7 inflation forecasts. International Journal of Forecasting, 36(4), 1318-1328.
+
+\chapter*{Appendix}
+\section{The Stochastic Volatility Model}
+
+We decide to organise this Appendix as an R package.
+All the R functions described in this Appendix take as reference a basic specification of Stochastic Volatility Model.\footnote{Actually, the function SVLWfun considers a SV model with an additional hierarchical level, as specified in section XX.}
+\begin{align*}
+    y_t|x_t & \sim N(0,e^{x_t})\\
+    x_t|x_{t-1},\alpha,\beta,\tau^2 & \sim N(\alpha+\beta x_{t-1},\tau^2)
+\end{align*}
+All the plots and tables of this work concerning stochastic volatility models use this functions.
+
+\hrule
+\hrule
+\
+\code{SVPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+\
+```{r message=FALSE, warning=FALSE, paged.print=FALSE}
+SVPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    x<-rnorm(N,alpha+beta*x,tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+\
+\hrule
+\hrule
+\
+\code{SVGPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVGPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    xprev<-x
+    x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+
+    w = w1/sum(w1)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+\
+\hrule
+\hrule
+\
+\code{SVAPFfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVAPFfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    weight = w*dnorm(data[t],0,exp(x/2))
+    k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+    x1   = rnorm(N,alpha+beta*x[k],tau)
+    lw  = dnorm(data[t],0,exp(x1/2),log=TRUE)-dnorm(data[t],0,exp(x[k]/2),log=TRUE)
+    w   = exp(lw)
+    w   = w/sum(w)
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      x<-x1[index]
+      w<-rep(1/N,N)
+    }else{}
+
+    x <- x1
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+```
+
+\hrule
+\hrule
+\
+\code{SVAPFoptfun(data,N,m0,C0,alpha,beta,tau,r)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{alpha} \ \ intercept $\alpha$ in the state equation
+\
+\code{beta} \ \ coefficient $\beta$ in the state equation
+\
+\code{tau} \ \ square root of $\tau^2$, the conditional variance of the state
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVAPFoptfun<-function(data,N,m0,C0,alpha,beta,tau,r){
+  if(missing(r)){r=2}else{}
+  xs<-NULL
+  ws<-NULL
+  ess<-NULL
+  x  = rnorm(N,m0,sqrt(C0))
+  w  = rep(1/N,N)
+
+  for(t in 1:length(data)){
+
+    ESS  = 1/sum(w^2)
+
+    if(ESS<(N/r)){
+      epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
+      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+      k   = sample(1:N,size=N,replace=TRUE,prob=weight)
+      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+      w   = rep(1/N,N)
+    }else{
+      weight = rep(1/N,N)
+      k   = sample(1:N,size=N,replace=FALSE,prob=weight)
+      x1   = rnorm(N,alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+      w   <-  w*dnorm(data[t],0,exp(x1/2))*dnorm(x1,alpha+beta*x[k],tau)/
+        dnorm(x1,mean=alpha+beta*x[k]+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x[k]))-2),tau)
+    }
+
+    w   = w/sum(w)
+    x <- x1
+
+    xs = rbind(xs,x)
+    ws = rbind(ws,w)
+    ess =rbind(ess,ESS)
+
+  }
+  return(list(xs=xs,ws=ws,ess=ess))
+}
+
+```
+
+\hrule
+\hrule
+\
+\code{SVLWfun(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda)}
+\
+\hrule
+\textbf{Arguments}
+
+\code{data} \ \ the observed process. It has to be a vector or a univariate time series.
+\
+\code{N} \ \ number of particles generated at each step
+\
+\code{m0} \ \ central value of the normal prior state distribution
+\
+\code{C0} \ \ variance of the normal prior state distribution
+\
+\code{ealpha} \ \ expected value of the intercept $\alpha$ in the state equation
+\
+\code{valpha} \ \ variance of the intercept $\alpha$ in the state equation
+\
+\code{ebeta} \ \ expected value of the coefficient $\beta$ in the state equation
+\
+\code{vbeta} \ \ variance of the coefficient $\beta$ in the state equation
+\
+\code{nu} \ \ when divided by 2, first hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
+\
+\code{lambda} \ \ when multiplied by $\nu/2$, second hyperparameter of the Inverse Gamma distribution on $\tau^2$, the conditional variance of the state
+\
+\code{r} \ \  if present the threshold is set equal to $N/r$ otherwise, if missing, the threshold is set equal to $N/2$
+\hrule
+\hrule
+
+```{r}
+SVLWfun<-function(data,N,m0,C0,ealpha,valpha,ebeta,vbeta,nu,lambda){
+
+  xs = rnorm(N,m0,sqrt(C0))
+  pars   = cbind(rnorm(N,ealpha,sqrt(valpha)),rnorm(N,ebeta,sqrt(vbeta)),
+                 log(1/rgamma(N,nu/2,nu*lambda/2)))
+  delta  = 0.75
+  a      = (3*delta-1)/(2*delta)
+  h2     = 1-a^2
+  parss  = array(0,c(N,3,n))
+  xss    = NULL
+  ws     = NULL
+  ESS    = NULL
+  w      = rep(1/N,N)
+  vpars   = NULL
+  for (t in 1:length(data)){
+
+    #mpar        = apply(pars,2,mean) #alternatively use this 
+    mpar<-c()
+    for(i in 1:3){
+      mpar[i]    = weighted.mean(pars[,i],w)
+    }
+
+    vpar     = var(pars)
+    ms          = a*pars+(1-a)*matrix(mpar,N,3,byrow=T)
+    mus         = pars[,1]+pars[,2]*xs 
+    weight      = w*dnorm(data[t],0,exp(mus/2))
+    k           = sample(1:N,size=N,replace=T,prob=weight)
+    ms1         = ms[k,] + matrix(rnorm(3*N),N,3)%*%chol(h2*vpar)
+    xt          = rnorm(N,ms1[,1]+ms1[,2]*xs[k],exp(ms1[,3]/2))
+    w           = dnorm(data[t],0,exp(xt/2))/dnorm(data[t],0,exp(mus[k]/2))
+    w           = w/sum(w)
+    ESS         = 1/sum(w^2)
+
+    if(ESS<(N/2)){
+      index<-sample(N,size=N,replace=T,prob=w)
+      xs<-xt[index]
+      pars<-ms1[index,]
+      w<-rep(1/N,N)
+    }else{
+      xs<-xt
+      pars<-ms1
+    }
+
+    vpars        = rbind(vpars,vpar)
+    xss         = rbind(xss,xs)
+    parss[,,t]  = pars 
+    ws          = rbind(ws,w)
+  }
+  return(list(xs=xss,pars=parss,ws=ws,vpars=vpars))
+}
 ```

### 2021-08-10 Real Dataset SV.Rmd
Aggiornate funzioni grafici

@@
 
-Filtplot1<-function(dataframe,title){
-  mean=sv_bpf1
+Filtplot1<-function(dataframe,title,fun){
+  mean=fun
   mean1=exp(mean/2)
@@
 }
-Filtplot2<-function(dataframe,title){
-  mean=sv_gpfopt1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot3<-function(dataframe,title){
-  mean=sv_apf1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot4<-function(dataframe,title){
-  mean=sv_lw1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot5<-function(dataframe,title){
-  mean=sv_bapf1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot6<-function(dataframe,title){
-  mean=sv_sis1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y1, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotobs1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
-plotobs2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
-plotobs3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
-plotobs4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
-plotobs5<-Filtplot5(dfsv5,"Always Resampling SP500")
-plotobs6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+plotobs1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500",sv_bpf1)
+plotobs2<-Filtplot1(dfsv2,"Opt Ker Guided Particle Filter SP500",sv_gpfopt1)
+plotobs3<-Filtplot1(dfsv3,"Auxiliary Particle Filter SP500",sv_apf1)
+plotobs4<-Filtplot1(dfsv4,"Liu and West Filter SP500",sv_lw1)
+plotobs5<-Filtplot1(dfsv5,"Always Resampling SP500",sv_bapf1)
+plotobs6<-Filtplot1(dfsv6,"No Resampling SP500",sv_sis1)
 
@@
 
-Filtplot7<-function(dataframe,title){
-  mean=sv_bpf2
+Filtplot2<-function(dataframe,title,fun){
+  mean=fun
   mean1=exp(mean/2)
@@
 }
-Filtplot8<-function(dataframe,title){
-  mean=sv_gpfopt2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot9<-function(dataframe,title){
-  mean=sv_apf2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot10<-function(dataframe,title){
-  mean=sv_lw2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot11<-function(dataframe,title){
-  mean=sv_bapf2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot12<-function(dataframe,title){
-  mean=sv_sis2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y2, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotobs7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX50E")
-plotobs8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX50E")
-plotobs9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX50E")
-plotobs10<-Filtplot10(dfsv10,"Liu and West Filter STOXX50E")
-plotobs11<-Filtplot11(dfsv11,"Always Resampling STOXX50E")
-plotobs12<-Filtplot12(dfsv12,"No Resampling STOXX50E")
+
+plotobs7<-Filtplot2(dfsv7,"Bootstrap Particle Filter STOXX50E",sv_bpf2)
+plotobs8<-Filtplot2(dfsv8,"Opt Ker Guided Particle Filter STOXX50E",sv_gpfopt2)
+plotobs9<-Filtplot2(dfsv9,"Auxiliary Particle Filter STOXX50E",sv_apf2)
+plotobs10<-Filtplot2(dfsv10,"Liu and West Filter STOXX50E",sv_lw2)
+plotobs11<-Filtplot2(dfsv11,"Always Resampling STOXX50E",sv_bapf2)
+plotobs12<-Filtplot2(dfsv12,"No Resampling STOXX50E",sv_sis2)
 
@@
 
-Filtplot13<-function(dataframe,title){
-  mean=sv_bpf3
+Filtplot3<-function(dataframe,title,fun){
+  mean=fun
   mean1=exp(mean/2)
@@
 }
-Filtplot14<-function(dataframe,title){
-  mean=sv_gpfopt3
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot15<-function(dataframe,title){
-  mean=sv_apf3
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot16<-function(dataframe,title){
-  mean=sv_lw3
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot17<-function(dataframe,title){
-  mean=sv_bapf3
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot18<-function(dataframe,title){
-  mean=sv_sis3
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=y3, col="Observations", linetype="Observations"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Observations" = "gray",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Observations" = 1,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotobs13<-Filtplot13(dfsv13,"Bootstrap Particle Filter DJI")
-plotobs14<-Filtplot14(dfsv14,"Opt Ker Guided Particle Filter DJI")
-plotobs15<-Filtplot15(dfsv15,"Auxiliary Particle Filter DJI")
-plotobs16<-Filtplot16(dfsv16,"Liu and West Filter DJI")
-plotobs17<-Filtplot17(dfsv17,"Always Resampling DJI")
-plotobs18<-Filtplot18(dfsv18,"No Resampling DJI")
+
+plotobs13<-Filtplot3(dfsv13,"Bootstrap Particle Filter DJI",sv_bpf3)
+plotobs14<-Filtplot3(dfsv14,"Opt Ker Guided Particle Filter DJI",sv_gpfopt3)
+plotobs15<-Filtplot3(dfsv15,"Auxiliary Particle Filter DJI",sv_apf3)
+plotobs16<-Filtplot3(dfsv16,"Liu and West Filter DJI",sv_lw3)
+plotobs17<-Filtplot3(dfsv17,"Always Resampling DJI",sv_bapf3)
+plotobs18<-Filtplot3(dfsv18,"No Resampling DJI",sv_sis3)
 
@@
 
-Filtplot1<-function(dataframe,title){
-  mean=sv_bpf1
-  sd=sv_bpf1sd
+Filtplot4<-function(dataframe,title,funmean,funsd){
+  mean=funmean
+  sd=funsd
   dataframe<-data.frame(dataframe,mean)
@@
 }
-Filtplot2<-function(dataframe,title){
-  mean=sv_gpfopt1
-  sd=sv_gpfopt1sd
-  dataframe<-data.frame(dataframe,mean)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
-    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
-    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
-                fill="red",alpha=0.16) +
-    scale_color_manual("",
-                       values=c("BT benchmark" = "black",
-                                "Filtered States"= "red"))+
-    scale_linetype_manual("",
-                          values=c("BT benchmark" = 1,
-                                   "Filtered States"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot3<-function(dataframe,title){
-  mean=sv_apf1
-  sd=sv_apf1sd
-  dataframe<-data.frame(dataframe,mean)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
-    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
-    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
-                fill="red",alpha=0.16) +
-    scale_color_manual("",
-                       values=c("BT benchmark" = "black",
-                                "Filtered States"= "red"))+
-    scale_linetype_manual("",
-                          values=c("BT benchmark" = 1,
-                                   "Filtered States"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-Filtplot5<-function(dataframe,title){
-  mean=sv_bapf1
-  sd=sv_bapf1sd
-  dataframe<-data.frame(dataframe,mean)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
-    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
-    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
-                fill="red",alpha=0.16) +
-    scale_color_manual("",
-                       values=c("BT benchmark" = "black",
-                                "Filtered States"= "red"))+
-    scale_linetype_manual("",
-                          values=c("BT benchmark" = 1,
-                                   "Filtered States"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot6<-function(dataframe,title){
-  mean=sv_sis1
-  sd=sv_sis1sd
-  dataframe<-data.frame(dataframe,mean)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=btbenchmark, col="BT benchmark", linetype="BT benchmark"))+
-    geom_line(aes(y=mean, col="Filtered States", linetype="Filtered States"))+
-    geom_ribbon(aes(ymin = mean -1.96*sd, ymax = mean +1.96*sd),
-                fill="red",alpha=0.16) +
-    scale_color_manual("",
-                       values=c("BT benchmark" = "black",
-                                "Filtered States"= "red"))+
-    scale_linetype_manual("",
-                          values=c("BT benchmark" = 1,
-                                   "Filtered States"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotcomp1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
-plotcomp2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
-plotcomp3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
-plotcomp5<-Filtplot5(dfsv5,"Always Resampling SP500")
-plotcomp6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+plotcomp1<-Filtplot4(dfsv1,"Bootstrap Particle Filter SP500",sv_bpf1,sv_bpf1sd)
+plotcomp2<-Filtplot4(dfsv2,"Opt Ker Guided Particle Filter SP500",sv_gpfopt1,sv_gpfopt1sd)
+plotcomp3<-Filtplot4(dfsv3,"Auxiliary Particle Filter SP500",sv_apf1,sv_apf1sd)
+plotcomp5<-Filtplot4(dfsv5,"Always Resampling SP500",sv_bapf1,sv_bapf1sd)
+plotcomp6<-Filtplot4(dfsv6,"No Resampling SP500",sv_sis1,sv_sis1sd)
 
@@
 
-Filtplot1<-function(dataframe,title){
-  mean=sv_bpf1
+Filtplot5<-function(dataframe,title,fun,realisedx){
+  mean=fun
   mean1=exp(mean/2)
@@
   ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
+    geom_line(aes(y=realisedx, col="Proxy", linetype="Proxy"))+
     geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
@@
 }
-Filtplot2<-function(dataframe,title){
-  mean=sv_gpfopt1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot3<-function(dataframe,title){
-  mean=sv_apf1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot4<-function(dataframe,title){
-  mean=sv_lw1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot5<-function(dataframe,title){
-  mean=sv_bapf1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot6<-function(dataframe,title){
-  mean=sv_sis1
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx1, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotproxy1<-Filtplot1(dfsv1,"Bootstrap Particle Filter SP500")
-plotproxy2<-Filtplot2(dfsv2,"Opt Ker Guided Particle Filter SP500")
-plotproxy3<-Filtplot3(dfsv3,"Auxiliary Particle Filter SP500")
-plotproxy4<-Filtplot4(dfsv4,"Liu and West Filter SP500")
-plotproxy5<-Filtplot5(dfsv5,"Always Resampling SP500")
-plotproxy6<-Filtplot6(dfsv6,"No Resampling SP500")
+
+plotproxy1<-Filtplot5(dfsv1,"Bootstrap Particle Filter SP500",sv_bpf1,realisedx1)
+plotproxy2<-Filtplot5(dfsv2,"Opt Ker Guided Particle Filter SP500",sv_gpfopt1,realisedx1)
+plotproxy3<-Filtplot5(dfsv3,"Auxiliary Particle Filter SP500",sv_apf1,realisedx1)
+plotproxy4<-Filtplot5(dfsv4,"Liu and West Filter SP500",sv_lw1,realisedx1)
+plotproxy5<-Filtplot5(dfsv5,"Always Resampling SP500",sv_bapf1,realisedx1)
+plotproxy6<-Filtplot5(dfsv6,"No Resampling SP500",sv_sis1,realisedx1)
 
@@
 
-Filtplot7<-function(dataframe,title){
-  mean=sv_bpf2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot8<-function(dataframe,title){
-  mean=sv_gpfopt2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot9<-function(dataframe,title){
-  mean=sv_apf2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot10<-function(dataframe,title){
-  mean=sv_lw2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot11<-function(dataframe,title){
-  mean=sv_bapf2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-Filtplot12<-function(dataframe,title){
-  mean=sv_sis2
-  mean1=exp(mean/2)
-  dataframe<-data.frame(dataframe,mean1)
-
-  ggplot(dataframe,aes(x=timeframe))+
-    geom_line(aes(y=realisedx2, col="Proxy", linetype="Proxy"))+
-    geom_line(aes(y=mean1, col="Filtered Volatility", linetype="Filtered Volatility"))+
-    scale_color_manual("",
-                       values=c("Proxy" = "black",
-                                "Filtered Volatility"= "red"))+
-    scale_linetype_manual("",
-                          values=c("Proxy" = 3,
-                                   "Filtered Volatility"=1))+
-    labs(x="Time",
-         y="")+
-    ggtitle(title)+
-    theme_bw()+
-    theme(legend.direction = "horizontal", legend.position = "bottom", legend.key = element_blank(), 
-          legend.background = element_rect(fill = "white", colour = "gray30")) +
-    theme(plot.title = element_text(hjust = 0.5))
-}
-
-plotproxy7<-Filtplot7(dfsv7,"Bootstrap Particle Filter STOXX")
-plotproxy8<-Filtplot8(dfsv8,"Opt Ker Guided Particle Filter STOXX")
-plotproxy9<-Filtplot9(dfsv9,"Auxiliary Particle Filter STOXX")
-plotproxy10<-Filtplot10(dfsv10,"Liu and West Filter STOXX")
-plotproxy11<-Filtplot11(dfsv11,"Always Resampling STOXX")
-plotproxy12<-Filtplot12(dfsv12,"No Resampling STOXX")
+plotproxy7<-Filtplot5(dfsv7,"Bootstrap Particle Filter STOXX",sv_bpf2,realisedx2)
+plotproxy8<-Filtplot5(dfsv8,"Opt Ker Guided Particle Filter STOXX",sv_gpfopt2,realisedx2)
+plotproxy9<-Filtplot5(dfsv9,"Auxiliary Particle Filter STOXX",sv_apf2,realisedx2)
+plotproxy10<-Filtplot5(dfsv10,"Liu and West Filter STOXX",sv_lw2,realisedx2)
+plotproxy11<-Filtplot5(dfsv11,"Always Resampling STOXX",sv_bapf2,realisedx2)
+plotproxy12<-Filtplot5(dfsv12,"No Resampling STOXX",sv_sis2,realisedx2)
 

### 2021-08-11 Real Dataset SV.Rmd
Riferimenti intratesto

@@
 Intuitively, such differences are due to the fact that now the latent process driving the Hidden Markov Model is not observed, nor do we know exactly its law of motion.\
-This part of the analysis is structured as follows: section \textit{fill with the number of the section} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \textit{fill with the number of the section} describes the data which we employ and the transformations that were applied to them.
-In section \textit{fill with the number of the section} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
-Section \textit{fill with the number of the section} features a forecast comparison between two SV specifications and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
-In section \textit{fill with the number of the section} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \textit{fill with the number of the section} for synthetic data.
-
-\section{A review of Stochastic Volatility models}
-
-In section \textit{fill with the number of the section} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
+This part of the analysis is structured as follows: section \ref{Literary review} provides a review of recent research in the stochastic volatility literature, including extensions of the basic specification and the use of particle filters; section \ref{Data description} describes the data which we employ and the transformations that were applied to them.
+In section \ref{Comparing the filters} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
+Section \ref{Forecast comparison} features a forecast comparison between a SV specification and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
+In section \ref{Realised volatility} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \ref{Simulated SV data} for synthetic data.
+
+\section{A review of Stochastic Volatility models}\label{Literary review}
+
+In section \ref{Simulated SV data} a simple specification of a Stochastic Volatility model was presented and used as the data generating process of a simulated dataset.
 Denoting by $y_t$ the log returns at time $t$ and by $x_t$ the latent volatility stochastic process, such state-space model was presented as follows,
@@
 
-\section{Data Description}
+\section{Data Description}\label{Data description}
 
@@
 
-As regards the series that we will employ in section \textit{fill with the number of the section}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \textit{fill with the number of the section}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
+As regards the series that we will employ in section \ref{Realised volatility}, namely the series of realised volatilities which we will employ as a proxy for the true latent volatilities, the data were retrieved from the Oxford-Man Institute's Realized Library, which provides several different daily non-parametric measures of past volatility.\footnote{Such class of measures will be better presented in section \ref{Realised volatility}, when reviewing the use of realised volatility estimators as proxies for latent volatilities.} We consider a proper rescaling of the \textit{rv5} series, \textit{i.e.} model-free daily volatility estimates based on 5 min intraday return intervals.
 Also in this case, the chosen interval spanned from June 1st, 2017 to May 30th, 2021.
@@
 
-\section{Comparing the filters}
+\section{Comparing the filters}\label{Comparing the filters}
 
@@
 \end{align*}
-where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part, see section \textit{fill with the number of the section}}.\
+where $\gamma$,$\zeta$,$\pi$,$\phi$,$\nu$ and $\lambda$ are assumed to be known.\footnote{In these empirical applications we keep the values of the parameter we employed in the synthetic data part.}.
+The initial state is drawn from a normal distribution with commonly used parameters for percentage log returns, as reported by Kastner (2019).\
+For simplicity, in $$y_t|x_t \sim N(0,e^{x_t})$$ $$x_t|x_{t-1},\alpha,\beta,\tau^2 \sim N(\alpha+\beta x_{t-1},\tau^2)$$ we set $\alpha=0$, a value that in general, as we argued earlier, also has an empirical justification for percentage log returns.
+We choose $\beta=0.99$, as one of the basic underlying ideas of SV models is considerable persistence in the volatility.
+Finally, we set $\tau^2=0.05$.\
+In the specification used by the LW filter, we have $\gamma=0$ and $\pi=0.99$, as to center the priors on the point guesses of the simpler model; $\zeta$ and $\phi$ are both set to $0.01$, then allowing a 0.1 standard deviation on both $\alpha$ and $\gamma$.
+For $\tau^2$, we choose an Inverse Gamma distribution whose expected value coincides with the point guess $0.05$, by setting $\nu=5$ and $\lambda=0.05*3/5$.\
 
@@
 We start by running the Bootstrap particle filter with N=50000, setting it as the benchmark.
-In table XX we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles.
+In the table below we report the RMSE and MAE measures one obtains by comparing the mean of the particles generated at each step by the Bootstrap PF with N=50000 and the mean\footnote{As a possible limitation of our exercise, here we do not consider in detail the variance of such generated particles.
 Nevertheless, one can get at least an idea of the differences in the generated variances by comparing the confidence intervals that will be shown in the graphs below.
@@
 Its performance ameliorates dramatically at N=1000, while it still shows some sizable improvements even when passing from N=5000 to N=10000, almost halving its errors.
-Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N$>>10000$.\
+Moreover, note that the mere fact that at N=10000 the Guided and the Bootstrap particle filters approximate better the APF benchmark suggests that the APF still has some margins for relative improvement even at high levels of N, pointing at higher computational costs than the competing filters: trivially, since at N=50000 its RMS and MA difference with respect to this benchmark must be zero, the APF overtakes the Guided and the Bootstrap PFs at N $>>10000$.\
 
@@
 
-\section{Realised volatility as benchmark}
-
-In section XX we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
+\section{Realised volatility as benchmark}\label{Realised volatility}
+
+In section \ref{Comparing the filters} we compared several filtered distributions, briefly pointing out three main behaviours, namely that of the SIS algorithm, that of the Liu and West particle filter and that of the remaining filters.
 However, we did not have metrics that could help us clearly point at one behaviour as the closest to reality.
@@
 In order to evaluate the different filters according to the evaluation tool represented by the realized volatility, we start once more from the visual inspection of the plots.
-For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \textit{fill with the number of the section}, we can now use such a proxy to determine the relative performance of the competing filters.\
+For any of the three indices, below graphs are shown including both the filtered volatility obtained from the filtered states (for any of the Sequential MC methods described in the theoretical section) and the corresponding rescaled realised volatility series.\footnote{For simplicity,troughout the rest of this section we will omit "rescaled" when refering to the rescaled realised volatility.} As compared to the visual analysis we conducted in section \ref{Comparing the filters}, we can now use such a proxy to determine the relative performance of the competing filters.\
 
@@
 
-The above intuitions are confirmed by looking at Table XX, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark.
+The above intuitions are confirmed by looking at the table below, where the RMSE and MAE are shown for each filter (run with a number of particles equal to 10000), considering the realised volatility as benchmark.
 Indeed, for any index, most of the filters give rather similar results in terms of RMSE and MAE.
@@
 
-Finally, we can see from Table XX the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices.
-We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section XX; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
+Finally, we can see from the table below the RMSE and MAE obtained with a number of particles equal to 50000, for the SP500 and the EuroStoxx50 indices.
+We consider the Bootstrap particle filter and the Auxiliary particle filter, which we employed in section \ref{Comparing the filters}; in addition to these, we add the Liu and West filter, which did not feature in that section of this work.\
 In the case of the SP500 index, although the number of particles increases, both RMSE and MAE increase (although by very small amounts) for the Auxiliary Particle Filter, while they slightly decrease for the Bootstrap PF.

### 2021-08-11 prova_knit_finale.Rmd
Riferimenti intratesto

@@
 
-\section{Implementations for Stochastic Volatility Models}
+\section{Implementations for Stochastic Volatility Models}\label{Simulated SV data}
 
@@
 
-\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}
+\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}\label{Forecast comparison}
 

### 2021-08-11 Real Dataset SV.Rmd
commit

@@
 Once more, most algorithms used (aside from SIS and Liu-West) seem to provide indistinguishable results.
-The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section XX, the degeneracy of its weights on the particles prevents it from delivering a good enough approximation of the target distribution (\textit{check whether this is the correct intuition}).
+The SIS algorithm appears completely inadequate to describe the "true" volatility process: as we argued in section \ref{pf_intro}, the diverging variance of its weights on the particles prevents it from delivering a good enough approximation of the target distribution.
 On top of that, it does not seem the case that the SIS algorithm, by falling short of the SV target distribution, ends up fitting the proxy better.\footnote{As we will now see, this seems instead to be the case for our preferred algorithms, namely those adopting adaptive resampling rules: a better approximation of the SV target filtering distribution yields slightly higher RMSEs and MAEs than those resulting from less accurate MC approximations.}\

### 2021-08-11 prova_knit_finale.Rmd
commit

@@
 
-```{r echo=FALSE, fig.align='center', fig.cap="Observations and filtered volatility, 3 indices", fig.height=4, fig.pos='H', fig.width=8, message=FALSE, warning=FALSE, results=F}
+```{r echo=FALSE, fig.align='center', fig.cap="Observations and filtered volatility, 3 indices", fig.height=7, fig.pos='H', fig.width=10, message=FALSE, warning=FALSE, results=F}
 
@@
 ggarrange(plotobs7,plotobs8)
-ggarrange(plotobs9,plotobs10)
-ggarrange(plotobs11,plotobs12)
+ggarrange(plotobs9,plotobs11)
+plotobs10
+plotobs12
 

### 2021-08-11 prova_knit_finale.Rmd
correzioni

@@
 \begin{itemize}
-    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y})$, and 
-    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X})),$
+    \item for all $x\in\mathcal{X}, P(x,\cdot)$ is a probability measure on $(\mathcal{Y}, \mathcal{B}(\mathcal{Y}))$, and 
+    \item for all subsets $A\in\mathcal{B}(\mathcal{Y}),$ the map $x\mapsto P(x,A)$ is measurable in $\mathcal{B}(\mathcal{X}),$
 \end{itemize}
@@
 Consider the initial distribution $\P_0(dx_0)$ and the probability kernels
-$P_t(x_{t-1},dx_t)$ and $F_t(x_t,d_t)$, where $t=1,2,\dots$.
+$P_t(x_{t-1},dx_t)$ and $F_t(x_t,dy_t)$, where $t=1,2,\dots$.
 \

### 2021-08-11 prova_knit_finale.Rmd
typo

--- 2021-08-11 prova_knit_finale.Rmd
@@
 \begin{lemma}\label{kalman-filter}
-  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,\Lambda)$, where $R$ and $Q_0$ are positive semi-definite matrices. 
+  Let $X\thicksim\N(m_0,C_0)$ and $Y\mid X\thicksim\N(BX,\Lambda)$, where $C_0$ and $\Lambda$ are positive semi-definite matrices. 
   Then, it holds that

### 2021-08-11 prova_knit_finale.Rmd
typo

@@
 \begin{align*}
-  \P_{t}(X_t\in dx_{t}\mid Y_{1:t-1}&=y_{1:t-1})=\N(m_t,C_t), \text{ where} \\ C_t&=R_t[I_{d_x} - B_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_tR_t] \\
+  \P_{t}(X_t\in dx_{t}\mid Y_{1:t}&=y_{1:t})=\N(m_t,C_t), \text{ where} \\ C_t&=R_t[I_{d_x} - B_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_tR_t] \\
   m_t &= [I_{d_x} - R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1} B_t]A_tm_{t-1}+R_tB_t'(B_tR_tB_t'+\Lambda_t)^{-1}y_t.

### 2021-08-11 prova_knit_finale.Rmd
bibliografia corretta
biblio
piccola modifica
cambia
nomi

@@
 author:
+  - Michele Bolla
+  - Benedetta Bruni
+  - Roberto Colarieti
   - Edoardo Marcelli
-  - Roberto Colarieti
+  - Gabriele Romano
   - Alexei Verkhovtsev
-  - Michele Bolla
-  - Gabriele Romano
-  - Benedetta Bruni
 
@@
 So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$.
-These results are of practical use because they provide a way to explicitly compute the variance of the PF estimates.
+These results are of practical use because they provide a way to explicitly compute the asymptotic variance of the PF estimates.
 
@@
 
-\section{Bibliography}\
+\chapter*{Bibliography}\
 
@@
 \chapter*{Appendix}
-\section{The Stochastic Volatility Model}
+\section*{The Stochastic Volatility Model}
 

### 2021-08-11 prova_knit_finale.Rmd
Model comparison: changes and move to end Other changes

@@
 
-```{r eval=FALSE, include=FALSE}
+```{r include=FALSE}
 truex<-x
@@
 
-```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, results="asis"}
+```{r echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
 kbl(estimatedx, align="c", longtable = T, booktabs = T, caption="RMSE", digits=4)
@@
 Since we do not have access to the exact filtering distribution of the SV model, when assessing how rapidly different particle filters converge to it, we actually need to approximate even such target distribution as to have a benchmark.\
-Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.\footnote{Riferimento alla sezione di convergenza?.} Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
+Fortunately, convergence results about particle filters ensure that when the number of particles diverges, the approximation converges to the target distribution.
+Thus, we run one of the particle filters with a very large number of particles, and assume that such approximation is close enough to the target to be used itself as the distribution the filters should tend to.\
 With such a benchmark, we can then compare how different filters approach it as the number of particles increases.
@@
 Even in light of these markedly more computationally expensive benchmarks, we then choose N=10000 as the number of particles which we routinely employ in this empirical application.
-
-\section{Comparing Constant and Stochastic Volatility Models of Equity Returns}\label{Forecast comparison}
-
-In this section, we apply the particle filter in a model-comparison
-exercise, ultimately underlining the importance of taking stochastic
-volatility into account in modeling equity returns. We follow XX, comparing density forecasts through relative predictive and cumulative likelihoods.
-
-We analyze a dataset of log close-to-close returns of the S&P500, from 2017 to 2021. Consider a stochastic volatility model
-$$
-\mathcal{M}_{SV}:\begin{array}{lc}
-r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
-\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
-\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
-$$
-
-and a constant volatility iid model 
-$$
-\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
-$$
-
-We calibrate the model parameters using the whole sample, and proceed
-by setting the return mean of both models to equal the sample mean,
-$\sigma^{2}$ equal to the return variance, and the parameters of
-the volatility equation in $\mathcal{M}_{SV}$ by running an $AR(1)$
-regression of the log squared residuals.
-
-In order to compare the predictive performance of the two, we compare
-the cumulative log likelihood of the observed returns up to time $t$
-given the considered model. In formulas, the cumulative likelihood
-can be computed as
-$$
-p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
-$$
-
-adopting the convention $y_{1:0}=\emptyset$. This is trivial to obtain
-in the iid case; considering the stohastic volatility model, the predictive
-likelihoods can be derived as
-
-$$
-p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
-$$
-$$
-=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
-$$
-
-Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
-is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
-with weighted particles obtained with particle filtering.
-
-Figure XX reports the time series of the cumulative log-likelihood
-given the constant volatility model $\mathcal{M}_{CV}$ relative to
-the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
-Notice that the first difference of this series is the relative log-likelihood
-of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.
-
-```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood", fig.height=3, fig.pos='H', fig.width=10, results=F}
-load("SV/cll_plot.Rda")
-cll_plot+
-  theme_bw()+
-  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
-  labs(x="Time (month/year)",y="")
-```
-Overall, this analysis shows how a model with stochastic
-volatility is better at describing and predicting the considered time-frame of S&P500 returns. As can be seen from the graph and perhaps unsurprisingly,
-the density forecast given the stochastic volatility model represents
-a particularly significant improvement in the period of extreme values of returns in March 2020.
 
@@
 
+\section{Comparing Constant and Stochastic Volatility Models of Equity Returns Using the BPF}\label{Forecast comparison}
+
+In this section, we round up the appplications chapter using the particle filter in a simple model-comparison
+exercise, underlining the importance of taking stochastic
+volatility into account in modeling equity returns. We follow Johannes, Korteweg and Polson (2014), comparing density forecasts through relative predictive and cumulative likelihoods.
+
+We analyze the previously mentioned dataset of log close-to-close returns of the S&P500, from 2017 to 2021. Consider a stochastic volatility model
+$$
+\mathcal{M}_{SV}:\begin{array}{lc}
+r_{t}=\alpha_{r}+\sigma_{r,t}\cdot v_{t}\\
+\log\sigma_{r,t}^{2}=\alpha_{\sigma}+\beta_{\sigma}\cdot\log\sigma_{r,t-1}^{2}+\sigma_{\sigma}\cdot w_{t}
+\end{array}\ (v_{t},w_{t})\sim\mathcal{N}_{2}(\boldsymbol{0},I_{2})
+$$
+
+and a constant volatility iid model 
+$$
+\mathcal{M}_{CV}:(r_{t})\overset{iid}{\sim}\mathcal{N}(\alpha,\sigma^{2}).
+$$
+
+We calibrate the model parameters using the whole sample, and proceed
+by setting the return mean of both models to equal the sample mean,
+$\sigma^{2}$ equal to the return variance, and the parameters of
+the volatility equation in $\mathcal{M}_{SV}$ by running an $AR(1)$
+regression of the log squared residuals.
+
+In order to compare the predictive performance of the two, we compare
+the cumulative log likelihood of the observed returns up to time $t$
+given the considered model. In formulas, the cumulative likelihood
+can be computed as
+$$
+p(y_{1:t}|\mathcal{M}_{i})=\prod_{s=0}^{t-1}p(y_{s+1}|y_{1:s},\mathcal{M}_{i}),i\in\{SV,CV\},
+$$
+
+adopting the convention $y_{1:0}=\emptyset$. This is trivial to obtain
+in the iid case; considering the stohastic volatility model, the predictive
+likelihoods can be derived as
+
+$$
+p(y_{t+1}|y_{1:t},\mathcal{M}_{SV})=\int p(y_{t+1}|\sigma_{r,t+1},y_{1:t},\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
+$$
+$$
+=\int p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})d(\sigma_{r,t+1})
+$$
+
+Whereas the emission distribution $p(y_{t+1}|\sigma_{r,t+1}\mathcal{M}_{SV})$
+is known, we approximate the predictive state distribution $p(\sigma_{r,t+1}|y_{1:t},\mathcal{M}_{SV})$
+with weighted particles obtained with a Bootstrap Particle Filter.
+
+Figure \ref{fig:cll_plot} reports the time series of the cumulative log-likelihood
+given the constant volatility model $\mathcal{M}_{CV}$ relative to
+the one given $\mathcal{M}_{SV}$, obtained as $\log p(y_{1:t}|\mathcal{M}_{CV})-\log p(y_{1:t}|\mathcal{M}_{SV})$.
+Notice that the first difference of this series is the relative log-likelihood
+of the incoming observation, $p(y_{t}|y_{1:t-1},\mathcal{M}_{CV})-p(y_{t}|y_{1:t-1},\mathcal{M}_{SV})$.
+
+```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood of S&P500 daily log returns, CV and SV", fig.height=3, fig.pos='H', fig.width=10, results=F}
+load("SV/cll_plot.Rda")
+cll_plot+
+  theme_bw()+
+  scale_x_date(date_breaks = "3 months",date_minor_breaks="1 month",date_labels = "%m/%y")+
+  labs(x="Time (month/year)",y="")
+```
+Overall, this analysis shows how a model with stochastic
+volatility is better at describing and predicting the considered time-frame of S&P500 returns. As can be seen from the graph and perhaps unsurprisingly,
+the density forecast given the stochastic volatility model represents
+a particularly significant improvement in the period of extreme values of returns in March 2020.
+
 \chapter*{Bibliography}\

### 2021-08-11 prova_knit_finale.Rmd
changes on chunks

@@
 
-First, a general point is that the sequence of percentage log returns are different across the three indices.
+First, a general point is that the sequences of percentage log returns are different across the three indices.
 The reasons why this holds can be associated with the index-construction method and the characteristics of the economy and industrial setting it represents.
@@
     x<-rnorm(N,alpha+beta*x+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*x))-2),tau)
-    w1<-w*dnorm(data[t],0,exp(x/2))*dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
+    w1<-w*dnorm(data[t],0,exp(x/2))* dnorm(x,alpha+beta*xprev,tau)/dnorm(x,mean=alpha+beta*xprev+(tau^2)/4*((data[t]^2)*exp(-(alpha+beta*xprev))-2),tau)
 
@@
       epsilon=alpha+beta*x+((tau^2)/4)*((data[t]^2)*exp(-alpha-beta*x)-2)
-      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*exp(-alpha-beta*x)*(1+alpha+beta*x))
+      weight = w*exp((1/(2*tau^2))*((epsilon)^2-(alpha+beta*x)^2)-((data[t]^2)/2)*
+      exp(-alpha-beta*x)*(1+alpha+beta*x))
       k   = sample(1:N,size=N,replace=TRUE,prob=weight)

### 2021-08-11 prova_knit_finale.Rmd
Final commit

@@
 \end{equation*}
+where by the inequality sign we mean that the matrix $Var\big[\psi^{A_t^{n}}\big]$ is positive definite.
 Hence, while the mean is preserved, the variance increases at each iteration.
@@
 In section \ref{Comparing the filters} we briefly compare the series of filtered states that are obtained from different particle filters; moreover, several PFs are compared in terms of computational cost in approximating the exact filtering distribution implied by our SV model.
-Section \ref{Forecast comparison} features a forecast comparison between a SV specification and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
-In section \ref{Realised volatility} we conclude by employing a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \ref{Simulated SV data} for synthetic data.
+Section \ref{Realised volatility} employs a non-parametric measure of realised volatility as a proxy for the latent volatility process, allowing a RMSE and MAE comparison between filters which is similar to that of section \ref{Simulated SV data} for synthetic data.
+In section \ref{Forecast comparison} we conclude by providing a forecast comparison between a SV specification and a model that assumes constant mean and variance for the financial returns, namely a forecasting exercise that would not be possible were it not for particle filtering techniques.
 
@@
 
-```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood of S&P500 daily log returns, CV and SV", fig.height=3, fig.pos='H', fig.width=10, results=F}
+```{r cll_plot, echo=FALSE, fig.align='center', fig.cap="Relative log-likelihood of SP500 daily log returns, CV and SV", fig.height=3, fig.pos='H', fig.width=10, results=F}
 load("SV/cll_plot.Rda")

### 2021-08-14 nuova-final-draft.Rmd
equality invece di inequality

--- 2021-07-20 nuova final draft.Rmd
@@
 \end{definition}
-From the second inequality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$
+From the second equality, it is apparent that $\{(X_t,Y_t)\}$ is a Markov process with initial distribution $[\P_0(dx_0)F_0(x_0,y_0)]$ and transition kernel $P_t(x_{t-1},dx_t)F_t(x_t,dy_t).$ The third equality, instead, shows that, $\{X_t\}$ has a marginal distribution $$\P_T(dx_{0:T})=\P_0(dx_0)\prod_{t=1}^T P_t(x_{t-1},dx_t),$$ meaning that $\{X_t\}$ is marginally distributed as a Markov process with initial distribution $\P_T(dx_{0:T})$ and transition kernel $P_t(x_{t-1},dx_t).$
 

### 2021-08-14 nuova-final-draft.Rmd
sigma algebra

@@
 $x_{0:t}=(x_0, x_1, \dots, x_t)$ for some $t \geq 0.$ The sets of outcomes are denoted by upper case calligraphic letters, e.g.
-$\mathcal{X}, \mathcal{Y}.$  The $\sigma$-algebra
+$\mathcal{X}, \mathcal{Y}.$  A generic $\sigma$-algebra
 of a set $\mathcal{X}$ reads $\mathcal{B}(\mathcal{X}).$ The pair $(\mathcal{X}, \mathcal{B}(\mathcal{X}))$ is a measurable space.

### 2021-08-15 prova_knit_finale.Rmd
correzioni clt

@@
     \begin{split}
-        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(\E_{M_t}(\varphi))+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
+        \hat{\mathcal{V}}_t(\varphi)&:=\mathcal{V}_{t-1}(M_t\varphi)+Var_{\Q_{t-1}M_t}(\varphi) \text{ }\forall t\geq1, \\
         \mathcal{V}_t(\varphi)&:=\hat{\mathcal{V}}_{t}\big( \bar{G}_t(\varphi-\E_{\Q_t}(\varphi)) \big) \text{ }\forall t\geq0.
@@
 The notation $\mid \sigma(X_{t-1}^{1:N})$ means conditioning on having sampled (from the multinomial distribution) $X_{t-1}^{1:N}$.
-Conditional on multinomial sampling, the $X_{t-1}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$.
+Conditional on multinomial sampling, the $X_{t}^n$ are i.i.d., with distribution  $\sum\limits_{n=1}^N W_{t-1}^n M_t(X_{t-1}^n, dx_t)$, that should converge to $\Q_{t-1}M_t$.
 So, it is natural to think that the second element of the sum converges to the variance of $\varphi$ with respect to the distribution $\Q_{t-1}M_t$.
